### 2023-01-24 manuscript.tex
even build tex? [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/51048f35d5fce97aa4194e6426fdd7c704f19199.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/51048f35d5fce97aa4194e6426fdd7c704f19199/checks
https://github.com/stain/2022-fdo-paper/actions/runs/3998374458

This manuscript
(\href{https://stain.github.io/2022-fdo-paper/v/51048f35d5fce97aa4194e6426fdd7c704f19199/}{permalink})
was automatically generated
from \href{https://github.com/stain/2022-fdo-paper/tree/51048f35d5fce97aa4194e6426fdd7c704f19199}{stain/2022-fdo-paper@51048f3}
on January 24, 2023.

\hypertarget{authors}{%
\section*{Authors}\label{authors}}
\addcontentsline{toc}{section}{Authors}

\begin{itemize}
\item
  \textbf{Stian Soiland-Reyes}
  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
  \href{https://github.com/stain}{stain}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
  \href{https://twitter.com/soilandreyes}{soilandreyes}
  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
\item
  \textbf{Carole Goble}
  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
  \href{https://github.com/carolegoble}{carolegoble}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
  Department of Computer Science, The University of Manchester, UK
  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
\item
  \textbf{Paul Groth}
  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
  \href{https://github.com/pgroth}{pgroth}
  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
  \href{https://twitter.com/pgroth}{pgroth}
  Informatics Institute, Faculty of Science, University of Amsterdam, NL
\end{itemize}

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

\textbf{TODO}: Rewrite from notes below to actual abstract.

FAIR Digital Object is an emerging concept from EOSC. This is important. Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape. Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.

What is the message of this paper?

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  These are the overlaps
\item
  This is what FDO is requiring but not commonly deployed
\item
  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
\item
  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
\item
  Lessons for Semantic Web community. What is missing in FDO and in SW.
\item
  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
\item
  Go back to the old SW layer cake. Explains why we picked these frameworks.
\item
  Lessons for SW: Parts of the stack that is less. FDO contributions.
\end{enumerate}

Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.

Have all the ingredients, but not cooking properly.

Systematic through frameworks.

Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?

\hypertarget{sec:introduction}{%
\section{Introduction}\label{sec:introduction}}

The FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures. They have been widely recognised by the research community and policy makers as a goal to strive for. In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures {[}\protect\hyperlink{ref-u8Cy0psL}{2}{]}. The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.

Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures. The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.

In other discourse, Linked Data {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles {[}\protect\hyperlink{ref-GJHYN6xW}{6},\protect\hyperlink{ref-EJDjHrUv}{7}{]}. Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services {[}\protect\hyperlink{ref-99MS5xE0}{9},\protect\hyperlink{ref-J2H8yssV}{10}{]}. While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability {[}\protect\hyperlink{ref-DNTCUjpC}{11},\protect\hyperlink{ref-Pa67pUtR}{12}{]}.

One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality {[}\protect\hyperlink{ref-uxWjmlio}{13},\protect\hyperlink{ref-v0WwFHgT}{14}{]} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves {[}\protect\hyperlink{ref-5V3GqSzR}{15}{]}.

These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data. Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.

Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.

\hypertarget{sec:background}{%
\section{Background and related work}\label{sec:background}}

In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.

\hypertarget{sec:fdo}{%
\subsection{FAIR Digital Object}\label{sec:fdo}}

\hypertarget{sec:next-step-fdo}{%
\subsubsection{Next steps for FDO}\label{sec:next-step-fdo}}

The FAIR Digital Object Forum {[}\protect\hyperlink{ref-DMEBhHID}{16}{]} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}. As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public. As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:

\textbf{FAIR Digital Object Overview and Specifications} {[}\protect\hyperlink{ref-11NpETLq1}{17}{]} is a comprehensive overview of FAIR Digital Object specifications listed below. It serves as a primer that introduces FDO concepts and the remaining documents.

The \textbf{FDO Forum Document Standards} {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions. In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.

The \textbf{FDO Requirement Specifications} {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} is an update of {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} as the foundational definition of FDO. This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations. The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.

The \textbf{Machine actionability} {[}\protect\hyperlink{ref-iJeys0T5}{21}{]} sets out to define what is meant by \emph{machine actionability} for FDOs. \emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar. The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.

\textbf{Configuration Types} {[}\protect\hyperlink{ref-SPd0uEqO}{22}{]} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs. Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}

\textbf{PID Profiles \& Attributes} {[}\protect\hyperlink{ref-ljVViWCl}{24}{]} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile. This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}. This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.

\textbf{Kernel Attributes \& Metadata} {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc. This document expands on RDA Recommendation on PID Kernel Information {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}. It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.

\textbf{Granularity, Versioning, Mutability} {[}\protect\hyperlink{ref-16W9AycSW}{27}{]} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements. The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.

\textbf{DOIP Endorsement Request} {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} is an endorsement of the DOIP v2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} specification as a potential FDO implementation, as it has been applied by several institutions {[}\protect\hyperlink{ref-jRWurmQc}{30}{]}. The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).

\textbf{Upload of FDO} {[}\protect\hyperlink{ref-1H7cmmr69}{31}{]} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve). ResourceSync {[}\protect\hyperlink{ref-eS5y9TRh}{32}{]} is suggested as one type of service to list FDOs. This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).

\textbf{Typing FAIR Digital Objects} {[}\protect\hyperlink{ref-ZFzPxCNB}{33}{]} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose. This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types. Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work. How to define such FDO operations is not specified.

\textbf{Implementation of Attributes, Types, Profiles and Registries} {[}\protect\hyperlink{ref-126uxr5pI}{34}{]} details how to establish FDO registries for types and FDO profiles, with their association with PID systems. This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices. Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} {[}\protect\hyperlink{ref-1CqIZ47pu}{35}{]}.

It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.

The concept of \textbf{FAIR Digital Objects} {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} has been introduced as way to expose research data as active objects that conform to the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}. This builds on the \emph{Digital Object} (DO) concept {[}\protect\hyperlink{ref-11MnuwJ4l}{36}{]}, first introduced in 1995 {[}\protect\hyperlink{ref-3Uqe3fuK}{37}{]} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles. DO was the inspiration for the ITU X.1255 framework {[}\protect\hyperlink{ref-103Hw8H43}{38}{]} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]}.

In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}. The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs). The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing {[}\protect\hyperlink{ref-tz0P3DTC}{40}{]}.

Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}. Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}. The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).

\hypertarget{fdo-approaches}{%
\subsubsection{FDO approaches}\label{fdo-approaches}}

FDO is an evolving concept. A set of FDO Demonstrators {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} highlight how current adapters are approaching implementations of FDO from different angles:

\begin{itemize}
\tightlist
\item
  Building on the Digital Object concept, using the simplified DOIP v2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, see {[}\protect\hyperlink{ref-xuQKRRx5}{41}{]}.} (usually TCP/IP over TLS). The main DOIP operations are retrieving, creating and updating digital objects. These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}. FDO types are registered in the local Cordra instance, where they are specified using JSON Schema {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]} and PIDs are assigned using the Handle system. Several type registries have been established.
\item
  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
\item
  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
\end{itemize}

From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.

\hypertarget{sec:ld}{%
\subsection{From the Semantic Web to Linked Data}\label{sec:ld}}

In order to describe \emph{Linked Data} as it is used today, we'll start with an (opinionated) description of the evolution of its foundation, the \emph{Semantic Web}.

\hypertarget{sec:semweb}{%
\subsubsection{A brief history of the Semantic Web}\label{sec:semweb}}

The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee {[}\protect\hyperlink{ref-Guy7cjVY}{43}{]}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs. The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}. The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.

Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) {[}\protect\hyperlink{ref-16WGmCih8}{44}{]} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format {[}\protect\hyperlink{ref-16wvFH4nh}{45}{]} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).

While triple-based knowledge representations were not new {[}\protect\hyperlink{ref-NxvhwSmG}{46}{]}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs {[}\protect\hyperlink{ref-RRtwPF7Y}{47}{]} are generalised forms of URLs that include locator-less identifiers
  such as ISBN book numbers (URNs). The distinction between locator-full and locator-less identifiers have weakened in recent years {[}\protect\hyperlink{ref-nW4HY8Nq}{48}{]}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) {[}\protect\hyperlink{ref-HqZ3r6J3}{49}{]}. RDF 1.1 formats use Unicode to support IRIs {[}\protect\hyperlink{ref-rqjuXkEf}{50}{]}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to). By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.

The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data). Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term. The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource. By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs. The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.

The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is). More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} {[}\protect\hyperlink{ref-yd2WE9eq}{52}{]} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.

Established Semantic Web services and standards include SPARQL {[}\protect\hyperlink{ref-xww8fb1B}{53}{]} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).

The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins {[}\protect\hyperlink{ref-bMSxa8XW}{54},\protect\hyperlink{ref-Qlrlr2TZ}{55}{]}. In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same. One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.

Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies. Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties). Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) {[}\protect\hyperlink{ref-FDDmgO6s}{57}{]} .

The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources. This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS {[}\protect\hyperlink{ref-191ZkYZMt}{58}{]}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).

With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources. This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps {[}\protect\hyperlink{ref-9CdJnsoJ}{59}{]}. Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication {[}\protect\hyperlink{ref-AGHR9oyr}{60}{]}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data {[}\protect\hyperlink{ref-5SuwLNdE}{61}{]}. Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies {[}\protect\hyperlink{ref-TUmA4KQT}{62}{]} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS {[}\protect\hyperlink{ref-ZwAcGQKY}{63}{]}, OWL {[}\protect\hyperlink{ref-1p4IWJpI}{64}{]}, OBO {[}\protect\hyperlink{ref-1DalZuKct}{65}{]}, SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]}), hash vs slash, HTTP status codes and PID redirection strategies {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.

\hypertarget{sec:ld-web}{%
\subsubsection{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}

The \textbf{Linked Data} concept {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} was kickstarted as a set of best practices {[}\protect\hyperlink{ref-OGXYtGZ8}{66}{]} to bring the Web aspect back into focus. Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers. This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.

Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data. In addition to several such embedded \emph{microformats} (Open Graph {[}\protect\hyperlink{ref-oN9D5LkD}{67}{]}, RDFa {[}\protect\hyperlink{ref-FKIF6ApB}{68}{]}, Microdata {[}\protect\hyperlink{ref-14Utscfe4}{69}{]}) we find JSON-LD {[}\protect\hyperlink{ref-X5SWPqSU}{70}{]} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications. JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites {[}\protect\hyperlink{ref-j3SqQLBR}{71}{]}.

Recently there has been a renewed emphasis to improve the \emph{Developer Experience} {[}\protect\hyperlink{ref-BFMoe8FC}{72}{]} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL {[}\protect\hyperlink{ref-GDrm9dgX}{73}{]} or ShEx {[}\protect\hyperlink{ref-YUzogaV}{74}{]}) can be used to validate RDF Data {[}\protect\hyperlink{ref-S5BI6Mwx}{75},\protect\hyperlink{ref-ZeOAPNHa}{76}{]} before consuming it programmatically, or reshaping data to fit other models. While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts {[}\protect\hyperlink{ref-o596f2JM}{77}{]}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.

A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}. One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies {[}\protect\hyperlink{ref-MhsCt4P3}{78}{]}, the use of Linked Data as part of REST Web APIs {[}\protect\hyperlink{ref-eS6a6tlz}{79}{]} , and as shown by the big uptake by publishers to annotate the Web using schema.org {[}\protect\hyperlink{ref-117tX9MEG}{81}{]}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.

\hypertarget{sec:comparing}{%
\section{Comparing FDO and existing approaches}\label{sec:comparing}}

To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  An Interoperability Framework and Distributed Platform for Fast Data Applications {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
\item
  The FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}, validated against its implementations for completeness.
\item
  A Comparison Framework for Middleware Infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
\item
  Cross-checks against RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-UzQhqk0M}{84}{]} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
\item
  EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
\end{enumerate}

The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).

Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices. For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} against Linked Data in general.

\hypertarget{sec:interoperability-compare}{%
\subsection{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}

The Interoperability Framework for Fast Data Applications {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).

We have chosen to investigate using this framework as it covers the higher levels of the OSI Model {[}\protect\hyperlink{ref-8ZRGFa6y}{85}{]} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles. In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]}.
\label{tbl:fdo-web-interoperability-framework}}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
\emph{Quality}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO w/ DOIP
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Web w/ Linked Data
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
\emph{Quality}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO w/ DOIP
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Web w/ Linked Data
\end{minipage} \\
\midrule
\endhead
\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types. FDO encourages research communities to develop common type registries to be shared across instances. In current DOIP practice, each service have their own types, attributes and operations. The wider symbiosis is consistent use of PIDs. & The Web is loosely coupled and encourages collaboration and linking by URL. In practice, REST APIs {[}\protect\hyperlink{ref-174AwcFUL}{86}{]} end up being mandated centrally by dominant (often commercial) providers {[}\protect\hyperlink{ref-11VxBOeZy}{87}{]}, which clients are required to use as-is with special code per service. Use of Linked Data enables common tooling and semantic mapping across differences. \\
\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP). \texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs). Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}). & ``Follow your nose'' crawler navigation, which may lead to frequent dead ends. Operational composition, typically within a single API provider, documented by OpenAPI 3 {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]}, schema.org Actions {[}\protect\hyperlink{ref-dKAekUjL}{89}{]}, WSDL/SOAP {[}\protect\hyperlink{ref-LmudecyN}{90}{]}, but frequently just as human-readable developer documentation/examples. \\
\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing. Every FDO has a type. Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}. No single type semantic, Type FDOs can link to existing vocabularies \& ontologies. JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} & Lightweight HTTP semantics for authenticity/navigation. Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata. Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit. \\
\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation} & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema. Multiple byte stream attachments of any media type. & Textual HTTP headers (including any signposting), single byte stream of any media type, e.g.~Linked Data formats (JSON-LD, Turtle, RDF/XML) or embedded in document (HTML with RDFa, JSON-LD or Microdata). XML was previously the main syntax used by APIs, JSON is now dominant. \\
\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}. Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable). Unique DOIP protocol can be hard to add to existing Web application frameworks, although proxy services have been developed (e.g.~B2SHARE adapter). & HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to (for static content) multi-cloud Content-Delivery Networks (CDN). Current scalable cloud technologies for Web hosting may not support HTTP features previously seen as important for Semantic Web, e.g.~content negotiation and semantic HTTP status codes. \\
\bottomrule
\end{longtable}

Based on the analysis shown in table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:

The Web has already showed us how one can compose workflows of hetereogeneous Web Services {[}\protect\hyperlink{ref-39K3yHT3}{91}{]}. However, this is mostly done via developer or human interaction {[}\protect\hyperlink{ref-iYh1wFBt}{92}{]}. Similiarly, FDO does not enable automatic composition because operation semantics are not well defined. There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.

A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics. Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON. On the semantic front, FDO requires that every object have a well-defined type and structured form. This is clearly not the case on the Web.

In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed. These have shown interoperability. The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles {[}\protect\hyperlink{ref-hxtgCGjY}{93}{]} to embrace openness, transparency, and broad consensus. In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference {[}\protect\hyperlink{ref-jrZe6Esu}{94}{]}. This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP {[}\protect\hyperlink{ref-p7nwRj8E}{95}{]}.

It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary. For example, DOIs are mostly resolved on the web {[}\protect\hyperlink{ref-1H9iQhQYq}{96}{]} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail {[}\protect\hyperlink{ref-68fzID2k}{97}{]}.

\hypertarget{mapping-of-metamodel-concepts}{%
\subsubsection{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}

The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics {[}\protect\hyperlink{ref-8HJqcF1Q}{98}{]}.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2923}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2769}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]} to equivalent concepts for FDO and Web.
\label{tbl:metamodel-concepts}}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
Metamodel concept
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO/DOIP concept
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Web/HTTP concept
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
Metamodel concept
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO/DOIP concept
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Web/HTTP concept
\end{minipage} \\
\midrule
\endhead
Resource & FDO/DO & Resource \\
Service & DOIP service & Server/endpoint \\
Transaction & (not supported) & Conditional requests, \texttt{409\ Conflict} \\
Process & Extended operations & (primarily stateless), \texttt{100\ Continue}, \texttt{202\ Accepted} \\
Operation & DOIP Operation & Method, query parameters \\
Request & DOIP Request & Request \\
Response & DOIP Response & Response \\
Message & Segment, \texttt{requestId} & Message, Representation \\
Channel & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures. & TCP/IP, TLS, UDP \\
Protocol & DOIP 2.0, ++ & HTTP/1.1, HTTP/2, HTTP/3 \\
Link & PID/Handle & URL \\
\bottomrule
\end{longtable}

From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology. Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \ref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.

\hypertarget{sec:doip-fdo-compare}{%
\subsection{Assessing FDO implementations}\label{sec:doip-fdo-compare}}

The FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} sets out recommendations for FDO implementations. In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} and using Linked Data Platform {[}\protect\hyperlink{ref-17OHlMRQA}{99}{]}, as proposed by {[}\protect\hyperlink{ref-7szz7dwO}{100}{]}.

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
\caption{Checking FDO guidelines {[}\protect\hyperlink{ref-yygVPoL0}{19},\protect\hyperlink{ref-RwvirqWg}{20}{]} against its current implementations as DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} and Linked Data Platform (LDP) {[}\protect\hyperlink{ref-7szz7dwO}{100}{]}, with suggestions for required additions.
\label{tbl:fdo-checks}}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\centering
\textbf{FDO Guideline}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
DOIP 2.0
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO suggestions
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Linked Data Platform
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
LDP suggestion
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\centering
\textbf{FDO Guideline}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
DOIP 2.0
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO suggestions
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Linked Data Platform
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
LDP suggestion
\end{minipage} \\
\midrule
\endhead
G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017. & Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years. \texttt{http://} URIs replaced by \texttt{https://}. & Keep flexibility of RDF serialisation formats which may change more frequently \\
G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories. DOIP relatively unknown, in effect only one implementation. & Further promote DOIP and justify its benefits. Build tutorials and OSI open source implementations. Standardise DOIP-over-HTTP alternative. & JSON-LD used by half of all websites {[}\protect\hyperlink{ref-j3SqQLBR}{71}{]}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering. Example-driven documentation. \\
G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples. & Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled. \\
G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup. Specify input/output expectations formally (e.g.~JSON Schema). & HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls. \\
G5: \emph{abstraction principle} & Handle PIDs as abstraction base. DOIP operations can use any transport protocol. & Document transport protocols as FDOs, recommend which transport to use. & URI as abstraction base. Does not specify PID requirements. & Give stronger deployment recommendations. \\
G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type. Unclear when metadata field is a PID or plain text. & Make datatype of fields explicit to support navigation. & Machine-navigation through URIs via properties and types. Unclear when URI should be followed or is just identifier, but always distinct from plain text. & \\
G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}). & Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608). & HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly. Unsupported methods reported using LDP constraints to human-readable text. & Declare supported methods in advance, e.g.~OpenAPI {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]} \\
G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations. No FDOF implementations. & Develop demonstrator of FDOF usage based on existing LDP server. \\
G9: \emph{standard compliance} & Handle {[}\protect\hyperlink{ref-2gXY2Nwh}{101}{]}, DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}. FDO requirements not standardised yet. & Formalise standard process of FDO requirements {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} & HTTP, LDP. FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
FDOF1: \emph{PID as basis} & Extensive use of Handle system. & Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve {[}\protect\hyperlink{ref-ljVViWCl}{24}{]}. Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs. & HTTP URLs as basis for identifiers, but they are frequently not persistent. & Add strong guidance for PID services like w3id and persistence policies. \\
FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata. Datacite DOIs content-negotiate to give registered metadata. & Add FAIR Signposting at PID provider for minimal PID record \\
FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure. No method/attribute defined for separate metadata, only directly in PID Record. Unclear meaning of multiple items and bytestream chunks. & Clarify expectations for multiple items. Recommend chunks to not be used. & URIs resolvable by default. Multiple ways to resolve metadata, unclear preference. & Add FAIR Signposting and preference order. \\
FDOF4: \emph{Additional attributes} & Freetext attribute keys. Attributes should be defined for FDO type (?). & Require that attribute keys should be PIDs (or have predefined mapping to PIDs). Explicitly allow attributes not already defined in type. & All attributes individually identified. Any Linked Data attributes can be used by URI or with mapped prefix. & Clarify type expectations of required/recommended/optional attributes. \\
FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles. & Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs. Clarify that operations can be mapped to protocol directly. & CRUD operations used directly in HTTP (e.g.~\texttt{PUT}). Unclear how to provide PID for additional operations. & Specify how additional operations should be called over HTTP. \\
FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}. & Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns {[}\protect\hyperlink{ref-i9Ang0rM}{102}{]} are used instead. & Document how operation resources can be discovered from an LPD container. Document search API. \\
FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type. PR-TypingFDOs leaves this open. & Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type. Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required. & Always make LDP types explicit in FDO profile. \\
FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record. Separate Metadata FDO need custom property. & Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs. & Assertions are always with semantics, using RDF vocabularies. Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common. & Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations. Metadata would need to be bundled within custom FDO types/attributes. Unclear how levels are separated within single FDO representation (need FDOF8?). & Declare which metadata are expected within response attribute or within FDO object. Require PIDs for custom attributes. Define how alternate metadata levels can be represented separately. & Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects. & Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema. & Require types to be FDOs with registered PIDs, implement shared types. & Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} {[}\protect\hyperlink{ref-CNP39tah}{103}{]} & Rather document better how individual ad-hoc schemas can be started for prototypes. \\
FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP. Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g. {[}\protect\hyperlink{ref-6LvydTsC}{104}{]}). & & LDP Containers required by specification, also user-created (eg. \texttt{BasicContainer}). & Clarify relation to other collections like DCAT 3 {[}\protect\hyperlink{ref-1EW4hLN8U}{105}{]}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE {[}\protect\hyperlink{ref-gtGuSevK}{106}{]} \\
FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP. \texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common. No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
\bottomrule
\end{longtable}

Note that the draft update to FDO specification {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.

A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines. Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals. The DOIP Endorsement {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}. Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.

From our evaluation, we can observe:

\begin{itemize}
\tightlist
\item
  G1 and G2 call for stability and trustworthiness. While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
\item
  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data. Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous. In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
\item
  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}. These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
\item
  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
\item
  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
\item
  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use. Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} {[}\protect\hyperlink{ref-9T4j3N4e}{107}{]}
\item
  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
\item
  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
\item
  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation. In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
\end{itemize}

\hypertarget{sec:middleware}{%
\subsection{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}

In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects. As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.

We argue that the Web, with its now ubiquitous use of REST API {[}\protect\hyperlink{ref-174AwcFUL}{86}{]}, can be compared as a similar global middleware. Note that while early moves for developing Semantic Web Services {[}\protect\hyperlink{ref-WzPECPDP}{108}{]} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} {[}\protect\hyperlink{ref-11dInoyVF}{109}{]}.

For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}) and Web technologies (using Linked Data) as middleware infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]}
\label{tbl:fdo-web-middleware}}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
\emph{Quality}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO w/ DOIP
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Web w/ Linked Data
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
\emph{Quality}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO w/ DOIP
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Web w/ Linked Data
\end{minipage} \\
\midrule
\endhead
\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints. Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server. PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}. & The Web is inherently open and made by cross-linked URLs. Participation requires DNS domain purchase (many free alternatives also exists). PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects. \\
\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up. HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend. \\
\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second {[}\protect\hyperlink{ref-18P9txgeB}{110}{]}. DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}). Multiple connections possible. Setup is typically through TCP and TLS which adds latency. & HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks {[}\protect\hyperlink{ref-F8VZ86hu}{111}{]}. HTTP 1 connections are serial and reusable, and concurrent connections is common. HTTP/2 adds asynchronous responses and multiplexed streams {[}\protect\hyperlink{ref-KbbW0kGT}{112}{]} but still has TCP+TLS startup costs. For reduced latency {[}\protect\hyperlink{ref-CJuBupCW}{113}{]}, HTTP/3 {[}\protect\hyperlink{ref-Amit86Tp}{114}{]} use QUIC {[}\protect\hyperlink{ref-1GWvyEmUf}{115}{]}) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic. \\
\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint). FDOs should provide the mandatory kernel metadata fields. FDOs of the same declared type typically share additional attributes (although that schema may not be declared). DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions. & Each URL accessed separately. Common HTTP headers provide basic metadata, although it is often not reliable. A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types. Metadata is not always machine findable, may need pre-agreed API URI Templates {[}\protect\hyperlink{ref-11hxwwuRt}{116}{]}, content-negotiation {[}\protect\hyperlink{ref-n2f3TfoI}{117}{]} or FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}. \\
\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup. Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers. & Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs. Multiple locations possible in Handle system, can expose geo-info. & PIDs and URL redirects. DNS aliases and IP routing can hide location. Geo-localised servers common for large cloud deployments. \\
\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures. FDO kernel metadata can include checksum and date. & HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications. Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource. \\
\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}. Custom codes uncommon. \\
\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently). No redirection from DOIP service. & HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints. \\
\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects. There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}). & URLs are not required to persist, although encouraged {[}\protect\hyperlink{ref-rbG9uRKw}{118}{]}. Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}. An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}. Memento {[}\protect\hyperlink{ref-DNjcMi4a}{119}{]} expose versioned snapshots. WebDAV \texttt{VERSION-CONTROL} method {[}\protect\hyperlink{ref-hRUsaTiz}{120}{]} (used by SVN). \\
\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP. Internal synchronisation possible in backend for Extended operations. & Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource. WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} {[}\protect\hyperlink{ref-xvMAyRAc}{121}{]} with \texttt{LOCK} and \texttt{UNLOCK} methods. \\
\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references. In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution. & The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space. In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend. \\
\textbf{Encapsulation}: \emph{separate interface from implementation. Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation. FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP). Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3. \texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade {[}\protect\hyperlink{ref-1B16WfR7q}{122}{]} to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}. Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns {[}\protect\hyperlink{ref-iME5l1kk}{123}{]} and PIDs via Persistent URLs (PURL). \\
\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited). FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain). Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]}) & Syntactically Media Type with multiple suffixes {[}\protect\hyperlink{ref-nlH2rm1s}{124}{]} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) {[}\protect\hyperlink{ref-nCzRO0pK}{125}{]}. In metadata, semantic type systems (RDFS {[}\protect\hyperlink{ref-ZwAcGQKY}{63}{]}, OWL2 {[}\protect\hyperlink{ref-1p4IWJpI}{64}{]}, SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]}). OpenAPI 3 {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]} \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}. XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} {[}\protect\hyperlink{ref-11oVTMsuZ}{126}{]}, JSON \texttt{\$schema} {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]}), JSON-LD \texttt{@context} {[}\protect\hyperlink{ref-by6CfOmv}{127}{]}. Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge. \\
\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined. Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} {[}\protect\hyperlink{ref-KbbW0kGT}{112}{]}, Web Sockets {[}\protect\hyperlink{ref-qgqADk45}{128}{]}, Linked Data Notifications {[}\protect\hyperlink{ref-zlGYiuWC}{129}{]}, AtomPub {[}\protect\hyperlink{ref-IbaebabD}{130}{]}, SWORD {[}\protect\hyperlink{ref-7DGllKzG}{131}{]}, Micropub, more typically ad-hoc jobs/futures REST resources \\
\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} {[}\protect\hyperlink{ref-HE7Ikwwl}{132}{]}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates {[}\protect\hyperlink{ref-11hxwwuRt}{116}{]}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State {[}\protect\hyperlink{ref-174AwcFUL}{86}{]}, an important element of the REST architectural style.} incl.~Hydra {[}\protect\hyperlink{ref-13PaKBrZe}{133}{]}, schema.org Actions {[}{[}\protect\hyperlink{ref-dKAekUjL}{89}{]}), JSON HAL {[}\protect\hyperlink{ref-nyrPYHoP}{134}{]} \& Link headers (RFC8288) {[}\protect\hyperlink{ref-ozdWB3O7}{135}{]} \\
\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO. DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 {[}\protect\hyperlink{ref-tIkSXrb5}{136}{]} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) {[}\protect\hyperlink{ref-PL512B0X}{137}{]}, MPEG-DASH (ISO/IEC 23009-1:2019) {[}\protect\hyperlink{ref-ikO7e0mh}{138}{]} \\
\bottomrule
\end{longtable}

Based on the analysis in Table \ref{tbl:fdo-web-middleware}, we make the following observations:

\begin{itemize}
\tightlist
\item
  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]} supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects). Multiplexing was removed for the much simplified DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}. Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel. It is unclear what is the purpose of splitting a binary into chunks on a channel which no longer can be multiplexed and the only property of a chunk is its size\footnote{Although it is possible with \texttt{0.DOIP/Op.Retrieve} to request only particular individual elements of an DO (e.g.~one file), unlike HTTP's \texttt{Range} request, it is not possible to select individual chunks of an element's bytestream.}.
\item
  HTTP has strong support for scalability and caching, but this mostly assumes read-operations from static resources. FDO has no view on immutability or validity of retrieved objects, but this should be taken into consideration to support large-scale usage.
\item
  HTTP optimisations for performance (e.g.~HTTP/2, multiplexing) is largely used for commercial media distribution (e.g.~Netflix), and not commonly used by providers of FAIR data
\item
  Cloud deployment of Web applications give many middleware benefits (Scalability, Distribution, Access transparancy, Location transparancy) -- it is unclear how DOIP as a custom protocol would perform in a cloud setting as most of this infrastructure assumes HTTP as the protocol.
\item
  Programmatically the Web is rather unstructured as middleware, as there are many implementation choices. Usually it is undeclared what to expect for a given URI/service, and programmers follow documented examples for a particular service rather than automated programmatic exploration across providers. This mean one can consider the Web as an ecosystem of smaller middlewares with commonalities.
\item
  Many providers of FAIR Linked Data also provide programmatic REST API endpoints, e.g.~\href{https://www.uniprot.org/help/programmatic_access}{UNIPROT}, \href{https://chembl.gitbook.io/chembl-interface-documentation/web-services}{ChEMBL}, but keeping the FAIR aspects such as retrieving metadata in such a scenario may require combining different services using multiple formats and identifier conventions.
\end{itemize}

\hypertarget{sec:fair-compare}{%
\subsection{Assessing FDO against FAIR}\label{sec:fair-compare}}

In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} {[}\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}{]}.

Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform {[}\protect\hyperlink{ref-7szz7dwO}{100}{]} comply with the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}. Here we've used the RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-1GrKTFaK2}{139}{]} as it has decomposed the FAIR principles to a structured list of FAIR indicators {[}\protect\hyperlink{ref-UzQhqk0M}{84}{]}, importantly considering \emph{Data} and \emph{Metadata} separately. In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
\caption{Assessing RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-UzQhqk0M}{84},\protect\hyperlink{ref-1GrKTFaK2}{139}{]} (first 2 columns) against the FDO guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}, FDO implemented with the protocol DOIPv2 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, Linked Data Platform (LDP) {[}\protect\hyperlink{ref-7szz7dwO}{100}{]} and examples from Linked Data practices in general. (--- indicates \emph{Unspecified}, may be possible with additional conventions)
\label{tbl:fair-data-maturity-model}}\tabularnewline
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
FAIR ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Indicator
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO guidelines
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO/DOIP
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO/LDP
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Linked Data examples
\end{minipage} \\
\midrule
\endfirsthead
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
FAIR ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Indicator
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO guidelines
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO/DOIP
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
FDO/LDP
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
Linked Data examples
\end{minipage} \\
\midrule
\endhead
RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1). Handle, DOI. & FDOF-IR (Identifier Record). PID can be any URI & ``Cool'' URIs {[}{[}\protect\hyperlink{ref-rbG9uRKw}{118}{]};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}. FAIR Signposting. \\
RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata. Unclear how to link to additional metadata. & FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting. Embedded in landing page (RDFa). \\
RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record. & PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries). Handle registry only searchable by PID. & Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g. {[}\protect\hyperlink{ref-1Gsq23e51}{140}{]}). Indexing by search engines if exposing HTML w/schema.org. \\
RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata. \texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3). Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP {[}\protect\hyperlink{ref-8HJqcF1Q}{98}{]} & HTTP/HTTPS, FTP (now less common), GridFTP {[}\protect\hyperlink{ref-18LsXvKGJ}{141}{]} (for large data), ARK {[}\protect\hyperlink{ref-voDQ6diM}{142}{]} \\
RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol {[}\protect\hyperlink{ref-I9Kx0Hjy}{143}{]}. One server implementation {[}\protect\hyperlink{ref-1CvYRVhy2}{144}{]}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved {[}\protect\hyperlink{ref-xVJMmGVr}{145}{]} as an open source license -- it includes usage restrictions and requires Service Agreements. It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}. One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source). & \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation. \href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}. & DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}. \\
RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF? & FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) {[}\protect\hyperlink{ref-1BlxdxOwa}{146}{]} \\
RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs {[}\protect\hyperlink{ref-o4h92LpS}{147}{]}), \texttt{rdfs:seeAlso}, FAIR signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]} \texttt{describedby} \\
RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles {[}\protect\hyperlink{ref-QpDmajc4}{148}{]}, \href{https://schema.org/Role}{schema.org/Role} \\
RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role {[}\protect\hyperlink{ref-10HA0mMBx}{149}{]}, \texttt{prov:specializationOf} {[}\protect\hyperlink{ref-ZUgwvAHE}{150}{]}, OAI-ORE Proxy {[}\protect\hyperlink{ref-ID1pTGry}{151}{]}) \\
RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required. Kernel metadata attributes desired {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} but not assigned PIDs yet. & Unspecified. Multiple metadata records can allow multiple semantic profiles. & Large number of general and domain-specific vocabularies can make it hard to find relevant attributes. Rough consensus on kernel metadata: schema.org {[}\protect\hyperlink{ref-10pzKCEUE}{152}{]}, Dublin Core Terms {[}\protect\hyperlink{ref-wAbZpOsr}{153}{]}, DCAT {[}\protect\hyperlink{ref-lnH0hEuh}{154}{]}, FOAF {[}\protect\hyperlink{ref-XmUY0cZB}{155}{]} \\
RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} {[}\protect\hyperlink{ref-lnH0hEuh}{154}{]} \\
RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods). PID Kernel attributes? Unspecified W3C PROV-O, PAV & & \\
RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O {[}\protect\hyperlink{ref-9T4j3N4e}{107}{]}, PAV {[}\protect\hyperlink{ref-1E4OfHyhI}{156}{]}, Dublin Core Terms {[}\protect\hyperlink{ref-hKblceo6}{157}{]} \\
RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen {[}\protect\hyperlink{ref-zZTtJruI}{158}{]}) & --- & Common, e.g.~DCAT 2 {[}\protect\hyperlink{ref-zt8AAjcZ}{159}{]}, BioSchemas {[}\protect\hyperlink{ref-1F8S9FeHf}{160}{]} \\
RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal {[}\protect\hyperlink{ref-1Gsq23e51}{140}{]}, Darwin Core {[}\protect\hyperlink{ref-9jaVzCbL}{161}{]} \\
RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
\bottomrule
\end{longtable}

From this evaluation we observe:

\begin{itemize}
\tightlist
\item
  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
\item
  FDO/DOIP are stronger on identifier indicators
\item
  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} has currently only a couple of implementations and is expressed informally. The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
\item
  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
\item
  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
\item
  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs. {[}\protect\hyperlink{ref-126uxr5pI}{34}{]} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised. Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
\item
  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata. For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information? FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource. In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
\item
  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications. \texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known. g
\end{itemize}

\hypertarget{eosc-interoperability-framework}{%
\section{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}

\textbf{TODO}: Introduce EOSC IF

The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} in section 3.6 recommends:

\begin{tablenos:no-prefix-table-caption}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0833}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
\toprule
\begin{minipage}[b]{\linewidth}\raggedright
Layer
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Recommendation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
FDO
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Linked Data
\end{minipage} \\
\midrule
\endhead
Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
Technical & Easy SLAs for service providers & Unspecified & None \\
Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
Semantic & Minimal metadata model for federated discovery & Kernel metadata {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} & DCAT, ++ \\
Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
Legal & List recommended licenses & Undefined & Best practice recommendations \\
Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
Legal & GDPR compliance for personal data & Undefined & Undefined \\
Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
Legal & Harmonised terms-of-use & Undefined & Undefined \\
Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
\bottomrule
\end{longtable}

\end{tablenos:no-prefix-table-caption}

Observations:

\begin{itemize}
\tightlist
\item
  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
\item
  Technical aspects highlighted by EOSC IF
\item
  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data. There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
\item
  FDO principles have strong recommendations for community development of organisational aspects.
\item
  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
\end{itemize}

\hypertarget{sec:discussion}{%
\section{Discussion}\label{sec:discussion}}

\textbf{TODO}

\begin{itemize}
\tightlist
\item
  Ramnifications of ide
\item
  Finish tables
\item
  Bullet points per table
\item
  Send in Google Docs
\end{itemize}

\hypertarget{what-does-it-mean-for-linked-data}{%
\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}

The FAIR Digital Object approach raises many important points for Linked Data practictioners.
At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake {[}\protect\hyperlink{ref-kYjzRJ0H}{162}{]}.
However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.

While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.

We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\ldots{}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-6DjakjNS}{}}%
\CSLLeftMargin{1. }
\CSLRightInline{\textbf{The FAIR Guiding Principles for scientific data management and stewardship}
\CSLBlock{Mark D Wilkinson, Michel Dumontier, IJsbrand Jan Aalbersberg, Gabrielle Appleton, Myles Axton, Arie Baak, Niklas Blomberg, Jan-Willem Boiten, Luiz Bonino da Silva Santos, Philip E Bourne, \ldots{} Barend Mons} \emph{Scientific Data} (2016-03-15)
\CSLBlock{DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/26978244}{26978244} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4792175}{PMC4792175}}}

\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
\CSLLeftMargin{2. }
\CSLRightInline{\textbf{Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud}
\CSLBlock{Barend Mons, Cameron Neylon, Jan Velterop, Michel Dumontier, Luiz Olavo Bonino da Silva Santos, Mark D Wilkinson} \emph{Information Services \& Use} (2017-03-07)
\CSLBlock{DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{isu\}-170824}}}

\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
\CSLLeftMargin{3. }
\CSLRightInline{\textbf{EOSC interoperability framework}
\CSLBlock{Oscar Corcho, Magnus Eriksson, Krzysztof Kurowski, Milan Ojsteršek, Christine Choirat, Mark van de Sanden, Frederik Coppens} \emph{Publications Office of the EU} (2021-02-05) \url{https://doi.org/10.2777/620649}
\CSLBlock{DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}}}

\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
\CSLLeftMargin{4. }
\CSLRightInline{\textbf{FAIR principles and digital objects: Accelerating convergence on a data infrastructure}
\CSLBlock{Erik Schultes, Peter Wittenburg} \emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers} (2019)
\CSLBlock{DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}} · ISBN: 978-3-030-23583-3}}

\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
\CSLLeftMargin{5. }
\CSLRightInline{\textbf{Linked data - the story so far}
\CSLBlock{Christian Bizer, Tom Heath, Tim Berners-Lee} \emph{International journal on Semantic Web and information systems} (2009-07) \url{http://services.igi-global.com/resolvedoi/resolve.aspx?doi=10.4018/jswis.2009081901}
\CSLBlock{DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}}}

\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
\CSLLeftMargin{6. }
\CSLRightInline{\textbf{Assessing FAIR data principles against the 5-star open data principles}
\CSLBlock{Ali Hasnain, Dietrich Rebholz-Schuhmann} \emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers} (2018) \url{http://link.springer.com/10.1007/978-3-319-98192-5/_60}
\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60} · ISBN: 978-3-319-98191-8}}

\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
\CSLLeftMargin{7. }
\CSLRightInline{\textbf{FAIR data points supporting big data interoperability}
\CSLBlock{Luiz Olavo Bonino Da Silva Santos, Mark D Wilkinson, Arnold Kuzniar, Rajaram Kaliyaperumal, Mark Thompson, Michel Dumontier, Kees Burger} \emph{Enterprise interoperability in the digitized and networked factory of the future} (2016) \url{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}
\CSLBlock{ISBN: 9781847040442}}

\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
\CSLLeftMargin{8. }
\CSLRightInline{\textbf{The semantic web identity crisis: In search of the trivialities that never were}
\CSLBlock{Ruben Verborgh, Miel Vander Sande} \emph{Satya Widya} (2020-01-31)
\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/sw-190372}}}

\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
\CSLLeftMargin{9. }
\CSLRightInline{\textbf{Conclusion and future challenges}
\CSLBlock{Albert Meroño-Peñuela, Pasquale Lisena, Carlos Martínez-Ortiz} \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers} (2021)
\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7} · ISBN: 978-3-031-00789-7}}

\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
\CSLLeftMargin{10. }
\CSLRightInline{\textbf{An analysis of public REST web service apis}
\CSLBlock{Andy Neumann, Nuno Laranjeiro, Jorge Bernardino} \emph{IEEE Transactions on Services Computing} (2021-07-01)
\CSLBlock{DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/tsc.2018.2847344}}}

\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
\CSLLeftMargin{11. }
\CSLRightInline{\textbf{Linked data, JSON-LD and the semantics of cultural and scientific heritage}
\CSLBlock{Ovidiu Turcoane} \emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} (2014-09-30)
\CSLBlock{DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}}}

\leavevmode\vadjust pre{\hypertarget{ref-Pa67pUtR}{}}%
\CSLLeftMargin{12. }
\CSLRightInline{\textbf{F*** workflows: when parts of FAIR are missing}
\CSLBlock{Sean R Wilkinson, Greg Eisenhauer, Anuj J Kapadia, Kathryn Knight, Jeremy Logan, Patrick Widener, Matthew Wolf} \emph{arXiv} (2022) \url{https://doi.org/grk3v5}
\CSLBlock{DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}}}

\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
\CSLLeftMargin{13. }
\CSLRightInline{\textbf{Do Developers Make Unbiased Decisions? - The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components}
\CSLBlock{Anisa Stefi} \emph{University of Münster, Münster, Germany} (2015) \url{https://doi.org/grcpwg}
\CSLBlock{DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}}}

\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
\CSLLeftMargin{14. }
\CSLRightInline{\textbf{To develop or to reuse? Two perspectives on external reuse in software projects}
\CSLBlock{Anisa Stefi, Thomas Hess} \emph{Software business} (2015)
\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18} · ISBN: 978-3-319-19592-6}}

\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
\CSLLeftMargin{15. }
\CSLRightInline{\textbf{Web data apis over SPARQL}
\CSLBlock{Albert Meroño-Peñuela, Pasquale Lisena, Carlos Martínez-Ortiz} \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers} (2021)
\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3} · ISBN: 978-3-031-00789-7}}

\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
\CSLLeftMargin{16. }
\CSLRightInline{\textbf{FAIR Digital Objects Forum \textbar{}} \url{https://fairdo.org/}}

\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
\CSLLeftMargin{17. }
\CSLRightInline{\textbf{FAIR digital object overview and specifications}
\CSLBlock{I Anders, Ch Blanchi, Daan Broeder, M Hellström, Sh Islam, Th Jejkal, L Lannom, K Peters, R Quick, A Schlemmer, \ldots{} Carlo Zwölf} \emph{FDO Forum} (2022-11-19) \url{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
\CSLLeftMargin{18. }
\CSLRightInline{\textbf{FDO Forum Document Standards}
\CSLBlock{C Weiland, U Schwardmann, P Wittenburg, C Kirkpatrick, R Hanisch, Z Trautt} \emph{FDO Forum} (2022-01-29) \url{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
\CSLLeftMargin{19. }
\CSLRightInline{\textbf{FDO Forum FDO requirement specifications}
\CSLBlock{G Strawn, P Wittenburg, St Soiland-Reyes, K Peters, L Lannom, U Schwardmann, Ch Blanch} \emph{FDO Forum} (2022-08-22) \url{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
\CSLLeftMargin{20. }
\CSLRightInline{\textbf{FAIR digital object framework}
\CSLBlock{Luiz Bonino, Oeter Wittenburg, Bonnie Carroll, Alex Hardisty, Mark Leggott, Carlo Zwölf} \emph{FDOF technical implementation guideline} (2019-11-22) \url{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}}

\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
\CSLLeftMargin{21. }
\CSLRightInline{\textbf{FDO machine actionability}
\CSLBlock{Claus Weiland, Sharif Islam, Daan Broder, Ivonne Anders, Peter Wittenburg} \emph{FDO Forum} (2022-06-11) \url{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
\CSLLeftMargin{22. }
\CSLRightInline{\textbf{FDO configuration types}
\CSLBlock{Larry Lannom, Karsten Peters-von Gehlen, Ivonne Anders, Andreas Pfeil, Alexander Schlemmer, Zach Trautt, Peter Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
\CSLLeftMargin{23. }
\CSLRightInline{\textbf{FAIR Signposting Profile}
\CSLBlock{Herbert Van de Sompel, Martin Klein, Shawn Jones, Michael L Nelson, Simeon Warner, Anusuriya Devaraju, Robert Huber, Wilko Steinhoff, Vyacheslav Tykhonov, Luc Boruta, \ldots{} Mark Wilkinson} (2022-07-27) \url{https://signposting.org/FAIR/}}

\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
\CSLLeftMargin{24. }
\CSLRightInline{\textbf{FDO PID profiles \& attributes}
\CSLBlock{Ivonne Anders, Maggie Hellström, Sharif Islam, Thomas Jejkal, Larry Lannom, Ulrich Schwardmann, Peter Wittenburg} \emph{FDO Forum} (2022-06-07) \url{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
\CSLLeftMargin{25. }
\CSLRightInline{\textbf{RDA Recommendation on PID Kernel Information}
\CSLBlock{Tobias Weigel, Beth Plale, Mark Parsons, Gabriel Zhou, Yu Luo, Ulrich Schwardmann, Robert Quick, Margareta Hellström, Kei Kurakawa} \emph{Research Data Alliance} (2018)
\CSLBlock{DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}}}

\leavevmode\vadjust pre{\hypertarget{ref-Fi62cJAo}{}}%
\CSLLeftMargin{26. }
\CSLRightInline{\textbf{FDO -- kernel attributes \& metadata}
\CSLBlock{FDO-TSIG Working Group, FDO-BIG Working Group} \emph{FDO Forum} (2022-08-28) \url{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
\CSLLeftMargin{27. }
\CSLRightInline{\textbf{FDO -- granularity, versioning, mutability}
\CSLBlock{FDO-TSIG Working Group} \emph{FDO Forum} (2022-08-26) \url{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
\CSLLeftMargin{28. }
\CSLRightInline{\textbf{DOIP endorsement request}
\CSLBlock{FDO-TSIG Working Group} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
\CSLLeftMargin{29. }
\CSLRightInline{\textbf{Digital object interface protocol specification, version 2.0}
\CSLBlock{DONA Foundation} \emph{DONA foundation} (2018-11-12) \url{https://hdl.handle.net/0.DOIP/DOIPV2.0}}

\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
\CSLLeftMargin{30. }
\CSLRightInline{\textbf{FAIR digital object demonstrators 2021}
\CSLBlock{Peter Wittenburg, Ivonne Anders, Christophe Blanchi, Merret Buurman, Carole Goble, Jonas Grieb, Alex Hardisty, Sharif Islam, Thomas Jejkal, Tibor Kálmán, \ldots{} Philipp Wieder} \emph{Zenodo} (2022)
\CSLBlock{DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}}}

\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
\CSLLeftMargin{31. }
\CSLRightInline{\textbf{FDO -- upload of FDO}
\CSLBlock{Christophe Blanchi, Daan Broeder, Thomas Jejkal, Islam Sharif, Alexander Schlemmer, Dieter van Uytvanck, Peter Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
\CSLLeftMargin{32. }
\CSLRightInline{\textbf{ResourceSync Framework Specification - Table of Contents} \url{http://www.openarchives.org/rs/toc}}

\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
\CSLLeftMargin{33. }
\CSLRightInline{\textbf{Typing FAIR digital objects}
\CSLBlock{Larry Lannom, U Schwardmann, C Blanchi, P Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
\CSLLeftMargin{34. }
\CSLRightInline{\textbf{Implementation of attributes, types, profiles and registries}
\CSLBlock{C Blanchi, M Hellström, L Lannom, U Schwardmann, P Wittenburg} \emph{FDO Forum} (2022-11-27) \url{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}}

\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
\CSLLeftMargin{35. }
\CSLRightInline{\textbf{Two Examples on How FDO Types can Support Machine and Human Readability}
\CSLBlock{Ulrich Schwardmann, Tibor Kálmán} \emph{Research Ideas and Outcomes} (2022-10-12) \url{https://doi.org/grcq23}
\CSLBlock{DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}}}

\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
\CSLLeftMargin{36. }
\CSLRightInline{\textbf{A framework for distributed digital object services}
\CSLBlock{Robert Kahn, Robert Wilensky} \emph{International Journal on Digital Libraries} (2006-04) \url{https://www.doi.org/topics/2006_05_02_Kahn_Framework.pdf}
\CSLBlock{DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}}}

\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
\CSLLeftMargin{37. }
\CSLRightInline{\textbf{A framework for distributed digital object services}
\CSLBlock{Robert Kahn, Robert Wilensky} \emph{CNRI} (1995-05-13) \url{http://www.cnri.reston.va.us/k-w.html}}

\leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
\CSLLeftMargin{38. }
\CSLRightInline{\textbf{X.1255 : Framework for discovery of identity management information} \url{https://www.itu.int/rec/T-REC-X.1255-201309-I}}

\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
\CSLLeftMargin{39. }
\CSLRightInline{\textbf{Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation} \url{https://www.dona.net/doipv1doc}}

\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
\CSLLeftMargin{40. }
\CSLRightInline{\textbf{Digital objects as drivers towards convergence in data infrastructures}
\CSLBlock{Peter Wittenburg, George Strawn, Barend Mons, Luiz Bonino, Erik Schultes} \emph{https://b2share.eudat.eu} (2019-01-06)
\CSLBlock{DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}}}

\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
\CSLLeftMargin{41. }
\CSLRightInline{\textbf{DOIP and Examples --- Cordra documentation} \url{https://www.cordra.org/documentation/api/doip.html}}

\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
\CSLLeftMargin{42. }
\CSLRightInline{\textbf{draft-bhutton-json-schema-00} \url{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}}

\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
\CSLLeftMargin{43. }
\CSLRightInline{\textbf{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}
\CSLBlock{Tim Berners-Lee, Mark Fischetti} \emph{HarperSanFrancisco} (1999)
\CSLBlock{ISBN: 9780062515865}}

\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
\CSLLeftMargin{44. }
\CSLRightInline{\textbf{RDF 1.1 Primer} \url{http://www.w3.org/TR/rdf11-primer/}}

\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
\CSLLeftMargin{45. }
\CSLRightInline{\textbf{Resource Description Framework (RDF) Model and Syntax Specification} \url{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}}

\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
\CSLLeftMargin{46. }
\CSLRightInline{\textbf{Process modelling for information system description}
\CSLBlock{Stefan K Stanczyk} \emph{The Open University} (1987)
\CSLBlock{DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}}}

\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
\CSLLeftMargin{47. }
\CSLRightInline{\textbf{Uniform Resource Identifier (URI): Generic Syntax}
\CSLBlock{T Berners-Lee, R Fielding, L Masinter} \emph{RFC Editor} (2005-01) \url{https://doi.org/ggqvpr}
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}}}

\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
\CSLLeftMargin{48. }
\CSLRightInline{\textbf{"info" URI Registry (Frozen)} \url{https://oclc-research.github.io/infoURI-Frozen/}}

\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
\CSLLeftMargin{49. }
\CSLRightInline{\textbf{Identifiers.org and MIRIAM Registry: community resources to provide persistent identification}
\CSLBlock{N Juty, N Le Novere, C Laibe} \emph{Nucleic Acids Research} (2011-12-02) \url{https://doi.org/cx2776}
\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22140103}{22140103} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3245029}{PMC3245029}}}

\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
\CSLLeftMargin{50. }
\CSLRightInline{\textbf{Internationalized Resource Identifiers (IRIs)}
\CSLBlock{M Duerst, M Suignard} \emph{RFC Editor} (2005-01) \url{https://doi.org/gjvnbg}
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}}}

\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
\CSLLeftMargin{51. }
\CSLRightInline{\textbf{Cool URIs for the semantic web}
\CSLBlock{Leo Sauermann, Richard Cyganiak, Max Völkel} \emph{Universität des Saarlandes} (2011-07-13)
\CSLBlock{DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}}}

\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
\CSLLeftMargin{52. }
\CSLRightInline{\textbf{The Semantic Web --- ISWC 2002}
\CSLBlock{Ian Horrocks, James Hendler (editors)} \emph{Lecture Notes in Computer Science} (2002) \url{https://doi.org/fwjmhp}
\CSLBlock{DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}}}

\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
\CSLLeftMargin{53. }
\CSLRightInline{\textbf{SPARQL 1.1 Overview} \url{http://www.w3.org/TR/sparql11-overview/}}

\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
\CSLLeftMargin{54. }
\CSLRightInline{\textbf{State of the nation in data integration for bioinformatics.}
\CSLBlock{Carole Goble, Robert Stevens} \emph{Journal of Biomedical Informatics} (2008-10-01)
\CSLBlock{DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/18358788}{18358788}}}

\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
\CSLLeftMargin{55. }
\CSLRightInline{\textbf{Open PHACTS: semantic interoperability for drug discovery}
\CSLBlock{Antony J Williams, Lee Harland, Paul Groth, Stephen Pettifer, Christine Chichester, Egon L Willighagen, Chris T Evelo, Niklas Blomberg, Gerhard Ecker, Carole Goble, Barend Mons} \emph{Drug Discovery Today} (2012-11) \url{https://doi.org/f4fdkd}
\CSLBlock{DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22683805}{22683805}}}

\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
\CSLLeftMargin{56. }
\CSLRightInline{\textbf{SKOS Simple Knowledge Organization System Primer} \url{http://www.w3.org/TR/skos-primer/}}

\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
\CSLLeftMargin{57. }
\CSLRightInline{\textbf{How matchable are four thousand ontologies on the semantic web}
\CSLBlock{Wei Hu, Jianfeng Chen, Hang Zhang, Yuzhong Qu} \emph{The semantic web: Research and applications} (2011)
\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20} · ISBN: 978-3-642-21033-4}}

\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
\CSLLeftMargin{58. }
\CSLRightInline{\textbf{API-centric Linked Data integration: The Open PHACTS Discovery Platform case study}
\CSLBlock{Paul Groth, Antonis Loizou, Alasdair JG Gray, Carole Goble, Lee Harland, Steve Pettifer} \emph{Journal of Web Semantics} (2014-12)
\CSLBlock{DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}}}

\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
\CSLLeftMargin{59. }
\CSLRightInline{\textbf{A systematic analysis of term reuse and term overlap across biomedical ontologies}
\CSLBlock{Maulik R Kamdar, Tania Tudorache, Mark A Musen} \emph{Semantic Web} (2017-08-07)
\CSLBlock{DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/28819351}{28819351} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5555235}{PMC5555235}}}

\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
\CSLLeftMargin{60. }
\CSLRightInline{\textbf{Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot}
\CSLBlock{Martin Klein, Herbert Van de Sompel, Robert Sanderson, Harihar Shankar, Lyudmila Balakireva, Ke Zhou, Richard Tobin} \emph{PLoS ONE} (2014-12-26)
\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/25541969}{25541969} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4277367}{PMC4277367}}}

\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
\CSLLeftMargin{61. }
\CSLRightInline{\textbf{A more decentralized vision for linked data}
\CSLBlock{Axel Polleres, Maulik Rajendra Kamdar, Javier David Fernández, Tania Tudorache, Mark Alan Musen} \emph{Satya Widya} (2020-01-31)
\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/sw-190380}}}

\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
\CSLLeftMargin{62. }
\CSLRightInline{\textbf{The landscape of ontology reuse approaches}
\CSLBlock{Valentina Anita Carriero, Marilena Daquino, Aldo Gangemi, Andrea Giovanni Nuzzolese, Silvio Peroni, Valentina Presutti, Francesca Tomasi} \emph{Applications and practices in ontology design, extraction, and reasoning} (2020-11-12)
\CSLBlock{DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/ssw200033} · ISBN: 9781643681429}}

\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
\CSLLeftMargin{63. }
\CSLRightInline{\textbf{RDF Schema 1.1} \url{http://www.w3.org/TR/rdf-schema/}}

\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
\CSLLeftMargin{64. }
\CSLRightInline{\textbf{OWL 2 Web Ontology Language Document Overview (Second Edition)} \url{http://www.w3.org/TR/owl2-overview/}}

\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
\CSLLeftMargin{65. }
\CSLRightInline{\textbf{Mapping between the OBO and OWL ontology languages}
\CSLBlock{Syed Tirmizi, Stuart Aitken, Dilvan A Moreira, Chris Mungall, Juan Sequeda, Nigam H Shah, Daniel P Miranker} \emph{Journal of Biomedical Semantics} (2011)
\CSLBlock{DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/21388572}{21388572} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3105495}{PMC3105495}}}

\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
\CSLLeftMargin{66. }
\CSLRightInline{\textbf{Linked Data - Design Issues} \url{https://www.w3.org/DesignIssues/LinkedData.html}}

\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
\CSLLeftMargin{67. }
\CSLRightInline{\textbf{The Open Graph protocol} \url{https://ogp.me/}}

\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
\CSLLeftMargin{68. }
\CSLRightInline{\textbf{RDFa 1.1 Primer - Third Edition} \url{http://www.w3.org/TR/rdfa-primer/}}

\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
\CSLLeftMargin{69. }
\CSLRightInline{\textbf{HTML Standard} \url{https://html.spec.whatwg.org/multipage/microdata.html}}

\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
\CSLLeftMargin{70. }
\CSLRightInline{\textbf{JSON-LD 1.1} \url{https://www.w3.org/TR/json-ld/}}

\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
\CSLLeftMargin{71. }
\CSLRightInline{\textbf{Usage Statistics of JSON-LD for Websites, May 2022} \url{https://w3techs.com/technologies/details/da-jsonld}}

\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
\CSLLeftMargin{72. }
\CSLRightInline{\textbf{Designing a Linked Data developer experience} (2018-12-28) \url{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}}

\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
\CSLLeftMargin{73. }
\CSLRightInline{\textbf{Shapes Constraint Language (SHACL)} \url{https://www.w3.org/TR/shacl/}}

\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
\CSLLeftMargin{74. }
\CSLRightInline{\textbf{Shape Expressions (ShEx) 2.1 Primer} \url{http://shex.io/shex-primer/}}

\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
\CSLLeftMargin{75. }
\CSLRightInline{\textbf{Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation}
\CSLBlock{Katherine Thornton, Harold Solbrig, Gregory S Stupp, Jose Emilio Labra Gayo, Daniel Mietchen, Eric Prud, Andra Waagmeester} \emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings} (2019)
\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39} · ISBN: 978-3-030-21347-3}}

\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
\CSLLeftMargin{76. }
\CSLRightInline{\textbf{Validating RDF Data}
\CSLBlock{Jose Emilio Labra Gayo, Eric Prud'hommeaux, Iovka Boneva, Dimitris Kontokostas} \emph{Synthesis Lectures on the Semantic Web: Theory and Technology} (2017-09-28)
\CSLBlock{DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}}}

\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
\CSLLeftMargin{77. }
\CSLRightInline{\textbf{Survey of tools for linked data consumption}
\CSLBlock{Jakub Klímek, Petr Škoda, Martin Nečaský} \emph{Satya Widya} (2019-05-23)
\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/sw-180316}}}

\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
\CSLLeftMargin{78. }
\CSLRightInline{\textbf{RightField: embedding ontology annotation in spreadsheets}
\CSLBlock{K Wolstencroft, S Owen, M Horridge, O Krebs, W Mueller, JL Snoep, F du Preez, C Goble} \emph{Bioinformatics} (2011-05-26)
\CSLBlock{DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/21622664}{21622664}}}

\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
\CSLLeftMargin{79. }
\CSLRightInline{\textbf{REST and Linked Data}
\CSLBlock{Kevin R Page, David C De Roure, Kirk Martinez} \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11} (2011) \url{http://eprints.soton.ac.uk/id/eprint/272098}
\CSLBlock{DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}}}

\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
\CSLLeftMargin{80. }
\CSLRightInline{\textbf{On Schema.org and Why It Matters for the Web}
\CSLBlock{Peter Mika} \emph{IEEE Internet Computing} (2015-07)
\CSLBlock{DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}}}

\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
\CSLLeftMargin{81. }
\CSLRightInline{\textbf{A new look at the semantic web}
\CSLBlock{Abraham Bernstein, James Hendler, Natalya Noy} \emph{Communications of the ACM} (2016-08-24)
\CSLBlock{DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}}}

\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
\CSLLeftMargin{82. }
\CSLRightInline{\textbf{An Interoperability Framework and Distributed Platform for Fast Data Applications}
\CSLBlock{José Carlos Martins Delgado} \emph{Data Science and Big Data Computing} (2016)
\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}}}

\leavevmode\vadjust pre{\hypertarget{ref-hRzcHhPD}{}}%
\CSLLeftMargin{83. }
\CSLRightInline{\textbf{A Comparison Framework for Middleware Infrastructures.}
\CSLBlock{Apostolos Zarras} \emph{The Journal of Object Technology} (2004)
\CSLBlock{DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}}}

\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
\CSLLeftMargin{84. }
\CSLRightInline{\textbf{The FAIR data maturity model: An approach to harmonise FAIR assessments}
\CSLBlock{Christophe Bahim, Carlos Casorrán-Amilburu, Makx Dekkers, Edit Herczog, Nicolas Loozen, Konstantinos Repanas, Keith Russell, Shelley Stall} \emph{Data Science Journal} (2020-10-27)
\CSLBlock{DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}}}

\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
\CSLLeftMargin{85. }
\CSLRightInline{\textbf{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}
\CSLBlock{William Stallings} \emph{Sams} (1990)
\CSLBlock{ISBN: 9780672226977}}

\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
\CSLLeftMargin{86. }
\CSLRightInline{\textbf{Architectural styles and the design of network-based software architectures}
\CSLBlock{Roy Thomas Fielding} \emph{University of California, Irvine} (2000) \url{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}}

\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
\CSLLeftMargin{87. }
\CSLRightInline{\textbf{Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award)}
\CSLBlock{Roy T Fielding, Richard N Taylor, Justin R Erenkrantz, Michael M Gorlick, Jim Whitehead, Rohit Khare, Peyman Oreizy} \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017} (2017-09-04) \url{http://dl.acm.org/citation.cfm?doid=3106237.3121282}
\CSLBlock{DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282} · ISBN: 9781450351058}}

\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
\CSLLeftMargin{88. }
\CSLRightInline{\textbf{OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More} \url{https://spec.openapis.org/oas/v3.1.0.html}}

\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
\CSLLeftMargin{89. }
\CSLRightInline{\textbf{Schema.org Actions - schema.org} \url{https://schema.org/docs/actions.html}}

\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
\CSLLeftMargin{90. }
\CSLRightInline{\textbf{Web Services Description Language (WSDL) Version 2.0 Part 0: Primer} \url{http://www.w3.org/TR/wsdl20-primer/}}

\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
\CSLLeftMargin{91. }
\CSLRightInline{\textbf{The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud}
\CSLBlock{Katherine Wolstencroft, Robert Haines, Donal Fellows, Alan Williams, David Withers, Stuart Owen, Stian Soiland-Reyes, Ian Dunlop, Aleksandra Nenadic, Paul Fisher, \ldots{} Carole Goble} \emph{Nucleic Acids Research} (2013-05-02) \url{https://doi.org/ggbwf4}
\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/23640334}{23640334} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3692062}{PMC3692062}}}

\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
\CSLLeftMargin{92. }
\CSLRightInline{\textbf{Perspectives on automated composition of workflows in the life sciences}
\CSLBlock{Anna-Lena Lamprecht, Magnus Palmblad, Jon Ison, Veit Schwämmle, Mohammad Sadnan Al Manir, Ilkay Altintas, Christopher JO Baker, Ammar Ben Hadj Amor, Salvador Capella-Gutierrez, Paulos Charonyktakis, \ldots{} Katherine Wolstencroft} \emph{F1000Research} (2021-09-07) \url{https://doi.org/gqprp7}
\CSLBlock{DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/34804501}{34804501} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8573700}{PMC8573700}}}

\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
\CSLLeftMargin{93. }
\CSLRightInline{\textbf{The Modern Standards Paradigm - Five Key Principles}
\CSLBlock{OpenStand} \url{https://open-stand.org/about-us/principles/}}

\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
\CSLLeftMargin{94. }
\CSLRightInline{\textbf{First International Conference on FAIR Digital Objects}
\CSLBlock{Tina Loo (editor)} \emph{Pensoft Publishers} (1970) \url{https://doi.org/grcqxh}
\CSLBlock{DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}}}

\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
\CSLLeftMargin{95. }
\CSLRightInline{\textbf{DOIP API for HTTP Clients --- Cordra documentation} \url{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}}

\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
\CSLLeftMargin{96. }
\CSLRightInline{\textbf{DOI Resolution Documentation} \url{https://www.doi.org/factsheets/DOIProxy.html}}

\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
\CSLLeftMargin{97. }
\CSLRightInline{\textbf{DOI Handbook - Resolution} \url{https://www.doi.org/doi_handbook/3_Resolution.html}}

\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
\CSLLeftMargin{98. }
\CSLRightInline{\textbf{HTTP Semantics}
\CSLBlock{R Fielding, M Nottingham, J Reschke (editors)} \emph{RFC Editor} (2022-06) \url{https://doi.org/gqprqb}
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}}}

\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
\CSLLeftMargin{99. }
\CSLRightInline{\textbf{Linked data platform 1.0}
\CSLBlock{Linked Data Platform Working Group} \emph{W3C} (2015-02-26) \url{https://www.w3.org/TR/2015/REC-ldp-20150226/}}

\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
\CSLLeftMargin{100. }
\CSLRightInline{\textbf{FAIR Digital Object Framework Documentation} \url{https://fairdigitalobjectframework.org/}}

\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
\CSLLeftMargin{101. }
\CSLRightInline{\textbf{Handle System Overview}
\CSLBlock{S Sun, L Lannom, B Boesch} \emph{RFC Editor} (2003-11) \url{https://doi.org/ggn83z}
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}}}

\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
\CSLLeftMargin{102. }
\CSLRightInline{\textbf{Web API design best practices - Azure Architecture Center}
\CSLBlock{martinekuan} \url{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}}

\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
\CSLLeftMargin{103. }
\CSLRightInline{\textbf{The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration}
\CSLBlock{Barry Smith, Michael Ashburner, Cornelius Rosse, Jonathan Bard, William Bug, Werner Ceusters, Louis J Goldberg, Karen Eilbeck, Amelia Ireland, \ldots{} Suzanna Lewis} \emph{Nature Biotechnology} (2007-11) \url{https://doi.org/bqng99}
\CSLBlock{DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/17989687}{17989687} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2814061}{PMC2814061}}}

\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
\CSLLeftMargin{104. }
\CSLRightInline{\textbf{Data Information View} \url{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}}

\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
\CSLLeftMargin{105. }
\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 3} \url{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}}

\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
\CSLLeftMargin{106. }
\CSLRightInline{\textbf{ORE User Guide - Primer} \url{http://www.openarchives.org/ore/1.0/primer}}

\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
\CSLLeftMargin{107. }
\CSLRightInline{\textbf{PROV-O: The PROV Ontology} \url{https://www.w3.org/TR/2013/REC-prov-o-20130430/}}

\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
\CSLLeftMargin{108. }
\CSLRightInline{\textbf{Semantic Web Services}
\CSLBlock{Dieter Fensel, Federico Michele Facca, Elena Simperl, Ioan Toma} \emph{Springer Berlin Heidelberg} (2011) \url{https://doi.org/bv7nnc}
\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}}}

\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
\CSLLeftMargin{109. }
\CSLRightInline{\textbf{5-star Open Data} \url{http://5stardata.info/en/}}

\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
\CSLLeftMargin{110. }
\CSLRightInline{\url{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}

\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
\CSLLeftMargin{111. }
\CSLRightInline{\textbf{Global Internet Phenomena Report 2022}
\CSLBlock{Sandvine} \url{https://www.sandvine.com/global-internet-phenomena-report-2022}}

\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
\CSLLeftMargin{112. }
\CSLRightInline{\textbf{Hypertext Transfer Protocol Version 2 (HTTP/2)}
\CSLBlock{M Belshe, R Peon} \emph{RFC Editor} (2015-05)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}}}

\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
\CSLLeftMargin{113. }
\CSLRightInline{\url{https://blog.cloudflare.com/http-3-vs-http-2/}}

\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
\CSLLeftMargin{114. }
\CSLRightInline{\textbf{draft-ietf-quic-http-34} \url{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}}

\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
\CSLLeftMargin{115. }
\CSLRightInline{\textbf{QUIC: A UDP-Based Multiplexed and Secure Transport}
\CSLBlock{J Iyengar, M Thomson (editors)} \emph{RFC Editor} (2021-05)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}}}

\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
\CSLLeftMargin{116. }
\CSLRightInline{\textbf{URI Template}
\CSLBlock{J Gregorio, R Fielding, M Hadley, M Nottingham, D Orchard} \emph{RFC Editor} (2012-03)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}}}

\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
\CSLLeftMargin{117. }
\CSLRightInline{\textbf{Content negotiation - HTTP \textbar{} MDN} \url{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}}

\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
\CSLLeftMargin{118. }
\CSLRightInline{\textbf{Hypertext Style: Cool URIs don't change.} \url{https://www.w3.org/Provider/Style/URI}}

\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
\CSLLeftMargin{119. }
\CSLRightInline{\textbf{HTTP Framework for Time-Based Access to Resource States -\/- Memento}
\CSLBlock{H Van de Sompel, M Nelson, R Sanderson} \emph{RFC Editor} (2013-12)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}}}

\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
\CSLLeftMargin{120. }
\CSLRightInline{\textbf{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}
\CSLBlock{G Clemm, J Amsden, T Ellison, C Kaler, J Whitehead} \emph{RFC Editor} (2002-03)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}}}

\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
\CSLLeftMargin{121. }
\CSLRightInline{\textbf{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}
\CSLBlock{L Dusseault (editor)} \emph{RFC Editor} (2007-06)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}}}

\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
\CSLLeftMargin{122. }
\CSLRightInline{\textbf{Upgrading to TLS Within HTTP/1.1}
\CSLBlock{R Khare, S Lawrence} \emph{RFC Editor} (2000-05)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}}}

\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
\CSLLeftMargin{123. }
\CSLRightInline{\textbf{Hypertext Style: Cool URIs don't change.} \url{https://www.w3.org/Provider/Style/URI.html}}

\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
\CSLLeftMargin{124. }
\CSLRightInline{\textbf{draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes} \url{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}}

\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
\CSLLeftMargin{125. }
\CSLRightInline{\textbf{The 'profile' Link Relation Type}
\CSLBlock{E Wilde} \emph{RFC Editor} (2013-03)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}}}

\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
\CSLLeftMargin{126. }
\CSLRightInline{\textbf{W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures} \url{http://www.w3.org/TR/xmlschema11-1/}}

\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
\CSLLeftMargin{127. }
\CSLRightInline{\textbf{JSON-LD 1.1} \url{http://www.w3.org/TR/json-ld/}}

\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
\CSLLeftMargin{128. }
\CSLRightInline{\textbf{WebSockets Standard} \url{https://websockets.spec.whatwg.org/}}

\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
\CSLLeftMargin{129. }
\CSLRightInline{\textbf{Linked Data Notifications} \url{https://www.w3.org/TR/ldn/}}

\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
\CSLLeftMargin{130. }
\CSLRightInline{\textbf{The Atom Publishing Protocol}
\CSLBlock{J Gregorio, B de hOra (editors)} \emph{RFC Editor} (2007-10)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}}}

\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
\CSLLeftMargin{131. }
\CSLRightInline{\textbf{SWORD 3.0 Specification} \url{https://swordapp.github.io/swordv3/swordv3.html}}

\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
\CSLLeftMargin{132. }
\CSLRightInline{\textbf{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}
\CSLBlock{R Fielding, J Reschke (editors)} \emph{RFC Editor} (2014-06)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}}}

\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
\CSLLeftMargin{133. }
\CSLRightInline{\textbf{Hydra W3C Community Group} \url{https://www.hydra-cg.com/}}

\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
\CSLLeftMargin{134. }
\CSLRightInline{\textbf{draft-kelly-json-hal-08} \url{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}}

\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
\CSLLeftMargin{135. }
\CSLRightInline{\textbf{Web Linking}
\CSLBlock{M Nottingham} \emph{RFC Editor} (2017-10)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}}}

\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
\CSLLeftMargin{136. }
\CSLRightInline{\textbf{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}
\CSLBlock{R Fielding, J Reschke (editors)} \emph{RFC Editor} (2014-06)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}}}

\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
\CSLLeftMargin{137. }
\CSLRightInline{\textbf{HTTP Live Streaming}
\CSLBlock{W May} \emph{RFC Editor} (2017-08)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}}}

\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
\CSLLeftMargin{138. }
\CSLRightInline{\textbf{ISO/IEC 23009-1:2019}
\CSLBlock{14:00-17:00} \emph{ISO} \url{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}}

\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
\CSLLeftMargin{139. }
\CSLRightInline{\textbf{FAIR data maturity model: Specification and guidelines}
\CSLBlock{Research Data Alliance FAIR Data Maturity Model Working Group} \emph{Research Data Alliance} (2020) \url{https://zenodo.org/record/3909563/\#.\%7BYGRNnq8za70\%7D}
\CSLBlock{DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}}}

\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
\CSLLeftMargin{140. }
\CSLRightInline{\textbf{NCBO BioPortal} \url{https://bioportal.bioontology.org/ontologies}}

\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
\CSLLeftMargin{141. }
\CSLRightInline{\textbf{The Globus Striped GridFTP Framework and Server}
\CSLBlock{W Allcock, J Bresnahan, R Kettimuthu, M Link} \emph{ACM/IEEE SC 2005 Conference (SC'05)} \url{https://doi.org/cgmc2b}
\CSLBlock{DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}}}

\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
\CSLLeftMargin{142. }
\CSLRightInline{\textbf{The ARK Identifier Scheme} \url{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}}

\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
\CSLLeftMargin{143. }
\CSLRightInline{\textbf{Handle System Protocol (ver 2.1) Specification}
\CSLBlock{S Sun, S Reilly, L Lannom, J Petrone} \emph{RFC Editor} (2003-11)
\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}}}

\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
\CSLLeftMargin{144. }
\CSLRightInline{\textbf{Handle.Net Registry} \url{https://www.handle.net/download_hnr.html}}

\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
\CSLLeftMargin{145. }
\CSLRightInline{\textbf{Licenses \& Standards \textbar{} Open Source Initiative} \url{https://opensource.org/licenses}}

\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
\CSLLeftMargin{146. }
\CSLRightInline{\textbf{ISO 16684-1:2019}
\CSLBlock{14:00-17:00} \emph{ISO} \url{https://www.iso.org/standard/75163.html}}

\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
\CSLLeftMargin{147. }
\CSLRightInline{\textbf{Data - W3C} \url{https://www.w3.org/standards/semanticweb/data}}

\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
\CSLLeftMargin{148. }
\CSLRightInline{\textbf{Linking Across Provenance Bundles} \url{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}}

\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
\CSLLeftMargin{149. }
\CSLRightInline{\textbf{Introducing 'Role'}
\CSLBlock{Unknown} \url{http://blog.schema.org/2014/06/introducing-role.html}}

\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
\CSLLeftMargin{150. }
\CSLRightInline{\textbf{PROV-O: The PROV Ontology} \url{https://www.w3.org/TR/prov-o/\#specializationOf}}

\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
\CSLLeftMargin{151. }
\CSLRightInline{\textbf{ORE Specification - Abstract Data Model} \url{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}}

\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
\CSLLeftMargin{152. }
\CSLRightInline{\textbf{Schema.org - Schema.org} \url{https://schema.org/}}

\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
\CSLLeftMargin{153. }
\CSLRightInline{\textbf{DCMI Metadata Terms} \url{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}}

\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
\CSLLeftMargin{154. }
\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 2} \url{https://www.w3.org/TR/vocab-dcat-2/}}

\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
\CSLLeftMargin{155. }
\CSLRightInline{\textbf{FOAF Vocabulary Specification} \url{http://xmlns.com/foaf/spec/}}

\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
\CSLLeftMargin{156. }
\CSLRightInline{\textbf{PAV ontology: provenance, authoring and versioning}
\CSLBlock{Paolo Ciccarese, Stian Soiland-Reyes, Khalid Belhajjame, Alasdair JG Gray, Carole Goble, Tim Clark} \emph{Journal of Biomedical Semantics} (2013) \url{https://doi.org/gftcpx}
\CSLBlock{DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/24267948}{24267948} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4177195}{PMC4177195}}}

\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
\CSLLeftMargin{157. }
\CSLRightInline{\textbf{DCMI Metadata Terms} (2020-01-20) \url{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}}

\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
\CSLLeftMargin{158. }
\CSLRightInline{\textbf{The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
Digital Object Approach to Speeding up Digital Mobilisation of Natural History
Collections}
\CSLBlock{Alex Hardisty, Paul Brack, Carole Goble, Laurence Livermore, Ben Scott, Quentin Groom, Stuart Owen, Stian Soiland-Reyes} \emph{Data Intelligence} (2022) \url{https://doi.org/grk33x}
\CSLBlock{DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}}}

\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
\CSLLeftMargin{159. }
\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 2} \url{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}}

\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
\CSLLeftMargin{160. }
\CSLRightInline{\textbf{Bioschemas - Bioschemas} \url{http://bioschemas.org/}}

\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
\CSLLeftMargin{161. }
\CSLRightInline{\textbf{Darwin Core: An Evolving Community-Developed Biodiversity Data Standard}
\CSLBlock{John Wieczorek, David Bloom, Robert Guralnick, Stan Blum, Markus Döring, Renato Giovanni, Tim Robertson, David Vieglais} \emph{PLoS ONE} (2012-01-06) \url{https://doi.org/fzrpwq}
\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22238640}{22238640} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3253084}{PMC3253084}}}

\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
\CSLLeftMargin{162. }
\CSLRightInline{\textbf{Semantic Web - XML2000 - slide "Architecture"} \url{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}}

\end{CSLReferences}

### 2023-01-24 manuscript.tex
peerj citation style [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/86bc7fe899c7464fad8f4ebe872efb61c3be1394.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/86bc7fe899c7464fad8f4ebe872efb61c3be1394/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4000951091

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/51048f35d5fce97aa4194e6426fdd7c704f19199/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/51048f35d5fce97aa4194e6426fdd7c704f19199}{stain/2022-fdo-paper@51048f3}
+from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
 on January 24, 2023.
@@
 
-The FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
+The FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}) encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
 They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures {[}\protect\hyperlink{ref-u8Cy0psL}{2}{]}.
-The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
-
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
+In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures (\protect\hyperlink{ref-u8Cy0psL}{Mons et al., 2017}).
+The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
+
+Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
 The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
 
-In other discourse, Linked Data {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles {[}\protect\hyperlink{ref-GJHYN6xW}{6},\protect\hyperlink{ref-EJDjHrUv}{7}{]}.
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services {[}\protect\hyperlink{ref-99MS5xE0}{9},\protect\hyperlink{ref-J2H8yssV}{10}{]}.
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability {[}\protect\hyperlink{ref-DNTCUjpC}{11},\protect\hyperlink{ref-Pa67pUtR}{12}{]}.
-
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality {[}\protect\hyperlink{ref-uxWjmlio}{13},\protect\hyperlink{ref-v0WwFHgT}{14}{]} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves {[}\protect\hyperlink{ref-5V3GqSzR}{15}{]}.
+In other discourse, Linked Data (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\protect\hyperlink{ref-EJDjHrUv}{Bonino Da Silva Santos et al., 2016}; \protect\hyperlink{ref-GJHYN6xW}{Hasnain \& Rebholz-Schuhmann, 2018}).
+Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}).
+While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).
+
+One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\protect\hyperlink{ref-uxWjmlio}{Stefi, 2015}; \protect\hyperlink{ref-v0WwFHgT}{Stefi \& Hess, 2015}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves (\protect\hyperlink{ref-5V3GqSzR}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021b}).
 
@@
 
-The FAIR Digital Object Forum {[}\protect\hyperlink{ref-DMEBhHID}{16}{]} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
+The FAIR Digital Object Forum (\protect\hyperlink{ref-DMEBhHID}{{`FAIR Digital Objects Forum \textbar{}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
 As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
@@
 
-\textbf{FAIR Digital Object Overview and Specifications} {[}\protect\hyperlink{ref-11NpETLq1}{17}{]} is a comprehensive overview of FAIR Digital Object specifications listed below.
+\textbf{FAIR Digital Object Overview and Specifications} (\protect\hyperlink{ref-11NpETLq1}{Anders et al., 2022a}) is a comprehensive overview of FAIR Digital Object specifications listed below.
 It serves as a primer that introduces FDO concepts and the remaining documents.
 
-The \textbf{FDO Forum Document Standards} {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
+The \textbf{FDO Forum Document Standards} (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
 In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
 
-The \textbf{FDO Requirement Specifications} {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} is an update of {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} as the foundational definition of FDO.
+The \textbf{FDO Requirement Specifications} (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) is an update of (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) as the foundational definition of FDO.
 This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
@@
 
-The \textbf{Machine actionability} {[}\protect\hyperlink{ref-iJeys0T5}{21}{]} sets out to define what is meant by \emph{machine actionability} for FDOs.
+The \textbf{Machine actionability} (\protect\hyperlink{ref-iJeys0T5}{Weiland et al., 2022a}) sets out to define what is meant by \emph{machine actionability} for FDOs.
 \emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
@@
 
-\textbf{Configuration Types} {[}\protect\hyperlink{ref-SPd0uEqO}{22}{]} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}
-
-\textbf{PID Profiles \& Attributes} {[}\protect\hyperlink{ref-ljVViWCl}{24}{]} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}.
+\textbf{Configuration Types} (\protect\hyperlink{ref-SPd0uEqO}{Lannom et al., 2022a}) classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
+Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022})
+
+\textbf{PID Profiles \& Attributes} (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}) specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
+This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
 This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
 
-\textbf{Kernel Attributes \& Metadata} {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}.
+\textbf{Kernel Attributes \& Metadata} (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
+This document expands on RDA Recommendation on PID Kernel Information (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
 It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
 
-\textbf{Granularity, Versioning, Mutability} {[}\protect\hyperlink{ref-16W9AycSW}{27}{]} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
+\textbf{Granularity, Versioning, Mutability} (\protect\hyperlink{ref-16W9AycSW}{Working Group, 2022b}) considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
 The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
 
-\textbf{DOIP Endorsement Request} {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} is an endorsement of the DOIP v2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} specification as a potential FDO implementation, as it has been applied by several institutions {[}\protect\hyperlink{ref-jRWurmQc}{30}{]}.
+\textbf{DOIP Endorsement Request} (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) is an endorsement of the DOIP v2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) specification as a potential FDO implementation, as it has been applied by several institutions (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}).
 The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
 
-\textbf{Upload of FDO} {[}\protect\hyperlink{ref-1H7cmmr69}{31}{]} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync {[}\protect\hyperlink{ref-eS5y9TRh}{32}{]} is suggested as one type of service to list FDOs.
+\textbf{Upload of FDO} (\protect\hyperlink{ref-1H7cmmr69}{Blanchi et al., 2022a}) illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
+ResourceSync (\protect\hyperlink{ref-eS5y9TRh}{{`ResourceSync Framework Specification - Table of Contents'}}) is suggested as one type of service to list FDOs.
 This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
 
-\textbf{Typing FAIR Digital Objects} {[}\protect\hyperlink{ref-ZFzPxCNB}{33}{]} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
+\textbf{Typing FAIR Digital Objects} (\protect\hyperlink{ref-ZFzPxCNB}{Lannom et al., 2022b}) defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
 This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
@@
 
-\textbf{Implementation of Attributes, Types, Profiles and Registries} {[}\protect\hyperlink{ref-126uxr5pI}{34}{]} details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
+\textbf{Implementation of Attributes, Types, Profiles and Registries} (\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
 This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} {[}\protect\hyperlink{ref-1CqIZ47pu}{35}{]}.
+Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} (\protect\hyperlink{ref-1CqIZ47pu}{Schwardmann \& Kálmán, 2022}).
 
@@
 
-The concept of \textbf{FAIR Digital Objects} {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} has been introduced as way to expose research data as active objects that conform to the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}.
-This builds on the \emph{Digital Object} (DO) concept {[}\protect\hyperlink{ref-11MnuwJ4l}{36}{]}, first introduced in 1995 {[}\protect\hyperlink{ref-3Uqe3fuK}{37}{]} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework {[}\protect\hyperlink{ref-103Hw8H43}{38}{]} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]}.
+The concept of \textbf{FAIR Digital Objects} (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) has been introduced as way to expose research data as active objects that conform to the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
+This builds on the \emph{Digital Object} (DO) concept (\protect\hyperlink{ref-11MnuwJ4l}{Kahn \& Wilensky, 2006}), first introduced in 1995 (\protect\hyperlink{ref-3Uqe3fuK}{Kahn \& Wilensky, 1995}) as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
+DO was the inspiration for the ITU X.1255 framework (\protect\hyperlink{ref-103Hw8H43}{{`X.1255 : Framework for discovery of identity management information'}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}).
 
@@
 The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing {[}\protect\hyperlink{ref-tz0P3DTC}{40}{]}.
-
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}.
+The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing (\protect\hyperlink{ref-tz0P3DTC}{Wittenburg et al., 2019}).
+
+Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}), in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
+Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}).
 The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
@@
 
-FDO is an evolving concept. A set of FDO Demonstrators {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} highlight how current adapters are approaching implementations of FDO from different angles:
+FDO is an evolving concept.
+A set of FDO Demonstrators (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) highlight how current adapters are approaching implementations of FDO from different angles:
 
@@
 \item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, see {[}\protect\hyperlink{ref-xuQKRRx5}{41}{]}.} (usually TCP/IP over TLS).
+  Building on the Digital Object concept, using the simplified DOIP v2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), see (\protect\hyperlink{ref-xuQKRRx5}{{`DOIP and Examples --- Cordra documentation'}}).} (usually TCP/IP over TLS).
 The main DOIP operations are retrieving, creating and updating digital objects.
 These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]} and PIDs are assigned using the Handle system.
+FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}}) and PIDs are assigned using the Handle system.
 Several type registries have been established.
@@
 
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee {[}\protect\hyperlink{ref-Guy7cjVY}{43}{]}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
+The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee (\protect\hyperlink{ref-Guy7cjVY}{Berners-Lee \& Fischetti, 1999}), at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
 The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
@@
 
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) {[}\protect\hyperlink{ref-16WGmCih8}{44}{]} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format {[}\protect\hyperlink{ref-16wvFH4nh}{45}{]} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
-
-While triple-based knowledge representations were not new {[}\protect\hyperlink{ref-NxvhwSmG}{46}{]}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs {[}\protect\hyperlink{ref-RRtwPF7Y}{47}{]} are generalised forms of URLs that include locator-less identifiers
+Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\protect\hyperlink{ref-16WGmCih8}{{`RDF 1.1 Primer'}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\protect\hyperlink{ref-16wvFH4nh}{{`Resource Description Framework (RDF) Model and Syntax Specification'}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
+
+While triple-based knowledge representations were not new (\protect\hyperlink{ref-NxvhwSmG}{Stanczyk, 1987}), the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs (\protect\hyperlink{ref-RRtwPF7Y}{Berners-Lee, Fielding \& Masinter, 2005}) are generalised forms of URLs that include locator-less identifiers
   such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years {[}\protect\hyperlink{ref-nW4HY8Nq}{48}{]}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) {[}\protect\hyperlink{ref-HqZ3r6J3}{49}{]}.
-RDF 1.1 formats use Unicode to support IRIs {[}\protect\hyperlink{ref-rqjuXkEf}{50}{]}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
+The distinction between locator-full and locator-less identifiers have weakened in recent years (\protect\hyperlink{ref-nW4HY8Nq}{{`"info" URI Registry (Frozen)'}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) (\protect\hyperlink{ref-HqZ3r6J3}{Juty, Le Novere \& Laibe, 2011}).
+RDF 1.1 formats use Unicode to support IRIs (\protect\hyperlink{ref-rqjuXkEf}{Duerst \& Suignard, 2005}), which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
+By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
 
@@
 The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} {[}\protect\hyperlink{ref-yd2WE9eq}{52}{]} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
-
-Established Semantic Web services and standards include SPARQL {[}\protect\hyperlink{ref-xww8fb1B}{53}{]} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
-
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins {[}\protect\hyperlink{ref-bMSxa8XW}{54},\protect\hyperlink{ref-Qlrlr2TZ}{55}{]}.
+More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} (\protect\hyperlink{ref-yd2WE9eq}{Horrocks \& Hendler, 2002}) gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
+
+Established Semantic Web services and standards include SPARQL (\protect\hyperlink{ref-xww8fb1B}{{`SPARQL 1.1 Overview'}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
+
+The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\protect\hyperlink{ref-bMSxa8XW}{Goble \& Stevens, 2008}; \protect\hyperlink{ref-Qlrlr2TZ}{Williams et al., 2012}).
 In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
@@
 Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) {[}\protect\hyperlink{ref-FDDmgO6s}{57}{]} .
+Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
+Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) (\protect\hyperlink{ref-FDDmgO6s}{Hu et al., 2011}) .
 
 The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS {[}\protect\hyperlink{ref-191ZkYZMt}{58}{]}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
+This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS (\protect\hyperlink{ref-191ZkYZMt}{Groth et al., 2014})), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
 
 With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps {[}\protect\hyperlink{ref-9CdJnsoJ}{59}{]}.
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication {[}\protect\hyperlink{ref-AGHR9oyr}{60}{]}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data {[}\protect\hyperlink{ref-5SuwLNdE}{61}{]}.
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies {[}\protect\hyperlink{ref-TUmA4KQT}{62}{]} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS {[}\protect\hyperlink{ref-ZwAcGQKY}{63}{]}, OWL {[}\protect\hyperlink{ref-1p4IWJpI}{64}{]}, OBO {[}\protect\hyperlink{ref-1DalZuKct}{65}{]}, SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]}), hash vs slash, HTTP status codes and PID redirection strategies {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.
+This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps (\protect\hyperlink{ref-9CdJnsoJ}{Kamdar, Tudorache \& Musen, 2017}).
+Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication (\protect\hyperlink{ref-AGHR9oyr}{Klein et al., 2014}), for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data (\protect\hyperlink{ref-5SuwLNdE}{Polleres et al., 2020, p.
+handle:20.500.11811/7183}).
+Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies (\protect\hyperlink{ref-TUmA4KQT}{Carriero et al., 2020}) , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), OBO (\protect\hyperlink{ref-1DalZuKct}{Tirmizi et al., 2011}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})), hash vs slash, HTTP status codes and PID redirection strategies (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).
 
@@
 
-The \textbf{Linked Data} concept {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} was kickstarted as a set of best practices {[}\protect\hyperlink{ref-OGXYtGZ8}{66}{]} to bring the Web aspect back into focus.
+The \textbf{Linked Data} concept (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) was kickstarted as a set of best practices (\protect\hyperlink{ref-OGXYtGZ8}{{`Linked Data - Design Issues'}}) to bring the Web aspect back into focus.
 Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
@@
 Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph {[}\protect\hyperlink{ref-oN9D5LkD}{67}{]}, RDFa {[}\protect\hyperlink{ref-FKIF6ApB}{68}{]}, Microdata {[}\protect\hyperlink{ref-14Utscfe4}{69}{]}) we find JSON-LD {[}\protect\hyperlink{ref-X5SWPqSU}{70}{]} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites {[}\protect\hyperlink{ref-j3SqQLBR}{71}{]}.
-
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} {[}\protect\hyperlink{ref-BFMoe8FC}{72}{]} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL {[}\protect\hyperlink{ref-GDrm9dgX}{73}{]} or ShEx {[}\protect\hyperlink{ref-YUzogaV}{74}{]}) can be used to validate RDF Data {[}\protect\hyperlink{ref-S5BI6Mwx}{75},\protect\hyperlink{ref-ZeOAPNHa}{76}{]} before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts {[}\protect\hyperlink{ref-o596f2JM}{77}{]}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
-
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}.
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies {[}\protect\hyperlink{ref-MhsCt4P3}{78}{]}, the use of Linked Data as part of REST Web APIs {[}\protect\hyperlink{ref-eS6a6tlz}{79}{]} , and as shown by the big uptake by publishers to annotate the Web using schema.org {[}\protect\hyperlink{ref-117tX9MEG}{81}{]}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
+In addition to several such embedded \emph{microformats} (Open Graph (\protect\hyperlink{ref-oN9D5LkD}{{`The Open Graph protocol'}}), RDFa (\protect\hyperlink{ref-FKIF6ApB}{{`RDFa 1.1 Primer - Third Edition'}}), Microdata (\protect\hyperlink{ref-14Utscfe4}{{`HTML Standard'}})) we find JSON-LD (\protect\hyperlink{ref-X5SWPqSU}{{`JSON-LD 1.1'}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
+JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}).
+
+Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\protect\hyperlink{ref-BFMoe8FC}{{`Designing a Linked Data developer experience'}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\protect\hyperlink{ref-GDrm9dgX}{{`Shapes Constraint Language (SHACL)'}}) or ShEx (\protect\hyperlink{ref-YUzogaV}{{`Shape Expressions (ShEx) 2.1 Primer'}})) can be used to validate RDF Data (\protect\hyperlink{ref-ZeOAPNHa}{Gayo et al., 2017}; \protect\hyperlink{ref-S5BI6Mwx}{Thornton et al., 2019}) before consuming it programmatically, or reshaping data to fit other models.
+While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts (\protect\hyperlink{ref-o596f2JM}{Klímek, Škoda \& Nečaský, 2019}), which then tend to prefer non-semantic two-dimensional formats such as CSV files.
+
+A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}).
+One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies (\protect\hyperlink{ref-MhsCt4P3}{Wolstencroft et al., 2011}), the use of Linked Data as part of REST Web APIs (\protect\hyperlink{ref-eS6a6tlz}{Page, De Roure \& Martinez, 2011}) , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy (\protect\hyperlink{ref-117tX9MEG}{2016}), with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
 
@@
 \item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
-\item
-  The FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}, validated against its implementations for completeness.
-\item
-  A Comparison Framework for Middleware Infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
-\item
-  Cross-checks against RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-UzQhqk0M}{84}{]} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
-\item
-  EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
+  An Interoperability Framework and Distributed Platform for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}), which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
+\item
+  The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), validated against its implementations for completeness.
+\item
+  A Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}), which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
+\item
+  Cross-checks against RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
+\item
+  EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
 \end{enumerate}
@@
 Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} against Linked Data in general.
+For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) against Linked Data in general.
 
@@
 
-The Interoperability Framework for Fast Data Applications {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
-
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model {[}\protect\hyperlink{ref-8ZRGFa6y}{85}{]} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
+The Interoperability Framework for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
+
+We have chosen to investigate using this framework as it covers the higher levels of the OSI Model (\protect\hyperlink{ref-8ZRGFa6y}{Stallings, 1990}) better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
 In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]}.
+\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}).
 \label{tbl:fdo-web-interoperability-framework}}\tabularnewline
@@
 & The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs {[}\protect\hyperlink{ref-174AwcFUL}{86}{]} end up being mandated centrally by dominant (often commercial) providers {[}\protect\hyperlink{ref-11VxBOeZy}{87}{]}, which clients are required to use as-is with special code per service.
+In practice, REST APIs (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}) end up being mandated centrally by dominant (often commercial) providers (\protect\hyperlink{ref-11VxBOeZy}{Fielding et al., 2017}), which clients are required to use as-is with special code per service.
 Use of Linked Data enables common tooling and semantic mapping across differences.
@@
 & ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]}, schema.org Actions {[}\protect\hyperlink{ref-dKAekUjL}{89}{]}, WSDL/SOAP {[}\protect\hyperlink{ref-LmudecyN}{90}{]}, but frequently just as human-readable developer documentation/examples.
+Operational composition, typically within a single API provider, documented by OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), schema.org Actions (\protect\hyperlink{ref-dKAekUjL}{{`Schema.org Actions - schema.org'}}), WSDL/SOAP (\protect\hyperlink{ref-LmudecyN}{{`Web Services Description Language (WSDL) Version 2.0 Part 0: Primer'}}), but frequently just as human-readable developer documentation/examples.
 \\
@@
 No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} & Lightweight HTTP semantics for authenticity/navigation.
+JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) & Lightweight HTTP semantics for authenticity/navigation.
 Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
@@
 \\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
+\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
 \textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
@@
 
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services {[}\protect\hyperlink{ref-39K3yHT3}{91}{]}.
-However, this is mostly done via developer or human interaction {[}\protect\hyperlink{ref-iYh1wFBt}{92}{]}.
+The Web has already showed us how one can compose workflows of hetereogeneous Web Services (\protect\hyperlink{ref-39K3yHT3}{Wolstencroft et al., 2013}).
+However, this is mostly done via developer or human interaction (\protect\hyperlink{ref-iYh1wFBt}{Lamprecht et al., 2021}).
 Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
@@
 These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles {[}\protect\hyperlink{ref-hxtgCGjY}{93}{]} to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference {[}\protect\hyperlink{ref-jrZe6Esu}{94}{]}.
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP {[}\protect\hyperlink{ref-p7nwRj8E}{95}{]}.
+The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\protect\hyperlink{ref-hxtgCGjY}{{`The Modern Standards Paradigm - Five Key Principles'}}) to embrace openness, transparency, and broad consensus.
+In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference (\protect\hyperlink{ref-jrZe6Esu}{Loo, 1970}).
+This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\protect\hyperlink{ref-p7nwRj8E}{{`DOIP API for HTTP Clients --- Cordra documentation'}}).
 
 It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web {[}\protect\hyperlink{ref-1H9iQhQYq}{96}{]} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail {[}\protect\hyperlink{ref-68fzID2k}{97}{]}.
+For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\protect\hyperlink{ref-68fzID2k}{{`DOI Handbook - Resolution'}}).
 
@@
 
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics {[}\protect\hyperlink{ref-8HJqcF1Q}{98}{]}.
+The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}).
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]} to equivalent concepts for FDO and Web.
+\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
 \label{tbl:metamodel-concepts}}\tabularnewline
@@
 
-The FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} and using Linked Data Platform {[}\protect\hyperlink{ref-17OHlMRQA}{99}{]}, as proposed by {[}\protect\hyperlink{ref-7szz7dwO}{100}{]}.
+The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) sets out recommendations for FDO implementations.
+In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and using Linked Data Platform (\protect\hyperlink{ref-17OHlMRQA}{Linked Data Platform Working Group, 2015}), as proposed by (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}).
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines {[}\protect\hyperlink{ref-yygVPoL0}{19},\protect\hyperlink{ref-RwvirqWg}{20}{]} against its current implementations as DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} and Linked Data Platform (LDP) {[}\protect\hyperlink{ref-7szz7dwO}{100}{]}, with suggestions for required additions.
+\caption{Checking FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}; \protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) against its current implementations as DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}), with suggestions for required additions.
 \label{tbl:fdo-checks}}\tabularnewline
@@
 Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites {[}\protect\hyperlink{ref-j3SqQLBR}{71}{]}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
+& JSON-LD used by half of all websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}), however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
 Example-driven documentation.
@@
 Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]} \\
+& Declare supported methods in advance, e.g.~OpenAPI (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \\
 G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
@@
 \\
-G9: \emph{standard compliance} & Handle {[}\protect\hyperlink{ref-2gXY2Nwh}{101}{]}, DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}.
+G9: \emph{standard compliance} & Handle (\protect\hyperlink{ref-2gXY2Nwh}{Sun, Lannom \& Boesch, 2003}), DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
 FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} & HTTP, LDP.
+& Formalise standard process of FDO requirements (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) & HTTP, LDP.
 FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
 FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve {[}\protect\hyperlink{ref-ljVViWCl}{24}{]}.
+& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}).
 Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
@@
 FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns {[}\protect\hyperlink{ref-i9Ang0rM}{102}{]} are used instead.
+& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan}) are used instead.
 & Document how operation resources can be discovered from an LPD container.
@@
 & Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} {[}\protect\hyperlink{ref-CNP39tah}{103}{]} & Rather document better how individual ad-hoc schemas can be started for prototypes.
+& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} (\protect\hyperlink{ref-CNP39tah}{Smith et al., 2007}) & Rather document better how individual ad-hoc schemas can be started for prototypes.
 \\
@@
 Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-{[}\protect\hyperlink{ref-6LvydTsC}{104}{]}).
+(\protect\hyperlink{ref-6LvydTsC}{{`Data Information View'}})).
 & & LDP Containers required by specification, also user-created (eg.
 \texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 {[}\protect\hyperlink{ref-1EW4hLN8U}{105}{]}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE {[}\protect\hyperlink{ref-gtGuSevK}{106}{]} \\
+& Clarify relation to other collections like DCAT 3 (\protect\hyperlink{ref-1EW4hLN8U}{{`Data Catalog Vocabulary (DCAT) - Version 3'}}), \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE (\protect\hyperlink{ref-gtGuSevK}{{`ORE User Guide - Primer'}}) \\
 FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
@@
 
-Note that the draft update to FDO specification {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
+Note that the draft update to FDO specification (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
 
@@
 Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
+The DOIP Endorsement (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
 Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
@@
   While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} {[}\protect\hyperlink{ref-9T4j3N4e}{107}{]}
+Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}})
 \item
@@
 
-We argue that the Web, with its now ubiquitous use of REST API {[}\protect\hyperlink{ref-174AwcFUL}{86}{]}, can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services {[}\protect\hyperlink{ref-WzPECPDP}{108}{]} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} {[}\protect\hyperlink{ref-11dInoyVF}{109}{]}.
-
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
+We argue that the Web, with its now ubiquitous use of REST API (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), can be compared as a similar global middleware.
+Note that while early moves for developing Semantic Web Services (\protect\hyperlink{ref-WzPECPDP}{Fensel et al., 2011}) attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\protect\hyperlink{ref-11dInoyVF}{{`5-star Open Data'}}).
+
+For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}) that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
 
@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}) and Web technologies (using Linked Data) as middleware infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]}
+\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018})) and Web technologies (using Linked Data) as middleware infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004})
 \label{tbl:fdo-web-middleware}}\tabularnewline
@@
 \\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second {[}\protect\hyperlink{ref-18P9txgeB}{110}{]}.
+\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
 DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
@@
 Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks {[}\protect\hyperlink{ref-F8VZ86hu}{111}{]}.
+& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks (\protect\hyperlink{ref-F8VZ86hu}{Sandvine}).
 HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams {[}\protect\hyperlink{ref-KbbW0kGT}{112}{]} but still has TCP+TLS startup costs.
-For reduced latency {[}\protect\hyperlink{ref-CJuBupCW}{113}{]}, HTTP/3 {[}\protect\hyperlink{ref-Amit86Tp}{114}{]} use QUIC {[}\protect\hyperlink{ref-1GWvyEmUf}{115}{]}) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
+HTTP/2 adds asynchronous responses and multiplexed streams (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}) but still has TCP+TLS startup costs.
+For reduced latency , HTTP/3 (\protect\hyperlink{ref-Amit86Tp}{{`draft-ietf-quic-http-34'}}) use QUIC (\protect\hyperlink{ref-1GWvyEmUf}{Iyengar \& Thomson, 2021})) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
 \\
@@
 A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates {[}\protect\hyperlink{ref-11hxwwuRt}{116}{]}, content-negotiation {[}\protect\hyperlink{ref-n2f3TfoI}{117}{]} or FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}.
+Metadata is not always machine findable, may need pre-agreed API URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), content-negotiation (\protect\hyperlink{ref-n2f3TfoI}{{`Content negotiation - HTTP \textbar{} MDN'}}) or FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}).
 \\
@@
 There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged {[}\protect\hyperlink{ref-rbG9uRKw}{118}{]}.
+& URLs are not required to persist, although encouraged (\protect\hyperlink{ref-rbG9uRKw}{{`Hypertext Style: Cool URIs don't change.'}}).
 Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
 An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento {[}\protect\hyperlink{ref-DNjcMi4a}{119}{]} expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method {[}\protect\hyperlink{ref-hRUsaTiz}{120}{]} (used by SVN).
+Memento (\protect\hyperlink{ref-DNjcMi4a}{Van de Sompel, Nelson \& Sanderson, 2013}) expose versioned snapshots.
+WebDAV \texttt{VERSION-CONTROL} method (\protect\hyperlink{ref-hRUsaTiz}{Clemm et al., 2002}) (used by SVN).
 \\
@@
 & Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} {[}\protect\hyperlink{ref-xvMAyRAc}{121}{]} with \texttt{LOCK} and \texttt{UNLOCK} methods.
+WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} (\protect\hyperlink{ref-xvMAyRAc}{Dusseault, 2007}) with \texttt{LOCK} and \texttt{UNLOCK} methods.
 \\
@@
 Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade {[}\protect\hyperlink{ref-1B16WfR7q}{122}{]} to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns {[}\protect\hyperlink{ref-iME5l1kk}{123}{]} and PIDs via Persistent URLs (PURL).
+\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade (\protect\hyperlink{ref-1B16WfR7q}{Khare \& Lawrence, 2000}) to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
+Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns (\protect\hyperlink{ref-iME5l1kk}{{`Hypertext Style: Cool URIs don't change.'}}) and PIDs via Persistent URLs (PURL).
 \\
@@
 FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]}) & Syntactically Media Type with multiple suffixes {[}\protect\hyperlink{ref-nlH2rm1s}{124}{]} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) {[}\protect\hyperlink{ref-nCzRO0pK}{125}{]}.
-In metadata, semantic type systems (RDFS {[}\protect\hyperlink{ref-ZwAcGQKY}{63}{]}, OWL2 {[}\protect\hyperlink{ref-1p4IWJpI}{64}{]}, SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]}).
-OpenAPI 3 {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]} \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} {[}\protect\hyperlink{ref-11oVTMsuZ}{126}{]}, JSON \texttt{\$schema} {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]}), JSON-LD \texttt{@context} {[}\protect\hyperlink{ref-by6CfOmv}{127}{]}.
+Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})) & Syntactically Media Type with multiple suffixes (\protect\hyperlink{ref-nlH2rm1s}{{`draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes'}}) (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) (\protect\hyperlink{ref-nCzRO0pK}{Wilde, 2013}).
+In metadata, semantic type systems (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL2 (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})).
+OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
+XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} (\protect\hyperlink{ref-11oVTMsuZ}{{`W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures'}}), JSON \texttt{\$schema} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})), JSON-LD \texttt{@context} (\protect\hyperlink{ref-by6CfOmv}{{`JSON-LD 1.1'}}).
 Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
@@
 \textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} {[}\protect\hyperlink{ref-KbbW0kGT}{112}{]}, Web Sockets {[}\protect\hyperlink{ref-qgqADk45}{128}{]}, Linked Data Notifications {[}\protect\hyperlink{ref-zlGYiuWC}{129}{]}, AtomPub {[}\protect\hyperlink{ref-IbaebabD}{130}{]}, SWORD {[}\protect\hyperlink{ref-7DGllKzG}{131}{]}, Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} {[}\protect\hyperlink{ref-HE7Ikwwl}{132}{]}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates {[}\protect\hyperlink{ref-11hxwwuRt}{116}{]}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State {[}\protect\hyperlink{ref-174AwcFUL}{86}{]}, an important element of the REST architectural style.} incl.~Hydra {[}\protect\hyperlink{ref-13PaKBrZe}{133}{]}, schema.org Actions {[}{[}\protect\hyperlink{ref-dKAekUjL}{89}{]}), JSON HAL {[}\protect\hyperlink{ref-nyrPYHoP}{134}{]} \& Link headers (RFC8288) {[}\protect\hyperlink{ref-ozdWB3O7}{135}{]} \\
+Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
+\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
 \textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 {[}\protect\hyperlink{ref-tIkSXrb5}{136}{]} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) {[}\protect\hyperlink{ref-PL512B0X}{137}{]}, MPEG-DASH (ISO/IEC 23009-1:2019) {[}\protect\hyperlink{ref-ikO7e0mh}{138}{]} \\
+DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 (\protect\hyperlink{ref-tIkSXrb5}{Fielding \& Reschke, 2014b}) \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) (\protect\hyperlink{ref-PL512B0X}{May, 2017}), MPEG-DASH (ISO/IEC 23009-1:2019) (\protect\hyperlink{ref-ikO7e0mh}{14:00-17:00}) \\
 \bottomrule
@@
 \item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]} supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}.
+  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}) supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
+Multiplexing was removed for the much simplified DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
 Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
@@
 
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} {[}\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}{]}.
-
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform {[}\protect\hyperlink{ref-7szz7dwO}{100}{]} comply with the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}.
-Here we've used the RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-1GrKTFaK2}{139}{]} as it has decomposed the FAIR principles to a structured list of FAIR indicators {[}\protect\hyperlink{ref-UzQhqk0M}{84}{]}, importantly considering \emph{Data} and \emph{Metadata} separately.
+In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}).
+
+Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) comply with the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
+Here we've used the RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}) as it has decomposed the FAIR principles to a structured list of FAIR indicators (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}), importantly considering \emph{Data} and \emph{Metadata} separately.
 In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
@@
   >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-UzQhqk0M}{84},\protect\hyperlink{ref-1GrKTFaK2}{139}{]} (first 2 columns) against the FDO guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}, FDO implemented with the protocol DOIPv2 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, Linked Data Platform (LDP) {[}\protect\hyperlink{ref-7szz7dwO}{100}{]} and examples from Linked Data practices in general.
+\caption{Assessing RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}; \protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) (first 2 columns) against the FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), FDO implemented with the protocol DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) and examples from Linked Data practices in general.
 (--- indicates \emph{Unspecified}, may be possible with additional conventions)
@@
 & FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{[}\protect\hyperlink{ref-rbG9uRKw}{118}{]};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
+PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
 RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
@@
 & Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-{[}\protect\hyperlink{ref-1Gsq23e51}{140}{]}).
+(\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}, pp.
+https://lod--cloud.net/})).
 Indexing by search engines if exposing HTML w/schema.org.
@@
 Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP {[}\protect\hyperlink{ref-8HJqcF1Q}{98}{]} & HTTP/HTTPS, FTP (now less common), GridFTP {[}\protect\hyperlink{ref-18LsXvKGJ}{141}{]} (for large data), ARK {[}\protect\hyperlink{ref-voDQ6diM}{142}{]} \\
+RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}) & HTTP/HTTPS, FTP (now less common), GridFTP (\protect\hyperlink{ref-18LsXvKGJ}{Allcock et al.}) (for large data), ARK (\protect\hyperlink{ref-voDQ6diM}{{`The ARK Identifier Scheme'}}) \\
 RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol {[}\protect\hyperlink{ref-I9Kx0Hjy}{143}{]}.
-One server implementation {[}\protect\hyperlink{ref-1CvYRVhy2}{144}{]}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved {[}\protect\hyperlink{ref-xVJMmGVr}{145}{]} as an open source license -- it includes usage restrictions and requires Service Agreements.
+RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol (\protect\hyperlink{ref-I9Kx0Hjy}{Sun et al., 2003}).
+One server implementation (\protect\hyperlink{ref-1CvYRVhy2}{{`Handle.Net Registry'}}), free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved (\protect\hyperlink{ref-xVJMmGVr}{{`Licenses \& Standards \textbar{} Open Source Initiative'}}) as an open source license -- it includes usage restrictions and requires Service Agreements.
 It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
@@
 & FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) {[}\protect\hyperlink{ref-1BlxdxOwa}{146}{]} \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs {[}\protect\hyperlink{ref-o4h92LpS}{147}{]}), \texttt{rdfs:seeAlso}, FAIR signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]} \texttt{describedby} \\
+RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) (\protect\hyperlink{ref-1BlxdxOwa}{14:00-17:00}) \\
+RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs (\protect\hyperlink{ref-o4h92LpS}{{`Data - W3C'}})), \texttt{rdfs:seeAlso}, FAIR signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) \texttt{describedby} \\
 RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
@@
 RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles {[}\protect\hyperlink{ref-QpDmajc4}{148}{]}, \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role {[}\protect\hyperlink{ref-10HA0mMBx}{149}{]}, \texttt{prov:specializationOf} {[}\protect\hyperlink{ref-ZUgwvAHE}{150}{]}, OAI-ORE Proxy {[}\protect\hyperlink{ref-ID1pTGry}{151}{]}) \\
+RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles (\protect\hyperlink{ref-QpDmajc4}{{`Linking Across Provenance Bundles'}}), \href{https://schema.org/Role}{schema.org/Role} \\
+RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role (\protect\hyperlink{ref-10HA0mMBx}{Unknown}), \texttt{prov:specializationOf} (\protect\hyperlink{ref-ZUgwvAHE}{{`PROV-O: The PROV Ontology'}}), OAI-ORE Proxy (\protect\hyperlink{ref-ID1pTGry}{{`ORE Specification - Abstract Data Model'}})) \\
 RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} but not assigned PIDs yet.
+Kernel metadata attributes desired (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) but not assigned PIDs yet.
 & Unspecified.
@@
 & Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org {[}\protect\hyperlink{ref-10pzKCEUE}{152}{]}, Dublin Core Terms {[}\protect\hyperlink{ref-wAbZpOsr}{153}{]}, DCAT {[}\protect\hyperlink{ref-lnH0hEuh}{154}{]}, FOAF {[}\protect\hyperlink{ref-XmUY0cZB}{155}{]} \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} {[}\protect\hyperlink{ref-lnH0hEuh}{154}{]} \\
+Rough consensus on kernel metadata: schema.org (\protect\hyperlink{ref-10pzKCEUE}{{`Schema.org - Schema.org'}}), Dublin Core Terms (\protect\hyperlink{ref-wAbZpOsr}{{`DCMI Metadata Terms'}}), DCAT (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), FOAF (\protect\hyperlink{ref-XmUY0cZB}{{`FOAF Vocabulary Specification'}}) \\
+RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}) \\
 RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
@@
 Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O {[}\protect\hyperlink{ref-9T4j3N4e}{107}{]}, PAV {[}\protect\hyperlink{ref-1E4OfHyhI}{156}{]}, Dublin Core Terms {[}\protect\hyperlink{ref-hKblceo6}{157}{]} \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen {[}\protect\hyperlink{ref-zZTtJruI}{158}{]}) & --- & Common, e.g.~DCAT 2 {[}\protect\hyperlink{ref-zt8AAjcZ}{159}{]}, BioSchemas {[}\protect\hyperlink{ref-1F8S9FeHf}{160}{]} \\
+RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}}), PAV (\protect\hyperlink{ref-1E4OfHyhI}{Ciccarese et al., 2013}), Dublin Core Terms (\protect\hyperlink{ref-hKblceo6}{{`DCMI Metadata Terms'}, 2020}) \\
+RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen (\protect\hyperlink{ref-zZTtJruI}{Hardisty et al., 2022})) & --- & Common, e.g.~DCAT 2 (\protect\hyperlink{ref-zt8AAjcZ}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), BioSchemas (\protect\hyperlink{ref-1F8S9FeHf}{{`Bioschemas - Bioschemas'}}) \\
 RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal {[}\protect\hyperlink{ref-1Gsq23e51}{140}{]}, Darwin Core {[}\protect\hyperlink{ref-9jaVzCbL}{161}{]} \\
+RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal (\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}}), Darwin Core (\protect\hyperlink{ref-9jaVzCbL}{Wieczorek et al., 2012}) \\
 RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
@@
 \item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} has currently only a couple of implementations and is expressed informally.
+  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) has currently only a couple of implementations and is expressed informally.
 The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
@@
   Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-{[}\protect\hyperlink{ref-126uxr5pI}{34}{]} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
+(\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
 Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
@@
 For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
+FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
 In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
@@
 
-The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} in section 3.6 recommends:
+The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) in section 3.6 recommends:
 
@@
 Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} & DCAT, ++ \\
+Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
 Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
@@
 The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake {[}\protect\hyperlink{ref-kYjzRJ0H}{162}{]}.
+At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\protect\hyperlink{ref-kYjzRJ0H}{{`Semantic Web - XML2000 - slide "Architecture"'}}).
 However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
+. \emph{Available at} \href{https://blog.cloudflare.com/http-3-vs-http-2/}{\emph{https://blog.cloudflare.com/http-3-vs-http-2/}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
+. \emph{Available at} \href{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}{\emph{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 January 2023b).
+
+\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
+14:00-17:00.ISO/IEC 23009-1:2019.
+\emph{Available at} \href{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}{\emph{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}} (accessed 26 May 2022a).
+
+\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
+Allcock W, Bresnahan J, Kettimuthu R, Link M.
+The Globus Striped GridFTP Framework and Server.
+In: \emph{ACM/IEEE SC 2005 Conference (SC'05)}.
+IEEE,.
+DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
+Anders I, Blanchi Ch, Broeder D, Hellström M, Islam Sh, Jejkal Th, Lannom L, Peters K, Quick R, Schlemmer A, Schwardmann U, Soiland-Reyes S, Strawn G, Uytvanck D van, Weiland C, Wittenburg P, Zwölf C.
+2022a.
+\emph{\href{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}{FAIR digital object overview and specifications}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
+Anders I, Hellström M, Islam S, Jejkal T, Lannom L, Schwardmann U, Wittenburg P.
+2022b.
+\emph{\href{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}{FDO PID profiles \& attributes}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
+Bahim C, Casorrán-Amilburu C, Dekkers M, Herczog E, Loozen N, Repanas K, Russell K, Stall S.
+2020.
+The FAIR data maturity model: An approach to harmonise FAIR assessments.
+\emph{Data Science Journal} 19.
+DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
+Belshe M, Peon R. 2015. \emph{Hypertext Transfer Protocol Version 2 (HTTP/2)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
+Berners-Lee T, Fielding R, Masinter L. 2005. \emph{Uniform Resource Identifier (URI): Generic Syntax}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
+Berners-Lee T, Fischetti M. 1999. \emph{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}. San Francisco: HarperSanFrancisco.
+
+\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
+Bernstein A, Hendler J, Noy N. 2016. A new look at the semantic web. \emph{Communications of the ACM} 59:35--37. DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
+Bioschemas - Bioschemas. \emph{Available at} \href{http://bioschemas.org/}{\emph{http://bioschemas.org/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
+Bizer C, Heath T, Berners-Lee T.
+2009.
+Linked data - the story so far.
+\emph{International journal on Semantic Web and information systems} 5:1--22.
+DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
+Blanchi C, Broeder D, Jejkal T, Sharif I, Schlemmer A, Uytvanck D van, Wittenburg P.
+2022a.
+\emph{\href{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}{FDO -- upload of FDO}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
+Blanchi C, Hellström M, Lannom L, Schwardmann U, Wittenburg P.
+2022b.
+\emph{\href{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}{Implementation of attributes, types, profiles and registries}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
+Bonino Da Silva Santos LO, Wilkinson MD, Kuzniar A, Kaliyaperumal R, Thompson M, Dumontier M, Burger K.
+2016.
+\href{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}{FAIR data points supporting big data interoperability}.
+In: Zelm M, Doumeingts G, Mendonça JP eds.
+\emph{Enterprise interoperability in the digitized and networked factory of the future}.
+iSTE Press, 270--279.
+
+\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
+Bonino L, Wittenburg O, Carroll B, Hardisty A, Leggott M, Zwölf C.
+2019.
+\emph{\href{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}{FAIR digital object framework}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
+Carriero VA, Daquino M, Gangemi A, Nuzzolese AG, Peroni S, Presutti V, Tomasi F.
+2020.
+The landscape of ontology reuse approaches.
+In: Cota G, Daquino M, Pozzato GL eds.
+\emph{Applications and practices in ontology design, extraction, and reasoning}.
+Studies on the semantic web.
+IOS Press,.
+DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/SSW200033}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
+Ciccarese P, Soiland-Reyes S, Belhajjame K, Gray AJ, Goble C, Clark T.
+2013.
+PAV ontology: provenance, authoring and versioning.
+\emph{Journal of Biomedical Semantics} 4:37.
+DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
+Clemm G, Amsden J, Ellison T, Kaler C, Whitehead J.
+2002.
+\emph{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}.
+RFC Editor.
+DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
+Content negotiation - HTTP \textbar{} MDN.
+\emph{Available at} \href{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}{\emph{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content\_negotiation}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
+Corcho O, Eriksson M, Kurowski K, Ojsteršek M, Choirat C, Sanden M van de, Coppens F.
+2021.
+\emph{EOSC interoperability framework}.
+Publications Office of the EU.
+DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
+Data - W3C. \emph{Available at} \href{https://www.w3.org/standards/semanticweb/data}{\emph{https://www.w3.org/standards/semanticweb/data}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
+Data Catalog Vocabulary (DCAT) - Version 2.
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 January 2023b).
+
+\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
+Data Catalog Vocabulary (DCAT) - Version 2. \emph{Available at} \href{https://www.w3.org/TR/vocab-dcat-2/}{\emph{https://www.w3.org/TR/vocab-dcat-2/}} (accessed 26 May 2022a).
+
+\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
+Data Catalog Vocabulary (DCAT) - Version 3.
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
+Data Information View.
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
+DCMI Metadata Terms.
+2020.
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
+DCMI Metadata Terms.
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
+Delgado JCM.
+2016.
+An Interoperability Framework and Distributed Platform for Fast Data Applications.
+In: \emph{Data Science and Big Data Computing}.
+Springer International Publishing, 3--39.
+DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
+Designing a Linked Data developer experience.
+2018.
+\emph{Available at} \href{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}{\emph{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
+Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation. \emph{Available at} \href{https://www.dona.net/doipv1doc}{\emph{https://www.dona.net/doipv1doc}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
+DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
+DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
+DOIP and Examples --- Cordra documentation.
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip.html}{\emph{https://www.cordra.org/documentation/api/doip.html}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
+DOIP API for HTTP Clients --- Cordra documentation.
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
+draft-bhutton-json-schema-00.
+\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}{\emph{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
+draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes.
+\emph{Available at} \href{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}{\emph{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
+draft-ietf-quic-http-34.
+\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}{\emph{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
+draft-kelly-json-hal-08.
+\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}{\emph{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
+Duerst M, Suignard M. 2005. \emph{Internationalized Resource Identifiers (IRIs)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
+Dusseault L (ed.). 2007. \emph{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
+FAIR Digital Object Framework Documentation. \emph{Available at} \href{https://fairdigitalobjectframework.org/}{\emph{https://fairdigitalobjectframework.org/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
+FAIR Digital Objects Forum \textbar. \emph{Available at} \href{https://fairdo.org/}{\emph{https://fairdo.org/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
+Fensel D, Facca FM, Simperl E, Toma I. 2011. \emph{Semantic Web Services}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
+Fielding RT.
+2000.
+\href{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}{Architectural styles and the design of network-based software architectures}.
+\{THESIS\}.\{DOCTORAL\} Thesis.
+University of California, Irvine.
+
+\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
+Fielding R, Nottingham M, Reschke J (eds.). 2022. \emph{HTTP Semantics}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
+Fielding R, Reschke J (eds.). 2014b. \emph{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
+Fielding R, Reschke J (eds.). 2014a. \emph{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
+Fielding RT, Taylor RN, Erenkrantz JR, Gorlick MM, Whitehead J, Khare R, Oreizy P.
+2017.
+Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award).
+In: \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017}.
+New York, New York, USA: ACM Press, 4--14.
+DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
+FOAF Vocabulary Specification. \emph{Available at} \href{http://xmlns.com/foaf/spec/}{\emph{http://xmlns.com/foaf/spec/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
+Foundation D. 2018. \emph{\href{https://hdl.handle.net/0.DOIP/DOIPV2.0}{Digital object interface protocol specification, version 2.0}}. DONA foundation.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
+Gayo JEL, Prud'hommeaux E, Boneva I, Kontokostas D.
+2017.
+Validating RDF Data.
+\emph{Synthesis Lectures on the Semantic Web: Theory and Technology} 7:1--328.
+DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
+Goble C, Stevens R.
+2008.
+State of the nation in data integration for bioinformatics.
+\emph{Journal of Biomedical Informatics} 41:687--693.
+DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
+Gregorio J, de hOra B (eds.). 2007. \emph{The Atom Publishing Protocol}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
+Gregorio J, Fielding R, Hadley M, Nottingham M, Orchard D. 2012. \emph{URI Template}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
+Groth P, Loizou A, Gray AJG, Goble C, Harland L, Pettifer S.
+2014.
+API-centric Linked Data integration: The Open PHACTS Discovery Platform case study.
+\emph{Journal of Web Semantics} 29:12--18.
+DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
+Group RDAFDMMW. 2020. FAIR data maturity model: Specification and guidelines. \emph{Research Data Alliance}. DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
+Hardisty A, Brack P, Goble C, Livermore L, Scott B, Groom Q, Owen S, Soiland-Reyes S. 2022. The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
+Digital Object Approach to Speeding up Digital Mobilisation of Natural History
+Collections. \emph{Data Intelligence} 4:320--341. DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
+Hasnain A, Rebholz-Schuhmann D.
+2018.
+Assessing FAIR data principles against the 5-star open data principles.
+In: Gangemi A, Gentile AL, Nuzzolese AG, Rudolph S, Maleshkova M, Paulheim H, Pan JZ, Alam M eds.
+\emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers}.
+Lecture notes in computer science.
+Cham: Springer International Publishing, 469--477.
+DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
+Horrocks I, Hendler J (eds.). 2002. \emph{The Semantic Web --- ISWC 2002}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
+HTML Standard. \emph{Available at} \href{https://html.spec.whatwg.org/multipage/microdata.html}{\emph{https://html.spec.whatwg.org/multipage/microdata.html}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
+Hu W, Chen J, Zhang H, Qu Y.
+2011.
+How matchable are four thousand ontologies on the semantic web.
+In: Antoniou G, Grobelnik M, Simperl E, Parsia B, Plexousakis D, De Leenheer P, Pan J eds.
+\emph{The semantic web: Research and applications}.
+Lecture notes in computer science.
+Berlin, Heidelberg: Springer Berlin Heidelberg, 290--304.
+DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
+Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI}{\emph{https://www.w3.org/Provider/Style/URI}} (accessed 26 May 2022a).
+
+\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
+Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI.html}{\emph{https://www.w3.org/Provider/Style/URI.html}} (accessed 26 May 2022b).
+
+\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
+Iyengar J, Thomson M (eds.). 2021. \emph{QUIC: A UDP-Based Multiplexed and Secure Transport}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
+JSON-LD 1.1. \emph{Available at} \href{http://www.w3.org/TR/json-ld/}{\emph{http://www.w3.org/TR/json-ld/}} (accessed 26 May 2022b).
+
+\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
+JSON-LD 1.1. \emph{Available at} \href{https://www.w3.org/TR/json-ld/}{\emph{https://www.w3.org/TR/json-ld/}} (accessed 26 May 2022a).
+
+\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
+Juty N, Le Novere N, Laibe C.
+2011.
+Identifiers.org and MIRIAM Registry: community resources to provide persistent identification.
+\emph{Nucleic Acids Research} 40:D580--D586.
+DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
+Kahn R, Wilensky R. 1995. \emph{\href{http://www.cnri.reston.va.us/k-w.html}{A framework for distributed digital object services}}. CNRI.
+
+\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
+Kahn R, Wilensky R.
+2006.
+A framework for distributed digital object services.
+\emph{International Journal on Digital Libraries} 6:115--123.
+DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
+Kamdar MR, Tudorache T, Musen MA.
+2017.
+A systematic analysis of term reuse and term overlap across biomedical ontologies.
+\emph{Semantic Web} 8:853--871.
+DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
+Khare R, Lawrence S. 2000. \emph{Upgrading to TLS Within HTTP/1.1}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
+Klein M, Van de Sompel H, Sanderson R, Shankar H, Balakireva L, Zhou K, Tobin R.
+2014.
+Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot.
+\emph{PLoS ONE} 9:e115253.
+DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
+Klímek J, Škoda P, Nečaský M. 2019. Survey of tools for linked data consumption. \emph{Satya Widya} 10:665--720. DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/SW-180316}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
+Lamprecht A-L, Palmblad M, Ison J, Schwämmle V, Al Manir MS, Altintas I, Baker CJO, Ben Hadj Amor A, Capella-Gutierrez S, Charonyktakis P, Crusoe MR, Gil Y, Goble C, Griffin TJ, Groth P, Ienasescu H, Jagtap P, Kalaš M, Kasalica V, Khanteymoori A, Kuhn T, Mei H, Ménager H, Möller S, Richardson RA, Robert V, Soiland-Reyes S, Stevens R, Szaniszlo S, Verberne S, Verhoeven A, Wolstencroft K.
+2021.
+Perspectives on automated composition of workflows in the life sciences.
+\emph{F1000Research} 10:897.
+DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
+Lannom L, Peters-von Gehlen K, Anders I, Pfeil A, Schlemmer A, Trautt Z, Wittenburg P.
+2022a.
+\emph{\href{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}{FDO configuration types}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
+Lannom L, Schwardmann U, Blanchi C, Wittenburg P.
+2022b.
+\emph{\href{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}{Typing FAIR digital objects}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
+Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
+Linked Data - Design Issues. \emph{Available at} \href{https://www.w3.org/DesignIssues/LinkedData.html}{\emph{https://www.w3.org/DesignIssues/LinkedData.html}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
+Linked Data Notifications. \emph{Available at} \href{https://www.w3.org/TR/ldn/}{\emph{https://www.w3.org/TR/ldn/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
+Linked Data Platform Working Group. 2015. \emph{\href{https://www.w3.org/TR/2015/REC-ldp-20150226/}{Linked data platform 1.0}}. W3C.
+
+\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
+Linking Across Provenance Bundles.
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
+Loo T (ed.). 1970. First International Conference on FAIR Digital Objects. DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
+martinekuan.Web API design best practices - Azure Architecture Center.
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
+May W. 2017. \emph{HTTP Live Streaming}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
+Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
+2021b.
+Web data apis over SPARQL.
+In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
+Synthesis lectures on data, semantics, and knowledge.
+Cham: Springer International Publishing, 27--38.
+DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
+Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
+2021a.
+Conclusion and future challenges.
+In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
+Synthesis lectures on data, semantics, and knowledge.
+Cham: Springer International Publishing, 77--80.
+DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
+Mika P. 2015. On Schema.org and Why It Matters for the Web. \emph{IEEE Internet Computing} 19:52--55. DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
+Mons B, Neylon C, Velterop J, Dumontier M, Silva Santos LOB da, Wilkinson MD.
+2017.
+Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud.
+\emph{Information Services \& Use} 37:49--56.
+DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{ISU\}-170824}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
+NCBO BioPortal. \emph{Available at} \href{https://bioportal.bioontology.org/ontologies}{\emph{https://bioportal.bioontology.org/ontologies}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
+Neumann A, Laranjeiro N, Bernardino J.
+2021.
+An analysis of public REST web service apis.
+\emph{IEEE Transactions on Services Computing} 14:957--970.
+DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/TSC.2018.2847344}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
+Nottingham M. 2017. \emph{Web Linking}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
+OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More.
+\emph{Available at} \href{https://spec.openapis.org/oas/v3.1.0.html}{\emph{https://spec.openapis.org/oas/v3.1.0.html}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
+ORE Specification - Abstract Data Model.
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
+OWL 2 Web Ontology Language Document Overview (Second Edition). \emph{Available at} \href{http://www.w3.org/TR/owl2-overview/}{\emph{http://www.w3.org/TR/owl2-overview/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
+Page KR, De Roure DC, Martinez K.
+2011.
+REST and Linked Data.
+In: \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11}.
+ACM Press,.
+DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
+Polleres A, Kamdar MR, Fernández JD, Tudorache T, Musen MA.
+2020.
+A more decentralized vision for linked data.
+\emph{Satya Widya} 11:101--113.
+DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/SW-190380}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 January 2023a).
+
+\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 January 2023b).
+
+\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
+RDF 1.1 Primer. \emph{Available at} \href{http://www.w3.org/TR/rdf11-primer/}{\emph{http://www.w3.org/TR/rdf11-primer/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
+RDF Schema 1.1. \emph{Available at} \href{http://www.w3.org/TR/rdf-schema/}{\emph{http://www.w3.org/TR/rdf-schema/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
+RDFa 1.1 Primer - Third Edition. \emph{Available at} \href{http://www.w3.org/TR/rdfa-primer/}{\emph{http://www.w3.org/TR/rdfa-primer/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
+Resource Description Framework (RDF) Model and Syntax Specification.
+\emph{Available at} \href{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}{\emph{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
+ResourceSync Framework Specification - Table of Contents. \emph{Available at} \href{http://www.openarchives.org/rs/toc}{\emph{http://www.openarchives.org/rs/toc}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
+Sandvine.Global Internet Phenomena Report 2022.
+\emph{Available at} \href{https://www.sandvine.com/global-internet-phenomena-report-2022}{\emph{https://www.sandvine.com/global-internet-phenomena-report-2022}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
+Sauermann L, Cyganiak R, Völkel M. 2011. \emph{Cool URIs for the semantic web}. Universität des Saarlandes. DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
+Schema.org - Schema.org. \emph{Available at} \href{https://schema.org/}{\emph{https://schema.org/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
+Schema.org Actions - schema.org. \emph{Available at} \href{https://schema.org/docs/actions.html}{\emph{https://schema.org/docs/actions.html}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
+Schultes E, Wittenburg P.
+2019.
+FAIR principles and digital objects: Accelerating convergence on a data infrastructure.
+In: Manolopoulos Y, Stupnikov S eds.
+\emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers}.
+Communications in computer and information science.
+Cham: Springer International Publishing, 3--16.
+DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
+Schwardmann U, Kálmán T.
+2022.
+Two Examples on How FDO Types can Support Machine and Human Readability.
+\emph{Research Ideas and Outcomes} 8.
+DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
+Semantic Web - XML2000 - slide "Architecture".
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
+Shape Expressions (ShEx) 2.1 Primer. \emph{Available at} \href{http://shex.io/shex-primer/}{\emph{http://shex.io/shex-primer/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
+Shapes Constraint Language (SHACL). \emph{Available at} \href{https://www.w3.org/TR/shacl/}{\emph{https://www.w3.org/TR/shacl/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
+SKOS Simple Knowledge Organization System Primer. \emph{Available at} \href{http://www.w3.org/TR/skos-primer/}{\emph{http://www.w3.org/TR/skos-primer/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
+Smith B, Ashburner M, Rosse C, Bard J, Bug W, Ceusters W, Goldberg LJ, Eilbeck K, Ireland A, Mungall CJ, Leontis N, Rocca-Serra P, Ruttenberg A, Sansone S-A, Scheuermann RH, Shah N, Whetzel PL, Lewis S.
+2007.
+The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration.
+\emph{Nature Biotechnology} 25:1251--1255.
+DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
+SPARQL 1.1 Overview. \emph{Available at} \href{http://www.w3.org/TR/sparql11-overview/}{\emph{http://www.w3.org/TR/sparql11-overview/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
+Stallings W. 1990. \emph{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}. Carmel, IN, USA: Sams.
+
+\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
+Stanczyk SK. 1987. Process modelling for information system description. \emph{The Open University}. DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
+Stefi A.
+2015.
+Do Developers Make Unbiased Decisions?
+- The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components.
+DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
+Stefi A, Hess T.
+2015.
+To develop or to reuse?
+Two perspectives on external reuse in software projects.
+In: Fernandes JM, Machado RJ, Wnuk K eds.
+\emph{Software business}.
+Lecture notes in business information processing.
+Cham: Springer International Publishing, 192--206.
+DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
+Strawn G, Wittenburg P, Soiland-Reyes St, Peters K, Lannom L, Schwardmann U, Blanch Ch.
+2022.
+\emph{\href{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}{FDO Forum FDO requirement specifications}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
+Sun S, Lannom L, Boesch B. 2003. \emph{Handle System Overview}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
+Sun S, Reilly S, Lannom L, Petrone J. 2003. \emph{Handle System Protocol (ver 2.1) Specification}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
+SWORD 3.0 Specification. \emph{Available at} \href{https://swordapp.github.io/swordv3/swordv3.html}{\emph{https://swordapp.github.io/swordv3/swordv3.html}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
+The ARK Identifier Scheme.
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
+The Modern Standards Paradigm - Five Key Principles.
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
+The Open Graph protocol. \emph{Available at} \href{https://ogp.me/}{\emph{https://ogp.me/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
+Thornton K, Solbrig H, Stupp GS, Labra Gayo JE, Mietchen D, Prud E, Waagmeester A.
+2019.
+Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation.
+In: Hitzler P, Fernández M, Janowicz K, Zaveri A, Gray AJG, Lopez V, Haller A, Hammar K eds.
+\emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings}.
+Lecture notes in computer science.
+Cham: Springer International Publishing, 606--620.
+DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
+Tirmizi S, Aitken S, Moreira DA, Mungall C, Sequeda J, Shah NH, Miranker DP.
+2011.
+Mapping between the OBO and OWL ontology languages.
+\emph{Journal of Biomedical Semantics} 2:S3.
+DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
+Turcoane O.
+2014.
+Linked data, JSON-LD and the semantics of cultural and scientific heritage.
+\emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} 4:95--105.
+DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
+Usage Statistics of JSON-LD for Websites, May 2022.
+\emph{Available at} \href{https://w3techs.com/technologies/details/da-jsonld}{\emph{https://w3techs.com/technologies/details/da-jsonld}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
+Van de Sompel H, Klein M, Jones S, Nelson ML, Warner S, Devaraju A, Huber R, Steinhoff W, Tykhonov V, Boruta L, Meijers E, Soiland-Reyes S, Wilkinson M.
+2022.FAIR Signposting Profile.
+\emph{Available at} \href{https://signposting.org/FAIR/}{\emph{https://signposting.org/FAIR/}} (accessed 5 January 2023).
+
+\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
+Van de Sompel H, Nelson M, Sanderson R. 2013. \emph{HTTP Framework for Time-Based Access to Resource States -\/- Memento}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
+Verborgh R, Vander Sande M.
+2020.
+The semantic web identity crisis: In search of the trivialities that never were.
+\emph{Satya Widya} 11:19--27.
+DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/SW-190372}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
+W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures. \emph{Available at} \href{http://www.w3.org/TR/xmlschema11-1/}{\emph{http://www.w3.org/TR/xmlschema11-1/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
+Web Services Description Language (WSDL) Version 2.0 Part 0: Primer. \emph{Available at} \href{http://www.w3.org/TR/wsdl20-primer/}{\emph{http://www.w3.org/TR/wsdl20-primer/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
+WebSockets Standard. \emph{Available at} \href{https://websockets.spec.whatwg.org/}{\emph{https://websockets.spec.whatwg.org/}} (accessed 26 May 2022).
+
+\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
+Weigel T, Plale B, Parsons M, Zhou G, Luo Y, Schwardmann U, Quick R, Hellström M, Kurakawa K.
+2018.
+RDA Recommendation on PID Kernel Information.
+\emph{Research Data Alliance}.
+DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
+Weiland C, Islam S, Broder D, Anders I, Wittenburg P.
+2022a.
+\emph{\href{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}{FDO machine actionability}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
+Weiland C, Schwardmann U, Wittenburg P, Kirkpatrick C, Hanisch R, Trautt Z.
+2022b.
+\emph{\href{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}{FDO Forum Document Standards}}.
+(internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
+Wieczorek J, Bloom D, Guralnick R, Blum S, Döring M, Giovanni R, Robertson T, Vieglais D.
+2012.
+Darwin Core: An Evolving Community-Developed Biodiversity Data Standard.
+\emph{PLoS ONE} 7:e29715.
+DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
+Wilde E. 2013. \emph{The 'profile' Link Relation Type}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-6DjakjNS}{}}%
-\CSLLeftMargin{1. }
-\CSLRightInline{\textbf{The FAIR Guiding Principles for scientific data management and stewardship}
-\CSLBlock{Mark D Wilkinson, Michel Dumontier, IJsbrand Jan Aalbersberg, Gabrielle Appleton, Myles Axton, Arie Baak, Niklas Blomberg, Jan-Willem Boiten, Luiz Bonino da Silva Santos, Philip E Bourne, \ldots{} Barend Mons} \emph{Scientific Data} (2016-03-15)
-\CSLBlock{DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/26978244}{26978244} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4792175}{PMC4792175}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
-\CSLLeftMargin{2. }
-\CSLRightInline{\textbf{Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud}
-\CSLBlock{Barend Mons, Cameron Neylon, Jan Velterop, Michel Dumontier, Luiz Olavo Bonino da Silva Santos, Mark D Wilkinson} \emph{Information Services \& Use} (2017-03-07)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{isu\}-170824}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
-\CSLLeftMargin{3. }
-\CSLRightInline{\textbf{EOSC interoperability framework}
-\CSLBlock{Oscar Corcho, Magnus Eriksson, Krzysztof Kurowski, Milan Ojsteršek, Christine Choirat, Mark van de Sanden, Frederik Coppens} \emph{Publications Office of the EU} (2021-02-05) \url{https://doi.org/10.2777/620649}
-\CSLBlock{DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
-\CSLLeftMargin{4. }
-\CSLRightInline{\textbf{FAIR principles and digital objects: Accelerating convergence on a data infrastructure}
-\CSLBlock{Erik Schultes, Peter Wittenburg} \emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers} (2019)
-\CSLBlock{DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}} · ISBN: 978-3-030-23583-3}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
-\CSLLeftMargin{5. }
-\CSLRightInline{\textbf{Linked data - the story so far}
-\CSLBlock{Christian Bizer, Tom Heath, Tim Berners-Lee} \emph{International journal on Semantic Web and information systems} (2009-07) \url{http://services.igi-global.com/resolvedoi/resolve.aspx?doi=10.4018/jswis.2009081901}
-\CSLBlock{DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
-\CSLLeftMargin{6. }
-\CSLRightInline{\textbf{Assessing FAIR data principles against the 5-star open data principles}
-\CSLBlock{Ali Hasnain, Dietrich Rebholz-Schuhmann} \emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers} (2018) \url{http://link.springer.com/10.1007/978-3-319-98192-5/_60}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60} · ISBN: 978-3-319-98191-8}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
-\CSLLeftMargin{7. }
-\CSLRightInline{\textbf{FAIR data points supporting big data interoperability}
-\CSLBlock{Luiz Olavo Bonino Da Silva Santos, Mark D Wilkinson, Arnold Kuzniar, Rajaram Kaliyaperumal, Mark Thompson, Michel Dumontier, Kees Burger} \emph{Enterprise interoperability in the digitized and networked factory of the future} (2016) \url{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}
-\CSLBlock{ISBN: 9781847040442}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
-\CSLLeftMargin{8. }
-\CSLRightInline{\textbf{The semantic web identity crisis: In search of the trivialities that never were}
-\CSLBlock{Ruben Verborgh, Miel Vander Sande} \emph{Satya Widya} (2020-01-31)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/sw-190372}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
-\CSLLeftMargin{9. }
-\CSLRightInline{\textbf{Conclusion and future challenges}
-\CSLBlock{Albert Meroño-Peñuela, Pasquale Lisena, Carlos Martínez-Ortiz} \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers} (2021)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7} · ISBN: 978-3-031-00789-7}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
-\CSLLeftMargin{10. }
-\CSLRightInline{\textbf{An analysis of public REST web service apis}
-\CSLBlock{Andy Neumann, Nuno Laranjeiro, Jorge Bernardino} \emph{IEEE Transactions on Services Computing} (2021-07-01)
-\CSLBlock{DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/tsc.2018.2847344}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
-\CSLLeftMargin{11. }
-\CSLRightInline{\textbf{Linked data, JSON-LD and the semantics of cultural and scientific heritage}
-\CSLBlock{Ovidiu Turcoane} \emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} (2014-09-30)
-\CSLBlock{DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}}}
+Wilkinson MD, Dumontier M, Aalbersberg IjJ, Appleton G, Axton M, Baak A, Blomberg N, Boiten J-W, da Silva Santos LB, Bourne PE, Bouwman J, Brookes AJ, Clark T, Crosas M, Dillo I, Dumon O, Edmunds S, Evelo CT, Finkers R, Gonzalez-Beltran A, Gray AJG, Groth P, Goble C, Grethe JS, Heringa J, 't Hoen PAC, Hooft R, Kuhn T, Kok R, Kok J, Lusher SJ, Martone ME, Mons A, Packer AL, Persson B, Rocca-Serra P, Roos M, van Schaik R, Sansone S-A, Schultes E, Sengstag T, Slater T, Strawn G, Swertz MA, Thompson M, van der Lei J, van Mulligen E, Velterop J, Waagmeester A, Wittenburg P, Wolstencroft K, Zhao J, Mons B.
+2016.
+The FAIR Guiding Principles for scientific data management and stewardship.
+\emph{Scientific Data} 3.
+DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Pa67pUtR}{}}%
-\CSLLeftMargin{12. }
-\CSLRightInline{\textbf{F*** workflows: when parts of FAIR are missing}
-\CSLBlock{Sean R Wilkinson, Greg Eisenhauer, Anuj J Kapadia, Kathryn Knight, Jeremy Logan, Patrick Widener, Matthew Wolf} \emph{arXiv} (2022) \url{https://doi.org/grk3v5}
-\CSLBlock{DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
-\CSLLeftMargin{13. }
-\CSLRightInline{\textbf{Do Developers Make Unbiased Decisions? - The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components}
-\CSLBlock{Anisa Stefi} \emph{University of Münster, Münster, Germany} (2015) \url{https://doi.org/grcpwg}
-\CSLBlock{DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
-\CSLLeftMargin{14. }
-\CSLRightInline{\textbf{To develop or to reuse? Two perspectives on external reuse in software projects}
-\CSLBlock{Anisa Stefi, Thomas Hess} \emph{Software business} (2015)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18} · ISBN: 978-3-319-19592-6}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
-\CSLLeftMargin{15. }
-\CSLRightInline{\textbf{Web data apis over SPARQL}
-\CSLBlock{Albert Meroño-Peñuela, Pasquale Lisena, Carlos Martínez-Ortiz} \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers} (2021)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3} · ISBN: 978-3-031-00789-7}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
-\CSLLeftMargin{16. }
-\CSLRightInline{\textbf{FAIR Digital Objects Forum \textbar{}} \url{https://fairdo.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
-\CSLLeftMargin{17. }
-\CSLRightInline{\textbf{FAIR digital object overview and specifications}
-\CSLBlock{I Anders, Ch Blanchi, Daan Broeder, M Hellström, Sh Islam, Th Jejkal, L Lannom, K Peters, R Quick, A Schlemmer, \ldots{} Carlo Zwölf} \emph{FDO Forum} (2022-11-19) \url{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
-\CSLLeftMargin{18. }
-\CSLRightInline{\textbf{FDO Forum Document Standards}
-\CSLBlock{C Weiland, U Schwardmann, P Wittenburg, C Kirkpatrick, R Hanisch, Z Trautt} \emph{FDO Forum} (2022-01-29) \url{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
-\CSLLeftMargin{19. }
-\CSLRightInline{\textbf{FDO Forum FDO requirement specifications}
-\CSLBlock{G Strawn, P Wittenburg, St Soiland-Reyes, K Peters, L Lannom, U Schwardmann, Ch Blanch} \emph{FDO Forum} (2022-08-22) \url{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
-\CSLLeftMargin{20. }
-\CSLRightInline{\textbf{FAIR digital object framework}
-\CSLBlock{Luiz Bonino, Oeter Wittenburg, Bonnie Carroll, Alex Hardisty, Mark Leggott, Carlo Zwölf} \emph{FDOF technical implementation guideline} (2019-11-22) \url{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
-\CSLLeftMargin{21. }
-\CSLRightInline{\textbf{FDO machine actionability}
-\CSLBlock{Claus Weiland, Sharif Islam, Daan Broder, Ivonne Anders, Peter Wittenburg} \emph{FDO Forum} (2022-06-11) \url{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
-\CSLLeftMargin{22. }
-\CSLRightInline{\textbf{FDO configuration types}
-\CSLBlock{Larry Lannom, Karsten Peters-von Gehlen, Ivonne Anders, Andreas Pfeil, Alexander Schlemmer, Zach Trautt, Peter Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
-\CSLLeftMargin{23. }
-\CSLRightInline{\textbf{FAIR Signposting Profile}
-\CSLBlock{Herbert Van de Sompel, Martin Klein, Shawn Jones, Michael L Nelson, Simeon Warner, Anusuriya Devaraju, Robert Huber, Wilko Steinhoff, Vyacheslav Tykhonov, Luc Boruta, \ldots{} Mark Wilkinson} (2022-07-27) \url{https://signposting.org/FAIR/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
-\CSLLeftMargin{24. }
-\CSLRightInline{\textbf{FDO PID profiles \& attributes}
-\CSLBlock{Ivonne Anders, Maggie Hellström, Sharif Islam, Thomas Jejkal, Larry Lannom, Ulrich Schwardmann, Peter Wittenburg} \emph{FDO Forum} (2022-06-07) \url{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
-\CSLLeftMargin{25. }
-\CSLRightInline{\textbf{RDA Recommendation on PID Kernel Information}
-\CSLBlock{Tobias Weigel, Beth Plale, Mark Parsons, Gabriel Zhou, Yu Luo, Ulrich Schwardmann, Robert Quick, Margareta Hellström, Kei Kurakawa} \emph{Research Data Alliance} (2018)
-\CSLBlock{DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}}}
+Wilkinson SR, Eisenhauer G, Kapadia AJ, Knight K, Logan J, Widener P, Wolf M.
+2022.
+F*** workflows: when parts of FAIR are missing.
+\emph{arXiv}.
+DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
+Williams AJ, Harland L, Groth P, Pettifer S, Chichester C, Willighagen EL, Evelo CT, Blomberg N, Ecker G, Goble C, Mons B.
+2012.
+Open PHACTS: semantic interoperability for drug discovery.
+\emph{Drug Discovery Today} 17:1188--1198.
+DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
+Wittenburg P, Anders I, Blanchi C, Buurman M, Goble C, Grieb J, Hardisty A, Islam S, Jejkal T, Kálmán T, Kirkpatrick C, Lannom L, Lauer T, Manepalli G, Peters-von Gehlen K, Pfeil A, Quick R, Sanden M van de, Schwardmann U, Soiland-Reyes S, Stotzka R, Trautt Z, Van Uytvanck D, Weiland C, Wieder P.
+2022.
+FAIR digital object demonstrators 2021.
+\emph{Zenodo}.
+DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
+Wittenburg P, Strawn G, Mons B, Bonino L, Schultes E.
+2019.
+Digital objects as drivers towards convergence in data infrastructures.
+\emph{https://b2share.eudat.eu}.
+DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
+Wolstencroft K, Haines R, Fellows D, Williams A, Withers D, Owen S, Soiland-Reyes S, Dunlop I, Nenadic A, Fisher P, Bhagat J, Belhajjame K, Bacall F, Hardisty A, Nieva de la Hidalga A, Balcazar Vargas MP, Sufi S, Goble C.
+2013.
+The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud.
+\emph{Nucleic Acids Research} 41:W557--W561.
+DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
+Wolstencroft K, Owen S, Horridge M, Krebs O, Mueller W, Snoep JL, du Preez F, Goble C.
+2011.
+RightField: embedding ontology annotation in spreadsheets.
+\emph{Bioinformatics} 27:2021--2022.
+DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
+Working Group FDO-TSIG. 2022a. \emph{\href{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}{DOIP endorsement request}}. (internal draft): FDO Forum.
+
+\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
+Working Group FDO-TSIG.
+2022b.
+\emph{\href{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}{FDO -- granularity, versioning, mutability}}.
+(internal draft): FDO Forum.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Fi62cJAo}{}}%
-\CSLLeftMargin{26. }
-\CSLRightInline{\textbf{FDO -- kernel attributes \& metadata}
-\CSLBlock{FDO-TSIG Working Group, FDO-BIG Working Group} \emph{FDO Forum} (2022-08-28) \url{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
-\CSLLeftMargin{27. }
-\CSLRightInline{\textbf{FDO -- granularity, versioning, mutability}
-\CSLBlock{FDO-TSIG Working Group} \emph{FDO Forum} (2022-08-26) \url{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
-\CSLLeftMargin{28. }
-\CSLRightInline{\textbf{DOIP endorsement request}
-\CSLBlock{FDO-TSIG Working Group} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
-\CSLLeftMargin{29. }
-\CSLRightInline{\textbf{Digital object interface protocol specification, version 2.0}
-\CSLBlock{DONA Foundation} \emph{DONA foundation} (2018-11-12) \url{https://hdl.handle.net/0.DOIP/DOIPV2.0}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
-\CSLLeftMargin{30. }
-\CSLRightInline{\textbf{FAIR digital object demonstrators 2021}
-\CSLBlock{Peter Wittenburg, Ivonne Anders, Christophe Blanchi, Merret Buurman, Carole Goble, Jonas Grieb, Alex Hardisty, Sharif Islam, Thomas Jejkal, Tibor Kálmán, \ldots{} Philipp Wieder} \emph{Zenodo} (2022)
-\CSLBlock{DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
-\CSLLeftMargin{31. }
-\CSLRightInline{\textbf{FDO -- upload of FDO}
-\CSLBlock{Christophe Blanchi, Daan Broeder, Thomas Jejkal, Islam Sharif, Alexander Schlemmer, Dieter van Uytvanck, Peter Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
-\CSLLeftMargin{32. }
-\CSLRightInline{\textbf{ResourceSync Framework Specification - Table of Contents} \url{http://www.openarchives.org/rs/toc}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
-\CSLLeftMargin{33. }
-\CSLRightInline{\textbf{Typing FAIR digital objects}
-\CSLBlock{Larry Lannom, U Schwardmann, C Blanchi, P Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
-\CSLLeftMargin{34. }
-\CSLRightInline{\textbf{Implementation of attributes, types, profiles and registries}
-\CSLBlock{C Blanchi, M Hellström, L Lannom, U Schwardmann, P Wittenburg} \emph{FDO Forum} (2022-11-27) \url{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
-\CSLLeftMargin{35. }
-\CSLRightInline{\textbf{Two Examples on How FDO Types can Support Machine and Human Readability}
-\CSLBlock{Ulrich Schwardmann, Tibor Kálmán} \emph{Research Ideas and Outcomes} (2022-10-12) \url{https://doi.org/grcq23}
-\CSLBlock{DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
-\CSLLeftMargin{36. }
-\CSLRightInline{\textbf{A framework for distributed digital object services}
-\CSLBlock{Robert Kahn, Robert Wilensky} \emph{International Journal on Digital Libraries} (2006-04) \url{https://www.doi.org/topics/2006_05_02_Kahn_Framework.pdf}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
-\CSLLeftMargin{37. }
-\CSLRightInline{\textbf{A framework for distributed digital object services}
-\CSLBlock{Robert Kahn, Robert Wilensky} \emph{CNRI} (1995-05-13) \url{http://www.cnri.reston.va.us/k-w.html}}
+Working Group FDO-TSIG, Working Group FDO-BIG.
+2022.
+\emph{\href{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}{FDO -- kernel attributes \& metadata}}.
+(internal draft): FDO Forum.
 
 \leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
-\CSLLeftMargin{38. }
-\CSLRightInline{\textbf{X.1255 : Framework for discovery of identity management information} \url{https://www.itu.int/rec/T-REC-X.1255-201309-I}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
-\CSLLeftMargin{39. }
-\CSLRightInline{\textbf{Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation} \url{https://www.dona.net/doipv1doc}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
-\CSLLeftMargin{40. }
-\CSLRightInline{\textbf{Digital objects as drivers towards convergence in data infrastructures}
-\CSLBlock{Peter Wittenburg, George Strawn, Barend Mons, Luiz Bonino, Erik Schultes} \emph{https://b2share.eudat.eu} (2019-01-06)
-\CSLBlock{DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
-\CSLLeftMargin{41. }
-\CSLRightInline{\textbf{DOIP and Examples --- Cordra documentation} \url{https://www.cordra.org/documentation/api/doip.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
-\CSLLeftMargin{42. }
-\CSLRightInline{\textbf{draft-bhutton-json-schema-00} \url{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
-\CSLLeftMargin{43. }
-\CSLRightInline{\textbf{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}
-\CSLBlock{Tim Berners-Lee, Mark Fischetti} \emph{HarperSanFrancisco} (1999)
-\CSLBlock{ISBN: 9780062515865}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
-\CSLLeftMargin{44. }
-\CSLRightInline{\textbf{RDF 1.1 Primer} \url{http://www.w3.org/TR/rdf11-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
-\CSLLeftMargin{45. }
-\CSLRightInline{\textbf{Resource Description Framework (RDF) Model and Syntax Specification} \url{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
-\CSLLeftMargin{46. }
-\CSLRightInline{\textbf{Process modelling for information system description}
-\CSLBlock{Stefan K Stanczyk} \emph{The Open University} (1987)
-\CSLBlock{DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
-\CSLLeftMargin{47. }
-\CSLRightInline{\textbf{Uniform Resource Identifier (URI): Generic Syntax}
-\CSLBlock{T Berners-Lee, R Fielding, L Masinter} \emph{RFC Editor} (2005-01) \url{https://doi.org/ggqvpr}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-\CSLLeftMargin{48. }
-\CSLRightInline{\textbf{"info" URI Registry (Frozen)} \url{https://oclc-research.github.io/infoURI-Frozen/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
-\CSLLeftMargin{49. }
-\CSLRightInline{\textbf{Identifiers.org and MIRIAM Registry: community resources to provide persistent identification}
-\CSLBlock{N Juty, N Le Novere, C Laibe} \emph{Nucleic Acids Research} (2011-12-02) \url{https://doi.org/cx2776}
-\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22140103}{22140103} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3245029}{PMC3245029}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
-\CSLLeftMargin{50. }
-\CSLRightInline{\textbf{Internationalized Resource Identifiers (IRIs)}
-\CSLBlock{M Duerst, M Suignard} \emph{RFC Editor} (2005-01) \url{https://doi.org/gjvnbg}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
-\CSLLeftMargin{51. }
-\CSLRightInline{\textbf{Cool URIs for the semantic web}
-\CSLBlock{Leo Sauermann, Richard Cyganiak, Max Völkel} \emph{Universität des Saarlandes} (2011-07-13)
-\CSLBlock{DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
-\CSLLeftMargin{52. }
-\CSLRightInline{\textbf{The Semantic Web --- ISWC 2002}
-\CSLBlock{Ian Horrocks, James Hendler (editors)} \emph{Lecture Notes in Computer Science} (2002) \url{https://doi.org/fwjmhp}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
-\CSLLeftMargin{53. }
-\CSLRightInline{\textbf{SPARQL 1.1 Overview} \url{http://www.w3.org/TR/sparql11-overview/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
-\CSLLeftMargin{54. }
-\CSLRightInline{\textbf{State of the nation in data integration for bioinformatics.}
-\CSLBlock{Carole Goble, Robert Stevens} \emph{Journal of Biomedical Informatics} (2008-10-01)
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/18358788}{18358788}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
-\CSLLeftMargin{55. }
-\CSLRightInline{\textbf{Open PHACTS: semantic interoperability for drug discovery}
-\CSLBlock{Antony J Williams, Lee Harland, Paul Groth, Stephen Pettifer, Christine Chichester, Egon L Willighagen, Chris T Evelo, Niklas Blomberg, Gerhard Ecker, Carole Goble, Barend Mons} \emph{Drug Discovery Today} (2012-11) \url{https://doi.org/f4fdkd}
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22683805}{22683805}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
-\CSLLeftMargin{56. }
-\CSLRightInline{\textbf{SKOS Simple Knowledge Organization System Primer} \url{http://www.w3.org/TR/skos-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
-\CSLLeftMargin{57. }
-\CSLRightInline{\textbf{How matchable are four thousand ontologies on the semantic web}
-\CSLBlock{Wei Hu, Jianfeng Chen, Hang Zhang, Yuzhong Qu} \emph{The semantic web: Research and applications} (2011)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20} · ISBN: 978-3-642-21033-4}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
-\CSLLeftMargin{58. }
-\CSLRightInline{\textbf{API-centric Linked Data integration: The Open PHACTS Discovery Platform case study}
-\CSLBlock{Paul Groth, Antonis Loizou, Alasdair JG Gray, Carole Goble, Lee Harland, Steve Pettifer} \emph{Journal of Web Semantics} (2014-12)
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
-\CSLLeftMargin{59. }
-\CSLRightInline{\textbf{A systematic analysis of term reuse and term overlap across biomedical ontologies}
-\CSLBlock{Maulik R Kamdar, Tania Tudorache, Mark A Musen} \emph{Semantic Web} (2017-08-07)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/28819351}{28819351} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5555235}{PMC5555235}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
-\CSLLeftMargin{60. }
-\CSLRightInline{\textbf{Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot}
-\CSLBlock{Martin Klein, Herbert Van de Sompel, Robert Sanderson, Harihar Shankar, Lyudmila Balakireva, Ke Zhou, Richard Tobin} \emph{PLoS ONE} (2014-12-26)
-\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/25541969}{25541969} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4277367}{PMC4277367}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
-\CSLLeftMargin{61. }
-\CSLRightInline{\textbf{A more decentralized vision for linked data}
-\CSLBlock{Axel Polleres, Maulik Rajendra Kamdar, Javier David Fernández, Tania Tudorache, Mark Alan Musen} \emph{Satya Widya} (2020-01-31)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/sw-190380}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
-\CSLLeftMargin{62. }
-\CSLRightInline{\textbf{The landscape of ontology reuse approaches}
-\CSLBlock{Valentina Anita Carriero, Marilena Daquino, Aldo Gangemi, Andrea Giovanni Nuzzolese, Silvio Peroni, Valentina Presutti, Francesca Tomasi} \emph{Applications and practices in ontology design, extraction, and reasoning} (2020-11-12)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/ssw200033} · ISBN: 9781643681429}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
-\CSLLeftMargin{63. }
-\CSLRightInline{\textbf{RDF Schema 1.1} \url{http://www.w3.org/TR/rdf-schema/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
-\CSLLeftMargin{64. }
-\CSLRightInline{\textbf{OWL 2 Web Ontology Language Document Overview (Second Edition)} \url{http://www.w3.org/TR/owl2-overview/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
-\CSLLeftMargin{65. }
-\CSLRightInline{\textbf{Mapping between the OBO and OWL ontology languages}
-\CSLBlock{Syed Tirmizi, Stuart Aitken, Dilvan A Moreira, Chris Mungall, Juan Sequeda, Nigam H Shah, Daniel P Miranker} \emph{Journal of Biomedical Semantics} (2011)
-\CSLBlock{DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/21388572}{21388572} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3105495}{PMC3105495}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
-\CSLLeftMargin{66. }
-\CSLRightInline{\textbf{Linked Data - Design Issues} \url{https://www.w3.org/DesignIssues/LinkedData.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
-\CSLLeftMargin{67. }
-\CSLRightInline{\textbf{The Open Graph protocol} \url{https://ogp.me/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
-\CSLLeftMargin{68. }
-\CSLRightInline{\textbf{RDFa 1.1 Primer - Third Edition} \url{http://www.w3.org/TR/rdfa-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
-\CSLLeftMargin{69. }
-\CSLRightInline{\textbf{HTML Standard} \url{https://html.spec.whatwg.org/multipage/microdata.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
-\CSLLeftMargin{70. }
-\CSLRightInline{\textbf{JSON-LD 1.1} \url{https://www.w3.org/TR/json-ld/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
-\CSLLeftMargin{71. }
-\CSLRightInline{\textbf{Usage Statistics of JSON-LD for Websites, May 2022} \url{https://w3techs.com/technologies/details/da-jsonld}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
-\CSLLeftMargin{72. }
-\CSLRightInline{\textbf{Designing a Linked Data developer experience} (2018-12-28) \url{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
-\CSLLeftMargin{73. }
-\CSLRightInline{\textbf{Shapes Constraint Language (SHACL)} \url{https://www.w3.org/TR/shacl/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
-\CSLLeftMargin{74. }
-\CSLRightInline{\textbf{Shape Expressions (ShEx) 2.1 Primer} \url{http://shex.io/shex-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
-\CSLLeftMargin{75. }
-\CSLRightInline{\textbf{Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation}
-\CSLBlock{Katherine Thornton, Harold Solbrig, Gregory S Stupp, Jose Emilio Labra Gayo, Daniel Mietchen, Eric Prud, Andra Waagmeester} \emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings} (2019)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39} · ISBN: 978-3-030-21347-3}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
-\CSLLeftMargin{76. }
-\CSLRightInline{\textbf{Validating RDF Data}
-\CSLBlock{Jose Emilio Labra Gayo, Eric Prud'hommeaux, Iovka Boneva, Dimitris Kontokostas} \emph{Synthesis Lectures on the Semantic Web: Theory and Technology} (2017-09-28)
-\CSLBlock{DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
-\CSLLeftMargin{77. }
-\CSLRightInline{\textbf{Survey of tools for linked data consumption}
-\CSLBlock{Jakub Klímek, Petr Škoda, Martin Nečaský} \emph{Satya Widya} (2019-05-23)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/sw-180316}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
-\CSLLeftMargin{78. }
-\CSLRightInline{\textbf{RightField: embedding ontology annotation in spreadsheets}
-\CSLBlock{K Wolstencroft, S Owen, M Horridge, O Krebs, W Mueller, JL Snoep, F du Preez, C Goble} \emph{Bioinformatics} (2011-05-26)
-\CSLBlock{DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/21622664}{21622664}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
-\CSLLeftMargin{79. }
-\CSLRightInline{\textbf{REST and Linked Data}
-\CSLBlock{Kevin R Page, David C De Roure, Kirk Martinez} \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11} (2011) \url{http://eprints.soton.ac.uk/id/eprint/272098}
-\CSLBlock{DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
-\CSLLeftMargin{80. }
-\CSLRightInline{\textbf{On Schema.org and Why It Matters for the Web}
-\CSLBlock{Peter Mika} \emph{IEEE Internet Computing} (2015-07)
-\CSLBlock{DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
-\CSLLeftMargin{81. }
-\CSLRightInline{\textbf{A new look at the semantic web}
-\CSLBlock{Abraham Bernstein, James Hendler, Natalya Noy} \emph{Communications of the ACM} (2016-08-24)
-\CSLBlock{DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
-\CSLLeftMargin{82. }
-\CSLRightInline{\textbf{An Interoperability Framework and Distributed Platform for Fast Data Applications}
-\CSLBlock{José Carlos Martins Delgado} \emph{Data Science and Big Data Computing} (2016)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}}}
+X.1255 : Framework for discovery of identity management information.
+\emph{Available at} \href{https://www.itu.int/rec/T-REC-X.1255-201309-I}{\emph{https://www.itu.int/rec/T-REC-X.1255-201309-I}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hRzcHhPD}{}}%
-\CSLLeftMargin{83. }
-\CSLRightInline{\textbf{A Comparison Framework for Middleware Infrastructures.}
-\CSLBlock{Apostolos Zarras} \emph{The Journal of Object Technology} (2004)
-\CSLBlock{DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
-\CSLLeftMargin{84. }
-\CSLRightInline{\textbf{The FAIR data maturity model: An approach to harmonise FAIR assessments}
-\CSLBlock{Christophe Bahim, Carlos Casorrán-Amilburu, Makx Dekkers, Edit Herczog, Nicolas Loozen, Konstantinos Repanas, Keith Russell, Shelley Stall} \emph{Data Science Journal} (2020-10-27)
-\CSLBlock{DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
-\CSLLeftMargin{85. }
-\CSLRightInline{\textbf{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}
-\CSLBlock{William Stallings} \emph{Sams} (1990)
-\CSLBlock{ISBN: 9780672226977}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
-\CSLLeftMargin{86. }
-\CSLRightInline{\textbf{Architectural styles and the design of network-based software architectures}
-\CSLBlock{Roy Thomas Fielding} \emph{University of California, Irvine} (2000) \url{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
-\CSLLeftMargin{87. }
-\CSLRightInline{\textbf{Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award)}
-\CSLBlock{Roy T Fielding, Richard N Taylor, Justin R Erenkrantz, Michael M Gorlick, Jim Whitehead, Rohit Khare, Peyman Oreizy} \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017} (2017-09-04) \url{http://dl.acm.org/citation.cfm?doid=3106237.3121282}
-\CSLBlock{DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282} · ISBN: 9781450351058}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
-\CSLLeftMargin{88. }
-\CSLRightInline{\textbf{OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More} \url{https://spec.openapis.org/oas/v3.1.0.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
-\CSLLeftMargin{89. }
-\CSLRightInline{\textbf{Schema.org Actions - schema.org} \url{https://schema.org/docs/actions.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
-\CSLLeftMargin{90. }
-\CSLRightInline{\textbf{Web Services Description Language (WSDL) Version 2.0 Part 0: Primer} \url{http://www.w3.org/TR/wsdl20-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
-\CSLLeftMargin{91. }
-\CSLRightInline{\textbf{The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud}
-\CSLBlock{Katherine Wolstencroft, Robert Haines, Donal Fellows, Alan Williams, David Withers, Stuart Owen, Stian Soiland-Reyes, Ian Dunlop, Aleksandra Nenadic, Paul Fisher, \ldots{} Carole Goble} \emph{Nucleic Acids Research} (2013-05-02) \url{https://doi.org/ggbwf4}
-\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/23640334}{23640334} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3692062}{PMC3692062}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
-\CSLLeftMargin{92. }
-\CSLRightInline{\textbf{Perspectives on automated composition of workflows in the life sciences}
-\CSLBlock{Anna-Lena Lamprecht, Magnus Palmblad, Jon Ison, Veit Schwämmle, Mohammad Sadnan Al Manir, Ilkay Altintas, Christopher JO Baker, Ammar Ben Hadj Amor, Salvador Capella-Gutierrez, Paulos Charonyktakis, \ldots{} Katherine Wolstencroft} \emph{F1000Research} (2021-09-07) \url{https://doi.org/gqprp7}
-\CSLBlock{DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/34804501}{34804501} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8573700}{PMC8573700}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-\CSLLeftMargin{93. }
-\CSLRightInline{\textbf{The Modern Standards Paradigm - Five Key Principles}
-\CSLBlock{OpenStand} \url{https://open-stand.org/about-us/principles/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
-\CSLLeftMargin{94. }
-\CSLRightInline{\textbf{First International Conference on FAIR Digital Objects}
-\CSLBlock{Tina Loo (editor)} \emph{Pensoft Publishers} (1970) \url{https://doi.org/grcqxh}
-\CSLBlock{DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
-\CSLLeftMargin{95. }
-\CSLRightInline{\textbf{DOIP API for HTTP Clients --- Cordra documentation} \url{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-\CSLLeftMargin{96. }
-\CSLRightInline{\textbf{DOI Resolution Documentation} \url{https://www.doi.org/factsheets/DOIProxy.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-\CSLLeftMargin{97. }
-\CSLRightInline{\textbf{DOI Handbook - Resolution} \url{https://www.doi.org/doi_handbook/3_Resolution.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
-\CSLLeftMargin{98. }
-\CSLRightInline{\textbf{HTTP Semantics}
-\CSLBlock{R Fielding, M Nottingham, J Reschke (editors)} \emph{RFC Editor} (2022-06) \url{https://doi.org/gqprqb}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
-\CSLLeftMargin{99. }
-\CSLRightInline{\textbf{Linked data platform 1.0}
-\CSLBlock{Linked Data Platform Working Group} \emph{W3C} (2015-02-26) \url{https://www.w3.org/TR/2015/REC-ldp-20150226/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
-\CSLLeftMargin{100. }
-\CSLRightInline{\textbf{FAIR Digital Object Framework Documentation} \url{https://fairdigitalobjectframework.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
-\CSLLeftMargin{101. }
-\CSLRightInline{\textbf{Handle System Overview}
-\CSLBlock{S Sun, L Lannom, B Boesch} \emph{RFC Editor} (2003-11) \url{https://doi.org/ggn83z}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
-\CSLLeftMargin{102. }
-\CSLRightInline{\textbf{Web API design best practices - Azure Architecture Center}
-\CSLBlock{martinekuan} \url{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
-\CSLLeftMargin{103. }
-\CSLRightInline{\textbf{The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration}
-\CSLBlock{Barry Smith, Michael Ashburner, Cornelius Rosse, Jonathan Bard, William Bug, Werner Ceusters, Louis J Goldberg, Karen Eilbeck, Amelia Ireland, \ldots{} Suzanna Lewis} \emph{Nature Biotechnology} (2007-11) \url{https://doi.org/bqng99}
-\CSLBlock{DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/17989687}{17989687} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2814061}{PMC2814061}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
-\CSLLeftMargin{104. }
-\CSLRightInline{\textbf{Data Information View} \url{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
-\CSLLeftMargin{105. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 3} \url{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-\CSLLeftMargin{106. }
-\CSLRightInline{\textbf{ORE User Guide - Primer} \url{http://www.openarchives.org/ore/1.0/primer}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-\CSLLeftMargin{107. }
-\CSLRightInline{\textbf{PROV-O: The PROV Ontology} \url{https://www.w3.org/TR/2013/REC-prov-o-20130430/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
-\CSLLeftMargin{108. }
-\CSLRightInline{\textbf{Semantic Web Services}
-\CSLBlock{Dieter Fensel, Federico Michele Facca, Elena Simperl, Ioan Toma} \emph{Springer Berlin Heidelberg} (2011) \url{https://doi.org/bv7nnc}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-\CSLLeftMargin{109. }
-\CSLRightInline{\textbf{5-star Open Data} \url{http://5stardata.info/en/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
-\CSLLeftMargin{110. }
-\CSLRightInline{\url{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
-\CSLLeftMargin{111. }
-\CSLRightInline{\textbf{Global Internet Phenomena Report 2022}
-\CSLBlock{Sandvine} \url{https://www.sandvine.com/global-internet-phenomena-report-2022}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
-\CSLLeftMargin{112. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol Version 2 (HTTP/2)}
-\CSLBlock{M Belshe, R Peon} \emph{RFC Editor} (2015-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
-\CSLLeftMargin{113. }
-\CSLRightInline{\url{https://blog.cloudflare.com/http-3-vs-http-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
-\CSLLeftMargin{114. }
-\CSLRightInline{\textbf{draft-ietf-quic-http-34} \url{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
-\CSLLeftMargin{115. }
-\CSLRightInline{\textbf{QUIC: A UDP-Based Multiplexed and Secure Transport}
-\CSLBlock{J Iyengar, M Thomson (editors)} \emph{RFC Editor} (2021-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
-\CSLLeftMargin{116. }
-\CSLRightInline{\textbf{URI Template}
-\CSLBlock{J Gregorio, R Fielding, M Hadley, M Nottingham, D Orchard} \emph{RFC Editor} (2012-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
-\CSLLeftMargin{117. }
-\CSLRightInline{\textbf{Content negotiation - HTTP \textbar{} MDN} \url{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
-\CSLLeftMargin{118. }
-\CSLRightInline{\textbf{Hypertext Style: Cool URIs don't change.} \url{https://www.w3.org/Provider/Style/URI}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
-\CSLLeftMargin{119. }
-\CSLRightInline{\textbf{HTTP Framework for Time-Based Access to Resource States -\/- Memento}
-\CSLBlock{H Van de Sompel, M Nelson, R Sanderson} \emph{RFC Editor} (2013-12)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
-\CSLLeftMargin{120. }
-\CSLRightInline{\textbf{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}
-\CSLBlock{G Clemm, J Amsden, T Ellison, C Kaler, J Whitehead} \emph{RFC Editor} (2002-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
-\CSLLeftMargin{121. }
-\CSLRightInline{\textbf{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}
-\CSLBlock{L Dusseault (editor)} \emph{RFC Editor} (2007-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
-\CSLLeftMargin{122. }
-\CSLRightInline{\textbf{Upgrading to TLS Within HTTP/1.1}
-\CSLBlock{R Khare, S Lawrence} \emph{RFC Editor} (2000-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
-\CSLLeftMargin{123. }
-\CSLRightInline{\textbf{Hypertext Style: Cool URIs don't change.} \url{https://www.w3.org/Provider/Style/URI.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
-\CSLLeftMargin{124. }
-\CSLRightInline{\textbf{draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes} \url{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
-\CSLLeftMargin{125. }
-\CSLRightInline{\textbf{The 'profile' Link Relation Type}
-\CSLBlock{E Wilde} \emph{RFC Editor} (2013-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
-\CSLLeftMargin{126. }
-\CSLRightInline{\textbf{W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures} \url{http://www.w3.org/TR/xmlschema11-1/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
-\CSLLeftMargin{127. }
-\CSLRightInline{\textbf{JSON-LD 1.1} \url{http://www.w3.org/TR/json-ld/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
-\CSLLeftMargin{128. }
-\CSLRightInline{\textbf{WebSockets Standard} \url{https://websockets.spec.whatwg.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
-\CSLLeftMargin{129. }
-\CSLRightInline{\textbf{Linked Data Notifications} \url{https://www.w3.org/TR/ldn/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
-\CSLLeftMargin{130. }
-\CSLRightInline{\textbf{The Atom Publishing Protocol}
-\CSLBlock{J Gregorio, B de hOra (editors)} \emph{RFC Editor} (2007-10)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
-\CSLLeftMargin{131. }
-\CSLRightInline{\textbf{SWORD 3.0 Specification} \url{https://swordapp.github.io/swordv3/swordv3.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
-\CSLLeftMargin{132. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}
-\CSLBlock{R Fielding, J Reschke (editors)} \emph{RFC Editor} (2014-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-\CSLLeftMargin{133. }
-\CSLRightInline{\textbf{Hydra W3C Community Group} \url{https://www.hydra-cg.com/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
-\CSLLeftMargin{134. }
-\CSLRightInline{\textbf{draft-kelly-json-hal-08} \url{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
-\CSLLeftMargin{135. }
-\CSLRightInline{\textbf{Web Linking}
-\CSLBlock{M Nottingham} \emph{RFC Editor} (2017-10)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
-\CSLLeftMargin{136. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}
-\CSLBlock{R Fielding, J Reschke (editors)} \emph{RFC Editor} (2014-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
-\CSLLeftMargin{137. }
-\CSLRightInline{\textbf{HTTP Live Streaming}
-\CSLBlock{W May} \emph{RFC Editor} (2017-08)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
-\CSLLeftMargin{138. }
-\CSLRightInline{\textbf{ISO/IEC 23009-1:2019}
-\CSLBlock{14:00-17:00} \emph{ISO} \url{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
-\CSLLeftMargin{139. }
-\CSLRightInline{\textbf{FAIR data maturity model: Specification and guidelines}
-\CSLBlock{Research Data Alliance FAIR Data Maturity Model Working Group} \emph{Research Data Alliance} (2020) \url{https://zenodo.org/record/3909563/\#.\%7BYGRNnq8za70\%7D}
-\CSLBlock{DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
-\CSLLeftMargin{140. }
-\CSLRightInline{\textbf{NCBO BioPortal} \url{https://bioportal.bioontology.org/ontologies}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
-\CSLLeftMargin{141. }
-\CSLRightInline{\textbf{The Globus Striped GridFTP Framework and Server}
-\CSLBlock{W Allcock, J Bresnahan, R Kettimuthu, M Link} \emph{ACM/IEEE SC 2005 Conference (SC'05)} \url{https://doi.org/cgmc2b}
-\CSLBlock{DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-\CSLLeftMargin{142. }
-\CSLRightInline{\textbf{The ARK Identifier Scheme} \url{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
-\CSLLeftMargin{143. }
-\CSLRightInline{\textbf{Handle System Protocol (ver 2.1) Specification}
-\CSLBlock{S Sun, S Reilly, L Lannom, J Petrone} \emph{RFC Editor} (2003-11)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-\CSLLeftMargin{144. }
-\CSLRightInline{\textbf{Handle.Net Registry} \url{https://www.handle.net/download_hnr.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-\CSLLeftMargin{145. }
-\CSLRightInline{\textbf{Licenses \& Standards \textbar{} Open Source Initiative} \url{https://opensource.org/licenses}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-\CSLLeftMargin{146. }
-\CSLRightInline{\textbf{ISO 16684-1:2019}
-\CSLBlock{14:00-17:00} \emph{ISO} \url{https://www.iso.org/standard/75163.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
-\CSLLeftMargin{147. }
-\CSLRightInline{\textbf{Data - W3C} \url{https://www.w3.org/standards/semanticweb/data}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-\CSLLeftMargin{148. }
-\CSLRightInline{\textbf{Linking Across Provenance Bundles} \url{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-\CSLLeftMargin{149. }
-\CSLRightInline{\textbf{Introducing 'Role'}
-\CSLBlock{Unknown} \url{http://blog.schema.org/2014/06/introducing-role.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-\CSLLeftMargin{150. }
-\CSLRightInline{\textbf{PROV-O: The PROV Ontology} \url{https://www.w3.org/TR/prov-o/\#specializationOf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
-\CSLLeftMargin{151. }
-\CSLRightInline{\textbf{ORE Specification - Abstract Data Model} \url{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
-\CSLLeftMargin{152. }
-\CSLRightInline{\textbf{Schema.org - Schema.org} \url{https://schema.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
-\CSLLeftMargin{153. }
-\CSLRightInline{\textbf{DCMI Metadata Terms} \url{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
-\CSLLeftMargin{154. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 2} \url{https://www.w3.org/TR/vocab-dcat-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
-\CSLLeftMargin{155. }
-\CSLRightInline{\textbf{FOAF Vocabulary Specification} \url{http://xmlns.com/foaf/spec/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
-\CSLLeftMargin{156. }
-\CSLRightInline{\textbf{PAV ontology: provenance, authoring and versioning}
-\CSLBlock{Paolo Ciccarese, Stian Soiland-Reyes, Khalid Belhajjame, Alasdair JG Gray, Carole Goble, Tim Clark} \emph{Journal of Biomedical Semantics} (2013) \url{https://doi.org/gftcpx}
-\CSLBlock{DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/24267948}{24267948} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4177195}{PMC4177195}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
-\CSLLeftMargin{157. }
-\CSLRightInline{\textbf{DCMI Metadata Terms} (2020-01-20) \url{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
-\CSLLeftMargin{158. }
-\CSLRightInline{\textbf{The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
-Digital Object Approach to Speeding up Digital Mobilisation of Natural History
-Collections}
-\CSLBlock{Alex Hardisty, Paul Brack, Carole Goble, Laurence Livermore, Ben Scott, Quentin Groom, Stuart Owen, Stian Soiland-Reyes} \emph{Data Intelligence} (2022) \url{https://doi.org/grk33x}
-\CSLBlock{DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
-\CSLLeftMargin{159. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 2} \url{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
-\CSLLeftMargin{160. }
-\CSLRightInline{\textbf{Bioschemas - Bioschemas} \url{http://bioschemas.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
-\CSLLeftMargin{161. }
-\CSLRightInline{\textbf{Darwin Core: An Evolving Community-Developed Biodiversity Data Standard}
-\CSLBlock{John Wieczorek, David Bloom, Robert Guralnick, Stan Blum, Markus Döring, Renato Giovanni, Tim Robertson, David Vieglais} \emph{PLoS ONE} (2012-01-06) \url{https://doi.org/fzrpwq}
-\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22238640}{22238640} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3253084}{PMC3253084}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
-\CSLLeftMargin{162. }
-\CSLRightInline{\textbf{Semantic Web - XML2000 - slide "Architecture"} \url{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}}
+Zarras A. 2004. A Comparison Framework for Middleware Infrastructures. \emph{The Journal of Object Technology} 3:103. DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}.
 

### 2023-01-24 latex/again.tex
converted to peerj overleaf template

--- 2023-01-24 manuscript.tex
@@
 \hypertarget{sec:introduction}{%
-\section{Introduction}\label{sec:introduction}}
+\section*{Introduction}\label{sec:introduction}}
 
@@
 \hypertarget{sec:background}{%
-\section{Background and related work}\label{sec:background}}
+\section*{Background and related work}\label{sec:background}}
 
@@
 \hypertarget{sec:fdo}{%
-\subsection{FAIR Digital Object}\label{sec:fdo}}
+\subsection*{FAIR Digital Object}\label{sec:fdo}}
 
 \hypertarget{sec:next-step-fdo}{%
-\subsubsection{Next steps for FDO}\label{sec:next-step-fdo}}
+\subsubsection*{Next steps for FDO}\label{sec:next-step-fdo}}
 
@@
 \hypertarget{fdo-approaches}{%
-\subsubsection{FDO approaches}\label{fdo-approaches}}
+\subsubsection*{FDO approaches}\label{fdo-approaches}}
 
@@
 \hypertarget{sec:ld}{%
-\subsection{From the Semantic Web to Linked Data}\label{sec:ld}}
+\subsection*{From the Semantic Web to Linked Data}\label{sec:ld}}
 
@@
 \hypertarget{sec:semweb}{%
-\subsubsection{A brief history of the Semantic Web}\label{sec:semweb}}
+\subsubsection*{A brief history of the Semantic Web}\label{sec:semweb}}
 
@@
 \hypertarget{sec:ld-web}{%
-\subsubsection{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
+\subsubsection*{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
 
@@
 \hypertarget{sec:comparing}{%
-\section{Comparing FDO and existing approaches}\label{sec:comparing}}
+\section*{Comparing FDO and existing approaches}\label{sec:comparing}}
 
@@
 \hypertarget{sec:interoperability-compare}{%
-\subsection{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
+\subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
 
@@
 \hypertarget{mapping-of-metamodel-concepts}{%
-\subsubsection{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
+\subsubsection*{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
 
@@
 \hypertarget{sec:doip-fdo-compare}{%
-\subsection{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
+\subsection*{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
 
@@
   G1 and G2 call for stability and trustworthiness.
-While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
+WhileC the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
 \item
@@
 \item
-  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
+  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-uCse.
 Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}})
@@
 \hypertarget{sec:middleware}{%
-\subsection{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
+\subsection*{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
 
@@
 \hypertarget{sec:fair-compare}{%
-\subsection{Assessing FDO against FAIR}\label{sec:fair-compare}}
+\subsection*{Assessing FDO against FAIR}\label{sec:fair-compare}}
 
@@
 \hypertarget{eosc-interoperability-framework}{%
-\section{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
+\section*{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
 
@@
 \hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
+\section*{Discussion}\label{sec:discussion}}
 
@@
 \hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
+\subsection*{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
 
@@
 \hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+\section*{Conclusion}\label{conclusion}}
 
@@
 \hypertarget{references}{%
-\section{References}\label{references}}
+\section*{References}\label{references}}
 

### 2023-01-24 latex/bibliography.tex
converted to peerj overleaf template

--- 2023-01-24 manuscript.tex
@@
-This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
-was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
-on January 24, 2023.
-
-\hypertarget{authors}{%
-\section*{Authors}\label{authors}}
-\addcontentsline{toc}{section}{Authors}
-
-\begin{itemize}
-\item
-  \textbf{Stian Soiland-Reyes}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/stain}{stain}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/soilandreyes}{soilandreyes}
-  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Carole Goble}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/carolegoble}{carolegoble}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
-  Department of Computer Science, The University of Manchester, UK
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Paul Groth}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/pgroth}{pgroth}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/pgroth}{pgroth}
-  Informatics Institute, Faculty of Science, University of Amsterdam, NL
-\end{itemize}
-
-\hypertarget{abstract}{%
-\section*{Abstract}\label{abstract}}
-\addcontentsline{toc}{section}{Abstract}
-
-\textbf{TODO}: Rewrite from notes below to actual abstract.
-
-FAIR Digital Object is an emerging concept from EOSC.
-This is important.
-Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
-Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
-
-What is the message of this paper?
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These are the overlaps
-\item
-  This is what FDO is requiring but not commonly deployed
-\item
-  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
-\item
-  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
-\item
-  Lessons for Semantic Web community. What is missing in FDO and in SW.
-\item
-  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
-\item
-  Go back to the old SW layer cake. Explains why we picked these frameworks.
-\item
-  Lessons for SW: Parts of the stack that is less. FDO contributions.
-\end{enumerate}
-
-Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.
-
-Have all the ingredients, but not cooking properly.
-
-Systematic through frameworks.
-
-Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?
-
-\hypertarget{sec:introduction}{%
-\section{Introduction}\label{sec:introduction}}
-
-The FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}) encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures (\protect\hyperlink{ref-u8Cy0psL}{Mons et al., 2017}).
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
-
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
-The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
-
-In other discourse, Linked Data (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\protect\hyperlink{ref-EJDjHrUv}{Bonino Da Silva Santos et al., 2016}; \protect\hyperlink{ref-GJHYN6xW}{Hasnain \& Rebholz-Schuhmann, 2018}).
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).
-
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\protect\hyperlink{ref-uxWjmlio}{Stefi, 2015}; \protect\hyperlink{ref-v0WwFHgT}{Stefi \& Hess, 2015}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves (\protect\hyperlink{ref-5V3GqSzR}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021b}).
-
-These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.
-
-Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.
-
-\hypertarget{sec:background}{%
-\section{Background and related work}\label{sec:background}}
-
-In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.
-
-\hypertarget{sec:fdo}{%
-\subsection{FAIR Digital Object}\label{sec:fdo}}
-
-\hypertarget{sec:next-step-fdo}{%
-\subsubsection{Next steps for FDO}\label{sec:next-step-fdo}}
-
-The FAIR Digital Object Forum (\protect\hyperlink{ref-DMEBhHID}{{`FAIR Digital Objects Forum \textbar{}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
-As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
-
-\textbf{FAIR Digital Object Overview and Specifications} (\protect\hyperlink{ref-11NpETLq1}{Anders et al., 2022a}) is a comprehensive overview of FAIR Digital Object specifications listed below.
-It serves as a primer that introduces FDO concepts and the remaining documents.
-
-The \textbf{FDO Forum Document Standards} (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
-In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
-
-The \textbf{FDO Requirement Specifications} (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) is an update of (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) as the foundational definition of FDO.
-This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
-The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
-
-The \textbf{Machine actionability} (\protect\hyperlink{ref-iJeys0T5}{Weiland et al., 2022a}) sets out to define what is meant by \emph{machine actionability} for FDOs.
-\emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
-The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.
-
-\textbf{Configuration Types} (\protect\hyperlink{ref-SPd0uEqO}{Lannom et al., 2022a}) classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022})
-
-\textbf{PID Profiles \& Attributes} (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}) specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
-
-\textbf{Kernel Attributes \& Metadata} (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
-It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
-
-\textbf{Granularity, Versioning, Mutability} (\protect\hyperlink{ref-16W9AycSW}{Working Group, 2022b}) considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
-The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
-
-\textbf{DOIP Endorsement Request} (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) is an endorsement of the DOIP v2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) specification as a potential FDO implementation, as it has been applied by several institutions (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}).
-The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
-
-\textbf{Upload of FDO} (\protect\hyperlink{ref-1H7cmmr69}{Blanchi et al., 2022a}) illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync (\protect\hyperlink{ref-eS5y9TRh}{{`ResourceSync Framework Specification - Table of Contents'}}) is suggested as one type of service to list FDOs.
-This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
-
-\textbf{Typing FAIR Digital Objects} (\protect\hyperlink{ref-ZFzPxCNB}{Lannom et al., 2022b}) defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
-This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
-Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work.
-How to define such FDO operations is not specified.
-
-\textbf{Implementation of Attributes, Types, Profiles and Registries} (\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
-This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} (\protect\hyperlink{ref-1CqIZ47pu}{Schwardmann \& Kálmán, 2022}).
-
-It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-
-The concept of \textbf{FAIR Digital Objects} (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) has been introduced as way to expose research data as active objects that conform to the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-This builds on the \emph{Digital Object} (DO) concept (\protect\hyperlink{ref-11MnuwJ4l}{Kahn \& Wilensky, 2006}), first introduced in 1995 (\protect\hyperlink{ref-3Uqe3fuK}{Kahn \& Wilensky, 1995}) as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework (\protect\hyperlink{ref-103Hw8H43}{{`X.1255 : Framework for discovery of identity management information'}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}).
-
-In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}.
-The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing (\protect\hyperlink{ref-tz0P3DTC}{Wittenburg et al., 2019}).
-
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}), in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}).
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
-
-\hypertarget{fdo-approaches}{%
-\subsubsection{FDO approaches}\label{fdo-approaches}}
-
-FDO is an evolving concept.
-A set of FDO Demonstrators (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) highlight how current adapters are approaching implementations of FDO from different angles:
-
-\begin{itemize}
-\tightlist
-\item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), see (\protect\hyperlink{ref-xuQKRRx5}{{`DOIP and Examples --- Cordra documentation'}}).} (usually TCP/IP over TLS).
-The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}}) and PIDs are assigned using the Handle system.
-Several type registries have been established.
-\item
-  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
-\item
-  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
-\end{itemize}
-
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
-
-\hypertarget{sec:ld}{%
-\subsection{From the Semantic Web to Linked Data}\label{sec:ld}}
-
-In order to describe \emph{Linked Data} as it is used today, we'll start with an (opinionated) description of the evolution of its foundation, the \emph{Semantic Web}.
-
-\hypertarget{sec:semweb}{%
-\subsubsection{A brief history of the Semantic Web}\label{sec:semweb}}
-
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee (\protect\hyperlink{ref-Guy7cjVY}{Berners-Lee \& Fischetti, 1999}), at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
-The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
-The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.
-
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\protect\hyperlink{ref-16WGmCih8}{{`RDF 1.1 Primer'}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\protect\hyperlink{ref-16wvFH4nh}{{`Resource Description Framework (RDF) Model and Syntax Specification'}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
-
-While triple-based knowledge representations were not new (\protect\hyperlink{ref-NxvhwSmG}{Stanczyk, 1987}), the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs (\protect\hyperlink{ref-RRtwPF7Y}{Berners-Lee, Fielding \& Masinter, 2005}) are generalised forms of URLs that include locator-less identifiers
-  such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years (\protect\hyperlink{ref-nW4HY8Nq}{{`"info" URI Registry (Frozen)'}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) (\protect\hyperlink{ref-HqZ3r6J3}{Juty, Le Novere \& Laibe, 2011}).
-RDF 1.1 formats use Unicode to support IRIs (\protect\hyperlink{ref-rqjuXkEf}{Duerst \& Suignard, 2005}), which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
-
-The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data).
-Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term.
-The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource.
-By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs.
-The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.
-
-The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} (\protect\hyperlink{ref-yd2WE9eq}{Horrocks \& Hendler, 2002}) gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
-
-Established Semantic Web services and standards include SPARQL (\protect\hyperlink{ref-xww8fb1B}{{`SPARQL 1.1 Overview'}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
-
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\protect\hyperlink{ref-bMSxa8XW}{Goble \& Stevens, 2008}; \protect\hyperlink{ref-Qlrlr2TZ}{Williams et al., 2012}).
-In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
-
-Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) (\protect\hyperlink{ref-FDDmgO6s}{Hu et al., 2011}) .
-
-The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS (\protect\hyperlink{ref-191ZkYZMt}{Groth et al., 2014})), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
-
-With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps (\protect\hyperlink{ref-9CdJnsoJ}{Kamdar, Tudorache \& Musen, 2017}).
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication (\protect\hyperlink{ref-AGHR9oyr}{Klein et al., 2014}), for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data (\protect\hyperlink{ref-5SuwLNdE}{Polleres et al., 2020, p.
-handle:20.500.11811/7183}).
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies (\protect\hyperlink{ref-TUmA4KQT}{Carriero et al., 2020}) , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), OBO (\protect\hyperlink{ref-1DalZuKct}{Tirmizi et al., 2011}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})), hash vs slash, HTTP status codes and PID redirection strategies (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).
-
-\hypertarget{sec:ld-web}{%
-\subsubsection{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
-
-The \textbf{Linked Data} concept (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) was kickstarted as a set of best practices (\protect\hyperlink{ref-OGXYtGZ8}{{`Linked Data - Design Issues'}}) to bring the Web aspect back into focus.
-Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
-This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
-
-Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph (\protect\hyperlink{ref-oN9D5LkD}{{`The Open Graph protocol'}}), RDFa (\protect\hyperlink{ref-FKIF6ApB}{{`RDFa 1.1 Primer - Third Edition'}}), Microdata (\protect\hyperlink{ref-14Utscfe4}{{`HTML Standard'}})) we find JSON-LD (\protect\hyperlink{ref-X5SWPqSU}{{`JSON-LD 1.1'}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}).
-
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\protect\hyperlink{ref-BFMoe8FC}{{`Designing a Linked Data developer experience'}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\protect\hyperlink{ref-GDrm9dgX}{{`Shapes Constraint Language (SHACL)'}}) or ShEx (\protect\hyperlink{ref-YUzogaV}{{`Shape Expressions (ShEx) 2.1 Primer'}})) can be used to validate RDF Data (\protect\hyperlink{ref-ZeOAPNHa}{Gayo et al., 2017}; \protect\hyperlink{ref-S5BI6Mwx}{Thornton et al., 2019}) before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts (\protect\hyperlink{ref-o596f2JM}{Klímek, Škoda \& Nečaský, 2019}), which then tend to prefer non-semantic two-dimensional formats such as CSV files.
-
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}).
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies (\protect\hyperlink{ref-MhsCt4P3}{Wolstencroft et al., 2011}), the use of Linked Data as part of REST Web APIs (\protect\hyperlink{ref-eS6a6tlz}{Page, De Roure \& Martinez, 2011}) , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy (\protect\hyperlink{ref-117tX9MEG}{2016}), with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
-
-\hypertarget{sec:comparing}{%
-\section{Comparing FDO and existing approaches}\label{sec:comparing}}
-
-To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}), which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
-\item
-  The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), validated against its implementations for completeness.
-\item
-  A Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}), which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
-\item
-  Cross-checks against RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
-\item
-  EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
-\end{enumerate}
-
-The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).
-
-Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) against Linked Data in general.
-
-\hypertarget{sec:interoperability-compare}{%
-\subsection{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
-
-The Interoperability Framework for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
-
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model (\protect\hyperlink{ref-8ZRGFa6y}{Stallings, 1990}) better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
-In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}).
-\label{tbl:fdo-web-interoperability-framework}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
-FDO encourages research communities to develop common type registries to be shared across instances.
-In current DOIP practice, each service have their own types, attributes and operations.
-The wider symbiosis is consistent use of PIDs.
-& The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}) end up being mandated centrally by dominant (often commercial) providers (\protect\hyperlink{ref-11VxBOeZy}{Fielding et al., 2017}), which clients are required to use as-is with special code per service.
-Use of Linked Data enables common tooling and semantic mapping across differences.
-\\
-\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP).
-\texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).
-Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}).
-& ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), schema.org Actions (\protect\hyperlink{ref-dKAekUjL}{{`Schema.org Actions - schema.org'}}), WSDL/SOAP (\protect\hyperlink{ref-LmudecyN}{{`Web Services Description Language (WSDL) Version 2.0 Part 0: Primer'}}), but frequently just as human-readable developer documentation/examples.
-\\
-\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing.
-Every FDO has a type.
-Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}.
-No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) & Lightweight HTTP semantics for authenticity/navigation.
-Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
-Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit.
-\\
-\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation} & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema.
-Multiple byte stream attachments of any media type.
-& Textual HTTP headers (including any signposting), single byte stream of any media type, e.g.~Linked Data formats (JSON-LD, Turtle, RDF/XML) or embedded in document (HTML with RDFa, JSON-LD or Microdata).
-XML was previously the main syntax used by APIs, JSON is now dominant.
-\\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
-\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
-Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable).
-Unique DOIP protocol can be hard to add to existing Web application frameworks, although proxy services have been developed (e.g.~B2SHARE adapter).
-& HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to (for static content) multi-cloud Content-Delivery Networks (CDN).
-Current scalable cloud technologies for Web hosting may not support HTTP features previously seen as important for Semantic Web, e.g.~content negotiation and semantic HTTP status codes.
-\\
-\bottomrule
-\end{longtable}
-
-Based on the analysis shown in table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
-
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services (\protect\hyperlink{ref-39K3yHT3}{Wolstencroft et al., 2013}).
-However, this is mostly done via developer or human interaction (\protect\hyperlink{ref-iYh1wFBt}{Lamprecht et al., 2021}).
-Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.
-
-A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics.
-Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON.
-On the semantic front, FDO requires that every object have a well-defined type and structured form.
-This is clearly not the case on the Web.
-
-In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed.
-These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\protect\hyperlink{ref-hxtgCGjY}{{`The Modern Standards Paradigm - Five Key Principles'}}) to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference (\protect\hyperlink{ref-jrZe6Esu}{Loo, 1970}).
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\protect\hyperlink{ref-p7nwRj8E}{{`DOIP API for HTTP Clients --- Cordra documentation'}}).
-
-It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\protect\hyperlink{ref-68fzID2k}{{`DOI Handbook - Resolution'}}).
-
-\hypertarget{mapping-of-metamodel-concepts}{%
-\subsubsection{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
-
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2923}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2769}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
-\label{tbl:metamodel-concepts}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endhead
-Resource & FDO/DO & Resource \\
-Service & DOIP service & Server/endpoint \\
-Transaction & (not supported) & Conditional requests, \texttt{409\ Conflict} \\
-Process & Extended operations & (primarily stateless), \texttt{100\ Continue}, \texttt{202\ Accepted} \\
-Operation & DOIP Operation & Method, query parameters \\
-Request & DOIP Request & Request \\
-Response & DOIP Response & Response \\
-Message & Segment, \texttt{requestId} & Message, Representation \\
-Channel & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures. & TCP/IP, TLS, UDP \\
-Protocol & DOIP 2.0, ++ & HTTP/1.1, HTTP/2, HTTP/3 \\
-Link & PID/Handle & URL \\
-\bottomrule
-\end{longtable}
-
-From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
-Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \ref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
-
-\hypertarget{sec:doip-fdo-compare}{%
-\subsection{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
-
-The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and using Linked Data Platform (\protect\hyperlink{ref-17OHlMRQA}{Linked Data Platform Working Group, 2015}), as proposed by (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}; \protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) against its current implementations as DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}), with suggestions for required additions.
-\label{tbl:fdo-checks}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endhead
-G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
-& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
-\texttt{http://} URIs replaced by \texttt{https://}.
-& Keep flexibility of RDF serialisation formats which may change more frequently \\
-G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
-DOIP relatively unknown, in effect only one implementation.
-& Further promote DOIP and justify its benefits.
-Build tutorials and OSI open source implementations.
-Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}), however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
-Example-driven documentation.
-\\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
-\\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
-Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
-\\
-G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
-DOIP operations can use any transport protocol.
-& Document transport protocols as FDOs, recommend which transport to use.
-& URI as abstraction base.
-Does not specify PID requirements.
-& Give stronger deployment recommendations.
-\\
-G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
-Unclear when metadata field is a PID or plain text.
-& Make datatype of fields explicit to support navigation.
-& Machine-navigation through URIs via properties and types.
-Unclear when URI should be followed or is just identifier, but always distinct from plain text.
-& \\
-G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
-& Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
-& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
-Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \\
-G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
-No FDOF implementations.
-& Develop demonstrator of FDOF usage based on existing LDP server.
-\\
-G9: \emph{standard compliance} & Handle (\protect\hyperlink{ref-2gXY2Nwh}{Sun, Lannom \& Boesch, 2003}), DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) & HTTP, LDP.
-FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
-FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}).
-Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-& HTTP URLs as basis for identifiers, but they are frequently not persistent.
-& Add strong guidance for PID services like w3id and persistence policies.
-\\
-FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
-Datacite DOIs content-negotiate to give registered metadata.
-& Add FAIR Signposting at PID provider for minimal PID record \\
-FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
-No method/attribute defined for separate metadata, only directly in PID Record.
-Unclear meaning of multiple items and bytestream chunks.
-& Clarify expectations for multiple items.
-Recommend chunks to not be used.
-& URIs resolvable by default.
-Multiple ways to resolve metadata, unclear preference.
-& Add FAIR Signposting and preference order.
-\\
-FDOF4: \emph{Additional attributes} & Freetext attribute keys.
-Attributes should be defined for FDO type (?).
-& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
-Explicitly allow attributes not already defined in type.
-& All attributes individually identified.
-Any Linked Data attributes can be used by URI or with mapped prefix.
-& Clarify type expectations of required/recommended/optional attributes.
-\\
-FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
-& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-Clarify that operations can be mapped to protocol directly.
-& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
-Unclear how to provide PID for additional operations.
-& Specify how additional operations should be called over HTTP.
-\\
-FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan}) are used instead.
-& Document how operation resources can be discovered from an LPD container.
-Document search API.
-\\
-FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
-PR-TypingFDOs leaves this open.
-& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
-Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
-& Always make LDP types explicit in FDO profile.
-\\
-FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
-Separate Metadata FDO need custom property.
-& Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
-& Assertions are always with semantics, using RDF vocabularies.
-Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
-& Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
-FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
-Metadata would need to be bundled within custom FDO types/attributes.
-Unclear how levels are separated within single FDO representation (need FDOF8?).
-& Declare which metadata are expected within response attribute or within FDO object.
-Require PIDs for custom attributes.
-Define how alternate metadata levels can be represented separately.
-& Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
-& Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
-FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
-& Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} (\protect\hyperlink{ref-CNP39tah}{Smith et al., 2007}) & Rather document better how individual ad-hoc schemas can be started for prototypes.
-\\
-FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
-Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-(\protect\hyperlink{ref-6LvydTsC}{{`Data Information View'}})).
-& & LDP Containers required by specification, also user-created (eg.
-\texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 (\protect\hyperlink{ref-1EW4hLN8U}{{`Data Catalog Vocabulary (DCAT) - Version 3'}}), \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE (\protect\hyperlink{ref-gtGuSevK}{{`ORE User Guide - Primer'}}) \\
-FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
-\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
-No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
-\bottomrule
-\end{longtable}
-
-Note that the draft update to FDO specification (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
-
-A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines.
-Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
-
-From our evaluation, we can observe:
-
-\begin{itemize}
-\tightlist
-\item
-  G1 and G2 call for stability and trustworthiness.
-While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
-\item
-  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data.
-Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous.
-In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
-\item
-  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}.
-These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
-\item
-  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
-\item
-  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
-\item
-  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}})
-\item
-  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
-\item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
-\item
-  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation.
-In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
-\end{itemize}
-
-\hypertarget{sec:middleware}{%
-\subsection{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
-
-In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
-As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.
-
-We argue that the Web, with its now ubiquitous use of REST API (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services (\protect\hyperlink{ref-WzPECPDP}{Fensel et al., 2011}) attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\protect\hyperlink{ref-11dInoyVF}{{`5-star Open Data'}}).
-
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}) that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018})) and Web technologies (using Linked Data) as middleware infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004})
-\label{tbl:fdo-web-middleware}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
-Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
-PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}.
-& The Web is inherently open and made by cross-linked URLs.
-Participation requires DNS domain purchase (many free alternatives also exists).
-PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects.
-\\
-\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
-HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
-\\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
-DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
-Multiple connections possible.
-Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks (\protect\hyperlink{ref-F8VZ86hu}{Sandvine}).
-HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}) but still has TCP+TLS startup costs.
-For reduced latency , HTTP/3 (\protect\hyperlink{ref-Amit86Tp}{{`draft-ietf-quic-http-34'}}) use QUIC (\protect\hyperlink{ref-1GWvyEmUf}{Iyengar \& Thomson, 2021})) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
-\\
-\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
-FDOs should provide the mandatory kernel metadata fields.
-FDOs of the same declared type typically share additional attributes (although that schema may not be declared).
-DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions.
-& Each URL accessed separately.
-Common HTTP headers provide basic metadata, although it is often not reliable.
-A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), content-negotiation (\protect\hyperlink{ref-n2f3TfoI}{{`Content negotiation - HTTP \textbar{} MDN'}}) or FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}).
-\\
-\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
-Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
-& Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
-\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
-Multiple locations possible in Handle system, can expose geo-info.
-& PIDs and URL redirects.
-DNS aliases and IP routing can hide location.
-Geo-localised servers common for large cloud deployments.
-\\
-\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
-FDO kernel metadata can include checksum and date.
-& HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
-Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
-\\
-\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
-Custom codes uncommon.
-\\
-\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
-No redirection from DOIP service.
-& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
-\\
-\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
-There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged (\protect\hyperlink{ref-rbG9uRKw}{{`Hypertext Style: Cool URIs don't change.'}}).
-Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
-An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento (\protect\hyperlink{ref-DNjcMi4a}{Van de Sompel, Nelson \& Sanderson, 2013}) expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method (\protect\hyperlink{ref-hRUsaTiz}{Clemm et al., 2002}) (used by SVN).
-\\
-\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
-Internal synchronisation possible in backend for Extended operations.
-& Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} (\protect\hyperlink{ref-xvMAyRAc}{Dusseault, 2007}) with \texttt{LOCK} and \texttt{UNLOCK} methods.
-\\
-\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
-In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
-& The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space.
-In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend.
-\\
-\textbf{Encapsulation}: \emph{separate interface from implementation.
-Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
-FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
-Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade (\protect\hyperlink{ref-1B16WfR7q}{Khare \& Lawrence, 2000}) to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns (\protect\hyperlink{ref-iME5l1kk}{{`Hypertext Style: Cool URIs don't change.'}}) and PIDs via Persistent URLs (PURL).
-\\
-\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
-FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})) & Syntactically Media Type with multiple suffixes (\protect\hyperlink{ref-nlH2rm1s}{{`draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes'}}) (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) (\protect\hyperlink{ref-nCzRO0pK}{Wilde, 2013}).
-In metadata, semantic type systems (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL2 (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})).
-OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} (\protect\hyperlink{ref-11oVTMsuZ}{{`W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures'}}), JSON \texttt{\$schema} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})), JSON-LD \texttt{@context} (\protect\hyperlink{ref-by6CfOmv}{{`JSON-LD 1.1'}}).
-Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
-\\
-\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
-\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 (\protect\hyperlink{ref-tIkSXrb5}{Fielding \& Reschke, 2014b}) \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) (\protect\hyperlink{ref-PL512B0X}{May, 2017}), MPEG-DASH (ISO/IEC 23009-1:2019) (\protect\hyperlink{ref-ikO7e0mh}{14:00-17:00}) \\
-\bottomrule
-\end{longtable}
-
-Based on the analysis in Table \ref{tbl:fdo-web-middleware}, we make the following observations:
-
-\begin{itemize}
-\tightlist
-\item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}) supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
-It is unclear what is the purpose of splitting a binary into chunks on a channel which no longer can be multiplexed and the only property of a chunk is its size\footnote{Although it is possible with \texttt{0.DOIP/Op.Retrieve} to request only particular individual elements of an DO (e.g.~one file), unlike HTTP's \texttt{Range} request, it is not possible to select individual chunks of an element's bytestream.}.
-\item
-  HTTP has strong support for scalability and caching, but this mostly assumes read-operations from static resources.
-FDO has no view on immutability or validity of retrieved objects, but this should be taken into consideration to support large-scale usage.
-\item
-  HTTP optimisations for performance (e.g.~HTTP/2, multiplexing) is largely used for commercial media distribution (e.g.~Netflix), and not commonly used by providers of FAIR data
-\item
-  Cloud deployment of Web applications give many middleware benefits (Scalability, Distribution, Access transparancy, Location transparancy) -- it is unclear how DOIP as a custom protocol would perform in a cloud setting as most of this infrastructure assumes HTTP as the protocol.
-\item
-  Programmatically the Web is rather unstructured as middleware, as there are many implementation choices.
-Usually it is undeclared what to expect for a given URI/service, and programmers follow documented examples for a particular service rather than automated programmatic exploration across providers.
-This mean one can consider the Web as an ecosystem of smaller middlewares with commonalities.
-\item
-  Many providers of FAIR Linked Data also provide programmatic REST API endpoints, e.g.~\href{https://www.uniprot.org/help/programmatic_access}{UNIPROT}, \href{https://chembl.gitbook.io/chembl-interface-documentation/web-services}{ChEMBL}, but keeping the FAIR aspects such as retrieving metadata in such a scenario may require combining different services using multiple formats and identifier conventions.
-\end{itemize}
-
-\hypertarget{sec:fair-compare}{%
-\subsection{Assessing FDO against FAIR}\label{sec:fair-compare}}
-
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}).
-
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) comply with the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-Here we've used the RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}) as it has decomposed the FAIR principles to a structured list of FAIR indicators (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}), importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}; \protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) (first 2 columns) against the FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), FDO implemented with the protocol DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) and examples from Linked Data practices in general.
-(--- indicates \emph{Unspecified}, may be possible with additional conventions)
-\label{tbl:fair-data-maturity-model}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endhead
-RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
-RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
-Handle, DOI.
-& FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
-RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
-FAIR Signposting.
-\\
-RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
-RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
-Unclear how to link to additional metadata.
-& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
-Embedded in landing page (RDFa).
-\\
-RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
-& PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
-RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
-Handle registry only searchable by PID.
-& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-(\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}, pp.
-https://lod--cloud.net/})).
-Indexing by search engines if exposing HTML w/schema.org.
-\\
-RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
-\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
-RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
-RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
-RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
-RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
-RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
-Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}) & HTTP/HTTPS, FTP (now less common), GridFTP (\protect\hyperlink{ref-18LsXvKGJ}{Allcock et al.}) (for large data), ARK (\protect\hyperlink{ref-voDQ6diM}{{`The ARK Identifier Scheme'}}) \\
-RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol (\protect\hyperlink{ref-I9Kx0Hjy}{Sun et al., 2003}).
-One server implementation (\protect\hyperlink{ref-1CvYRVhy2}{{`Handle.Net Registry'}}), free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved (\protect\hyperlink{ref-xVJMmGVr}{{`Licenses \& Standards \textbar{} Open Source Initiative'}}) as an open source license -- it includes usage restrictions and requires Service Agreements.
-It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
-One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
-\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
-& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
-\\
-RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
-RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
-RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
-RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
-RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
-RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
-RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
-RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
-& FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) (\protect\hyperlink{ref-1BlxdxOwa}{14:00-17:00}) \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs (\protect\hyperlink{ref-o4h92LpS}{{`Data - W3C'}})), \texttt{rdfs:seeAlso}, FAIR signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) \texttt{describedby} \\
-RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
-RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
-RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles (\protect\hyperlink{ref-QpDmajc4}{{`Linking Across Provenance Bundles'}}), \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role (\protect\hyperlink{ref-10HA0mMBx}{Unknown}), \texttt{prov:specializationOf} (\protect\hyperlink{ref-ZUgwvAHE}{{`PROV-O: The PROV Ontology'}}), OAI-ORE Proxy (\protect\hyperlink{ref-ID1pTGry}{{`ORE Specification - Abstract Data Model'}})) \\
-RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) but not assigned PIDs yet.
-& Unspecified.
-Multiple metadata records can allow multiple semantic profiles.
-& Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org (\protect\hyperlink{ref-10pzKCEUE}{{`Schema.org - Schema.org'}}), Dublin Core Terms (\protect\hyperlink{ref-wAbZpOsr}{{`DCMI Metadata Terms'}}), DCAT (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), FOAF (\protect\hyperlink{ref-XmUY0cZB}{{`FOAF Vocabulary Specification'}}) \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}) \\
-RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
-RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
-RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
-PID Kernel attributes?
-Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}}), PAV (\protect\hyperlink{ref-1E4OfHyhI}{Ciccarese et al., 2013}), Dublin Core Terms (\protect\hyperlink{ref-hKblceo6}{{`DCMI Metadata Terms'}, 2020}) \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen (\protect\hyperlink{ref-zZTtJruI}{Hardisty et al., 2022})) & --- & Common, e.g.~DCAT 2 (\protect\hyperlink{ref-zt8AAjcZ}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), BioSchemas (\protect\hyperlink{ref-1F8S9FeHf}{{`Bioschemas - Bioschemas'}}) \\
-RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal (\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}}), Darwin Core (\protect\hyperlink{ref-9jaVzCbL}{Wieczorek et al., 2012}) \\
-RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
-\bottomrule
-\end{longtable}
-
-From this evaluation we observe:
-
-\begin{itemize}
-\tightlist
-\item
-  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
-\item
-  FDO/DOIP are stronger on identifier indicators
-\item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) has currently only a couple of implementations and is expressed informally.
-The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
-\item
-  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
-\item
-  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
-\item
-  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-(\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
-Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
-\item
-  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata.
-For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
-In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
-\item
-  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications.
-\texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known.
-g
-\end{itemize}
-
-\hypertarget{eosc-interoperability-framework}{%
-\section{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
-
-\textbf{TODO}: Introduce EOSC IF
-
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) in section 3.6 recommends:
-
-\begin{tablenos:no-prefix-table-caption}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Layer
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Recommendation
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-FDO
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
-Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
-Technical & Easy SLAs for service providers & Unspecified & None \\
-Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
-Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
-Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
-Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
-Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
-Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
-Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
-Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
-Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
-Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
-Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
-Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
-Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
-Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
-Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
-Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
-Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
-Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
-Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
-Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
-Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
-Legal & List recommended licenses & Undefined & Best practice recommendations \\
-Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
-Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
-Legal & GDPR compliance for personal data & Undefined & Undefined \\
-Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
-Legal & Harmonised terms-of-use & Undefined & Undefined \\
-Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
-\bottomrule
-\end{longtable}
-
-\end{tablenos:no-prefix-table-caption}
-
-Observations:
-
-\begin{itemize}
-\tightlist
-\item
-  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
-\item
-  Technical aspects highlighted by EOSC IF
-\item
-  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data.
-There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
-\item
-  FDO principles have strong recommendations for community development of organisational aspects.
-\item
-  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
-\end{itemize}
-
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
-
-\textbf{TODO}
-
-\begin{itemize}
-\tightlist
-\item
-  Ramnifications of ide
-\item
-  Finish tables
-\item
-  Bullet points per table
-\item
-  Send in Google Docs
-\end{itemize}
-
-\hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
-
-The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\protect\hyperlink{ref-kYjzRJ0H}{{`Semantic Web - XML2000 - slide "Architecture"'}}).
-However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
-
-While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
-
-We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-\ldots{}
-
 \hypertarget{references}{%

### 2023-01-24 latex/table1.tex
converted to peerj overleaf template

--- 2023-01-24 manuscript.tex
@@
-This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
-was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
-on January 24, 2023.
-
-\hypertarget{authors}{%
-\section*{Authors}\label{authors}}
-\addcontentsline{toc}{section}{Authors}
-
-\begin{itemize}
-\item
-  \textbf{Stian Soiland-Reyes}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/stain}{stain}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/soilandreyes}{soilandreyes}
-  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Carole Goble}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/carolegoble}{carolegoble}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
-  Department of Computer Science, The University of Manchester, UK
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Paul Groth}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/pgroth}{pgroth}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/pgroth}{pgroth}
-  Informatics Institute, Faculty of Science, University of Amsterdam, NL
-\end{itemize}
-
-\hypertarget{abstract}{%
-\section*{Abstract}\label{abstract}}
-\addcontentsline{toc}{section}{Abstract}
-
-\textbf{TODO}: Rewrite from notes below to actual abstract.
-
-FAIR Digital Object is an emerging concept from EOSC.
-This is important.
-Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
-Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
-
-What is the message of this paper?
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These are the overlaps
-\item
-  This is what FDO is requiring but not commonly deployed
-\item
-  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
-\item
-  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
-\item
-  Lessons for Semantic Web community. What is missing in FDO and in SW.
-\item
-  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
-\item
-  Go back to the old SW layer cake. Explains why we picked these frameworks.
-\item
-  Lessons for SW: Parts of the stack that is less. FDO contributions.
-\end{enumerate}
-
-Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.
-
-Have all the ingredients, but not cooking properly.
-
-Systematic through frameworks.
-
-Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?
-
-\hypertarget{sec:introduction}{%
-\section{Introduction}\label{sec:introduction}}
-
-The FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}) encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures (\protect\hyperlink{ref-u8Cy0psL}{Mons et al., 2017}).
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
-
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
-The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
-
-In other discourse, Linked Data (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\protect\hyperlink{ref-EJDjHrUv}{Bonino Da Silva Santos et al., 2016}; \protect\hyperlink{ref-GJHYN6xW}{Hasnain \& Rebholz-Schuhmann, 2018}).
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).
-
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\protect\hyperlink{ref-uxWjmlio}{Stefi, 2015}; \protect\hyperlink{ref-v0WwFHgT}{Stefi \& Hess, 2015}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves (\protect\hyperlink{ref-5V3GqSzR}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021b}).
-
-These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.
-
-Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.
-
-\hypertarget{sec:background}{%
-\section{Background and related work}\label{sec:background}}
-
-In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.
-
-\hypertarget{sec:fdo}{%
-\subsection{FAIR Digital Object}\label{sec:fdo}}
-
-\hypertarget{sec:next-step-fdo}{%
-\subsubsection{Next steps for FDO}\label{sec:next-step-fdo}}
-
-The FAIR Digital Object Forum (\protect\hyperlink{ref-DMEBhHID}{{`FAIR Digital Objects Forum \textbar{}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
-As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
-
-\textbf{FAIR Digital Object Overview and Specifications} (\protect\hyperlink{ref-11NpETLq1}{Anders et al., 2022a}) is a comprehensive overview of FAIR Digital Object specifications listed below.
-It serves as a primer that introduces FDO concepts and the remaining documents.
-
-The \textbf{FDO Forum Document Standards} (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
-In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
-
-The \textbf{FDO Requirement Specifications} (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) is an update of (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) as the foundational definition of FDO.
-This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
-The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
-
-The \textbf{Machine actionability} (\protect\hyperlink{ref-iJeys0T5}{Weiland et al., 2022a}) sets out to define what is meant by \emph{machine actionability} for FDOs.
-\emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
-The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.
-
-\textbf{Configuration Types} (\protect\hyperlink{ref-SPd0uEqO}{Lannom et al., 2022a}) classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022})
-
-\textbf{PID Profiles \& Attributes} (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}) specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
-
-\textbf{Kernel Attributes \& Metadata} (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
-It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
-
-\textbf{Granularity, Versioning, Mutability} (\protect\hyperlink{ref-16W9AycSW}{Working Group, 2022b}) considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
-The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
-
-\textbf{DOIP Endorsement Request} (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) is an endorsement of the DOIP v2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) specification as a potential FDO implementation, as it has been applied by several institutions (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}).
-The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
-
-\textbf{Upload of FDO} (\protect\hyperlink{ref-1H7cmmr69}{Blanchi et al., 2022a}) illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync (\protect\hyperlink{ref-eS5y9TRh}{{`ResourceSync Framework Specification - Table of Contents'}}) is suggested as one type of service to list FDOs.
-This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
-
-\textbf{Typing FAIR Digital Objects} (\protect\hyperlink{ref-ZFzPxCNB}{Lannom et al., 2022b}) defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
-This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
-Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work.
-How to define such FDO operations is not specified.
-
-\textbf{Implementation of Attributes, Types, Profiles and Registries} (\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
-This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} (\protect\hyperlink{ref-1CqIZ47pu}{Schwardmann \& Kálmán, 2022}).
-
-It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-
-The concept of \textbf{FAIR Digital Objects} (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) has been introduced as way to expose research data as active objects that conform to the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-This builds on the \emph{Digital Object} (DO) concept (\protect\hyperlink{ref-11MnuwJ4l}{Kahn \& Wilensky, 2006}), first introduced in 1995 (\protect\hyperlink{ref-3Uqe3fuK}{Kahn \& Wilensky, 1995}) as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework (\protect\hyperlink{ref-103Hw8H43}{{`X.1255 : Framework for discovery of identity management information'}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}).
-
-In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}.
-The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing (\protect\hyperlink{ref-tz0P3DTC}{Wittenburg et al., 2019}).
-
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}), in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}).
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
-
-\hypertarget{fdo-approaches}{%
-\subsubsection{FDO approaches}\label{fdo-approaches}}
-
-FDO is an evolving concept.
-A set of FDO Demonstrators (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) highlight how current adapters are approaching implementations of FDO from different angles:
-
-\begin{itemize}
-\tightlist
-\item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), see (\protect\hyperlink{ref-xuQKRRx5}{{`DOIP and Examples --- Cordra documentation'}}).} (usually TCP/IP over TLS).
-The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}}) and PIDs are assigned using the Handle system.
-Several type registries have been established.
-\item
-  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
-\item
-  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
-\end{itemize}
-
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
-
-\hypertarget{sec:ld}{%
-\subsection{From the Semantic Web to Linked Data}\label{sec:ld}}
-
-In order to describe \emph{Linked Data} as it is used today, we'll start with an (opinionated) description of the evolution of its foundation, the \emph{Semantic Web}.
-
-\hypertarget{sec:semweb}{%
-\subsubsection{A brief history of the Semantic Web}\label{sec:semweb}}
-
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee (\protect\hyperlink{ref-Guy7cjVY}{Berners-Lee \& Fischetti, 1999}), at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
-The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
-The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.
-
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\protect\hyperlink{ref-16WGmCih8}{{`RDF 1.1 Primer'}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\protect\hyperlink{ref-16wvFH4nh}{{`Resource Description Framework (RDF) Model and Syntax Specification'}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
-
-While triple-based knowledge representations were not new (\protect\hyperlink{ref-NxvhwSmG}{Stanczyk, 1987}), the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs (\protect\hyperlink{ref-RRtwPF7Y}{Berners-Lee, Fielding \& Masinter, 2005}) are generalised forms of URLs that include locator-less identifiers
-  such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years (\protect\hyperlink{ref-nW4HY8Nq}{{`"info" URI Registry (Frozen)'}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) (\protect\hyperlink{ref-HqZ3r6J3}{Juty, Le Novere \& Laibe, 2011}).
-RDF 1.1 formats use Unicode to support IRIs (\protect\hyperlink{ref-rqjuXkEf}{Duerst \& Suignard, 2005}), which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
-
-The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data).
-Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term.
-The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource.
-By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs.
-The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.
-
-The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} (\protect\hyperlink{ref-yd2WE9eq}{Horrocks \& Hendler, 2002}) gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
-
-Established Semantic Web services and standards include SPARQL (\protect\hyperlink{ref-xww8fb1B}{{`SPARQL 1.1 Overview'}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
-
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\protect\hyperlink{ref-bMSxa8XW}{Goble \& Stevens, 2008}; \protect\hyperlink{ref-Qlrlr2TZ}{Williams et al., 2012}).
-In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
-
-Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) (\protect\hyperlink{ref-FDDmgO6s}{Hu et al., 2011}) .
-
-The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS (\protect\hyperlink{ref-191ZkYZMt}{Groth et al., 2014})), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
-
-With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps (\protect\hyperlink{ref-9CdJnsoJ}{Kamdar, Tudorache \& Musen, 2017}).
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication (\protect\hyperlink{ref-AGHR9oyr}{Klein et al., 2014}), for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data (\protect\hyperlink{ref-5SuwLNdE}{Polleres et al., 2020, p.
-handle:20.500.11811/7183}).
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies (\protect\hyperlink{ref-TUmA4KQT}{Carriero et al., 2020}) , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), OBO (\protect\hyperlink{ref-1DalZuKct}{Tirmizi et al., 2011}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})), hash vs slash, HTTP status codes and PID redirection strategies (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).
-
-\hypertarget{sec:ld-web}{%
-\subsubsection{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
-
-The \textbf{Linked Data} concept (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) was kickstarted as a set of best practices (\protect\hyperlink{ref-OGXYtGZ8}{{`Linked Data - Design Issues'}}) to bring the Web aspect back into focus.
-Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
-This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
-
-Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph (\protect\hyperlink{ref-oN9D5LkD}{{`The Open Graph protocol'}}), RDFa (\protect\hyperlink{ref-FKIF6ApB}{{`RDFa 1.1 Primer - Third Edition'}}), Microdata (\protect\hyperlink{ref-14Utscfe4}{{`HTML Standard'}})) we find JSON-LD (\protect\hyperlink{ref-X5SWPqSU}{{`JSON-LD 1.1'}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}).
-
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\protect\hyperlink{ref-BFMoe8FC}{{`Designing a Linked Data developer experience'}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\protect\hyperlink{ref-GDrm9dgX}{{`Shapes Constraint Language (SHACL)'}}) or ShEx (\protect\hyperlink{ref-YUzogaV}{{`Shape Expressions (ShEx) 2.1 Primer'}})) can be used to validate RDF Data (\protect\hyperlink{ref-ZeOAPNHa}{Gayo et al., 2017}; \protect\hyperlink{ref-S5BI6Mwx}{Thornton et al., 2019}) before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts (\protect\hyperlink{ref-o596f2JM}{Klímek, Škoda \& Nečaský, 2019}), which then tend to prefer non-semantic two-dimensional formats such as CSV files.
-
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}).
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies (\protect\hyperlink{ref-MhsCt4P3}{Wolstencroft et al., 2011}), the use of Linked Data as part of REST Web APIs (\protect\hyperlink{ref-eS6a6tlz}{Page, De Roure \& Martinez, 2011}) , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy (\protect\hyperlink{ref-117tX9MEG}{2016}), with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
-
-\hypertarget{sec:comparing}{%
-\section{Comparing FDO and existing approaches}\label{sec:comparing}}
-
-To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}), which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
-\item
-  The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), validated against its implementations for completeness.
-\item
-  A Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}), which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
-\item
-  Cross-checks against RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
-\item
-  EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
-\end{enumerate}
-
-The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).
-
-Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) against Linked Data in general.
-
-\hypertarget{sec:interoperability-compare}{%
-\subsection{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
-
-The Interoperability Framework for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
-
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model (\protect\hyperlink{ref-8ZRGFa6y}{Stallings, 1990}) better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
-In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
-
+\renewcommand*{\arraystretch}{1.4}
 \begin{longtable}[]{@{}
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
+  >{\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
+  >{\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
 \caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}).
@@
 \end{longtable}
-
-Based on the analysis shown in table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
-
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services (\protect\hyperlink{ref-39K3yHT3}{Wolstencroft et al., 2013}).
-However, this is mostly done via developer or human interaction (\protect\hyperlink{ref-iYh1wFBt}{Lamprecht et al., 2021}).
-Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.
-
-A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics.
-Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON.
-On the semantic front, FDO requires that every object have a well-defined type and structured form.
-This is clearly not the case on the Web.
-
-In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed.
-These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\protect\hyperlink{ref-hxtgCGjY}{{`The Modern Standards Paradigm - Five Key Principles'}}) to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference (\protect\hyperlink{ref-jrZe6Esu}{Loo, 1970}).
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\protect\hyperlink{ref-p7nwRj8E}{{`DOIP API for HTTP Clients --- Cordra documentation'}}).
-
-It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\protect\hyperlink{ref-68fzID2k}{{`DOI Handbook - Resolution'}}).
-
-\hypertarget{mapping-of-metamodel-concepts}{%
-\subsubsection{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
-
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2923}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2769}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
-\label{tbl:metamodel-concepts}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endhead
-Resource & FDO/DO & Resource \\
-Service & DOIP service & Server/endpoint \\
-Transaction & (not supported) & Conditional requests, \texttt{409\ Conflict} \\
-Process & Extended operations & (primarily stateless), \texttt{100\ Continue}, \texttt{202\ Accepted} \\
-Operation & DOIP Operation & Method, query parameters \\
-Request & DOIP Request & Request \\
-Response & DOIP Response & Response \\
-Message & Segment, \texttt{requestId} & Message, Representation \\
-Channel & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures. & TCP/IP, TLS, UDP \\
-Protocol & DOIP 2.0, ++ & HTTP/1.1, HTTP/2, HTTP/3 \\
-Link & PID/Handle & URL \\
-\bottomrule
-\end{longtable}
-
-From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
-Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \ref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
-
-\hypertarget{sec:doip-fdo-compare}{%
-\subsection{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
-
-The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and using Linked Data Platform (\protect\hyperlink{ref-17OHlMRQA}{Linked Data Platform Working Group, 2015}), as proposed by (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}; \protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) against its current implementations as DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}), with suggestions for required additions.
-\label{tbl:fdo-checks}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endhead
-G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
-& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
-\texttt{http://} URIs replaced by \texttt{https://}.
-& Keep flexibility of RDF serialisation formats which may change more frequently \\
-G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
-DOIP relatively unknown, in effect only one implementation.
-& Further promote DOIP and justify its benefits.
-Build tutorials and OSI open source implementations.
-Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}), however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
-Example-driven documentation.
-\\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
-\\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
-Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
-\\
-G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
-DOIP operations can use any transport protocol.
-& Document transport protocols as FDOs, recommend which transport to use.
-& URI as abstraction base.
-Does not specify PID requirements.
-& Give stronger deployment recommendations.
-\\
-G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
-Unclear when metadata field is a PID or plain text.
-& Make datatype of fields explicit to support navigation.
-& Machine-navigation through URIs via properties and types.
-Unclear when URI should be followed or is just identifier, but always distinct from plain text.
-& \\
-G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
-& Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
-& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
-Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \\
-G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
-No FDOF implementations.
-& Develop demonstrator of FDOF usage based on existing LDP server.
-\\
-G9: \emph{standard compliance} & Handle (\protect\hyperlink{ref-2gXY2Nwh}{Sun, Lannom \& Boesch, 2003}), DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) & HTTP, LDP.
-FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
-FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}).
-Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-& HTTP URLs as basis for identifiers, but they are frequently not persistent.
-& Add strong guidance for PID services like w3id and persistence policies.
-\\
-FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
-Datacite DOIs content-negotiate to give registered metadata.
-& Add FAIR Signposting at PID provider for minimal PID record \\
-FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
-No method/attribute defined for separate metadata, only directly in PID Record.
-Unclear meaning of multiple items and bytestream chunks.
-& Clarify expectations for multiple items.
-Recommend chunks to not be used.
-& URIs resolvable by default.
-Multiple ways to resolve metadata, unclear preference.
-& Add FAIR Signposting and preference order.
-\\
-FDOF4: \emph{Additional attributes} & Freetext attribute keys.
-Attributes should be defined for FDO type (?).
-& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
-Explicitly allow attributes not already defined in type.
-& All attributes individually identified.
-Any Linked Data attributes can be used by URI or with mapped prefix.
-& Clarify type expectations of required/recommended/optional attributes.
-\\
-FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
-& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-Clarify that operations can be mapped to protocol directly.
-& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
-Unclear how to provide PID for additional operations.
-& Specify how additional operations should be called over HTTP.
-\\
-FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan}) are used instead.
-& Document how operation resources can be discovered from an LPD container.
-Document search API.
-\\
-FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
-PR-TypingFDOs leaves this open.
-& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
-Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
-& Always make LDP types explicit in FDO profile.
-\\
-FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
-Separate Metadata FDO need custom property.
-& Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
-& Assertions are always with semantics, using RDF vocabularies.
-Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
-& Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
-FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
-Metadata would need to be bundled within custom FDO types/attributes.
-Unclear how levels are separated within single FDO representation (need FDOF8?).
-& Declare which metadata are expected within response attribute or within FDO object.
-Require PIDs for custom attributes.
-Define how alternate metadata levels can be represented separately.
-& Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
-& Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
-FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
-& Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} (\protect\hyperlink{ref-CNP39tah}{Smith et al., 2007}) & Rather document better how individual ad-hoc schemas can be started for prototypes.
-\\
-FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
-Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-(\protect\hyperlink{ref-6LvydTsC}{{`Data Information View'}})).
-& & LDP Containers required by specification, also user-created (eg.
-\texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 (\protect\hyperlink{ref-1EW4hLN8U}{{`Data Catalog Vocabulary (DCAT) - Version 3'}}), \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE (\protect\hyperlink{ref-gtGuSevK}{{`ORE User Guide - Primer'}}) \\
-FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
-\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
-No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
-\bottomrule
-\end{longtable}
-
-Note that the draft update to FDO specification (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
-
-A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines.
-Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
-
-From our evaluation, we can observe:
-
-\begin{itemize}
-\tightlist
-\item
-  G1 and G2 call for stability and trustworthiness.
-While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
-\item
-  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data.
-Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous.
-In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
-\item
-  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}.
-These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
-\item
-  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
-\item
-  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
-\item
-  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}})
-\item
-  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
-\item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
-\item
-  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation.
-In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
-\end{itemize}
-
-\hypertarget{sec:middleware}{%
-\subsection{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
-
-In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
-As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.
-
-We argue that the Web, with its now ubiquitous use of REST API (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services (\protect\hyperlink{ref-WzPECPDP}{Fensel et al., 2011}) attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\protect\hyperlink{ref-11dInoyVF}{{`5-star Open Data'}}).
-
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}) that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018})) and Web technologies (using Linked Data) as middleware infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004})
-\label{tbl:fdo-web-middleware}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
-Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
-PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}.
-& The Web is inherently open and made by cross-linked URLs.
-Participation requires DNS domain purchase (many free alternatives also exists).
-PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects.
-\\
-\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
-HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
-\\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
-DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
-Multiple connections possible.
-Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks (\protect\hyperlink{ref-F8VZ86hu}{Sandvine}).
-HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}) but still has TCP+TLS startup costs.
-For reduced latency , HTTP/3 (\protect\hyperlink{ref-Amit86Tp}{{`draft-ietf-quic-http-34'}}) use QUIC (\protect\hyperlink{ref-1GWvyEmUf}{Iyengar \& Thomson, 2021})) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
-\\
-\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
-FDOs should provide the mandatory kernel metadata fields.
-FDOs of the same declared type typically share additional attributes (although that schema may not be declared).
-DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions.
-& Each URL accessed separately.
-Common HTTP headers provide basic metadata, although it is often not reliable.
-A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), content-negotiation (\protect\hyperlink{ref-n2f3TfoI}{{`Content negotiation - HTTP \textbar{} MDN'}}) or FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}).
-\\
-\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
-Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
-& Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
-\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
-Multiple locations possible in Handle system, can expose geo-info.
-& PIDs and URL redirects.
-DNS aliases and IP routing can hide location.
-Geo-localised servers common for large cloud deployments.
-\\
-\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
-FDO kernel metadata can include checksum and date.
-& HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
-Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
-\\
-\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
-Custom codes uncommon.
-\\
-\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
-No redirection from DOIP service.
-& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
-\\
-\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
-There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged (\protect\hyperlink{ref-rbG9uRKw}{{`Hypertext Style: Cool URIs don't change.'}}).
-Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
-An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento (\protect\hyperlink{ref-DNjcMi4a}{Van de Sompel, Nelson \& Sanderson, 2013}) expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method (\protect\hyperlink{ref-hRUsaTiz}{Clemm et al., 2002}) (used by SVN).
-\\
-\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
-Internal synchronisation possible in backend for Extended operations.
-& Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} (\protect\hyperlink{ref-xvMAyRAc}{Dusseault, 2007}) with \texttt{LOCK} and \texttt{UNLOCK} methods.
-\\
-\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
-In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
-& The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space.
-In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend.
-\\
-\textbf{Encapsulation}: \emph{separate interface from implementation.
-Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
-FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
-Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade (\protect\hyperlink{ref-1B16WfR7q}{Khare \& Lawrence, 2000}) to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns (\protect\hyperlink{ref-iME5l1kk}{{`Hypertext Style: Cool URIs don't change.'}}) and PIDs via Persistent URLs (PURL).
-\\
-\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
-FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})) & Syntactically Media Type with multiple suffixes (\protect\hyperlink{ref-nlH2rm1s}{{`draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes'}}) (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) (\protect\hyperlink{ref-nCzRO0pK}{Wilde, 2013}).
-In metadata, semantic type systems (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL2 (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})).
-OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} (\protect\hyperlink{ref-11oVTMsuZ}{{`W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures'}}), JSON \texttt{\$schema} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})), JSON-LD \texttt{@context} (\protect\hyperlink{ref-by6CfOmv}{{`JSON-LD 1.1'}}).
-Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
-\\
-\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
-\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 (\protect\hyperlink{ref-tIkSXrb5}{Fielding \& Reschke, 2014b}) \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) (\protect\hyperlink{ref-PL512B0X}{May, 2017}), MPEG-DASH (ISO/IEC 23009-1:2019) (\protect\hyperlink{ref-ikO7e0mh}{14:00-17:00}) \\
-\bottomrule
-\end{longtable}
-
-Based on the analysis in Table \ref{tbl:fdo-web-middleware}, we make the following observations:
-
-\begin{itemize}
-\tightlist
-\item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}) supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
-It is unclear what is the purpose of splitting a binary into chunks on a channel which no longer can be multiplexed and the only property of a chunk is its size\footnote{Although it is possible with \texttt{0.DOIP/Op.Retrieve} to request only particular individual elements of an DO (e.g.~one file), unlike HTTP's \texttt{Range} request, it is not possible to select individual chunks of an element's bytestream.}.
-\item
-  HTTP has strong support for scalability and caching, but this mostly assumes read-operations from static resources.
-FDO has no view on immutability or validity of retrieved objects, but this should be taken into consideration to support large-scale usage.
-\item
-  HTTP optimisations for performance (e.g.~HTTP/2, multiplexing) is largely used for commercial media distribution (e.g.~Netflix), and not commonly used by providers of FAIR data
-\item
-  Cloud deployment of Web applications give many middleware benefits (Scalability, Distribution, Access transparancy, Location transparancy) -- it is unclear how DOIP as a custom protocol would perform in a cloud setting as most of this infrastructure assumes HTTP as the protocol.
-\item
-  Programmatically the Web is rather unstructured as middleware, as there are many implementation choices.
-Usually it is undeclared what to expect for a given URI/service, and programmers follow documented examples for a particular service rather than automated programmatic exploration across providers.
-This mean one can consider the Web as an ecosystem of smaller middlewares with commonalities.
-\item
-  Many providers of FAIR Linked Data also provide programmatic REST API endpoints, e.g.~\href{https://www.uniprot.org/help/programmatic_access}{UNIPROT}, \href{https://chembl.gitbook.io/chembl-interface-documentation/web-services}{ChEMBL}, but keeping the FAIR aspects such as retrieving metadata in such a scenario may require combining different services using multiple formats and identifier conventions.
-\end{itemize}
-
-\hypertarget{sec:fair-compare}{%
-\subsection{Assessing FDO against FAIR}\label{sec:fair-compare}}
-
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}).
-
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) comply with the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-Here we've used the RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}) as it has decomposed the FAIR principles to a structured list of FAIR indicators (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}), importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}; \protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) (first 2 columns) against the FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), FDO implemented with the protocol DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) and examples from Linked Data practices in general.
-(--- indicates \emph{Unspecified}, may be possible with additional conventions)
-\label{tbl:fair-data-maturity-model}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endhead
-RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
-RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
-Handle, DOI.
-& FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
-RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
-FAIR Signposting.
-\\
-RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
-RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
-Unclear how to link to additional metadata.
-& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
-Embedded in landing page (RDFa).
-\\
-RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
-& PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
-RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
-Handle registry only searchable by PID.
-& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-(\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}, pp.
-https://lod--cloud.net/})).
-Indexing by search engines if exposing HTML w/schema.org.
-\\
-RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
-\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
-RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
-RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
-RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
-RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
-RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
-Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}) & HTTP/HTTPS, FTP (now less common), GridFTP (\protect\hyperlink{ref-18LsXvKGJ}{Allcock et al.}) (for large data), ARK (\protect\hyperlink{ref-voDQ6diM}{{`The ARK Identifier Scheme'}}) \\
-RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol (\protect\hyperlink{ref-I9Kx0Hjy}{Sun et al., 2003}).
-One server implementation (\protect\hyperlink{ref-1CvYRVhy2}{{`Handle.Net Registry'}}), free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved (\protect\hyperlink{ref-xVJMmGVr}{{`Licenses \& Standards \textbar{} Open Source Initiative'}}) as an open source license -- it includes usage restrictions and requires Service Agreements.
-It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
-One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
-\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
-& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
-\\
-RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
-RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
-RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
-RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
-RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
-RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
-RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
-RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
-& FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) (\protect\hyperlink{ref-1BlxdxOwa}{14:00-17:00}) \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs (\protect\hyperlink{ref-o4h92LpS}{{`Data - W3C'}})), \texttt{rdfs:seeAlso}, FAIR signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) \texttt{describedby} \\
-RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
-RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
-RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles (\protect\hyperlink{ref-QpDmajc4}{{`Linking Across Provenance Bundles'}}), \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role (\protect\hyperlink{ref-10HA0mMBx}{Unknown}), \texttt{prov:specializationOf} (\protect\hyperlink{ref-ZUgwvAHE}{{`PROV-O: The PROV Ontology'}}), OAI-ORE Proxy (\protect\hyperlink{ref-ID1pTGry}{{`ORE Specification - Abstract Data Model'}})) \\
-RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) but not assigned PIDs yet.
-& Unspecified.
-Multiple metadata records can allow multiple semantic profiles.
-& Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org (\protect\hyperlink{ref-10pzKCEUE}{{`Schema.org - Schema.org'}}), Dublin Core Terms (\protect\hyperlink{ref-wAbZpOsr}{{`DCMI Metadata Terms'}}), DCAT (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), FOAF (\protect\hyperlink{ref-XmUY0cZB}{{`FOAF Vocabulary Specification'}}) \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}) \\
-RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
-RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
-RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
-PID Kernel attributes?
-Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}}), PAV (\protect\hyperlink{ref-1E4OfHyhI}{Ciccarese et al., 2013}), Dublin Core Terms (\protect\hyperlink{ref-hKblceo6}{{`DCMI Metadata Terms'}, 2020}) \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen (\protect\hyperlink{ref-zZTtJruI}{Hardisty et al., 2022})) & --- & Common, e.g.~DCAT 2 (\protect\hyperlink{ref-zt8AAjcZ}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), BioSchemas (\protect\hyperlink{ref-1F8S9FeHf}{{`Bioschemas - Bioschemas'}}) \\
-RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal (\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}}), Darwin Core (\protect\hyperlink{ref-9jaVzCbL}{Wieczorek et al., 2012}) \\
-RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
-\bottomrule
-\end{longtable}
-
-From this evaluation we observe:
-
-\begin{itemize}
-\tightlist
-\item
-  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
-\item
-  FDO/DOIP are stronger on identifier indicators
-\item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) has currently only a couple of implementations and is expressed informally.
-The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
-\item
-  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
-\item
-  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
-\item
-  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-(\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
-Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
-\item
-  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata.
-For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
-In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
-\item
-  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications.
-\texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known.
-g
-\end{itemize}
-
-\hypertarget{eosc-interoperability-framework}{%
-\section{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
-
-\textbf{TODO}: Introduce EOSC IF
-
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) in section 3.6 recommends:
-
-\begin{tablenos:no-prefix-table-caption}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Layer
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Recommendation
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-FDO
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
-Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
-Technical & Easy SLAs for service providers & Unspecified & None \\
-Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
-Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
-Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
-Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
-Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
-Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
-Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
-Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
-Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
-Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
-Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
-Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
-Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
-Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
-Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
-Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
-Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
-Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
-Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
-Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
-Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
-Legal & List recommended licenses & Undefined & Best practice recommendations \\
-Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
-Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
-Legal & GDPR compliance for personal data & Undefined & Undefined \\
-Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
-Legal & Harmonised terms-of-use & Undefined & Undefined \\
-Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
-\bottomrule
-\end{longtable}
-
-\end{tablenos:no-prefix-table-caption}
-
-Observations:
-
-\begin{itemize}
-\tightlist
-\item
-  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
-\item
-  Technical aspects highlighted by EOSC IF
-\item
-  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data.
-There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
-\item
-  FDO principles have strong recommendations for community development of organisational aspects.
-\item
-  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
-\end{itemize}
-
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
-
-\textbf{TODO}
-
-\begin{itemize}
-\tightlist
-\item
-  Ramnifications of ide
-\item
-  Finish tables
-\item
-  Bullet points per table
-\item
-  Send in Google Docs
-\end{itemize}
-
-\hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
-
-The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\protect\hyperlink{ref-kYjzRJ0H}{{`Semantic Web - XML2000 - slide "Architecture"'}}).
-However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
-
-While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
-
-We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-\ldots{}
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
-. \emph{Available at} \href{https://blog.cloudflare.com/http-3-vs-http-2/}{\emph{https://blog.cloudflare.com/http-3-vs-http-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
-. \emph{Available at} \href{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}{\emph{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
-14:00-17:00.ISO/IEC 23009-1:2019.
-\emph{Available at} \href{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}{\emph{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
-Allcock W, Bresnahan J, Kettimuthu R, Link M.
-The Globus Striped GridFTP Framework and Server.
-In: \emph{ACM/IEEE SC 2005 Conference (SC'05)}.
-IEEE,.
-DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
-Anders I, Blanchi Ch, Broeder D, Hellström M, Islam Sh, Jejkal Th, Lannom L, Peters K, Quick R, Schlemmer A, Schwardmann U, Soiland-Reyes S, Strawn G, Uytvanck D van, Weiland C, Wittenburg P, Zwölf C.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}{FAIR digital object overview and specifications}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
-Anders I, Hellström M, Islam S, Jejkal T, Lannom L, Schwardmann U, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}{FDO PID profiles \& attributes}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
-Bahim C, Casorrán-Amilburu C, Dekkers M, Herczog E, Loozen N, Repanas K, Russell K, Stall S.
-2020.
-The FAIR data maturity model: An approach to harmonise FAIR assessments.
-\emph{Data Science Journal} 19.
-DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
-Belshe M, Peon R. 2015. \emph{Hypertext Transfer Protocol Version 2 (HTTP/2)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
-Berners-Lee T, Fielding R, Masinter L. 2005. \emph{Uniform Resource Identifier (URI): Generic Syntax}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
-Berners-Lee T, Fischetti M. 1999. \emph{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}. San Francisco: HarperSanFrancisco.
-
-\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
-Bernstein A, Hendler J, Noy N. 2016. A new look at the semantic web. \emph{Communications of the ACM} 59:35--37. DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
-Bioschemas - Bioschemas. \emph{Available at} \href{http://bioschemas.org/}{\emph{http://bioschemas.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
-Bizer C, Heath T, Berners-Lee T.
-2009.
-Linked data - the story so far.
-\emph{International journal on Semantic Web and information systems} 5:1--22.
-DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
-Blanchi C, Broeder D, Jejkal T, Sharif I, Schlemmer A, Uytvanck D van, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}{FDO -- upload of FDO}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
-Blanchi C, Hellström M, Lannom L, Schwardmann U, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}{Implementation of attributes, types, profiles and registries}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
-Bonino Da Silva Santos LO, Wilkinson MD, Kuzniar A, Kaliyaperumal R, Thompson M, Dumontier M, Burger K.
-2016.
-\href{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}{FAIR data points supporting big data interoperability}.
-In: Zelm M, Doumeingts G, Mendonça JP eds.
-\emph{Enterprise interoperability in the digitized and networked factory of the future}.
-iSTE Press, 270--279.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
-Bonino L, Wittenburg O, Carroll B, Hardisty A, Leggott M, Zwölf C.
-2019.
-\emph{\href{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}{FAIR digital object framework}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
-Carriero VA, Daquino M, Gangemi A, Nuzzolese AG, Peroni S, Presutti V, Tomasi F.
-2020.
-The landscape of ontology reuse approaches.
-In: Cota G, Daquino M, Pozzato GL eds.
-\emph{Applications and practices in ontology design, extraction, and reasoning}.
-Studies on the semantic web.
-IOS Press,.
-DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/SSW200033}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
-Ciccarese P, Soiland-Reyes S, Belhajjame K, Gray AJ, Goble C, Clark T.
-2013.
-PAV ontology: provenance, authoring and versioning.
-\emph{Journal of Biomedical Semantics} 4:37.
-DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
-Clemm G, Amsden J, Ellison T, Kaler C, Whitehead J.
-2002.
-\emph{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}.
-RFC Editor.
-DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
-Content negotiation - HTTP \textbar{} MDN.
-\emph{Available at} \href{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}{\emph{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content\_negotiation}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
-Corcho O, Eriksson M, Kurowski K, Ojsteršek M, Choirat C, Sanden M van de, Coppens F.
-2021.
-\emph{EOSC interoperability framework}.
-Publications Office of the EU.
-DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
-Data - W3C. \emph{Available at} \href{https://www.w3.org/standards/semanticweb/data}{\emph{https://www.w3.org/standards/semanticweb/data}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2. \emph{Available at} \href{https://www.w3.org/TR/vocab-dcat-2/}{\emph{https://www.w3.org/TR/vocab-dcat-2/}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
-Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
-Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
-DCMI Metadata Terms.
-2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
-DCMI Metadata Terms.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
-Delgado JCM.
-2016.
-An Interoperability Framework and Distributed Platform for Fast Data Applications.
-In: \emph{Data Science and Big Data Computing}.
-Springer International Publishing, 3--39.
-DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
-Designing a Linked Data developer experience.
-2018.
-\emph{Available at} \href{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}{\emph{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
-Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation. \emph{Available at} \href{https://www.dona.net/doipv1doc}{\emph{https://www.dona.net/doipv1doc}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
-DOIP and Examples --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip.html}{\emph{https://www.cordra.org/documentation/api/doip.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
-DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
-draft-bhutton-json-schema-00.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}{\emph{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
-draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}{\emph{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
-draft-ietf-quic-http-34.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}{\emph{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
-draft-kelly-json-hal-08.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}{\emph{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
-Duerst M, Suignard M. 2005. \emph{Internationalized Resource Identifiers (IRIs)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
-Dusseault L (ed.). 2007. \emph{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
-FAIR Digital Object Framework Documentation. \emph{Available at} \href{https://fairdigitalobjectframework.org/}{\emph{https://fairdigitalobjectframework.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
-FAIR Digital Objects Forum \textbar. \emph{Available at} \href{https://fairdo.org/}{\emph{https://fairdo.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
-Fensel D, Facca FM, Simperl E, Toma I. 2011. \emph{Semantic Web Services}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
-Fielding RT.
-2000.
-\href{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}{Architectural styles and the design of network-based software architectures}.
-\{THESIS\}.\{DOCTORAL\} Thesis.
-University of California, Irvine.
-
-\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
-Fielding R, Nottingham M, Reschke J (eds.). 2022. \emph{HTTP Semantics}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
-Fielding R, Reschke J (eds.). 2014b. \emph{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
-Fielding R, Reschke J (eds.). 2014a. \emph{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
-Fielding RT, Taylor RN, Erenkrantz JR, Gorlick MM, Whitehead J, Khare R, Oreizy P.
-2017.
-Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award).
-In: \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017}.
-New York, New York, USA: ACM Press, 4--14.
-DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
-FOAF Vocabulary Specification. \emph{Available at} \href{http://xmlns.com/foaf/spec/}{\emph{http://xmlns.com/foaf/spec/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
-Foundation D. 2018. \emph{\href{https://hdl.handle.net/0.DOIP/DOIPV2.0}{Digital object interface protocol specification, version 2.0}}. DONA foundation.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
-Gayo JEL, Prud'hommeaux E, Boneva I, Kontokostas D.
-2017.
-Validating RDF Data.
-\emph{Synthesis Lectures on the Semantic Web: Theory and Technology} 7:1--328.
-DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
-Goble C, Stevens R.
-2008.
-State of the nation in data integration for bioinformatics.
-\emph{Journal of Biomedical Informatics} 41:687--693.
-DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
-Gregorio J, de hOra B (eds.). 2007. \emph{The Atom Publishing Protocol}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
-Gregorio J, Fielding R, Hadley M, Nottingham M, Orchard D. 2012. \emph{URI Template}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
-Groth P, Loizou A, Gray AJG, Goble C, Harland L, Pettifer S.
-2014.
-API-centric Linked Data integration: The Open PHACTS Discovery Platform case study.
-\emph{Journal of Web Semantics} 29:12--18.
-DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
-Group RDAFDMMW. 2020. FAIR data maturity model: Specification and guidelines. \emph{Research Data Alliance}. DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
-Hardisty A, Brack P, Goble C, Livermore L, Scott B, Groom Q, Owen S, Soiland-Reyes S. 2022. The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
-Digital Object Approach to Speeding up Digital Mobilisation of Natural History
-Collections. \emph{Data Intelligence} 4:320--341. DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
-Hasnain A, Rebholz-Schuhmann D.
-2018.
-Assessing FAIR data principles against the 5-star open data principles.
-In: Gangemi A, Gentile AL, Nuzzolese AG, Rudolph S, Maleshkova M, Paulheim H, Pan JZ, Alam M eds.
-\emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers}.
-Lecture notes in computer science.
-Cham: Springer International Publishing, 469--477.
-DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
-Horrocks I, Hendler J (eds.). 2002. \emph{The Semantic Web --- ISWC 2002}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
-HTML Standard. \emph{Available at} \href{https://html.spec.whatwg.org/multipage/microdata.html}{\emph{https://html.spec.whatwg.org/multipage/microdata.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
-Hu W, Chen J, Zhang H, Qu Y.
-2011.
-How matchable are four thousand ontologies on the semantic web.
-In: Antoniou G, Grobelnik M, Simperl E, Parsia B, Plexousakis D, De Leenheer P, Pan J eds.
-\emph{The semantic web: Research and applications}.
-Lecture notes in computer science.
-Berlin, Heidelberg: Springer Berlin Heidelberg, 290--304.
-DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI}{\emph{https://www.w3.org/Provider/Style/URI}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI.html}{\emph{https://www.w3.org/Provider/Style/URI.html}} (accessed 26 May 2022b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
-Iyengar J, Thomson M (eds.). 2021. \emph{QUIC: A UDP-Based Multiplexed and Secure Transport}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
-JSON-LD 1.1. \emph{Available at} \href{http://www.w3.org/TR/json-ld/}{\emph{http://www.w3.org/TR/json-ld/}} (accessed 26 May 2022b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
-JSON-LD 1.1. \emph{Available at} \href{https://www.w3.org/TR/json-ld/}{\emph{https://www.w3.org/TR/json-ld/}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
-Juty N, Le Novere N, Laibe C.
-2011.
-Identifiers.org and MIRIAM Registry: community resources to provide persistent identification.
-\emph{Nucleic Acids Research} 40:D580--D586.
-DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
-Kahn R, Wilensky R. 1995. \emph{\href{http://www.cnri.reston.va.us/k-w.html}{A framework for distributed digital object services}}. CNRI.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
-Kahn R, Wilensky R.
-2006.
-A framework for distributed digital object services.
-\emph{International Journal on Digital Libraries} 6:115--123.
-DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
-Kamdar MR, Tudorache T, Musen MA.
-2017.
-A systematic analysis of term reuse and term overlap across biomedical ontologies.
-\emph{Semantic Web} 8:853--871.
-DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
-Khare R, Lawrence S. 2000. \emph{Upgrading to TLS Within HTTP/1.1}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
-Klein M, Van de Sompel H, Sanderson R, Shankar H, Balakireva L, Zhou K, Tobin R.
-2014.
-Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot.
-\emph{PLoS ONE} 9:e115253.
-DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
-Klímek J, Škoda P, Nečaský M. 2019. Survey of tools for linked data consumption. \emph{Satya Widya} 10:665--720. DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/SW-180316}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
-Lamprecht A-L, Palmblad M, Ison J, Schwämmle V, Al Manir MS, Altintas I, Baker CJO, Ben Hadj Amor A, Capella-Gutierrez S, Charonyktakis P, Crusoe MR, Gil Y, Goble C, Griffin TJ, Groth P, Ienasescu H, Jagtap P, Kalaš M, Kasalica V, Khanteymoori A, Kuhn T, Mei H, Ménager H, Möller S, Richardson RA, Robert V, Soiland-Reyes S, Stevens R, Szaniszlo S, Verberne S, Verhoeven A, Wolstencroft K.
-2021.
-Perspectives on automated composition of workflows in the life sciences.
-\emph{F1000Research} 10:897.
-DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
-Lannom L, Peters-von Gehlen K, Anders I, Pfeil A, Schlemmer A, Trautt Z, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}{FDO configuration types}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
-Lannom L, Schwardmann U, Blanchi C, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}{Typing FAIR digital objects}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
-Linked Data - Design Issues. \emph{Available at} \href{https://www.w3.org/DesignIssues/LinkedData.html}{\emph{https://www.w3.org/DesignIssues/LinkedData.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
-Linked Data Notifications. \emph{Available at} \href{https://www.w3.org/TR/ldn/}{\emph{https://www.w3.org/TR/ldn/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
-Linked Data Platform Working Group. 2015. \emph{\href{https://www.w3.org/TR/2015/REC-ldp-20150226/}{Linked data platform 1.0}}. W3C.
-
-\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
-Loo T (ed.). 1970. First International Conference on FAIR Digital Objects. DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
-martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
-May W. 2017. \emph{HTTP Live Streaming}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
-Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
-2021b.
-Web data apis over SPARQL.
-In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
-Synthesis lectures on data, semantics, and knowledge.
-Cham: Springer International Publishing, 27--38.
-DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
-Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
-2021a.
-Conclusion and future challenges.
-In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
-Synthesis lectures on data, semantics, and knowledge.
-Cham: Springer International Publishing, 77--80.
-DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
-Mika P. 2015. On Schema.org and Why It Matters for the Web. \emph{IEEE Internet Computing} 19:52--55. DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
-Mons B, Neylon C, Velterop J, Dumontier M, Silva Santos LOB da, Wilkinson MD.
-2017.
-Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud.
-\emph{Information Services \& Use} 37:49--56.
-DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{ISU\}-170824}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
-NCBO BioPortal. \emph{Available at} \href{https://bioportal.bioontology.org/ontologies}{\emph{https://bioportal.bioontology.org/ontologies}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
-Neumann A, Laranjeiro N, Bernardino J.
-2021.
-An analysis of public REST web service apis.
-\emph{IEEE Transactions on Services Computing} 14:957--970.
-DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/TSC.2018.2847344}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
-Nottingham M. 2017. \emph{Web Linking}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
-OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More.
-\emph{Available at} \href{https://spec.openapis.org/oas/v3.1.0.html}{\emph{https://spec.openapis.org/oas/v3.1.0.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
-ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
-OWL 2 Web Ontology Language Document Overview (Second Edition). \emph{Available at} \href{http://www.w3.org/TR/owl2-overview/}{\emph{http://www.w3.org/TR/owl2-overview/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
-Page KR, De Roure DC, Martinez K.
-2011.
-REST and Linked Data.
-In: \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11}.
-ACM Press,.
-DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
-Polleres A, Kamdar MR, Fernández JD, Tudorache T, Musen MA.
-2020.
-A more decentralized vision for linked data.
-\emph{Satya Widya} 11:101--113.
-DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/SW-190380}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 January 2023a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
-RDF 1.1 Primer. \emph{Available at} \href{http://www.w3.org/TR/rdf11-primer/}{\emph{http://www.w3.org/TR/rdf11-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
-RDF Schema 1.1. \emph{Available at} \href{http://www.w3.org/TR/rdf-schema/}{\emph{http://www.w3.org/TR/rdf-schema/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
-RDFa 1.1 Primer - Third Edition. \emph{Available at} \href{http://www.w3.org/TR/rdfa-primer/}{\emph{http://www.w3.org/TR/rdfa-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
-Resource Description Framework (RDF) Model and Syntax Specification.
-\emph{Available at} \href{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}{\emph{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
-ResourceSync Framework Specification - Table of Contents. \emph{Available at} \href{http://www.openarchives.org/rs/toc}{\emph{http://www.openarchives.org/rs/toc}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
-Sandvine.Global Internet Phenomena Report 2022.
-\emph{Available at} \href{https://www.sandvine.com/global-internet-phenomena-report-2022}{\emph{https://www.sandvine.com/global-internet-phenomena-report-2022}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
-Sauermann L, Cyganiak R, Völkel M. 2011. \emph{Cool URIs for the semantic web}. Universität des Saarlandes. DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
-Schema.org - Schema.org. \emph{Available at} \href{https://schema.org/}{\emph{https://schema.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
-Schema.org Actions - schema.org. \emph{Available at} \href{https://schema.org/docs/actions.html}{\emph{https://schema.org/docs/actions.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
-Schultes E, Wittenburg P.
-2019.
-FAIR principles and digital objects: Accelerating convergence on a data infrastructure.
-In: Manolopoulos Y, Stupnikov S eds.
-\emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers}.
-Communications in computer and information science.
-Cham: Springer International Publishing, 3--16.
-DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
-Schwardmann U, Kálmán T.
-2022.
-Two Examples on How FDO Types can Support Machine and Human Readability.
-\emph{Research Ideas and Outcomes} 8.
-DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
-Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
-Shape Expressions (ShEx) 2.1 Primer. \emph{Available at} \href{http://shex.io/shex-primer/}{\emph{http://shex.io/shex-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
-Shapes Constraint Language (SHACL). \emph{Available at} \href{https://www.w3.org/TR/shacl/}{\emph{https://www.w3.org/TR/shacl/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
-SKOS Simple Knowledge Organization System Primer. \emph{Available at} \href{http://www.w3.org/TR/skos-primer/}{\emph{http://www.w3.org/TR/skos-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
-Smith B, Ashburner M, Rosse C, Bard J, Bug W, Ceusters W, Goldberg LJ, Eilbeck K, Ireland A, Mungall CJ, Leontis N, Rocca-Serra P, Ruttenberg A, Sansone S-A, Scheuermann RH, Shah N, Whetzel PL, Lewis S.
-2007.
-The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration.
-\emph{Nature Biotechnology} 25:1251--1255.
-DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
-SPARQL 1.1 Overview. \emph{Available at} \href{http://www.w3.org/TR/sparql11-overview/}{\emph{http://www.w3.org/TR/sparql11-overview/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
-Stallings W. 1990. \emph{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}. Carmel, IN, USA: Sams.
-
-\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
-Stanczyk SK. 1987. Process modelling for information system description. \emph{The Open University}. DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
-Stefi A.
-2015.
-Do Developers Make Unbiased Decisions?
-- The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components.
-DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
-Stefi A, Hess T.
-2015.
-To develop or to reuse?
-Two perspectives on external reuse in software projects.
-In: Fernandes JM, Machado RJ, Wnuk K eds.
-\emph{Software business}.
-Lecture notes in business information processing.
-Cham: Springer International Publishing, 192--206.
-DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
-Strawn G, Wittenburg P, Soiland-Reyes St, Peters K, Lannom L, Schwardmann U, Blanch Ch.
-2022.
-\emph{\href{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}{FDO Forum FDO requirement specifications}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
-Sun S, Lannom L, Boesch B. 2003. \emph{Handle System Overview}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
-Sun S, Reilly S, Lannom L, Petrone J. 2003. \emph{Handle System Protocol (ver 2.1) Specification}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
-SWORD 3.0 Specification. \emph{Available at} \href{https://swordapp.github.io/swordv3/swordv3.html}{\emph{https://swordapp.github.io/swordv3/swordv3.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
-The Open Graph protocol. \emph{Available at} \href{https://ogp.me/}{\emph{https://ogp.me/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
-Thornton K, Solbrig H, Stupp GS, Labra Gayo JE, Mietchen D, Prud E, Waagmeester A.
-2019.
-Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation.
-In: Hitzler P, Fernández M, Janowicz K, Zaveri A, Gray AJG, Lopez V, Haller A, Hammar K eds.
-\emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings}.
-Lecture notes in computer science.
-Cham: Springer International Publishing, 606--620.
-DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
-Tirmizi S, Aitken S, Moreira DA, Mungall C, Sequeda J, Shah NH, Miranker DP.
-2011.
-Mapping between the OBO and OWL ontology languages.
-\emph{Journal of Biomedical Semantics} 2:S3.
-DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
-Turcoane O.
-2014.
-Linked data, JSON-LD and the semantics of cultural and scientific heritage.
-\emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} 4:95--105.
-DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
-Usage Statistics of JSON-LD for Websites, May 2022.
-\emph{Available at} \href{https://w3techs.com/technologies/details/da-jsonld}{\emph{https://w3techs.com/technologies/details/da-jsonld}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
-Van de Sompel H, Klein M, Jones S, Nelson ML, Warner S, Devaraju A, Huber R, Steinhoff W, Tykhonov V, Boruta L, Meijers E, Soiland-Reyes S, Wilkinson M.
-2022.FAIR Signposting Profile.
-\emph{Available at} \href{https://signposting.org/FAIR/}{\emph{https://signposting.org/FAIR/}} (accessed 5 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
-Van de Sompel H, Nelson M, Sanderson R. 2013. \emph{HTTP Framework for Time-Based Access to Resource States -\/- Memento}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
-Verborgh R, Vander Sande M.
-2020.
-The semantic web identity crisis: In search of the trivialities that never were.
-\emph{Satya Widya} 11:19--27.
-DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/SW-190372}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
-W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures. \emph{Available at} \href{http://www.w3.org/TR/xmlschema11-1/}{\emph{http://www.w3.org/TR/xmlschema11-1/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
-Web Services Description Language (WSDL) Version 2.0 Part 0: Primer. \emph{Available at} \href{http://www.w3.org/TR/wsdl20-primer/}{\emph{http://www.w3.org/TR/wsdl20-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
-WebSockets Standard. \emph{Available at} \href{https://websockets.spec.whatwg.org/}{\emph{https://websockets.spec.whatwg.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
-Weigel T, Plale B, Parsons M, Zhou G, Luo Y, Schwardmann U, Quick R, Hellström M, Kurakawa K.
-2018.
-RDA Recommendation on PID Kernel Information.
-\emph{Research Data Alliance}.
-DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
-Weiland C, Islam S, Broder D, Anders I, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}{FDO machine actionability}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
-Weiland C, Schwardmann U, Wittenburg P, Kirkpatrick C, Hanisch R, Trautt Z.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}{FDO Forum Document Standards}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
-Wieczorek J, Bloom D, Guralnick R, Blum S, Döring M, Giovanni R, Robertson T, Vieglais D.
-2012.
-Darwin Core: An Evolving Community-Developed Biodiversity Data Standard.
-\emph{PLoS ONE} 7:e29715.
-DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
-Wilde E. 2013. \emph{The 'profile' Link Relation Type}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-6DjakjNS}{}}%
-Wilkinson MD, Dumontier M, Aalbersberg IjJ, Appleton G, Axton M, Baak A, Blomberg N, Boiten J-W, da Silva Santos LB, Bourne PE, Bouwman J, Brookes AJ, Clark T, Crosas M, Dillo I, Dumon O, Edmunds S, Evelo CT, Finkers R, Gonzalez-Beltran A, Gray AJG, Groth P, Goble C, Grethe JS, Heringa J, 't Hoen PAC, Hooft R, Kuhn T, Kok R, Kok J, Lusher SJ, Martone ME, Mons A, Packer AL, Persson B, Rocca-Serra P, Roos M, van Schaik R, Sansone S-A, Schultes E, Sengstag T, Slater T, Strawn G, Swertz MA, Thompson M, van der Lei J, van Mulligen E, Velterop J, Waagmeester A, Wittenburg P, Wolstencroft K, Zhao J, Mons B.
-2016.
-The FAIR Guiding Principles for scientific data management and stewardship.
-\emph{Scientific Data} 3.
-DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pa67pUtR}{}}%
-Wilkinson SR, Eisenhauer G, Kapadia AJ, Knight K, Logan J, Widener P, Wolf M.
-2022.
-F*** workflows: when parts of FAIR are missing.
-\emph{arXiv}.
-DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
-Williams AJ, Harland L, Groth P, Pettifer S, Chichester C, Willighagen EL, Evelo CT, Blomberg N, Ecker G, Goble C, Mons B.
-2012.
-Open PHACTS: semantic interoperability for drug discovery.
-\emph{Drug Discovery Today} 17:1188--1198.
-DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
-Wittenburg P, Anders I, Blanchi C, Buurman M, Goble C, Grieb J, Hardisty A, Islam S, Jejkal T, Kálmán T, Kirkpatrick C, Lannom L, Lauer T, Manepalli G, Peters-von Gehlen K, Pfeil A, Quick R, Sanden M van de, Schwardmann U, Soiland-Reyes S, Stotzka R, Trautt Z, Van Uytvanck D, Weiland C, Wieder P.
-2022.
-FAIR digital object demonstrators 2021.
-\emph{Zenodo}.
-DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
-Wittenburg P, Strawn G, Mons B, Bonino L, Schultes E.
-2019.
-Digital objects as drivers towards convergence in data infrastructures.
-\emph{https://b2share.eudat.eu}.
-DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
-Wolstencroft K, Haines R, Fellows D, Williams A, Withers D, Owen S, Soiland-Reyes S, Dunlop I, Nenadic A, Fisher P, Bhagat J, Belhajjame K, Bacall F, Hardisty A, Nieva de la Hidalga A, Balcazar Vargas MP, Sufi S, Goble C.
-2013.
-The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud.
-\emph{Nucleic Acids Research} 41:W557--W561.
-DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
-Wolstencroft K, Owen S, Horridge M, Krebs O, Mueller W, Snoep JL, du Preez F, Goble C.
-2011.
-RightField: embedding ontology annotation in spreadsheets.
-\emph{Bioinformatics} 27:2021--2022.
-DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
-Working Group FDO-TSIG. 2022a. \emph{\href{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}{DOIP endorsement request}}. (internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
-Working Group FDO-TSIG.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}{FDO -- granularity, versioning, mutability}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Fi62cJAo}{}}%
-Working Group FDO-TSIG, Working Group FDO-BIG.
-2022.
-\emph{\href{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}{FDO -- kernel attributes \& metadata}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
-X.1255 : Framework for discovery of identity management information.
-\emph{Available at} \href{https://www.itu.int/rec/T-REC-X.1255-201309-I}{\emph{https://www.itu.int/rec/T-REC-X.1255-201309-I}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRzcHhPD}{}}%
-Zarras A. 2004. A Comparison Framework for Middleware Infrastructures. \emph{The Journal of Object Technology} 3:103. DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}.
-
-\end{CSLReferences}

### 2023-01-24 latex/table2.tex
converted to peerj overleaf template

--- 2023-01-24 manuscript.tex
@@
-This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/51048f35d5fce97aa4194e6426fdd7c704f19199/}{permalink})
-was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/51048f35d5fce97aa4194e6426fdd7c704f19199}{stain/2022-fdo-paper@51048f3}
-on January 24, 2023.
-
-\hypertarget{authors}{%
-\section*{Authors}\label{authors}}
-\addcontentsline{toc}{section}{Authors}
-
-\begin{itemize}
-\item
-  \textbf{Stian Soiland-Reyes}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/stain}{stain}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/soilandreyes}{soilandreyes}
-  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Carole Goble}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/carolegoble}{carolegoble}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
-  Department of Computer Science, The University of Manchester, UK
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Paul Groth}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/pgroth}{pgroth}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/pgroth}{pgroth}
-  Informatics Institute, Faculty of Science, University of Amsterdam, NL
-\end{itemize}
-
-\hypertarget{abstract}{%
-\section*{Abstract}\label{abstract}}
-\addcontentsline{toc}{section}{Abstract}
-
-\textbf{TODO}: Rewrite from notes below to actual abstract.
-
-FAIR Digital Object is an emerging concept from EOSC.
-This is important.
-Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
-Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
-
-What is the message of this paper?
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These are the overlaps
-\item
-  This is what FDO is requiring but not commonly deployed
-\item
-  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
-\item
-  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
-\item
-  Lessons for Semantic Web community. What is missing in FDO and in SW.
-\item
-  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
-\item
-  Go back to the old SW layer cake. Explains why we picked these frameworks.
-\item
-  Lessons for SW: Parts of the stack that is less. FDO contributions.
-\end{enumerate}
-
-Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.
-
-Have all the ingredients, but not cooking properly.
-
-Systematic through frameworks.
-
-Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?
-
-\hypertarget{sec:introduction}{%
-\section{Introduction}\label{sec:introduction}}
-
-The FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures {[}\protect\hyperlink{ref-u8Cy0psL}{2}{]}.
-The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
-
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
-The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
-
-In other discourse, Linked Data {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles {[}\protect\hyperlink{ref-GJHYN6xW}{6},\protect\hyperlink{ref-EJDjHrUv}{7}{]}.
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services {[}\protect\hyperlink{ref-99MS5xE0}{9},\protect\hyperlink{ref-J2H8yssV}{10}{]}.
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability {[}\protect\hyperlink{ref-DNTCUjpC}{11},\protect\hyperlink{ref-Pa67pUtR}{12}{]}.
-
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality {[}\protect\hyperlink{ref-uxWjmlio}{13},\protect\hyperlink{ref-v0WwFHgT}{14}{]} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves {[}\protect\hyperlink{ref-5V3GqSzR}{15}{]}.
-
-These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.
-
-Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.
-
-\hypertarget{sec:background}{%
-\section{Background and related work}\label{sec:background}}
-
-In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.
-
-\hypertarget{sec:fdo}{%
-\subsection{FAIR Digital Object}\label{sec:fdo}}
-
-\hypertarget{sec:next-step-fdo}{%
-\subsubsection{Next steps for FDO}\label{sec:next-step-fdo}}
-
-The FAIR Digital Object Forum {[}\protect\hyperlink{ref-DMEBhHID}{16}{]} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
-As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
-
-\textbf{FAIR Digital Object Overview and Specifications} {[}\protect\hyperlink{ref-11NpETLq1}{17}{]} is a comprehensive overview of FAIR Digital Object specifications listed below.
-It serves as a primer that introduces FDO concepts and the remaining documents.
-
-The \textbf{FDO Forum Document Standards} {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
-In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
-
-The \textbf{FDO Requirement Specifications} {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} is an update of {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} as the foundational definition of FDO.
-This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
-The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
-
-The \textbf{Machine actionability} {[}\protect\hyperlink{ref-iJeys0T5}{21}{]} sets out to define what is meant by \emph{machine actionability} for FDOs.
-\emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
-The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.
-
-\textbf{Configuration Types} {[}\protect\hyperlink{ref-SPd0uEqO}{22}{]} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}
-
-\textbf{PID Profiles \& Attributes} {[}\protect\hyperlink{ref-ljVViWCl}{24}{]} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}.
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
-
-\textbf{Kernel Attributes \& Metadata} {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}.
-It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
-
-\textbf{Granularity, Versioning, Mutability} {[}\protect\hyperlink{ref-16W9AycSW}{27}{]} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
-The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
-
-\textbf{DOIP Endorsement Request} {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} is an endorsement of the DOIP v2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} specification as a potential FDO implementation, as it has been applied by several institutions {[}\protect\hyperlink{ref-jRWurmQc}{30}{]}.
-The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
-
-\textbf{Upload of FDO} {[}\protect\hyperlink{ref-1H7cmmr69}{31}{]} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync {[}\protect\hyperlink{ref-eS5y9TRh}{32}{]} is suggested as one type of service to list FDOs.
-This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
-
-\textbf{Typing FAIR Digital Objects} {[}\protect\hyperlink{ref-ZFzPxCNB}{33}{]} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
-This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
-Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work.
-How to define such FDO operations is not specified.
-
-\textbf{Implementation of Attributes, Types, Profiles and Registries} {[}\protect\hyperlink{ref-126uxr5pI}{34}{]} details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
-This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} {[}\protect\hyperlink{ref-1CqIZ47pu}{35}{]}.
-
-It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-
-The concept of \textbf{FAIR Digital Objects} {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} has been introduced as way to expose research data as active objects that conform to the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}.
-This builds on the \emph{Digital Object} (DO) concept {[}\protect\hyperlink{ref-11MnuwJ4l}{36}{]}, first introduced in 1995 {[}\protect\hyperlink{ref-3Uqe3fuK}{37}{]} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework {[}\protect\hyperlink{ref-103Hw8H43}{38}{]} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]}.
-
-In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}.
-The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing {[}\protect\hyperlink{ref-tz0P3DTC}{40}{]}.
-
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}.
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
-
-\hypertarget{fdo-approaches}{%
-\subsubsection{FDO approaches}\label{fdo-approaches}}
-
-FDO is an evolving concept. A set of FDO Demonstrators {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} highlight how current adapters are approaching implementations of FDO from different angles:
-
-\begin{itemize}
-\tightlist
-\item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, see {[}\protect\hyperlink{ref-xuQKRRx5}{41}{]}.} (usually TCP/IP over TLS).
-The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]} and PIDs are assigned using the Handle system.
-Several type registries have been established.
-\item
-  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
-\item
-  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
-\end{itemize}
-
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
-
-\hypertarget{sec:ld}{%
-\subsection{From the Semantic Web to Linked Data}\label{sec:ld}}
-
-In order to describe \emph{Linked Data} as it is used today, we'll start with an (opinionated) description of the evolution of its foundation, the \emph{Semantic Web}.
-
-\hypertarget{sec:semweb}{%
-\subsubsection{A brief history of the Semantic Web}\label{sec:semweb}}
-
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee {[}\protect\hyperlink{ref-Guy7cjVY}{43}{]}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
-The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
-The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.
-
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) {[}\protect\hyperlink{ref-16WGmCih8}{44}{]} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format {[}\protect\hyperlink{ref-16wvFH4nh}{45}{]} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
-
-While triple-based knowledge representations were not new {[}\protect\hyperlink{ref-NxvhwSmG}{46}{]}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs {[}\protect\hyperlink{ref-RRtwPF7Y}{47}{]} are generalised forms of URLs that include locator-less identifiers
-  such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years {[}\protect\hyperlink{ref-nW4HY8Nq}{48}{]}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) {[}\protect\hyperlink{ref-HqZ3r6J3}{49}{]}.
-RDF 1.1 formats use Unicode to support IRIs {[}\protect\hyperlink{ref-rqjuXkEf}{50}{]}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
-
-The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data).
-Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term.
-The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource.
-By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs.
-The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.
-
-The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} {[}\protect\hyperlink{ref-yd2WE9eq}{52}{]} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
-
-Established Semantic Web services and standards include SPARQL {[}\protect\hyperlink{ref-xww8fb1B}{53}{]} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
-
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins {[}\protect\hyperlink{ref-bMSxa8XW}{54},\protect\hyperlink{ref-Qlrlr2TZ}{55}{]}.
-In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
-
-Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) {[}\protect\hyperlink{ref-FDDmgO6s}{57}{]} .
-
-The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS {[}\protect\hyperlink{ref-191ZkYZMt}{58}{]}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
-
-With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps {[}\protect\hyperlink{ref-9CdJnsoJ}{59}{]}.
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication {[}\protect\hyperlink{ref-AGHR9oyr}{60}{]}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data {[}\protect\hyperlink{ref-5SuwLNdE}{61}{]}.
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies {[}\protect\hyperlink{ref-TUmA4KQT}{62}{]} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS {[}\protect\hyperlink{ref-ZwAcGQKY}{63}{]}, OWL {[}\protect\hyperlink{ref-1p4IWJpI}{64}{]}, OBO {[}\protect\hyperlink{ref-1DalZuKct}{65}{]}, SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]}), hash vs slash, HTTP status codes and PID redirection strategies {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.
-
-\hypertarget{sec:ld-web}{%
-\subsubsection{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
-
-The \textbf{Linked Data} concept {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} was kickstarted as a set of best practices {[}\protect\hyperlink{ref-OGXYtGZ8}{66}{]} to bring the Web aspect back into focus.
-Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
-This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
-
-Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph {[}\protect\hyperlink{ref-oN9D5LkD}{67}{]}, RDFa {[}\protect\hyperlink{ref-FKIF6ApB}{68}{]}, Microdata {[}\protect\hyperlink{ref-14Utscfe4}{69}{]}) we find JSON-LD {[}\protect\hyperlink{ref-X5SWPqSU}{70}{]} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites {[}\protect\hyperlink{ref-j3SqQLBR}{71}{]}.
-
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} {[}\protect\hyperlink{ref-BFMoe8FC}{72}{]} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL {[}\protect\hyperlink{ref-GDrm9dgX}{73}{]} or ShEx {[}\protect\hyperlink{ref-YUzogaV}{74}{]}) can be used to validate RDF Data {[}\protect\hyperlink{ref-S5BI6Mwx}{75},\protect\hyperlink{ref-ZeOAPNHa}{76}{]} before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts {[}\protect\hyperlink{ref-o596f2JM}{77}{]}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
-
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}.
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies {[}\protect\hyperlink{ref-MhsCt4P3}{78}{]}, the use of Linked Data as part of REST Web APIs {[}\protect\hyperlink{ref-eS6a6tlz}{79}{]} , and as shown by the big uptake by publishers to annotate the Web using schema.org {[}\protect\hyperlink{ref-117tX9MEG}{81}{]}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
-
-\hypertarget{sec:comparing}{%
-\section{Comparing FDO and existing approaches}\label{sec:comparing}}
-
-To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
-\item
-  The FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}, validated against its implementations for completeness.
-\item
-  A Comparison Framework for Middleware Infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
-\item
-  Cross-checks against RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-UzQhqk0M}{84}{]} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
-\item
-  EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
-\end{enumerate}
-
-The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).
-
-Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} against Linked Data in general.
-
-\hypertarget{sec:interoperability-compare}{%
-\subsection{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
-
-The Interoperability Framework for Fast Data Applications {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
-
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model {[}\protect\hyperlink{ref-8ZRGFa6y}{85}{]} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
-In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]}.
-\label{tbl:fdo-web-interoperability-framework}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
-FDO encourages research communities to develop common type registries to be shared across instances.
-In current DOIP practice, each service have their own types, attributes and operations.
-The wider symbiosis is consistent use of PIDs.
-& The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs {[}\protect\hyperlink{ref-174AwcFUL}{86}{]} end up being mandated centrally by dominant (often commercial) providers {[}\protect\hyperlink{ref-11VxBOeZy}{87}{]}, which clients are required to use as-is with special code per service.
-Use of Linked Data enables common tooling and semantic mapping across differences.
-\\
-\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP).
-\texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).
-Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}).
-& ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]}, schema.org Actions {[}\protect\hyperlink{ref-dKAekUjL}{89}{]}, WSDL/SOAP {[}\protect\hyperlink{ref-LmudecyN}{90}{]}, but frequently just as human-readable developer documentation/examples.
-\\
-\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing.
-Every FDO has a type.
-Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}.
-No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} & Lightweight HTTP semantics for authenticity/navigation.
-Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
-Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit.
-\\
-\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation} & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema.
-Multiple byte stream attachments of any media type.
-& Textual HTTP headers (including any signposting), single byte stream of any media type, e.g.~Linked Data formats (JSON-LD, Turtle, RDF/XML) or embedded in document (HTML with RDFa, JSON-LD or Microdata).
-XML was previously the main syntax used by APIs, JSON is now dominant.
-\\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
-\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
-Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable).
-Unique DOIP protocol can be hard to add to existing Web application frameworks, although proxy services have been developed (e.g.~B2SHARE adapter).
-& HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to (for static content) multi-cloud Content-Delivery Networks (CDN).
-Current scalable cloud technologies for Web hosting may not support HTTP features previously seen as important for Semantic Web, e.g.~content negotiation and semantic HTTP status codes.
-\\
-\bottomrule
-\end{longtable}
-
-Based on the analysis shown in table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
-
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services {[}\protect\hyperlink{ref-39K3yHT3}{91}{]}.
-However, this is mostly done via developer or human interaction {[}\protect\hyperlink{ref-iYh1wFBt}{92}{]}.
-Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.
-
-A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics.
-Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON.
-On the semantic front, FDO requires that every object have a well-defined type and structured form.
-This is clearly not the case on the Web.
-
-In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed.
-These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles {[}\protect\hyperlink{ref-hxtgCGjY}{93}{]} to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference {[}\protect\hyperlink{ref-jrZe6Esu}{94}{]}.
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP {[}\protect\hyperlink{ref-p7nwRj8E}{95}{]}.
-
-It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web {[}\protect\hyperlink{ref-1H9iQhQYq}{96}{]} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail {[}\protect\hyperlink{ref-68fzID2k}{97}{]}.
-
-\hypertarget{mapping-of-metamodel-concepts}{%
-\subsubsection{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
-
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics {[}\protect\hyperlink{ref-8HJqcF1Q}{98}{]}.
-
 \begin{longtable}[]{@{}
@@
 \end{longtable}
-
-From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
-Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \ref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
-
-\hypertarget{sec:doip-fdo-compare}{%
-\subsection{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
-
-The FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} and using Linked Data Platform {[}\protect\hyperlink{ref-17OHlMRQA}{99}{]}, as proposed by {[}\protect\hyperlink{ref-7szz7dwO}{100}{]}.
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines {[}\protect\hyperlink{ref-yygVPoL0}{19},\protect\hyperlink{ref-RwvirqWg}{20}{]} against its current implementations as DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} and Linked Data Platform (LDP) {[}\protect\hyperlink{ref-7szz7dwO}{100}{]}, with suggestions for required additions.
-\label{tbl:fdo-checks}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endhead
-G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
-& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
-\texttt{http://} URIs replaced by \texttt{https://}.
-& Keep flexibility of RDF serialisation formats which may change more frequently \\
-G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
-DOIP relatively unknown, in effect only one implementation.
-& Further promote DOIP and justify its benefits.
-Build tutorials and OSI open source implementations.
-Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites {[}\protect\hyperlink{ref-j3SqQLBR}{71}{]}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
-Example-driven documentation.
-\\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
-\\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
-Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
-\\
-G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
-DOIP operations can use any transport protocol.
-& Document transport protocols as FDOs, recommend which transport to use.
-& URI as abstraction base.
-Does not specify PID requirements.
-& Give stronger deployment recommendations.
-\\
-G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
-Unclear when metadata field is a PID or plain text.
-& Make datatype of fields explicit to support navigation.
-& Machine-navigation through URIs via properties and types.
-Unclear when URI should be followed or is just identifier, but always distinct from plain text.
-& \\
-G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
-& Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
-& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
-Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]} \\
-G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
-No FDOF implementations.
-& Develop demonstrator of FDOF usage based on existing LDP server.
-\\
-G9: \emph{standard compliance} & Handle {[}\protect\hyperlink{ref-2gXY2Nwh}{101}{]}, DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}.
-FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} & HTTP, LDP.
-FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
-FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve {[}\protect\hyperlink{ref-ljVViWCl}{24}{]}.
-Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-& HTTP URLs as basis for identifiers, but they are frequently not persistent.
-& Add strong guidance for PID services like w3id and persistence policies.
-\\
-FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
-Datacite DOIs content-negotiate to give registered metadata.
-& Add FAIR Signposting at PID provider for minimal PID record \\
-FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
-No method/attribute defined for separate metadata, only directly in PID Record.
-Unclear meaning of multiple items and bytestream chunks.
-& Clarify expectations for multiple items.
-Recommend chunks to not be used.
-& URIs resolvable by default.
-Multiple ways to resolve metadata, unclear preference.
-& Add FAIR Signposting and preference order.
-\\
-FDOF4: \emph{Additional attributes} & Freetext attribute keys.
-Attributes should be defined for FDO type (?).
-& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
-Explicitly allow attributes not already defined in type.
-& All attributes individually identified.
-Any Linked Data attributes can be used by URI or with mapped prefix.
-& Clarify type expectations of required/recommended/optional attributes.
-\\
-FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
-& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-Clarify that operations can be mapped to protocol directly.
-& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
-Unclear how to provide PID for additional operations.
-& Specify how additional operations should be called over HTTP.
-\\
-FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns {[}\protect\hyperlink{ref-i9Ang0rM}{102}{]} are used instead.
-& Document how operation resources can be discovered from an LPD container.
-Document search API.
-\\
-FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
-PR-TypingFDOs leaves this open.
-& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
-Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
-& Always make LDP types explicit in FDO profile.
-\\
-FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
-Separate Metadata FDO need custom property.
-& Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
-& Assertions are always with semantics, using RDF vocabularies.
-Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
-& Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
-FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
-Metadata would need to be bundled within custom FDO types/attributes.
-Unclear how levels are separated within single FDO representation (need FDOF8?).
-& Declare which metadata are expected within response attribute or within FDO object.
-Require PIDs for custom attributes.
-Define how alternate metadata levels can be represented separately.
-& Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
-& Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
-FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
-& Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} {[}\protect\hyperlink{ref-CNP39tah}{103}{]} & Rather document better how individual ad-hoc schemas can be started for prototypes.
-\\
-FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
-Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-{[}\protect\hyperlink{ref-6LvydTsC}{104}{]}).
-& & LDP Containers required by specification, also user-created (eg.
-\texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 {[}\protect\hyperlink{ref-1EW4hLN8U}{105}{]}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE {[}\protect\hyperlink{ref-gtGuSevK}{106}{]} \\
-FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
-\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
-No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
-\bottomrule
-\end{longtable}
-
-Note that the draft update to FDO specification {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
-
-A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines.
-Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
-
-From our evaluation, we can observe:
-
-\begin{itemize}
-\tightlist
-\item
-  G1 and G2 call for stability and trustworthiness.
-While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
-\item
-  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data.
-Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous.
-In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
-\item
-  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}.
-These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
-\item
-  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
-\item
-  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
-\item
-  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} {[}\protect\hyperlink{ref-9T4j3N4e}{107}{]}
-\item
-  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
-\item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
-\item
-  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation.
-In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
-\end{itemize}
-
-\hypertarget{sec:middleware}{%
-\subsection{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
-
-In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
-As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.
-
-We argue that the Web, with its now ubiquitous use of REST API {[}\protect\hyperlink{ref-174AwcFUL}{86}{]}, can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services {[}\protect\hyperlink{ref-WzPECPDP}{108}{]} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} {[}\protect\hyperlink{ref-11dInoyVF}{109}{]}.
-
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}) and Web technologies (using Linked Data) as middleware infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]}
-\label{tbl:fdo-web-middleware}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
-Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
-PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}.
-& The Web is inherently open and made by cross-linked URLs.
-Participation requires DNS domain purchase (many free alternatives also exists).
-PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects.
-\\
-\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
-HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
-\\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second {[}\protect\hyperlink{ref-18P9txgeB}{110}{]}.
-DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
-Multiple connections possible.
-Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks {[}\protect\hyperlink{ref-F8VZ86hu}{111}{]}.
-HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams {[}\protect\hyperlink{ref-KbbW0kGT}{112}{]} but still has TCP+TLS startup costs.
-For reduced latency {[}\protect\hyperlink{ref-CJuBupCW}{113}{]}, HTTP/3 {[}\protect\hyperlink{ref-Amit86Tp}{114}{]} use QUIC {[}\protect\hyperlink{ref-1GWvyEmUf}{115}{]}) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
-\\
-\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
-FDOs should provide the mandatory kernel metadata fields.
-FDOs of the same declared type typically share additional attributes (although that schema may not be declared).
-DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions.
-& Each URL accessed separately.
-Common HTTP headers provide basic metadata, although it is often not reliable.
-A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates {[}\protect\hyperlink{ref-11hxwwuRt}{116}{]}, content-negotiation {[}\protect\hyperlink{ref-n2f3TfoI}{117}{]} or FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}.
-\\
-\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
-Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
-& Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
-\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
-Multiple locations possible in Handle system, can expose geo-info.
-& PIDs and URL redirects.
-DNS aliases and IP routing can hide location.
-Geo-localised servers common for large cloud deployments.
-\\
-\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
-FDO kernel metadata can include checksum and date.
-& HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
-Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
-\\
-\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
-Custom codes uncommon.
-\\
-\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
-No redirection from DOIP service.
-& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
-\\
-\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
-There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged {[}\protect\hyperlink{ref-rbG9uRKw}{118}{]}.
-Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
-An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento {[}\protect\hyperlink{ref-DNjcMi4a}{119}{]} expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method {[}\protect\hyperlink{ref-hRUsaTiz}{120}{]} (used by SVN).
-\\
-\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
-Internal synchronisation possible in backend for Extended operations.
-& Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} {[}\protect\hyperlink{ref-xvMAyRAc}{121}{]} with \texttt{LOCK} and \texttt{UNLOCK} methods.
-\\
-\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
-In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
-& The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space.
-In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend.
-\\
-\textbf{Encapsulation}: \emph{separate interface from implementation.
-Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
-FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
-Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade {[}\protect\hyperlink{ref-1B16WfR7q}{122}{]} to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns {[}\protect\hyperlink{ref-iME5l1kk}{123}{]} and PIDs via Persistent URLs (PURL).
-\\
-\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
-FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]}) & Syntactically Media Type with multiple suffixes {[}\protect\hyperlink{ref-nlH2rm1s}{124}{]} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) {[}\protect\hyperlink{ref-nCzRO0pK}{125}{]}.
-In metadata, semantic type systems (RDFS {[}\protect\hyperlink{ref-ZwAcGQKY}{63}{]}, OWL2 {[}\protect\hyperlink{ref-1p4IWJpI}{64}{]}, SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]}).
-OpenAPI 3 {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]} \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} {[}\protect\hyperlink{ref-11oVTMsuZ}{126}{]}, JSON \texttt{\$schema} {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]}), JSON-LD \texttt{@context} {[}\protect\hyperlink{ref-by6CfOmv}{127}{]}.
-Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
-\\
-\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} {[}\protect\hyperlink{ref-KbbW0kGT}{112}{]}, Web Sockets {[}\protect\hyperlink{ref-qgqADk45}{128}{]}, Linked Data Notifications {[}\protect\hyperlink{ref-zlGYiuWC}{129}{]}, AtomPub {[}\protect\hyperlink{ref-IbaebabD}{130}{]}, SWORD {[}\protect\hyperlink{ref-7DGllKzG}{131}{]}, Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} {[}\protect\hyperlink{ref-HE7Ikwwl}{132}{]}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates {[}\protect\hyperlink{ref-11hxwwuRt}{116}{]}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State {[}\protect\hyperlink{ref-174AwcFUL}{86}{]}, an important element of the REST architectural style.} incl.~Hydra {[}\protect\hyperlink{ref-13PaKBrZe}{133}{]}, schema.org Actions {[}{[}\protect\hyperlink{ref-dKAekUjL}{89}{]}), JSON HAL {[}\protect\hyperlink{ref-nyrPYHoP}{134}{]} \& Link headers (RFC8288) {[}\protect\hyperlink{ref-ozdWB3O7}{135}{]} \\
-\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 {[}\protect\hyperlink{ref-tIkSXrb5}{136}{]} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) {[}\protect\hyperlink{ref-PL512B0X}{137}{]}, MPEG-DASH (ISO/IEC 23009-1:2019) {[}\protect\hyperlink{ref-ikO7e0mh}{138}{]} \\
-\bottomrule
-\end{longtable}
-
-Based on the analysis in Table \ref{tbl:fdo-web-middleware}, we make the following observations:
-
-\begin{itemize}
-\tightlist
-\item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]} supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}.
-Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
-It is unclear what is the purpose of splitting a binary into chunks on a channel which no longer can be multiplexed and the only property of a chunk is its size\footnote{Although it is possible with \texttt{0.DOIP/Op.Retrieve} to request only particular individual elements of an DO (e.g.~one file), unlike HTTP's \texttt{Range} request, it is not possible to select individual chunks of an element's bytestream.}.
-\item
-  HTTP has strong support for scalability and caching, but this mostly assumes read-operations from static resources.
-FDO has no view on immutability or validity of retrieved objects, but this should be taken into consideration to support large-scale usage.
-\item
-  HTTP optimisations for performance (e.g.~HTTP/2, multiplexing) is largely used for commercial media distribution (e.g.~Netflix), and not commonly used by providers of FAIR data
-\item
-  Cloud deployment of Web applications give many middleware benefits (Scalability, Distribution, Access transparancy, Location transparancy) -- it is unclear how DOIP as a custom protocol would perform in a cloud setting as most of this infrastructure assumes HTTP as the protocol.
-\item
-  Programmatically the Web is rather unstructured as middleware, as there are many implementation choices.
-Usually it is undeclared what to expect for a given URI/service, and programmers follow documented examples for a particular service rather than automated programmatic exploration across providers.
-This mean one can consider the Web as an ecosystem of smaller middlewares with commonalities.
-\item
-  Many providers of FAIR Linked Data also provide programmatic REST API endpoints, e.g.~\href{https://www.uniprot.org/help/programmatic_access}{UNIPROT}, \href{https://chembl.gitbook.io/chembl-interface-documentation/web-services}{ChEMBL}, but keeping the FAIR aspects such as retrieving metadata in such a scenario may require combining different services using multiple formats and identifier conventions.
-\end{itemize}
-
-\hypertarget{sec:fair-compare}{%
-\subsection{Assessing FDO against FAIR}\label{sec:fair-compare}}
-
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} {[}\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}{]}.
-
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform {[}\protect\hyperlink{ref-7szz7dwO}{100}{]} comply with the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}.
-Here we've used the RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-1GrKTFaK2}{139}{]} as it has decomposed the FAIR principles to a structured list of FAIR indicators {[}\protect\hyperlink{ref-UzQhqk0M}{84}{]}, importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-UzQhqk0M}{84},\protect\hyperlink{ref-1GrKTFaK2}{139}{]} (first 2 columns) against the FDO guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}, FDO implemented with the protocol DOIPv2 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, Linked Data Platform (LDP) {[}\protect\hyperlink{ref-7szz7dwO}{100}{]} and examples from Linked Data practices in general.
-(--- indicates \emph{Unspecified}, may be possible with additional conventions)
-\label{tbl:fair-data-maturity-model}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endhead
-RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
-RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
-Handle, DOI.
-& FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{[}\protect\hyperlink{ref-rbG9uRKw}{118}{]};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
-RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
-FAIR Signposting.
-\\
-RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
-RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
-Unclear how to link to additional metadata.
-& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
-Embedded in landing page (RDFa).
-\\
-RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
-& PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
-RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
-Handle registry only searchable by PID.
-& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-{[}\protect\hyperlink{ref-1Gsq23e51}{140}{]}).
-Indexing by search engines if exposing HTML w/schema.org.
-\\
-RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
-\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
-RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
-RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
-RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
-RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
-RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
-Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP {[}\protect\hyperlink{ref-8HJqcF1Q}{98}{]} & HTTP/HTTPS, FTP (now less common), GridFTP {[}\protect\hyperlink{ref-18LsXvKGJ}{141}{]} (for large data), ARK {[}\protect\hyperlink{ref-voDQ6diM}{142}{]} \\
-RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol {[}\protect\hyperlink{ref-I9Kx0Hjy}{143}{]}.
-One server implementation {[}\protect\hyperlink{ref-1CvYRVhy2}{144}{]}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved {[}\protect\hyperlink{ref-xVJMmGVr}{145}{]} as an open source license -- it includes usage restrictions and requires Service Agreements.
-It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
-One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
-\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
-& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
-\\
-RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
-RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
-RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
-RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
-RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
-RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
-RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
-RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
-& FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) {[}\protect\hyperlink{ref-1BlxdxOwa}{146}{]} \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs {[}\protect\hyperlink{ref-o4h92LpS}{147}{]}), \texttt{rdfs:seeAlso}, FAIR signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]} \texttt{describedby} \\
-RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
-RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
-RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles {[}\protect\hyperlink{ref-QpDmajc4}{148}{]}, \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role {[}\protect\hyperlink{ref-10HA0mMBx}{149}{]}, \texttt{prov:specializationOf} {[}\protect\hyperlink{ref-ZUgwvAHE}{150}{]}, OAI-ORE Proxy {[}\protect\hyperlink{ref-ID1pTGry}{151}{]}) \\
-RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} but not assigned PIDs yet.
-& Unspecified.
-Multiple metadata records can allow multiple semantic profiles.
-& Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org {[}\protect\hyperlink{ref-10pzKCEUE}{152}{]}, Dublin Core Terms {[}\protect\hyperlink{ref-wAbZpOsr}{153}{]}, DCAT {[}\protect\hyperlink{ref-lnH0hEuh}{154}{]}, FOAF {[}\protect\hyperlink{ref-XmUY0cZB}{155}{]} \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} {[}\protect\hyperlink{ref-lnH0hEuh}{154}{]} \\
-RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
-RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
-RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
-PID Kernel attributes?
-Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O {[}\protect\hyperlink{ref-9T4j3N4e}{107}{]}, PAV {[}\protect\hyperlink{ref-1E4OfHyhI}{156}{]}, Dublin Core Terms {[}\protect\hyperlink{ref-hKblceo6}{157}{]} \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen {[}\protect\hyperlink{ref-zZTtJruI}{158}{]}) & --- & Common, e.g.~DCAT 2 {[}\protect\hyperlink{ref-zt8AAjcZ}{159}{]}, BioSchemas {[}\protect\hyperlink{ref-1F8S9FeHf}{160}{]} \\
-RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal {[}\protect\hyperlink{ref-1Gsq23e51}{140}{]}, Darwin Core {[}\protect\hyperlink{ref-9jaVzCbL}{161}{]} \\
-RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
-\bottomrule
-\end{longtable}
-
-From this evaluation we observe:
-
-\begin{itemize}
-\tightlist
-\item
-  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
-\item
-  FDO/DOIP are stronger on identifier indicators
-\item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} has currently only a couple of implementations and is expressed informally.
-The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
-\item
-  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
-\item
-  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
-\item
-  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-{[}\protect\hyperlink{ref-126uxr5pI}{34}{]} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
-Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
-\item
-  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata.
-For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
-In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
-\item
-  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications.
-\texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known.
-g
-\end{itemize}
-
-\hypertarget{eosc-interoperability-framework}{%
-\section{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
-
-\textbf{TODO}: Introduce EOSC IF
-
-The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} in section 3.6 recommends:
-
-\begin{tablenos:no-prefix-table-caption}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Layer
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Recommendation
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-FDO
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
-Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
-Technical & Easy SLAs for service providers & Unspecified & None \\
-Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
-Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
-Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
-Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
-Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
-Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
-Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
-Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} & DCAT, ++ \\
-Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
-Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
-Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
-Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
-Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
-Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
-Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
-Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
-Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
-Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
-Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
-Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
-Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
-Legal & List recommended licenses & Undefined & Best practice recommendations \\
-Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
-Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
-Legal & GDPR compliance for personal data & Undefined & Undefined \\
-Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
-Legal & Harmonised terms-of-use & Undefined & Undefined \\
-Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
-\bottomrule
-\end{longtable}
-
-\end{tablenos:no-prefix-table-caption}
-
-Observations:
-
-\begin{itemize}
-\tightlist
-\item
-  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
-\item
-  Technical aspects highlighted by EOSC IF
-\item
-  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data.
-There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
-\item
-  FDO principles have strong recommendations for community development of organisational aspects.
-\item
-  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
-\end{itemize}
-
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
-
-\textbf{TODO}
-
-\begin{itemize}
-\tightlist
-\item
-  Ramnifications of ide
-\item
-  Finish tables
-\item
-  Bullet points per table
-\item
-  Send in Google Docs
-\end{itemize}
-
-\hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
-
-The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake {[}\protect\hyperlink{ref-kYjzRJ0H}{162}{]}.
-However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
-
-While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
-
-We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-\ldots{}
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-6DjakjNS}{}}%
-\CSLLeftMargin{1. }
-\CSLRightInline{\textbf{The FAIR Guiding Principles for scientific data management and stewardship}
-\CSLBlock{Mark D Wilkinson, Michel Dumontier, IJsbrand Jan Aalbersberg, Gabrielle Appleton, Myles Axton, Arie Baak, Niklas Blomberg, Jan-Willem Boiten, Luiz Bonino da Silva Santos, Philip E Bourne, \ldots{} Barend Mons} \emph{Scientific Data} (2016-03-15)
-\CSLBlock{DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/26978244}{26978244} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4792175}{PMC4792175}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
-\CSLLeftMargin{2. }
-\CSLRightInline{\textbf{Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud}
-\CSLBlock{Barend Mons, Cameron Neylon, Jan Velterop, Michel Dumontier, Luiz Olavo Bonino da Silva Santos, Mark D Wilkinson} \emph{Information Services \& Use} (2017-03-07)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{isu\}-170824}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
-\CSLLeftMargin{3. }
-\CSLRightInline{\textbf{EOSC interoperability framework}
-\CSLBlock{Oscar Corcho, Magnus Eriksson, Krzysztof Kurowski, Milan Ojsteršek, Christine Choirat, Mark van de Sanden, Frederik Coppens} \emph{Publications Office of the EU} (2021-02-05) \url{https://doi.org/10.2777/620649}
-\CSLBlock{DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
-\CSLLeftMargin{4. }
-\CSLRightInline{\textbf{FAIR principles and digital objects: Accelerating convergence on a data infrastructure}
-\CSLBlock{Erik Schultes, Peter Wittenburg} \emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers} (2019)
-\CSLBlock{DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}} · ISBN: 978-3-030-23583-3}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
-\CSLLeftMargin{5. }
-\CSLRightInline{\textbf{Linked data - the story so far}
-\CSLBlock{Christian Bizer, Tom Heath, Tim Berners-Lee} \emph{International journal on Semantic Web and information systems} (2009-07) \url{http://services.igi-global.com/resolvedoi/resolve.aspx?doi=10.4018/jswis.2009081901}
-\CSLBlock{DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
-\CSLLeftMargin{6. }
-\CSLRightInline{\textbf{Assessing FAIR data principles against the 5-star open data principles}
-\CSLBlock{Ali Hasnain, Dietrich Rebholz-Schuhmann} \emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers} (2018) \url{http://link.springer.com/10.1007/978-3-319-98192-5/_60}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60} · ISBN: 978-3-319-98191-8}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
-\CSLLeftMargin{7. }
-\CSLRightInline{\textbf{FAIR data points supporting big data interoperability}
-\CSLBlock{Luiz Olavo Bonino Da Silva Santos, Mark D Wilkinson, Arnold Kuzniar, Rajaram Kaliyaperumal, Mark Thompson, Michel Dumontier, Kees Burger} \emph{Enterprise interoperability in the digitized and networked factory of the future} (2016) \url{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}
-\CSLBlock{ISBN: 9781847040442}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
-\CSLLeftMargin{8. }
-\CSLRightInline{\textbf{The semantic web identity crisis: In search of the trivialities that never were}
-\CSLBlock{Ruben Verborgh, Miel Vander Sande} \emph{Satya Widya} (2020-01-31)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/sw-190372}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
-\CSLLeftMargin{9. }
-\CSLRightInline{\textbf{Conclusion and future challenges}
-\CSLBlock{Albert Meroño-Peñuela, Pasquale Lisena, Carlos Martínez-Ortiz} \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers} (2021)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7} · ISBN: 978-3-031-00789-7}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
-\CSLLeftMargin{10. }
-\CSLRightInline{\textbf{An analysis of public REST web service apis}
-\CSLBlock{Andy Neumann, Nuno Laranjeiro, Jorge Bernardino} \emph{IEEE Transactions on Services Computing} (2021-07-01)
-\CSLBlock{DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/tsc.2018.2847344}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
-\CSLLeftMargin{11. }
-\CSLRightInline{\textbf{Linked data, JSON-LD and the semantics of cultural and scientific heritage}
-\CSLBlock{Ovidiu Turcoane} \emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} (2014-09-30)
-\CSLBlock{DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pa67pUtR}{}}%
-\CSLLeftMargin{12. }
-\CSLRightInline{\textbf{F*** workflows: when parts of FAIR are missing}
-\CSLBlock{Sean R Wilkinson, Greg Eisenhauer, Anuj J Kapadia, Kathryn Knight, Jeremy Logan, Patrick Widener, Matthew Wolf} \emph{arXiv} (2022) \url{https://doi.org/grk3v5}
-\CSLBlock{DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
-\CSLLeftMargin{13. }
-\CSLRightInline{\textbf{Do Developers Make Unbiased Decisions? - The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components}
-\CSLBlock{Anisa Stefi} \emph{University of Münster, Münster, Germany} (2015) \url{https://doi.org/grcpwg}
-\CSLBlock{DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
-\CSLLeftMargin{14. }
-\CSLRightInline{\textbf{To develop or to reuse? Two perspectives on external reuse in software projects}
-\CSLBlock{Anisa Stefi, Thomas Hess} \emph{Software business} (2015)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18} · ISBN: 978-3-319-19592-6}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
-\CSLLeftMargin{15. }
-\CSLRightInline{\textbf{Web data apis over SPARQL}
-\CSLBlock{Albert Meroño-Peñuela, Pasquale Lisena, Carlos Martínez-Ortiz} \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers} (2021)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3} · ISBN: 978-3-031-00789-7}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
-\CSLLeftMargin{16. }
-\CSLRightInline{\textbf{FAIR Digital Objects Forum \textbar{}} \url{https://fairdo.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
-\CSLLeftMargin{17. }
-\CSLRightInline{\textbf{FAIR digital object overview and specifications}
-\CSLBlock{I Anders, Ch Blanchi, Daan Broeder, M Hellström, Sh Islam, Th Jejkal, L Lannom, K Peters, R Quick, A Schlemmer, \ldots{} Carlo Zwölf} \emph{FDO Forum} (2022-11-19) \url{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
-\CSLLeftMargin{18. }
-\CSLRightInline{\textbf{FDO Forum Document Standards}
-\CSLBlock{C Weiland, U Schwardmann, P Wittenburg, C Kirkpatrick, R Hanisch, Z Trautt} \emph{FDO Forum} (2022-01-29) \url{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
-\CSLLeftMargin{19. }
-\CSLRightInline{\textbf{FDO Forum FDO requirement specifications}
-\CSLBlock{G Strawn, P Wittenburg, St Soiland-Reyes, K Peters, L Lannom, U Schwardmann, Ch Blanch} \emph{FDO Forum} (2022-08-22) \url{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
-\CSLLeftMargin{20. }
-\CSLRightInline{\textbf{FAIR digital object framework}
-\CSLBlock{Luiz Bonino, Oeter Wittenburg, Bonnie Carroll, Alex Hardisty, Mark Leggott, Carlo Zwölf} \emph{FDOF technical implementation guideline} (2019-11-22) \url{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
-\CSLLeftMargin{21. }
-\CSLRightInline{\textbf{FDO machine actionability}
-\CSLBlock{Claus Weiland, Sharif Islam, Daan Broder, Ivonne Anders, Peter Wittenburg} \emph{FDO Forum} (2022-06-11) \url{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
-\CSLLeftMargin{22. }
-\CSLRightInline{\textbf{FDO configuration types}
-\CSLBlock{Larry Lannom, Karsten Peters-von Gehlen, Ivonne Anders, Andreas Pfeil, Alexander Schlemmer, Zach Trautt, Peter Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
-\CSLLeftMargin{23. }
-\CSLRightInline{\textbf{FAIR Signposting Profile}
-\CSLBlock{Herbert Van de Sompel, Martin Klein, Shawn Jones, Michael L Nelson, Simeon Warner, Anusuriya Devaraju, Robert Huber, Wilko Steinhoff, Vyacheslav Tykhonov, Luc Boruta, \ldots{} Mark Wilkinson} (2022-07-27) \url{https://signposting.org/FAIR/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
-\CSLLeftMargin{24. }
-\CSLRightInline{\textbf{FDO PID profiles \& attributes}
-\CSLBlock{Ivonne Anders, Maggie Hellström, Sharif Islam, Thomas Jejkal, Larry Lannom, Ulrich Schwardmann, Peter Wittenburg} \emph{FDO Forum} (2022-06-07) \url{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
-\CSLLeftMargin{25. }
-\CSLRightInline{\textbf{RDA Recommendation on PID Kernel Information}
-\CSLBlock{Tobias Weigel, Beth Plale, Mark Parsons, Gabriel Zhou, Yu Luo, Ulrich Schwardmann, Robert Quick, Margareta Hellström, Kei Kurakawa} \emph{Research Data Alliance} (2018)
-\CSLBlock{DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Fi62cJAo}{}}%
-\CSLLeftMargin{26. }
-\CSLRightInline{\textbf{FDO -- kernel attributes \& metadata}
-\CSLBlock{FDO-TSIG Working Group, FDO-BIG Working Group} \emph{FDO Forum} (2022-08-28) \url{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
-\CSLLeftMargin{27. }
-\CSLRightInline{\textbf{FDO -- granularity, versioning, mutability}
-\CSLBlock{FDO-TSIG Working Group} \emph{FDO Forum} (2022-08-26) \url{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
-\CSLLeftMargin{28. }
-\CSLRightInline{\textbf{DOIP endorsement request}
-\CSLBlock{FDO-TSIG Working Group} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
-\CSLLeftMargin{29. }
-\CSLRightInline{\textbf{Digital object interface protocol specification, version 2.0}
-\CSLBlock{DONA Foundation} \emph{DONA foundation} (2018-11-12) \url{https://hdl.handle.net/0.DOIP/DOIPV2.0}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
-\CSLLeftMargin{30. }
-\CSLRightInline{\textbf{FAIR digital object demonstrators 2021}
-\CSLBlock{Peter Wittenburg, Ivonne Anders, Christophe Blanchi, Merret Buurman, Carole Goble, Jonas Grieb, Alex Hardisty, Sharif Islam, Thomas Jejkal, Tibor Kálmán, \ldots{} Philipp Wieder} \emph{Zenodo} (2022)
-\CSLBlock{DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
-\CSLLeftMargin{31. }
-\CSLRightInline{\textbf{FDO -- upload of FDO}
-\CSLBlock{Christophe Blanchi, Daan Broeder, Thomas Jejkal, Islam Sharif, Alexander Schlemmer, Dieter van Uytvanck, Peter Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
-\CSLLeftMargin{32. }
-\CSLRightInline{\textbf{ResourceSync Framework Specification - Table of Contents} \url{http://www.openarchives.org/rs/toc}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
-\CSLLeftMargin{33. }
-\CSLRightInline{\textbf{Typing FAIR digital objects}
-\CSLBlock{Larry Lannom, U Schwardmann, C Blanchi, P Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
-\CSLLeftMargin{34. }
-\CSLRightInline{\textbf{Implementation of attributes, types, profiles and registries}
-\CSLBlock{C Blanchi, M Hellström, L Lannom, U Schwardmann, P Wittenburg} \emph{FDO Forum} (2022-11-27) \url{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
-\CSLLeftMargin{35. }
-\CSLRightInline{\textbf{Two Examples on How FDO Types can Support Machine and Human Readability}
-\CSLBlock{Ulrich Schwardmann, Tibor Kálmán} \emph{Research Ideas and Outcomes} (2022-10-12) \url{https://doi.org/grcq23}
-\CSLBlock{DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
-\CSLLeftMargin{36. }
-\CSLRightInline{\textbf{A framework for distributed digital object services}
-\CSLBlock{Robert Kahn, Robert Wilensky} \emph{International Journal on Digital Libraries} (2006-04) \url{https://www.doi.org/topics/2006_05_02_Kahn_Framework.pdf}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
-\CSLLeftMargin{37. }
-\CSLRightInline{\textbf{A framework for distributed digital object services}
-\CSLBlock{Robert Kahn, Robert Wilensky} \emph{CNRI} (1995-05-13) \url{http://www.cnri.reston.va.us/k-w.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
-\CSLLeftMargin{38. }
-\CSLRightInline{\textbf{X.1255 : Framework for discovery of identity management information} \url{https://www.itu.int/rec/T-REC-X.1255-201309-I}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
-\CSLLeftMargin{39. }
-\CSLRightInline{\textbf{Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation} \url{https://www.dona.net/doipv1doc}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
-\CSLLeftMargin{40. }
-\CSLRightInline{\textbf{Digital objects as drivers towards convergence in data infrastructures}
-\CSLBlock{Peter Wittenburg, George Strawn, Barend Mons, Luiz Bonino, Erik Schultes} \emph{https://b2share.eudat.eu} (2019-01-06)
-\CSLBlock{DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
-\CSLLeftMargin{41. }
-\CSLRightInline{\textbf{DOIP and Examples --- Cordra documentation} \url{https://www.cordra.org/documentation/api/doip.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
-\CSLLeftMargin{42. }
-\CSLRightInline{\textbf{draft-bhutton-json-schema-00} \url{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
-\CSLLeftMargin{43. }
-\CSLRightInline{\textbf{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}
-\CSLBlock{Tim Berners-Lee, Mark Fischetti} \emph{HarperSanFrancisco} (1999)
-\CSLBlock{ISBN: 9780062515865}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
-\CSLLeftMargin{44. }
-\CSLRightInline{\textbf{RDF 1.1 Primer} \url{http://www.w3.org/TR/rdf11-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
-\CSLLeftMargin{45. }
-\CSLRightInline{\textbf{Resource Description Framework (RDF) Model and Syntax Specification} \url{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
-\CSLLeftMargin{46. }
-\CSLRightInline{\textbf{Process modelling for information system description}
-\CSLBlock{Stefan K Stanczyk} \emph{The Open University} (1987)
-\CSLBlock{DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
-\CSLLeftMargin{47. }
-\CSLRightInline{\textbf{Uniform Resource Identifier (URI): Generic Syntax}
-\CSLBlock{T Berners-Lee, R Fielding, L Masinter} \emph{RFC Editor} (2005-01) \url{https://doi.org/ggqvpr}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-\CSLLeftMargin{48. }
-\CSLRightInline{\textbf{"info" URI Registry (Frozen)} \url{https://oclc-research.github.io/infoURI-Frozen/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
-\CSLLeftMargin{49. }
-\CSLRightInline{\textbf{Identifiers.org and MIRIAM Registry: community resources to provide persistent identification}
-\CSLBlock{N Juty, N Le Novere, C Laibe} \emph{Nucleic Acids Research} (2011-12-02) \url{https://doi.org/cx2776}
-\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22140103}{22140103} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3245029}{PMC3245029}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
-\CSLLeftMargin{50. }
-\CSLRightInline{\textbf{Internationalized Resource Identifiers (IRIs)}
-\CSLBlock{M Duerst, M Suignard} \emph{RFC Editor} (2005-01) \url{https://doi.org/gjvnbg}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
-\CSLLeftMargin{51. }
-\CSLRightInline{\textbf{Cool URIs for the semantic web}
-\CSLBlock{Leo Sauermann, Richard Cyganiak, Max Völkel} \emph{Universität des Saarlandes} (2011-07-13)
-\CSLBlock{DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
-\CSLLeftMargin{52. }
-\CSLRightInline{\textbf{The Semantic Web --- ISWC 2002}
-\CSLBlock{Ian Horrocks, James Hendler (editors)} \emph{Lecture Notes in Computer Science} (2002) \url{https://doi.org/fwjmhp}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
-\CSLLeftMargin{53. }
-\CSLRightInline{\textbf{SPARQL 1.1 Overview} \url{http://www.w3.org/TR/sparql11-overview/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
-\CSLLeftMargin{54. }
-\CSLRightInline{\textbf{State of the nation in data integration for bioinformatics.}
-\CSLBlock{Carole Goble, Robert Stevens} \emph{Journal of Biomedical Informatics} (2008-10-01)
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/18358788}{18358788}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
-\CSLLeftMargin{55. }
-\CSLRightInline{\textbf{Open PHACTS: semantic interoperability for drug discovery}
-\CSLBlock{Antony J Williams, Lee Harland, Paul Groth, Stephen Pettifer, Christine Chichester, Egon L Willighagen, Chris T Evelo, Niklas Blomberg, Gerhard Ecker, Carole Goble, Barend Mons} \emph{Drug Discovery Today} (2012-11) \url{https://doi.org/f4fdkd}
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22683805}{22683805}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
-\CSLLeftMargin{56. }
-\CSLRightInline{\textbf{SKOS Simple Knowledge Organization System Primer} \url{http://www.w3.org/TR/skos-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
-\CSLLeftMargin{57. }
-\CSLRightInline{\textbf{How matchable are four thousand ontologies on the semantic web}
-\CSLBlock{Wei Hu, Jianfeng Chen, Hang Zhang, Yuzhong Qu} \emph{The semantic web: Research and applications} (2011)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20} · ISBN: 978-3-642-21033-4}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
-\CSLLeftMargin{58. }
-\CSLRightInline{\textbf{API-centric Linked Data integration: The Open PHACTS Discovery Platform case study}
-\CSLBlock{Paul Groth, Antonis Loizou, Alasdair JG Gray, Carole Goble, Lee Harland, Steve Pettifer} \emph{Journal of Web Semantics} (2014-12)
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
-\CSLLeftMargin{59. }
-\CSLRightInline{\textbf{A systematic analysis of term reuse and term overlap across biomedical ontologies}
-\CSLBlock{Maulik R Kamdar, Tania Tudorache, Mark A Musen} \emph{Semantic Web} (2017-08-07)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/28819351}{28819351} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5555235}{PMC5555235}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
-\CSLLeftMargin{60. }
-\CSLRightInline{\textbf{Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot}
-\CSLBlock{Martin Klein, Herbert Van de Sompel, Robert Sanderson, Harihar Shankar, Lyudmila Balakireva, Ke Zhou, Richard Tobin} \emph{PLoS ONE} (2014-12-26)
-\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/25541969}{25541969} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4277367}{PMC4277367}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
-\CSLLeftMargin{61. }
-\CSLRightInline{\textbf{A more decentralized vision for linked data}
-\CSLBlock{Axel Polleres, Maulik Rajendra Kamdar, Javier David Fernández, Tania Tudorache, Mark Alan Musen} \emph{Satya Widya} (2020-01-31)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/sw-190380}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
-\CSLLeftMargin{62. }
-\CSLRightInline{\textbf{The landscape of ontology reuse approaches}
-\CSLBlock{Valentina Anita Carriero, Marilena Daquino, Aldo Gangemi, Andrea Giovanni Nuzzolese, Silvio Peroni, Valentina Presutti, Francesca Tomasi} \emph{Applications and practices in ontology design, extraction, and reasoning} (2020-11-12)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/ssw200033} · ISBN: 9781643681429}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
-\CSLLeftMargin{63. }
-\CSLRightInline{\textbf{RDF Schema 1.1} \url{http://www.w3.org/TR/rdf-schema/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
-\CSLLeftMargin{64. }
-\CSLRightInline{\textbf{OWL 2 Web Ontology Language Document Overview (Second Edition)} \url{http://www.w3.org/TR/owl2-overview/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
-\CSLLeftMargin{65. }
-\CSLRightInline{\textbf{Mapping between the OBO and OWL ontology languages}
-\CSLBlock{Syed Tirmizi, Stuart Aitken, Dilvan A Moreira, Chris Mungall, Juan Sequeda, Nigam H Shah, Daniel P Miranker} \emph{Journal of Biomedical Semantics} (2011)
-\CSLBlock{DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/21388572}{21388572} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3105495}{PMC3105495}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
-\CSLLeftMargin{66. }
-\CSLRightInline{\textbf{Linked Data - Design Issues} \url{https://www.w3.org/DesignIssues/LinkedData.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
-\CSLLeftMargin{67. }
-\CSLRightInline{\textbf{The Open Graph protocol} \url{https://ogp.me/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
-\CSLLeftMargin{68. }
-\CSLRightInline{\textbf{RDFa 1.1 Primer - Third Edition} \url{http://www.w3.org/TR/rdfa-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
-\CSLLeftMargin{69. }
-\CSLRightInline{\textbf{HTML Standard} \url{https://html.spec.whatwg.org/multipage/microdata.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
-\CSLLeftMargin{70. }
-\CSLRightInline{\textbf{JSON-LD 1.1} \url{https://www.w3.org/TR/json-ld/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
-\CSLLeftMargin{71. }
-\CSLRightInline{\textbf{Usage Statistics of JSON-LD for Websites, May 2022} \url{https://w3techs.com/technologies/details/da-jsonld}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
-\CSLLeftMargin{72. }
-\CSLRightInline{\textbf{Designing a Linked Data developer experience} (2018-12-28) \url{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
-\CSLLeftMargin{73. }
-\CSLRightInline{\textbf{Shapes Constraint Language (SHACL)} \url{https://www.w3.org/TR/shacl/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
-\CSLLeftMargin{74. }
-\CSLRightInline{\textbf{Shape Expressions (ShEx) 2.1 Primer} \url{http://shex.io/shex-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
-\CSLLeftMargin{75. }
-\CSLRightInline{\textbf{Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation}
-\CSLBlock{Katherine Thornton, Harold Solbrig, Gregory S Stupp, Jose Emilio Labra Gayo, Daniel Mietchen, Eric Prud, Andra Waagmeester} \emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings} (2019)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39} · ISBN: 978-3-030-21347-3}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
-\CSLLeftMargin{76. }
-\CSLRightInline{\textbf{Validating RDF Data}
-\CSLBlock{Jose Emilio Labra Gayo, Eric Prud'hommeaux, Iovka Boneva, Dimitris Kontokostas} \emph{Synthesis Lectures on the Semantic Web: Theory and Technology} (2017-09-28)
-\CSLBlock{DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
-\CSLLeftMargin{77. }
-\CSLRightInline{\textbf{Survey of tools for linked data consumption}
-\CSLBlock{Jakub Klímek, Petr Škoda, Martin Nečaský} \emph{Satya Widya} (2019-05-23)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/sw-180316}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
-\CSLLeftMargin{78. }
-\CSLRightInline{\textbf{RightField: embedding ontology annotation in spreadsheets}
-\CSLBlock{K Wolstencroft, S Owen, M Horridge, O Krebs, W Mueller, JL Snoep, F du Preez, C Goble} \emph{Bioinformatics} (2011-05-26)
-\CSLBlock{DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/21622664}{21622664}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
-\CSLLeftMargin{79. }
-\CSLRightInline{\textbf{REST and Linked Data}
-\CSLBlock{Kevin R Page, David C De Roure, Kirk Martinez} \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11} (2011) \url{http://eprints.soton.ac.uk/id/eprint/272098}
-\CSLBlock{DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
-\CSLLeftMargin{80. }
-\CSLRightInline{\textbf{On Schema.org and Why It Matters for the Web}
-\CSLBlock{Peter Mika} \emph{IEEE Internet Computing} (2015-07)
-\CSLBlock{DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
-\CSLLeftMargin{81. }
-\CSLRightInline{\textbf{A new look at the semantic web}
-\CSLBlock{Abraham Bernstein, James Hendler, Natalya Noy} \emph{Communications of the ACM} (2016-08-24)
-\CSLBlock{DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
-\CSLLeftMargin{82. }
-\CSLRightInline{\textbf{An Interoperability Framework and Distributed Platform for Fast Data Applications}
-\CSLBlock{José Carlos Martins Delgado} \emph{Data Science and Big Data Computing} (2016)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRzcHhPD}{}}%
-\CSLLeftMargin{83. }
-\CSLRightInline{\textbf{A Comparison Framework for Middleware Infrastructures.}
-\CSLBlock{Apostolos Zarras} \emph{The Journal of Object Technology} (2004)
-\CSLBlock{DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
-\CSLLeftMargin{84. }
-\CSLRightInline{\textbf{The FAIR data maturity model: An approach to harmonise FAIR assessments}
-\CSLBlock{Christophe Bahim, Carlos Casorrán-Amilburu, Makx Dekkers, Edit Herczog, Nicolas Loozen, Konstantinos Repanas, Keith Russell, Shelley Stall} \emph{Data Science Journal} (2020-10-27)
-\CSLBlock{DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
-\CSLLeftMargin{85. }
-\CSLRightInline{\textbf{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}
-\CSLBlock{William Stallings} \emph{Sams} (1990)
-\CSLBlock{ISBN: 9780672226977}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
-\CSLLeftMargin{86. }
-\CSLRightInline{\textbf{Architectural styles and the design of network-based software architectures}
-\CSLBlock{Roy Thomas Fielding} \emph{University of California, Irvine} (2000) \url{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
-\CSLLeftMargin{87. }
-\CSLRightInline{\textbf{Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award)}
-\CSLBlock{Roy T Fielding, Richard N Taylor, Justin R Erenkrantz, Michael M Gorlick, Jim Whitehead, Rohit Khare, Peyman Oreizy} \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017} (2017-09-04) \url{http://dl.acm.org/citation.cfm?doid=3106237.3121282}
-\CSLBlock{DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282} · ISBN: 9781450351058}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
-\CSLLeftMargin{88. }
-\CSLRightInline{\textbf{OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More} \url{https://spec.openapis.org/oas/v3.1.0.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
-\CSLLeftMargin{89. }
-\CSLRightInline{\textbf{Schema.org Actions - schema.org} \url{https://schema.org/docs/actions.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
-\CSLLeftMargin{90. }
-\CSLRightInline{\textbf{Web Services Description Language (WSDL) Version 2.0 Part 0: Primer} \url{http://www.w3.org/TR/wsdl20-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
-\CSLLeftMargin{91. }
-\CSLRightInline{\textbf{The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud}
-\CSLBlock{Katherine Wolstencroft, Robert Haines, Donal Fellows, Alan Williams, David Withers, Stuart Owen, Stian Soiland-Reyes, Ian Dunlop, Aleksandra Nenadic, Paul Fisher, \ldots{} Carole Goble} \emph{Nucleic Acids Research} (2013-05-02) \url{https://doi.org/ggbwf4}
-\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/23640334}{23640334} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3692062}{PMC3692062}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
-\CSLLeftMargin{92. }
-\CSLRightInline{\textbf{Perspectives on automated composition of workflows in the life sciences}
-\CSLBlock{Anna-Lena Lamprecht, Magnus Palmblad, Jon Ison, Veit Schwämmle, Mohammad Sadnan Al Manir, Ilkay Altintas, Christopher JO Baker, Ammar Ben Hadj Amor, Salvador Capella-Gutierrez, Paulos Charonyktakis, \ldots{} Katherine Wolstencroft} \emph{F1000Research} (2021-09-07) \url{https://doi.org/gqprp7}
-\CSLBlock{DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/34804501}{34804501} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8573700}{PMC8573700}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-\CSLLeftMargin{93. }
-\CSLRightInline{\textbf{The Modern Standards Paradigm - Five Key Principles}
-\CSLBlock{OpenStand} \url{https://open-stand.org/about-us/principles/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
-\CSLLeftMargin{94. }
-\CSLRightInline{\textbf{First International Conference on FAIR Digital Objects}
-\CSLBlock{Tina Loo (editor)} \emph{Pensoft Publishers} (1970) \url{https://doi.org/grcqxh}
-\CSLBlock{DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
-\CSLLeftMargin{95. }
-\CSLRightInline{\textbf{DOIP API for HTTP Clients --- Cordra documentation} \url{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-\CSLLeftMargin{96. }
-\CSLRightInline{\textbf{DOI Resolution Documentation} \url{https://www.doi.org/factsheets/DOIProxy.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-\CSLLeftMargin{97. }
-\CSLRightInline{\textbf{DOI Handbook - Resolution} \url{https://www.doi.org/doi_handbook/3_Resolution.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
-\CSLLeftMargin{98. }
-\CSLRightInline{\textbf{HTTP Semantics}
-\CSLBlock{R Fielding, M Nottingham, J Reschke (editors)} \emph{RFC Editor} (2022-06) \url{https://doi.org/gqprqb}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
-\CSLLeftMargin{99. }
-\CSLRightInline{\textbf{Linked data platform 1.0}
-\CSLBlock{Linked Data Platform Working Group} \emph{W3C} (2015-02-26) \url{https://www.w3.org/TR/2015/REC-ldp-20150226/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
-\CSLLeftMargin{100. }
-\CSLRightInline{\textbf{FAIR Digital Object Framework Documentation} \url{https://fairdigitalobjectframework.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
-\CSLLeftMargin{101. }
-\CSLRightInline{\textbf{Handle System Overview}
-\CSLBlock{S Sun, L Lannom, B Boesch} \emph{RFC Editor} (2003-11) \url{https://doi.org/ggn83z}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
-\CSLLeftMargin{102. }
-\CSLRightInline{\textbf{Web API design best practices - Azure Architecture Center}
-\CSLBlock{martinekuan} \url{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
-\CSLLeftMargin{103. }
-\CSLRightInline{\textbf{The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration}
-\CSLBlock{Barry Smith, Michael Ashburner, Cornelius Rosse, Jonathan Bard, William Bug, Werner Ceusters, Louis J Goldberg, Karen Eilbeck, Amelia Ireland, \ldots{} Suzanna Lewis} \emph{Nature Biotechnology} (2007-11) \url{https://doi.org/bqng99}
-\CSLBlock{DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/17989687}{17989687} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2814061}{PMC2814061}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
-\CSLLeftMargin{104. }
-\CSLRightInline{\textbf{Data Information View} \url{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
-\CSLLeftMargin{105. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 3} \url{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-\CSLLeftMargin{106. }
-\CSLRightInline{\textbf{ORE User Guide - Primer} \url{http://www.openarchives.org/ore/1.0/primer}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-\CSLLeftMargin{107. }
-\CSLRightInline{\textbf{PROV-O: The PROV Ontology} \url{https://www.w3.org/TR/2013/REC-prov-o-20130430/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
-\CSLLeftMargin{108. }
-\CSLRightInline{\textbf{Semantic Web Services}
-\CSLBlock{Dieter Fensel, Federico Michele Facca, Elena Simperl, Ioan Toma} \emph{Springer Berlin Heidelberg} (2011) \url{https://doi.org/bv7nnc}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-\CSLLeftMargin{109. }
-\CSLRightInline{\textbf{5-star Open Data} \url{http://5stardata.info/en/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
-\CSLLeftMargin{110. }
-\CSLRightInline{\url{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
-\CSLLeftMargin{111. }
-\CSLRightInline{\textbf{Global Internet Phenomena Report 2022}
-\CSLBlock{Sandvine} \url{https://www.sandvine.com/global-internet-phenomena-report-2022}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
-\CSLLeftMargin{112. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol Version 2 (HTTP/2)}
-\CSLBlock{M Belshe, R Peon} \emph{RFC Editor} (2015-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
-\CSLLeftMargin{113. }
-\CSLRightInline{\url{https://blog.cloudflare.com/http-3-vs-http-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
-\CSLLeftMargin{114. }
-\CSLRightInline{\textbf{draft-ietf-quic-http-34} \url{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
-\CSLLeftMargin{115. }
-\CSLRightInline{\textbf{QUIC: A UDP-Based Multiplexed and Secure Transport}
-\CSLBlock{J Iyengar, M Thomson (editors)} \emph{RFC Editor} (2021-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
-\CSLLeftMargin{116. }
-\CSLRightInline{\textbf{URI Template}
-\CSLBlock{J Gregorio, R Fielding, M Hadley, M Nottingham, D Orchard} \emph{RFC Editor} (2012-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
-\CSLLeftMargin{117. }
-\CSLRightInline{\textbf{Content negotiation - HTTP \textbar{} MDN} \url{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
-\CSLLeftMargin{118. }
-\CSLRightInline{\textbf{Hypertext Style: Cool URIs don't change.} \url{https://www.w3.org/Provider/Style/URI}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
-\CSLLeftMargin{119. }
-\CSLRightInline{\textbf{HTTP Framework for Time-Based Access to Resource States -\/- Memento}
-\CSLBlock{H Van de Sompel, M Nelson, R Sanderson} \emph{RFC Editor} (2013-12)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
-\CSLLeftMargin{120. }
-\CSLRightInline{\textbf{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}
-\CSLBlock{G Clemm, J Amsden, T Ellison, C Kaler, J Whitehead} \emph{RFC Editor} (2002-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
-\CSLLeftMargin{121. }
-\CSLRightInline{\textbf{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}
-\CSLBlock{L Dusseault (editor)} \emph{RFC Editor} (2007-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
-\CSLLeftMargin{122. }
-\CSLRightInline{\textbf{Upgrading to TLS Within HTTP/1.1}
-\CSLBlock{R Khare, S Lawrence} \emph{RFC Editor} (2000-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
-\CSLLeftMargin{123. }
-\CSLRightInline{\textbf{Hypertext Style: Cool URIs don't change.} \url{https://www.w3.org/Provider/Style/URI.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
-\CSLLeftMargin{124. }
-\CSLRightInline{\textbf{draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes} \url{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
-\CSLLeftMargin{125. }
-\CSLRightInline{\textbf{The 'profile' Link Relation Type}
-\CSLBlock{E Wilde} \emph{RFC Editor} (2013-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
-\CSLLeftMargin{126. }
-\CSLRightInline{\textbf{W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures} \url{http://www.w3.org/TR/xmlschema11-1/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
-\CSLLeftMargin{127. }
-\CSLRightInline{\textbf{JSON-LD 1.1} \url{http://www.w3.org/TR/json-ld/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
-\CSLLeftMargin{128. }
-\CSLRightInline{\textbf{WebSockets Standard} \url{https://websockets.spec.whatwg.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
-\CSLLeftMargin{129. }
-\CSLRightInline{\textbf{Linked Data Notifications} \url{https://www.w3.org/TR/ldn/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
-\CSLLeftMargin{130. }
-\CSLRightInline{\textbf{The Atom Publishing Protocol}
-\CSLBlock{J Gregorio, B de hOra (editors)} \emph{RFC Editor} (2007-10)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
-\CSLLeftMargin{131. }
-\CSLRightInline{\textbf{SWORD 3.0 Specification} \url{https://swordapp.github.io/swordv3/swordv3.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
-\CSLLeftMargin{132. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}
-\CSLBlock{R Fielding, J Reschke (editors)} \emph{RFC Editor} (2014-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-\CSLLeftMargin{133. }
-\CSLRightInline{\textbf{Hydra W3C Community Group} \url{https://www.hydra-cg.com/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
-\CSLLeftMargin{134. }
-\CSLRightInline{\textbf{draft-kelly-json-hal-08} \url{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
-\CSLLeftMargin{135. }
-\CSLRightInline{\textbf{Web Linking}
-\CSLBlock{M Nottingham} \emph{RFC Editor} (2017-10)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
-\CSLLeftMargin{136. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}
-\CSLBlock{R Fielding, J Reschke (editors)} \emph{RFC Editor} (2014-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
-\CSLLeftMargin{137. }
-\CSLRightInline{\textbf{HTTP Live Streaming}
-\CSLBlock{W May} \emph{RFC Editor} (2017-08)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
-\CSLLeftMargin{138. }
-\CSLRightInline{\textbf{ISO/IEC 23009-1:2019}
-\CSLBlock{14:00-17:00} \emph{ISO} \url{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
-\CSLLeftMargin{139. }
-\CSLRightInline{\textbf{FAIR data maturity model: Specification and guidelines}
-\CSLBlock{Research Data Alliance FAIR Data Maturity Model Working Group} \emph{Research Data Alliance} (2020) \url{https://zenodo.org/record/3909563/\#.\%7BYGRNnq8za70\%7D}
-\CSLBlock{DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
-\CSLLeftMargin{140. }
-\CSLRightInline{\textbf{NCBO BioPortal} \url{https://bioportal.bioontology.org/ontologies}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
-\CSLLeftMargin{141. }
-\CSLRightInline{\textbf{The Globus Striped GridFTP Framework and Server}
-\CSLBlock{W Allcock, J Bresnahan, R Kettimuthu, M Link} \emph{ACM/IEEE SC 2005 Conference (SC'05)} \url{https://doi.org/cgmc2b}
-\CSLBlock{DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-\CSLLeftMargin{142. }
-\CSLRightInline{\textbf{The ARK Identifier Scheme} \url{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
-\CSLLeftMargin{143. }
-\CSLRightInline{\textbf{Handle System Protocol (ver 2.1) Specification}
-\CSLBlock{S Sun, S Reilly, L Lannom, J Petrone} \emph{RFC Editor} (2003-11)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-\CSLLeftMargin{144. }
-\CSLRightInline{\textbf{Handle.Net Registry} \url{https://www.handle.net/download_hnr.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-\CSLLeftMargin{145. }
-\CSLRightInline{\textbf{Licenses \& Standards \textbar{} Open Source Initiative} \url{https://opensource.org/licenses}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-\CSLLeftMargin{146. }
-\CSLRightInline{\textbf{ISO 16684-1:2019}
-\CSLBlock{14:00-17:00} \emph{ISO} \url{https://www.iso.org/standard/75163.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
-\CSLLeftMargin{147. }
-\CSLRightInline{\textbf{Data - W3C} \url{https://www.w3.org/standards/semanticweb/data}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-\CSLLeftMargin{148. }
-\CSLRightInline{\textbf{Linking Across Provenance Bundles} \url{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-\CSLLeftMargin{149. }
-\CSLRightInline{\textbf{Introducing 'Role'}
-\CSLBlock{Unknown} \url{http://blog.schema.org/2014/06/introducing-role.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-\CSLLeftMargin{150. }
-\CSLRightInline{\textbf{PROV-O: The PROV Ontology} \url{https://www.w3.org/TR/prov-o/\#specializationOf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
-\CSLLeftMargin{151. }
-\CSLRightInline{\textbf{ORE Specification - Abstract Data Model} \url{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
-\CSLLeftMargin{152. }
-\CSLRightInline{\textbf{Schema.org - Schema.org} \url{https://schema.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
-\CSLLeftMargin{153. }
-\CSLRightInline{\textbf{DCMI Metadata Terms} \url{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
-\CSLLeftMargin{154. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 2} \url{https://www.w3.org/TR/vocab-dcat-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
-\CSLLeftMargin{155. }
-\CSLRightInline{\textbf{FOAF Vocabulary Specification} \url{http://xmlns.com/foaf/spec/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
-\CSLLeftMargin{156. }
-\CSLRightInline{\textbf{PAV ontology: provenance, authoring and versioning}
-\CSLBlock{Paolo Ciccarese, Stian Soiland-Reyes, Khalid Belhajjame, Alasdair JG Gray, Carole Goble, Tim Clark} \emph{Journal of Biomedical Semantics} (2013) \url{https://doi.org/gftcpx}
-\CSLBlock{DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/24267948}{24267948} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4177195}{PMC4177195}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
-\CSLLeftMargin{157. }
-\CSLRightInline{\textbf{DCMI Metadata Terms} (2020-01-20) \url{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
-\CSLLeftMargin{158. }
-\CSLRightInline{\textbf{The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
-Digital Object Approach to Speeding up Digital Mobilisation of Natural History
-Collections}
-\CSLBlock{Alex Hardisty, Paul Brack, Carole Goble, Laurence Livermore, Ben Scott, Quentin Groom, Stuart Owen, Stian Soiland-Reyes} \emph{Data Intelligence} (2022) \url{https://doi.org/grk33x}
-\CSLBlock{DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
-\CSLLeftMargin{159. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 2} \url{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
-\CSLLeftMargin{160. }
-\CSLRightInline{\textbf{Bioschemas - Bioschemas} \url{http://bioschemas.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
-\CSLLeftMargin{161. }
-\CSLRightInline{\textbf{Darwin Core: An Evolving Community-Developed Biodiversity Data Standard}
-\CSLBlock{John Wieczorek, David Bloom, Robert Guralnick, Stan Blum, Markus Döring, Renato Giovanni, Tim Robertson, David Vieglais} \emph{PLoS ONE} (2012-01-06) \url{https://doi.org/fzrpwq}
-\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22238640}{22238640} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3253084}{PMC3253084}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
-\CSLLeftMargin{162. }
-\CSLRightInline{\textbf{Semantic Web - XML2000 - slide "Architecture"} \url{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}}
-
-\end{CSLReferences}

### 2023-01-24 latex/table3.tex
converted to peerj overleaf template

--- 2023-01-24 manuscript.tex
@@
-This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
-was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
-on January 24, 2023.
-
-\hypertarget{authors}{%
-\section*{Authors}\label{authors}}
-\addcontentsline{toc}{section}{Authors}
-
-\begin{itemize}
-\item
-  \textbf{Stian Soiland-Reyes}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/stain}{stain}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/soilandreyes}{soilandreyes}
-  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Carole Goble}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/carolegoble}{carolegoble}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
-  Department of Computer Science, The University of Manchester, UK
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Paul Groth}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/pgroth}{pgroth}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/pgroth}{pgroth}
-  Informatics Institute, Faculty of Science, University of Amsterdam, NL
-\end{itemize}
-
-\hypertarget{abstract}{%
-\section*{Abstract}\label{abstract}}
-\addcontentsline{toc}{section}{Abstract}
-
-\textbf{TODO}: Rewrite from notes below to actual abstract.
-
-FAIR Digital Object is an emerging concept from EOSC.
-This is important.
-Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
-Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
-
-What is the message of this paper?
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These are the overlaps
-\item
-  This is what FDO is requiring but not commonly deployed
-\item
-  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
-\item
-  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
-\item
-  Lessons for Semantic Web community. What is missing in FDO and in SW.
-\item
-  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
-\item
-  Go back to the old SW layer cake. Explains why we picked these frameworks.
-\item
-  Lessons for SW: Parts of the stack that is less. FDO contributions.
-\end{enumerate}
-
-Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.
-
-Have all the ingredients, but not cooking properly.
-
-Systematic through frameworks.
-
-Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?
-
-\hypertarget{sec:introduction}{%
-\section{Introduction}\label{sec:introduction}}
-
-The FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}) encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures (\protect\hyperlink{ref-u8Cy0psL}{Mons et al., 2017}).
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
-
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
-The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
-
-In other discourse, Linked Data (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\protect\hyperlink{ref-EJDjHrUv}{Bonino Da Silva Santos et al., 2016}; \protect\hyperlink{ref-GJHYN6xW}{Hasnain \& Rebholz-Schuhmann, 2018}).
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).
-
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\protect\hyperlink{ref-uxWjmlio}{Stefi, 2015}; \protect\hyperlink{ref-v0WwFHgT}{Stefi \& Hess, 2015}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves (\protect\hyperlink{ref-5V3GqSzR}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021b}).
-
-These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.
-
-Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.
-
-\hypertarget{sec:background}{%
-\section{Background and related work}\label{sec:background}}
-
-In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.
-
-\hypertarget{sec:fdo}{%
-\subsection{FAIR Digital Object}\label{sec:fdo}}
-
-\hypertarget{sec:next-step-fdo}{%
-\subsubsection{Next steps for FDO}\label{sec:next-step-fdo}}
-
-The FAIR Digital Object Forum (\protect\hyperlink{ref-DMEBhHID}{{`FAIR Digital Objects Forum \textbar{}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
-As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
-
-\textbf{FAIR Digital Object Overview and Specifications} (\protect\hyperlink{ref-11NpETLq1}{Anders et al., 2022a}) is a comprehensive overview of FAIR Digital Object specifications listed below.
-It serves as a primer that introduces FDO concepts and the remaining documents.
-
-The \textbf{FDO Forum Document Standards} (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
-In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
-
-The \textbf{FDO Requirement Specifications} (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) is an update of (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) as the foundational definition of FDO.
-This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
-The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
-
-The \textbf{Machine actionability} (\protect\hyperlink{ref-iJeys0T5}{Weiland et al., 2022a}) sets out to define what is meant by \emph{machine actionability} for FDOs.
-\emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
-The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.
-
-\textbf{Configuration Types} (\protect\hyperlink{ref-SPd0uEqO}{Lannom et al., 2022a}) classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022})
-
-\textbf{PID Profiles \& Attributes} (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}) specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
-
-\textbf{Kernel Attributes \& Metadata} (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
-It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
-
-\textbf{Granularity, Versioning, Mutability} (\protect\hyperlink{ref-16W9AycSW}{Working Group, 2022b}) considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
-The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
-
-\textbf{DOIP Endorsement Request} (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) is an endorsement of the DOIP v2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) specification as a potential FDO implementation, as it has been applied by several institutions (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}).
-The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
-
-\textbf{Upload of FDO} (\protect\hyperlink{ref-1H7cmmr69}{Blanchi et al., 2022a}) illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync (\protect\hyperlink{ref-eS5y9TRh}{{`ResourceSync Framework Specification - Table of Contents'}}) is suggested as one type of service to list FDOs.
-This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
-
-\textbf{Typing FAIR Digital Objects} (\protect\hyperlink{ref-ZFzPxCNB}{Lannom et al., 2022b}) defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
-This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
-Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work.
-How to define such FDO operations is not specified.
-
-\textbf{Implementation of Attributes, Types, Profiles and Registries} (\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
-This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} (\protect\hyperlink{ref-1CqIZ47pu}{Schwardmann \& Kálmán, 2022}).
-
-It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-
-The concept of \textbf{FAIR Digital Objects} (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) has been introduced as way to expose research data as active objects that conform to the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-This builds on the \emph{Digital Object} (DO) concept (\protect\hyperlink{ref-11MnuwJ4l}{Kahn \& Wilensky, 2006}), first introduced in 1995 (\protect\hyperlink{ref-3Uqe3fuK}{Kahn \& Wilensky, 1995}) as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework (\protect\hyperlink{ref-103Hw8H43}{{`X.1255 : Framework for discovery of identity management information'}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}).
-
-In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}.
-The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing (\protect\hyperlink{ref-tz0P3DTC}{Wittenburg et al., 2019}).
-
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}), in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}).
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
-
-\hypertarget{fdo-approaches}{%
-\subsubsection{FDO approaches}\label{fdo-approaches}}
-
-FDO is an evolving concept.
-A set of FDO Demonstrators (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) highlight how current adapters are approaching implementations of FDO from different angles:
-
-\begin{itemize}
-\tightlist
-\item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), see (\protect\hyperlink{ref-xuQKRRx5}{{`DOIP and Examples --- Cordra documentation'}}).} (usually TCP/IP over TLS).
-The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}}) and PIDs are assigned using the Handle system.
-Several type registries have been established.
-\item
-  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
-\item
-  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
-\end{itemize}
-
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
-
-\hypertarget{sec:ld}{%
-\subsection{From the Semantic Web to Linked Data}\label{sec:ld}}
-
-In order to describe \emph{Linked Data} as it is used today, we'll start with an (opinionated) description of the evolution of its foundation, the \emph{Semantic Web}.
-
-\hypertarget{sec:semweb}{%
-\subsubsection{A brief history of the Semantic Web}\label{sec:semweb}}
-
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee (\protect\hyperlink{ref-Guy7cjVY}{Berners-Lee \& Fischetti, 1999}), at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
-The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
-The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.
-
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\protect\hyperlink{ref-16WGmCih8}{{`RDF 1.1 Primer'}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\protect\hyperlink{ref-16wvFH4nh}{{`Resource Description Framework (RDF) Model and Syntax Specification'}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
-
-While triple-based knowledge representations were not new (\protect\hyperlink{ref-NxvhwSmG}{Stanczyk, 1987}), the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs (\protect\hyperlink{ref-RRtwPF7Y}{Berners-Lee, Fielding \& Masinter, 2005}) are generalised forms of URLs that include locator-less identifiers
-  such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years (\protect\hyperlink{ref-nW4HY8Nq}{{`"info" URI Registry (Frozen)'}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) (\protect\hyperlink{ref-HqZ3r6J3}{Juty, Le Novere \& Laibe, 2011}).
-RDF 1.1 formats use Unicode to support IRIs (\protect\hyperlink{ref-rqjuXkEf}{Duerst \& Suignard, 2005}), which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
-
-The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data).
-Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term.
-The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource.
-By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs.
-The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.
-
-The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} (\protect\hyperlink{ref-yd2WE9eq}{Horrocks \& Hendler, 2002}) gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
-
-Established Semantic Web services and standards include SPARQL (\protect\hyperlink{ref-xww8fb1B}{{`SPARQL 1.1 Overview'}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
-
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\protect\hyperlink{ref-bMSxa8XW}{Goble \& Stevens, 2008}; \protect\hyperlink{ref-Qlrlr2TZ}{Williams et al., 2012}).
-In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
-
-Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) (\protect\hyperlink{ref-FDDmgO6s}{Hu et al., 2011}) .
-
-The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS (\protect\hyperlink{ref-191ZkYZMt}{Groth et al., 2014})), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
-
-With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps (\protect\hyperlink{ref-9CdJnsoJ}{Kamdar, Tudorache \& Musen, 2017}).
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication (\protect\hyperlink{ref-AGHR9oyr}{Klein et al., 2014}), for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data (\protect\hyperlink{ref-5SuwLNdE}{Polleres et al., 2020, p.
-handle:20.500.11811/7183}).
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies (\protect\hyperlink{ref-TUmA4KQT}{Carriero et al., 2020}) , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), OBO (\protect\hyperlink{ref-1DalZuKct}{Tirmizi et al., 2011}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})), hash vs slash, HTTP status codes and PID redirection strategies (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).
-
-\hypertarget{sec:ld-web}{%
-\subsubsection{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
-
-The \textbf{Linked Data} concept (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) was kickstarted as a set of best practices (\protect\hyperlink{ref-OGXYtGZ8}{{`Linked Data - Design Issues'}}) to bring the Web aspect back into focus.
-Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
-This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
-
-Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph (\protect\hyperlink{ref-oN9D5LkD}{{`The Open Graph protocol'}}), RDFa (\protect\hyperlink{ref-FKIF6ApB}{{`RDFa 1.1 Primer - Third Edition'}}), Microdata (\protect\hyperlink{ref-14Utscfe4}{{`HTML Standard'}})) we find JSON-LD (\protect\hyperlink{ref-X5SWPqSU}{{`JSON-LD 1.1'}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}).
-
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\protect\hyperlink{ref-BFMoe8FC}{{`Designing a Linked Data developer experience'}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\protect\hyperlink{ref-GDrm9dgX}{{`Shapes Constraint Language (SHACL)'}}) or ShEx (\protect\hyperlink{ref-YUzogaV}{{`Shape Expressions (ShEx) 2.1 Primer'}})) can be used to validate RDF Data (\protect\hyperlink{ref-ZeOAPNHa}{Gayo et al., 2017}; \protect\hyperlink{ref-S5BI6Mwx}{Thornton et al., 2019}) before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts (\protect\hyperlink{ref-o596f2JM}{Klímek, Škoda \& Nečaský, 2019}), which then tend to prefer non-semantic two-dimensional formats such as CSV files.
-
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}).
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies (\protect\hyperlink{ref-MhsCt4P3}{Wolstencroft et al., 2011}), the use of Linked Data as part of REST Web APIs (\protect\hyperlink{ref-eS6a6tlz}{Page, De Roure \& Martinez, 2011}) , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy (\protect\hyperlink{ref-117tX9MEG}{2016}), with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
-
-\hypertarget{sec:comparing}{%
-\section{Comparing FDO and existing approaches}\label{sec:comparing}}
-
-To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}), which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
-\item
-  The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), validated against its implementations for completeness.
-\item
-  A Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}), which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
-\item
-  Cross-checks against RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
-\item
-  EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
-\end{enumerate}
-
-The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).
-
-Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) against Linked Data in general.
-
-\hypertarget{sec:interoperability-compare}{%
-\subsection{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
-
-The Interoperability Framework for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
-
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model (\protect\hyperlink{ref-8ZRGFa6y}{Stallings, 1990}) better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
-In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}).
-\label{tbl:fdo-web-interoperability-framework}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
-FDO encourages research communities to develop common type registries to be shared across instances.
-In current DOIP practice, each service have their own types, attributes and operations.
-The wider symbiosis is consistent use of PIDs.
-& The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}) end up being mandated centrally by dominant (often commercial) providers (\protect\hyperlink{ref-11VxBOeZy}{Fielding et al., 2017}), which clients are required to use as-is with special code per service.
-Use of Linked Data enables common tooling and semantic mapping across differences.
-\\
-\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP).
-\texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).
-Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}).
-& ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), schema.org Actions (\protect\hyperlink{ref-dKAekUjL}{{`Schema.org Actions - schema.org'}}), WSDL/SOAP (\protect\hyperlink{ref-LmudecyN}{{`Web Services Description Language (WSDL) Version 2.0 Part 0: Primer'}}), but frequently just as human-readable developer documentation/examples.
-\\
-\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing.
-Every FDO has a type.
-Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}.
-No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) & Lightweight HTTP semantics for authenticity/navigation.
-Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
-Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit.
-\\
-\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation} & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema.
-Multiple byte stream attachments of any media type.
-& Textual HTTP headers (including any signposting), single byte stream of any media type, e.g.~Linked Data formats (JSON-LD, Turtle, RDF/XML) or embedded in document (HTML with RDFa, JSON-LD or Microdata).
-XML was previously the main syntax used by APIs, JSON is now dominant.
-\\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
-\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
-Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable).
-Unique DOIP protocol can be hard to add to existing Web application frameworks, although proxy services have been developed (e.g.~B2SHARE adapter).
-& HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to (for static content) multi-cloud Content-Delivery Networks (CDN).
-Current scalable cloud technologies for Web hosting may not support HTTP features previously seen as important for Semantic Web, e.g.~content negotiation and semantic HTTP status codes.
-\\
-\bottomrule
-\end{longtable}
-
-Based on the analysis shown in table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
-
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services (\protect\hyperlink{ref-39K3yHT3}{Wolstencroft et al., 2013}).
-However, this is mostly done via developer or human interaction (\protect\hyperlink{ref-iYh1wFBt}{Lamprecht et al., 2021}).
-Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.
-
-A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics.
-Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON.
-On the semantic front, FDO requires that every object have a well-defined type and structured form.
-This is clearly not the case on the Web.
-
-In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed.
-These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\protect\hyperlink{ref-hxtgCGjY}{{`The Modern Standards Paradigm - Five Key Principles'}}) to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference (\protect\hyperlink{ref-jrZe6Esu}{Loo, 1970}).
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\protect\hyperlink{ref-p7nwRj8E}{{`DOIP API for HTTP Clients --- Cordra documentation'}}).
-
-It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\protect\hyperlink{ref-68fzID2k}{{`DOI Handbook - Resolution'}}).
-
-\hypertarget{mapping-of-metamodel-concepts}{%
-\subsubsection{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
-
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2923}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2769}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
-\label{tbl:metamodel-concepts}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endhead
-Resource & FDO/DO & Resource \\
-Service & DOIP service & Server/endpoint \\
-Transaction & (not supported) & Conditional requests, \texttt{409\ Conflict} \\
-Process & Extended operations & (primarily stateless), \texttt{100\ Continue}, \texttt{202\ Accepted} \\
-Operation & DOIP Operation & Method, query parameters \\
-Request & DOIP Request & Request \\
-Response & DOIP Response & Response \\
-Message & Segment, \texttt{requestId} & Message, Representation \\
-Channel & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures. & TCP/IP, TLS, UDP \\
-Protocol & DOIP 2.0, ++ & HTTP/1.1, HTTP/2, HTTP/3 \\
-Link & PID/Handle & URL \\
-\bottomrule
-\end{longtable}
-
-From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
-Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \ref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
-
-\hypertarget{sec:doip-fdo-compare}{%
-\subsection{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
-
-The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and using Linked Data Platform (\protect\hyperlink{ref-17OHlMRQA}{Linked Data Platform Working Group, 2015}), as proposed by (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}; \protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) against its current implementations as DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}), with suggestions for required additions.
-\label{tbl:fdo-checks}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endhead
-G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
-& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
-\texttt{http://} URIs replaced by \texttt{https://}.
-& Keep flexibility of RDF serialisation formats which may change more frequently \\
-G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
-DOIP relatively unknown, in effect only one implementation.
-& Further promote DOIP and justify its benefits.
-Build tutorials and OSI open source implementations.
-Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}), however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
-Example-driven documentation.
-\\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
-\\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
-Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
-\\
-G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
-DOIP operations can use any transport protocol.
-& Document transport protocols as FDOs, recommend which transport to use.
-& URI as abstraction base.
-Does not specify PID requirements.
-& Give stronger deployment recommendations.
-\\
-G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
-Unclear when metadata field is a PID or plain text.
-& Make datatype of fields explicit to support navigation.
-& Machine-navigation through URIs via properties and types.
-Unclear when URI should be followed or is just identifier, but always distinct from plain text.
-& \\
-G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
-& Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
-& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
-Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \\
-G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
-No FDOF implementations.
-& Develop demonstrator of FDOF usage based on existing LDP server.
-\\
-G9: \emph{standard compliance} & Handle (\protect\hyperlink{ref-2gXY2Nwh}{Sun, Lannom \& Boesch, 2003}), DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) & HTTP, LDP.
-FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
-FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}).
-Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-& HTTP URLs as basis for identifiers, but they are frequently not persistent.
-& Add strong guidance for PID services like w3id and persistence policies.
-\\
-FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
-Datacite DOIs content-negotiate to give registered metadata.
-& Add FAIR Signposting at PID provider for minimal PID record \\
-FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
-No method/attribute defined for separate metadata, only directly in PID Record.
-Unclear meaning of multiple items and bytestream chunks.
-& Clarify expectations for multiple items.
-Recommend chunks to not be used.
-& URIs resolvable by default.
-Multiple ways to resolve metadata, unclear preference.
-& Add FAIR Signposting and preference order.
-\\
-FDOF4: \emph{Additional attributes} & Freetext attribute keys.
-Attributes should be defined for FDO type (?).
-& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
-Explicitly allow attributes not already defined in type.
-& All attributes individually identified.
-Any Linked Data attributes can be used by URI or with mapped prefix.
-& Clarify type expectations of required/recommended/optional attributes.
-\\
-FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
-& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-Clarify that operations can be mapped to protocol directly.
-& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
-Unclear how to provide PID for additional operations.
-& Specify how additional operations should be called over HTTP.
-\\
-FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan}) are used instead.
-& Document how operation resources can be discovered from an LPD container.
-Document search API.
-\\
-FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
-PR-TypingFDOs leaves this open.
-& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
-Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
-& Always make LDP types explicit in FDO profile.
-\\
-FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
-Separate Metadata FDO need custom property.
-& Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
-& Assertions are always with semantics, using RDF vocabularies.
-Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
-& Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
-FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
-Metadata would need to be bundled within custom FDO types/attributes.
-Unclear how levels are separated within single FDO representation (need FDOF8?).
-& Declare which metadata are expected within response attribute or within FDO object.
-Require PIDs for custom attributes.
-Define how alternate metadata levels can be represented separately.
-& Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
-& Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
-FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
-& Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} (\protect\hyperlink{ref-CNP39tah}{Smith et al., 2007}) & Rather document better how individual ad-hoc schemas can be started for prototypes.
-\\
-FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
-Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-(\protect\hyperlink{ref-6LvydTsC}{{`Data Information View'}})).
-& & LDP Containers required by specification, also user-created (eg.
-\texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 (\protect\hyperlink{ref-1EW4hLN8U}{{`Data Catalog Vocabulary (DCAT) - Version 3'}}), \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE (\protect\hyperlink{ref-gtGuSevK}{{`ORE User Guide - Primer'}}) \\
-FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
-\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
-No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
-\bottomrule
-\end{longtable}
-
-Note that the draft update to FDO specification (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
-
-A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines.
-Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
-
-From our evaluation, we can observe:
-
-\begin{itemize}
-\tightlist
-\item
-  G1 and G2 call for stability and trustworthiness.
-While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
-\item
-  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data.
-Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous.
-In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
-\item
-  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}.
-These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
-\item
-  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
-\item
-  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
-\item
-  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}})
-\item
-  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
-\item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
-\item
-  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation.
-In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
-\end{itemize}
-
-\hypertarget{sec:middleware}{%
-\subsection{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
-
-In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
-As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.
-
-We argue that the Web, with its now ubiquitous use of REST API (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services (\protect\hyperlink{ref-WzPECPDP}{Fensel et al., 2011}) attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\protect\hyperlink{ref-11dInoyVF}{{`5-star Open Data'}}).
-
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}) that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
-
 \begin{longtable}[]{@{}
@@
 \end{longtable}
-
-Based on the analysis in Table \ref{tbl:fdo-web-middleware}, we make the following observations:
-
-\begin{itemize}
-\tightlist
-\item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}) supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
-It is unclear what is the purpose of splitting a binary into chunks on a channel which no longer can be multiplexed and the only property of a chunk is its size\footnote{Although it is possible with \texttt{0.DOIP/Op.Retrieve} to request only particular individual elements of an DO (e.g.~one file), unlike HTTP's \texttt{Range} request, it is not possible to select individual chunks of an element's bytestream.}.
-\item
-  HTTP has strong support for scalability and caching, but this mostly assumes read-operations from static resources.
-FDO has no view on immutability or validity of retrieved objects, but this should be taken into consideration to support large-scale usage.
-\item
-  HTTP optimisations for performance (e.g.~HTTP/2, multiplexing) is largely used for commercial media distribution (e.g.~Netflix), and not commonly used by providers of FAIR data
-\item
-  Cloud deployment of Web applications give many middleware benefits (Scalability, Distribution, Access transparancy, Location transparancy) -- it is unclear how DOIP as a custom protocol would perform in a cloud setting as most of this infrastructure assumes HTTP as the protocol.
-\item
-  Programmatically the Web is rather unstructured as middleware, as there are many implementation choices.
-Usually it is undeclared what to expect for a given URI/service, and programmers follow documented examples for a particular service rather than automated programmatic exploration across providers.
-This mean one can consider the Web as an ecosystem of smaller middlewares with commonalities.
-\item
-  Many providers of FAIR Linked Data also provide programmatic REST API endpoints, e.g.~\href{https://www.uniprot.org/help/programmatic_access}{UNIPROT}, \href{https://chembl.gitbook.io/chembl-interface-documentation/web-services}{ChEMBL}, but keeping the FAIR aspects such as retrieving metadata in such a scenario may require combining different services using multiple formats and identifier conventions.
-\end{itemize}
-
-\hypertarget{sec:fair-compare}{%
-\subsection{Assessing FDO against FAIR}\label{sec:fair-compare}}
-
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}).
-
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) comply with the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-Here we've used the RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}) as it has decomposed the FAIR principles to a structured list of FAIR indicators (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}), importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}; \protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) (first 2 columns) against the FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), FDO implemented with the protocol DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) and examples from Linked Data practices in general.
-(--- indicates \emph{Unspecified}, may be possible with additional conventions)
-\label{tbl:fair-data-maturity-model}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endhead
-RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
-RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
-Handle, DOI.
-& FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
-RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
-FAIR Signposting.
-\\
-RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
-RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
-Unclear how to link to additional metadata.
-& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
-Embedded in landing page (RDFa).
-\\
-RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
-& PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
-RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
-Handle registry only searchable by PID.
-& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-(\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}, pp.
-https://lod--cloud.net/})).
-Indexing by search engines if exposing HTML w/schema.org.
-\\
-RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
-\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
-RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
-RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
-RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
-RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
-RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
-Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}) & HTTP/HTTPS, FTP (now less common), GridFTP (\protect\hyperlink{ref-18LsXvKGJ}{Allcock et al.}) (for large data), ARK (\protect\hyperlink{ref-voDQ6diM}{{`The ARK Identifier Scheme'}}) \\
-RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol (\protect\hyperlink{ref-I9Kx0Hjy}{Sun et al., 2003}).
-One server implementation (\protect\hyperlink{ref-1CvYRVhy2}{{`Handle.Net Registry'}}), free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved (\protect\hyperlink{ref-xVJMmGVr}{{`Licenses \& Standards \textbar{} Open Source Initiative'}}) as an open source license -- it includes usage restrictions and requires Service Agreements.
-It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
-One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
-\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
-& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
-\\
-RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
-RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
-RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
-RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
-RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
-RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
-RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
-RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
-& FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) (\protect\hyperlink{ref-1BlxdxOwa}{14:00-17:00}) \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs (\protect\hyperlink{ref-o4h92LpS}{{`Data - W3C'}})), \texttt{rdfs:seeAlso}, FAIR signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) \texttt{describedby} \\
-RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
-RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
-RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles (\protect\hyperlink{ref-QpDmajc4}{{`Linking Across Provenance Bundles'}}), \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role (\protect\hyperlink{ref-10HA0mMBx}{Unknown}), \texttt{prov:specializationOf} (\protect\hyperlink{ref-ZUgwvAHE}{{`PROV-O: The PROV Ontology'}}), OAI-ORE Proxy (\protect\hyperlink{ref-ID1pTGry}{{`ORE Specification - Abstract Data Model'}})) \\
-RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) but not assigned PIDs yet.
-& Unspecified.
-Multiple metadata records can allow multiple semantic profiles.
-& Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org (\protect\hyperlink{ref-10pzKCEUE}{{`Schema.org - Schema.org'}}), Dublin Core Terms (\protect\hyperlink{ref-wAbZpOsr}{{`DCMI Metadata Terms'}}), DCAT (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), FOAF (\protect\hyperlink{ref-XmUY0cZB}{{`FOAF Vocabulary Specification'}}) \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}) \\
-RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
-RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
-RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
-PID Kernel attributes?
-Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}}), PAV (\protect\hyperlink{ref-1E4OfHyhI}{Ciccarese et al., 2013}), Dublin Core Terms (\protect\hyperlink{ref-hKblceo6}{{`DCMI Metadata Terms'}, 2020}) \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen (\protect\hyperlink{ref-zZTtJruI}{Hardisty et al., 2022})) & --- & Common, e.g.~DCAT 2 (\protect\hyperlink{ref-zt8AAjcZ}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), BioSchemas (\protect\hyperlink{ref-1F8S9FeHf}{{`Bioschemas - Bioschemas'}}) \\
-RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal (\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}}), Darwin Core (\protect\hyperlink{ref-9jaVzCbL}{Wieczorek et al., 2012}) \\
-RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
-\bottomrule
-\end{longtable}
-
-From this evaluation we observe:
-
-\begin{itemize}
-\tightlist
-\item
-  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
-\item
-  FDO/DOIP are stronger on identifier indicators
-\item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) has currently only a couple of implementations and is expressed informally.
-The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
-\item
-  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
-\item
-  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
-\item
-  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-(\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
-Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
-\item
-  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata.
-For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
-In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
-\item
-  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications.
-\texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known.
-g
-\end{itemize}
-
-\hypertarget{eosc-interoperability-framework}{%
-\section{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
-
-\textbf{TODO}: Introduce EOSC IF
-
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) in section 3.6 recommends:
-
-\begin{tablenos:no-prefix-table-caption}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Layer
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Recommendation
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-FDO
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
-Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
-Technical & Easy SLAs for service providers & Unspecified & None \\
-Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
-Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
-Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
-Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
-Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
-Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
-Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
-Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
-Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
-Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
-Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
-Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
-Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
-Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
-Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
-Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
-Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
-Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
-Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
-Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
-Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
-Legal & List recommended licenses & Undefined & Best practice recommendations \\
-Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
-Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
-Legal & GDPR compliance for personal data & Undefined & Undefined \\
-Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
-Legal & Harmonised terms-of-use & Undefined & Undefined \\
-Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
-\bottomrule
-\end{longtable}
-
-\end{tablenos:no-prefix-table-caption}
-
-Observations:
-
-\begin{itemize}
-\tightlist
-\item
-  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
-\item
-  Technical aspects highlighted by EOSC IF
-\item
-  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data.
-There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
-\item
-  FDO principles have strong recommendations for community development of organisational aspects.
-\item
-  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
-\end{itemize}
-
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
-
-\textbf{TODO}
-
-\begin{itemize}
-\tightlist
-\item
-  Ramnifications of ide
-\item
-  Finish tables
-\item
-  Bullet points per table
-\item
-  Send in Google Docs
-\end{itemize}
-
-\hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
-
-The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\protect\hyperlink{ref-kYjzRJ0H}{{`Semantic Web - XML2000 - slide "Architecture"'}}).
-However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
-
-While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
-
-We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-\ldots{}
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
-. \emph{Available at} \href{https://blog.cloudflare.com/http-3-vs-http-2/}{\emph{https://blog.cloudflare.com/http-3-vs-http-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
-. \emph{Available at} \href{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}{\emph{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
-14:00-17:00.ISO/IEC 23009-1:2019.
-\emph{Available at} \href{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}{\emph{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
-Allcock W, Bresnahan J, Kettimuthu R, Link M.
-The Globus Striped GridFTP Framework and Server.
-In: \emph{ACM/IEEE SC 2005 Conference (SC'05)}.
-IEEE,.
-DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
-Anders I, Blanchi Ch, Broeder D, Hellström M, Islam Sh, Jejkal Th, Lannom L, Peters K, Quick R, Schlemmer A, Schwardmann U, Soiland-Reyes S, Strawn G, Uytvanck D van, Weiland C, Wittenburg P, Zwölf C.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}{FAIR digital object overview and specifications}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
-Anders I, Hellström M, Islam S, Jejkal T, Lannom L, Schwardmann U, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}{FDO PID profiles \& attributes}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
-Bahim C, Casorrán-Amilburu C, Dekkers M, Herczog E, Loozen N, Repanas K, Russell K, Stall S.
-2020.
-The FAIR data maturity model: An approach to harmonise FAIR assessments.
-\emph{Data Science Journal} 19.
-DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
-Belshe M, Peon R. 2015. \emph{Hypertext Transfer Protocol Version 2 (HTTP/2)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
-Berners-Lee T, Fielding R, Masinter L. 2005. \emph{Uniform Resource Identifier (URI): Generic Syntax}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
-Berners-Lee T, Fischetti M. 1999. \emph{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}. San Francisco: HarperSanFrancisco.
-
-\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
-Bernstein A, Hendler J, Noy N. 2016. A new look at the semantic web. \emph{Communications of the ACM} 59:35--37. DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
-Bioschemas - Bioschemas. \emph{Available at} \href{http://bioschemas.org/}{\emph{http://bioschemas.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
-Bizer C, Heath T, Berners-Lee T.
-2009.
-Linked data - the story so far.
-\emph{International journal on Semantic Web and information systems} 5:1--22.
-DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
-Blanchi C, Broeder D, Jejkal T, Sharif I, Schlemmer A, Uytvanck D van, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}{FDO -- upload of FDO}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
-Blanchi C, Hellström M, Lannom L, Schwardmann U, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}{Implementation of attributes, types, profiles and registries}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
-Bonino Da Silva Santos LO, Wilkinson MD, Kuzniar A, Kaliyaperumal R, Thompson M, Dumontier M, Burger K.
-2016.
-\href{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}{FAIR data points supporting big data interoperability}.
-In: Zelm M, Doumeingts G, Mendonça JP eds.
-\emph{Enterprise interoperability in the digitized and networked factory of the future}.
-iSTE Press, 270--279.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
-Bonino L, Wittenburg O, Carroll B, Hardisty A, Leggott M, Zwölf C.
-2019.
-\emph{\href{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}{FAIR digital object framework}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
-Carriero VA, Daquino M, Gangemi A, Nuzzolese AG, Peroni S, Presutti V, Tomasi F.
-2020.
-The landscape of ontology reuse approaches.
-In: Cota G, Daquino M, Pozzato GL eds.
-\emph{Applications and practices in ontology design, extraction, and reasoning}.
-Studies on the semantic web.
-IOS Press,.
-DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/SSW200033}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
-Ciccarese P, Soiland-Reyes S, Belhajjame K, Gray AJ, Goble C, Clark T.
-2013.
-PAV ontology: provenance, authoring and versioning.
-\emph{Journal of Biomedical Semantics} 4:37.
-DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
-Clemm G, Amsden J, Ellison T, Kaler C, Whitehead J.
-2002.
-\emph{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}.
-RFC Editor.
-DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
-Content negotiation - HTTP \textbar{} MDN.
-\emph{Available at} \href{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}{\emph{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content\_negotiation}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
-Corcho O, Eriksson M, Kurowski K, Ojsteršek M, Choirat C, Sanden M van de, Coppens F.
-2021.
-\emph{EOSC interoperability framework}.
-Publications Office of the EU.
-DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
-Data - W3C. \emph{Available at} \href{https://www.w3.org/standards/semanticweb/data}{\emph{https://www.w3.org/standards/semanticweb/data}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2. \emph{Available at} \href{https://www.w3.org/TR/vocab-dcat-2/}{\emph{https://www.w3.org/TR/vocab-dcat-2/}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
-Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
-Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
-DCMI Metadata Terms.
-2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
-DCMI Metadata Terms.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
-Delgado JCM.
-2016.
-An Interoperability Framework and Distributed Platform for Fast Data Applications.
-In: \emph{Data Science and Big Data Computing}.
-Springer International Publishing, 3--39.
-DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
-Designing a Linked Data developer experience.
-2018.
-\emph{Available at} \href{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}{\emph{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
-Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation. \emph{Available at} \href{https://www.dona.net/doipv1doc}{\emph{https://www.dona.net/doipv1doc}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
-DOIP and Examples --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip.html}{\emph{https://www.cordra.org/documentation/api/doip.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
-DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
-draft-bhutton-json-schema-00.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}{\emph{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
-draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}{\emph{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
-draft-ietf-quic-http-34.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}{\emph{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
-draft-kelly-json-hal-08.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}{\emph{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
-Duerst M, Suignard M. 2005. \emph{Internationalized Resource Identifiers (IRIs)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
-Dusseault L (ed.). 2007. \emph{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
-FAIR Digital Object Framework Documentation. \emph{Available at} \href{https://fairdigitalobjectframework.org/}{\emph{https://fairdigitalobjectframework.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
-FAIR Digital Objects Forum \textbar. \emph{Available at} \href{https://fairdo.org/}{\emph{https://fairdo.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
-Fensel D, Facca FM, Simperl E, Toma I. 2011. \emph{Semantic Web Services}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
-Fielding RT.
-2000.
-\href{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}{Architectural styles and the design of network-based software architectures}.
-\{THESIS\}.\{DOCTORAL\} Thesis.
-University of California, Irvine.
-
-\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
-Fielding R, Nottingham M, Reschke J (eds.). 2022. \emph{HTTP Semantics}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
-Fielding R, Reschke J (eds.). 2014b. \emph{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
-Fielding R, Reschke J (eds.). 2014a. \emph{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
-Fielding RT, Taylor RN, Erenkrantz JR, Gorlick MM, Whitehead J, Khare R, Oreizy P.
-2017.
-Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award).
-In: \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017}.
-New York, New York, USA: ACM Press, 4--14.
-DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
-FOAF Vocabulary Specification. \emph{Available at} \href{http://xmlns.com/foaf/spec/}{\emph{http://xmlns.com/foaf/spec/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
-Foundation D. 2018. \emph{\href{https://hdl.handle.net/0.DOIP/DOIPV2.0}{Digital object interface protocol specification, version 2.0}}. DONA foundation.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
-Gayo JEL, Prud'hommeaux E, Boneva I, Kontokostas D.
-2017.
-Validating RDF Data.
-\emph{Synthesis Lectures on the Semantic Web: Theory and Technology} 7:1--328.
-DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
-Goble C, Stevens R.
-2008.
-State of the nation in data integration for bioinformatics.
-\emph{Journal of Biomedical Informatics} 41:687--693.
-DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
-Gregorio J, de hOra B (eds.). 2007. \emph{The Atom Publishing Protocol}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
-Gregorio J, Fielding R, Hadley M, Nottingham M, Orchard D. 2012. \emph{URI Template}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
-Groth P, Loizou A, Gray AJG, Goble C, Harland L, Pettifer S.
-2014.
-API-centric Linked Data integration: The Open PHACTS Discovery Platform case study.
-\emph{Journal of Web Semantics} 29:12--18.
-DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
-Group RDAFDMMW. 2020. FAIR data maturity model: Specification and guidelines. \emph{Research Data Alliance}. DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
-Hardisty A, Brack P, Goble C, Livermore L, Scott B, Groom Q, Owen S, Soiland-Reyes S. 2022. The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
-Digital Object Approach to Speeding up Digital Mobilisation of Natural History
-Collections. \emph{Data Intelligence} 4:320--341. DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
-Hasnain A, Rebholz-Schuhmann D.
-2018.
-Assessing FAIR data principles against the 5-star open data principles.
-In: Gangemi A, Gentile AL, Nuzzolese AG, Rudolph S, Maleshkova M, Paulheim H, Pan JZ, Alam M eds.
-\emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers}.
-Lecture notes in computer science.
-Cham: Springer International Publishing, 469--477.
-DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
-Horrocks I, Hendler J (eds.). 2002. \emph{The Semantic Web --- ISWC 2002}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
-HTML Standard. \emph{Available at} \href{https://html.spec.whatwg.org/multipage/microdata.html}{\emph{https://html.spec.whatwg.org/multipage/microdata.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
-Hu W, Chen J, Zhang H, Qu Y.
-2011.
-How matchable are four thousand ontologies on the semantic web.
-In: Antoniou G, Grobelnik M, Simperl E, Parsia B, Plexousakis D, De Leenheer P, Pan J eds.
-\emph{The semantic web: Research and applications}.
-Lecture notes in computer science.
-Berlin, Heidelberg: Springer Berlin Heidelberg, 290--304.
-DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI}{\emph{https://www.w3.org/Provider/Style/URI}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI.html}{\emph{https://www.w3.org/Provider/Style/URI.html}} (accessed 26 May 2022b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
-Iyengar J, Thomson M (eds.). 2021. \emph{QUIC: A UDP-Based Multiplexed and Secure Transport}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
-JSON-LD 1.1. \emph{Available at} \href{http://www.w3.org/TR/json-ld/}{\emph{http://www.w3.org/TR/json-ld/}} (accessed 26 May 2022b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
-JSON-LD 1.1. \emph{Available at} \href{https://www.w3.org/TR/json-ld/}{\emph{https://www.w3.org/TR/json-ld/}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
-Juty N, Le Novere N, Laibe C.
-2011.
-Identifiers.org and MIRIAM Registry: community resources to provide persistent identification.
-\emph{Nucleic Acids Research} 40:D580--D586.
-DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
-Kahn R, Wilensky R. 1995. \emph{\href{http://www.cnri.reston.va.us/k-w.html}{A framework for distributed digital object services}}. CNRI.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
-Kahn R, Wilensky R.
-2006.
-A framework for distributed digital object services.
-\emph{International Journal on Digital Libraries} 6:115--123.
-DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
-Kamdar MR, Tudorache T, Musen MA.
-2017.
-A systematic analysis of term reuse and term overlap across biomedical ontologies.
-\emph{Semantic Web} 8:853--871.
-DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
-Khare R, Lawrence S. 2000. \emph{Upgrading to TLS Within HTTP/1.1}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
-Klein M, Van de Sompel H, Sanderson R, Shankar H, Balakireva L, Zhou K, Tobin R.
-2014.
-Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot.
-\emph{PLoS ONE} 9:e115253.
-DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
-Klímek J, Škoda P, Nečaský M. 2019. Survey of tools for linked data consumption. \emph{Satya Widya} 10:665--720. DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/SW-180316}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
-Lamprecht A-L, Palmblad M, Ison J, Schwämmle V, Al Manir MS, Altintas I, Baker CJO, Ben Hadj Amor A, Capella-Gutierrez S, Charonyktakis P, Crusoe MR, Gil Y, Goble C, Griffin TJ, Groth P, Ienasescu H, Jagtap P, Kalaš M, Kasalica V, Khanteymoori A, Kuhn T, Mei H, Ménager H, Möller S, Richardson RA, Robert V, Soiland-Reyes S, Stevens R, Szaniszlo S, Verberne S, Verhoeven A, Wolstencroft K.
-2021.
-Perspectives on automated composition of workflows in the life sciences.
-\emph{F1000Research} 10:897.
-DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
-Lannom L, Peters-von Gehlen K, Anders I, Pfeil A, Schlemmer A, Trautt Z, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}{FDO configuration types}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
-Lannom L, Schwardmann U, Blanchi C, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}{Typing FAIR digital objects}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
-Linked Data - Design Issues. \emph{Available at} \href{https://www.w3.org/DesignIssues/LinkedData.html}{\emph{https://www.w3.org/DesignIssues/LinkedData.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
-Linked Data Notifications. \emph{Available at} \href{https://www.w3.org/TR/ldn/}{\emph{https://www.w3.org/TR/ldn/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
-Linked Data Platform Working Group. 2015. \emph{\href{https://www.w3.org/TR/2015/REC-ldp-20150226/}{Linked data platform 1.0}}. W3C.
-
-\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
-Loo T (ed.). 1970. First International Conference on FAIR Digital Objects. DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
-martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
-May W. 2017. \emph{HTTP Live Streaming}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
-Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
-2021b.
-Web data apis over SPARQL.
-In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
-Synthesis lectures on data, semantics, and knowledge.
-Cham: Springer International Publishing, 27--38.
-DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
-Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
-2021a.
-Conclusion and future challenges.
-In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
-Synthesis lectures on data, semantics, and knowledge.
-Cham: Springer International Publishing, 77--80.
-DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
-Mika P. 2015. On Schema.org and Why It Matters for the Web. \emph{IEEE Internet Computing} 19:52--55. DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
-Mons B, Neylon C, Velterop J, Dumontier M, Silva Santos LOB da, Wilkinson MD.
-2017.
-Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud.
-\emph{Information Services \& Use} 37:49--56.
-DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{ISU\}-170824}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
-NCBO BioPortal. \emph{Available at} \href{https://bioportal.bioontology.org/ontologies}{\emph{https://bioportal.bioontology.org/ontologies}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
-Neumann A, Laranjeiro N, Bernardino J.
-2021.
-An analysis of public REST web service apis.
-\emph{IEEE Transactions on Services Computing} 14:957--970.
-DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/TSC.2018.2847344}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
-Nottingham M. 2017. \emph{Web Linking}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
-OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More.
-\emph{Available at} \href{https://spec.openapis.org/oas/v3.1.0.html}{\emph{https://spec.openapis.org/oas/v3.1.0.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
-ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
-OWL 2 Web Ontology Language Document Overview (Second Edition). \emph{Available at} \href{http://www.w3.org/TR/owl2-overview/}{\emph{http://www.w3.org/TR/owl2-overview/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
-Page KR, De Roure DC, Martinez K.
-2011.
-REST and Linked Data.
-In: \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11}.
-ACM Press,.
-DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
-Polleres A, Kamdar MR, Fernández JD, Tudorache T, Musen MA.
-2020.
-A more decentralized vision for linked data.
-\emph{Satya Widya} 11:101--113.
-DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/SW-190380}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 January 2023a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
-RDF 1.1 Primer. \emph{Available at} \href{http://www.w3.org/TR/rdf11-primer/}{\emph{http://www.w3.org/TR/rdf11-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
-RDF Schema 1.1. \emph{Available at} \href{http://www.w3.org/TR/rdf-schema/}{\emph{http://www.w3.org/TR/rdf-schema/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
-RDFa 1.1 Primer - Third Edition. \emph{Available at} \href{http://www.w3.org/TR/rdfa-primer/}{\emph{http://www.w3.org/TR/rdfa-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
-Resource Description Framework (RDF) Model and Syntax Specification.
-\emph{Available at} \href{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}{\emph{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
-ResourceSync Framework Specification - Table of Contents. \emph{Available at} \href{http://www.openarchives.org/rs/toc}{\emph{http://www.openarchives.org/rs/toc}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
-Sandvine.Global Internet Phenomena Report 2022.
-\emph{Available at} \href{https://www.sandvine.com/global-internet-phenomena-report-2022}{\emph{https://www.sandvine.com/global-internet-phenomena-report-2022}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
-Sauermann L, Cyganiak R, Völkel M. 2011. \emph{Cool URIs for the semantic web}. Universität des Saarlandes. DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
-Schema.org - Schema.org. \emph{Available at} \href{https://schema.org/}{\emph{https://schema.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
-Schema.org Actions - schema.org. \emph{Available at} \href{https://schema.org/docs/actions.html}{\emph{https://schema.org/docs/actions.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
-Schultes E, Wittenburg P.
-2019.
-FAIR principles and digital objects: Accelerating convergence on a data infrastructure.
-In: Manolopoulos Y, Stupnikov S eds.
-\emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers}.
-Communications in computer and information science.
-Cham: Springer International Publishing, 3--16.
-DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
-Schwardmann U, Kálmán T.
-2022.
-Two Examples on How FDO Types can Support Machine and Human Readability.
-\emph{Research Ideas and Outcomes} 8.
-DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
-Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
-Shape Expressions (ShEx) 2.1 Primer. \emph{Available at} \href{http://shex.io/shex-primer/}{\emph{http://shex.io/shex-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
-Shapes Constraint Language (SHACL). \emph{Available at} \href{https://www.w3.org/TR/shacl/}{\emph{https://www.w3.org/TR/shacl/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
-SKOS Simple Knowledge Organization System Primer. \emph{Available at} \href{http://www.w3.org/TR/skos-primer/}{\emph{http://www.w3.org/TR/skos-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
-Smith B, Ashburner M, Rosse C, Bard J, Bug W, Ceusters W, Goldberg LJ, Eilbeck K, Ireland A, Mungall CJ, Leontis N, Rocca-Serra P, Ruttenberg A, Sansone S-A, Scheuermann RH, Shah N, Whetzel PL, Lewis S.
-2007.
-The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration.
-\emph{Nature Biotechnology} 25:1251--1255.
-DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
-SPARQL 1.1 Overview. \emph{Available at} \href{http://www.w3.org/TR/sparql11-overview/}{\emph{http://www.w3.org/TR/sparql11-overview/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
-Stallings W. 1990. \emph{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}. Carmel, IN, USA: Sams.
-
-\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
-Stanczyk SK. 1987. Process modelling for information system description. \emph{The Open University}. DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
-Stefi A.
-2015.
-Do Developers Make Unbiased Decisions?
-- The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components.
-DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
-Stefi A, Hess T.
-2015.
-To develop or to reuse?
-Two perspectives on external reuse in software projects.
-In: Fernandes JM, Machado RJ, Wnuk K eds.
-\emph{Software business}.
-Lecture notes in business information processing.
-Cham: Springer International Publishing, 192--206.
-DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
-Strawn G, Wittenburg P, Soiland-Reyes St, Peters K, Lannom L, Schwardmann U, Blanch Ch.
-2022.
-\emph{\href{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}{FDO Forum FDO requirement specifications}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
-Sun S, Lannom L, Boesch B. 2003. \emph{Handle System Overview}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
-Sun S, Reilly S, Lannom L, Petrone J. 2003. \emph{Handle System Protocol (ver 2.1) Specification}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
-SWORD 3.0 Specification. \emph{Available at} \href{https://swordapp.github.io/swordv3/swordv3.html}{\emph{https://swordapp.github.io/swordv3/swordv3.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
-The Open Graph protocol. \emph{Available at} \href{https://ogp.me/}{\emph{https://ogp.me/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
-Thornton K, Solbrig H, Stupp GS, Labra Gayo JE, Mietchen D, Prud E, Waagmeester A.
-2019.
-Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation.
-In: Hitzler P, Fernández M, Janowicz K, Zaveri A, Gray AJG, Lopez V, Haller A, Hammar K eds.
-\emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings}.
-Lecture notes in computer science.
-Cham: Springer International Publishing, 606--620.
-DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
-Tirmizi S, Aitken S, Moreira DA, Mungall C, Sequeda J, Shah NH, Miranker DP.
-2011.
-Mapping between the OBO and OWL ontology languages.
-\emph{Journal of Biomedical Semantics} 2:S3.
-DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
-Turcoane O.
-2014.
-Linked data, JSON-LD and the semantics of cultural and scientific heritage.
-\emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} 4:95--105.
-DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
-Usage Statistics of JSON-LD for Websites, May 2022.
-\emph{Available at} \href{https://w3techs.com/technologies/details/da-jsonld}{\emph{https://w3techs.com/technologies/details/da-jsonld}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
-Van de Sompel H, Klein M, Jones S, Nelson ML, Warner S, Devaraju A, Huber R, Steinhoff W, Tykhonov V, Boruta L, Meijers E, Soiland-Reyes S, Wilkinson M.
-2022.FAIR Signposting Profile.
-\emph{Available at} \href{https://signposting.org/FAIR/}{\emph{https://signposting.org/FAIR/}} (accessed 5 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
-Van de Sompel H, Nelson M, Sanderson R. 2013. \emph{HTTP Framework for Time-Based Access to Resource States -\/- Memento}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
-Verborgh R, Vander Sande M.
-2020.
-The semantic web identity crisis: In search of the trivialities that never were.
-\emph{Satya Widya} 11:19--27.
-DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/SW-190372}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
-W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures. \emph{Available at} \href{http://www.w3.org/TR/xmlschema11-1/}{\emph{http://www.w3.org/TR/xmlschema11-1/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
-Web Services Description Language (WSDL) Version 2.0 Part 0: Primer. \emph{Available at} \href{http://www.w3.org/TR/wsdl20-primer/}{\emph{http://www.w3.org/TR/wsdl20-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
-WebSockets Standard. \emph{Available at} \href{https://websockets.spec.whatwg.org/}{\emph{https://websockets.spec.whatwg.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
-Weigel T, Plale B, Parsons M, Zhou G, Luo Y, Schwardmann U, Quick R, Hellström M, Kurakawa K.
-2018.
-RDA Recommendation on PID Kernel Information.
-\emph{Research Data Alliance}.
-DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
-Weiland C, Islam S, Broder D, Anders I, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}{FDO machine actionability}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
-Weiland C, Schwardmann U, Wittenburg P, Kirkpatrick C, Hanisch R, Trautt Z.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}{FDO Forum Document Standards}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
-Wieczorek J, Bloom D, Guralnick R, Blum S, Döring M, Giovanni R, Robertson T, Vieglais D.
-2012.
-Darwin Core: An Evolving Community-Developed Biodiversity Data Standard.
-\emph{PLoS ONE} 7:e29715.
-DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
-Wilde E. 2013. \emph{The 'profile' Link Relation Type}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-6DjakjNS}{}}%
-Wilkinson MD, Dumontier M, Aalbersberg IjJ, Appleton G, Axton M, Baak A, Blomberg N, Boiten J-W, da Silva Santos LB, Bourne PE, Bouwman J, Brookes AJ, Clark T, Crosas M, Dillo I, Dumon O, Edmunds S, Evelo CT, Finkers R, Gonzalez-Beltran A, Gray AJG, Groth P, Goble C, Grethe JS, Heringa J, 't Hoen PAC, Hooft R, Kuhn T, Kok R, Kok J, Lusher SJ, Martone ME, Mons A, Packer AL, Persson B, Rocca-Serra P, Roos M, van Schaik R, Sansone S-A, Schultes E, Sengstag T, Slater T, Strawn G, Swertz MA, Thompson M, van der Lei J, van Mulligen E, Velterop J, Waagmeester A, Wittenburg P, Wolstencroft K, Zhao J, Mons B.
-2016.
-The FAIR Guiding Principles for scientific data management and stewardship.
-\emph{Scientific Data} 3.
-DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pa67pUtR}{}}%
-Wilkinson SR, Eisenhauer G, Kapadia AJ, Knight K, Logan J, Widener P, Wolf M.
-2022.
-F*** workflows: when parts of FAIR are missing.
-\emph{arXiv}.
-DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
-Williams AJ, Harland L, Groth P, Pettifer S, Chichester C, Willighagen EL, Evelo CT, Blomberg N, Ecker G, Goble C, Mons B.
-2012.
-Open PHACTS: semantic interoperability for drug discovery.
-\emph{Drug Discovery Today} 17:1188--1198.
-DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
-Wittenburg P, Anders I, Blanchi C, Buurman M, Goble C, Grieb J, Hardisty A, Islam S, Jejkal T, Kálmán T, Kirkpatrick C, Lannom L, Lauer T, Manepalli G, Peters-von Gehlen K, Pfeil A, Quick R, Sanden M van de, Schwardmann U, Soiland-Reyes S, Stotzka R, Trautt Z, Van Uytvanck D, Weiland C, Wieder P.
-2022.
-FAIR digital object demonstrators 2021.
-\emph{Zenodo}.
-DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
-Wittenburg P, Strawn G, Mons B, Bonino L, Schultes E.
-2019.
-Digital objects as drivers towards convergence in data infrastructures.
-\emph{https://b2share.eudat.eu}.
-DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
-Wolstencroft K, Haines R, Fellows D, Williams A, Withers D, Owen S, Soiland-Reyes S, Dunlop I, Nenadic A, Fisher P, Bhagat J, Belhajjame K, Bacall F, Hardisty A, Nieva de la Hidalga A, Balcazar Vargas MP, Sufi S, Goble C.
-2013.
-The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud.
-\emph{Nucleic Acids Research} 41:W557--W561.
-DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
-Wolstencroft K, Owen S, Horridge M, Krebs O, Mueller W, Snoep JL, du Preez F, Goble C.
-2011.
-RightField: embedding ontology annotation in spreadsheets.
-\emph{Bioinformatics} 27:2021--2022.
-DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
-Working Group FDO-TSIG. 2022a. \emph{\href{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}{DOIP endorsement request}}. (internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
-Working Group FDO-TSIG.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}{FDO -- granularity, versioning, mutability}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Fi62cJAo}{}}%
-Working Group FDO-TSIG, Working Group FDO-BIG.
-2022.
-\emph{\href{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}{FDO -- kernel attributes \& metadata}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
-X.1255 : Framework for discovery of identity management information.
-\emph{Available at} \href{https://www.itu.int/rec/T-REC-X.1255-201309-I}{\emph{https://www.itu.int/rec/T-REC-X.1255-201309-I}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRzcHhPD}{}}%
-Zarras A. 2004. A Comparison Framework for Middleware Infrastructures. \emph{The Journal of Object Technology} 3:103. DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}.
-
-\end{CSLReferences}

### 2023-01-24 latex/table4.tex
converted to peerj overleaf template

--- 2023-01-24 manuscript.tex
@@
-This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
-was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
-on January 24, 2023.
-
-\hypertarget{authors}{%
-\section*{Authors}\label{authors}}
-\addcontentsline{toc}{section}{Authors}
-
-\begin{itemize}
-\item
-  \textbf{Stian Soiland-Reyes}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/stain}{stain}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/soilandreyes}{soilandreyes}
-  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Carole Goble}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/carolegoble}{carolegoble}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
-  Department of Computer Science, The University of Manchester, UK
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Paul Groth}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/pgroth}{pgroth}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/pgroth}{pgroth}
-  Informatics Institute, Faculty of Science, University of Amsterdam, NL
-\end{itemize}
-
-\hypertarget{abstract}{%
-\section*{Abstract}\label{abstract}}
-\addcontentsline{toc}{section}{Abstract}
-
-\textbf{TODO}: Rewrite from notes below to actual abstract.
-
-FAIR Digital Object is an emerging concept from EOSC.
-This is important.
-Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
-Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
-
-What is the message of this paper?
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These are the overlaps
-\item
-  This is what FDO is requiring but not commonly deployed
-\item
-  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
-\item
-  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
-\item
-  Lessons for Semantic Web community. What is missing in FDO and in SW.
-\item
-  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
-\item
-  Go back to the old SW layer cake. Explains why we picked these frameworks.
-\item
-  Lessons for SW: Parts of the stack that is less. FDO contributions.
-\end{enumerate}
-
-Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.
-
-Have all the ingredients, but not cooking properly.
-
-Systematic through frameworks.
-
-Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?
-
-\hypertarget{sec:introduction}{%
-\section{Introduction}\label{sec:introduction}}
-
-The FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}) encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures (\protect\hyperlink{ref-u8Cy0psL}{Mons et al., 2017}).
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
-
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
-The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
-
-In other discourse, Linked Data (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\protect\hyperlink{ref-EJDjHrUv}{Bonino Da Silva Santos et al., 2016}; \protect\hyperlink{ref-GJHYN6xW}{Hasnain \& Rebholz-Schuhmann, 2018}).
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).
-
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\protect\hyperlink{ref-uxWjmlio}{Stefi, 2015}; \protect\hyperlink{ref-v0WwFHgT}{Stefi \& Hess, 2015}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves (\protect\hyperlink{ref-5V3GqSzR}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021b}).
-
-These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.
-
-Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.
-
-\hypertarget{sec:background}{%
-\section{Background and related work}\label{sec:background}}
-
-In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.
-
-\hypertarget{sec:fdo}{%
-\subsection{FAIR Digital Object}\label{sec:fdo}}
-
-\hypertarget{sec:next-step-fdo}{%
-\subsubsection{Next steps for FDO}\label{sec:next-step-fdo}}
-
-The FAIR Digital Object Forum (\protect\hyperlink{ref-DMEBhHID}{{`FAIR Digital Objects Forum \textbar{}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
-As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
-
-\textbf{FAIR Digital Object Overview and Specifications} (\protect\hyperlink{ref-11NpETLq1}{Anders et al., 2022a}) is a comprehensive overview of FAIR Digital Object specifications listed below.
-It serves as a primer that introduces FDO concepts and the remaining documents.
-
-The \textbf{FDO Forum Document Standards} (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
-In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
-
-The \textbf{FDO Requirement Specifications} (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) is an update of (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) as the foundational definition of FDO.
-This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
-The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
-
-The \textbf{Machine actionability} (\protect\hyperlink{ref-iJeys0T5}{Weiland et al., 2022a}) sets out to define what is meant by \emph{machine actionability} for FDOs.
-\emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
-The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.
-
-\textbf{Configuration Types} (\protect\hyperlink{ref-SPd0uEqO}{Lannom et al., 2022a}) classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022})
-
-\textbf{PID Profiles \& Attributes} (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}) specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
-
-\textbf{Kernel Attributes \& Metadata} (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
-It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
-
-\textbf{Granularity, Versioning, Mutability} (\protect\hyperlink{ref-16W9AycSW}{Working Group, 2022b}) considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
-The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
-
-\textbf{DOIP Endorsement Request} (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) is an endorsement of the DOIP v2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) specification as a potential FDO implementation, as it has been applied by several institutions (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}).
-The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
-
-\textbf{Upload of FDO} (\protect\hyperlink{ref-1H7cmmr69}{Blanchi et al., 2022a}) illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync (\protect\hyperlink{ref-eS5y9TRh}{{`ResourceSync Framework Specification - Table of Contents'}}) is suggested as one type of service to list FDOs.
-This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
-
-\textbf{Typing FAIR Digital Objects} (\protect\hyperlink{ref-ZFzPxCNB}{Lannom et al., 2022b}) defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
-This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
-Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work.
-How to define such FDO operations is not specified.
-
-\textbf{Implementation of Attributes, Types, Profiles and Registries} (\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
-This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} (\protect\hyperlink{ref-1CqIZ47pu}{Schwardmann \& Kálmán, 2022}).
-
-It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-
-The concept of \textbf{FAIR Digital Objects} (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) has been introduced as way to expose research data as active objects that conform to the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-This builds on the \emph{Digital Object} (DO) concept (\protect\hyperlink{ref-11MnuwJ4l}{Kahn \& Wilensky, 2006}), first introduced in 1995 (\protect\hyperlink{ref-3Uqe3fuK}{Kahn \& Wilensky, 1995}) as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework (\protect\hyperlink{ref-103Hw8H43}{{`X.1255 : Framework for discovery of identity management information'}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}).
-
-In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}.
-The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing (\protect\hyperlink{ref-tz0P3DTC}{Wittenburg et al., 2019}).
-
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}), in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}).
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
-
-\hypertarget{fdo-approaches}{%
-\subsubsection{FDO approaches}\label{fdo-approaches}}
-
-FDO is an evolving concept.
-A set of FDO Demonstrators (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) highlight how current adapters are approaching implementations of FDO from different angles:
-
-\begin{itemize}
-\tightlist
-\item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), see (\protect\hyperlink{ref-xuQKRRx5}{{`DOIP and Examples --- Cordra documentation'}}).} (usually TCP/IP over TLS).
-The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}}) and PIDs are assigned using the Handle system.
-Several type registries have been established.
-\item
-  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
-\item
-  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
-\end{itemize}
-
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
-
-\hypertarget{sec:ld}{%
-\subsection{From the Semantic Web to Linked Data}\label{sec:ld}}
-
-In order to describe \emph{Linked Data} as it is used today, we'll start with an (opinionated) description of the evolution of its foundation, the \emph{Semantic Web}.
-
-\hypertarget{sec:semweb}{%
-\subsubsection{A brief history of the Semantic Web}\label{sec:semweb}}
-
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee (\protect\hyperlink{ref-Guy7cjVY}{Berners-Lee \& Fischetti, 1999}), at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
-The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
-The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.
-
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\protect\hyperlink{ref-16WGmCih8}{{`RDF 1.1 Primer'}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\protect\hyperlink{ref-16wvFH4nh}{{`Resource Description Framework (RDF) Model and Syntax Specification'}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
-
-While triple-based knowledge representations were not new (\protect\hyperlink{ref-NxvhwSmG}{Stanczyk, 1987}), the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs (\protect\hyperlink{ref-RRtwPF7Y}{Berners-Lee, Fielding \& Masinter, 2005}) are generalised forms of URLs that include locator-less identifiers
-  such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years (\protect\hyperlink{ref-nW4HY8Nq}{{`"info" URI Registry (Frozen)'}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) (\protect\hyperlink{ref-HqZ3r6J3}{Juty, Le Novere \& Laibe, 2011}).
-RDF 1.1 formats use Unicode to support IRIs (\protect\hyperlink{ref-rqjuXkEf}{Duerst \& Suignard, 2005}), which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
-
-The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data).
-Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term.
-The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource.
-By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs.
-The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.
-
-The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} (\protect\hyperlink{ref-yd2WE9eq}{Horrocks \& Hendler, 2002}) gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
-
-Established Semantic Web services and standards include SPARQL (\protect\hyperlink{ref-xww8fb1B}{{`SPARQL 1.1 Overview'}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
-
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\protect\hyperlink{ref-bMSxa8XW}{Goble \& Stevens, 2008}; \protect\hyperlink{ref-Qlrlr2TZ}{Williams et al., 2012}).
-In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
-
-Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) (\protect\hyperlink{ref-FDDmgO6s}{Hu et al., 2011}) .
-
-The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS (\protect\hyperlink{ref-191ZkYZMt}{Groth et al., 2014})), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
-
-With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps (\protect\hyperlink{ref-9CdJnsoJ}{Kamdar, Tudorache \& Musen, 2017}).
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication (\protect\hyperlink{ref-AGHR9oyr}{Klein et al., 2014}), for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data (\protect\hyperlink{ref-5SuwLNdE}{Polleres et al., 2020, p.
-handle:20.500.11811/7183}).
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies (\protect\hyperlink{ref-TUmA4KQT}{Carriero et al., 2020}) , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), OBO (\protect\hyperlink{ref-1DalZuKct}{Tirmizi et al., 2011}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})), hash vs slash, HTTP status codes and PID redirection strategies (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).
-
-\hypertarget{sec:ld-web}{%
-\subsubsection{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
-
-The \textbf{Linked Data} concept (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) was kickstarted as a set of best practices (\protect\hyperlink{ref-OGXYtGZ8}{{`Linked Data - Design Issues'}}) to bring the Web aspect back into focus.
-Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
-This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
-
-Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph (\protect\hyperlink{ref-oN9D5LkD}{{`The Open Graph protocol'}}), RDFa (\protect\hyperlink{ref-FKIF6ApB}{{`RDFa 1.1 Primer - Third Edition'}}), Microdata (\protect\hyperlink{ref-14Utscfe4}{{`HTML Standard'}})) we find JSON-LD (\protect\hyperlink{ref-X5SWPqSU}{{`JSON-LD 1.1'}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}).
-
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\protect\hyperlink{ref-BFMoe8FC}{{`Designing a Linked Data developer experience'}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\protect\hyperlink{ref-GDrm9dgX}{{`Shapes Constraint Language (SHACL)'}}) or ShEx (\protect\hyperlink{ref-YUzogaV}{{`Shape Expressions (ShEx) 2.1 Primer'}})) can be used to validate RDF Data (\protect\hyperlink{ref-ZeOAPNHa}{Gayo et al., 2017}; \protect\hyperlink{ref-S5BI6Mwx}{Thornton et al., 2019}) before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts (\protect\hyperlink{ref-o596f2JM}{Klímek, Škoda \& Nečaský, 2019}), which then tend to prefer non-semantic two-dimensional formats such as CSV files.
-
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}).
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies (\protect\hyperlink{ref-MhsCt4P3}{Wolstencroft et al., 2011}), the use of Linked Data as part of REST Web APIs (\protect\hyperlink{ref-eS6a6tlz}{Page, De Roure \& Martinez, 2011}) , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy (\protect\hyperlink{ref-117tX9MEG}{2016}), with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
-
-\hypertarget{sec:comparing}{%
-\section{Comparing FDO and existing approaches}\label{sec:comparing}}
-
-To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}), which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
-\item
-  The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), validated against its implementations for completeness.
-\item
-  A Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}), which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
-\item
-  Cross-checks against RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
-\item
-  EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
-\end{enumerate}
-
-The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).
-
-Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) against Linked Data in general.
-
-\hypertarget{sec:interoperability-compare}{%
-\subsection{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
-
-The Interoperability Framework for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
-
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model (\protect\hyperlink{ref-8ZRGFa6y}{Stallings, 1990}) better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
-In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}).
-\label{tbl:fdo-web-interoperability-framework}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
-FDO encourages research communities to develop common type registries to be shared across instances.
-In current DOIP practice, each service have their own types, attributes and operations.
-The wider symbiosis is consistent use of PIDs.
-& The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}) end up being mandated centrally by dominant (often commercial) providers (\protect\hyperlink{ref-11VxBOeZy}{Fielding et al., 2017}), which clients are required to use as-is with special code per service.
-Use of Linked Data enables common tooling and semantic mapping across differences.
-\\
-\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP).
-\texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).
-Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}).
-& ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), schema.org Actions (\protect\hyperlink{ref-dKAekUjL}{{`Schema.org Actions - schema.org'}}), WSDL/SOAP (\protect\hyperlink{ref-LmudecyN}{{`Web Services Description Language (WSDL) Version 2.0 Part 0: Primer'}}), but frequently just as human-readable developer documentation/examples.
-\\
-\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing.
-Every FDO has a type.
-Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}.
-No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) & Lightweight HTTP semantics for authenticity/navigation.
-Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
-Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit.
-\\
-\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation} & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema.
-Multiple byte stream attachments of any media type.
-& Textual HTTP headers (including any signposting), single byte stream of any media type, e.g.~Linked Data formats (JSON-LD, Turtle, RDF/XML) or embedded in document (HTML with RDFa, JSON-LD or Microdata).
-XML was previously the main syntax used by APIs, JSON is now dominant.
-\\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
-\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
-Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable).
-Unique DOIP protocol can be hard to add to existing Web application frameworks, although proxy services have been developed (e.g.~B2SHARE adapter).
-& HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to (for static content) multi-cloud Content-Delivery Networks (CDN).
-Current scalable cloud technologies for Web hosting may not support HTTP features previously seen as important for Semantic Web, e.g.~content negotiation and semantic HTTP status codes.
-\\
-\bottomrule
-\end{longtable}
-
-Based on the analysis shown in table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
-
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services (\protect\hyperlink{ref-39K3yHT3}{Wolstencroft et al., 2013}).
-However, this is mostly done via developer or human interaction (\protect\hyperlink{ref-iYh1wFBt}{Lamprecht et al., 2021}).
-Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.
-
-A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics.
-Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON.
-On the semantic front, FDO requires that every object have a well-defined type and structured form.
-This is clearly not the case on the Web.
-
-In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed.
-These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\protect\hyperlink{ref-hxtgCGjY}{{`The Modern Standards Paradigm - Five Key Principles'}}) to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference (\protect\hyperlink{ref-jrZe6Esu}{Loo, 1970}).
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\protect\hyperlink{ref-p7nwRj8E}{{`DOIP API for HTTP Clients --- Cordra documentation'}}).
-
-It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\protect\hyperlink{ref-68fzID2k}{{`DOI Handbook - Resolution'}}).
-
-\hypertarget{mapping-of-metamodel-concepts}{%
-\subsubsection{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
-
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2923}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2769}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
-\label{tbl:metamodel-concepts}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endhead
-Resource & FDO/DO & Resource \\
-Service & DOIP service & Server/endpoint \\
-Transaction & (not supported) & Conditional requests, \texttt{409\ Conflict} \\
-Process & Extended operations & (primarily stateless), \texttt{100\ Continue}, \texttt{202\ Accepted} \\
-Operation & DOIP Operation & Method, query parameters \\
-Request & DOIP Request & Request \\
-Response & DOIP Response & Response \\
-Message & Segment, \texttt{requestId} & Message, Representation \\
-Channel & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures. & TCP/IP, TLS, UDP \\
-Protocol & DOIP 2.0, ++ & HTTP/1.1, HTTP/2, HTTP/3 \\
-Link & PID/Handle & URL \\
-\bottomrule
-\end{longtable}
-
-From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
-Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \ref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
-
-\hypertarget{sec:doip-fdo-compare}{%
-\subsection{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
-
-The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and using Linked Data Platform (\protect\hyperlink{ref-17OHlMRQA}{Linked Data Platform Working Group, 2015}), as proposed by (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}; \protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) against its current implementations as DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}), with suggestions for required additions.
-\label{tbl:fdo-checks}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endhead
-G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
-& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
-\texttt{http://} URIs replaced by \texttt{https://}.
-& Keep flexibility of RDF serialisation formats which may change more frequently \\
-G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
-DOIP relatively unknown, in effect only one implementation.
-& Further promote DOIP and justify its benefits.
-Build tutorials and OSI open source implementations.
-Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}), however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
-Example-driven documentation.
-\\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
-\\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
-Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
-\\
-G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
-DOIP operations can use any transport protocol.
-& Document transport protocols as FDOs, recommend which transport to use.
-& URI as abstraction base.
-Does not specify PID requirements.
-& Give stronger deployment recommendations.
-\\
-G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
-Unclear when metadata field is a PID or plain text.
-& Make datatype of fields explicit to support navigation.
-& Machine-navigation through URIs via properties and types.
-Unclear when URI should be followed or is just identifier, but always distinct from plain text.
-& \\
-G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
-& Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
-& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
-Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \\
-G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
-No FDOF implementations.
-& Develop demonstrator of FDOF usage based on existing LDP server.
-\\
-G9: \emph{standard compliance} & Handle (\protect\hyperlink{ref-2gXY2Nwh}{Sun, Lannom \& Boesch, 2003}), DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) & HTTP, LDP.
-FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
-FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}).
-Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-& HTTP URLs as basis for identifiers, but they are frequently not persistent.
-& Add strong guidance for PID services like w3id and persistence policies.
-\\
-FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
-Datacite DOIs content-negotiate to give registered metadata.
-& Add FAIR Signposting at PID provider for minimal PID record \\
-FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
-No method/attribute defined for separate metadata, only directly in PID Record.
-Unclear meaning of multiple items and bytestream chunks.
-& Clarify expectations for multiple items.
-Recommend chunks to not be used.
-& URIs resolvable by default.
-Multiple ways to resolve metadata, unclear preference.
-& Add FAIR Signposting and preference order.
-\\
-FDOF4: \emph{Additional attributes} & Freetext attribute keys.
-Attributes should be defined for FDO type (?).
-& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
-Explicitly allow attributes not already defined in type.
-& All attributes individually identified.
-Any Linked Data attributes can be used by URI or with mapped prefix.
-& Clarify type expectations of required/recommended/optional attributes.
-\\
-FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
-& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-Clarify that operations can be mapped to protocol directly.
-& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
-Unclear how to provide PID for additional operations.
-& Specify how additional operations should be called over HTTP.
-\\
-FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan}) are used instead.
-& Document how operation resources can be discovered from an LPD container.
-Document search API.
-\\
-FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
-PR-TypingFDOs leaves this open.
-& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
-Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
-& Always make LDP types explicit in FDO profile.
-\\
-FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
-Separate Metadata FDO need custom property.
-& Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
-& Assertions are always with semantics, using RDF vocabularies.
-Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
-& Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
-FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
-Metadata would need to be bundled within custom FDO types/attributes.
-Unclear how levels are separated within single FDO representation (need FDOF8?).
-& Declare which metadata are expected within response attribute or within FDO object.
-Require PIDs for custom attributes.
-Define how alternate metadata levels can be represented separately.
-& Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
-& Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
-FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
-& Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} (\protect\hyperlink{ref-CNP39tah}{Smith et al., 2007}) & Rather document better how individual ad-hoc schemas can be started for prototypes.
-\\
-FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
-Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-(\protect\hyperlink{ref-6LvydTsC}{{`Data Information View'}})).
-& & LDP Containers required by specification, also user-created (eg.
-\texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 (\protect\hyperlink{ref-1EW4hLN8U}{{`Data Catalog Vocabulary (DCAT) - Version 3'}}), \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE (\protect\hyperlink{ref-gtGuSevK}{{`ORE User Guide - Primer'}}) \\
-FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
-\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
-No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
-\bottomrule
-\end{longtable}
-
-Note that the draft update to FDO specification (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
-
-A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines.
-Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
-
-From our evaluation, we can observe:
-
-\begin{itemize}
-\tightlist
-\item
-  G1 and G2 call for stability and trustworthiness.
-While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
-\item
-  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data.
-Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous.
-In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
-\item
-  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}.
-These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
-\item
-  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
-\item
-  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
-\item
-  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}})
-\item
-  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
-\item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
-\item
-  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation.
-In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
-\end{itemize}
-
-\hypertarget{sec:middleware}{%
-\subsection{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
-
-In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
-As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.
-
-We argue that the Web, with its now ubiquitous use of REST API (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services (\protect\hyperlink{ref-WzPECPDP}{Fensel et al., 2011}) attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\protect\hyperlink{ref-11dInoyVF}{{`5-star Open Data'}}).
-
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}) that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018})) and Web technologies (using Linked Data) as middleware infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004})
-\label{tbl:fdo-web-middleware}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
-Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
-PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}.
-& The Web is inherently open and made by cross-linked URLs.
-Participation requires DNS domain purchase (many free alternatives also exists).
-PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects.
-\\
-\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
-HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
-\\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
-DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
-Multiple connections possible.
-Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks (\protect\hyperlink{ref-F8VZ86hu}{Sandvine}).
-HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}) but still has TCP+TLS startup costs.
-For reduced latency , HTTP/3 (\protect\hyperlink{ref-Amit86Tp}{{`draft-ietf-quic-http-34'}}) use QUIC (\protect\hyperlink{ref-1GWvyEmUf}{Iyengar \& Thomson, 2021})) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
-\\
-\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
-FDOs should provide the mandatory kernel metadata fields.
-FDOs of the same declared type typically share additional attributes (although that schema may not be declared).
-DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions.
-& Each URL accessed separately.
-Common HTTP headers provide basic metadata, although it is often not reliable.
-A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), content-negotiation (\protect\hyperlink{ref-n2f3TfoI}{{`Content negotiation - HTTP \textbar{} MDN'}}) or FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}).
-\\
-\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
-Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
-& Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
-\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
-Multiple locations possible in Handle system, can expose geo-info.
-& PIDs and URL redirects.
-DNS aliases and IP routing can hide location.
-Geo-localised servers common for large cloud deployments.
-\\
-\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
-FDO kernel metadata can include checksum and date.
-& HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
-Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
-\\
-\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
-Custom codes uncommon.
-\\
-\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
-No redirection from DOIP service.
-& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
-\\
-\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
-There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged (\protect\hyperlink{ref-rbG9uRKw}{{`Hypertext Style: Cool URIs don't change.'}}).
-Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
-An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento (\protect\hyperlink{ref-DNjcMi4a}{Van de Sompel, Nelson \& Sanderson, 2013}) expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method (\protect\hyperlink{ref-hRUsaTiz}{Clemm et al., 2002}) (used by SVN).
-\\
-\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
-Internal synchronisation possible in backend for Extended operations.
-& Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} (\protect\hyperlink{ref-xvMAyRAc}{Dusseault, 2007}) with \texttt{LOCK} and \texttt{UNLOCK} methods.
-\\
-\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
-In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
-& The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space.
-In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend.
-\\
-\textbf{Encapsulation}: \emph{separate interface from implementation.
-Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
-FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
-Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade (\protect\hyperlink{ref-1B16WfR7q}{Khare \& Lawrence, 2000}) to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns (\protect\hyperlink{ref-iME5l1kk}{{`Hypertext Style: Cool URIs don't change.'}}) and PIDs via Persistent URLs (PURL).
-\\
-\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
-FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})) & Syntactically Media Type with multiple suffixes (\protect\hyperlink{ref-nlH2rm1s}{{`draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes'}}) (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) (\protect\hyperlink{ref-nCzRO0pK}{Wilde, 2013}).
-In metadata, semantic type systems (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL2 (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})).
-OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} (\protect\hyperlink{ref-11oVTMsuZ}{{`W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures'}}), JSON \texttt{\$schema} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})), JSON-LD \texttt{@context} (\protect\hyperlink{ref-by6CfOmv}{{`JSON-LD 1.1'}}).
-Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
-\\
-\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
-\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 (\protect\hyperlink{ref-tIkSXrb5}{Fielding \& Reschke, 2014b}) \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) (\protect\hyperlink{ref-PL512B0X}{May, 2017}), MPEG-DASH (ISO/IEC 23009-1:2019) (\protect\hyperlink{ref-ikO7e0mh}{14:00-17:00}) \\
-\bottomrule
-\end{longtable}
-
-Based on the analysis in Table \ref{tbl:fdo-web-middleware}, we make the following observations:
-
-\begin{itemize}
-\tightlist
-\item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}) supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
-It is unclear what is the purpose of splitting a binary into chunks on a channel which no longer can be multiplexed and the only property of a chunk is its size\footnote{Although it is possible with \texttt{0.DOIP/Op.Retrieve} to request only particular individual elements of an DO (e.g.~one file), unlike HTTP's \texttt{Range} request, it is not possible to select individual chunks of an element's bytestream.}.
-\item
-  HTTP has strong support for scalability and caching, but this mostly assumes read-operations from static resources.
-FDO has no view on immutability or validity of retrieved objects, but this should be taken into consideration to support large-scale usage.
-\item
-  HTTP optimisations for performance (e.g.~HTTP/2, multiplexing) is largely used for commercial media distribution (e.g.~Netflix), and not commonly used by providers of FAIR data
-\item
-  Cloud deployment of Web applications give many middleware benefits (Scalability, Distribution, Access transparancy, Location transparancy) -- it is unclear how DOIP as a custom protocol would perform in a cloud setting as most of this infrastructure assumes HTTP as the protocol.
-\item
-  Programmatically the Web is rather unstructured as middleware, as there are many implementation choices.
-Usually it is undeclared what to expect for a given URI/service, and programmers follow documented examples for a particular service rather than automated programmatic exploration across providers.
-This mean one can consider the Web as an ecosystem of smaller middlewares with commonalities.
-\item
-  Many providers of FAIR Linked Data also provide programmatic REST API endpoints, e.g.~\href{https://www.uniprot.org/help/programmatic_access}{UNIPROT}, \href{https://chembl.gitbook.io/chembl-interface-documentation/web-services}{ChEMBL}, but keeping the FAIR aspects such as retrieving metadata in such a scenario may require combining different services using multiple formats and identifier conventions.
-\end{itemize}
-
-\hypertarget{sec:fair-compare}{%
-\subsection{Assessing FDO against FAIR}\label{sec:fair-compare}}
-
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}).
-
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) comply with the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-Here we've used the RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}) as it has decomposed the FAIR principles to a structured list of FAIR indicators (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}), importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
-
 \begin{longtable}[]{@{}
@@
 \end{longtable}
-
-From this evaluation we observe:
-
-\begin{itemize}
-\tightlist
-\item
-  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
-\item
-  FDO/DOIP are stronger on identifier indicators
-\item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) has currently only a couple of implementations and is expressed informally.
-The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
-\item
-  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
-\item
-  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
-\item
-  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-(\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
-Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
-\item
-  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata.
-For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
-In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
-\item
-  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications.
-\texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known.
-g
-\end{itemize}
-
-\hypertarget{eosc-interoperability-framework}{%
-\section{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
-
-\textbf{TODO}: Introduce EOSC IF
-
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) in section 3.6 recommends:
-
-\begin{tablenos:no-prefix-table-caption}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Layer
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Recommendation
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-FDO
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
-Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
-Technical & Easy SLAs for service providers & Unspecified & None \\
-Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
-Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
-Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
-Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
-Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
-Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
-Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
-Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
-Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
-Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
-Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
-Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
-Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
-Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
-Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
-Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
-Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
-Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
-Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
-Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
-Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
-Legal & List recommended licenses & Undefined & Best practice recommendations \\
-Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
-Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
-Legal & GDPR compliance for personal data & Undefined & Undefined \\
-Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
-Legal & Harmonised terms-of-use & Undefined & Undefined \\
-Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
-\bottomrule
-\end{longtable}
-
-\end{tablenos:no-prefix-table-caption}
-
-Observations:
-
-\begin{itemize}
-\tightlist
-\item
-  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
-\item
-  Technical aspects highlighted by EOSC IF
-\item
-  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data.
-There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
-\item
-  FDO principles have strong recommendations for community development of organisational aspects.
-\item
-  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
-\end{itemize}
-
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
-
-\textbf{TODO}
-
-\begin{itemize}
-\tightlist
-\item
-  Ramnifications of ide
-\item
-  Finish tables
-\item
-  Bullet points per table
-\item
-  Send in Google Docs
-\end{itemize}
-
-\hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
-
-The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\protect\hyperlink{ref-kYjzRJ0H}{{`Semantic Web - XML2000 - slide "Architecture"'}}).
-However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
-
-While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
-
-We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-\ldots{}
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
-. \emph{Available at} \href{https://blog.cloudflare.com/http-3-vs-http-2/}{\emph{https://blog.cloudflare.com/http-3-vs-http-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
-. \emph{Available at} \href{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}{\emph{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
-14:00-17:00.ISO/IEC 23009-1:2019.
-\emph{Available at} \href{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}{\emph{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
-Allcock W, Bresnahan J, Kettimuthu R, Link M.
-The Globus Striped GridFTP Framework and Server.
-In: \emph{ACM/IEEE SC 2005 Conference (SC'05)}.
-IEEE,.
-DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
-Anders I, Blanchi Ch, Broeder D, Hellström M, Islam Sh, Jejkal Th, Lannom L, Peters K, Quick R, Schlemmer A, Schwardmann U, Soiland-Reyes S, Strawn G, Uytvanck D van, Weiland C, Wittenburg P, Zwölf C.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}{FAIR digital object overview and specifications}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
-Anders I, Hellström M, Islam S, Jejkal T, Lannom L, Schwardmann U, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}{FDO PID profiles \& attributes}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
-Bahim C, Casorrán-Amilburu C, Dekkers M, Herczog E, Loozen N, Repanas K, Russell K, Stall S.
-2020.
-The FAIR data maturity model: An approach to harmonise FAIR assessments.
-\emph{Data Science Journal} 19.
-DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
-Belshe M, Peon R. 2015. \emph{Hypertext Transfer Protocol Version 2 (HTTP/2)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
-Berners-Lee T, Fielding R, Masinter L. 2005. \emph{Uniform Resource Identifier (URI): Generic Syntax}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
-Berners-Lee T, Fischetti M. 1999. \emph{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}. San Francisco: HarperSanFrancisco.
-
-\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
-Bernstein A, Hendler J, Noy N. 2016. A new look at the semantic web. \emph{Communications of the ACM} 59:35--37. DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
-Bioschemas - Bioschemas. \emph{Available at} \href{http://bioschemas.org/}{\emph{http://bioschemas.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
-Bizer C, Heath T, Berners-Lee T.
-2009.
-Linked data - the story so far.
-\emph{International journal on Semantic Web and information systems} 5:1--22.
-DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
-Blanchi C, Broeder D, Jejkal T, Sharif I, Schlemmer A, Uytvanck D van, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}{FDO -- upload of FDO}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
-Blanchi C, Hellström M, Lannom L, Schwardmann U, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}{Implementation of attributes, types, profiles and registries}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
-Bonino Da Silva Santos LO, Wilkinson MD, Kuzniar A, Kaliyaperumal R, Thompson M, Dumontier M, Burger K.
-2016.
-\href{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}{FAIR data points supporting big data interoperability}.
-In: Zelm M, Doumeingts G, Mendonça JP eds.
-\emph{Enterprise interoperability in the digitized and networked factory of the future}.
-iSTE Press, 270--279.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
-Bonino L, Wittenburg O, Carroll B, Hardisty A, Leggott M, Zwölf C.
-2019.
-\emph{\href{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}{FAIR digital object framework}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
-Carriero VA, Daquino M, Gangemi A, Nuzzolese AG, Peroni S, Presutti V, Tomasi F.
-2020.
-The landscape of ontology reuse approaches.
-In: Cota G, Daquino M, Pozzato GL eds.
-\emph{Applications and practices in ontology design, extraction, and reasoning}.
-Studies on the semantic web.
-IOS Press,.
-DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/SSW200033}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
-Ciccarese P, Soiland-Reyes S, Belhajjame K, Gray AJ, Goble C, Clark T.
-2013.
-PAV ontology: provenance, authoring and versioning.
-\emph{Journal of Biomedical Semantics} 4:37.
-DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
-Clemm G, Amsden J, Ellison T, Kaler C, Whitehead J.
-2002.
-\emph{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}.
-RFC Editor.
-DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
-Content negotiation - HTTP \textbar{} MDN.
-\emph{Available at} \href{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}{\emph{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content\_negotiation}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
-Corcho O, Eriksson M, Kurowski K, Ojsteršek M, Choirat C, Sanden M van de, Coppens F.
-2021.
-\emph{EOSC interoperability framework}.
-Publications Office of the EU.
-DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
-Data - W3C. \emph{Available at} \href{https://www.w3.org/standards/semanticweb/data}{\emph{https://www.w3.org/standards/semanticweb/data}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2. \emph{Available at} \href{https://www.w3.org/TR/vocab-dcat-2/}{\emph{https://www.w3.org/TR/vocab-dcat-2/}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
-Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
-Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
-DCMI Metadata Terms.
-2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
-DCMI Metadata Terms.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
-Delgado JCM.
-2016.
-An Interoperability Framework and Distributed Platform for Fast Data Applications.
-In: \emph{Data Science and Big Data Computing}.
-Springer International Publishing, 3--39.
-DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
-Designing a Linked Data developer experience.
-2018.
-\emph{Available at} \href{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}{\emph{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
-Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation. \emph{Available at} \href{https://www.dona.net/doipv1doc}{\emph{https://www.dona.net/doipv1doc}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
-DOIP and Examples --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip.html}{\emph{https://www.cordra.org/documentation/api/doip.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
-DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
-draft-bhutton-json-schema-00.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}{\emph{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
-draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}{\emph{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
-draft-ietf-quic-http-34.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}{\emph{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
-draft-kelly-json-hal-08.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}{\emph{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
-Duerst M, Suignard M. 2005. \emph{Internationalized Resource Identifiers (IRIs)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
-Dusseault L (ed.). 2007. \emph{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
-FAIR Digital Object Framework Documentation. \emph{Available at} \href{https://fairdigitalobjectframework.org/}{\emph{https://fairdigitalobjectframework.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
-FAIR Digital Objects Forum \textbar. \emph{Available at} \href{https://fairdo.org/}{\emph{https://fairdo.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
-Fensel D, Facca FM, Simperl E, Toma I. 2011. \emph{Semantic Web Services}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
-Fielding RT.
-2000.
-\href{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}{Architectural styles and the design of network-based software architectures}.
-\{THESIS\}.\{DOCTORAL\} Thesis.
-University of California, Irvine.
-
-\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
-Fielding R, Nottingham M, Reschke J (eds.). 2022. \emph{HTTP Semantics}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
-Fielding R, Reschke J (eds.). 2014b. \emph{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
-Fielding R, Reschke J (eds.). 2014a. \emph{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
-Fielding RT, Taylor RN, Erenkrantz JR, Gorlick MM, Whitehead J, Khare R, Oreizy P.
-2017.
-Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award).
-In: \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017}.
-New York, New York, USA: ACM Press, 4--14.
-DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
-FOAF Vocabulary Specification. \emph{Available at} \href{http://xmlns.com/foaf/spec/}{\emph{http://xmlns.com/foaf/spec/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
-Foundation D. 2018. \emph{\href{https://hdl.handle.net/0.DOIP/DOIPV2.0}{Digital object interface protocol specification, version 2.0}}. DONA foundation.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
-Gayo JEL, Prud'hommeaux E, Boneva I, Kontokostas D.
-2017.
-Validating RDF Data.
-\emph{Synthesis Lectures on the Semantic Web: Theory and Technology} 7:1--328.
-DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
-Goble C, Stevens R.
-2008.
-State of the nation in data integration for bioinformatics.
-\emph{Journal of Biomedical Informatics} 41:687--693.
-DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
-Gregorio J, de hOra B (eds.). 2007. \emph{The Atom Publishing Protocol}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
-Gregorio J, Fielding R, Hadley M, Nottingham M, Orchard D. 2012. \emph{URI Template}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
-Groth P, Loizou A, Gray AJG, Goble C, Harland L, Pettifer S.
-2014.
-API-centric Linked Data integration: The Open PHACTS Discovery Platform case study.
-\emph{Journal of Web Semantics} 29:12--18.
-DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
-Group RDAFDMMW. 2020. FAIR data maturity model: Specification and guidelines. \emph{Research Data Alliance}. DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
-Hardisty A, Brack P, Goble C, Livermore L, Scott B, Groom Q, Owen S, Soiland-Reyes S. 2022. The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
-Digital Object Approach to Speeding up Digital Mobilisation of Natural History
-Collections. \emph{Data Intelligence} 4:320--341. DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
-Hasnain A, Rebholz-Schuhmann D.
-2018.
-Assessing FAIR data principles against the 5-star open data principles.
-In: Gangemi A, Gentile AL, Nuzzolese AG, Rudolph S, Maleshkova M, Paulheim H, Pan JZ, Alam M eds.
-\emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers}.
-Lecture notes in computer science.
-Cham: Springer International Publishing, 469--477.
-DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
-Horrocks I, Hendler J (eds.). 2002. \emph{The Semantic Web --- ISWC 2002}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
-HTML Standard. \emph{Available at} \href{https://html.spec.whatwg.org/multipage/microdata.html}{\emph{https://html.spec.whatwg.org/multipage/microdata.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
-Hu W, Chen J, Zhang H, Qu Y.
-2011.
-How matchable are four thousand ontologies on the semantic web.
-In: Antoniou G, Grobelnik M, Simperl E, Parsia B, Plexousakis D, De Leenheer P, Pan J eds.
-\emph{The semantic web: Research and applications}.
-Lecture notes in computer science.
-Berlin, Heidelberg: Springer Berlin Heidelberg, 290--304.
-DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI}{\emph{https://www.w3.org/Provider/Style/URI}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI.html}{\emph{https://www.w3.org/Provider/Style/URI.html}} (accessed 26 May 2022b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
-Iyengar J, Thomson M (eds.). 2021. \emph{QUIC: A UDP-Based Multiplexed and Secure Transport}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
-JSON-LD 1.1. \emph{Available at} \href{http://www.w3.org/TR/json-ld/}{\emph{http://www.w3.org/TR/json-ld/}} (accessed 26 May 2022b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
-JSON-LD 1.1. \emph{Available at} \href{https://www.w3.org/TR/json-ld/}{\emph{https://www.w3.org/TR/json-ld/}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
-Juty N, Le Novere N, Laibe C.
-2011.
-Identifiers.org and MIRIAM Registry: community resources to provide persistent identification.
-\emph{Nucleic Acids Research} 40:D580--D586.
-DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
-Kahn R, Wilensky R. 1995. \emph{\href{http://www.cnri.reston.va.us/k-w.html}{A framework for distributed digital object services}}. CNRI.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
-Kahn R, Wilensky R.
-2006.
-A framework for distributed digital object services.
-\emph{International Journal on Digital Libraries} 6:115--123.
-DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
-Kamdar MR, Tudorache T, Musen MA.
-2017.
-A systematic analysis of term reuse and term overlap across biomedical ontologies.
-\emph{Semantic Web} 8:853--871.
-DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
-Khare R, Lawrence S. 2000. \emph{Upgrading to TLS Within HTTP/1.1}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
-Klein M, Van de Sompel H, Sanderson R, Shankar H, Balakireva L, Zhou K, Tobin R.
-2014.
-Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot.
-\emph{PLoS ONE} 9:e115253.
-DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
-Klímek J, Škoda P, Nečaský M. 2019. Survey of tools for linked data consumption. \emph{Satya Widya} 10:665--720. DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/SW-180316}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
-Lamprecht A-L, Palmblad M, Ison J, Schwämmle V, Al Manir MS, Altintas I, Baker CJO, Ben Hadj Amor A, Capella-Gutierrez S, Charonyktakis P, Crusoe MR, Gil Y, Goble C, Griffin TJ, Groth P, Ienasescu H, Jagtap P, Kalaš M, Kasalica V, Khanteymoori A, Kuhn T, Mei H, Ménager H, Möller S, Richardson RA, Robert V, Soiland-Reyes S, Stevens R, Szaniszlo S, Verberne S, Verhoeven A, Wolstencroft K.
-2021.
-Perspectives on automated composition of workflows in the life sciences.
-\emph{F1000Research} 10:897.
-DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
-Lannom L, Peters-von Gehlen K, Anders I, Pfeil A, Schlemmer A, Trautt Z, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}{FDO configuration types}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
-Lannom L, Schwardmann U, Blanchi C, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}{Typing FAIR digital objects}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
-Linked Data - Design Issues. \emph{Available at} \href{https://www.w3.org/DesignIssues/LinkedData.html}{\emph{https://www.w3.org/DesignIssues/LinkedData.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
-Linked Data Notifications. \emph{Available at} \href{https://www.w3.org/TR/ldn/}{\emph{https://www.w3.org/TR/ldn/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
-Linked Data Platform Working Group. 2015. \emph{\href{https://www.w3.org/TR/2015/REC-ldp-20150226/}{Linked data platform 1.0}}. W3C.
-
-\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
-Loo T (ed.). 1970. First International Conference on FAIR Digital Objects. DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
-martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
-May W. 2017. \emph{HTTP Live Streaming}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
-Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
-2021b.
-Web data apis over SPARQL.
-In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
-Synthesis lectures on data, semantics, and knowledge.
-Cham: Springer International Publishing, 27--38.
-DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
-Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
-2021a.
-Conclusion and future challenges.
-In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
-Synthesis lectures on data, semantics, and knowledge.
-Cham: Springer International Publishing, 77--80.
-DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
-Mika P. 2015. On Schema.org and Why It Matters for the Web. \emph{IEEE Internet Computing} 19:52--55. DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
-Mons B, Neylon C, Velterop J, Dumontier M, Silva Santos LOB da, Wilkinson MD.
-2017.
-Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud.
-\emph{Information Services \& Use} 37:49--56.
-DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{ISU\}-170824}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
-NCBO BioPortal. \emph{Available at} \href{https://bioportal.bioontology.org/ontologies}{\emph{https://bioportal.bioontology.org/ontologies}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
-Neumann A, Laranjeiro N, Bernardino J.
-2021.
-An analysis of public REST web service apis.
-\emph{IEEE Transactions on Services Computing} 14:957--970.
-DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/TSC.2018.2847344}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
-Nottingham M. 2017. \emph{Web Linking}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
-OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More.
-\emph{Available at} \href{https://spec.openapis.org/oas/v3.1.0.html}{\emph{https://spec.openapis.org/oas/v3.1.0.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
-ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
-OWL 2 Web Ontology Language Document Overview (Second Edition). \emph{Available at} \href{http://www.w3.org/TR/owl2-overview/}{\emph{http://www.w3.org/TR/owl2-overview/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
-Page KR, De Roure DC, Martinez K.
-2011.
-REST and Linked Data.
-In: \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11}.
-ACM Press,.
-DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
-Polleres A, Kamdar MR, Fernández JD, Tudorache T, Musen MA.
-2020.
-A more decentralized vision for linked data.
-\emph{Satya Widya} 11:101--113.
-DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/SW-190380}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 January 2023a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
-RDF 1.1 Primer. \emph{Available at} \href{http://www.w3.org/TR/rdf11-primer/}{\emph{http://www.w3.org/TR/rdf11-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
-RDF Schema 1.1. \emph{Available at} \href{http://www.w3.org/TR/rdf-schema/}{\emph{http://www.w3.org/TR/rdf-schema/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
-RDFa 1.1 Primer - Third Edition. \emph{Available at} \href{http://www.w3.org/TR/rdfa-primer/}{\emph{http://www.w3.org/TR/rdfa-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
-Resource Description Framework (RDF) Model and Syntax Specification.
-\emph{Available at} \href{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}{\emph{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
-ResourceSync Framework Specification - Table of Contents. \emph{Available at} \href{http://www.openarchives.org/rs/toc}{\emph{http://www.openarchives.org/rs/toc}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
-Sandvine.Global Internet Phenomena Report 2022.
-\emph{Available at} \href{https://www.sandvine.com/global-internet-phenomena-report-2022}{\emph{https://www.sandvine.com/global-internet-phenomena-report-2022}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
-Sauermann L, Cyganiak R, Völkel M. 2011. \emph{Cool URIs for the semantic web}. Universität des Saarlandes. DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
-Schema.org - Schema.org. \emph{Available at} \href{https://schema.org/}{\emph{https://schema.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
-Schema.org Actions - schema.org. \emph{Available at} \href{https://schema.org/docs/actions.html}{\emph{https://schema.org/docs/actions.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
-Schultes E, Wittenburg P.
-2019.
-FAIR principles and digital objects: Accelerating convergence on a data infrastructure.
-In: Manolopoulos Y, Stupnikov S eds.
-\emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers}.
-Communications in computer and information science.
-Cham: Springer International Publishing, 3--16.
-DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
-Schwardmann U, Kálmán T.
-2022.
-Two Examples on How FDO Types can Support Machine and Human Readability.
-\emph{Research Ideas and Outcomes} 8.
-DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
-Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
-Shape Expressions (ShEx) 2.1 Primer. \emph{Available at} \href{http://shex.io/shex-primer/}{\emph{http://shex.io/shex-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
-Shapes Constraint Language (SHACL). \emph{Available at} \href{https://www.w3.org/TR/shacl/}{\emph{https://www.w3.org/TR/shacl/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
-SKOS Simple Knowledge Organization System Primer. \emph{Available at} \href{http://www.w3.org/TR/skos-primer/}{\emph{http://www.w3.org/TR/skos-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
-Smith B, Ashburner M, Rosse C, Bard J, Bug W, Ceusters W, Goldberg LJ, Eilbeck K, Ireland A, Mungall CJ, Leontis N, Rocca-Serra P, Ruttenberg A, Sansone S-A, Scheuermann RH, Shah N, Whetzel PL, Lewis S.
-2007.
-The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration.
-\emph{Nature Biotechnology} 25:1251--1255.
-DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
-SPARQL 1.1 Overview. \emph{Available at} \href{http://www.w3.org/TR/sparql11-overview/}{\emph{http://www.w3.org/TR/sparql11-overview/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
-Stallings W. 1990. \emph{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}. Carmel, IN, USA: Sams.
-
-\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
-Stanczyk SK. 1987. Process modelling for information system description. \emph{The Open University}. DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
-Stefi A.
-2015.
-Do Developers Make Unbiased Decisions?
-- The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components.
-DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
-Stefi A, Hess T.
-2015.
-To develop or to reuse?
-Two perspectives on external reuse in software projects.
-In: Fernandes JM, Machado RJ, Wnuk K eds.
-\emph{Software business}.
-Lecture notes in business information processing.
-Cham: Springer International Publishing, 192--206.
-DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
-Strawn G, Wittenburg P, Soiland-Reyes St, Peters K, Lannom L, Schwardmann U, Blanch Ch.
-2022.
-\emph{\href{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}{FDO Forum FDO requirement specifications}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
-Sun S, Lannom L, Boesch B. 2003. \emph{Handle System Overview}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
-Sun S, Reilly S, Lannom L, Petrone J. 2003. \emph{Handle System Protocol (ver 2.1) Specification}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
-SWORD 3.0 Specification. \emph{Available at} \href{https://swordapp.github.io/swordv3/swordv3.html}{\emph{https://swordapp.github.io/swordv3/swordv3.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
-The Open Graph protocol. \emph{Available at} \href{https://ogp.me/}{\emph{https://ogp.me/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
-Thornton K, Solbrig H, Stupp GS, Labra Gayo JE, Mietchen D, Prud E, Waagmeester A.
-2019.
-Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation.
-In: Hitzler P, Fernández M, Janowicz K, Zaveri A, Gray AJG, Lopez V, Haller A, Hammar K eds.
-\emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings}.
-Lecture notes in computer science.
-Cham: Springer International Publishing, 606--620.
-DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
-Tirmizi S, Aitken S, Moreira DA, Mungall C, Sequeda J, Shah NH, Miranker DP.
-2011.
-Mapping between the OBO and OWL ontology languages.
-\emph{Journal of Biomedical Semantics} 2:S3.
-DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
-Turcoane O.
-2014.
-Linked data, JSON-LD and the semantics of cultural and scientific heritage.
-\emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} 4:95--105.
-DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
-Usage Statistics of JSON-LD for Websites, May 2022.
-\emph{Available at} \href{https://w3techs.com/technologies/details/da-jsonld}{\emph{https://w3techs.com/technologies/details/da-jsonld}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
-Van de Sompel H, Klein M, Jones S, Nelson ML, Warner S, Devaraju A, Huber R, Steinhoff W, Tykhonov V, Boruta L, Meijers E, Soiland-Reyes S, Wilkinson M.
-2022.FAIR Signposting Profile.
-\emph{Available at} \href{https://signposting.org/FAIR/}{\emph{https://signposting.org/FAIR/}} (accessed 5 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
-Van de Sompel H, Nelson M, Sanderson R. 2013. \emph{HTTP Framework for Time-Based Access to Resource States -\/- Memento}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
-Verborgh R, Vander Sande M.
-2020.
-The semantic web identity crisis: In search of the trivialities that never were.
-\emph{Satya Widya} 11:19--27.
-DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/SW-190372}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
-W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures. \emph{Available at} \href{http://www.w3.org/TR/xmlschema11-1/}{\emph{http://www.w3.org/TR/xmlschema11-1/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
-Web Services Description Language (WSDL) Version 2.0 Part 0: Primer. \emph{Available at} \href{http://www.w3.org/TR/wsdl20-primer/}{\emph{http://www.w3.org/TR/wsdl20-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
-WebSockets Standard. \emph{Available at} \href{https://websockets.spec.whatwg.org/}{\emph{https://websockets.spec.whatwg.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
-Weigel T, Plale B, Parsons M, Zhou G, Luo Y, Schwardmann U, Quick R, Hellström M, Kurakawa K.
-2018.
-RDA Recommendation on PID Kernel Information.
-\emph{Research Data Alliance}.
-DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
-Weiland C, Islam S, Broder D, Anders I, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}{FDO machine actionability}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
-Weiland C, Schwardmann U, Wittenburg P, Kirkpatrick C, Hanisch R, Trautt Z.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}{FDO Forum Document Standards}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
-Wieczorek J, Bloom D, Guralnick R, Blum S, Döring M, Giovanni R, Robertson T, Vieglais D.
-2012.
-Darwin Core: An Evolving Community-Developed Biodiversity Data Standard.
-\emph{PLoS ONE} 7:e29715.
-DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
-Wilde E. 2013. \emph{The 'profile' Link Relation Type}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-6DjakjNS}{}}%
-Wilkinson MD, Dumontier M, Aalbersberg IjJ, Appleton G, Axton M, Baak A, Blomberg N, Boiten J-W, da Silva Santos LB, Bourne PE, Bouwman J, Brookes AJ, Clark T, Crosas M, Dillo I, Dumon O, Edmunds S, Evelo CT, Finkers R, Gonzalez-Beltran A, Gray AJG, Groth P, Goble C, Grethe JS, Heringa J, 't Hoen PAC, Hooft R, Kuhn T, Kok R, Kok J, Lusher SJ, Martone ME, Mons A, Packer AL, Persson B, Rocca-Serra P, Roos M, van Schaik R, Sansone S-A, Schultes E, Sengstag T, Slater T, Strawn G, Swertz MA, Thompson M, van der Lei J, van Mulligen E, Velterop J, Waagmeester A, Wittenburg P, Wolstencroft K, Zhao J, Mons B.
-2016.
-The FAIR Guiding Principles for scientific data management and stewardship.
-\emph{Scientific Data} 3.
-DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pa67pUtR}{}}%
-Wilkinson SR, Eisenhauer G, Kapadia AJ, Knight K, Logan J, Widener P, Wolf M.
-2022.
-F*** workflows: when parts of FAIR are missing.
-\emph{arXiv}.
-DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
-Williams AJ, Harland L, Groth P, Pettifer S, Chichester C, Willighagen EL, Evelo CT, Blomberg N, Ecker G, Goble C, Mons B.
-2012.
-Open PHACTS: semantic interoperability for drug discovery.
-\emph{Drug Discovery Today} 17:1188--1198.
-DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
-Wittenburg P, Anders I, Blanchi C, Buurman M, Goble C, Grieb J, Hardisty A, Islam S, Jejkal T, Kálmán T, Kirkpatrick C, Lannom L, Lauer T, Manepalli G, Peters-von Gehlen K, Pfeil A, Quick R, Sanden M van de, Schwardmann U, Soiland-Reyes S, Stotzka R, Trautt Z, Van Uytvanck D, Weiland C, Wieder P.
-2022.
-FAIR digital object demonstrators 2021.
-\emph{Zenodo}.
-DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
-Wittenburg P, Strawn G, Mons B, Bonino L, Schultes E.
-2019.
-Digital objects as drivers towards convergence in data infrastructures.
-\emph{https://b2share.eudat.eu}.
-DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
-Wolstencroft K, Haines R, Fellows D, Williams A, Withers D, Owen S, Soiland-Reyes S, Dunlop I, Nenadic A, Fisher P, Bhagat J, Belhajjame K, Bacall F, Hardisty A, Nieva de la Hidalga A, Balcazar Vargas MP, Sufi S, Goble C.
-2013.
-The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud.
-\emph{Nucleic Acids Research} 41:W557--W561.
-DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
-Wolstencroft K, Owen S, Horridge M, Krebs O, Mueller W, Snoep JL, du Preez F, Goble C.
-2011.
-RightField: embedding ontology annotation in spreadsheets.
-\emph{Bioinformatics} 27:2021--2022.
-DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
-Working Group FDO-TSIG. 2022a. \emph{\href{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}{DOIP endorsement request}}. (internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
-Working Group FDO-TSIG.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}{FDO -- granularity, versioning, mutability}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Fi62cJAo}{}}%
-Working Group FDO-TSIG, Working Group FDO-BIG.
-2022.
-\emph{\href{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}{FDO -- kernel attributes \& metadata}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
-X.1255 : Framework for discovery of identity management information.
-\emph{Available at} \href{https://www.itu.int/rec/T-REC-X.1255-201309-I}{\emph{https://www.itu.int/rec/T-REC-X.1255-201309-I}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRzcHhPD}{}}%
-Zarras A. 2004. A Comparison Framework for Middleware Infrastructures. \emph{The Journal of Object Technology} 3:103. DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}.
-
-\end{CSLReferences}

### 2023-01-24 latex/table5.tex
converted to peerj overleaf template

--- 2023-01-24 manuscript.tex
@@
-This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/51048f35d5fce97aa4194e6426fdd7c704f19199/}{permalink})
-was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/51048f35d5fce97aa4194e6426fdd7c704f19199}{stain/2022-fdo-paper@51048f3}
-on January 24, 2023.
-
-\hypertarget{authors}{%
-\section*{Authors}\label{authors}}
-\addcontentsline{toc}{section}{Authors}
-
-\begin{itemize}
-\item
-  \textbf{Stian Soiland-Reyes}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/stain}{stain}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/soilandreyes}{soilandreyes}
-  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Carole Goble}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/carolegoble}{carolegoble}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
-  Department of Computer Science, The University of Manchester, UK
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Paul Groth}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/pgroth}{pgroth}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/pgroth}{pgroth}
-  Informatics Institute, Faculty of Science, University of Amsterdam, NL
-\end{itemize}
-
-\hypertarget{abstract}{%
-\section*{Abstract}\label{abstract}}
-\addcontentsline{toc}{section}{Abstract}
-
-\textbf{TODO}: Rewrite from notes below to actual abstract.
-
-FAIR Digital Object is an emerging concept from EOSC.
-This is important.
-Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
-Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
-
-What is the message of this paper?
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These are the overlaps
-\item
-  This is what FDO is requiring but not commonly deployed
-\item
-  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
-\item
-  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
-\item
-  Lessons for Semantic Web community. What is missing in FDO and in SW.
-\item
-  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
-\item
-  Go back to the old SW layer cake. Explains why we picked these frameworks.
-\item
-  Lessons for SW: Parts of the stack that is less. FDO contributions.
-\end{enumerate}
-
-Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.
-
-Have all the ingredients, but not cooking properly.
-
-Systematic through frameworks.
-
-Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?
-
-\hypertarget{sec:introduction}{%
-\section{Introduction}\label{sec:introduction}}
-
-The FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures {[}\protect\hyperlink{ref-u8Cy0psL}{2}{]}.
-The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
-
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
-The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
-
-In other discourse, Linked Data {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles {[}\protect\hyperlink{ref-GJHYN6xW}{6},\protect\hyperlink{ref-EJDjHrUv}{7}{]}.
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services {[}\protect\hyperlink{ref-99MS5xE0}{9},\protect\hyperlink{ref-J2H8yssV}{10}{]}.
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability {[}\protect\hyperlink{ref-DNTCUjpC}{11},\protect\hyperlink{ref-Pa67pUtR}{12}{]}.
-
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality {[}\protect\hyperlink{ref-uxWjmlio}{13},\protect\hyperlink{ref-v0WwFHgT}{14}{]} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves {[}\protect\hyperlink{ref-5V3GqSzR}{15}{]}.
-
-These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.
-
-Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.
-
-\hypertarget{sec:background}{%
-\section{Background and related work}\label{sec:background}}
-
-In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.
-
-\hypertarget{sec:fdo}{%
-\subsection{FAIR Digital Object}\label{sec:fdo}}
-
-\hypertarget{sec:next-step-fdo}{%
-\subsubsection{Next steps for FDO}\label{sec:next-step-fdo}}
-
-The FAIR Digital Object Forum {[}\protect\hyperlink{ref-DMEBhHID}{16}{]} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
-As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
-
-\textbf{FAIR Digital Object Overview and Specifications} {[}\protect\hyperlink{ref-11NpETLq1}{17}{]} is a comprehensive overview of FAIR Digital Object specifications listed below.
-It serves as a primer that introduces FDO concepts and the remaining documents.
-
-The \textbf{FDO Forum Document Standards} {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
-In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
-
-The \textbf{FDO Requirement Specifications} {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} is an update of {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} as the foundational definition of FDO.
-This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
-The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
-
-The \textbf{Machine actionability} {[}\protect\hyperlink{ref-iJeys0T5}{21}{]} sets out to define what is meant by \emph{machine actionability} for FDOs.
-\emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
-The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.
-
-\textbf{Configuration Types} {[}\protect\hyperlink{ref-SPd0uEqO}{22}{]} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}
-
-\textbf{PID Profiles \& Attributes} {[}\protect\hyperlink{ref-ljVViWCl}{24}{]} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}.
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
-
-\textbf{Kernel Attributes \& Metadata} {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information {[}\protect\hyperlink{ref-lCoshHXg}{25}{]}.
-It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
-
-\textbf{Granularity, Versioning, Mutability} {[}\protect\hyperlink{ref-16W9AycSW}{27}{]} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
-The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
-
-\textbf{DOIP Endorsement Request} {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} is an endorsement of the DOIP v2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} specification as a potential FDO implementation, as it has been applied by several institutions {[}\protect\hyperlink{ref-jRWurmQc}{30}{]}.
-The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
-
-\textbf{Upload of FDO} {[}\protect\hyperlink{ref-1H7cmmr69}{31}{]} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync {[}\protect\hyperlink{ref-eS5y9TRh}{32}{]} is suggested as one type of service to list FDOs.
-This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
-
-\textbf{Typing FAIR Digital Objects} {[}\protect\hyperlink{ref-ZFzPxCNB}{33}{]} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
-This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
-Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work.
-How to define such FDO operations is not specified.
-
-\textbf{Implementation of Attributes, Types, Profiles and Registries} {[}\protect\hyperlink{ref-126uxr5pI}{34}{]} details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
-This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} {[}\protect\hyperlink{ref-1CqIZ47pu}{35}{]}.
-
-It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-
-The concept of \textbf{FAIR Digital Objects} {[}\protect\hyperlink{ref-IHLT6hye}{4}{]} has been introduced as way to expose research data as active objects that conform to the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}.
-This builds on the \emph{Digital Object} (DO) concept {[}\protect\hyperlink{ref-11MnuwJ4l}{36}{]}, first introduced in 1995 {[}\protect\hyperlink{ref-3Uqe3fuK}{37}{]} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework {[}\protect\hyperlink{ref-103Hw8H43}{38}{]} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]}.
-
-In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}.
-The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing {[}\protect\hyperlink{ref-tz0P3DTC}{40}{]}.
-
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}.
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
-
-\hypertarget{fdo-approaches}{%
-\subsubsection{FDO approaches}\label{fdo-approaches}}
-
-FDO is an evolving concept. A set of FDO Demonstrators {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} highlight how current adapters are approaching implementations of FDO from different angles:
-
-\begin{itemize}
-\tightlist
-\item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, see {[}\protect\hyperlink{ref-xuQKRRx5}{41}{]}.} (usually TCP/IP over TLS).
-The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]} and PIDs are assigned using the Handle system.
-Several type registries have been established.
-\item
-  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
-\item
-  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
-\end{itemize}
-
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
-
-\hypertarget{sec:ld}{%
-\subsection{From the Semantic Web to Linked Data}\label{sec:ld}}
-
-In order to describe \emph{Linked Data} as it is used today, we'll start with an (opinionated) description of the evolution of its foundation, the \emph{Semantic Web}.
-
-\hypertarget{sec:semweb}{%
-\subsubsection{A brief history of the Semantic Web}\label{sec:semweb}}
-
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee {[}\protect\hyperlink{ref-Guy7cjVY}{43}{]}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
-The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
-The goal of Semantic Web was to further develop the machine-readable aspects of the Web, in particular adding \emph{meaning} (or semantics) to not just the link relations, but also to the \emph{resources} that the URLs identified, and for machines thus being able to meaningfully navigate across such resources, e.g.~to answer a particular query.
-
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) {[}\protect\hyperlink{ref-16WGmCih8}{44}{]} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format {[}\protect\hyperlink{ref-16wvFH4nh}{45}{]} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
-
-While triple-based knowledge representations were not new {[}\protect\hyperlink{ref-NxvhwSmG}{46}{]}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs {[}\protect\hyperlink{ref-RRtwPF7Y}{47}{]} are generalised forms of URLs that include locator-less identifiers
-  such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years {[}\protect\hyperlink{ref-nW4HY8Nq}{48}{]}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) {[}\protect\hyperlink{ref-HqZ3r6J3}{49}{]}.
-RDF 1.1 formats use Unicode to support IRIs {[}\protect\hyperlink{ref-rqjuXkEf}{50}{]}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
-
-The early days of the Semantic Web saw fairly lightweight approaches with the establishment of vocabularies such as FOAF (to describe people and their affiliations) and Dublin Core (for bibliographic data).
-Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term.
-The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource.
-By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs.
-The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.
-
-The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} {[}\protect\hyperlink{ref-yd2WE9eq}{52}{]} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
-
-Established Semantic Web services and standards include SPARQL {[}\protect\hyperlink{ref-xww8fb1B}{53}{]} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
-
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins {[}\protect\hyperlink{ref-bMSxa8XW}{54},\protect\hyperlink{ref-Qlrlr2TZ}{55}{]}.
-In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
-
-Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) {[}\protect\hyperlink{ref-FDDmgO6s}{57}{]} .
-
-The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS {[}\protect\hyperlink{ref-191ZkYZMt}{58}{]}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
-
-With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps {[}\protect\hyperlink{ref-9CdJnsoJ}{59}{]}.
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication {[}\protect\hyperlink{ref-AGHR9oyr}{60}{]}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data {[}\protect\hyperlink{ref-5SuwLNdE}{61}{]}.
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies {[}\protect\hyperlink{ref-TUmA4KQT}{62}{]} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS {[}\protect\hyperlink{ref-ZwAcGQKY}{63}{]}, OWL {[}\protect\hyperlink{ref-1p4IWJpI}{64}{]}, OBO {[}\protect\hyperlink{ref-1DalZuKct}{65}{]}, SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]}), hash vs slash, HTTP status codes and PID redirection strategies {[}\protect\hyperlink{ref-rnzJ4khD}{51}{]}.
-
-\hypertarget{sec:ld-web}{%
-\subsubsection{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
-
-The \textbf{Linked Data} concept {[}\protect\hyperlink{ref-19s3yyxDn}{5}{]} was kickstarted as a set of best practices {[}\protect\hyperlink{ref-OGXYtGZ8}{66}{]} to bring the Web aspect back into focus.
-Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
-This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
-
-Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph {[}\protect\hyperlink{ref-oN9D5LkD}{67}{]}, RDFa {[}\protect\hyperlink{ref-FKIF6ApB}{68}{]}, Microdata {[}\protect\hyperlink{ref-14Utscfe4}{69}{]}) we find JSON-LD {[}\protect\hyperlink{ref-X5SWPqSU}{70}{]} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites {[}\protect\hyperlink{ref-j3SqQLBR}{71}{]}.
-
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} {[}\protect\hyperlink{ref-BFMoe8FC}{72}{]} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL {[}\protect\hyperlink{ref-GDrm9dgX}{73}{]} or ShEx {[}\protect\hyperlink{ref-YUzogaV}{74}{]}) can be used to validate RDF Data {[}\protect\hyperlink{ref-S5BI6Mwx}{75},\protect\hyperlink{ref-ZeOAPNHa}{76}{]} before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts {[}\protect\hyperlink{ref-o596f2JM}{77}{]}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
-
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers {[}\protect\hyperlink{ref-jHwttlTs}{8}{]}.
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies {[}\protect\hyperlink{ref-MhsCt4P3}{78}{]}, the use of Linked Data as part of REST Web APIs {[}\protect\hyperlink{ref-eS6a6tlz}{79}{]} , and as shown by the big uptake by publishers to annotate the Web using schema.org {[}\protect\hyperlink{ref-117tX9MEG}{81}{]}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
-
-\hypertarget{sec:comparing}{%
-\section{Comparing FDO and existing approaches}\label{sec:comparing}}
-
-To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
-\item
-  The FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}, validated against its implementations for completeness.
-\item
-  A Comparison Framework for Middleware Infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
-\item
-  Cross-checks against RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-UzQhqk0M}{84}{]} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
-\item
-  EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
-\end{enumerate}
-
-The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).
-
-Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} against Linked Data in general.
-
-\hypertarget{sec:interoperability-compare}{%
-\subsection{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
-
-The Interoperability Framework for Fast Data Applications {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
-
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model {[}\protect\hyperlink{ref-8ZRGFa6y}{85}{]} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
-In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]}.
-\label{tbl:fdo-web-interoperability-framework}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
-FDO encourages research communities to develop common type registries to be shared across instances.
-In current DOIP practice, each service have their own types, attributes and operations.
-The wider symbiosis is consistent use of PIDs.
-& The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs {[}\protect\hyperlink{ref-174AwcFUL}{86}{]} end up being mandated centrally by dominant (often commercial) providers {[}\protect\hyperlink{ref-11VxBOeZy}{87}{]}, which clients are required to use as-is with special code per service.
-Use of Linked Data enables common tooling and semantic mapping across differences.
-\\
-\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP).
-\texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).
-Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}).
-& ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]}, schema.org Actions {[}\protect\hyperlink{ref-dKAekUjL}{89}{]}, WSDL/SOAP {[}\protect\hyperlink{ref-LmudecyN}{90}{]}, but frequently just as human-readable developer documentation/examples.
-\\
-\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing.
-Every FDO has a type.
-Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}.
-No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) {[}\protect\hyperlink{ref-jRWurmQc}{30}{]} & Lightweight HTTP semantics for authenticity/navigation.
-Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
-Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit.
-\\
-\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation} & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema.
-Multiple byte stream attachments of any media type.
-& Textual HTTP headers (including any signposting), single byte stream of any media type, e.g.~Linked Data formats (JSON-LD, Turtle, RDF/XML) or embedded in document (HTML with RDFa, JSON-LD or Microdata).
-XML was previously the main syntax used by APIs, JSON is now dominant.
-\\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
-\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
-Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable).
-Unique DOIP protocol can be hard to add to existing Web application frameworks, although proxy services have been developed (e.g.~B2SHARE adapter).
-& HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to (for static content) multi-cloud Content-Delivery Networks (CDN).
-Current scalable cloud technologies for Web hosting may not support HTTP features previously seen as important for Semantic Web, e.g.~content negotiation and semantic HTTP status codes.
-\\
-\bottomrule
-\end{longtable}
-
-Based on the analysis shown in table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
-
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services {[}\protect\hyperlink{ref-39K3yHT3}{91}{]}.
-However, this is mostly done via developer or human interaction {[}\protect\hyperlink{ref-iYh1wFBt}{92}{]}.
-Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.
-
-A difference between Web technologies and FDO is the stringency of the requirements for both syntax and semantics.
-Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON.
-On the semantic front, FDO requires that every object have a well-defined type and structured form.
-This is clearly not the case on the Web.
-
-In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed.
-These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles {[}\protect\hyperlink{ref-hxtgCGjY}{93}{]} to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference {[}\protect\hyperlink{ref-jrZe6Esu}{94}{]}.
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP {[}\protect\hyperlink{ref-p7nwRj8E}{95}{]}.
-
-It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web {[}\protect\hyperlink{ref-1H9iQhQYq}{96}{]} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail {[}\protect\hyperlink{ref-68fzID2k}{97}{]}.
-
-\hypertarget{mapping-of-metamodel-concepts}{%
-\subsubsection{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
-
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics {[}\protect\hyperlink{ref-8HJqcF1Q}{98}{]}.
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2923}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2769}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data {[}\protect\hyperlink{ref-KlCFFFyL}{82}{]} to equivalent concepts for FDO and Web.
-\label{tbl:metamodel-concepts}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endhead
-Resource & FDO/DO & Resource \\
-Service & DOIP service & Server/endpoint \\
-Transaction & (not supported) & Conditional requests, \texttt{409\ Conflict} \\
-Process & Extended operations & (primarily stateless), \texttt{100\ Continue}, \texttt{202\ Accepted} \\
-Operation & DOIP Operation & Method, query parameters \\
-Request & DOIP Request & Request \\
-Response & DOIP Response & Response \\
-Message & Segment, \texttt{requestId} & Message, Representation \\
-Channel & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures. & TCP/IP, TLS, UDP \\
-Protocol & DOIP 2.0, ++ & HTTP/1.1, HTTP/2, HTTP/3 \\
-Link & PID/Handle & URL \\
-\bottomrule
-\end{longtable}
-
-From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
-Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \ref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
-
-\hypertarget{sec:doip-fdo-compare}{%
-\subsection{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
-
-The FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} and using Linked Data Platform {[}\protect\hyperlink{ref-17OHlMRQA}{99}{]}, as proposed by {[}\protect\hyperlink{ref-7szz7dwO}{100}{]}.
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines {[}\protect\hyperlink{ref-yygVPoL0}{19},\protect\hyperlink{ref-RwvirqWg}{20}{]} against its current implementations as DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} and Linked Data Platform (LDP) {[}\protect\hyperlink{ref-7szz7dwO}{100}{]}, with suggestions for required additions.
-\label{tbl:fdo-checks}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endhead
-G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
-& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
-\texttt{http://} URIs replaced by \texttt{https://}.
-& Keep flexibility of RDF serialisation formats which may change more frequently \\
-G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
-DOIP relatively unknown, in effect only one implementation.
-& Further promote DOIP and justify its benefits.
-Build tutorials and OSI open source implementations.
-Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites {[}\protect\hyperlink{ref-j3SqQLBR}{71}{]}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
-Example-driven documentation.
-\\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
-\\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
-Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
-\\
-G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
-DOIP operations can use any transport protocol.
-& Document transport protocols as FDOs, recommend which transport to use.
-& URI as abstraction base.
-Does not specify PID requirements.
-& Give stronger deployment recommendations.
-\\
-G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
-Unclear when metadata field is a PID or plain text.
-& Make datatype of fields explicit to support navigation.
-& Machine-navigation through URIs via properties and types.
-Unclear when URI should be followed or is just identifier, but always distinct from plain text.
-& \\
-G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
-& Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
-& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
-Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]} \\
-G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
-No FDOF implementations.
-& Develop demonstrator of FDOF usage based on existing LDP server.
-\\
-G9: \emph{standard compliance} & Handle {[}\protect\hyperlink{ref-2gXY2Nwh}{101}{]}, DOIP {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}.
-FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements {[}\protect\hyperlink{ref-TQdku4YF}{18}{]} & HTTP, LDP.
-FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
-FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve {[}\protect\hyperlink{ref-ljVViWCl}{24}{]}.
-Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-& HTTP URLs as basis for identifiers, but they are frequently not persistent.
-& Add strong guidance for PID services like w3id and persistence policies.
-\\
-FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
-Datacite DOIs content-negotiate to give registered metadata.
-& Add FAIR Signposting at PID provider for minimal PID record \\
-FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
-No method/attribute defined for separate metadata, only directly in PID Record.
-Unclear meaning of multiple items and bytestream chunks.
-& Clarify expectations for multiple items.
-Recommend chunks to not be used.
-& URIs resolvable by default.
-Multiple ways to resolve metadata, unclear preference.
-& Add FAIR Signposting and preference order.
-\\
-FDOF4: \emph{Additional attributes} & Freetext attribute keys.
-Attributes should be defined for FDO type (?).
-& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
-Explicitly allow attributes not already defined in type.
-& All attributes individually identified.
-Any Linked Data attributes can be used by URI or with mapped prefix.
-& Clarify type expectations of required/recommended/optional attributes.
-\\
-FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
-& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-Clarify that operations can be mapped to protocol directly.
-& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
-Unclear how to provide PID for additional operations.
-& Specify how additional operations should be called over HTTP.
-\\
-FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns {[}\protect\hyperlink{ref-i9Ang0rM}{102}{]} are used instead.
-& Document how operation resources can be discovered from an LPD container.
-Document search API.
-\\
-FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
-PR-TypingFDOs leaves this open.
-& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
-Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
-& Always make LDP types explicit in FDO profile.
-\\
-FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
-Separate Metadata FDO need custom property.
-& Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
-& Assertions are always with semantics, using RDF vocabularies.
-Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
-& Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
-FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
-Metadata would need to be bundled within custom FDO types/attributes.
-Unclear how levels are separated within single FDO representation (need FDOF8?).
-& Declare which metadata are expected within response attribute or within FDO object.
-Require PIDs for custom attributes.
-Define how alternate metadata levels can be represented separately.
-& Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
-& Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
-FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
-& Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} {[}\protect\hyperlink{ref-CNP39tah}{103}{]} & Rather document better how individual ad-hoc schemas can be started for prototypes.
-\\
-FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
-Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-{[}\protect\hyperlink{ref-6LvydTsC}{104}{]}).
-& & LDP Containers required by specification, also user-created (eg.
-\texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 {[}\protect\hyperlink{ref-1EW4hLN8U}{105}{]}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE {[}\protect\hyperlink{ref-gtGuSevK}{106}{]} \\
-FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
-\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
-No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
-\bottomrule
-\end{longtable}
-
-Note that the draft update to FDO specification {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer {[}\protect\hyperlink{ref-yygVPoL0}{19}{]} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
-
-A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines.
-Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement {[}\protect\hyperlink{ref-15yGpJ0wh}{28}{]} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
-
-From our evaluation, we can observe:
-
-\begin{itemize}
-\tightlist
-\item
-  G1 and G2 call for stability and trustworthiness.
-While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
-\item
-  Machine actionability (G4, G6) is a core feature of both FDOs and Linked Data.
-Conceptually they differ in the which way types and operations are discovered, with FDO seemingly more rigorous.
-In practice, however, we see that DOIP also relies on dynamic discovery of operations and that operation expectations for types (FDOF7) have not yet been defined.
-\item
-  FDO proposes that types can have additional operations beyond CRUD (FDOF5, FDOF6), while Linked Data mainly achieves this with RESTful patterns using CRUD on additional resources, e.g.~\texttt{order/152/items}.
-These are mainly stylistics but affect the architectural view -- FDOs have more of an object-oriented approach.
-\item
-  FDO puts strong emphasis on the use of PIDs (FDOF1, FDOF2, FDOF3, FDOF5), but in current practice DOIP use local types, local extended operations (FDOF5) and attributes (FDOF4) that are not bound to any global namespace.
-\item
-  Linked Data have a strong emphasis on semantics (FDOF8), and metadata schemas developed by community agreements (FDOF10). FDO types need to be made reusable across servers.
-\item
-  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} {[}\protect\hyperlink{ref-9T4j3N4e}{107}{]}
-\item
-  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
-\item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
-\item
-  DOIP supports multiple chunks of data for an object (FDOF3), while Linked Data can support content-negotiation.
-In either case it can be unclear to clients what is the meaning or equivalence of any additional chunks.
-\end{itemize}
-
-\hypertarget{sec:middleware}{%
-\subsection{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
-
-In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
-As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.
-
-We argue that the Web, with its now ubiquitous use of REST API {[}\protect\hyperlink{ref-174AwcFUL}{86}{]}, can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services {[}\protect\hyperlink{ref-WzPECPDP}{108}{]} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} {[}\protect\hyperlink{ref-11dInoyVF}{109}{]}.
-
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}) and Web technologies (using Linked Data) as middleware infrastructures {[}\protect\hyperlink{ref-hRzcHhPD}{83}{]}
-\label{tbl:fdo-web-middleware}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
-Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
-PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}.
-& The Web is inherently open and made by cross-linked URLs.
-Participation requires DNS domain purchase (many free alternatives also exists).
-PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects.
-\\
-\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
-HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
-\\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second {[}\protect\hyperlink{ref-18P9txgeB}{110}{]}.
-DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
-Multiple connections possible.
-Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks {[}\protect\hyperlink{ref-F8VZ86hu}{111}{]}.
-HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams {[}\protect\hyperlink{ref-KbbW0kGT}{112}{]} but still has TCP+TLS startup costs.
-For reduced latency {[}\protect\hyperlink{ref-CJuBupCW}{113}{]}, HTTP/3 {[}\protect\hyperlink{ref-Amit86Tp}{114}{]} use QUIC {[}\protect\hyperlink{ref-1GWvyEmUf}{115}{]}) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
-\\
-\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
-FDOs should provide the mandatory kernel metadata fields.
-FDOs of the same declared type typically share additional attributes (although that schema may not be declared).
-DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions.
-& Each URL accessed separately.
-Common HTTP headers provide basic metadata, although it is often not reliable.
-A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates {[}\protect\hyperlink{ref-11hxwwuRt}{116}{]}, content-negotiation {[}\protect\hyperlink{ref-n2f3TfoI}{117}{]} or FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]}.
-\\
-\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
-Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
-& Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
-\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
-Multiple locations possible in Handle system, can expose geo-info.
-& PIDs and URL redirects.
-DNS aliases and IP routing can hide location.
-Geo-localised servers common for large cloud deployments.
-\\
-\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
-FDO kernel metadata can include checksum and date.
-& HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
-Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
-\\
-\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
-Custom codes uncommon.
-\\
-\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
-No redirection from DOIP service.
-& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
-\\
-\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
-There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged {[}\protect\hyperlink{ref-rbG9uRKw}{118}{]}.
-Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
-An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento {[}\protect\hyperlink{ref-DNjcMi4a}{119}{]} expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method {[}\protect\hyperlink{ref-hRUsaTiz}{120}{]} (used by SVN).
-\\
-\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
-Internal synchronisation possible in backend for Extended operations.
-& Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} {[}\protect\hyperlink{ref-xvMAyRAc}{121}{]} with \texttt{LOCK} and \texttt{UNLOCK} methods.
-\\
-\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
-In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
-& The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space.
-In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend.
-\\
-\textbf{Encapsulation}: \emph{separate interface from implementation.
-Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
-FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
-Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade {[}\protect\hyperlink{ref-1B16WfR7q}{122}{]} to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns {[}\protect\hyperlink{ref-iME5l1kk}{123}{]} and PIDs via Persistent URLs (PURL).
-\\
-\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
-FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]}) & Syntactically Media Type with multiple suffixes {[}\protect\hyperlink{ref-nlH2rm1s}{124}{]} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) {[}\protect\hyperlink{ref-nCzRO0pK}{125}{]}.
-In metadata, semantic type systems (RDFS {[}\protect\hyperlink{ref-ZwAcGQKY}{63}{]}, OWL2 {[}\protect\hyperlink{ref-1p4IWJpI}{64}{]}, SKOS {[}\protect\hyperlink{ref-15gQDya5B}{56}{]}).
-OpenAPI 3 {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]} \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} {[}\protect\hyperlink{ref-11oVTMsuZ}{126}{]}, JSON \texttt{\$schema} {[}\protect\hyperlink{ref-15EZ2D0Rm}{42}{]}), JSON-LD \texttt{@context} {[}\protect\hyperlink{ref-by6CfOmv}{127}{]}.
-Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
-\\
-\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} {[}\protect\hyperlink{ref-KbbW0kGT}{112}{]}, Web Sockets {[}\protect\hyperlink{ref-qgqADk45}{128}{]}, Linked Data Notifications {[}\protect\hyperlink{ref-zlGYiuWC}{129}{]}, AtomPub {[}\protect\hyperlink{ref-IbaebabD}{130}{]}, SWORD {[}\protect\hyperlink{ref-7DGllKzG}{131}{]}, Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} {[}\protect\hyperlink{ref-HE7Ikwwl}{132}{]}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates {[}\protect\hyperlink{ref-11hxwwuRt}{116}{]}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} {[}\protect\hyperlink{ref-k0AfCGzw}{88}{]}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State {[}\protect\hyperlink{ref-174AwcFUL}{86}{]}, an important element of the REST architectural style.} incl.~Hydra {[}\protect\hyperlink{ref-13PaKBrZe}{133}{]}, schema.org Actions {[}{[}\protect\hyperlink{ref-dKAekUjL}{89}{]}), JSON HAL {[}\protect\hyperlink{ref-nyrPYHoP}{134}{]} \& Link headers (RFC8288) {[}\protect\hyperlink{ref-ozdWB3O7}{135}{]} \\
-\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 {[}\protect\hyperlink{ref-tIkSXrb5}{136}{]} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) {[}\protect\hyperlink{ref-PL512B0X}{137}{]}, MPEG-DASH (ISO/IEC 23009-1:2019) {[}\protect\hyperlink{ref-ikO7e0mh}{138}{]} \\
-\bottomrule
-\end{longtable}
-
-Based on the analysis in Table \ref{tbl:fdo-web-middleware}, we make the following observations:
-
-\begin{itemize}
-\tightlist
-\item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP {[}\protect\hyperlink{ref-16uB3jxpa}{39}{]} supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}.
-Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
-It is unclear what is the purpose of splitting a binary into chunks on a channel which no longer can be multiplexed and the only property of a chunk is its size\footnote{Although it is possible with \texttt{0.DOIP/Op.Retrieve} to request only particular individual elements of an DO (e.g.~one file), unlike HTTP's \texttt{Range} request, it is not possible to select individual chunks of an element's bytestream.}.
-\item
-  HTTP has strong support for scalability and caching, but this mostly assumes read-operations from static resources.
-FDO has no view on immutability or validity of retrieved objects, but this should be taken into consideration to support large-scale usage.
-\item
-  HTTP optimisations for performance (e.g.~HTTP/2, multiplexing) is largely used for commercial media distribution (e.g.~Netflix), and not commonly used by providers of FAIR data
-\item
-  Cloud deployment of Web applications give many middleware benefits (Scalability, Distribution, Access transparancy, Location transparancy) -- it is unclear how DOIP as a custom protocol would perform in a cloud setting as most of this infrastructure assumes HTTP as the protocol.
-\item
-  Programmatically the Web is rather unstructured as middleware, as there are many implementation choices.
-Usually it is undeclared what to expect for a given URI/service, and programmers follow documented examples for a particular service rather than automated programmatic exploration across providers.
-This mean one can consider the Web as an ecosystem of smaller middlewares with commonalities.
-\item
-  Many providers of FAIR Linked Data also provide programmatic REST API endpoints, e.g.~\href{https://www.uniprot.org/help/programmatic_access}{UNIPROT}, \href{https://chembl.gitbook.io/chembl-interface-documentation/web-services}{ChEMBL}, but keeping the FAIR aspects such as retrieving metadata in such a scenario may require combining different services using multiple formats and identifier conventions.
-\end{itemize}
-
-\hypertarget{sec:fair-compare}{%
-\subsection{Assessing FDO against FAIR}\label{sec:fair-compare}}
-
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} {[}\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}{]}.
-
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform {[}\protect\hyperlink{ref-7szz7dwO}{100}{]} comply with the FAIR principles {[}\protect\hyperlink{ref-6DjakjNS}{1}{]}.
-Here we've used the RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-1GrKTFaK2}{139}{]} as it has decomposed the FAIR principles to a structured list of FAIR indicators {[}\protect\hyperlink{ref-UzQhqk0M}{84}{]}, importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model {[}\protect\hyperlink{ref-UzQhqk0M}{84},\protect\hyperlink{ref-1GrKTFaK2}{139}{]} (first 2 columns) against the FDO guidelines {[}\protect\hyperlink{ref-RwvirqWg}{20}{]}, FDO implemented with the protocol DOIPv2 {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]}, Linked Data Platform (LDP) {[}\protect\hyperlink{ref-7szz7dwO}{100}{]} and examples from Linked Data practices in general.
-(--- indicates \emph{Unspecified}, may be possible with additional conventions)
-\label{tbl:fair-data-maturity-model}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endhead
-RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
-RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
-Handle, DOI.
-& FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{[}\protect\hyperlink{ref-rbG9uRKw}{118}{]};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
-RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
-FAIR Signposting.
-\\
-RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
-RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
-Unclear how to link to additional metadata.
-& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
-Embedded in landing page (RDFa).
-\\
-RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
-& PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
-RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
-Handle registry only searchable by PID.
-& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-{[}\protect\hyperlink{ref-1Gsq23e51}{140}{]}).
-Indexing by search engines if exposing HTML w/schema.org.
-\\
-RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
-\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
-RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
-RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
-RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
-RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
-RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
-Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP {[}\protect\hyperlink{ref-8HJqcF1Q}{98}{]} & HTTP/HTTPS, FTP (now less common), GridFTP {[}\protect\hyperlink{ref-18LsXvKGJ}{141}{]} (for large data), ARK {[}\protect\hyperlink{ref-voDQ6diM}{142}{]} \\
-RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol {[}\protect\hyperlink{ref-I9Kx0Hjy}{143}{]}.
-One server implementation {[}\protect\hyperlink{ref-1CvYRVhy2}{144}{]}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved {[}\protect\hyperlink{ref-xVJMmGVr}{145}{]} as an open source license -- it includes usage restrictions and requires Service Agreements.
-It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
-One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
-\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
-& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
-\\
-RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
-RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
-RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
-RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
-RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
-RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
-RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
-RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
-& FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) {[}\protect\hyperlink{ref-1BlxdxOwa}{146}{]} \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs {[}\protect\hyperlink{ref-o4h92LpS}{147}{]}), \texttt{rdfs:seeAlso}, FAIR signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]} \texttt{describedby} \\
-RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
-RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
-RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles {[}\protect\hyperlink{ref-QpDmajc4}{148}{]}, \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role {[}\protect\hyperlink{ref-10HA0mMBx}{149}{]}, \texttt{prov:specializationOf} {[}\protect\hyperlink{ref-ZUgwvAHE}{150}{]}, OAI-ORE Proxy {[}\protect\hyperlink{ref-ID1pTGry}{151}{]}) \\
-RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} but not assigned PIDs yet.
-& Unspecified.
-Multiple metadata records can allow multiple semantic profiles.
-& Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org {[}\protect\hyperlink{ref-10pzKCEUE}{152}{]}, Dublin Core Terms {[}\protect\hyperlink{ref-wAbZpOsr}{153}{]}, DCAT {[}\protect\hyperlink{ref-lnH0hEuh}{154}{]}, FOAF {[}\protect\hyperlink{ref-XmUY0cZB}{155}{]} \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata {[}\protect\hyperlink{ref-Fi62cJAo}{26}{]} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} {[}\protect\hyperlink{ref-lnH0hEuh}{154}{]} \\
-RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
-RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
-RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
-PID Kernel attributes?
-Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O {[}\protect\hyperlink{ref-9T4j3N4e}{107}{]}, PAV {[}\protect\hyperlink{ref-1E4OfHyhI}{156}{]}, Dublin Core Terms {[}\protect\hyperlink{ref-hKblceo6}{157}{]} \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen {[}\protect\hyperlink{ref-zZTtJruI}{158}{]}) & --- & Common, e.g.~DCAT 2 {[}\protect\hyperlink{ref-zt8AAjcZ}{159}{]}, BioSchemas {[}\protect\hyperlink{ref-1F8S9FeHf}{160}{]} \\
-RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal {[}\protect\hyperlink{ref-1Gsq23e51}{140}{]}, Darwin Core {[}\protect\hyperlink{ref-9jaVzCbL}{161}{]} \\
-RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
-\bottomrule
-\end{longtable}
-
-From this evaluation we observe:
-
-\begin{itemize}
-\tightlist
-\item
-  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
-\item
-  FDO/DOIP are stronger on identifier indicators
-\item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification {[}\protect\hyperlink{ref-13TcbsZF6}{29}{]} has currently only a couple of implementations and is expressed informally.
-The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
-\item
-  RDA-A1-02M and RDA-A1-02D highlights access by manual intervention, which is common for http/https URIs, but also using above PID resolvers for DOIP implementation \href{https://www.cordra.org/}{CORDRA} (e.g.~\url{https://hdl.handle.net/21.14100/90ec1c7b-6f5e-4e12-9137-0cedd16d1bce}), yet neither LDP, FDO nor DOIP specifications recommends human-readable representations to be provided
-\item
-  Neither DOIP nor LDP require license to be expressed (RDA-R1.1-01M, RDA-R1.1-02M, RDA-R1.1-03M), yet this is crucial for re-use and machine actionability of FAIR data and metadata to be legal
-\item
-  Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-{[}\protect\hyperlink{ref-126uxr5pI}{34}{]} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
-Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
-\item
-  Indicators for FAIR data are weak for either approach, as too much reliance is put on metadata.
-For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting {[}\protect\hyperlink{ref-snykkm7R}{23}{]} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
-In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
-\item
-  Resolving FDOs via Handle PIDs to the corresponding DOIP server is currently undefined by FDO and DOIP specifications.
-\texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known.
-g
-\end{itemize}
-
-\hypertarget{eosc-interoperability-framework}{%
-\section{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
-
-\textbf{TODO}: Introduce EOSC IF
-
-The EOSC Interoperability Framework {[}\protect\hyperlink{ref-aCye3KpE}{3}{]} in section 3.6 recommends:
-
-\begin{tablenos:no-prefix-table-caption}
-
 \begin{longtable}[]{@{}
@@
 \end{longtable}
-
-\end{tablenos:no-prefix-table-caption}
-
-Observations:
-
-\begin{itemize}
-\tightlist
-\item
-  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
-\item
-  Technical aspects highlighted by EOSC IF
-\item
-  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data.
-There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
-\item
-  FDO principles have strong recommendations for community development of organisational aspects.
-\item
-  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
-\end{itemize}
-
-\hypertarget{sec:discussion}{%
-\section{Discussion}\label{sec:discussion}}
-
-\textbf{TODO}
-
-\begin{itemize}
-\tightlist
-\item
-  Ramnifications of ide
-\item
-  Finish tables
-\item
-  Bullet points per table
-\item
-  Send in Google Docs
-\end{itemize}
-
-\hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
-
-The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake {[}\protect\hyperlink{ref-kYjzRJ0H}{162}{]}.
-However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
-
-While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
-
-We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
-
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
-
-\ldots{}
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{0}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-6DjakjNS}{}}%
-\CSLLeftMargin{1. }
-\CSLRightInline{\textbf{The FAIR Guiding Principles for scientific data management and stewardship}
-\CSLBlock{Mark D Wilkinson, Michel Dumontier, IJsbrand Jan Aalbersberg, Gabrielle Appleton, Myles Axton, Arie Baak, Niklas Blomberg, Jan-Willem Boiten, Luiz Bonino da Silva Santos, Philip E Bourne, \ldots{} Barend Mons} \emph{Scientific Data} (2016-03-15)
-\CSLBlock{DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/26978244}{26978244} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4792175}{PMC4792175}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
-\CSLLeftMargin{2. }
-\CSLRightInline{\textbf{Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud}
-\CSLBlock{Barend Mons, Cameron Neylon, Jan Velterop, Michel Dumontier, Luiz Olavo Bonino da Silva Santos, Mark D Wilkinson} \emph{Information Services \& Use} (2017-03-07)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{isu\}-170824}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
-\CSLLeftMargin{3. }
-\CSLRightInline{\textbf{EOSC interoperability framework}
-\CSLBlock{Oscar Corcho, Magnus Eriksson, Krzysztof Kurowski, Milan Ojsteršek, Christine Choirat, Mark van de Sanden, Frederik Coppens} \emph{Publications Office of the EU} (2021-02-05) \url{https://doi.org/10.2777/620649}
-\CSLBlock{DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
-\CSLLeftMargin{4. }
-\CSLRightInline{\textbf{FAIR principles and digital objects: Accelerating convergence on a data infrastructure}
-\CSLBlock{Erik Schultes, Peter Wittenburg} \emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers} (2019)
-\CSLBlock{DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}} · ISBN: 978-3-030-23583-3}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
-\CSLLeftMargin{5. }
-\CSLRightInline{\textbf{Linked data - the story so far}
-\CSLBlock{Christian Bizer, Tom Heath, Tim Berners-Lee} \emph{International journal on Semantic Web and information systems} (2009-07) \url{http://services.igi-global.com/resolvedoi/resolve.aspx?doi=10.4018/jswis.2009081901}
-\CSLBlock{DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
-\CSLLeftMargin{6. }
-\CSLRightInline{\textbf{Assessing FAIR data principles against the 5-star open data principles}
-\CSLBlock{Ali Hasnain, Dietrich Rebholz-Schuhmann} \emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers} (2018) \url{http://link.springer.com/10.1007/978-3-319-98192-5/_60}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60} · ISBN: 978-3-319-98191-8}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
-\CSLLeftMargin{7. }
-\CSLRightInline{\textbf{FAIR data points supporting big data interoperability}
-\CSLBlock{Luiz Olavo Bonino Da Silva Santos, Mark D Wilkinson, Arnold Kuzniar, Rajaram Kaliyaperumal, Mark Thompson, Michel Dumontier, Kees Burger} \emph{Enterprise interoperability in the digitized and networked factory of the future} (2016) \url{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}
-\CSLBlock{ISBN: 9781847040442}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
-\CSLLeftMargin{8. }
-\CSLRightInline{\textbf{The semantic web identity crisis: In search of the trivialities that never were}
-\CSLBlock{Ruben Verborgh, Miel Vander Sande} \emph{Satya Widya} (2020-01-31)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/sw-190372}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
-\CSLLeftMargin{9. }
-\CSLRightInline{\textbf{Conclusion and future challenges}
-\CSLBlock{Albert Meroño-Peñuela, Pasquale Lisena, Carlos Martínez-Ortiz} \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers} (2021)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7} · ISBN: 978-3-031-00789-7}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
-\CSLLeftMargin{10. }
-\CSLRightInline{\textbf{An analysis of public REST web service apis}
-\CSLBlock{Andy Neumann, Nuno Laranjeiro, Jorge Bernardino} \emph{IEEE Transactions on Services Computing} (2021-07-01)
-\CSLBlock{DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/tsc.2018.2847344}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
-\CSLLeftMargin{11. }
-\CSLRightInline{\textbf{Linked data, JSON-LD and the semantics of cultural and scientific heritage}
-\CSLBlock{Ovidiu Turcoane} \emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} (2014-09-30)
-\CSLBlock{DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pa67pUtR}{}}%
-\CSLLeftMargin{12. }
-\CSLRightInline{\textbf{F*** workflows: when parts of FAIR are missing}
-\CSLBlock{Sean R Wilkinson, Greg Eisenhauer, Anuj J Kapadia, Kathryn Knight, Jeremy Logan, Patrick Widener, Matthew Wolf} \emph{arXiv} (2022) \url{https://doi.org/grk3v5}
-\CSLBlock{DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
-\CSLLeftMargin{13. }
-\CSLRightInline{\textbf{Do Developers Make Unbiased Decisions? - The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components}
-\CSLBlock{Anisa Stefi} \emph{University of Münster, Münster, Germany} (2015) \url{https://doi.org/grcpwg}
-\CSLBlock{DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
-\CSLLeftMargin{14. }
-\CSLRightInline{\textbf{To develop or to reuse? Two perspectives on external reuse in software projects}
-\CSLBlock{Anisa Stefi, Thomas Hess} \emph{Software business} (2015)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18} · ISBN: 978-3-319-19592-6}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
-\CSLLeftMargin{15. }
-\CSLRightInline{\textbf{Web data apis over SPARQL}
-\CSLBlock{Albert Meroño-Peñuela, Pasquale Lisena, Carlos Martínez-Ortiz} \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers} (2021)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3} · ISBN: 978-3-031-00789-7}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
-\CSLLeftMargin{16. }
-\CSLRightInline{\textbf{FAIR Digital Objects Forum \textbar{}} \url{https://fairdo.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
-\CSLLeftMargin{17. }
-\CSLRightInline{\textbf{FAIR digital object overview and specifications}
-\CSLBlock{I Anders, Ch Blanchi, Daan Broeder, M Hellström, Sh Islam, Th Jejkal, L Lannom, K Peters, R Quick, A Schlemmer, \ldots{} Carlo Zwölf} \emph{FDO Forum} (2022-11-19) \url{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
-\CSLLeftMargin{18. }
-\CSLRightInline{\textbf{FDO Forum Document Standards}
-\CSLBlock{C Weiland, U Schwardmann, P Wittenburg, C Kirkpatrick, R Hanisch, Z Trautt} \emph{FDO Forum} (2022-01-29) \url{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
-\CSLLeftMargin{19. }
-\CSLRightInline{\textbf{FDO Forum FDO requirement specifications}
-\CSLBlock{G Strawn, P Wittenburg, St Soiland-Reyes, K Peters, L Lannom, U Schwardmann, Ch Blanch} \emph{FDO Forum} (2022-08-22) \url{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
-\CSLLeftMargin{20. }
-\CSLRightInline{\textbf{FAIR digital object framework}
-\CSLBlock{Luiz Bonino, Oeter Wittenburg, Bonnie Carroll, Alex Hardisty, Mark Leggott, Carlo Zwölf} \emph{FDOF technical implementation guideline} (2019-11-22) \url{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
-\CSLLeftMargin{21. }
-\CSLRightInline{\textbf{FDO machine actionability}
-\CSLBlock{Claus Weiland, Sharif Islam, Daan Broder, Ivonne Anders, Peter Wittenburg} \emph{FDO Forum} (2022-06-11) \url{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
-\CSLLeftMargin{22. }
-\CSLRightInline{\textbf{FDO configuration types}
-\CSLBlock{Larry Lannom, Karsten Peters-von Gehlen, Ivonne Anders, Andreas Pfeil, Alexander Schlemmer, Zach Trautt, Peter Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
-\CSLLeftMargin{23. }
-\CSLRightInline{\textbf{FAIR Signposting Profile}
-\CSLBlock{Herbert Van de Sompel, Martin Klein, Shawn Jones, Michael L Nelson, Simeon Warner, Anusuriya Devaraju, Robert Huber, Wilko Steinhoff, Vyacheslav Tykhonov, Luc Boruta, \ldots{} Mark Wilkinson} (2022-07-27) \url{https://signposting.org/FAIR/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
-\CSLLeftMargin{24. }
-\CSLRightInline{\textbf{FDO PID profiles \& attributes}
-\CSLBlock{Ivonne Anders, Maggie Hellström, Sharif Islam, Thomas Jejkal, Larry Lannom, Ulrich Schwardmann, Peter Wittenburg} \emph{FDO Forum} (2022-06-07) \url{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
-\CSLLeftMargin{25. }
-\CSLRightInline{\textbf{RDA Recommendation on PID Kernel Information}
-\CSLBlock{Tobias Weigel, Beth Plale, Mark Parsons, Gabriel Zhou, Yu Luo, Ulrich Schwardmann, Robert Quick, Margareta Hellström, Kei Kurakawa} \emph{Research Data Alliance} (2018)
-\CSLBlock{DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Fi62cJAo}{}}%
-\CSLLeftMargin{26. }
-\CSLRightInline{\textbf{FDO -- kernel attributes \& metadata}
-\CSLBlock{FDO-TSIG Working Group, FDO-BIG Working Group} \emph{FDO Forum} (2022-08-28) \url{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
-\CSLLeftMargin{27. }
-\CSLRightInline{\textbf{FDO -- granularity, versioning, mutability}
-\CSLBlock{FDO-TSIG Working Group} \emph{FDO Forum} (2022-08-26) \url{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
-\CSLLeftMargin{28. }
-\CSLRightInline{\textbf{DOIP endorsement request}
-\CSLBlock{FDO-TSIG Working Group} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
-\CSLLeftMargin{29. }
-\CSLRightInline{\textbf{Digital object interface protocol specification, version 2.0}
-\CSLBlock{DONA Foundation} \emph{DONA foundation} (2018-11-12) \url{https://hdl.handle.net/0.DOIP/DOIPV2.0}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
-\CSLLeftMargin{30. }
-\CSLRightInline{\textbf{FAIR digital object demonstrators 2021}
-\CSLBlock{Peter Wittenburg, Ivonne Anders, Christophe Blanchi, Merret Buurman, Carole Goble, Jonas Grieb, Alex Hardisty, Sharif Islam, Thomas Jejkal, Tibor Kálmán, \ldots{} Philipp Wieder} \emph{Zenodo} (2022)
-\CSLBlock{DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
-\CSLLeftMargin{31. }
-\CSLRightInline{\textbf{FDO -- upload of FDO}
-\CSLBlock{Christophe Blanchi, Daan Broeder, Thomas Jejkal, Islam Sharif, Alexander Schlemmer, Dieter van Uytvanck, Peter Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
-\CSLLeftMargin{32. }
-\CSLRightInline{\textbf{ResourceSync Framework Specification - Table of Contents} \url{http://www.openarchives.org/rs/toc}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
-\CSLLeftMargin{33. }
-\CSLRightInline{\textbf{Typing FAIR digital objects}
-\CSLBlock{Larry Lannom, U Schwardmann, C Blanchi, P Wittenburg} \emph{FDO Forum} (2022-06-08) \url{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
-\CSLLeftMargin{34. }
-\CSLRightInline{\textbf{Implementation of attributes, types, profiles and registries}
-\CSLBlock{C Blanchi, M Hellström, L Lannom, U Schwardmann, P Wittenburg} \emph{FDO Forum} (2022-11-27) \url{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
-\CSLLeftMargin{35. }
-\CSLRightInline{\textbf{Two Examples on How FDO Types can Support Machine and Human Readability}
-\CSLBlock{Ulrich Schwardmann, Tibor Kálmán} \emph{Research Ideas and Outcomes} (2022-10-12) \url{https://doi.org/grcq23}
-\CSLBlock{DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
-\CSLLeftMargin{36. }
-\CSLRightInline{\textbf{A framework for distributed digital object services}
-\CSLBlock{Robert Kahn, Robert Wilensky} \emph{International Journal on Digital Libraries} (2006-04) \url{https://www.doi.org/topics/2006_05_02_Kahn_Framework.pdf}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
-\CSLLeftMargin{37. }
-\CSLRightInline{\textbf{A framework for distributed digital object services}
-\CSLBlock{Robert Kahn, Robert Wilensky} \emph{CNRI} (1995-05-13) \url{http://www.cnri.reston.va.us/k-w.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
-\CSLLeftMargin{38. }
-\CSLRightInline{\textbf{X.1255 : Framework for discovery of identity management information} \url{https://www.itu.int/rec/T-REC-X.1255-201309-I}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
-\CSLLeftMargin{39. }
-\CSLRightInline{\textbf{Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation} \url{https://www.dona.net/doipv1doc}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
-\CSLLeftMargin{40. }
-\CSLRightInline{\textbf{Digital objects as drivers towards convergence in data infrastructures}
-\CSLBlock{Peter Wittenburg, George Strawn, Barend Mons, Luiz Bonino, Erik Schultes} \emph{https://b2share.eudat.eu} (2019-01-06)
-\CSLBlock{DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
-\CSLLeftMargin{41. }
-\CSLRightInline{\textbf{DOIP and Examples --- Cordra documentation} \url{https://www.cordra.org/documentation/api/doip.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
-\CSLLeftMargin{42. }
-\CSLRightInline{\textbf{draft-bhutton-json-schema-00} \url{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
-\CSLLeftMargin{43. }
-\CSLRightInline{\textbf{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}
-\CSLBlock{Tim Berners-Lee, Mark Fischetti} \emph{HarperSanFrancisco} (1999)
-\CSLBlock{ISBN: 9780062515865}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
-\CSLLeftMargin{44. }
-\CSLRightInline{\textbf{RDF 1.1 Primer} \url{http://www.w3.org/TR/rdf11-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
-\CSLLeftMargin{45. }
-\CSLRightInline{\textbf{Resource Description Framework (RDF) Model and Syntax Specification} \url{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
-\CSLLeftMargin{46. }
-\CSLRightInline{\textbf{Process modelling for information system description}
-\CSLBlock{Stefan K Stanczyk} \emph{The Open University} (1987)
-\CSLBlock{DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
-\CSLLeftMargin{47. }
-\CSLRightInline{\textbf{Uniform Resource Identifier (URI): Generic Syntax}
-\CSLBlock{T Berners-Lee, R Fielding, L Masinter} \emph{RFC Editor} (2005-01) \url{https://doi.org/ggqvpr}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-\CSLLeftMargin{48. }
-\CSLRightInline{\textbf{"info" URI Registry (Frozen)} \url{https://oclc-research.github.io/infoURI-Frozen/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
-\CSLLeftMargin{49. }
-\CSLRightInline{\textbf{Identifiers.org and MIRIAM Registry: community resources to provide persistent identification}
-\CSLBlock{N Juty, N Le Novere, C Laibe} \emph{Nucleic Acids Research} (2011-12-02) \url{https://doi.org/cx2776}
-\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22140103}{22140103} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3245029}{PMC3245029}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
-\CSLLeftMargin{50. }
-\CSLRightInline{\textbf{Internationalized Resource Identifiers (IRIs)}
-\CSLBlock{M Duerst, M Suignard} \emph{RFC Editor} (2005-01) \url{https://doi.org/gjvnbg}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
-\CSLLeftMargin{51. }
-\CSLRightInline{\textbf{Cool URIs for the semantic web}
-\CSLBlock{Leo Sauermann, Richard Cyganiak, Max Völkel} \emph{Universität des Saarlandes} (2011-07-13)
-\CSLBlock{DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
-\CSLLeftMargin{52. }
-\CSLRightInline{\textbf{The Semantic Web --- ISWC 2002}
-\CSLBlock{Ian Horrocks, James Hendler (editors)} \emph{Lecture Notes in Computer Science} (2002) \url{https://doi.org/fwjmhp}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
-\CSLLeftMargin{53. }
-\CSLRightInline{\textbf{SPARQL 1.1 Overview} \url{http://www.w3.org/TR/sparql11-overview/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
-\CSLLeftMargin{54. }
-\CSLRightInline{\textbf{State of the nation in data integration for bioinformatics.}
-\CSLBlock{Carole Goble, Robert Stevens} \emph{Journal of Biomedical Informatics} (2008-10-01)
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/18358788}{18358788}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
-\CSLLeftMargin{55. }
-\CSLRightInline{\textbf{Open PHACTS: semantic interoperability for drug discovery}
-\CSLBlock{Antony J Williams, Lee Harland, Paul Groth, Stephen Pettifer, Christine Chichester, Egon L Willighagen, Chris T Evelo, Niklas Blomberg, Gerhard Ecker, Carole Goble, Barend Mons} \emph{Drug Discovery Today} (2012-11) \url{https://doi.org/f4fdkd}
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22683805}{22683805}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
-\CSLLeftMargin{56. }
-\CSLRightInline{\textbf{SKOS Simple Knowledge Organization System Primer} \url{http://www.w3.org/TR/skos-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
-\CSLLeftMargin{57. }
-\CSLRightInline{\textbf{How matchable are four thousand ontologies on the semantic web}
-\CSLBlock{Wei Hu, Jianfeng Chen, Hang Zhang, Yuzhong Qu} \emph{The semantic web: Research and applications} (2011)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20} · ISBN: 978-3-642-21033-4}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
-\CSLLeftMargin{58. }
-\CSLRightInline{\textbf{API-centric Linked Data integration: The Open PHACTS Discovery Platform case study}
-\CSLBlock{Paul Groth, Antonis Loizou, Alasdair JG Gray, Carole Goble, Lee Harland, Steve Pettifer} \emph{Journal of Web Semantics} (2014-12)
-\CSLBlock{DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
-\CSLLeftMargin{59. }
-\CSLRightInline{\textbf{A systematic analysis of term reuse and term overlap across biomedical ontologies}
-\CSLBlock{Maulik R Kamdar, Tania Tudorache, Mark A Musen} \emph{Semantic Web} (2017-08-07)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/28819351}{28819351} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5555235}{PMC5555235}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
-\CSLLeftMargin{60. }
-\CSLRightInline{\textbf{Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot}
-\CSLBlock{Martin Klein, Herbert Van de Sompel, Robert Sanderson, Harihar Shankar, Lyudmila Balakireva, Ke Zhou, Richard Tobin} \emph{PLoS ONE} (2014-12-26)
-\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/25541969}{25541969} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4277367}{PMC4277367}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
-\CSLLeftMargin{61. }
-\CSLRightInline{\textbf{A more decentralized vision for linked data}
-\CSLBlock{Axel Polleres, Maulik Rajendra Kamdar, Javier David Fernández, Tania Tudorache, Mark Alan Musen} \emph{Satya Widya} (2020-01-31)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/sw-190380}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
-\CSLLeftMargin{62. }
-\CSLRightInline{\textbf{The landscape of ontology reuse approaches}
-\CSLBlock{Valentina Anita Carriero, Marilena Daquino, Aldo Gangemi, Andrea Giovanni Nuzzolese, Silvio Peroni, Valentina Presutti, Francesca Tomasi} \emph{Applications and practices in ontology design, extraction, and reasoning} (2020-11-12)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/ssw200033} · ISBN: 9781643681429}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
-\CSLLeftMargin{63. }
-\CSLRightInline{\textbf{RDF Schema 1.1} \url{http://www.w3.org/TR/rdf-schema/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
-\CSLLeftMargin{64. }
-\CSLRightInline{\textbf{OWL 2 Web Ontology Language Document Overview (Second Edition)} \url{http://www.w3.org/TR/owl2-overview/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
-\CSLLeftMargin{65. }
-\CSLRightInline{\textbf{Mapping between the OBO and OWL ontology languages}
-\CSLBlock{Syed Tirmizi, Stuart Aitken, Dilvan A Moreira, Chris Mungall, Juan Sequeda, Nigam H Shah, Daniel P Miranker} \emph{Journal of Biomedical Semantics} (2011)
-\CSLBlock{DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/21388572}{21388572} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3105495}{PMC3105495}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
-\CSLLeftMargin{66. }
-\CSLRightInline{\textbf{Linked Data - Design Issues} \url{https://www.w3.org/DesignIssues/LinkedData.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
-\CSLLeftMargin{67. }
-\CSLRightInline{\textbf{The Open Graph protocol} \url{https://ogp.me/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
-\CSLLeftMargin{68. }
-\CSLRightInline{\textbf{RDFa 1.1 Primer - Third Edition} \url{http://www.w3.org/TR/rdfa-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
-\CSLLeftMargin{69. }
-\CSLRightInline{\textbf{HTML Standard} \url{https://html.spec.whatwg.org/multipage/microdata.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
-\CSLLeftMargin{70. }
-\CSLRightInline{\textbf{JSON-LD 1.1} \url{https://www.w3.org/TR/json-ld/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
-\CSLLeftMargin{71. }
-\CSLRightInline{\textbf{Usage Statistics of JSON-LD for Websites, May 2022} \url{https://w3techs.com/technologies/details/da-jsonld}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
-\CSLLeftMargin{72. }
-\CSLRightInline{\textbf{Designing a Linked Data developer experience} (2018-12-28) \url{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
-\CSLLeftMargin{73. }
-\CSLRightInline{\textbf{Shapes Constraint Language (SHACL)} \url{https://www.w3.org/TR/shacl/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
-\CSLLeftMargin{74. }
-\CSLRightInline{\textbf{Shape Expressions (ShEx) 2.1 Primer} \url{http://shex.io/shex-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
-\CSLLeftMargin{75. }
-\CSLRightInline{\textbf{Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation}
-\CSLBlock{Katherine Thornton, Harold Solbrig, Gregory S Stupp, Jose Emilio Labra Gayo, Daniel Mietchen, Eric Prud, Andra Waagmeester} \emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings} (2019)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39} · ISBN: 978-3-030-21347-3}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
-\CSLLeftMargin{76. }
-\CSLRightInline{\textbf{Validating RDF Data}
-\CSLBlock{Jose Emilio Labra Gayo, Eric Prud'hommeaux, Iovka Boneva, Dimitris Kontokostas} \emph{Synthesis Lectures on the Semantic Web: Theory and Technology} (2017-09-28)
-\CSLBlock{DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
-\CSLLeftMargin{77. }
-\CSLRightInline{\textbf{Survey of tools for linked data consumption}
-\CSLBlock{Jakub Klímek, Petr Škoda, Martin Nečaský} \emph{Satya Widya} (2019-05-23)
-\CSLBlock{DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/sw-180316}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
-\CSLLeftMargin{78. }
-\CSLRightInline{\textbf{RightField: embedding ontology annotation in spreadsheets}
-\CSLBlock{K Wolstencroft, S Owen, M Horridge, O Krebs, W Mueller, JL Snoep, F du Preez, C Goble} \emph{Bioinformatics} (2011-05-26)
-\CSLBlock{DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/21622664}{21622664}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
-\CSLLeftMargin{79. }
-\CSLRightInline{\textbf{REST and Linked Data}
-\CSLBlock{Kevin R Page, David C De Roure, Kirk Martinez} \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11} (2011) \url{http://eprints.soton.ac.uk/id/eprint/272098}
-\CSLBlock{DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
-\CSLLeftMargin{80. }
-\CSLRightInline{\textbf{On Schema.org and Why It Matters for the Web}
-\CSLBlock{Peter Mika} \emph{IEEE Internet Computing} (2015-07)
-\CSLBlock{DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
-\CSLLeftMargin{81. }
-\CSLRightInline{\textbf{A new look at the semantic web}
-\CSLBlock{Abraham Bernstein, James Hendler, Natalya Noy} \emph{Communications of the ACM} (2016-08-24)
-\CSLBlock{DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
-\CSLLeftMargin{82. }
-\CSLRightInline{\textbf{An Interoperability Framework and Distributed Platform for Fast Data Applications}
-\CSLBlock{José Carlos Martins Delgado} \emph{Data Science and Big Data Computing} (2016)
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRzcHhPD}{}}%
-\CSLLeftMargin{83. }
-\CSLRightInline{\textbf{A Comparison Framework for Middleware Infrastructures.}
-\CSLBlock{Apostolos Zarras} \emph{The Journal of Object Technology} (2004)
-\CSLBlock{DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
-\CSLLeftMargin{84. }
-\CSLRightInline{\textbf{The FAIR data maturity model: An approach to harmonise FAIR assessments}
-\CSLBlock{Christophe Bahim, Carlos Casorrán-Amilburu, Makx Dekkers, Edit Herczog, Nicolas Loozen, Konstantinos Repanas, Keith Russell, Shelley Stall} \emph{Data Science Journal} (2020-10-27)
-\CSLBlock{DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
-\CSLLeftMargin{85. }
-\CSLRightInline{\textbf{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}
-\CSLBlock{William Stallings} \emph{Sams} (1990)
-\CSLBlock{ISBN: 9780672226977}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
-\CSLLeftMargin{86. }
-\CSLRightInline{\textbf{Architectural styles and the design of network-based software architectures}
-\CSLBlock{Roy Thomas Fielding} \emph{University of California, Irvine} (2000) \url{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
-\CSLLeftMargin{87. }
-\CSLRightInline{\textbf{Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award)}
-\CSLBlock{Roy T Fielding, Richard N Taylor, Justin R Erenkrantz, Michael M Gorlick, Jim Whitehead, Rohit Khare, Peyman Oreizy} \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017} (2017-09-04) \url{http://dl.acm.org/citation.cfm?doid=3106237.3121282}
-\CSLBlock{DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282} · ISBN: 9781450351058}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
-\CSLLeftMargin{88. }
-\CSLRightInline{\textbf{OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More} \url{https://spec.openapis.org/oas/v3.1.0.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
-\CSLLeftMargin{89. }
-\CSLRightInline{\textbf{Schema.org Actions - schema.org} \url{https://schema.org/docs/actions.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
-\CSLLeftMargin{90. }
-\CSLRightInline{\textbf{Web Services Description Language (WSDL) Version 2.0 Part 0: Primer} \url{http://www.w3.org/TR/wsdl20-primer/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
-\CSLLeftMargin{91. }
-\CSLRightInline{\textbf{The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud}
-\CSLBlock{Katherine Wolstencroft, Robert Haines, Donal Fellows, Alan Williams, David Withers, Stuart Owen, Stian Soiland-Reyes, Ian Dunlop, Aleksandra Nenadic, Paul Fisher, \ldots{} Carole Goble} \emph{Nucleic Acids Research} (2013-05-02) \url{https://doi.org/ggbwf4}
-\CSLBlock{DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/23640334}{23640334} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3692062}{PMC3692062}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
-\CSLLeftMargin{92. }
-\CSLRightInline{\textbf{Perspectives on automated composition of workflows in the life sciences}
-\CSLBlock{Anna-Lena Lamprecht, Magnus Palmblad, Jon Ison, Veit Schwämmle, Mohammad Sadnan Al Manir, Ilkay Altintas, Christopher JO Baker, Ammar Ben Hadj Amor, Salvador Capella-Gutierrez, Paulos Charonyktakis, \ldots{} Katherine Wolstencroft} \emph{F1000Research} (2021-09-07) \url{https://doi.org/gqprp7}
-\CSLBlock{DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/34804501}{34804501} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8573700}{PMC8573700}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-\CSLLeftMargin{93. }
-\CSLRightInline{\textbf{The Modern Standards Paradigm - Five Key Principles}
-\CSLBlock{OpenStand} \url{https://open-stand.org/about-us/principles/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
-\CSLLeftMargin{94. }
-\CSLRightInline{\textbf{First International Conference on FAIR Digital Objects}
-\CSLBlock{Tina Loo (editor)} \emph{Pensoft Publishers} (1970) \url{https://doi.org/grcqxh}
-\CSLBlock{DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
-\CSLLeftMargin{95. }
-\CSLRightInline{\textbf{DOIP API for HTTP Clients --- Cordra documentation} \url{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-\CSLLeftMargin{96. }
-\CSLRightInline{\textbf{DOI Resolution Documentation} \url{https://www.doi.org/factsheets/DOIProxy.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-\CSLLeftMargin{97. }
-\CSLRightInline{\textbf{DOI Handbook - Resolution} \url{https://www.doi.org/doi_handbook/3_Resolution.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
-\CSLLeftMargin{98. }
-\CSLRightInline{\textbf{HTTP Semantics}
-\CSLBlock{R Fielding, M Nottingham, J Reschke (editors)} \emph{RFC Editor} (2022-06) \url{https://doi.org/gqprqb}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
-\CSLLeftMargin{99. }
-\CSLRightInline{\textbf{Linked data platform 1.0}
-\CSLBlock{Linked Data Platform Working Group} \emph{W3C} (2015-02-26) \url{https://www.w3.org/TR/2015/REC-ldp-20150226/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
-\CSLLeftMargin{100. }
-\CSLRightInline{\textbf{FAIR Digital Object Framework Documentation} \url{https://fairdigitalobjectframework.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
-\CSLLeftMargin{101. }
-\CSLRightInline{\textbf{Handle System Overview}
-\CSLBlock{S Sun, L Lannom, B Boesch} \emph{RFC Editor} (2003-11) \url{https://doi.org/ggn83z}
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
-\CSLLeftMargin{102. }
-\CSLRightInline{\textbf{Web API design best practices - Azure Architecture Center}
-\CSLBlock{martinekuan} \url{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
-\CSLLeftMargin{103. }
-\CSLRightInline{\textbf{The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration}
-\CSLBlock{Barry Smith, Michael Ashburner, Cornelius Rosse, Jonathan Bard, William Bug, Werner Ceusters, Louis J Goldberg, Karen Eilbeck, Amelia Ireland, \ldots{} Suzanna Lewis} \emph{Nature Biotechnology} (2007-11) \url{https://doi.org/bqng99}
-\CSLBlock{DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/17989687}{17989687} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2814061}{PMC2814061}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
-\CSLLeftMargin{104. }
-\CSLRightInline{\textbf{Data Information View} \url{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
-\CSLLeftMargin{105. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 3} \url{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-\CSLLeftMargin{106. }
-\CSLRightInline{\textbf{ORE User Guide - Primer} \url{http://www.openarchives.org/ore/1.0/primer}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-\CSLLeftMargin{107. }
-\CSLRightInline{\textbf{PROV-O: The PROV Ontology} \url{https://www.w3.org/TR/2013/REC-prov-o-20130430/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
-\CSLLeftMargin{108. }
-\CSLRightInline{\textbf{Semantic Web Services}
-\CSLBlock{Dieter Fensel, Federico Michele Facca, Elena Simperl, Ioan Toma} \emph{Springer Berlin Heidelberg} (2011) \url{https://doi.org/bv7nnc}
-\CSLBlock{DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-\CSLLeftMargin{109. }
-\CSLRightInline{\textbf{5-star Open Data} \url{http://5stardata.info/en/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
-\CSLLeftMargin{110. }
-\CSLRightInline{\url{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
-\CSLLeftMargin{111. }
-\CSLRightInline{\textbf{Global Internet Phenomena Report 2022}
-\CSLBlock{Sandvine} \url{https://www.sandvine.com/global-internet-phenomena-report-2022}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
-\CSLLeftMargin{112. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol Version 2 (HTTP/2)}
-\CSLBlock{M Belshe, R Peon} \emph{RFC Editor} (2015-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
-\CSLLeftMargin{113. }
-\CSLRightInline{\url{https://blog.cloudflare.com/http-3-vs-http-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
-\CSLLeftMargin{114. }
-\CSLRightInline{\textbf{draft-ietf-quic-http-34} \url{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
-\CSLLeftMargin{115. }
-\CSLRightInline{\textbf{QUIC: A UDP-Based Multiplexed and Secure Transport}
-\CSLBlock{J Iyengar, M Thomson (editors)} \emph{RFC Editor} (2021-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
-\CSLLeftMargin{116. }
-\CSLRightInline{\textbf{URI Template}
-\CSLBlock{J Gregorio, R Fielding, M Hadley, M Nottingham, D Orchard} \emph{RFC Editor} (2012-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
-\CSLLeftMargin{117. }
-\CSLRightInline{\textbf{Content negotiation - HTTP \textbar{} MDN} \url{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
-\CSLLeftMargin{118. }
-\CSLRightInline{\textbf{Hypertext Style: Cool URIs don't change.} \url{https://www.w3.org/Provider/Style/URI}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
-\CSLLeftMargin{119. }
-\CSLRightInline{\textbf{HTTP Framework for Time-Based Access to Resource States -\/- Memento}
-\CSLBlock{H Van de Sompel, M Nelson, R Sanderson} \emph{RFC Editor} (2013-12)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
-\CSLLeftMargin{120. }
-\CSLRightInline{\textbf{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}
-\CSLBlock{G Clemm, J Amsden, T Ellison, C Kaler, J Whitehead} \emph{RFC Editor} (2002-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
-\CSLLeftMargin{121. }
-\CSLRightInline{\textbf{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}
-\CSLBlock{L Dusseault (editor)} \emph{RFC Editor} (2007-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
-\CSLLeftMargin{122. }
-\CSLRightInline{\textbf{Upgrading to TLS Within HTTP/1.1}
-\CSLBlock{R Khare, S Lawrence} \emph{RFC Editor} (2000-05)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
-\CSLLeftMargin{123. }
-\CSLRightInline{\textbf{Hypertext Style: Cool URIs don't change.} \url{https://www.w3.org/Provider/Style/URI.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
-\CSLLeftMargin{124. }
-\CSLRightInline{\textbf{draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes} \url{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
-\CSLLeftMargin{125. }
-\CSLRightInline{\textbf{The 'profile' Link Relation Type}
-\CSLBlock{E Wilde} \emph{RFC Editor} (2013-03)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
-\CSLLeftMargin{126. }
-\CSLRightInline{\textbf{W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures} \url{http://www.w3.org/TR/xmlschema11-1/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
-\CSLLeftMargin{127. }
-\CSLRightInline{\textbf{JSON-LD 1.1} \url{http://www.w3.org/TR/json-ld/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
-\CSLLeftMargin{128. }
-\CSLRightInline{\textbf{WebSockets Standard} \url{https://websockets.spec.whatwg.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
-\CSLLeftMargin{129. }
-\CSLRightInline{\textbf{Linked Data Notifications} \url{https://www.w3.org/TR/ldn/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
-\CSLLeftMargin{130. }
-\CSLRightInline{\textbf{The Atom Publishing Protocol}
-\CSLBlock{J Gregorio, B de hOra (editors)} \emph{RFC Editor} (2007-10)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
-\CSLLeftMargin{131. }
-\CSLRightInline{\textbf{SWORD 3.0 Specification} \url{https://swordapp.github.io/swordv3/swordv3.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
-\CSLLeftMargin{132. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}
-\CSLBlock{R Fielding, J Reschke (editors)} \emph{RFC Editor} (2014-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-\CSLLeftMargin{133. }
-\CSLRightInline{\textbf{Hydra W3C Community Group} \url{https://www.hydra-cg.com/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
-\CSLLeftMargin{134. }
-\CSLRightInline{\textbf{draft-kelly-json-hal-08} \url{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
-\CSLLeftMargin{135. }
-\CSLRightInline{\textbf{Web Linking}
-\CSLBlock{M Nottingham} \emph{RFC Editor} (2017-10)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
-\CSLLeftMargin{136. }
-\CSLRightInline{\textbf{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}
-\CSLBlock{R Fielding, J Reschke (editors)} \emph{RFC Editor} (2014-06)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
-\CSLLeftMargin{137. }
-\CSLRightInline{\textbf{HTTP Live Streaming}
-\CSLBlock{W May} \emph{RFC Editor} (2017-08)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
-\CSLLeftMargin{138. }
-\CSLRightInline{\textbf{ISO/IEC 23009-1:2019}
-\CSLBlock{14:00-17:00} \emph{ISO} \url{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
-\CSLLeftMargin{139. }
-\CSLRightInline{\textbf{FAIR data maturity model: Specification and guidelines}
-\CSLBlock{Research Data Alliance FAIR Data Maturity Model Working Group} \emph{Research Data Alliance} (2020) \url{https://zenodo.org/record/3909563/\#.\%7BYGRNnq8za70\%7D}
-\CSLBlock{DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
-\CSLLeftMargin{140. }
-\CSLRightInline{\textbf{NCBO BioPortal} \url{https://bioportal.bioontology.org/ontologies}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
-\CSLLeftMargin{141. }
-\CSLRightInline{\textbf{The Globus Striped GridFTP Framework and Server}
-\CSLBlock{W Allcock, J Bresnahan, R Kettimuthu, M Link} \emph{ACM/IEEE SC 2005 Conference (SC'05)} \url{https://doi.org/cgmc2b}
-\CSLBlock{DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-\CSLLeftMargin{142. }
-\CSLRightInline{\textbf{The ARK Identifier Scheme} \url{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
-\CSLLeftMargin{143. }
-\CSLRightInline{\textbf{Handle System Protocol (ver 2.1) Specification}
-\CSLBlock{S Sun, S Reilly, L Lannom, J Petrone} \emph{RFC Editor} (2003-11)
-\CSLBlock{DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-\CSLLeftMargin{144. }
-\CSLRightInline{\textbf{Handle.Net Registry} \url{https://www.handle.net/download_hnr.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-\CSLLeftMargin{145. }
-\CSLRightInline{\textbf{Licenses \& Standards \textbar{} Open Source Initiative} \url{https://opensource.org/licenses}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-\CSLLeftMargin{146. }
-\CSLRightInline{\textbf{ISO 16684-1:2019}
-\CSLBlock{14:00-17:00} \emph{ISO} \url{https://www.iso.org/standard/75163.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
-\CSLLeftMargin{147. }
-\CSLRightInline{\textbf{Data - W3C} \url{https://www.w3.org/standards/semanticweb/data}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-\CSLLeftMargin{148. }
-\CSLRightInline{\textbf{Linking Across Provenance Bundles} \url{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-\CSLLeftMargin{149. }
-\CSLRightInline{\textbf{Introducing 'Role'}
-\CSLBlock{Unknown} \url{http://blog.schema.org/2014/06/introducing-role.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-\CSLLeftMargin{150. }
-\CSLRightInline{\textbf{PROV-O: The PROV Ontology} \url{https://www.w3.org/TR/prov-o/\#specializationOf}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
-\CSLLeftMargin{151. }
-\CSLRightInline{\textbf{ORE Specification - Abstract Data Model} \url{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
-\CSLLeftMargin{152. }
-\CSLRightInline{\textbf{Schema.org - Schema.org} \url{https://schema.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
-\CSLLeftMargin{153. }
-\CSLRightInline{\textbf{DCMI Metadata Terms} \url{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
-\CSLLeftMargin{154. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 2} \url{https://www.w3.org/TR/vocab-dcat-2/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
-\CSLLeftMargin{155. }
-\CSLRightInline{\textbf{FOAF Vocabulary Specification} \url{http://xmlns.com/foaf/spec/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
-\CSLLeftMargin{156. }
-\CSLRightInline{\textbf{PAV ontology: provenance, authoring and versioning}
-\CSLBlock{Paolo Ciccarese, Stian Soiland-Reyes, Khalid Belhajjame, Alasdair JG Gray, Carole Goble, Tim Clark} \emph{Journal of Biomedical Semantics} (2013) \url{https://doi.org/gftcpx}
-\CSLBlock{DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/24267948}{24267948} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4177195}{PMC4177195}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
-\CSLLeftMargin{157. }
-\CSLRightInline{\textbf{DCMI Metadata Terms} (2020-01-20) \url{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
-\CSLLeftMargin{158. }
-\CSLRightInline{\textbf{The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
-Digital Object Approach to Speeding up Digital Mobilisation of Natural History
-Collections}
-\CSLBlock{Alex Hardisty, Paul Brack, Carole Goble, Laurence Livermore, Ben Scott, Quentin Groom, Stuart Owen, Stian Soiland-Reyes} \emph{Data Intelligence} (2022) \url{https://doi.org/grk33x}
-\CSLBlock{DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
-\CSLLeftMargin{159. }
-\CSLRightInline{\textbf{Data Catalog Vocabulary (DCAT) - Version 2} \url{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
-\CSLLeftMargin{160. }
-\CSLRightInline{\textbf{Bioschemas - Bioschemas} \url{http://bioschemas.org/}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
-\CSLLeftMargin{161. }
-\CSLRightInline{\textbf{Darwin Core: An Evolving Community-Developed Biodiversity Data Standard}
-\CSLBlock{John Wieczorek, David Bloom, Robert Guralnick, Stan Blum, Markus Döring, Renato Giovanni, Tim Robertson, David Vieglais} \emph{PLoS ONE} (2012-01-06) \url{https://doi.org/fzrpwq}
-\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/22238640}{22238640} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3253084}{PMC3253084}}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
-\CSLLeftMargin{162. }
-\CSLRightInline{\textbf{Semantic Web - XML2000 - slide "Architecture"} \url{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}}
-
-\end{CSLReferences}

### 2023-01-24 build/build.sh
overleaf compatibility

@@ -12,14 +12,20 @@ export TZ=Etc/UTC
 export LC_ALL=en_US.UTF-8
 
 
+# Overleaf workaround, where symlinks are not permitted in git
+ln -fs v/latest/manuscript.pdf webpage/
+ln -fs v/latest/images webpage/
+ln -fs v/latest/index.html webpage/
+ln -s ../latex/ro-crate.bib content/manual-references-ro-crate.bib
+
 # Set DOCKER_RUNNING to true if docker is running, otherwise false.
 DOCKER_RUNNING="$(docker info &> /dev/null && echo "true" || (true && echo "false"))"
 
 # Set option defaults
 CI="${CI:-false}"
-BUILD_PDF="${BUILD_PDF:-true}"
+BUILD_PDF="${BUILD_PDF:-false}"
 BUILD_DOCX="${BUILD_DOCX:-false}"
-BUILD_LATEX="${BUILD_LATEX:-false}"
+BUILD_LATEX="${BUILD_LATEX:-true}"
 SPELLCHECK="${SPELLCHECK:-false}"
 MANUBOT_USE_DOCKER="${MANUBOT_USE_DOCKER:-$DOCKER_RUNNING}"
 # Pandoc's configuration is specified via files of option defaults

### 2023-01-24 manuscript.tex
converted to peerj overleaf template [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/64d3fd6d7e3b49d20343cf1a1b33a9e52fb8d9de.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/64d3fd6d7e3b49d20343cf1a1b33a9e52fb8d9de/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4001304378

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/64d3fd6d7e3b49d20343cf1a1b33a9e52fb8d9de/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
+from \href{https://github.com/stain/2022-fdo-paper/tree/64d3fd6d7e3b49d20343cf1a1b33a9e52fb8d9de}{stain/2022-fdo-paper@64d3fd6}
 on January 24, 2023.

### 2023-01-24 manuscript.tex
overleaf compatibility [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/fa45efa5c79f23abe61e9b521cf4269293651d36.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/fa45efa5c79f23abe61e9b521cf4269293651d36/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4001325241

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/64d3fd6d7e3b49d20343cf1a1b33a9e52fb8d9de/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/fa45efa5c79f23abe61e9b521cf4269293651d36/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/64d3fd6d7e3b49d20343cf1a1b33a9e52fb8d9de}{stain/2022-fdo-paper@64d3fd6}
+from \href{https://github.com/stain/2022-fdo-paper/tree/fa45efa5c79f23abe61e9b521cf4269293651d36}{stain/2022-fdo-paper@fa45efa}
 on January 24, 2023.

### 2023-01-24 manuscript.tex
STIX2 fonts [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/1c02a3ca58bd5ba01fcd0dd7266edfe5376fb0c4.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/1c02a3ca58bd5ba01fcd0dd7266edfe5376fb0c4/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4001413120

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/fa45efa5c79f23abe61e9b521cf4269293651d36/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/1c02a3ca58bd5ba01fcd0dd7266edfe5376fb0c4/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/fa45efa5c79f23abe61e9b521cf4269293651d36}{stain/2022-fdo-paper@fa45efa}
+from \href{https://github.com/stain/2022-fdo-paper/tree/1c02a3ca58bd5ba01fcd0dd7266edfe5376fb0c4}{stain/2022-fdo-paper@1c02a3c}
 on January 24, 2023.

### 2023-01-25 latex/again.tex
Fixing references to use bibliography

@@
 In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and using Linked Data Platform (\protect\hyperlink{ref-17OHlMRQA}{Linked Data Platform Working Group, 2015}), as proposed by (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}).
-
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}; \protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) against its current implementations as DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}), with suggestions for required additions.
-\label{tbl:fdo-checks}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
-\midrule
-\endhead
-G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
-& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
-\texttt{http://} URIs replaced by \texttt{https://}.
-& Keep flexibility of RDF serialisation formats which may change more frequently \\
-G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
-DOIP relatively unknown, in effect only one implementation.
-& Further promote DOIP and justify its benefits.
-Build tutorials and OSI open source implementations.
-Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}), however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
-Example-driven documentation.
-\\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
-\\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
-Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
-\\
-G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
-DOIP operations can use any transport protocol.
-& Document transport protocols as FDOs, recommend which transport to use.
-& URI as abstraction base.
-Does not specify PID requirements.
-& Give stronger deployment recommendations.
-\\
-G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
-Unclear when metadata field is a PID or plain text.
-& Make datatype of fields explicit to support navigation.
-& Machine-navigation through URIs via properties and types.
-Unclear when URI should be followed or is just identifier, but always distinct from plain text.
-& \\
-G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
-& Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
-& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
-Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \\
-G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
-No FDOF implementations.
-& Develop demonstrator of FDOF usage based on existing LDP server.
-\\
-G9: \emph{standard compliance} & Handle (\protect\hyperlink{ref-2gXY2Nwh}{Sun, Lannom \& Boesch, 2003}), DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
-FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) & HTTP, LDP.
-FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
-FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}).
-Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-& HTTP URLs as basis for identifiers, but they are frequently not persistent.
-& Add strong guidance for PID services like w3id and persistence policies.
-\\
-FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
-Datacite DOIs content-negotiate to give registered metadata.
-& Add FAIR Signposting at PID provider for minimal PID record \\
-FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
-No method/attribute defined for separate metadata, only directly in PID Record.
-Unclear meaning of multiple items and bytestream chunks.
-& Clarify expectations for multiple items.
-Recommend chunks to not be used.
-& URIs resolvable by default.
-Multiple ways to resolve metadata, unclear preference.
-& Add FAIR Signposting and preference order.
-\\
-FDOF4: \emph{Additional attributes} & Freetext attribute keys.
-Attributes should be defined for FDO type (?).
-& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
-Explicitly allow attributes not already defined in type.
-& All attributes individually identified.
-Any Linked Data attributes can be used by URI or with mapped prefix.
-& Clarify type expectations of required/recommended/optional attributes.
-\\
-FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
-& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-Clarify that operations can be mapped to protocol directly.
-& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
-Unclear how to provide PID for additional operations.
-& Specify how additional operations should be called over HTTP.
-\\
-FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan}) are used instead.
-& Document how operation resources can be discovered from an LPD container.
-Document search API.
-\\
-FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
-PR-TypingFDOs leaves this open.
-& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
-Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
-& Always make LDP types explicit in FDO profile.
-\\
-FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
-Separate Metadata FDO need custom property.
-& Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
-& Assertions are always with semantics, using RDF vocabularies.
-Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
-& Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
-FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
-Metadata would need to be bundled within custom FDO types/attributes.
-Unclear how levels are separated within single FDO representation (need FDOF8?).
-& Declare which metadata are expected within response attribute or within FDO object.
-Require PIDs for custom attributes.
-Define how alternate metadata levels can be represented separately.
-& Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
-& Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
-FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
-& Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} (\protect\hyperlink{ref-CNP39tah}{Smith et al., 2007}) & Rather document better how individual ad-hoc schemas can be started for prototypes.
-\\
-FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
-Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-(\protect\hyperlink{ref-6LvydTsC}{{`Data Information View'}})).
-& & LDP Containers required by specification, also user-created (eg.
-\texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 (\protect\hyperlink{ref-1EW4hLN8U}{{`Data Catalog Vocabulary (DCAT) - Version 3'}}), \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE (\protect\hyperlink{ref-gtGuSevK}{{`ORE User Guide - Primer'}}) \\
-FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
-\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
-No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
-\bottomrule
-\end{longtable}
 

### 2023-01-25 latex/main.tex
Fixing references to use bibliography

--- 2023-01-25 latex/again.tex
@@
-This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/86bc7fe899c7464fad8f4ebe872efb61c3be1394/}{permalink})
-was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/86bc7fe899c7464fad8f4ebe872efb61c3be1394}{stain/2022-fdo-paper@86bc7fe}
-on January 24, 2023.
+%% Submissions for peer-review must enable line-numbering 
+%% using the lineno option in the \documentclass command.
+%%
+%% Camera-ready submissions do not need line numbers, and
+%% should have this option removed.
+%%
+%% Please note that the line numbering option requires
+%% version 1.1 or newer of the wlpeerj.cls file, and
+%% the corresponding author info requires v1.2
 
-\hypertarget{authors}{%
-\section*{Authors}\label{authors}}
-\addcontentsline{toc}{section}{Authors}
+\documentclass[fleqn,10pt,lineno]{wlpeerj}
 
-\begin{itemize}
-\item
-  \textbf{Stian Soiland-Reyes}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0001-9842-9718}{0000-0001-9842-9718}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/stain}{stain}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/soilandreyes}{soilandreyes}
-  Department of Computer Science, The University of Manchester, UK; Informatics Institute, Faculty of Science, University of Amsterdam, NL
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Carole Goble}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-1219-2137}{0000-0003-1219-2137}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/carolegoble}{carolegoble}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/CaroleAnneGoble}{CaroleAnneGoble}
-  Department of Computer Science, The University of Manchester, UK
-  · Funded by \href{https://bioexcel.eu/}{BioExcel-2} (European Commission \href{https://cordis.europa.eu/project/id/823830}{H2020-INFRAEDI-2018-1 823830}); \href{https://www.eosc-life.eu/}{EOSC-Life} (European Commission \href{https://cordis.europa.eu/project/id/824087}{H2020-INFRAEOSC-2018-2 824087}); \href{https://by-covid.eu/}{BY-COVID} (European Commission \href{https://cordis.europa.eu/project/id/101046203}{HORIZON-INFRA-2021-EMERGENCY-01 101046203}); \href{https://biodt.org/}{BioDT} (European Commission \href{https://cordis.europa.eu/project/id/101057437}{HORIZON-INFRA-2021-TECH-01-01 101057437}); \href{http://www.fair-impact.eu/}{FAIR-IMPACT} (European Commission (European Commission \href{https://cordis.europa.eu/project/id/101057344}{HORIZON-INFRA-2021-EOSC-01 101057344})
-\item
-  \textbf{Paul Groth}
-  \includegraphics[width=0.16667in,height=0.16667in]{images/orcid.svg}
-  \href{https://orcid.org/0000-0003-0183-6910}{0000-0003-0183-6910}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/github.svg}
-  \href{https://github.com/pgroth}{pgroth}
-  · \includegraphics[width=0.16667in,height=0.16667in]{images/twitter.svg}
-  \href{https://twitter.com/pgroth}{pgroth}
-  Informatics Institute, Faculty of Science, University of Amsterdam, NL
-\end{itemize}
+%% Definitions needed for pandoc's silly output
 
-\hypertarget{abstract}{%
-\section*{Abstract}\label{abstract}}
-\addcontentsline{toc}{section}{Abstract}
+\usepackage{hyperref}
+\hypersetup{
+  pdflang={en-GB},
+  hidelinks
+}
 
-\textbf{TODO}: Rewrite from notes below to actual abstract.
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\usepackage{longtable,booktabs,array}
 
+%% CSL definitions for pandoc's generated references
+%% https://github.com/jgm/pandoc/blob/2.19.2/data/templates/default.latex#L302
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{0.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{2em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+%% END
+
+%% STTYX2Fonts for lualatex
+% https://www.overleaf.com/learn/latex/Articles%2FOpenType-based_math_typesetting%3A_An_introduction_to_the_STIX2_OpenType_fonts
+\usepackage{amsmath}
+%\usepackage{hyperref}
+\usepackage{fontspec}
+\usepackage{unicode-math}
+%\usepackage{hologo}
+%\usepackage{graphicx}
+\setmathfont{STIX2Math}[
+Extension={.otf},
+Path=./STIX2fonts/,
+Scale=1]
+\setmainfont{STIX2Text}[
+Extension={.otf},
+Path=./STIX2fonts/,
+UprightFont={*-Regular},
+BoldFont={*-Bold},
+ItalicFont={*-Italic},
+BoldItalicFont={*-BoldItalic}]
+%% END
+
+%%%% END pandoc defs
+%%% Additional packages/definitions
+\usepackage{citation-style-language}
+\cslsetup{
+style = peerj
+}
+\addbibresource{bibliography.json}
+
+\usepackage{pdflscape} 
+%%% END own imports
+
+\title{Evaluating FAIR Digital Object as a distributed object system}
+
+\author[1,2]{Stian Soiland-Reyes}
+\author[1]{Carole Goble}
+\author[2]{Paul Groth}
+\affil[1]{Department of Computer Science, The University of Manchester, UK}
+\affil[2]{Informatics Institute, Faculty of Science, University of Amsterdam, NL }
+\corrauthor[1]{Stian Soiland-Reyes}{soiland-reyes@manchester.ac.uk}
+
+% \keywords{Keyword1, Keyword2, Keyword3}
+
+\begin{abstract}
 FAIR Digital Object is an emerging concept from EOSC.
@@
 Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
+\end{abstract}
 
-What is the message of this paper?
+\begin{document}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These are the overlaps
-\item
-  This is what FDO is requiring but not commonly deployed
-\item
-  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
-\item
-  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
-\item
-  Lessons for Semantic Web community. What is missing in FDO and in SW.
-\item
-  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
-\item
-  Go back to the old SW layer cake. Explains why we picked these frameworks.
-\item
-  Lessons for SW: Parts of the stack that is less. FDO contributions.
-\end{enumerate}
+\flushbottom
+\maketitle
+\thispagestyle{empty}
 
-Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.
+\section*{Introduction}
 
-Have all the ingredients, but not cooking properly.
+\subsection*{About PeerJ}
 
-Systematic through frameworks.
-
-Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?
+Here goes
 
@@
 
-The FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}) encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
+The FAIR principles \cite{6DjakjNS} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
 They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures (\protect\hyperlink{ref-u8Cy0psL}{Mons et al., 2017}).
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
+In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \cite{u8Cy0psL}.
+The EOSC Interoperability Framework \cite{aCye3KpE} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
 
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
+Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object \cite{IHLT6hye} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
 The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
 
-In other discourse, Linked Data (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\protect\hyperlink{ref-EJDjHrUv}{Bonino Da Silva Santos et al., 2016}; \protect\hyperlink{ref-GJHYN6xW}{Hasnain \& Rebholz-Schuhmann, 2018}).
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).
+In other discourse, Linked Data \cite{19s3yyxDn} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\cite{EJDjHrUv}; \cite{GJHYN6xW}).
+Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{jHwttlTs}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\cite{99MS5xE0}; \cite{J2H8yssV}).
+While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\cite{DNTCUjpC}; \cite{Pa67pUtR}).
 
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\protect\hyperlink{ref-uxWjmlio}{Stefi, 2015}; \protect\hyperlink{ref-v0WwFHgT}{Stefi \& Hess, 2015}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves (\protect\hyperlink{ref-5V3GqSzR}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021b}).
+One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\cite{uxWjmlio}; \cite{v0WwFHgT}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{5V3GqSzR}.
 
@@
 
-The FAIR Digital Object Forum (\protect\hyperlink{ref-DMEBhHID}{{`FAIR Digital Objects Forum \textbar{}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
+The FAIR Digital Object Forum (\cite{DMEBhHID}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
 As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
@@
 
-\textbf{FAIR Digital Object Overview and Specifications} (\protect\hyperlink{ref-11NpETLq1}{Anders et al., 2022a}) is a comprehensive overview of FAIR Digital Object specifications listed below.
+\textbf{FAIR Digital Object Overview and Specifications} \cite{11NpETLq1} is a comprehensive overview of FAIR Digital Object specifications listed below.
 It serves as a primer that introduces FDO concepts and the remaining documents.
 
-The \textbf{FDO Forum Document Standards} (\protect\hyperlink{ref-TQdku4YF}{Weiland et al., 2022b}) documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
+The \textbf{FDO Forum Document Standards} \cite{TQdku4YF} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
 In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
 
-The \textbf{FDO Requirement Specifications} (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) is an update of (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) as the foundational definition of FDO.
+The \textbf{FDO Requirement Specifications} \cite{yygVPoL0} is an update of \cite{RwvirqWg} as the foundational definition of FDO.
 This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
@@
 
-The \textbf{Machine actionability} (\protect\hyperlink{ref-iJeys0T5}{Weiland et al., 2022a}) sets out to define what is meant by \emph{machine actionability} for FDOs.
+The \textbf{Machine actionability} \cite{iJeys0T5} sets out to define what is meant by \emph{machine actionability} for FDOs.
 \emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
@@
 
-\textbf{Configuration Types} (\protect\hyperlink{ref-SPd0uEqO}{Lannom et al., 2022a}) classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022})
+\textbf{Configuration Types} \cite{SPd0uEqO} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
+Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \cite{snykkm7R}
 
-\textbf{PID Profiles \& Attributes} (\protect\hyperlink{ref-ljVViWCl}{Anders et al., 2022b}) specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
+\textbf{PID Profiles \& Attributes} \cite{ljVViWCl} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
+This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \cite{lCoshHXg}.
 This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
 
-\textbf{Kernel Attributes \& Metadata} (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information (\protect\hyperlink{ref-lCoshHXg}{Weigel et al., 2018}).
+\textbf{Kernel Attributes \& Metadata} \cite{Fi62cJAo} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
+This document expands on RDA Recommendation on PID Kernel Information \cite{lCoshHXg}.
 It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
 
-\textbf{Granularity, Versioning, Mutability} (\protect\hyperlink{ref-16W9AycSW}{Working Group, 2022b}) considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
+\textbf{Granularity, Versioning, Mutability} \cite{16W9AycSW} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
 The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
 
-\textbf{DOIP Endorsement Request} (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) is an endorsement of the DOIP v2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) specification as a potential FDO implementation, as it has been applied by several institutions (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}).
+\textbf{DOIP Endorsement Request} \cite{15yGpJ0wh} is an endorsement of the DOIP v2.0 \cite{13TcbsZF6} specification as a potential FDO implementation, as it has been applied by several institutions \cite{jRWurmQc}.
 The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
 
-\textbf{Upload of FDO} (\protect\hyperlink{ref-1H7cmmr69}{Blanchi et al., 2022a}) illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync (\protect\hyperlink{ref-eS5y9TRh}{{`ResourceSync Framework Specification - Table of Contents'}}) is suggested as one type of service to list FDOs.
+\textbf{Upload of FDO} \cite{1H7cmmr69} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
+ResourceSync (\cite{eS5y9TRh}}) is suggested as one type of service to list FDOs.
 This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
 
-\textbf{Typing FAIR Digital Objects} (\protect\hyperlink{ref-ZFzPxCNB}{Lannom et al., 2022b}) defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
+\textbf{Typing FAIR Digital Objects} \cite{ZFzPxCNB} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
 This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
@@
 
-\textbf{Implementation of Attributes, Types, Profiles and Registries} (\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
+\textbf{Implementation of Attributes, Types, Profiles and Registries} \cite{126uxr5pI} details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
 This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} (\protect\hyperlink{ref-1CqIZ47pu}{Schwardmann \& Kálmán, 2022}).
+Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} \cite{1CqIZ47pu}.
 
@@
 
-The concept of \textbf{FAIR Digital Objects} (\protect\hyperlink{ref-IHLT6hye}{Schultes \& Wittenburg, 2019}) has been introduced as way to expose research data as active objects that conform to the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-This builds on the \emph{Digital Object} (DO) concept (\protect\hyperlink{ref-11MnuwJ4l}{Kahn \& Wilensky, 2006}), first introduced in 1995 (\protect\hyperlink{ref-3Uqe3fuK}{Kahn \& Wilensky, 1995}) as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework (\protect\hyperlink{ref-103Hw8H43}{{`X.1255 : Framework for discovery of identity management information'}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}).
+The concept of \textbf{FAIR Digital Objects} \cite{IHLT6hye} has been introduced as way to expose research data as active objects that conform to the FAIR principles \cite{6DjakjNS}.
+This builds on the \emph{Digital Object} (DO) concept \cite{11MnuwJ4l}, first introduced in 1995 \cite{3Uqe3fuK} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
+DO was the inspiration for the ITU X.1255 framework (\cite{103Hw8H43}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\cite{16uB3jxpa} DONA Foundation'}}).
 
@@
 The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing (\protect\hyperlink{ref-tz0P3DTC}{Wittenburg et al., 2019}).
+The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing \cite{tz0P3DTC}.
 
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}), in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}).
+Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \cite{aCye3KpE}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
+Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \cite{RwvirqWg}.
 The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
@@
 
-FDO is an evolving concept.
-A set of FDO Demonstrators (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) highlight how current adapters are approaching implementations of FDO from different angles:
+FDO is an evolving concept. A set of FDO Demonstrators \cite{jRWurmQc} highlight how current adapters are approaching implementations of FDO from different angles:
 
@@
 \item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), see (\protect\hyperlink{ref-xuQKRRx5}{{`DOIP and Examples --- Cordra documentation'}}).} (usually TCP/IP over TLS).
+  Building on the Digital Object concept, using the simplified DOIP v2 specification \cite{13TcbsZF6}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 \cite{13TcbsZF6}, see (\cite{xuQKRRx5}}).} (usually TCP/IP over TLS).
 The main DOIP operations are retrieving, creating and updating digital objects.
 These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}}) and PIDs are assigned using the Handle system.
+FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\cite{15EZ2D0Rm}}) and PIDs are assigned using the Handle system.
 Several type registries have been established.
@@
 
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee (\protect\hyperlink{ref-Guy7cjVY}{Berners-Lee \& Fischetti, 1999}), at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
+The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \cite{Guy7cjVY}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
 The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
@@
 
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\protect\hyperlink{ref-16WGmCih8}{{`RDF 1.1 Primer'}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\protect\hyperlink{ref-16wvFH4nh}{{`Resource Description Framework (RDF) Model and Syntax Specification'}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
+Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\cite{16WGmCih8}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\cite{16wvFH4nh}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
 
-While triple-based knowledge representations were not new (\protect\hyperlink{ref-NxvhwSmG}{Stanczyk, 1987}), the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs (\protect\hyperlink{ref-RRtwPF7Y}{Berners-Lee, Fielding \& Masinter, 2005}) are generalised forms of URLs that include locator-less identifiers
+While triple-based knowledge representations were not new \cite{NxvhwSmG}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \cite{RRtwPF7Y} are generalised forms of URLs that include locator-less identifiers
   such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years (\protect\hyperlink{ref-nW4HY8Nq}{{`"info" URI Registry (Frozen)'}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) (\protect\hyperlink{ref-HqZ3r6J3}{Juty, Le Novere \& Laibe, 2011}).
-RDF 1.1 formats use Unicode to support IRIs (\protect\hyperlink{ref-rqjuXkEf}{Duerst \& Suignard, 2005}), which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
+The distinction between locator-full and locator-less identifiers have weakened in recent years (\cite{nW4HY8Nq}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \cite{HqZ3r6J3}.
+RDF 1.1 formats use Unicode to support IRIs \cite{rqjuXkEf}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
+By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \cite{rnzJ4khD}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
 
@@
 The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} (\protect\hyperlink{ref-yd2WE9eq}{Horrocks \& Hendler, 2002}) gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
+More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \cite{yd2WE9eq} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
 
-Established Semantic Web services and standards include SPARQL (\protect\hyperlink{ref-xww8fb1B}{{`SPARQL 1.1 Overview'}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
+Established Semantic Web services and standards include SPARQL (\cite{xww8fb1B}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
 
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\protect\hyperlink{ref-bMSxa8XW}{Goble \& Stevens, 2008}; \protect\hyperlink{ref-Qlrlr2TZ}{Williams et al., 2012}).
+The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\cite{bMSxa8XW}; \cite{Qlrlr2TZ}).
 In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
@@
 Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) (\protect\hyperlink{ref-FDDmgO6s}{Hu et al., 2011}) .
+Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\cite{15gQDya5B}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
+Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \cite{FDDmgO6s} .
 
 The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS (\protect\hyperlink{ref-191ZkYZMt}{Groth et al., 2014})), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
+This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \cite{191ZkYZMt}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
 
 With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps (\protect\hyperlink{ref-9CdJnsoJ}{Kamdar, Tudorache \& Musen, 2017}).
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication (\protect\hyperlink{ref-AGHR9oyr}{Klein et al., 2014}), for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data (\protect\hyperlink{ref-5SuwLNdE}{Polleres et al., 2020, p.
-handle:20.500.11811/7183}).
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies (\protect\hyperlink{ref-TUmA4KQT}{Carriero et al., 2020}) , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), OBO (\protect\hyperlink{ref-1DalZuKct}{Tirmizi et al., 2011}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})), hash vs slash, HTTP status codes and PID redirection strategies (\protect\hyperlink{ref-rnzJ4khD}{Sauermann, Cyganiak \& Völkel, 2011}).
+This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \cite{9CdJnsoJ}.
+Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}.
+Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\cite{ZwAcGQKY}}), OWL (\cite{1p4IWJpI}}), OBO \cite{1DalZuKct}, SKOS (\cite{15gQDya5B}})), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.
 
@@
 
-The \textbf{Linked Data} concept (\protect\hyperlink{ref-19s3yyxDn}{Bizer, Heath \& Berners-Lee, 2009}) was kickstarted as a set of best practices (\protect\hyperlink{ref-OGXYtGZ8}{{`Linked Data - Design Issues'}}) to bring the Web aspect back into focus.
+The \textbf{Linked Data} concept \cite{19s3yyxDn} was kickstarted as a set of best practices (\cite{OGXYtGZ8}}) to bring the Web aspect back into focus.
 Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
@@
 Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph (\protect\hyperlink{ref-oN9D5LkD}{{`The Open Graph protocol'}}), RDFa (\protect\hyperlink{ref-FKIF6ApB}{{`RDFa 1.1 Primer - Third Edition'}}), Microdata (\protect\hyperlink{ref-14Utscfe4}{{`HTML Standard'}})) we find JSON-LD (\protect\hyperlink{ref-X5SWPqSU}{{`JSON-LD 1.1'}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\protect\hyperlink{ref-j3SqQLBR}{{`Usage Statistics of JSON-LD for Websites, May 2022'}}).
+In addition to several such embedded \emph{microformats} (Open Graph (\cite{oN9D5LkD}}), RDFa (\cite{FKIF6ApB}}), Microdata (\cite{14Utscfe4}})) we find JSON-LD (\cite{X5SWPqSU}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
+JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\cite{j3SqQLBR}}).
 
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\protect\hyperlink{ref-BFMoe8FC}{{`Designing a Linked Data developer experience'}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\protect\hyperlink{ref-GDrm9dgX}{{`Shapes Constraint Language (SHACL)'}}) or ShEx (\protect\hyperlink{ref-YUzogaV}{{`Shape Expressions (ShEx) 2.1 Primer'}})) can be used to validate RDF Data (\protect\hyperlink{ref-ZeOAPNHa}{Gayo et al., 2017}; \protect\hyperlink{ref-S5BI6Mwx}{Thornton et al., 2019}) before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts (\protect\hyperlink{ref-o596f2JM}{Klímek, Škoda \& Nečaský, 2019}), which then tend to prefer non-semantic two-dimensional formats such as CSV files.
+Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\cite{BFMoe8FC}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\cite{GDrm9dgX}}) or ShEx (\cite{YUzogaV}})) can be used to validate RDF Data (\cite{ZeOAPNHa}; \cite{S5BI6Mwx}) before consuming it programmatically, or reshaping data to fit other models.
+While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \cite{o596f2JM}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
 
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}).
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies (\protect\hyperlink{ref-MhsCt4P3}{Wolstencroft et al., 2011}), the use of Linked Data as part of REST Web APIs (\protect\hyperlink{ref-eS6a6tlz}{Page, De Roure \& Martinez, 2011}) , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy (\protect\hyperlink{ref-117tX9MEG}{2016}), with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
+A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \cite{jHwttlTs}.
+One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \cite{MhsCt4P3}, the use of Linked Data as part of REST Web APIs \cite{eS6a6tlz} , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy \cite{117tX9MEG}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
 
@@
 \item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}), which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
+  An Interoperability Framework and Distributed Platform for Fast Data Applications \cite{KlCFFFyL}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
 \item
-  The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), validated against its implementations for completeness.
+  The FAIR Digital Object guidelines \cite{RwvirqWg}, validated against its implementations for completeness.
 \item
-  A Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}), which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
+  A Comparison Framework for Middleware Infrastructures \cite{hRzcHhPD}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
 \item
-  Cross-checks against RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
+  Cross-checks against RDA's FAIR Data Maturity Model \cite{UzQhqk0M} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
 \item
-  EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
+  EOSC Interoperability Framework \cite{aCye3KpE} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
 \end{enumerate}
@@
 Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) against Linked Data in general.
+For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \cite{13TcbsZF6} against Linked Data in general.
 
@@
 
-The Interoperability Framework for Fast Data Applications (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
+The Interoperability Framework for Fast Data Applications \cite{KlCFFFyL} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
 
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model (\protect\hyperlink{ref-8ZRGFa6y}{Stallings, 1990}) better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
+We have chosen to investigate using this framework as it covers the higher levels of the OSI Model \cite{8ZRGFa6y} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
 In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
 
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}).
-\label{tbl:fdo-web-interoperability-framework}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
-FDO encourages research communities to develop common type registries to be shared across instances.
-In current DOIP practice, each service have their own types, attributes and operations.
-The wider symbiosis is consistent use of PIDs.
-& The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}) end up being mandated centrally by dominant (often commercial) providers (\protect\hyperlink{ref-11VxBOeZy}{Fielding et al., 2017}), which clients are required to use as-is with special code per service.
-Use of Linked Data enables common tooling and semantic mapping across differences.
-\\
-\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP).
-\texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).
-Semantics/purpose of operations not formalised (similar operations can be grouped with \texttt{0.DOIP/OperationReference}).
-& ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), schema.org Actions (\protect\hyperlink{ref-dKAekUjL}{{`Schema.org Actions - schema.org'}}), WSDL/SOAP (\protect\hyperlink{ref-LmudecyN}{{`Web Services Description Language (WSDL) Version 2.0 Part 0: Primer'}}), but frequently just as human-readable developer documentation/examples.
-\\
-\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing.
-Every FDO has a type.
-Types maintained in FDO Type registries, which may add additional semantics, e.g.~the ePIC \href{https://hdl.handle.net/21.11104/c1a0ec5ad347427f25d6}{PID-InfoType for Model}.
-No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) & Lightweight HTTP semantics for authenticity/navigation.
-Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
-Semantic of type implied by Linked Data formats (e.g.~OWL2, RDFS), although choice of type system may not be explicit.
-\\
-\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation} & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema.
-Multiple byte stream attachments of any media type.
-& Textual HTTP headers (including any signposting), single byte stream of any media type, e.g.~Linked Data formats (JSON-LD, Turtle, RDF/XML) or embedded in document (HTML with RDFa, JSON-LD or Microdata).
-XML was previously the main syntax used by APIs, JSON is now dominant.
-\\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
-\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
-Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable).
-Unique DOIP protocol can be hard to add to existing Web application frameworks, although proxy services have been developed (e.g.~B2SHARE adapter).
-& HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to (for static content) multi-cloud Content-Delivery Networks (CDN).
-Current scalable cloud technologies for Web hosting may not support HTTP features previously seen as important for Semantic Web, e.g.~content negotiation and semantic HTTP status codes.
-\\
-\bottomrule
-\end{longtable}
+\include{table1}
 
@@
 
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services (\protect\hyperlink{ref-39K3yHT3}{Wolstencroft et al., 2013}).
-However, this is mostly done via developer or human interaction (\protect\hyperlink{ref-iYh1wFBt}{Lamprecht et al., 2021}).
+The Web has already showed us how one can compose workflows of hetereogeneous Web Services \cite{39K3yHT3}.
+However, this is mostly done via developer or human interaction \cite{iYh1wFBt}.
 Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
@@
 These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\protect\hyperlink{ref-hxtgCGjY}{{`The Modern Standards Paradigm - Five Key Principles'}}) to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference (\protect\hyperlink{ref-jrZe6Esu}{Loo, 1970}).
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\protect\hyperlink{ref-p7nwRj8E}{{`DOIP API for HTTP Clients --- Cordra documentation'}}).
+The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\cite{hxtgCGjY}}) to embrace openness, transparency, and broad consensus.
+In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \cite{jrZe6Esu}.
+This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\cite{p7nwRj8E}}).
 
 It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\protect\hyperlink{ref-68fzID2k}{{`DOI Handbook - Resolution'}}).
+For example, DOIs are mostly resolved on the web (\cite{1H9iQhQYq}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\cite{68fzID2k}}).
 
@@
 
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}).
+The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \cite{8HJqcF1Q}.
 
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2923}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2769}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
-\label{tbl:metamodel-concepts}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endhead
-Resource & FDO/DO & Resource \\
-Service & DOIP service & Server/endpoint \\
-Transaction & (not supported) & Conditional requests, \texttt{409\ Conflict} \\
-Process & Extended operations & (primarily stateless), \texttt{100\ Continue}, \texttt{202\ Accepted} \\
-Operation & DOIP Operation & Method, query parameters \\
-Request & DOIP Request & Request \\
-Response & DOIP Response & Response \\
-Message & Segment, \texttt{requestId} & Message, Representation \\
-Channel & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures. & TCP/IP, TLS, UDP \\
-Protocol & DOIP 2.0, ++ & HTTP/1.1, HTTP/2, HTTP/3 \\
-Link & PID/Handle & URL \\
-\bottomrule
-\end{longtable}
+\include{table2}
 
@@
 
-The FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) and using Linked Data Platform (\protect\hyperlink{ref-17OHlMRQA}{Linked Data Platform Working Group, 2015}), as proposed by (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}).
+The FAIR Digital Object guidelines \cite{RwvirqWg} sets out recommendations for FDO implementations.
+In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \cite{13TcbsZF6} and using Linked Data Platform \cite{17OHlMRQA}, as proposed by (\cite{7szz7dwO}}).
 
-Note that the draft update to FDO specification (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer (\protect\hyperlink{ref-yygVPoL0}{Strawn et al., 2022}) renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
+\include{table3}
+
+Note that the draft update to FDO specification \cite{yygVPoL0} (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer \cite{yygVPoL0} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
 
@@
 Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement (\protect\hyperlink{ref-15yGpJ0wh}{Working Group, 2022a}) sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
+The DOIP Endorsement \cite{15yGpJ0wh} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
 Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
@@
   G1 and G2 call for stability and trustworthiness.
-WhileC the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
+While the foundations of both DOIP and Linked Data approaches are now well established -- the FDO requirements and in particular how they can be implemented are still taking shape and subject to change.
 \item
@@
 \item
-  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-uCse.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}})
+  While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
+Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\cite{9T4j3N4e}})
 \item
@@
 
-We argue that the Web, with its now ubiquitous use of REST API (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services (\protect\hyperlink{ref-WzPECPDP}{Fensel et al., 2011}) attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\protect\hyperlink{ref-11dInoyVF}{{`5-star Open Data'}}).
+We argue that the Web, with its now ubiquitous use of REST API \cite{174AwcFUL}, can be compared as a similar global middleware.
+Note that while early moves for developing Semantic Web Services \cite{WzPECPDP} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\cite{11dInoyVF}}).
 
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004}) that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
+For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures \cite{hRzcHhPD} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
 
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018})) and Web technologies (using Linked Data) as middleware infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004})
-\label{tbl:fdo-web-middleware}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
-Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
-PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}.
-& The Web is inherently open and made by cross-linked URLs.
-Participation requires DNS domain purchase (many free alternatives also exists).
-PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects.
-\\
-\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
-HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
-\\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
-DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
-Multiple connections possible.
-Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks (\protect\hyperlink{ref-F8VZ86hu}{Sandvine}).
-HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}) but still has TCP+TLS startup costs.
-For reduced latency , HTTP/3 (\protect\hyperlink{ref-Amit86Tp}{{`draft-ietf-quic-http-34'}}) use QUIC (\protect\hyperlink{ref-1GWvyEmUf}{Iyengar \& Thomson, 2021})) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
-\\
-\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
-FDOs should provide the mandatory kernel metadata fields.
-FDOs of the same declared type typically share additional attributes (although that schema may not be declared).
-DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions.
-& Each URL accessed separately.
-Common HTTP headers provide basic metadata, although it is often not reliable.
-A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), content-negotiation (\protect\hyperlink{ref-n2f3TfoI}{{`Content negotiation - HTTP \textbar{} MDN'}}) or FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}).
-\\
-\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
-Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
-& Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
-\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
-Multiple locations possible in Handle system, can expose geo-info.
-& PIDs and URL redirects.
-DNS aliases and IP routing can hide location.
-Geo-localised servers common for large cloud deployments.
-\\
-\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
-FDO kernel metadata can include checksum and date.
-& HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
-Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
-\\
-\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
-Custom codes uncommon.
-\\
-\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
-No redirection from DOIP service.
-& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
-\\
-\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
-There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged (\protect\hyperlink{ref-rbG9uRKw}{{`Hypertext Style: Cool URIs don't change.'}}).
-Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
-An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento (\protect\hyperlink{ref-DNjcMi4a}{Van de Sompel, Nelson \& Sanderson, 2013}) expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method (\protect\hyperlink{ref-hRUsaTiz}{Clemm et al., 2002}) (used by SVN).
-\\
-\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
-Internal synchronisation possible in backend for Extended operations.
-& Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} (\protect\hyperlink{ref-xvMAyRAc}{Dusseault, 2007}) with \texttt{LOCK} and \texttt{UNLOCK} methods.
-\\
-\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
-In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
-& The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space.
-In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend.
-\\
-\textbf{Encapsulation}: \emph{separate interface from implementation.
-Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
-FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
-Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade (\protect\hyperlink{ref-1B16WfR7q}{Khare \& Lawrence, 2000}) to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns (\protect\hyperlink{ref-iME5l1kk}{{`Hypertext Style: Cool URIs don't change.'}}) and PIDs via Persistent URLs (PURL).
-\\
-\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
-FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})) & Syntactically Media Type with multiple suffixes (\protect\hyperlink{ref-nlH2rm1s}{{`draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes'}}) (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) (\protect\hyperlink{ref-nCzRO0pK}{Wilde, 2013}).
-In metadata, semantic type systems (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL2 (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})).
-OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} (\protect\hyperlink{ref-11oVTMsuZ}{{`W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures'}}), JSON \texttt{\$schema} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})), JSON-LD \texttt{@context} (\protect\hyperlink{ref-by6CfOmv}{{`JSON-LD 1.1'}}).
-Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
-\\
-\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
-\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 (\protect\hyperlink{ref-tIkSXrb5}{Fielding \& Reschke, 2014b}) \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) (\protect\hyperlink{ref-PL512B0X}{May, 2017}), MPEG-DASH (ISO/IEC 23009-1:2019) (\protect\hyperlink{ref-ikO7e0mh}{14:00-17:00}) \\
-\bottomrule
-\end{longtable}
+\include{table4}
 
@@
 \item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP (\protect\hyperlink{ref-16uB3jxpa}{{`Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation'}}) supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}).
+  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP (\cite{16uB3jxpa} DONA Foundation'}}) supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
+Multiplexing was removed for the much simplified DOIP 2.0 \cite{13TcbsZF6}.
 Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
@@
 
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}) also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\protect\hyperlink{ref-PR-RequirementSpec-2.0}{\textbf{PR-RequirementSpec-2.0?}}).
+In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \cite{RwvirqWg} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\cite{PR-RequirementSpec-2.0}}).
 
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) comply with the FAIR principles (\protect\hyperlink{ref-6DjakjNS}{Wilkinson et al., 2016}).
-Here we've used the RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}) as it has decomposed the FAIR principles to a structured list of FAIR indicators (\protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}), importantly considering \emph{Data} and \emph{Metadata} separately.
+Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\cite{7szz7dwO}}) comply with the FAIR principles \cite{6DjakjNS}.
+Here we've used the RDA's FAIR Data Maturity Model \cite{1GrKTFaK2} as it has decomposed the FAIR principles to a structured list of FAIR indicators \cite{UzQhqk0M}, importantly considering \emph{Data} and \emph{Metadata} separately.
 In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
 
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}; \protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) (first 2 columns) against the FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), FDO implemented with the protocol DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) and examples from Linked Data practices in general.
-(--- indicates \emph{Unspecified}, may be possible with additional conventions)
-\label{tbl:fair-data-maturity-model}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
-\midrule
-\endhead
-RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
-RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
-Handle, DOI.
-& FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
-RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
-FAIR Signposting.
-\\
-RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
-RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
-Unclear how to link to additional metadata.
-& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
-Embedded in landing page (RDFa).
-\\
-RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
-& PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
-RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
-Handle registry only searchable by PID.
-& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-(\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}, pp.
-https://lod--cloud.net/})).
-Indexing by search engines if exposing HTML w/schema.org.
-\\
-RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
-\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
-RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
-RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
-RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
-RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
-RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
-Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}) & HTTP/HTTPS, FTP (now less common), GridFTP (\protect\hyperlink{ref-18LsXvKGJ}{Allcock et al.}) (for large data), ARK (\protect\hyperlink{ref-voDQ6diM}{{`The ARK Identifier Scheme'}}) \\
-RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol (\protect\hyperlink{ref-I9Kx0Hjy}{Sun et al., 2003}).
-One server implementation (\protect\hyperlink{ref-1CvYRVhy2}{{`Handle.Net Registry'}}), free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved (\protect\hyperlink{ref-xVJMmGVr}{{`Licenses \& Standards \textbar{} Open Source Initiative'}}) as an open source license -- it includes usage restrictions and requires Service Agreements.
-It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
-One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
-\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
-& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
-\\
-RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
-RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
-RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
-RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
-RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
-RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
-RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
-RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
-& FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) (\protect\hyperlink{ref-1BlxdxOwa}{14:00-17:00}) \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs (\protect\hyperlink{ref-o4h92LpS}{{`Data - W3C'}})), \texttt{rdfs:seeAlso}, FAIR signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) \texttt{describedby} \\
-RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
-RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
-RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles (\protect\hyperlink{ref-QpDmajc4}{{`Linking Across Provenance Bundles'}}), \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role (\protect\hyperlink{ref-10HA0mMBx}{Unknown}), \texttt{prov:specializationOf} (\protect\hyperlink{ref-ZUgwvAHE}{{`PROV-O: The PROV Ontology'}}), OAI-ORE Proxy (\protect\hyperlink{ref-ID1pTGry}{{`ORE Specification - Abstract Data Model'}})) \\
-RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) but not assigned PIDs yet.
-& Unspecified.
-Multiple metadata records can allow multiple semantic profiles.
-& Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org (\protect\hyperlink{ref-10pzKCEUE}{{`Schema.org - Schema.org'}}), Dublin Core Terms (\protect\hyperlink{ref-wAbZpOsr}{{`DCMI Metadata Terms'}}), DCAT (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), FOAF (\protect\hyperlink{ref-XmUY0cZB}{{`FOAF Vocabulary Specification'}}) \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}) \\
-RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
-RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
-RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
-PID Kernel attributes?
-Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}}), PAV (\protect\hyperlink{ref-1E4OfHyhI}{Ciccarese et al., 2013}), Dublin Core Terms (\protect\hyperlink{ref-hKblceo6}{{`DCMI Metadata Terms'}, 2020}) \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen (\protect\hyperlink{ref-zZTtJruI}{Hardisty et al., 2022})) & --- & Common, e.g.~DCAT 2 (\protect\hyperlink{ref-zt8AAjcZ}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), BioSchemas (\protect\hyperlink{ref-1F8S9FeHf}{{`Bioschemas - Bioschemas'}}) \\
-RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal (\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}}), Darwin Core (\protect\hyperlink{ref-9jaVzCbL}{Wieczorek et al., 2012}) \\
-RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
-\bottomrule
-\end{longtable}
+\include{table5}
 
@@
 \item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) has currently only a couple of implementations and is expressed informally.
+  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification \cite{13TcbsZF6} has currently only a couple of implementations and is expressed informally.
 The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
@@
   Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-(\protect\hyperlink{ref-126uxr5pI}{Blanchi et al., 2022b}) explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
+\cite{126uxr5pI} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
 Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
@@
 For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
+FAIR Signposting \cite{snykkm7R} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
 In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
@@
 
-The EOSC Interoperability Framework (\protect\hyperlink{ref-aCye3KpE}{Corcho et al., 2021}) in section 3.6 recommends:
+The EOSC Interoperability Framework \cite{aCye3KpE} in section 3.6 recommends:
 
-\begin{tablenos:no-prefix-table-caption}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
-\toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Layer
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Recommendation
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-FDO
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Linked Data
-\end{minipage} \\
-\midrule
-\endhead
-Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
-Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
-Technical & Easy SLAs for service providers & Unspecified & None \\
-Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
-Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
-Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
-Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
-Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
-Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
-Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
-Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
-Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
-Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
-Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
-Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
-Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
-Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
-Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
-Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
-Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
-Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
-Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
-Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
-Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
-Legal & List recommended licenses & Undefined & Best practice recommendations \\
-Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
-Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
-Legal & GDPR compliance for personal data & Undefined & Undefined \\
-Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
-Legal & Harmonised terms-of-use & Undefined & Undefined \\
-Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
-\bottomrule
-\end{longtable}
-
-\end{tablenos:no-prefix-table-caption}
+\include{table6}
 
@@
 The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\protect\hyperlink{ref-kYjzRJ0H}{{`Semantic Web - XML2000 - slide "Architecture"'}}).
+At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\cite{kYjzRJ0H}}).
 However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
@@
 
-\ldots{}
+%\include{bibliography}
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
+\section*{Methods}
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
-. \emph{Available at} \href{https://blog.cloudflare.com/http-3-vs-http-2/}{\emph{https://blog.cloudflare.com/http-3-vs-http-2/}}
+\section*{Results}
 
-\leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
-. \emph{Available at} \href{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}{\emph{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
+\section*{Discussion}
 
-\leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 January 2023b).
+\section*{Conclusions}
 
-\leavevmode\vadjust pre{\hypertarget{ref-ikO7e0mh}{}}%
-14:00-17:00.ISO/IEC 23009-1:2019.
-\emph{Available at} \href{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}{\emph{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}} (accessed 26 May 2022a).
+\section*{Acknowledgments}
 
-\leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
+So long and thanks for all the fish.
 
-\leavevmode\vadjust pre{\hypertarget{ref-18LsXvKGJ}{}}%
-Allcock W, Bresnahan J, Kettimuthu R, Link M.
-The Globus Striped GridFTP Framework and Server.
-In: \emph{ACM/IEEE SC 2005 Conference (SC'05)}.
-IEEE,.
-DOI: \href{https://doi.org/10.1109/sc.2005.72}{10.1109/sc.2005.72}.
+%%\bibliography{sample}
+\printbibliography
 
-\leavevmode\vadjust pre{\hypertarget{ref-11NpETLq1}{}}%
-Anders I, Blanchi Ch, Broeder D, Hellström M, Islam Sh, Jejkal Th, Lannom L, Peters K, Quick R, Schlemmer A, Schwardmann U, Soiland-Reyes S, Strawn G, Uytvanck D van, Weiland C, Wittenburg P, Zwölf C.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1GAj-1owAAPDF7hVis2dYPPCOHiPHrIXIo4j-S5AdEFI/edit}{FAIR digital object overview and specifications}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ljVViWCl}{}}%
-Anders I, Hellström M, Islam S, Jejkal T, Lannom L, Schwardmann U, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1c2mZziq5pIPmLxMHLcYqlWrjYsc2ezGMXvp0E46iljo/edit}{FDO PID profiles \& attributes}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-UzQhqk0M}{}}%
-Bahim C, Casorrán-Amilburu C, Dekkers M, Herczog E, Loozen N, Repanas K, Russell K, Stall S.
-2020.
-The FAIR data maturity model: An approach to harmonise FAIR assessments.
-\emph{Data Science Journal} 19.
-DOI: \href{https://doi.org/10.5334/dsj-2020-041}{10.5334/dsj-2020-041}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-KbbW0kGT}{}}%
-Belshe M, Peon R. 2015. \emph{Hypertext Transfer Protocol Version 2 (HTTP/2)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7540}{10.17487/rfc7540}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RRtwPF7Y}{}}%
-Berners-Lee T, Fielding R, Masinter L. 2005. \emph{Uniform Resource Identifier (URI): Generic Syntax}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3986}{10.17487/rfc3986}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Guy7cjVY}{}}%
-Berners-Lee T, Fischetti M. 1999. \emph{Weaving the Web: the original design and ultimate destiny of the World Wide Web by its inventor}. San Francisco: HarperSanFrancisco.
-
-\leavevmode\vadjust pre{\hypertarget{ref-117tX9MEG}{}}%
-Bernstein A, Hendler J, Noy N. 2016. A new look at the semantic web. \emph{Communications of the ACM} 59:35--37. DOI: \href{https://doi.org/10.1145/2890489}{10.1145/2890489}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
-Bioschemas - Bioschemas. \emph{Available at} \href{http://bioschemas.org/}{\emph{http://bioschemas.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-19s3yyxDn}{}}%
-Bizer C, Heath T, Berners-Lee T.
-2009.
-Linked data - the story so far.
-\emph{International journal on Semantic Web and information systems} 5:1--22.
-DOI: \href{https://doi.org/10.4018/jswis.2009081901}{10.4018/jswis.2009081901}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H7cmmr69}{}}%
-Blanchi C, Broeder D, Jejkal T, Sharif I, Schlemmer A, Uytvanck D van, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1fDR5VHbVIa2AbLsBR58idrfn_Ib3x6Fk-_LJ4c_Ftt4/edit}{FDO -- upload of FDO}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-126uxr5pI}{}}%
-Blanchi C, Hellström M, Lannom L, Schwardmann U, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1RrOiwMhkl-hRzWmlluA2iXCzHK-bj7_80LlMXgWx4w4/edit}{Implementation of attributes, types, profiles and registries}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-EJDjHrUv}{}}%
-Bonino Da Silva Santos LO, Wilkinson MD, Kuzniar A, Kaliyaperumal R, Thompson M, Dumontier M, Burger K.
-2016.
-\href{https://www.researchgate.net/publication/\%7B309468587/_FAIR/_Data/_Points/_Supporting/_Big/_Data/_Interoperability\%7D}{FAIR data points supporting big data interoperability}.
-In: Zelm M, Doumeingts G, Mendonça JP eds.
-\emph{Enterprise interoperability in the digitized and networked factory of the future}.
-iSTE Press, 270--279.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RwvirqWg}{}}%
-Bonino L, Wittenburg O, Carroll B, Hardisty A, Leggott M, Zwölf C.
-2019.
-\emph{\href{https://github.com/GEDE-RDA-Europe/GEDE/blob/master/FAIR\%20Digital\%20Objects/FDOF/FAIR\%20Digital\%20Object\%20Framework-v1-02.docx}{FAIR digital object framework}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TUmA4KQT}{}}%
-Carriero VA, Daquino M, Gangemi A, Nuzzolese AG, Peroni S, Presutti V, Tomasi F.
-2020.
-The landscape of ontology reuse approaches.
-In: Cota G, Daquino M, Pozzato GL eds.
-\emph{Applications and practices in ontology design, extraction, and reasoning}.
-Studies on the semantic web.
-IOS Press,.
-DOI: \href{https://doi.org/10.3233/SSW200033}{10.3233/SSW200033}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1E4OfHyhI}{}}%
-Ciccarese P, Soiland-Reyes S, Belhajjame K, Gray AJ, Goble C, Clark T.
-2013.
-PAV ontology: provenance, authoring and versioning.
-\emph{Journal of Biomedical Semantics} 4:37.
-DOI: \href{https://doi.org/10.1186/2041-1480-4-37}{10.1186/2041-1480-4-37}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRUsaTiz}{}}%
-Clemm G, Amsden J, Ellison T, Kaler C, Whitehead J.
-2002.
-\emph{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}.
-RFC Editor.
-DOI: \href{https://doi.org/10.17487/rfc3253}{10.17487/rfc3253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-n2f3TfoI}{}}%
-Content negotiation - HTTP \textbar{} MDN.
-\emph{Available at} \href{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}{\emph{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content\_negotiation}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-aCye3KpE}{}}%
-Corcho O, Eriksson M, Kurowski K, Ojsteršek M, Choirat C, Sanden M van de, Coppens F.
-2021.
-\emph{EOSC interoperability framework}.
-Publications Office of the EU.
-DOI: \href{https://doi.org/10.2777/620649}{10.2777/620649}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
-Data - W3C. \emph{Available at} \href{https://www.w3.org/standards/semanticweb/data}{\emph{https://www.w3.org/standards/semanticweb/data}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2. \emph{Available at} \href{https://www.w3.org/TR/vocab-dcat-2/}{\emph{https://www.w3.org/TR/vocab-dcat-2/}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
-Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-6LvydTsC}{}}%
-Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hKblceo6}{}}%
-DCMI Metadata Terms.
-2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
-DCMI Metadata Terms.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-KlCFFFyL}{}}%
-Delgado JCM.
-2016.
-An Interoperability Framework and Distributed Platform for Fast Data Applications.
-In: \emph{Data Science and Big Data Computing}.
-Springer International Publishing, 3--39.
-DOI: \href{https://doi.org/10.1007/978-3-319-31861-5_1}{10.1007/978-3-319-31861-5\_1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-BFMoe8FC}{}}%
-Designing a Linked Data developer experience.
-2018.
-\emph{Available at} \href{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}{\emph{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
-Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation. \emph{Available at} \href{https://www.dona.net/doipv1doc}{\emph{https://www.dona.net/doipv1doc}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
-DOIP and Examples --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip.html}{\emph{https://www.cordra.org/documentation/api/doip.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-p7nwRj8E}{}}%
-DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-15EZ2D0Rm}{}}%
-draft-bhutton-json-schema-00.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}{\emph{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlH2rm1s}{}}%
-draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}{\emph{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
-draft-ietf-quic-http-34.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}{\emph{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
-draft-kelly-json-hal-08.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}{\emph{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rqjuXkEf}{}}%
-Duerst M, Suignard M. 2005. \emph{Internationalized Resource Identifiers (IRIs)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3987}{10.17487/rfc3987}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xvMAyRAc}{}}%
-Dusseault L (ed.). 2007. \emph{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc4918}{10.17487/rfc4918}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
-FAIR Digital Object Framework Documentation. \emph{Available at} \href{https://fairdigitalobjectframework.org/}{\emph{https://fairdigitalobjectframework.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
-FAIR Digital Objects Forum \textbar. \emph{Available at} \href{https://fairdo.org/}{\emph{https://fairdo.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-WzPECPDP}{}}%
-Fensel D, Facca FM, Simperl E, Toma I. 2011. \emph{Semantic Web Services}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/978-3-642-19193-0}{10.1007/978-3-642-19193-0}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-174AwcFUL}{}}%
-Fielding RT.
-2000.
-\href{https://www.ics.uci.edu//~fielding/pubs/dissertation/top.htm}{Architectural styles and the design of network-based software architectures}.
-\{THESIS\}.\{DOCTORAL\} Thesis.
-University of California, Irvine.
-
-\leavevmode\vadjust pre{\hypertarget{ref-8HJqcF1Q}{}}%
-Fielding R, Nottingham M, Reschke J (eds.). 2022. \emph{HTTP Semantics}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9110}{10.17487/rfc9110}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tIkSXrb5}{}}%
-Fielding R, Reschke J (eds.). 2014b. \emph{Hypertext Transfer Protocol (HTTP/1.1): Message Syntax and Routing}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7230}{10.17487/rfc7230}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-HE7Ikwwl}{}}%
-Fielding R, Reschke J (eds.). 2014a. \emph{Hypertext Transfer Protocol (HTTP/1.1): Semantics and Content}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7231}{10.17487/rfc7231}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11VxBOeZy}{}}%
-Fielding RT, Taylor RN, Erenkrantz JR, Gorlick MM, Whitehead J, Khare R, Oreizy P.
-2017.
-Reflections on the REST architectural style and "principled design of the modern web architecture" (impact paper award).
-In: \emph{Proceedings of the 2017 11th joint meeting on foundations of software engineering - ESEC/FSE 2017}.
-New York, New York, USA: ACM Press, 4--14.
-DOI: \href{https://doi.org/10.1145/3106237.3121282}{10.1145/3106237.3121282}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
-FOAF Vocabulary Specification. \emph{Available at} \href{http://xmlns.com/foaf/spec/}{\emph{http://xmlns.com/foaf/spec/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-13TcbsZF6}{}}%
-Foundation D. 2018. \emph{\href{https://hdl.handle.net/0.DOIP/DOIPV2.0}{Digital object interface protocol specification, version 2.0}}. DONA foundation.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZeOAPNHa}{}}%
-Gayo JEL, Prud'hommeaux E, Boneva I, Kontokostas D.
-2017.
-Validating RDF Data.
-\emph{Synthesis Lectures on the Semantic Web: Theory and Technology} 7:1--328.
-DOI: \href{https://doi.org/10.2200/s00786ed1v01y201707wbe016}{10.2200/s00786ed1v01y201707wbe016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-bMSxa8XW}{}}%
-Goble C, Stevens R.
-2008.
-State of the nation in data integration for bioinformatics.
-\emph{Journal of Biomedical Informatics} 41:687--693.
-DOI: \href{https://doi.org/10.1016/j.jbi.2008.01.008}{10.1016/j.jbi.2008.01.008}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-IbaebabD}{}}%
-Gregorio J, de hOra B (eds.). 2007. \emph{The Atom Publishing Protocol}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc5023}{10.17487/rfc5023}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11hxwwuRt}{}}%
-Gregorio J, Fielding R, Hadley M, Nottingham M, Orchard D. 2012. \emph{URI Template}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6570}{10.17487/rfc6570}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-191ZkYZMt}{}}%
-Groth P, Loizou A, Gray AJG, Goble C, Harland L, Pettifer S.
-2014.
-API-centric Linked Data integration: The Open PHACTS Discovery Platform case study.
-\emph{Journal of Web Semantics} 29:12--18.
-DOI: \href{https://doi.org/10.1016/j.websem.2014.03.003}{10.1016/j.websem.2014.03.003}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GrKTFaK2}{}}%
-Group RDAFDMMW. 2020. FAIR data maturity model: Specification and guidelines. \emph{Research Data Alliance}. DOI: \href{https://doi.org/10.15497/rda00050}{10.15497/rda00050}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
-Hardisty A, Brack P, Goble C, Livermore L, Scott B, Groom Q, Owen S, Soiland-Reyes S. 2022. The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
-Digital Object Approach to Speeding up Digital Mobilisation of Natural History
-Collections. \emph{Data Intelligence} 4:320--341. DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-GJHYN6xW}{}}%
-Hasnain A, Rebholz-Schuhmann D.
-2018.
-Assessing FAIR data principles against the 5-star open data principles.
-In: Gangemi A, Gentile AL, Nuzzolese AG, Rudolph S, Maleshkova M, Paulheim H, Pan JZ, Alam M eds.
-\emph{The semantic web: ESWC 2018 satellite events: ESWC 2018 satellite events, heraklion, crete, greece, june 3-7, 2018, revised selected papers}.
-Lecture notes in computer science.
-Cham: Springer International Publishing, 469--477.
-DOI: \href{https://doi.org/10.1007/978-3-319-98192-5_60}{10.1007/978-3-319-98192-5\_60}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yd2WE9eq}{}}%
-Horrocks I, Hendler J (eds.). 2002. \emph{The Semantic Web --- ISWC 2002}. Springer Berlin Heidelberg. DOI: \href{https://doi.org/10.1007/3-540-48005-6}{10.1007/3-540-48005-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
-HTML Standard. \emph{Available at} \href{https://html.spec.whatwg.org/multipage/microdata.html}{\emph{https://html.spec.whatwg.org/multipage/microdata.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-FDDmgO6s}{}}%
-Hu W, Chen J, Zhang H, Qu Y.
-2011.
-How matchable are four thousand ontologies on the semantic web.
-In: Antoniou G, Grobelnik M, Simperl E, Parsia B, Plexousakis D, De Leenheer P, Pan J eds.
-\emph{The semantic web: Research and applications}.
-Lecture notes in computer science.
-Berlin, Heidelberg: Springer Berlin Heidelberg, 290--304.
-DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI}{\emph{https://www.w3.org/Provider/Style/URI}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI.html}{\emph{https://www.w3.org/Provider/Style/URI.html}} (accessed 26 May 2022b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1GWvyEmUf}{}}%
-Iyengar J, Thomson M (eds.). 2021. \emph{QUIC: A UDP-Based Multiplexed and Secure Transport}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc9000}{10.17487/rfc9000}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
-JSON-LD 1.1. \emph{Available at} \href{http://www.w3.org/TR/json-ld/}{\emph{http://www.w3.org/TR/json-ld/}} (accessed 26 May 2022b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
-JSON-LD 1.1. \emph{Available at} \href{https://www.w3.org/TR/json-ld/}{\emph{https://www.w3.org/TR/json-ld/}} (accessed 26 May 2022a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-HqZ3r6J3}{}}%
-Juty N, Le Novere N, Laibe C.
-2011.
-Identifiers.org and MIRIAM Registry: community resources to provide persistent identification.
-\emph{Nucleic Acids Research} 40:D580--D586.
-DOI: \href{https://doi.org/10.1093/nar/gkr1097}{10.1093/nar/gkr1097}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-3Uqe3fuK}{}}%
-Kahn R, Wilensky R. 1995. \emph{\href{http://www.cnri.reston.va.us/k-w.html}{A framework for distributed digital object services}}. CNRI.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11MnuwJ4l}{}}%
-Kahn R, Wilensky R.
-2006.
-A framework for distributed digital object services.
-\emph{International Journal on Digital Libraries} 6:115--123.
-DOI: \href{https://doi.org/10.1007/s00799-005-0128-x}{10.1007/s00799-005-0128-x}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9CdJnsoJ}{}}%
-Kamdar MR, Tudorache T, Musen MA.
-2017.
-A systematic analysis of term reuse and term overlap across biomedical ontologies.
-\emph{Semantic Web} 8:853--871.
-DOI: \href{https://doi.org/10.3233/sw-160238}{10.3233/sw-160238}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1B16WfR7q}{}}%
-Khare R, Lawrence S. 2000. \emph{Upgrading to TLS Within HTTP/1.1}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc2817}{10.17487/rfc2817}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-AGHR9oyr}{}}%
-Klein M, Van de Sompel H, Sanderson R, Shankar H, Balakireva L, Zhou K, Tobin R.
-2014.
-Scholarly Context Not Found: One in Five Articles Suffers from Reference Rot.
-\emph{PLoS ONE} 9:e115253.
-DOI: \href{https://doi.org/10.1371/journal.pone.0115253}{10.1371/journal.pone.0115253}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-o596f2JM}{}}%
-Klímek J, Škoda P, Nečaský M. 2019. Survey of tools for linked data consumption. \emph{Satya Widya} 10:665--720. DOI: \href{https://doi.org/10.3233/SW-180316}{10.3233/SW-180316}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iYh1wFBt}{}}%
-Lamprecht A-L, Palmblad M, Ison J, Schwämmle V, Al Manir MS, Altintas I, Baker CJO, Ben Hadj Amor A, Capella-Gutierrez S, Charonyktakis P, Crusoe MR, Gil Y, Goble C, Griffin TJ, Groth P, Ienasescu H, Jagtap P, Kalaš M, Kasalica V, Khanteymoori A, Kuhn T, Mei H, Ménager H, Möller S, Richardson RA, Robert V, Soiland-Reyes S, Stevens R, Szaniszlo S, Verberne S, Verhoeven A, Wolstencroft K.
-2021.
-Perspectives on automated composition of workflows in the life sciences.
-\emph{F1000Research} 10:897.
-DOI: \href{https://doi.org/10.12688/f1000research.54159.1}{10.12688/f1000research.54159.1}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SPd0uEqO}{}}%
-Lannom L, Peters-von Gehlen K, Anders I, Pfeil A, Schlemmer A, Trautt Z, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1ivvg3C_QWSO9PIQwkKT89xG4fBhNAs7_6b0Dz11EwDg/edit}{FDO configuration types}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZFzPxCNB}{}}%
-Lannom L, Schwardmann U, Blanchi C, Wittenburg P.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1X0hcOVIqP7iYIJf9u-7x3RwcXK8ecsauy0FZg_6-Bg0/edit}{Typing FAIR digital objects}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
-Linked Data - Design Issues. \emph{Available at} \href{https://www.w3.org/DesignIssues/LinkedData.html}{\emph{https://www.w3.org/DesignIssues/LinkedData.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
-Linked Data Notifications. \emph{Available at} \href{https://www.w3.org/TR/ldn/}{\emph{https://www.w3.org/TR/ldn/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-17OHlMRQA}{}}%
-Linked Data Platform Working Group. 2015. \emph{\href{https://www.w3.org/TR/2015/REC-ldp-20150226/}{Linked data platform 1.0}}. W3C.
-
-\leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-jrZe6Esu}{}}%
-Loo T (ed.). 1970. First International Conference on FAIR Digital Objects. DOI: \href{https://doi.org/10.3897/rio.coll.190}{10.3897/rio.coll.190}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
-martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-PL512B0X}{}}%
-May W. 2017. \emph{HTTP Live Streaming}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8216}{10.17487/rfc8216}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-5V3GqSzR}{}}%
-Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
-2021b.
-Web data apis over SPARQL.
-In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
-Synthesis lectures on data, semantics, and knowledge.
-Cham: Springer International Publishing, 27--38.
-DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_3}{10.1007/978-3-031-01917-3\_3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-99MS5xE0}{}}%
-Meroño-Peñuela A, Lisena P, Martínez-Ortiz C.
-2021a.
-Conclusion and future challenges.
-In: \emph{Web data apis for knowledge graphs: Easing access to semantic data for application developers}.
-Synthesis lectures on data, semantics, and knowledge.
-Cham: Springer International Publishing, 77--80.
-DOI: \href{https://doi.org/10.1007/978-3-031-01917-3_7}{10.1007/978-3-031-01917-3\_7}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tZOPTKnJ}{}}%
-Mika P. 2015. On Schema.org and Why It Matters for the Web. \emph{IEEE Internet Computing} 19:52--55. DOI: \href{https://doi.org/10.1109/mic.2015.81}{10.1109/mic.2015.81}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-u8Cy0psL}{}}%
-Mons B, Neylon C, Velterop J, Dumontier M, Silva Santos LOB da, Wilkinson MD.
-2017.
-Cloudy, increasingly FAIR; revisiting the FAIR data guiding principles for the european open science cloud.
-\emph{Information Services \& Use} 37:49--56.
-DOI: \href{https://doi.org/10.3233/\%7BISU\%7D-170824}{10.3233/\{ISU\}-170824}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
-NCBO BioPortal. \emph{Available at} \href{https://bioportal.bioontology.org/ontologies}{\emph{https://bioportal.bioontology.org/ontologies}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-J2H8yssV}{}}%
-Neumann A, Laranjeiro N, Bernardino J.
-2021.
-An analysis of public REST web service apis.
-\emph{IEEE Transactions on Services Computing} 14:957--970.
-DOI: \href{https://doi.org/10.1109/TSC.2018.2847344}{10.1109/TSC.2018.2847344}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ozdWB3O7}{}}%
-Nottingham M. 2017. \emph{Web Linking}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc8288}{10.17487/rfc8288}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
-OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More.
-\emph{Available at} \href{https://spec.openapis.org/oas/v3.1.0.html}{\emph{https://spec.openapis.org/oas/v3.1.0.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
-ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
-OWL 2 Web Ontology Language Document Overview (Second Edition). \emph{Available at} \href{http://www.w3.org/TR/owl2-overview/}{\emph{http://www.w3.org/TR/owl2-overview/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS6a6tlz}{}}%
-Page KR, De Roure DC, Martinez K.
-2011.
-REST and Linked Data.
-In: \emph{Proceedings of the Second International Workshop on RESTful Design - WS-REST '11}.
-ACM Press,.
-DOI: \href{https://doi.org/10.1145/1967428.1967435}{10.1145/1967428.1967435}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-5SuwLNdE}{}}%
-Polleres A, Kamdar MR, Fernández JD, Tudorache T, Musen MA.
-2020.
-A more decentralized vision for linked data.
-\emph{Satya Widya} 11:101--113.
-DOI: \href{https://doi.org/10.3233/SW-190380}{10.3233/SW-190380}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 January 2023a).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 January 2023b).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
-RDF 1.1 Primer. \emph{Available at} \href{http://www.w3.org/TR/rdf11-primer/}{\emph{http://www.w3.org/TR/rdf11-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
-RDF Schema 1.1. \emph{Available at} \href{http://www.w3.org/TR/rdf-schema/}{\emph{http://www.w3.org/TR/rdf-schema/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
-RDFa 1.1 Primer - Third Edition. \emph{Available at} \href{http://www.w3.org/TR/rdfa-primer/}{\emph{http://www.w3.org/TR/rdfa-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-16wvFH4nh}{}}%
-Resource Description Framework (RDF) Model and Syntax Specification.
-\emph{Available at} \href{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}{\emph{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
-ResourceSync Framework Specification - Table of Contents. \emph{Available at} \href{http://www.openarchives.org/rs/toc}{\emph{http://www.openarchives.org/rs/toc}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-F8VZ86hu}{}}%
-Sandvine.Global Internet Phenomena Report 2022.
-\emph{Available at} \href{https://www.sandvine.com/global-internet-phenomena-report-2022}{\emph{https://www.sandvine.com/global-internet-phenomena-report-2022}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-rnzJ4khD}{}}%
-Sauermann L, Cyganiak R, Völkel M. 2011. \emph{Cool URIs for the semantic web}. Universität des Saarlandes. DOI: \href{https://doi.org/10.22028/d291-25086}{10.22028/d291-25086}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
-Schema.org - Schema.org. \emph{Available at} \href{https://schema.org/}{\emph{https://schema.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
-Schema.org Actions - schema.org. \emph{Available at} \href{https://schema.org/docs/actions.html}{\emph{https://schema.org/docs/actions.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-IHLT6hye}{}}%
-Schultes E, Wittenburg P.
-2019.
-FAIR principles and digital objects: Accelerating convergence on a data infrastructure.
-In: Manolopoulos Y, Stupnikov S eds.
-\emph{Data analytics and management in data intensive domains: 20th international conference, DAMDID/RCDL 2018, moscow, russia, october 9--12, 2018, revised selected papers}.
-Communications in computer and information science.
-Cham: Springer International Publishing, 3--16.
-DOI: \href{https://doi.org/\%7B10.1007/978-3-030-23584-0_1\%7D}{\{10.1007/978-3-030-23584-0\_1\}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1CqIZ47pu}{}}%
-Schwardmann U, Kálmán T.
-2022.
-Two Examples on How FDO Types can Support Machine and Human Readability.
-\emph{Research Ideas and Outcomes} 8.
-DOI: \href{https://doi.org/10.3897/rio.8.e96014}{10.3897/rio.8.e96014}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
-Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
-Shape Expressions (ShEx) 2.1 Primer. \emph{Available at} \href{http://shex.io/shex-primer/}{\emph{http://shex.io/shex-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
-Shapes Constraint Language (SHACL). \emph{Available at} \href{https://www.w3.org/TR/shacl/}{\emph{https://www.w3.org/TR/shacl/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
-SKOS Simple Knowledge Organization System Primer. \emph{Available at} \href{http://www.w3.org/TR/skos-primer/}{\emph{http://www.w3.org/TR/skos-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-CNP39tah}{}}%
-Smith B, Ashburner M, Rosse C, Bard J, Bug W, Ceusters W, Goldberg LJ, Eilbeck K, Ireland A, Mungall CJ, Leontis N, Rocca-Serra P, Ruttenberg A, Sansone S-A, Scheuermann RH, Shah N, Whetzel PL, Lewis S.
-2007.
-The OBO Foundry: coordinated evolution of ontologies to support biomedical data integration.
-\emph{Nature Biotechnology} 25:1251--1255.
-DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
-SPARQL 1.1 Overview. \emph{Available at} \href{http://www.w3.org/TR/sparql11-overview/}{\emph{http://www.w3.org/TR/sparql11-overview/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-8ZRGFa6y}{}}%
-Stallings W. 1990. \emph{Handbook of computer-communications standards: The open systems (OSI) model and OSI-related standards}. Carmel, IN, USA: Sams.
-
-\leavevmode\vadjust pre{\hypertarget{ref-NxvhwSmG}{}}%
-Stanczyk SK. 1987. Process modelling for information system description. \emph{The Open University}. DOI: \href{https://doi.org/10.21954/ou.ro.0000f821}{10.21954/ou.ro.0000f821}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-uxWjmlio}{}}%
-Stefi A.
-2015.
-Do Developers Make Unbiased Decisions?
-- The Effect of Mindfulness and Not-Invented-Here Bias on the Adoption of Software Components.
-DOI: \href{https://doi.org/10.18151/7217489}{10.18151/7217489}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-v0WwFHgT}{}}%
-Stefi A, Hess T.
-2015.
-To develop or to reuse?
-Two perspectives on external reuse in software projects.
-In: Fernandes JM, Machado RJ, Wnuk K eds.
-\emph{Software business}.
-Lecture notes in business information processing.
-Cham: Springer International Publishing, 192--206.
-DOI: \href{https://doi.org/10.1007/978-3-319-19593-3_18}{10.1007/978-3-319-19593-3\_18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-yygVPoL0}{}}%
-Strawn G, Wittenburg P, Soiland-Reyes St, Peters K, Lannom L, Schwardmann U, Blanch Ch.
-2022.
-\emph{\href{https://docs.google.com/document/d/1aGA-TBr4XpORhMPtnf_--Nb4FYJccgeSvGmGh68jNws/edit}{FDO Forum FDO requirement specifications}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-2gXY2Nwh}{}}%
-Sun S, Lannom L, Boesch B. 2003. \emph{Handle System Overview}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3650}{10.17487/rfc3650}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-I9Kx0Hjy}{}}%
-Sun S, Reilly S, Lannom L, Petrone J. 2003. \emph{Handle System Protocol (ver 2.1) Specification}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc3652}{10.17487/rfc3652}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
-SWORD 3.0 Specification. \emph{Available at} \href{https://swordapp.github.io/swordv3/swordv3.html}{\emph{https://swordapp.github.io/swordv3/swordv3.html}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
-The Open Graph protocol. \emph{Available at} \href{https://ogp.me/}{\emph{https://ogp.me/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-S5BI6Mwx}{}}%
-Thornton K, Solbrig H, Stupp GS, Labra Gayo JE, Mietchen D, Prud E, Waagmeester A.
-2019.
-Using shape expressions (ShEx) to share RDF data models and to guide curation with rigorous validation.
-In: Hitzler P, Fernández M, Janowicz K, Zaveri A, Gray AJG, Lopez V, Haller A, Hammar K eds.
-\emph{The semantic web: 16th international conference, ESWC 2019, portorož, slovenia, june 2--6, 2019, proceedings}.
-Lecture notes in computer science.
-Cham: Springer International Publishing, 606--620.
-DOI: \href{https://doi.org/10.1007/978-3-030-21348-0_39}{10.1007/978-3-030-21348-0\_39}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-1DalZuKct}{}}%
-Tirmizi S, Aitken S, Moreira DA, Mungall C, Sequeda J, Shah NH, Miranker DP.
-2011.
-Mapping between the OBO and OWL ontology languages.
-\emph{Journal of Biomedical Semantics} 2:S3.
-DOI: \href{https://doi.org/10.1186/2041-1480-2-s1-s3}{10.1186/2041-1480-2-s1-s3}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNTCUjpC}{}}%
-Turcoane O.
-2014.
-Linked data, JSON-LD and the semantics of cultural and scientific heritage.
-\emph{Digital Presentation and Preservation of Cultural and Scientific Heritage} 4:95--105.
-DOI: \href{https://doi.org/10.55630/dipp.2014.4.11}{10.55630/dipp.2014.4.11}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
-Usage Statistics of JSON-LD for Websites, May 2022.
-\emph{Available at} \href{https://w3techs.com/technologies/details/da-jsonld}{\emph{https://w3techs.com/technologies/details/da-jsonld}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-snykkm7R}{}}%
-Van de Sompel H, Klein M, Jones S, Nelson ML, Warner S, Devaraju A, Huber R, Steinhoff W, Tykhonov V, Boruta L, Meijers E, Soiland-Reyes S, Wilkinson M.
-2022.FAIR Signposting Profile.
-\emph{Available at} \href{https://signposting.org/FAIR/}{\emph{https://signposting.org/FAIR/}} (accessed 5 January 2023).
-
-\leavevmode\vadjust pre{\hypertarget{ref-DNjcMi4a}{}}%
-Van de Sompel H, Nelson M, Sanderson R. 2013. \emph{HTTP Framework for Time-Based Access to Resource States -\/- Memento}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc7089}{10.17487/rfc7089}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jHwttlTs}{}}%
-Verborgh R, Vander Sande M.
-2020.
-The semantic web identity crisis: In search of the trivialities that never were.
-\emph{Satya Widya} 11:19--27.
-DOI: \href{https://doi.org/10.3233/SW-190372}{10.3233/SW-190372}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
-W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures. \emph{Available at} \href{http://www.w3.org/TR/xmlschema11-1/}{\emph{http://www.w3.org/TR/xmlschema11-1/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
-Web Services Description Language (WSDL) Version 2.0 Part 0: Primer. \emph{Available at} \href{http://www.w3.org/TR/wsdl20-primer/}{\emph{http://www.w3.org/TR/wsdl20-primer/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
-WebSockets Standard. \emph{Available at} \href{https://websockets.spec.whatwg.org/}{\emph{https://websockets.spec.whatwg.org/}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-lCoshHXg}{}}%
-Weigel T, Plale B, Parsons M, Zhou G, Luo Y, Schwardmann U, Quick R, Hellström M, Kurakawa K.
-2018.
-RDA Recommendation on PID Kernel Information.
-\emph{Research Data Alliance}.
-DOI: \href{https://doi.org/10.15497/rda00031}{10.15497/rda00031}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-iJeys0T5}{}}%
-Weiland C, Islam S, Broder D, Anders I, Wittenburg P.
-2022a.
-\emph{\href{https://docs.google.com/document/d/1GHFPAUGpNvYaxctkx-CpvY1vKf_aGZpSlWGOWvyXSiQ/edit}{FDO machine actionability}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-TQdku4YF}{}}%
-Weiland C, Schwardmann U, Wittenburg P, Kirkpatrick C, Hanisch R, Trautt Z.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1lPNBBROjEoZ6fTfrtdqcMa3Q2G27PoC_/edit}{FDO Forum Document Standards}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-9jaVzCbL}{}}%
-Wieczorek J, Bloom D, Guralnick R, Blum S, Döring M, Giovanni R, Robertson T, Vieglais D.
-2012.
-Darwin Core: An Evolving Community-Developed Biodiversity Data Standard.
-\emph{PLoS ONE} 7:e29715.
-DOI: \href{https://doi.org/10.1371/journal.pone.0029715}{10.1371/journal.pone.0029715}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nCzRO0pK}{}}%
-Wilde E. 2013. \emph{The 'profile' Link Relation Type}. RFC Editor. DOI: \href{https://doi.org/10.17487/rfc6906}{10.17487/rfc6906}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-6DjakjNS}{}}%
-Wilkinson MD, Dumontier M, Aalbersberg IjJ, Appleton G, Axton M, Baak A, Blomberg N, Boiten J-W, da Silva Santos LB, Bourne PE, Bouwman J, Brookes AJ, Clark T, Crosas M, Dillo I, Dumon O, Edmunds S, Evelo CT, Finkers R, Gonzalez-Beltran A, Gray AJG, Groth P, Goble C, Grethe JS, Heringa J, 't Hoen PAC, Hooft R, Kuhn T, Kok R, Kok J, Lusher SJ, Martone ME, Mons A, Packer AL, Persson B, Rocca-Serra P, Roos M, van Schaik R, Sansone S-A, Schultes E, Sengstag T, Slater T, Strawn G, Swertz MA, Thompson M, van der Lei J, van Mulligen E, Velterop J, Waagmeester A, Wittenburg P, Wolstencroft K, Zhao J, Mons B.
-2016.
-The FAIR Guiding Principles for scientific data management and stewardship.
-\emph{Scientific Data} 3.
-DOI: \href{https://doi.org/10.1038/sdata.2016.18}{10.1038/sdata.2016.18}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Pa67pUtR}{}}%
-Wilkinson SR, Eisenhauer G, Kapadia AJ, Knight K, Logan J, Widener P, Wolf M.
-2022.
-F*** workflows: when parts of FAIR are missing.
-\emph{arXiv}.
-DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Qlrlr2TZ}{}}%
-Williams AJ, Harland L, Groth P, Pettifer S, Chichester C, Willighagen EL, Evelo CT, Blomberg N, Ecker G, Goble C, Mons B.
-2012.
-Open PHACTS: semantic interoperability for drug discovery.
-\emph{Drug Discovery Today} 17:1188--1198.
-DOI: \href{https://doi.org/10.1016/j.drudis.2012.05.016}{10.1016/j.drudis.2012.05.016}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jRWurmQc}{}}%
-Wittenburg P, Anders I, Blanchi C, Buurman M, Goble C, Grieb J, Hardisty A, Islam S, Jejkal T, Kálmán T, Kirkpatrick C, Lannom L, Lauer T, Manepalli G, Peters-von Gehlen K, Pfeil A, Quick R, Sanden M van de, Schwardmann U, Soiland-Reyes S, Stotzka R, Trautt Z, Van Uytvanck D, Weiland C, Wieder P.
-2022.
-FAIR digital object demonstrators 2021.
-\emph{Zenodo}.
-DOI: \href{https://doi.org/10.5281/zenodo.5872645}{10.5281/zenodo.5872645}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tz0P3DTC}{}}%
-Wittenburg P, Strawn G, Mons B, Bonino L, Schultes E.
-2019.
-Digital objects as drivers towards convergence in data infrastructures.
-\emph{https://b2share.eudat.eu}.
-DOI: \href{https://doi.org/10.23728/b2share.b605d85809ca45679b110719b6c6cb11}{10.23728/b2share.b605d85809ca45679b110719b6c6cb11}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-39K3yHT3}{}}%
-Wolstencroft K, Haines R, Fellows D, Williams A, Withers D, Owen S, Soiland-Reyes S, Dunlop I, Nenadic A, Fisher P, Bhagat J, Belhajjame K, Bacall F, Hardisty A, Nieva de la Hidalga A, Balcazar Vargas MP, Sufi S, Goble C.
-2013.
-The Taverna workflow suite: designing and executing workflows of Web Services on the desktop, web or in the cloud.
-\emph{Nucleic Acids Research} 41:W557--W561.
-DOI: \href{https://doi.org/10.1093/nar/gkt328}{10.1093/nar/gkt328}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-MhsCt4P3}{}}%
-Wolstencroft K, Owen S, Horridge M, Krebs O, Mueller W, Snoep JL, du Preez F, Goble C.
-2011.
-RightField: embedding ontology annotation in spreadsheets.
-\emph{Bioinformatics} 27:2021--2022.
-DOI: \href{https://doi.org/10.1093/bioinformatics/btr312}{10.1093/bioinformatics/btr312}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-15yGpJ0wh}{}}%
-Working Group FDO-TSIG. 2022a. \emph{\href{https://docs.google.com/document/d/1me0L8C5yDe39cYP1Sxud4Y10hxhphOimLB-K-KgHQUk/edit}{DOIP endorsement request}}. (internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-16W9AycSW}{}}%
-Working Group FDO-TSIG.
-2022b.
-\emph{\href{https://docs.google.com/document/d/1VJTYHxLXIbjHSoFnMEkKaTbpTz9CagoIJyoSafkVh_I/edit}{FDO -- granularity, versioning, mutability}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Fi62cJAo}{}}%
-Working Group FDO-TSIG, Working Group FDO-BIG.
-2022.
-\emph{\href{https://docs.google.com/document/d/11CuSoNOpg3vYaoHqi4Yg6ncP3qOclJIC/edit}{FDO -- kernel attributes \& metadata}}.
-(internal draft): FDO Forum.
-
-\leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
-X.1255 : Framework for discovery of identity management information.
-\emph{Available at} \href{https://www.itu.int/rec/T-REC-X.1255-201309-I}{\emph{https://www.itu.int/rec/T-REC-X.1255-201309-I}} (accessed 26 May 2022).
-
-\leavevmode\vadjust pre{\hypertarget{ref-hRzcHhPD}{}}%
-Zarras A. 2004. A Comparison Framework for Middleware Infrastructures. \emph{The Journal of Object Technology} 3:103. DOI: \href{https://doi.org/10.5381/jot.2004.3.5.a2}{10.5381/jot.2004.3.5.a2}.
-
-\end{CSLReferences}
+\end{document}

### 2023-01-25 latex/table1.tex
Fixing references to use bibliography

@@
   >{\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}).
+\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data \cite{KlCFFFyL}.
 \label{tbl:fdo-web-interoperability-framework}}\tabularnewline
@@
 & The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}) end up being mandated centrally by dominant (often commercial) providers (\protect\hyperlink{ref-11VxBOeZy}{Fielding et al., 2017}), which clients are required to use as-is with special code per service.
+In practice, REST APIs \cite{174AwcFUL} end up being mandated centrally by dominant (often commercial) providers \cite{11VxBOeZy}, which clients are required to use as-is with special code per service.
 Use of Linked Data enables common tooling and semantic mapping across differences.
@@
 & ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), schema.org Actions (\protect\hyperlink{ref-dKAekUjL}{{`Schema.org Actions - schema.org'}}), WSDL/SOAP (\protect\hyperlink{ref-LmudecyN}{{`Web Services Description Language (WSDL) Version 2.0 Part 0: Primer'}}), but frequently just as human-readable developer documentation/examples.
+Operational composition, typically within a single API provider, documented by OpenAPI 3 \cite{k0AfCGzw}, schema.org Actions \cite{dKAekUjL}, WSDL/SOAP \cite{LmudecyN}, but frequently just as human-readable developer documentation/examples.
 \\
@@
 No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) (\protect\hyperlink{ref-jRWurmQc}{Wittenburg et al., 2022}) & Lightweight HTTP semantics for authenticity/navigation.
+JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) \cite{jRWurmQc} & Lightweight HTTP semantics for authenticity/navigation.
 Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
@@
 \\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}) is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
+\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP \cite{13TcbsZF6} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
 \textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.

### 2023-01-25 latex/table3.tex
Fixing references to use bibliography

@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018})) and Web technologies (using Linked Data) as middleware infrastructures (\protect\hyperlink{ref-hRzcHhPD}{Zarras, 2004})
-\label{tbl:fdo-web-middleware}}\tabularnewline
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
+\caption{Checking FDO guidelines \cite{RwvirqWg, yygVPoL0} against its current implementations as DOIP \cite{13TcbsZF6} and Linked Data Platform (LDP) \cite{7szz7dwO}, with suggestions for required additions.
+\label{tbl:fdo-checks}}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
+\begin{minipage}[b]{\linewidth}\centering
+\textbf{FDO Guideline}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
+DOIP 2.0
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
+FDO suggestions
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Linked Data Platform
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+LDP suggestion
 \end{minipage} \\
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
+\begin{minipage}[b]{\linewidth}\centering
+\textbf{FDO Guideline}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
+DOIP 2.0
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
+FDO suggestions
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Linked Data Platform
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+LDP suggestion
 \end{minipage} \\
@@
 \endhead
-\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
-Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
-PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}.
-& The Web is inherently open and made by cross-linked URLs.
-Participation requires DNS domain purchase (many free alternatives also exists).
-PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects.
+G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
+& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
+\texttt{http://} URIs replaced by \texttt{https://}.
+& Keep flexibility of RDF serialisation formats which may change more frequently \\
+G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
+DOIP relatively unknown, in effect only one implementation.
+& Further promote DOIP and justify its benefits.
+Build tutorials and OSI open source implementations.
+Standardise DOIP-over-HTTP alternative.
+& JSON-LD used by half of all websites \cite{j3SqQLBR}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
+Example-driven documentation.
 \\
-\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
-HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
+G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
+& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
 \\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
-DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
-Multiple connections possible.
-Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks (\protect\hyperlink{ref-F8VZ86hu}{Sandvine}).
-HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}) but still has TCP+TLS startup costs.
-For reduced latency , HTTP/3 (\protect\hyperlink{ref-Amit86Tp}{{`draft-ietf-quic-http-34'}}) use QUIC (\protect\hyperlink{ref-1GWvyEmUf}{Iyengar \& Thomson, 2021})) rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
+G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
+Specify input/output expectations formally (e.g.~JSON Schema).
+& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
 \\
-\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
-FDOs should provide the mandatory kernel metadata fields.
-FDOs of the same declared type typically share additional attributes (although that schema may not be declared).
-DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions.
-& Each URL accessed separately.
-Common HTTP headers provide basic metadata, although it is often not reliable.
-A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), content-negotiation (\protect\hyperlink{ref-n2f3TfoI}{{`Content negotiation - HTTP \textbar{} MDN'}}) or FAIR Signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}).
+G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
+DOIP operations can use any transport protocol.
+& Document transport protocols as FDOs, recommend which transport to use.
+& URI as abstraction base.
+Does not specify PID requirements.
+& Give stronger deployment recommendations.
 \\
-\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
-Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
-& Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
-\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
-Multiple locations possible in Handle system, can expose geo-info.
-& PIDs and URL redirects.
-DNS aliases and IP routing can hide location.
-Geo-localised servers common for large cloud deployments.
+G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
+Unclear when metadata field is a PID or plain text.
+& Make datatype of fields explicit to support navigation.
+& Machine-navigation through URIs via properties and types.
+Unclear when URI should be followed or is just identifier, but always distinct from plain text.
+& \\
+G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
+& Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
+& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
+Unsupported methods reported using LDP constraints to human-readable text.
+& Declare supported methods in advance, e.g.~OpenAPI \cite{k0AfCGzw} \\
+G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
+No FDOF implementations.
+& Develop demonstrator of FDOF usage based on existing LDP server.
 \\
-\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
-FDO kernel metadata can include checksum and date.
-& HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
-Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
+G9: \emph{standard compliance} & Handle \cite{2gXY2Nwh}, DOIP \cite{13TcbsZF6}.
+FDO requirements not standardised yet.
+& Formalise standard process of FDO requirements \cite{TQdku4YF} & HTTP, LDP.
+FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
+FDOF1: \emph{PID as basis} & Extensive use of Handle system.
+& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve \cite{ljVViWCl}.
+Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
+& HTTP URLs as basis for identifiers, but they are frequently not persistent.
+& Add strong guidance for PID services like w3id and persistence policies.
 \\
-\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
-Custom codes uncommon.
+FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
+Datacite DOIs content-negotiate to give registered metadata.
+& Add FAIR Signposting at PID provider for minimal PID record \\
+FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
+No method/attribute defined for separate metadata, only directly in PID Record.
+Unclear meaning of multiple items and bytestream chunks.
+& Clarify expectations for multiple items.
+Recommend chunks to not be used.
+& URIs resolvable by default.
+Multiple ways to resolve metadata, unclear preference.
+& Add FAIR Signposting and preference order.
 \\
-\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
-No redirection from DOIP service.
-& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
+FDOF4: \emph{Additional attributes} & Freetext attribute keys.
+Attributes should be defined for FDO type (?).
+& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
+Explicitly allow attributes not already defined in type.
+& All attributes individually identified.
+Any Linked Data attributes can be used by URI or with mapped prefix.
+& Clarify type expectations of required/recommended/optional attributes.
 \\
-\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
-There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged (\protect\hyperlink{ref-rbG9uRKw}{{`Hypertext Style: Cool URIs don't change.'}}).
-Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
-An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento (\protect\hyperlink{ref-DNjcMi4a}{Van de Sompel, Nelson \& Sanderson, 2013}) expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method (\protect\hyperlink{ref-hRUsaTiz}{Clemm et al., 2002}) (used by SVN).
+FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
+& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
+Clarify that operations can be mapped to protocol directly.
+& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
+Unclear how to provide PID for additional operations.
+& Specify how additional operations should be called over HTTP.
 \\
-\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
-Internal synchronisation possible in backend for Extended operations.
-& Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} (\protect\hyperlink{ref-xvMAyRAc}{Dusseault, 2007}) with \texttt{LOCK} and \texttt{UNLOCK} methods.
+FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
+& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns \cite{i9Ang0rM} are used instead.
+& Document how operation resources can be discovered from an LPD container.
+Document search API.
 \\
-\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
-In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
-& The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space.
-In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend.
+FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
+PR-TypingFDOs leaves this open.
+& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
+Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
+& Always make LDP types explicit in FDO profile.
 \\
-\textbf{Encapsulation}: \emph{separate interface from implementation.
-Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
-FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
-Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade (\protect\hyperlink{ref-1B16WfR7q}{Khare \& Lawrence, 2000}) to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns (\protect\hyperlink{ref-iME5l1kk}{{`Hypertext Style: Cool URIs don't change.'}}) and PIDs via Persistent URLs (PURL).
+FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
+Separate Metadata FDO need custom property.
+& Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
+& Assertions are always with semantics, using RDF vocabularies.
+Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
+& Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
+FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
+Metadata would need to be bundled within custom FDO types/attributes.
+Unclear how levels are separated within single FDO representation (need FDOF8?).
+& Declare which metadata are expected within response attribute or within FDO object.
+Require PIDs for custom attributes.
+Define how alternate metadata levels can be represented separately.
+& Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
+& Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
+FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
+& Require types to be FDOs with registered PIDs, implement shared types.
+& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} \cite{CNP39tah} & Rather document better how individual ad-hoc schemas can be started for prototypes.
 \\
-\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
-FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})) & Syntactically Media Type with multiple suffixes (\protect\hyperlink{ref-nlH2rm1s}{{`draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes'}}) (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) (\protect\hyperlink{ref-nCzRO0pK}{Wilde, 2013}).
-In metadata, semantic type systems (RDFS (\protect\hyperlink{ref-ZwAcGQKY}{{`RDF Schema 1.1'}}), OWL2 (\protect\hyperlink{ref-1p4IWJpI}{{`OWL 2 Web Ontology Language Document Overview (Second Edition)'}}), SKOS (\protect\hyperlink{ref-15gQDya5B}{{`SKOS Simple Knowledge Organization System Primer'}})).
-OpenAPI 3 (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}) \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} (\protect\hyperlink{ref-11oVTMsuZ}{{`W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures'}}), JSON \texttt{\$schema} (\protect\hyperlink{ref-15EZ2D0Rm}{{`draft-bhutton-json-schema-00'}})), JSON-LD \texttt{@context} (\protect\hyperlink{ref-by6CfOmv}{{`JSON-LD 1.1'}}).
-Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
-\\
-\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
-\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 (\protect\hyperlink{ref-tIkSXrb5}{Fielding \& Reschke, 2014b}) \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) (\protect\hyperlink{ref-PL512B0X}{May, 2017}), MPEG-DASH (ISO/IEC 23009-1:2019) (\protect\hyperlink{ref-ikO7e0mh}{14:00-17:00}) \\
+FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
+Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
+\cite{6LvydTsC}).
+& & LDP Containers required by specification, also user-created (eg.
+\texttt{BasicContainer}).
+& Clarify relation to other collections like DCAT 3 \cite{1EW4hLN8U}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE \cite{gtGuSevK} \\
+FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
+\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
+No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
 \bottomrule

### 2023-01-25 latex/table4.tex
Fixing references to use bibliography

@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model (\protect\hyperlink{ref-1GrKTFaK2}{Group, 2020}; \protect\hyperlink{ref-UzQhqk0M}{Bahim et al., 2020}) (first 2 columns) against the FDO guidelines (\protect\hyperlink{ref-RwvirqWg}{Bonino et al., 2019}), FDO implemented with the protocol DOIPv2 (\protect\hyperlink{ref-13TcbsZF6}{Foundation, 2018}), Linked Data Platform (LDP) (\protect\hyperlink{ref-7szz7dwO}{{`FAIR Digital Object Framework Documentation'}}) and examples from Linked Data practices in general.
-(--- indicates \emph{Unspecified}, may be possible with additional conventions)
-\label{tbl:fair-data-maturity-model}}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
+  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
+\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \cite{13TcbsZF6}) and Web technologies (using Linked Data) as middleware infrastructures \cite{hRzcHhPD}
+\label{tbl:fdo-web-middleware}}\tabularnewline
 \toprule
 \begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
+\emph{Quality}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
+FDO w/ DOIP
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
+Web w/ Linked Data
 \end{minipage} \\
@@
 \begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
+\emph{Quality}
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
+FDO w/ DOIP
 \end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
+Web w/ Linked Data
 \end{minipage} \\
@@
 \endhead
-RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
-RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
-Handle, DOI.
-& FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
-RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
-FAIR Signposting.
+\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
+Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
+PID minting requires Handle.net prefix subscription, or use of services like \href{https://datacite.org/}{Datacite}, \href{https://eudat.eu/services/userdoc/b2handle}{B2Handle}.
+& The Web is inherently open and made by cross-linked URLs.
+Participation requires DNS domain purchase (many free alternatives also exists).
+PID minting can be free using PURL/ARK services, or can use DOI/Handle with HTTP redirects.
 \\
-RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
-RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
-Unclear how to link to additional metadata.
-& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
-Embedded in landing page (RDFa).
+\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
+HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
 \\
-RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
-& PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
-RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
-Handle registry only searchable by PID.
-& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-(\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}, pp.
-https://lod--cloud.net/})).
-Indexing by search engines if exposing HTML w/schema.org.
+\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
+DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
+Multiple connections possible.
+Setup is typically through TCP and TLS which adds latency.
+& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks \cite{F8VZ86hu}.
+HTTP 1 connections are serial and reusable, and concurrent connections is common.
+HTTP/2 adds asynchronous responses and multiplexed streams \cite{KbbW0kGT} but still has TCP+TLS startup costs.
+For reduced latency, HTTP/3 \cite{Amit86Tp} use QUIC \cite{1GWvyEmUf} rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
 \\
-RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
-\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
-RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
-RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
-RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
-RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
-RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
-Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP (\protect\hyperlink{ref-8HJqcF1Q}{Fielding, Nottingham \& Reschke, 2022}) & HTTP/HTTPS, FTP (now less common), GridFTP (\protect\hyperlink{ref-18LsXvKGJ}{Allcock et al.}) (for large data), ARK (\protect\hyperlink{ref-voDQ6diM}{{`The ARK Identifier Scheme'}}) \\
-RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol (\protect\hyperlink{ref-I9Kx0Hjy}{Sun et al., 2003}).
-One server implementation (\protect\hyperlink{ref-1CvYRVhy2}{{`Handle.Net Registry'}}), free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved (\protect\hyperlink{ref-xVJMmGVr}{{`Licenses \& Standards \textbar{} Open Source Initiative'}}) as an open source license -- it includes usage restrictions and requires Service Agreements.
-It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
-One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
-\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
-& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
+\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
+FDOs should provide the mandatory kernel metadata fields.
+FDOs of the same declared type typically share additional attributes (although that schema may not be declared).
+DOIP does not enforce metadata typing constraints, this need to be established as FDO conventions.
+& Each URL accessed separately.
+Common HTTP headers provide basic metadata, although it is often not reliable.
+A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
+Metadata is not always machine findable, may need pre-agreed API URI Templates \cite{11hxwwuRt}, content-negotiation \cite{n2f3TfoI} or FAIR Signposting \cite{snykkm7R}.
 \\
-RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
-RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
-RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
-RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
-RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
-RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
-RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
-RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
-& FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) (\protect\hyperlink{ref-1BlxdxOwa}{14:00-17:00}) \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs (\protect\hyperlink{ref-o4h92LpS}{{`Data - W3C'}})), \texttt{rdfs:seeAlso}, FAIR signposting (\protect\hyperlink{ref-snykkm7R}{Van de Sompel et al., 2022}) \texttt{describedby} \\
-RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
-RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
-RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles (\protect\hyperlink{ref-QpDmajc4}{{`Linking Across Provenance Bundles'}}), \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role (\protect\hyperlink{ref-10HA0mMBx}{Unknown}), \texttt{prov:specializationOf} (\protect\hyperlink{ref-ZUgwvAHE}{{`PROV-O: The PROV Ontology'}}), OAI-ORE Proxy (\protect\hyperlink{ref-ID1pTGry}{{`ORE Specification - Abstract Data Model'}})) \\
-RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) but not assigned PIDs yet.
-& Unspecified.
-Multiple metadata records can allow multiple semantic profiles.
-& Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org (\protect\hyperlink{ref-10pzKCEUE}{{`Schema.org - Schema.org'}}), Dublin Core Terms (\protect\hyperlink{ref-wAbZpOsr}{{`DCMI Metadata Terms'}}), DCAT (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), FOAF (\protect\hyperlink{ref-XmUY0cZB}{{`FOAF Vocabulary Specification'}}) \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} (\protect\hyperlink{ref-lnH0hEuh}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}) \\
-RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
-RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
-RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
-PID Kernel attributes?
-Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O (\protect\hyperlink{ref-9T4j3N4e}{{`PROV-O: The PROV Ontology'}}), PAV (\protect\hyperlink{ref-1E4OfHyhI}{Ciccarese et al., 2013}), Dublin Core Terms (\protect\hyperlink{ref-hKblceo6}{{`DCMI Metadata Terms'}, 2020}) \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen (\protect\hyperlink{ref-zZTtJruI}{Hardisty et al., 2022})) & --- & Common, e.g.~DCAT 2 (\protect\hyperlink{ref-zt8AAjcZ}{{`Data Catalog Vocabulary (DCAT) - Version 2'}}), BioSchemas (\protect\hyperlink{ref-1F8S9FeHf}{{`Bioschemas - Bioschemas'}}) \\
-RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal (\protect\hyperlink{ref-1Gsq23e51}{{`NCBO BioPortal'}}), Darwin Core (\protect\hyperlink{ref-9jaVzCbL}{Wieczorek et al., 2012}) \\
-RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
+\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
+Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
+& Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
+\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
+Multiple locations possible in Handle system, can expose geo-info.
+& PIDs and URL redirects.
+DNS aliases and IP routing can hide location.
+Geo-localised servers common for large cloud deployments.
+\\
+\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
+FDO kernel metadata can include checksum and date.
+& HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
+Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
+\\
+\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
+Custom codes uncommon.
+\\
+\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
+No redirection from DOIP service.
+& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
+\\
+\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
+There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
+& URLs are not required to persist, although encouraged \cite{rbG9uRKw}.
+Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
+An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
+Memento \cite{DNjcMi4a} expose versioned snapshots.
+WebDAV \texttt{VERSION-CONTROL} method \cite{hRUsaTiz} (used by SVN).
+\\
+\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
+Internal synchronisation possible in backend for Extended operations.
+& Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
+WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} \cite{xvMAyRAc} with \texttt{LOCK} and \texttt{UNLOCK} methods.
+\\
+\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
+In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
+& The Web is inheritently modular in that distributed objects are cross-referenced within a global URI space.
+In practice, an API's set of resources will be exposed through a single HTTP service, but modularity enables fine-grained scalability in backend.
+\\
+\textbf{Encapsulation}: \emph{separate interface from implementation.
+Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
+FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
+Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
+\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade \cite{1B16WfR7q} to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
+Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns \cite{iME5l1kk}) and PIDs via Persistent URLs (PURL).
+\\
+\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
+FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
+Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} \cite{15EZ2D0Rm}) & Syntactically Media Type with multiple suffixes \cite{nlH2rm1s} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) \cite{nCzRO0pK}.
+In metadata, semantic type systems (RDFS \cite{ZwAcGQKY}), OWL2 \cite{1p4IWJpI}, SKOS \cite{15gQDya5B}).
+OpenAPI 3 \cite{k0AfCGzw} \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
+XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} \cite{11oVTMsuZ}, JSON \texttt{\$schema} \cite{15EZ2D0Rm}), JSON-LD \texttt{@context} \cite{by6CfOmv}.
+Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
+\\
+\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
+Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} \cite{KbbW0kGT}, Web Sockets \cite{qgqADk45}, Linked Data Notifications \cite{zlGYiuWC}, AtomPub \cite{IbaebabD}, SWORD \cite{7DGllKzG}, Micropub, more typically ad-hoc jobs/futures REST resources \\
+\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \cite{HE7Ikwwl}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \cite{11hxwwuRt}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \cite{k0AfCGzw}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \cite{174AwcFUL}, an important element of the REST architectural style.} incl.~Hydra \cite{13PaKBrZe}, schema.org Actions \cite{dKAekUjL}, JSON HAL \cite{nyrPYHoP}) \& Link headers (RFC8288) \cite{ozdWB3O7} \\
+\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
+DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 \cite{tIkSXrb5} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) \cite{PL512B0X}, MPEG-DASH (ISO/IEC 23009-1:2019) \cite{ikO7e0mh} \\
 \bottomrule

### 2023-01-25 latex/table5.tex
Fixing references to use bibliography

@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.0833}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
+  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
+  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
+  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
+\caption{Assessing RDA's FAIR Data Maturity Model \cite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \cite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \cite{13TcbsZF6}, Linked Data Platform (LDP) \cite{7szz7dwO} and examples from Linked Data practices in general.
+(--- indicates \emph{Unspecified}, may be possible with additional conventions)
+\label{tbl:fair-data-maturity-model}}\tabularnewline
 \toprule
 \begin{minipage}[b]{\linewidth}\raggedright
-Layer
+FAIR ID
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Recommendation
+Indicator
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+FDO guidelines
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+FDO/DOIP
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+FDO/LDP
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Linked Data examples
+\end{minipage} \\
+\midrule
+\endfirsthead
+\toprule
+\begin{minipage}[b]{\linewidth}\raggedright
+FAIR ID
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-FDO
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Linked Data
+Indicator
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+FDO guidelines
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+FDO/DOIP
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+FDO/LDP
+\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+Linked Data examples
 \end{minipage} \\
@@
 \endhead
-Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
-Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
-Technical & Easy SLAs for service providers & Unspecified & None \\
-Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
-Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
-Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
-Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
-Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
-Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
-Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
-Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
-Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
-Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
-Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
-Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
-Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
-Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
-Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
-Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
-Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
-Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
-Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
-Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
-Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
-Legal & List recommended licenses & Undefined & Best practice recommendations \\
-Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
-Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
-Legal & GDPR compliance for personal data & Undefined & Undefined \\
-Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
-Legal & Harmonised terms-of-use & Undefined & Undefined \\
-Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
+RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
+RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
+Handle, DOI.
+& FDOF-IR (Identifier Record).
+PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
+RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
+FAIR Signposting.
+\\
+RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
+RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
+Unclear how to link to additional metadata.
+& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
+Embedded in landing page (RDFa).
+\\
+RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
+& PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
+RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
+Handle registry only searchable by PID.
+& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
+\cite{1Gsq23e51}).
+Indexing by search engines if exposing HTML w/schema.org.
+\\
+RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
+\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
+RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
+RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
+RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
+RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
+RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
+Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
+RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP \cite{8HJqcF1Q} & HTTP/HTTPS, FTP (now less common), GridFTP \cite{18LsXvKGJ} (for large data), ARK \cite{voDQ6diM} \\
+RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
+RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol \cite{I9Kx0Hjy}.
+One server implementation \cite{1CvYRVhy2}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved \cite{xVJMmGVr}  as an open source license -- it includes usage restrictions and requires Service Agreements.
+It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
+One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
+& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
+\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
+& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
+\\
+RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
+RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
+RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
+RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
+RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
+RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
+RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
+RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
+& FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
+RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \cite{1BlxdxOwa} \\
+RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs \cite{o4h92LpS}), \texttt{rdfs:seeAlso}, FAIR signposting \cite{snykkm7R} \texttt{describedby} \\
+RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
+RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
+RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
+RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles \cite{QpDmajc4}, \href{https://schema.org/Role}{schema.org/Role} \\
+RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role \cite{10HA0mMBx}, \texttt{prov:specializationOf} \cite{ZUgwvAHE}, OAI-ORE Proxy \cite{ID1pTGry}) \\
+RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
+Kernel metadata attributes desired \cite{Fi62cJAo} but not assigned PIDs yet.
+& Unspecified.
+Multiple metadata records can allow multiple semantic profiles.
+& Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
+Rough consensus on kernel metadata: schema.org \cite{10pzKCEUE}, Dublin Core Terms \cite{wAbZpOsr}, DCAT \cite{lnH0hEuh}, FOAF \cite{XmUY0cZB} \\
+RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata \cite{Fi62cJAo} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \cite{lnH0hEuh} \\
+RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
+RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
+RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
+PID Kernel attributes?
+Unspecified W3C PROV-O, PAV & & \\
+RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O \cite{9T4j3N4e}, PAV \cite{1E4OfHyhI}, Dublin Core Terms \cite{hKblceo6} \\
+RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen \cite{zZTtJruI}) & --- & Common, e.g.~DCAT 2 \cite{zt8AAjcZ}, BioSchemas \cite{1F8S9FeHf} \\
+RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
+RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal \cite{1Gsq23e51}, Darwin Core \cite{9jaVzCbL} \\
+RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
 \bottomrule

### 2023-01-25 latex/main.tex
more regex!

@@
 
-The FAIR Digital Object Forum (\cite{DMEBhHID}'}}) working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
+The FAIR Digital Object Forum \cite{DMEBhHID} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
 As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
@@
 \textbf{Upload of FDO} \cite{1H7cmmr69} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync (\cite{eS5y9TRh}}) is suggested as one type of service to list FDOs.
+ResourceSync \cite{eS5y9TRh} is suggested as one type of service to list FDOs.
 This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
@@
 This builds on the \emph{Digital Object} (DO) concept \cite{11MnuwJ4l}, first introduced in 1995 \cite{3Uqe3fuK} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework (\cite{103Hw8H43}}) which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 (\cite{16uB3jxpa} DONA Foundation'}}).
+DO was the inspiration for the ITU X.1255 framework \cite{103Hw8H43} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 \cite{16uB3jxpa}.
 
@@
 \item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification \cite{13TcbsZF6}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0 \cite{13TcbsZF6}, see (\cite{xuQKRRx5}}).} (usually TCP/IP over TLS).
+  Building on the Digital Object concept, using the simplified DOIP v2 specification \cite{13TcbsZF6}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0, see \cite{xuQKRRx5}} (usually TCP/IP over TLS).
 The main DOIP operations are retrieving, creating and updating digital objects.
 These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema (\cite{15EZ2D0Rm}}) and PIDs are assigned using the Handle system.
+FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \cite{15EZ2D0Rm} and PIDs are assigned using the Handle system.
 Several type registries have been established.
@@
 
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) (\cite{16WGmCih8}}) that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format (\cite{16wvFH4nh}}) being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
+Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \cite{16WGmCih8} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \cite{16wvFH4nh} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
 
@@
   such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years (\cite{nW4HY8Nq}}), for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \cite{HqZ3r6J3}.
+The distinction between locator-full and locator-less identifiers have weakened in recent years \cite{nW4HY8Nq}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \cite{HqZ3r6J3}.
 RDF 1.1 formats use Unicode to support IRIs \cite{rqjuXkEf}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
@@
 
-Established Semantic Web services and standards include SPARQL (\cite{xww8fb1B}}) (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
+Established Semantic Web services and standards include SPARQL \cite{xww8fb1B} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
 
@@
 Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS (\cite{15gQDya5B}}) (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
+Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \cite{15gQDya5B} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
 Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \cite{FDDmgO6s} .
@@
 Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}.
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS (\cite{ZwAcGQKY}}), OWL (\cite{1p4IWJpI}}), OBO \cite{1DalZuKct}, SKOS (\cite{15gQDya5B}})), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.
+Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \cite{ZwAcGQKY}, OWL \cite{1p4IWJpI}, OBO \cite{1DalZuKct}, SKOS \cite{15gQDya5B}), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.
 
@@
 
-The \textbf{Linked Data} concept \cite{19s3yyxDn} was kickstarted as a set of best practices (\cite{OGXYtGZ8}}) to bring the Web aspect back into focus.
+The \textbf{Linked Data} concept \cite{19s3yyxDn} was kickstarted as a set of best practices \cite{OGXYtGZ8} to bring the Web aspect back into focus.
 Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
@@
 Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph (\cite{oN9D5LkD}}), RDFa (\cite{FKIF6ApB}}), Microdata (\cite{14Utscfe4}})) we find JSON-LD (\cite{X5SWPqSU}}) as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites (\cite{j3SqQLBR}}).
+In addition to several such embedded \emph{microformats} (Open Graph \cite{oN9D5LkD}, RDFa \cite{FKIF6ApB}, Microdata \cite{14Utscfe4}) we find JSON-LD \cite{X5SWPqSU} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
+JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites \cite{j3SqQLBR}.
 
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\cite{BFMoe8FC}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL (\cite{GDrm9dgX}}) or ShEx (\cite{YUzogaV}})) can be used to validate RDF Data (\cite{ZeOAPNHa}; \cite{S5BI6Mwx}) before consuming it programmatically, or reshaping data to fit other models.
+Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\cite{BFMoe8FC}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \cite{GDrm9dgX} or ShEx \cite{YUzogaV}) can be used to validate RDF Data (\cite{ZeOAPNHa}; \cite{S5BI6Mwx}) before consuming it programmatically, or reshaping data to fit other models.
 While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \cite{o596f2JM}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
@@
 These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles (\cite{hxtgCGjY}}) to embrace openness, transparency, and broad consensus.
+The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \cite{hxtgCGjY} to embrace openness, transparency, and broad consensus.
 In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \cite{jrZe6Esu}.
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP (\cite{p7nwRj8E}}).
+This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \cite{p7nwRj8E}.
 
 It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web (\cite{1H9iQhQYq}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\cite{68fzID2k}}).
+For example, DOIs are mostly resolved on the web \cite{1H9iQhQYq} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \cite{68fzID2k}.
 
@@
 The FAIR Digital Object guidelines \cite{RwvirqWg} sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \cite{13TcbsZF6} and using Linked Data Platform \cite{17OHlMRQA}, as proposed by (\cite{7szz7dwO}}).
+In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \cite{13TcbsZF6} and using Linked Data Platform \cite{17OHlMRQA}, as proposed by \cite{7szz7dwO}.
 
@@
   While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} (\cite{9T4j3N4e}})
+Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} \cite{9T4j3N4e}
 \item
@@
 We argue that the Web, with its now ubiquitous use of REST API \cite{174AwcFUL}, can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services \cite{WzPECPDP} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} (\cite{11dInoyVF}}).
+Note that while early moves for developing Semantic Web Services \cite{WzPECPDP} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} \cite{11dInoyVF}.
 
@@
 
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \cite{RwvirqWg} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} (\cite{PR-RequirementSpec-2.0}}).
+In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \cite{RwvirqWg} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} \cite{PR-RequirementSpec-2.0}.
 
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform (\cite{7szz7dwO}}) comply with the FAIR principles \cite{6DjakjNS}.
+Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform \cite{7szz7dwO} comply with the FAIR principles \cite{6DjakjNS}.
 Here we've used the RDA's FAIR Data Maturity Model \cite{1GrKTFaK2} as it has decomposed the FAIR principles to a structured list of FAIR indicators \cite{UzQhqk0M}, importantly considering \emph{Data} and \emph{Metadata} separately.
@@
 The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\cite{kYjzRJ0H}}).
+At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \cite{kYjzRJ0H}.
 However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.

### 2023-01-25 manuscript.tex
Fixing references to use bibliography [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/76f75247d47f0d152d3ef69cfed06a9feb6ff958.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/76f75247d47f0d152d3ef69cfed06a9feb6ff958/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4001679725

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/1c02a3ca58bd5ba01fcd0dd7266edfe5376fb0c4/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/76f75247d47f0d152d3ef69cfed06a9feb6ff958/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/1c02a3ca58bd5ba01fcd0dd7266edfe5376fb0c4}{stain/2022-fdo-paper@1c02a3c}
-on January 24, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/76f75247d47f0d152d3ef69cfed06a9feb6ff958}{stain/2022-fdo-paper@76f7524}
+on January 25, 2023.
 

### 2023-01-25 manuscript.tex
more regex! [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/83ec084ea920c4a2ffc6631608fbfbe21aaa0737.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/83ec084ea920c4a2ffc6631608fbfbe21aaa0737/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4001720206

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/76f75247d47f0d152d3ef69cfed06a9feb6ff958/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/83ec084ea920c4a2ffc6631608fbfbe21aaa0737/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/76f75247d47f0d152d3ef69cfed06a9feb6ff958}{stain/2022-fdo-paper@76f7524}
+from \href{https://github.com/stain/2022-fdo-paper/tree/83ec084ea920c4a2ffc6631608fbfbe21aaa0737}{stain/2022-fdo-paper@83ec084}
 on January 25, 2023.

### 2023-01-25 latex/main.tex
latex syntax fun

@@
 
-In other discourse, Linked Data \cite{19s3yyxDn} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles (\cite{EJDjHrUv}; \cite{GJHYN6xW}).
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{jHwttlTs}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\cite{99MS5xE0}; \cite{J2H8yssV}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\cite{DNTCUjpC}; \cite{Pa67pUtR}).
+In other discourse, Linked Data \cite{19s3yyxDn} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \cite{EJDjHrUv,GJHYN6xW}.
+Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{jHwttlTs}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \cite{99MS5xE0,J2H8yssV}).
+While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \cite{DNTCUjpC,Pa67pUtR}.
 
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality (\cite{uxWjmlio}; \cite{v0WwFHgT}) when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{5V3GqSzR}.
+One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality \cite{uxWjmlio,v0WwFHgT} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{5V3GqSzR}.
 
@@
 
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins (\cite{bMSxa8XW}; \cite{Qlrlr2TZ}).
+The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins \cite{bMSxa8XW,Qlrlr2TZ}.
 In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
@@
 Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}.
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT} , but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \cite{ZwAcGQKY}, OWL \cite{1p4IWJpI}, OBO \cite{1DalZuKct}, SKOS \cite{15gQDya5B}), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.
+Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \cite{ZwAcGQKY}, OWL \cite{1p4IWJpI}, OBO \cite{1DalZuKct}, SKOS \cite{15gQDya5B}), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.
 
@@
 
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} (\cite{BFMoe8FC}, 2018}) for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \cite{GDrm9dgX} or ShEx \cite{YUzogaV}) can be used to validate RDF Data (\cite{ZeOAPNHa}; \cite{S5BI6Mwx}) before consuming it programmatically, or reshaping data to fit other models.
+Recently there has been a renewed emphasis to improve the \emph{Developer Experience} \cite{BFMoe8FC} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \cite{GDrm9dgX} or ShEx \cite{YUzogaV}) can be used to validate RDF Data \cite{ZeOAPNHa,S5BI6Mwx} before consuming it programmatically, or reshaping data to fit other models.
 While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \cite{o596f2JM}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
@@
 A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \cite{jHwttlTs}.
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \cite{MhsCt4P3}, the use of Linked Data as part of REST Web APIs \cite{eS6a6tlz} , and as shown by the big uptake by publishers to annotate the Web using schema.org Bernstein, Hendler \& Noy \cite{117tX9MEG}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
+One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \cite{MhsCt4P3}, the use of Linked Data as part of REST Web APIs \cite{eS6a6tlz}, and as shown by the big uptake by publishers to annotate the Web using schema.org \cite{117tX9MEG}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
 
@@
 \texttt{0.TYPE/DOIPServiceInfo} lookup is only possible once DOIP server is known.
-g
 \end{itemize}

### 2023-01-25 manuscript.tex
latex syntax fun [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/9a7ff254f386b2d4c6b95bea2ca71d9e215395b0.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/9a7ff254f386b2d4c6b95bea2ca71d9e215395b0/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4001755816

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/83ec084ea920c4a2ffc6631608fbfbe21aaa0737/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/9a7ff254f386b2d4c6b95bea2ca71d9e215395b0/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/83ec084ea920c4a2ffc6631608fbfbe21aaa0737}{stain/2022-fdo-paper@83ec084}
+from \href{https://github.com/stain/2022-fdo-paper/tree/9a7ff254f386b2d4c6b95bea2ca71d9e215395b0}{stain/2022-fdo-paper@9a7ff25}
 on January 25, 2023.

### 2023-01-25 manuscript.tex
lualatex [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/2a8b4df20b46456e0a64f573b633682c1b8abbd8.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/2a8b4df20b46456e0a64f573b633682c1b8abbd8/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4001955882

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/9a7ff254f386b2d4c6b95bea2ca71d9e215395b0/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/2a8b4df20b46456e0a64f573b633682c1b8abbd8/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/9a7ff254f386b2d4c6b95bea2ca71d9e215395b0}{stain/2022-fdo-paper@9a7ff25}
+from \href{https://github.com/stain/2022-fdo-paper/tree/2a8b4df20b46456e0a64f573b633682c1b8abbd8}{stain/2022-fdo-paper@2a8b4df}
 on January 25, 2023.

### 2023-01-27 latex/main.tex
lualatex
Use wlpeerjlua

@@
 
-\documentclass[fleqn,10pt,lineno]{wlpeerj}
+\documentclass[fleqn,10pt,lineno]{wlpeerjlua}
 
@@
 
-While triple-based knowledge representations were not new \cite{NxvhwSmG}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \cite{RRtwPF7Y} are generalised forms of URLs that include locator-less identifiers
-  such as ISBN book numbers (URNs).
+While triple-based knowledge representations were not new \cite{NxvhwSmG}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \cite{RRtwPF7Y} are generalised forms of URLs that include locator-less identifiers such as ISBN book numbers (URNs).
 The distinction between locator-full and locator-less identifiers have weakened in recent years \cite{nW4HY8Nq}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \cite{HqZ3r6J3}.
@@
 \item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP (\cite{16uB3jxpa} DONA Foundation'}}) supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
+  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP \cite{16uB3jxpa} supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
 Multiplexing was removed for the much simplified DOIP 2.0 \cite{13TcbsZF6}.

### 2023-01-27 manuscript.tex
Merge overleaf-2023-01-27-0107 into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/6539cec27b37606a106bbd772e4c0e6ebcaf9329.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/6539cec27b37606a106bbd772e4c0e6ebcaf9329/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4020595803

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/2a8b4df20b46456e0a64f573b633682c1b8abbd8/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/6539cec27b37606a106bbd772e4c0e6ebcaf9329/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/2a8b4df20b46456e0a64f573b633682c1b8abbd8}{stain/2022-fdo-paper@2a8b4df}
-on January 25, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/6539cec27b37606a106bbd772e4c0e6ebcaf9329}{stain/2022-fdo-paper@6539cec}
+on January 27, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 January 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 26 January 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 26 January 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 January 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 26 January 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 January 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 26 January 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 January 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 26 January 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 January 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 26 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 24 January 2023).
+DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 26 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 24 January 2023).
+DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 26 January 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 January 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 26 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 January 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 26 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 January 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 26 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 January 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 26 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 24 January 2023).
+Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 26 January 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 January 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 26 January 2023).
 
@@
 martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 January 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 26 January 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 January 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 26 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 January 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 26 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 January 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 26 January 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 January 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 26 January 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 January 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 26 January 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 January 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 26 January 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 January 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 26 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 January 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 26 January 2023).
 

### 2023-01-30 latex/main.tex
Updates from Overleaf
table tidy

--- 2023-01-25 latex/main.tex
@@
 
+%% Note: This latex project needs lualatex from texlive 2022 or later.
+
 %% Definitions needed for pandoc's silly output
-
 \usepackage{hyperref}
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \usepackage{longtable,booktabs,array}
-
-%% CSL definitions for pandoc's generated references
-%% https://github.com/jgm/pandoc/blob/2.19.2/data/templates/default.latex#L302
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{0.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{2em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-%% END
-
-%% STTYX2Fonts for lualatex
-% https://www.overleaf.com/learn/latex/Articles%2FOpenType-based_math_typesetting%3A_An_introduction_to_the_STIX2_OpenType_fonts
-\usepackage{amsmath}
-%\usepackage{hyperref}
-\usepackage{fontspec}
-\usepackage{unicode-math}
-%\usepackage{hologo}
-%\usepackage{graphicx}
-\setmathfont{STIX2Math}[
-Extension={.otf},
-Path=./STIX2fonts/,
-Scale=1]
-\setmainfont{STIX2Text}[
-Extension={.otf},
-Path=./STIX2fonts/,
-UprightFont={*-Regular},
-BoldFont={*-Bold},
-ItalicFont={*-Italic},
-BoldItalicFont={*-BoldItalic}]
-%% END
+%% END pandoc
 
@@
 %%% Additional packages/definitions
-\usepackage{citation-style-language}
-\cslsetup{
-style = peerj
-}
-\addbibresource{bibliography.json}
+%\usepackage{citation-style-language}
+%\cslsetup{
+%style = peerj
+%}
+%\addbibresource{bibliography.json}
 
@@
 
-The FAIR principles \cite{6DjakjNS} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
+The FAIR principles \cite{wilkinson_fair_2016} and \cite{working_group_doip_2022} and \cite{noauthor_info_nodate} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
 They have been widely recognised by the research community and policy makers as a goal to strive for.
@@
 
-%%\bibliography{sample}
-\printbibliography
+\bibliography{bibliography}
+%\printbibliography
 

### 2023-01-30 latex/table1.tex
table tidy

@@
 \endhead
-\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact/align/collaborate/cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
+\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact, align, collaborate or cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
 FDO encourages research communities to develop common type registries to be shared across instances.

### 2023-01-30 manuscript.tex
table tidy [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/6dae9e62c88e9c06696451de74996e91a9a05be7.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/6dae9e62c88e9c06696451de74996e91a9a05be7/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4042087802

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/6539cec27b37606a106bbd772e4c0e6ebcaf9329/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/6dae9e62c88e9c06696451de74996e91a9a05be7/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/6539cec27b37606a106bbd772e4c0e6ebcaf9329}{stain/2022-fdo-paper@6539cec}
-on January 27, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/6dae9e62c88e9c06696451de74996e91a9a05be7}{stain/2022-fdo-paper@6dae9e6}
+on January 30, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 26 January 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 30 January 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 26 January 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 30 January 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 26 January 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 30 January 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 26 January 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 30 January 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 26 January 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 30 January 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 26 January 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 30 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 26 January 2023).
+DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 30 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 26 January 2023).
+DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 30 January 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 26 January 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 30 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 26 January 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 30 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 26 January 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 30 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 26 January 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 30 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 26 January 2023).
+Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 30 January 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 26 January 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 30 January 2023).
 
@@
 martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 26 January 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 30 January 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 26 January 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 30 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 26 January 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 30 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 26 January 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 30 January 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 26 January 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 30 January 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 26 January 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 30 January 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 26 January 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 30 January 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 26 January 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 30 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 26 January 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 30 January 2023).
 

### 2023-01-30 latex/bibliography.tex
no "available at"

@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
+%\begin{CSLReferences}{1}{0}
+\begin{thebibliography}{9}
+
+\bibitem[5stardata.info]{ref-11dInoyVF}
+5-star Open Data. \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
+
+\end{thebibliography}
+
 \leavevmode\vadjust pre{\hypertarget{ref-CJuBupCW}{}}%
-. \emph{Available at} \href{https://blog.cloudflare.com/http-3-vs-http-2/}{\emph{https://blog.cloudflare.com/http-3-vs-http-2/}}
+. \href{https://blog.cloudflare.com/http-3-vs-http-2/}{\emph{https://blog.cloudflare.com/http-3-vs-http-2/}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
-. \emph{Available at} \href{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}{\emph{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
+. \href{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}{\emph{https://www.rd-alliance.org/sites/default/files/Cordra.2022.pdf}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 January 2023b).
+14:00-17:00.ISO 16684-1:2019. \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 January 2023b).
 
@@
 14:00-17:00.ISO/IEC 23009-1:2019.
-\emph{Available at} \href{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}{\emph{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}} (accessed 26 May 2022a).
+\href{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}{\emph{https://www.iso.org/cms/render/live/en/sites/isoorg/contents/data/standard/07/93/79329.html}} (accessed 26 May 2022a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
+5-star Open Data. \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1F8S9FeHf}{}}%
-Bioschemas - Bioschemas. \emph{Available at} \href{http://bioschemas.org/}{\emph{http://bioschemas.org/}} (accessed 26 May 2022).
+Bioschemas - Bioschemas. \href{http://bioschemas.org/}{\emph{http://bioschemas.org/}} (accessed 26 May 2022).
 
@@
 Content negotiation - HTTP \textbar{} MDN.
-\emph{Available at} \href{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}{\emph{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content\_negotiation}} (accessed 26 May 2022).
+\href{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}{\emph{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content\_negotiation}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-o4h92LpS}{}}%
-Data - W3C. \emph{Available at} \href{https://www.w3.org/standards/semanticweb/data}{\emph{https://www.w3.org/standards/semanticweb/data}} (accessed 26 May 2022).
+Data - W3C. \href{https://www.w3.org/standards/semanticweb/data}{\emph{https://www.w3.org/standards/semanticweb/data}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-zt8AAjcZ}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 January 2023b).
+Data Catalog Vocabulary (DCAT) - Version 2. \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 January 2023b).
 
 \leavevmode\vadjust pre{\hypertarget{ref-lnH0hEuh}{}}%
-Data Catalog Vocabulary (DCAT) - Version 2. \emph{Available at} \href{https://www.w3.org/TR/vocab-dcat-2/}{\emph{https://www.w3.org/TR/vocab-dcat-2/}} (accessed 26 May 2022a).
+Data Catalog Vocabulary (DCAT) - Version 2. \href{https://www.w3.org/TR/vocab-dcat-2/}{\emph{https://www.w3.org/TR/vocab-dcat-2/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1EW4hLN8U}{}}%
-Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 January 2023).
+Data Catalog Vocabulary (DCAT) - Version 3. \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 January 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 January 2023).
+\href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 January 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 January 2023).
+\href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-wAbZpOsr}{}}%
-DCMI Metadata Terms.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}} (accessed 26 May 2022).
+DCMI Metadata Terms. \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/}} (accessed 26 May 2022).
 
@@
 2018.
-\emph{Available at} \href{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}{\emph{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}} (accessed 26 May 2022).
+\href{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}{\emph{https://ruben.verborgh.org/blog/2018/12/28/designing-a-linked-data-developer-experience/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-16uB3jxpa}{}}%
-Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation. \emph{Available at} \href{https://www.dona.net/doipv1doc}{\emph{https://www.dona.net/doipv1doc}} (accessed 26 May 2022).
+Digital Object Interface Protocol Version 1.0 \textbar{} DONA Foundation. \href{https://www.dona.net/doipv1doc}{\emph{https://www.dona.net/doipv1doc}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 24 January 2023).
+DOI Handbook - Resolution. \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 24 January 2023).
+DOI Resolution Documentation. \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-xuQKRRx5}{}}%
-DOIP and Examples --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip.html}{\emph{https://www.cordra.org/documentation/api/doip.html}} (accessed 26 May 2022).
+DOIP and Examples --- Cordra documentation. \href{https://www.cordra.org/documentation/api/doip.html}{\emph{https://www.cordra.org/documentation/api/doip.html}} (accessed 26 May 2022).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 January 2023).
+\href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 January 2023).
 
@@
 draft-bhutton-json-schema-00.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}{\emph{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}} (accessed 26 May 2022).
+\href{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}{\emph{https://datatracker.ietf.org/doc/html/draft-bhutton-json-schema-00}} (accessed 26 May 2022).
 
@@
 draft-ietf-mediaman-suffixes-00 - Media Types with Multiple Suffixes.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}{\emph{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}} (accessed 26 May 2022).
+\href{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}{\emph{https://datatracker.ietf.org/doc/draft-ietf-mediaman-suffixes/00/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-Amit86Tp}{}}%
-draft-ietf-quic-http-34.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}{\emph{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}} (accessed 26 May 2022).
+draft-ietf-quic-http-34. \href{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}{\emph{https://datatracker.ietf.org/doc/html/draft-ietf-quic-http-34}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-nyrPYHoP}{}}%
-draft-kelly-json-hal-08.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}{\emph{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}} (accessed 26 May 2022).
+draft-kelly-json-hal-08. \href{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}{\emph{https://datatracker.ietf.org/doc/html/draft-kelly-json-hal-08}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-7szz7dwO}{}}%
-FAIR Digital Object Framework Documentation. \emph{Available at} \href{https://fairdigitalobjectframework.org/}{\emph{https://fairdigitalobjectframework.org/}} (accessed 26 May 2022).
+FAIR Digital Object Framework Documentation. \href{https://fairdigitalobjectframework.org/}{\emph{https://fairdigitalobjectframework.org/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-DMEBhHID}{}}%
-FAIR Digital Objects Forum \textbar. \emph{Available at} \href{https://fairdo.org/}{\emph{https://fairdo.org/}} (accessed 26 May 2022).
+FAIR Digital Objects Forum \textbar. \href{https://fairdo.org/}{\emph{https://fairdo.org/}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-XmUY0cZB}{}}%
-FOAF Vocabulary Specification. \emph{Available at} \href{http://xmlns.com/foaf/spec/}{\emph{http://xmlns.com/foaf/spec/}} (accessed 26 May 2022).
+FOAF Vocabulary Specification. \href{http://xmlns.com/foaf/spec/}{\emph{http://xmlns.com/foaf/spec/}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 January 2023).
+Handle.Net Registry. \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-14Utscfe4}{}}%
-HTML Standard. \emph{Available at} \href{https://html.spec.whatwg.org/multipage/microdata.html}{\emph{https://html.spec.whatwg.org/multipage/microdata.html}} (accessed 26 May 2022).
+HTML Standard. \href{https://html.spec.whatwg.org/multipage/microdata.html}{\emph{https://html.spec.whatwg.org/multipage/microdata.html}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 January 2023).
+Hydra W3C Community Group. \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI}{\emph{https://www.w3.org/Provider/Style/URI}} (accessed 26 May 2022a).
+Hypertext Style: Cool URIs don't change. \href{https://www.w3.org/Provider/Style/URI}{\emph{https://www.w3.org/Provider/Style/URI}} (accessed 26 May 2022a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-iME5l1kk}{}}%
-Hypertext Style: Cool URIs don't change. \emph{Available at} \href{https://www.w3.org/Provider/Style/URI.html}{\emph{https://www.w3.org/Provider/Style/URI.html}} (accessed 26 May 2022b).
+Hypertext Style: Cool URIs don't change. \href{https://www.w3.org/Provider/Style/URI.html}{\emph{https://www.w3.org/Provider/Style/URI.html}} (accessed 26 May 2022b).
 
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 January 2023).
+"info" URI Registry (Frozen). \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-by6CfOmv}{}}%
-JSON-LD 1.1. \emph{Available at} \href{http://www.w3.org/TR/json-ld/}{\emph{http://www.w3.org/TR/json-ld/}} (accessed 26 May 2022b).
+JSON-LD 1.1. \href{http://www.w3.org/TR/json-ld/}{\emph{http://www.w3.org/TR/json-ld/}} (accessed 26 May 2022b).
 
 \leavevmode\vadjust pre{\hypertarget{ref-X5SWPqSU}{}}%
-JSON-LD 1.1. \emph{Available at} \href{https://www.w3.org/TR/json-ld/}{\emph{https://www.w3.org/TR/json-ld/}} (accessed 26 May 2022a).
+JSON-LD 1.1. \href{https://www.w3.org/TR/json-ld/}{\emph{https://www.w3.org/TR/json-ld/}} (accessed 26 May 2022a).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 24 January 2023).
+Licenses \& Standards \textbar{} Open Source Initiative. \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-OGXYtGZ8}{}}%
-Linked Data - Design Issues. \emph{Available at} \href{https://www.w3.org/DesignIssues/LinkedData.html}{\emph{https://www.w3.org/DesignIssues/LinkedData.html}} (accessed 26 May 2022).
+Linked Data - Design Issues. \href{https://www.w3.org/DesignIssues/LinkedData.html}{\emph{https://www.w3.org/DesignIssues/LinkedData.html}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-zlGYiuWC}{}}%
-Linked Data Notifications. \emph{Available at} \href{https://www.w3.org/TR/ldn/}{\emph{https://www.w3.org/TR/ldn/}} (accessed 26 May 2022).
+Linked Data Notifications. \href{https://www.w3.org/TR/ldn/}{\emph{https://www.w3.org/TR/ldn/}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 January 2023).
+Linking Across Provenance Bundles. \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 January 2023).
 
@@
 martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 January 2023).
+\href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1Gsq23e51}{}}%
-NCBO BioPortal. \emph{Available at} \href{https://bioportal.bioontology.org/ontologies}{\emph{https://bioportal.bioontology.org/ontologies}} (accessed 26 May 2022).
+NCBO BioPortal. \href{https://bioportal.bioontology.org/ontologies}{\emph{https://bioportal.bioontology.org/ontologies}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-k0AfCGzw}{}}%
-OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More.
-\emph{Available at} \href{https://spec.openapis.org/oas/v3.1.0.html}{\emph{https://spec.openapis.org/oas/v3.1.0.html}} (accessed 26 May 2022).
+OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More. \href{https://spec.openapis.org/oas/v3.1.0.html}{\emph{https://spec.openapis.org/oas/v3.1.0.html}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ID1pTGry}{}}%
-ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 January 2023).
+ORE Specification - Abstract Data Model. \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 January 2023).
+ORE User Guide - Primer. \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1p4IWJpI}{}}%
-OWL 2 Web Ontology Language Document Overview (Second Edition). \emph{Available at} \href{http://www.w3.org/TR/owl2-overview/}{\emph{http://www.w3.org/TR/owl2-overview/}} (accessed 26 May 2022).
+OWL 2 Web Ontology Language Document Overview (Second Edition). \href{http://www.w3.org/TR/owl2-overview/}{\emph{http://www.w3.org/TR/owl2-overview/}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 January 2023a).
+PROV-O: The PROV Ontology. \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 January 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 January 2023b).
+PROV-O: The PROV Ontology. \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 January 2023b).
 
 \leavevmode\vadjust pre{\hypertarget{ref-16WGmCih8}{}}%
-RDF 1.1 Primer. \emph{Available at} \href{http://www.w3.org/TR/rdf11-primer/}{\emph{http://www.w3.org/TR/rdf11-primer/}} (accessed 26 May 2022).
+RDF 1.1 Primer. \href{http://www.w3.org/TR/rdf11-primer/}{\emph{http://www.w3.org/TR/rdf11-primer/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZwAcGQKY}{}}%
-RDF Schema 1.1. \emph{Available at} \href{http://www.w3.org/TR/rdf-schema/}{\emph{http://www.w3.org/TR/rdf-schema/}} (accessed 26 May 2022).
+RDF Schema 1.1. \href{http://www.w3.org/TR/rdf-schema/}{\emph{http://www.w3.org/TR/rdf-schema/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-FKIF6ApB}{}}%
-RDFa 1.1 Primer - Third Edition. \emph{Available at} \href{http://www.w3.org/TR/rdfa-primer/}{\emph{http://www.w3.org/TR/rdfa-primer/}} (accessed 26 May 2022).
+RDFa 1.1 Primer - Third Edition. \href{http://www.w3.org/TR/rdfa-primer/}{\emph{http://www.w3.org/TR/rdfa-primer/}} (accessed 26 May 2022).
 
@@
 Resource Description Framework (RDF) Model and Syntax Specification.
-\emph{Available at} \href{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}{\emph{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}} (accessed 26 May 2022).
+\href{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}{\emph{https://www.w3.org/TR/1999/REC-rdf-syntax-19990222/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-eS5y9TRh}{}}%
-ResourceSync Framework Specification - Table of Contents. \emph{Available at} \href{http://www.openarchives.org/rs/toc}{\emph{http://www.openarchives.org/rs/toc}} (accessed 26 May 2022).
+ResourceSync Framework Specification - Table of Contents. \href{http://www.openarchives.org/rs/toc}{\emph{http://www.openarchives.org/rs/toc}} (accessed 26 May 2022).
 
@@
 Sandvine.Global Internet Phenomena Report 2022.
-\emph{Available at} \href{https://www.sandvine.com/global-internet-phenomena-report-2022}{\emph{https://www.sandvine.com/global-internet-phenomena-report-2022}} (accessed 26 May 2022).
+\href{https://www.sandvine.com/global-internet-phenomena-report-2022}{\emph{https://www.sandvine.com/global-internet-phenomena-report-2022}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10pzKCEUE}{}}%
-Schema.org - Schema.org. \emph{Available at} \href{https://schema.org/}{\emph{https://schema.org/}} (accessed 26 May 2022).
+Schema.org - Schema.org. \href{https://schema.org/}{\emph{https://schema.org/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-dKAekUjL}{}}%
-Schema.org Actions - schema.org. \emph{Available at} \href{https://schema.org/docs/actions.html}{\emph{https://schema.org/docs/actions.html}} (accessed 26 May 2022).
+Schema.org Actions - schema.org. \href{https://schema.org/docs/actions.html}{\emph{https://schema.org/docs/actions.html}} (accessed 26 May 2022).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 January 2023).
+\href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
-Shape Expressions (ShEx) 2.1 Primer. \emph{Available at} \href{http://shex.io/shex-primer/}{\emph{http://shex.io/shex-primer/}} (accessed 26 May 2022).
+Shape Expressions (ShEx) 2.1 Primer. \href{http://shex.io/shex-primer/}{\emph{http://shex.io/shex-primer/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-GDrm9dgX}{}}%
-Shapes Constraint Language (SHACL). \emph{Available at} \href{https://www.w3.org/TR/shacl/}{\emph{https://www.w3.org/TR/shacl/}} (accessed 26 May 2022).
+Shapes Constraint Language (SHACL). \href{https://www.w3.org/TR/shacl/}{\emph{https://www.w3.org/TR/shacl/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-15gQDya5B}{}}%
-SKOS Simple Knowledge Organization System Primer. \emph{Available at} \href{http://www.w3.org/TR/skos-primer/}{\emph{http://www.w3.org/TR/skos-primer/}} (accessed 26 May 2022).
+SKOS Simple Knowledge Organization System Primer. \href{http://www.w3.org/TR/skos-primer/}{\emph{http://www.w3.org/TR/skos-primer/}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xww8fb1B}{}}%
-SPARQL 1.1 Overview. \emph{Available at} \href{http://www.w3.org/TR/sparql11-overview/}{\emph{http://www.w3.org/TR/sparql11-overview/}} (accessed 26 May 2022).
+SPARQL 1.1 Overview. \href{http://www.w3.org/TR/sparql11-overview/}{\emph{http://www.w3.org/TR/sparql11-overview/}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-7DGllKzG}{}}%
-SWORD 3.0 Specification. \emph{Available at} \href{https://swordapp.github.io/swordv3/swordv3.html}{\emph{https://swordapp.github.io/swordv3/swordv3.html}} (accessed 26 May 2022).
+SWORD 3.0 Specification. \href{https://swordapp.github.io/swordv3/swordv3.html}{\emph{https://swordapp.github.io/swordv3/swordv3.html}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 January 2023).
+The ARK Identifier Scheme. \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 January 2023).
+The Modern Standards Paradigm - Five Key Principles. \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-oN9D5LkD}{}}%
-The Open Graph protocol. \emph{Available at} \href{https://ogp.me/}{\emph{https://ogp.me/}} (accessed 26 May 2022).
+The Open Graph protocol. \href{https://ogp.me/}{\emph{https://ogp.me/}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 January 2023).
+Unknown.Introducing 'Role'. \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-j3SqQLBR}{}}%
-Usage Statistics of JSON-LD for Websites, May 2022.
-\emph{Available at} \href{https://w3techs.com/technologies/details/da-jsonld}{\emph{https://w3techs.com/technologies/details/da-jsonld}} (accessed 26 May 2022).
+Usage Statistics of JSON-LD for Websites, May 2022. \href{https://w3techs.com/technologies/details/da-jsonld}{\emph{https://w3techs.com/technologies/details/da-jsonld}} (accessed 26 May 2022).
 
@@
 2022.FAIR Signposting Profile.
-\emph{Available at} \href{https://signposting.org/FAIR/}{\emph{https://signposting.org/FAIR/}} (accessed 5 January 2023).
+\href{https://signposting.org/FAIR/}{\emph{https://signposting.org/FAIR/}} (accessed 5 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
-W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures. \emph{Available at} \href{http://www.w3.org/TR/xmlschema11-1/}{\emph{http://www.w3.org/TR/xmlschema11-1/}} (accessed 26 May 2022).
+W3C XML Schema Definition Language (XSD) 1.1 Part 1: Structures. \href{http://www.w3.org/TR/xmlschema11-1/}{\emph{http://www.w3.org/TR/xmlschema11-1/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-LmudecyN}{}}%
-Web Services Description Language (WSDL) Version 2.0 Part 0: Primer. \emph{Available at} \href{http://www.w3.org/TR/wsdl20-primer/}{\emph{http://www.w3.org/TR/wsdl20-primer/}} (accessed 26 May 2022).
+Web Services Description Language (WSDL) Version 2.0 Part 0: Primer. \href{http://www.w3.org/TR/wsdl20-primer/}{\emph{http://www.w3.org/TR/wsdl20-primer/}} (accessed 26 May 2022).
 
 \leavevmode\vadjust pre{\hypertarget{ref-qgqADk45}{}}%
-WebSockets Standard. \emph{Available at} \href{https://websockets.spec.whatwg.org/}{\emph{https://websockets.spec.whatwg.org/}} (accessed 26 May 2022).
+WebSockets Standard. \href{https://websockets.spec.whatwg.org/}{\emph{https://websockets.spec.whatwg.org/}} (accessed 26 May 2022).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-103Hw8H43}{}}%
-X.1255 : Framework for discovery of identity management information.
-\emph{Available at} \href{https://www.itu.int/rec/T-REC-X.1255-201309-I}{\emph{https://www.itu.int/rec/T-REC-X.1255-201309-I}} (accessed 26 May 2022).
+X.1255 : Framework for discovery of identity management information. \href{https://www.itu.int/rec/T-REC-X.1255-201309-I}{\emph{https://www.itu.int/rec/T-REC-X.1255-201309-I}} (accessed 26 May 2022).
 

### 2023-01-31 manuscript.tex
re-export from Zotero using Better BibTex [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/77065153bd22894490553066987dc5130d835eea.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/77065153bd22894490553066987dc5130d835eea/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4054552800

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/6dae9e62c88e9c06696451de74996e91a9a05be7/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/77065153bd22894490553066987dc5130d835eea/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/6dae9e62c88e9c06696451de74996e91a9a05be7}{stain/2022-fdo-paper@6dae9e6}
-on January 30, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/77065153bd22894490553066987dc5130d835eea}{stain/2022-fdo-paper@7706515}
+on January 31, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 30 January 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 31 January 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 30 January 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 31 January 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 30 January 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 31 January 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 30 January 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 31 January 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 30 January 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 31 January 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 30 January 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 31 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 30 January 2023).
+DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 31 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 30 January 2023).
+DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 31 January 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 30 January 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 31 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 30 January 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 31 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 30 January 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 31 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 30 January 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 31 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 30 January 2023).
+Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 31 January 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 30 January 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 31 January 2023).
 
@@
 martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 30 January 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 31 January 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 30 January 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 31 January 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 30 January 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 31 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 30 January 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 31 January 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 30 January 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 31 January 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 30 January 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 31 January 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 30 January 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 31 January 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 30 January 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 31 January 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 30 January 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 31 January 2023).
 

### 2023-01-31 manuscript.tex
Merge branch 'main' of github.com:stain/2022-fdo-paper into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/484ae8b4b9e9611eeeb1e0e67f96d5b8bd8b70a4.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/484ae8b4b9e9611eeeb1e0e67f96d5b8bd8b70a4/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4054549790

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/77065153bd22894490553066987dc5130d835eea/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/484ae8b4b9e9611eeeb1e0e67f96d5b8bd8b70a4/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/77065153bd22894490553066987dc5130d835eea}{stain/2022-fdo-paper@7706515}
+from \href{https://github.com/stain/2022-fdo-paper/tree/484ae8b4b9e9611eeeb1e0e67f96d5b8bd8b70a4}{stain/2022-fdo-paper@484ae8b}
 on January 31, 2023.

### 2023-01-31 latex/main.tex
some citation keys fixed

@@
 
-The FAIR principles \cite{wilkinson_fair_2016} and \cite{working_group_doip_2022} and \cite{noauthor_info_nodate} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
+The FAIR principles \cite{wilkinsonFAIRGuidingPrinciples2016e} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
 They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \cite{u8Cy0psL}.
-The EOSC Interoperability Framework \cite{aCye3KpE} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
+In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \cite{monsCloudyIncreasinglyFAIR2017b}.
+The EOSC Interoperability Framework \cite{corchoEOSCInteroperabilityFramework2021b} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
 
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object \cite{IHLT6hye} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
+Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object \cite{schultesFAIRPrinciplesDigital2019a} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
 The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
 
-In other discourse, Linked Data \cite{19s3yyxDn} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \cite{EJDjHrUv,GJHYN6xW}.
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{jHwttlTs}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \cite{99MS5xE0,J2H8yssV}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \cite{DNTCUjpC,Pa67pUtR}.
+In other discourse, Linked Data \cite{bizerLinkedDataStory2009a} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \cite{boninodasilvasantosFAIRDataPoints2016a,hasnainAssessingFAIRData2018a}.
+Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \cite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}).
+While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \cite{turcoaneLinkedDataJSONLD2014a,wilkinsonWorkflowsWhenParts2022b}.
 
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality \cite{uxWjmlio,v0WwFHgT} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{5V3GqSzR}.
+One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality \cite{stefiDevelopersMakeUnbiased2015,stefiDevelopReuseTwo2015a} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{merono-penuelaWebDataApis2021b}.
 

### 2023-01-31 manuscript.tex
some citation keys fixed [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/bfaaab005d18e6d1a45e17f5adbda0cf18649697.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/bfaaab005d18e6d1a45e17f5adbda0cf18649697/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4054891245

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/484ae8b4b9e9611eeeb1e0e67f96d5b8bd8b70a4/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/bfaaab005d18e6d1a45e17f5adbda0cf18649697/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/484ae8b4b9e9611eeeb1e0e67f96d5b8bd8b70a4}{stain/2022-fdo-paper@484ae8b}
+from \href{https://github.com/stain/2022-fdo-paper/tree/bfaaab005d18e6d1a45e17f5adbda0cf18649697}{stain/2022-fdo-paper@bfaaab0}
 on January 31, 2023.

### 2023-01-31 latex/main.tex

--- 2023-01-31 latex/main.tex
@@
 \usepackage{pdflscape}
+
+% https://www.overleaf.com/latex/examples/bibliographies-with-biber-and-biblatex/ccrkczqwnywf
+% use biblatex/biber as natbib is from 1990s and have no DOIs
+\usepackage[autostyle]{csquotes}
+\usepackage[backend=biber,style=authoryear]{biblatex}
+\addbibresource{bibliography.bib}
+
 %%% END own imports
@@
 \section*{Introduction}
+Hello
 
@@
 
-The FAIR principles \cite{wilkinson_fair_2016} and \cite{working_group_doip_2022} and \cite{noauthor_info_nodate} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
+The FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
 They have been widely recognised by the research community and policy makers as a goal to strive for.
-In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \cite{u8Cy0psL}.
-The EOSC Interoperability Framework \cite{aCye3KpE} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
+In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \autocite{monsCloudyIncreasinglyFAIR2017b}.
+The EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
 
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object \cite{IHLT6hye} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
+Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object \autocite{schultesFAIRPrinciplesDigital2019a} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
 The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
 
-In other discourse, Linked Data \cite{19s3yyxDn} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \cite{EJDjHrUv,GJHYN6xW}.
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \cite{jHwttlTs}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \cite{99MS5xE0,J2H8yssV}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \cite{DNTCUjpC,Pa67pUtR}.
+In other discourse, Linked Data \autocite{bizerLinkedDataStory2009a} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \autocite{boninodasilvasantosFAIRDataPoints2016a,hasnainAssessingFAIRData2018a}.
+Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \autocite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \autocite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}).
+While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \autocite{turcoaneLinkedDataJSONLD2014a,wilkinsonWorkflowsWhenParts2022b}.
 
-One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality \cite{uxWjmlio,v0WwFHgT} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \cite{5V3GqSzR}.
+One challenge that may, perhaps counter-intuitively, steer developers towards a not-invented-here mentality \autocite{stefiDevelopersMakeUnbiased2015,stefiDevelopReuseTwo2015a} when exposing their data on the Web is the heterogeneity and apparent complexity of Semantic Web approaches themselves \autocite{merono-penuelaWebDataApis2021b}.
 
@@
 
-\bibliography{bibliography}
-%\printbibliography
+%\bibliography{bibliography}
+\printbibliography
 

### 2023-01-31 manuscript.tex
now using biblatex [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/a78fcc264323c8aac5b5620d946ef6d07686c498.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/a78fcc264323c8aac5b5620d946ef6d07686c498/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4055058960

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/bfaaab005d18e6d1a45e17f5adbda0cf18649697/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/a78fcc264323c8aac5b5620d946ef6d07686c498/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/bfaaab005d18e6d1a45e17f5adbda0cf18649697}{stain/2022-fdo-paper@bfaaab0}
+from \href{https://github.com/stain/2022-fdo-paper/tree/a78fcc264323c8aac5b5620d946ef6d07686c498}{stain/2022-fdo-paper@a78fcc2}
 on January 31, 2023.

### 2023-01-31 manuscript.tex
Merge branch 'main' of github.com:stain/2022-fdo-paper into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/c0dc59e999a0bc215e5d5e0509a2cf6927bcbba4.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/c0dc59e999a0bc215e5d5e0509a2cf6927bcbba4/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4055104070

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/a78fcc264323c8aac5b5620d946ef6d07686c498/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/c0dc59e999a0bc215e5d5e0509a2cf6927bcbba4/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/a78fcc264323c8aac5b5620d946ef6d07686c498}{stain/2022-fdo-paper@a78fcc2}
+from \href{https://github.com/stain/2022-fdo-paper/tree/c0dc59e999a0bc215e5d5e0509a2cf6927bcbba4}{stain/2022-fdo-paper@c0dc59e}
 on January 31, 2023.

### 2023-01-31 manuscript.tex
Some citations [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/91171dc6c031928ebbaa06767feaef4bcede30f4.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/91171dc6c031928ebbaa06767feaef4bcede30f4/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4055611488

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/c0dc59e999a0bc215e5d5e0509a2cf6927bcbba4/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/91171dc6c031928ebbaa06767feaef4bcede30f4/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/c0dc59e999a0bc215e5d5e0509a2cf6927bcbba4}{stain/2022-fdo-paper@c0dc59e}
+from \href{https://github.com/stain/2022-fdo-paper/tree/91171dc6c031928ebbaa06767feaef4bcede30f4}{stain/2022-fdo-paper@91171dc}
 on January 31, 2023.

### 2023-01-31 manuscript.tex
avoid {{\_ bugs [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/8bbdd67883c9ed6306479264d368472288bd67d0.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/8bbdd67883c9ed6306479264d368472288bd67d0/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4056063800

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/91171dc6c031928ebbaa06767feaef4bcede30f4/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/8bbdd67883c9ed6306479264d368472288bd67d0/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/91171dc6c031928ebbaa06767feaef4bcede30f4}{stain/2022-fdo-paper@91171dc}
+from \href{https://github.com/stain/2022-fdo-paper/tree/8bbdd67883c9ed6306479264d368472288bd67d0}{stain/2022-fdo-paper@8bbdd67}
 on January 31, 2023.

### 2023-01-31 manuscript.tex
FDO citations [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/36d2cc87aca77995c39b6fea4ee9a6aa69587b8e.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/36d2cc87aca77995c39b6fea4ee9a6aa69587b8e/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4056315166

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/8bbdd67883c9ed6306479264d368472288bd67d0/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/36d2cc87aca77995c39b6fea4ee9a6aa69587b8e/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/8bbdd67883c9ed6306479264d368472288bd67d0}{stain/2022-fdo-paper@8bbdd67}
+from \href{https://github.com/stain/2022-fdo-paper/tree/36d2cc87aca77995c39b6fea4ee9a6aa69587b8e}{stain/2022-fdo-paper@36d2cc8}
 on January 31, 2023.

### 2023-01-31 latex/main.tex
Some citations
some FDO citations

@@
 \addbibresource{bibliography.bib}
+\DeclareFieldFormat{doi}{%
+  \url{https://doi.org/#1}}
 
@@
 \thispagestyle{empty}
-
-\section*{Introduction}
-Hello
-
-\subsection*{About PeerJ}
-
-Here goes
 
@@
 
-The FAIR Digital Object Forum \cite{DMEBhHID} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
+The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
 As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
@@
 
-\textbf{FAIR Digital Object Overview and Specifications} \cite{11NpETLq1} is a comprehensive overview of FAIR Digital Object specifications listed below.
+\textbf{FAIR Digital Object Overview and Specifications} \autocite{fdo-Overview} is a comprehensive overview of FAIR Digital Object specifications listed below.
 It serves as a primer that introduces FDO concepts and the remaining documents.
 
-The \textbf{FDO Forum Document Standards} \cite{TQdku4YF} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
+The \textbf{FDO Forum Document Standards} \autocite{fdo-DocProcessStd} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
 In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
 
-The \textbf{FDO Requirement Specifications} \cite{yygVPoL0} is an update of \cite{RwvirqWg} as the foundational definition of FDO.
+The \textbf{FDO Requirement Specifications} \autocite{fdo-RequirementSpec} is an update of \autocite{boninoFAIRDigitalObject} as the foundational definition of FDO.
 This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
@@
 
-The \textbf{Machine actionability} \cite{iJeys0T5} sets out to define what is meant by \emph{machine actionability} for FDOs.
+The \textbf{Machine actionability} \autocite{fdo-MachineActionDef} sets out to define what is meant by \emph{machine actionability} for FDOs.
 \emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
@@
 
-\textbf{Configuration Types} \cite{SPd0uEqO} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
+\textbf{Configuration Types} \autocite{fdo-ConfigurationTypes} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
 Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \cite{snykkm7R}
 
-\textbf{PID Profiles \& Attributes} \cite{ljVViWCl} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \cite{lCoshHXg}.
+\textbf{PID Profiles \& Attributes} \autocite{fdo-PIDProfileAttributes} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
+This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \autocite{weigelRDARecommendationPID2018}.
 This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
 
-\textbf{Kernel Attributes \& Metadata} \cite{Fi62cJAo} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
+\textbf{Kernel Attributes \& Metadata} \autocite{fdo-KernelAttributes} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
 This document expands on RDA Recommendation on PID Kernel Information \cite{lCoshHXg}.

### 2023-01-31 latex/main.tex
more FDO references

@@
 \textbf{Configuration Types} \autocite{fdo-ConfigurationTypes} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \cite{snykkm7R}
+Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022}
 
@@
 \textbf{Kernel Attributes \& Metadata} \autocite{fdo-KernelAttributes} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information \cite{lCoshHXg}.
+This document expands on RDA Recommendation on PID Kernel Information \autocite{weigelRDARecommendationPID2018}.
 It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
 
-\textbf{Granularity, Versioning, Mutability} \cite{16W9AycSW} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
+\textbf{Granularity, Versioning, Mutability} \autocite{fdo-Granularity} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
 The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
 
-\textbf{DOIP Endorsement Request} \cite{15yGpJ0wh} is an endorsement of the DOIP v2.0 \cite{13TcbsZF6} specification as a potential FDO implementation, as it has been applied by several institutions \cite{jRWurmQc}.
+\textbf{DOIP Endorsement Request} \autocite{fdo-DOIPEndorsement} is an endorsement of the DOIP v2.0 \autocite{foundationDigitalObjectInterface} specification as a potential FDO implementation, as it has been applied by several institutions \autocite{wittenburgFAIRDigitalObject2022b}.
 The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
 
-\textbf{Upload of FDO} \cite{1H7cmmr69} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync \cite{eS5y9TRh} is suggested as one type of service to list FDOs.
+\textbf{Upload of FDO} \autocite{fdo-FDO-Upload} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
+ResourceSync \autocite{ResourceSyncFrameworkSpecification} is suggested as one type of service to list FDOs.
 This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
 
-\textbf{Typing FAIR Digital Objects} \cite{ZFzPxCNB} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
+\textbf{Typing FAIR Digital Objects} \autocite{fdo-TypingFDOs} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
 This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
@@
 
-\textbf{Implementation of Attributes, Types, Profiles and Registries} \cite{126uxr5pI} details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
+\textbf{Implementation of Attributes, Types, Profiles and Registries} \autocite{fdo-ImplAttributesTypesProfiles} details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
 This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} \cite{1CqIZ47pu}.
+Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} \autocite{schwardmannTwoExamplesHow2022}.
 
@@
 
-The concept of \textbf{FAIR Digital Objects} \cite{IHLT6hye} has been introduced as way to expose research data as active objects that conform to the FAIR principles \cite{6DjakjNS}.
-This builds on the \emph{Digital Object} (DO) concept \cite{11MnuwJ4l}, first introduced in 1995 \cite{3Uqe3fuK} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework \cite{103Hw8H43} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 \cite{16uB3jxpa}.
+The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
+This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced in 1995 \autocite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
+DO was the inspiration for the ITU X.1255 framework \autocite{x1255FrameworkDiscovery} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 \autocite{DigitalObjectInterface}.
 
@@
 The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing \cite{tz0P3DTC}.
+The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing \autocite{wittenburgDigitalObjectsDrivers2019a}.
 
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \cite{aCye3KpE}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \cite{RwvirqWg}.
+Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
+Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}.
 The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).

### 2023-01-31 manuscript.tex
more FDO references [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/8ed1966988bc19593b6e558205553033af6267a7.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/8ed1966988bc19593b6e558205553033af6267a7/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4058378332

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/36d2cc87aca77995c39b6fea4ee9a6aa69587b8e/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/8ed1966988bc19593b6e558205553033af6267a7/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/36d2cc87aca77995c39b6fea4ee9a6aa69587b8e}{stain/2022-fdo-paper@36d2cc8}
+from \href{https://github.com/stain/2022-fdo-paper/tree/8ed1966988bc19593b6e558205553033af6267a7}{stain/2022-fdo-paper@8ed1966}
 on January 31, 2023.

### 2023-01-31 manuscript.tex
tiny URL fonts [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/f600bf52e21065865fb4f3a0f1fb349145c9e9ce.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/f600bf52e21065865fb4f3a0f1fb349145c9e9ce/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4059272240

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/8ed1966988bc19593b6e558205553033af6267a7/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/f600bf52e21065865fb4f3a0f1fb349145c9e9ce/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/8ed1966988bc19593b6e558205553033af6267a7}{stain/2022-fdo-paper@8ed1966}
+from \href{https://github.com/stain/2022-fdo-paper/tree/f600bf52e21065865fb4f3a0f1fb349145c9e9ce}{stain/2022-fdo-paper@f600bf5}
 on January 31, 2023.

### 2023-01-31 manuscript.tex
please give me xelatex.. [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/a9e59367a493c3366c2f044abed56115cd182b08.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/a9e59367a493c3366c2f044abed56115cd182b08/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4059405651

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/f600bf52e21065865fb4f3a0f1fb349145c9e9ce/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/a9e59367a493c3366c2f044abed56115cd182b08/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/f600bf52e21065865fb4f3a0f1fb349145c9e9ce}{stain/2022-fdo-paper@f600bf5}
+from \href{https://github.com/stain/2022-fdo-paper/tree/a9e59367a493c3366c2f044abed56115cd182b08}{stain/2022-fdo-paper@a9e5936}
 on January 31, 2023.

### 2023-02-01 latex/main.tex
tiny URL fonts
please give me xelatex..
reordered FDO section

@@
 
+The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
+This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced in 1995 \autocite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
+DO was the inspiration for the ITU X.1255 framework \autocite{x1255FrameworkDiscovery} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 \autocite{DigitalObjectInterface}.
+
+In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}.
+The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
+The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing \autocite{wittenburgDigitalObjectsDrivers2019a}.
+
+Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
+Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}.
+The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
+
+\hypertarget{fdo-approaches}{%
+\subsubsection*{FDO approaches}\label{fdo-approaches}}
+
+FDO is an evolving concept. A set of FDO Demonstrators \autocite{wittenburgFAIRDigitalObject2022b} highlight how current adapters are approaching implementations of FDO from different angles:
+
+\begin{itemize}
+\tightlist
+\item
+  Building on the Digital Object concept, using the simplified DOIP v2 specification \autocite{foundationDigitalObjectInterface}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0, see \autocite{DOIPExamplesCordraa}} (usually TCP/IP over TLS).
+The main DOIP operations are retrieving, creating and updating digital objects.
+These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
+FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \autocite{Draftbhuttonjsonschema00} and PIDs are assigned using the Handle system.
+Several type registries have been established.
+\item
+  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
+\item
+  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
+\end{itemize}
+
+From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
+
 \hypertarget{sec:next-step-fdo}{%
@@
 It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-
-The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
-This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced in 1995 \autocite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework \autocite{x1255FrameworkDiscovery} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 \autocite{DigitalObjectInterface}.
-
-In brief, the structure of a FAIR Digital Object (FDO) is to, given a \emph{persistent identifier} (PID) such as a DOI, resolve to a \emph{PID Record} that gives the object a \emph{type} along with a mechanism to retrieve its \emph{bit sequences}, \emph{metadata} and references to further programmatic \emph{operations}.
-The type of an FDO (itself an FDO) defines attributes to semantically describe and relate such FDOs to other concepts (typically other FDOs referenced by PIDs).
-The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing \autocite{wittenburgDigitalObjectsDrivers2019a}.
-
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
-Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}.
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
-
-\hypertarget{fdo-approaches}{%
-\subsubsection*{FDO approaches}\label{fdo-approaches}}
-
-FDO is an evolving concept. A set of FDO Demonstrators \cite{jRWurmQc} highlight how current adapters are approaching implementations of FDO from different angles:
-
-\begin{itemize}
-\tightlist
-\item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification \cite{13TcbsZF6}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0, see \cite{xuQKRRx5}} (usually TCP/IP over TLS).
-The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \cite{15EZ2D0Rm} and PIDs are assigned using the Handle system.
-Several type registries have been established.
-\item
-  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
-\item
-  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
-\end{itemize}
-
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
 

### 2023-02-01 manuscript.tex
Merge branch 'main' of github.com:stain/2022-fdo-paper into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/5c660e0d407c65b4f669383e77cc59e770e73c12.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/5c660e0d407c65b4f669383e77cc59e770e73c12/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4063796569

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/a9e59367a493c3366c2f044abed56115cd182b08/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/5c660e0d407c65b4f669383e77cc59e770e73c12/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/a9e59367a493c3366c2f044abed56115cd182b08}{stain/2022-fdo-paper@a9e5936}
-on January 31, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/5c660e0d407c65b4f669383e77cc59e770e73c12}{stain/2022-fdo-paper@5c660e0}
+on February 1, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 31 January 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 1 February 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 31 January 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 1 February 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 31 January 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 1 February 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 31 January 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 1 February 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 31 January 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 1 February 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 31 January 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 1 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 31 January 2023).
+DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 1 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 31 January 2023).
+DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 1 February 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 31 January 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 1 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 31 January 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 1 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 31 January 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 1 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 31 January 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 1 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 31 January 2023).
+Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 1 February 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 31 January 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 1 February 2023).
 
@@
 martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 31 January 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 1 February 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 31 January 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 1 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 31 January 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 1 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 31 January 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 1 February 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 31 January 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 1 February 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 31 January 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 1 February 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 31 January 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 1 February 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 31 January 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 1 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 31 January 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 1 February 2023).
 

### 2023-02-01 latex/main.tex
more references sorted

@@
 
+% Make tt font not so massive
+\renewcommand*\ttdefault{cmvtt}
+
 % https://www.overleaf.com/latex/examples/bibliographies-with-biber-and-biblatex/ccrkczqwnywf
@@
   \url{https://doi.org/#1}}
+%\urlstyle{tt}
 
@@
 
-\author[1,2]{Stian Soiland-Reyes}
-\author[1]{Carole Goble}
-\author[2]{Paul Groth}
+\author[1,2]{Stian Soiland-Reyes} % https://orcid.org/0000-0001-9842-9718
+\author[1]{Carole Goble}          % https://orcid.org/0000-0003-1219-2137
+\author[2]{Paul Groth}            % https://orcid.org/0000-0003-0183-6910
+
 \affil[1]{Department of Computer Science, The University of Manchester, UK}
@@
 
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \cite{Guy7cjVY}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
+The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \autocite{berners-leeWeavingWebOriginal1999}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
 The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
@@
 
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \cite{16WGmCih8} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \cite{16wvFH4nh} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
+Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \autocite{w3-rdf11-primer} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \autocite{w3-rdf-syntax} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
 
-While triple-based knowledge representations were not new \cite{NxvhwSmG}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \cite{RRtwPF7Y} are generalised forms of URLs that include locator-less identifiers such as ISBN book numbers (URNs).
-The distinction between locator-full and locator-less identifiers have weakened in recent years \cite{nW4HY8Nq}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \cite{HqZ3r6J3}.
-RDF 1.1 formats use Unicode to support IRIs \cite{rqjuXkEf}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \cite{rnzJ4khD}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
+While triple-based knowledge representations were not new \autocite{stanczykProcessModellingInformation1987}, the main innovation of RDF was the use of global identifiers in the form of URIs\footnote{URIs \autocite{rfc3986} are generalised forms of URLs that include locator-less identifiers such as ISBN book numbers (URNs).
+The distinction between locator-full and locator-less identifiers have weakened in recent years \autocite{InfoURIRegistry}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \autocite{jutyIdentifiersOrgMIRIAM2011}.
+RDF 1.1 formats use Unicode to support IRIs \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
+By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \autocite{sauermannCoolURIsSemantic2011}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
 
@@
 The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \cite{yd2WE9eq} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
+More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
 
-Established Semantic Web services and standards include SPARQL \cite{xww8fb1B} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
+Established Semantic Web services and standards include SPARQL \autocite{w3-sparql11-overview} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
 
-The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins \cite{bMSxa8XW,Qlrlr2TZ}.
+The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins \autocite{gobleStateNationData2008c,williamsOpenPHACTSSemantic2012c}.
 In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
@@
 Facing these challenges, thanks to the use of universal identifiers in the form of URIs, \emph{mappings} could retrospectively be developed not just between resources, but also across vocabularies.
-Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \cite{15gQDya5B} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \cite{FDDmgO6s} .
+Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \autocite{w3-skos-primer} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
+Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \autocite{huHowMatchableAre2011a} .
 
 The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
-This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \cite{191ZkYZMt}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
+This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \autocite{grothAPIcentricLinkedData2014b}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
 
 With these trends, an emerging problem was that adopters of the Semantic Web primarily utillised it as a set of graph technologies, with little consideration to existing Web resources.
-This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \cite{9CdJnsoJ}.
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \cite{AGHR9oyr}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \cite{5SuwLNdE}.
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \cite{TUmA4KQT}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \cite{ZwAcGQKY}, OWL \cite{1p4IWJpI}, OBO \cite{1DalZuKct}, SKOS \cite{15gQDya5B}), hash vs slash, HTTP status codes and PID redirection strategies \cite{rnzJ4khD}.
+This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \autocite{kamdarSystematicAnalysisTerm2017a}.
+Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \autocite{kleinScholarlyContextNot2014a}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \autocite{polleresMoreDecentralizedVision2020a}.
+Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), hash vs slash, HTTP status codes and PID redirection strategies \autocite{sauermannCoolURIsSemantic2011}.
 
@@
 %\bibliography{bibliography}
+\def\UrlFont{\small}
 \printbibliography

### 2023-02-01 latex/main.tex
more references

@@
 
-The \textbf{Linked Data} concept \cite{19s3yyxDn} was kickstarted as a set of best practices \cite{OGXYtGZ8} to bring the Web aspect back into focus.
+The \textbf{Linked Data} concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect back into focus.
 Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
@@
 Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph \cite{oN9D5LkD}, RDFa \cite{FKIF6ApB}, Microdata \cite{14Utscfe4}) we find JSON-LD \cite{X5SWPqSU} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 42.7\% of the top 10 million websites \cite{j3SqQLBR}.
+In addition to several such embedded \emph{microformats} (Open Graph \autocite{OpenGraphProtocol}, RDFa \autocite{w3-rdfa-primer}, Microdata \autocite{HTMLStandard}) we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
+JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 44.4\% of the top 10 million websites \autocite{UsageStatisticsJSONLD}.
 
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} \cite{BFMoe8FC} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \cite{GDrm9dgX} or ShEx \cite{YUzogaV}) can be used to validate RDF Data \cite{ZeOAPNHa,S5BI6Mwx} before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \cite{o596f2JM}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
+Recently there has been a renewed emphasis to improve the \emph{Developer Experience} \autocite{DesigningLinkedData2018} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \autocite{w3-shacl} or ShEx \autocite{ShapeExpressionsShEx}) can be used to validate RDF Data \autocite{gayoValidatingRDFData2017a,thorntonUsingShapeExpressions2019a} before consuming it programmatically, or reshaping data to fit other models.
+While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \autocite{klimekSurveyToolsLinked2019a}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
 
-A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \cite{jHwttlTs}.
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \cite{MhsCt4P3}, the use of Linked Data as part of REST Web APIs \cite{eS6a6tlz}, and as shown by the big uptake by publishers to annotate the Web using schema.org \cite{117tX9MEG}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
+A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \autocite{verborghSemanticWebIdentity2020a}.
+One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \autocite{wolstencroftRightFieldEmbeddingOntology2011b}, the use of Linked Data as part of REST Web APIs \autocite{pageRESTLinkedData2011}, and as shown by the big uptake by publishers to annotate the Web using schema.org \autocite{bernsteinNewLookSemantic2016a}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
 
@@
 \item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications \cite{KlCFFFyL}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
+  An Interoperability Framework and Distributed Platform for Fast Data Applications \cite{delgadoInteroperabilityFrameworkDistributed2016a}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
 \item
-  The FAIR Digital Object guidelines \cite{RwvirqWg}, validated against its implementations for completeness.
+  The FAIR Digital Object guidelines \cite{boninoFAIRDigitalObject}, validated against its implementations for completeness.
 \item
-  A Comparison Framework for Middleware Infrastructures \cite{hRzcHhPD}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
+  A Comparison Framework for Middleware Infrastructures \cite{zarrasComparisonFrameworkMiddleware2004a}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
 \item
-  Cross-checks against RDA's FAIR Data Maturity Model \cite{UzQhqk0M} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
+  Cross-checks against RDA's FAIR Data Maturity Model \cite{bahimFAIRDataMaturity2020a} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
 \item
-  EOSC Interoperability Framework \cite{aCye3KpE} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
+  EOSC Interoperability Framework \cite{corchoEOSCInteroperabilityFramework2021b} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
 \end{enumerate}
@@
 Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \cite{13TcbsZF6} against Linked Data in general.
+For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \cite{foundationDigitalObjectInterface} against Linked Data in general.
 
@@
 
-The Interoperability Framework for Fast Data Applications \cite{KlCFFFyL} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
+The Interoperability Framework for Fast Data Applications \cite{delgadoInteroperabilityFrameworkDistributed2016a} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
 
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model \cite{8ZRGFa6y} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
+We have chosen to investigate using this framework as it covers the higher levels of the OSI Model \cite{stallingsHandbookComputercommunicationsStandards1990} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
 In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
@@
 
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services \cite{39K3yHT3}.
-However, this is mostly done via developer or human interaction \cite{iYh1wFBt}.
+The Web has already showed us how one can compose workflows of hetereogeneous Web Services \cite{wolstencroftTavernaWorkflowSuite2013d}.
+However, this is mostly done via developer or human interaction \cite{lamprechtPerspectivesAutomatedComposition2021b}.
 Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
@@
 These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \cite{hxtgCGjY} to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \cite{jrZe6Esu}.
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \cite{p7nwRj8E}.
+The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \cite{ModernStandardsParadigm} to embrace openness, transparency, and broad consensus.
+In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \cite{looFirstInternationalConference2022}.
+This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \cite{DOIPAPIHTTPa}.
 
 It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web \cite{1H9iQhQYq} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \cite{68fzID2k}.
+For example, DOIs are mostly resolved on the web \cite{DOIResolutionDocumentation} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \cite{DOIHandbookResolution}.
 
@@
 
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \cite{8HJqcF1Q}.
+The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \cite{rfc9110}.
 
@@
 
-The FAIR Digital Object guidelines \cite{RwvirqWg} sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \cite{13TcbsZF6} and using Linked Data Platform \cite{17OHlMRQA}, as proposed by \cite{7szz7dwO}.
+The FAIR Digital Object guidelines \cite{boninoFAIRDigitalObject} sets out recommendations for FDO implementations.
+In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \cite{foundationDigitalObjectInterface} and using Linked Data Platform \cite{w3-ldp}, as proposed by \cite{FDOFramework}.
 
@@
 
-Note that the draft update to FDO specification \cite{yygVPoL0} (see box \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer \cite{yygVPoL0} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
+Note that the draft update to FDO specification \cite{fdo-RequirementSpec} (see section \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer \cite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
 
@@
 Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement \cite{15yGpJ0wh} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see box \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
+The DOIP Endorsement \cite{fdo-DOIPEndorsement} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see section \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
 Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
@@
   While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} \cite{9T4j3N4e}
+Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} \cite{w3-prov-o}
 \item
@@
 
-We argue that the Web, with its now ubiquitous use of REST API \cite{174AwcFUL}, can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services \cite{WzPECPDP} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} \cite{11dInoyVF}.
+We argue that the Web, with its now ubiquitous use of REST API \cite{fieldingArchitecturalStylesDesign2000a}, can be compared as a similar global middleware.
+Note that while early moves for developing Semantic Web Services \cite{fenselSemanticWebServices2011} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} \cite{OpenData}.
 
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures \cite{hRzcHhPD} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
+For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures \cite{zarrasComparisonFrameworkMiddleware2004a} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
 

### 2023-02-01 latex/main.tex
all references now updated to biblatex

@@
 \item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications \cite{delgadoInteroperabilityFrameworkDistributed2016a}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
+  An Interoperability Framework and Distributed Platform for Fast Data Applications \autocite{delgadoInteroperabilityFrameworkDistributed2016a}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
 \item
-  The FAIR Digital Object guidelines \cite{boninoFAIRDigitalObject}, validated against its implementations for completeness.
+  The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject}, validated against its implementations for completeness.
 \item
-  A Comparison Framework for Middleware Infrastructures \cite{zarrasComparisonFrameworkMiddleware2004a}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
+  A Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
 \item
-  Cross-checks against RDA's FAIR Data Maturity Model \cite{bahimFAIRDataMaturity2020a} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
+  Cross-checks against RDA's FAIR Data Maturity Model \autocite{bahimFAIRDataMaturity2020a} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
 \item
-  EOSC Interoperability Framework \cite{corchoEOSCInteroperabilityFramework2021b} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
+  EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
 \end{enumerate}
@@
 Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \cite{foundationDigitalObjectInterface} against Linked Data in general.
+For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data in general.
 
@@
 
-The Interoperability Framework for Fast Data Applications \cite{delgadoInteroperabilityFrameworkDistributed2016a} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
+The Interoperability Framework for Fast Data Applications \autocite{delgadoInteroperabilityFrameworkDistributed2016a} categorises interoperability between applications along 6 strands, covering different architectural levels: from \emph{symbiotic} (agreement to cooperate) and \emph{pragmatic} (ability to choreograph processes), through \emph{semantic} (common understanding) and \emph{syntactic} (common message formats), to low-level \emph{connective} (transport-level) and \emph{environmental} (deployment practices).
 
-We have chosen to investigate using this framework as it covers the higher levels of the OSI Model \cite{stallingsHandbookComputercommunicationsStandards1990} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
+We have chosen to investigate using this framework as it covers the higher levels of the OSI Model \autocite{stallingsHandbookComputercommunicationsStandards1990} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
 In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
@@
 
-The Web has already showed us how one can compose workflows of hetereogeneous Web Services \cite{wolstencroftTavernaWorkflowSuite2013d}.
-However, this is mostly done via developer or human interaction \cite{lamprechtPerspectivesAutomatedComposition2021b}.
+The Web has already showed us how one can compose workflows of hetereogeneous Web Services \autocite{wolstencroftTavernaWorkflowSuite2013d}.
+However, this is mostly done via developer or human interaction \autocite{lamprechtPerspectivesAutomatedComposition2021b}.
 Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
@@
 These have shown interoperability.
-The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \cite{ModernStandardsParadigm} to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \cite{looFirstInternationalConference2022}.
-This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \cite{DOIPAPIHTTPa}.
+The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \autocite{ModernStandardsParadigm} to embrace openness, transparency, and broad consensus.
+In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \autocite{looFirstInternationalConference2022}.
+This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \autocite{DOIPAPIHTTPa}.
 
 It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web \cite{DOIResolutionDocumentation} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \cite{DOIHandbookResolution}.
+For example, DOIs are mostly resolved on the web \autocite{DOIResolutionDocumentation} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \autocite{DOIHandbookResolution}.
 
@@
 
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \cite{rfc9110}.
+The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \autocite{rfc9110}.
 
@@
 
-The FAIR Digital Object guidelines \cite{boninoFAIRDigitalObject} sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \cite{foundationDigitalObjectInterface} and using Linked Data Platform \cite{w3-ldp}, as proposed by \cite{FDOFramework}.
+The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject} sets out recommendations for FDO implementations.
+In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \autocite{FDOFramework}.
 
@@
 
-Note that the draft update to FDO specification \cite{fdo-RequirementSpec} (see section \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer \cite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
+Note that the draft update to FDO specification \autocite{fdo-RequirementSpec} (see section \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer \autocite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
 
@@
 Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement \cite{fdo-DOIPEndorsement} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see section \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
+The DOIP Endorsement \autocite{fdo-DOIPEndorsement} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see section \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
 Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
@@
   While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} \cite{w3-prov-o}
+Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} \autocite{w3-prov-o}
 \item
@@
 
-We argue that the Web, with its now ubiquitous use of REST API \cite{fieldingArchitecturalStylesDesign2000a}, can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services \cite{fenselSemanticWebServices2011} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} \cite{OpenData}.
+We argue that the Web, with its now ubiquitous use of REST API \autocite{fieldingArchitecturalStylesDesign2000a}, can be compared as a similar global middleware.
+Note that while early moves for developing Semantic Web Services \autocite{fenselSemanticWebServices2011} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} \autocite{OpenData}.
 
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures \cite{zarrasComparisonFrameworkMiddleware2004a} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
+For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inhexieritance.
 
@@
 \item
-  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP \cite{16uB3jxpa} supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
-Multiplexing was removed for the much simplified DOIP 2.0 \cite{13TcbsZF6}.
+  With respect to the aspect of \emph{Performance}, it is interesting to note that while the first version of DOIP \autocite{DigitalObjectInterface} supported multiplexed channels similar to HTTP/2 (allowing concurrent transfer of several digital objects).
+Multiplexing was removed for the much simplified DOIP 2.0 \autocite{foundationDigitalObjectInterface}.
 Unlike DOIP 1.0, DOIP 2.0 will require a DO response to be sent back completely, as a series of segments (which again can be split the bytes of each binary \emph{element} into sized \emph{chunks}), before transmission of another DO response can start on the transport channel.
@@
 
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \cite{RwvirqWg} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} \cite{PR-RequirementSpec-2.0}.
+In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \autocite{fdo-RequirementSpec} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} .
 
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform \cite{7szz7dwO} comply with the FAIR principles \cite{6DjakjNS}.
-Here we've used the RDA's FAIR Data Maturity Model \cite{1GrKTFaK2} as it has decomposed the FAIR principles to a structured list of FAIR indicators \cite{UzQhqk0M}, importantly considering \emph{Data} and \emph{Metadata} separately.
+Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform \autocite{FDOFramework} comply with the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
+Here we've used the RDA's FAIR Data Maturity Model \autocite{groupFAIRDataMaturity2020} as it has decomposed the FAIR principles to a structured list of FAIR indicators \autocite{bahimFAIRDataMaturity2020a}, importantly considering \emph{Data} and \emph{Metadata} separately.
 In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
@@
 \item
-  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification \cite{13TcbsZF6} has currently only a couple of implementations and is expressed informally.
+  Indicators on standard protocols (RDA-A1-04M, RDA-A1-04D, RDA-A1.1-01M, RDA-A1.1-01D) favour LDP's mature standards (HTTP, URI) -- the DOIPv2 specification \autocite{foundationDigitalObjectInterface} has currently only a couple of implementations and is expressed informally.
 The underlying Handle system for PIDs is arguably mature and commonly used by researchers (this article alone references about 80 DOIs), however DOIs are more commonly accessed as HTTP redirects through resolvers like \url{https://doi.org/} and \url{http://hdl.handle.net/} rather than the Handle protocol.
@@
   Machine-understandable types, provenance and data/metadata standards (RDA-R1.1-03M RDA-R1.3-02M, RDA-R1.3-02M, RDA-R1.3-02D) are important for machine actionability, but are currently unspecified for FDOs.
-\cite{126uxr5pI} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
+\autocite{fdo-ImplAttributesTypesProfiles} explores possible machine-readable FDO types, however the type systems themselves have not yet been formalised.
 Linked Data on the other side have too many semantic and syntactic type systems, making it difficult to write consistent clients.
@@
 For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting \cite{snykkm7R} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
+FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
 In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.
@@
 
-The EOSC Interoperability Framework \cite{aCye3KpE} in section 3.6 recommends:
+The EOSC Interoperability Framework \autocite[][section 3.6]{corchoEOSCInteroperabilityFramework2021b} recommends:
 
@@
 The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \cite{kYjzRJ0H}.
+At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite{SemanticWebXML2000}.
 However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.

### 2023-02-01 manuscript.tex
more references [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/66cf80eb0cdc5217862aa9b9080e957b1073a1f9.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/66cf80eb0cdc5217862aa9b9080e957b1073a1f9/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4067158862

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/5c660e0d407c65b4f669383e77cc59e770e73c12/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/66cf80eb0cdc5217862aa9b9080e957b1073a1f9/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/5c660e0d407c65b4f669383e77cc59e770e73c12}{stain/2022-fdo-paper@5c660e0}
+from \href{https://github.com/stain/2022-fdo-paper/tree/66cf80eb0cdc5217862aa9b9080e957b1073a1f9}{stain/2022-fdo-paper@66cf80e}
 on February 1, 2023.

### 2023-02-01 manuscript.tex
all references now updated to biblatex [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/17d1434fb82116f7f28b84838b370611d75d3569.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/17d1434fb82116f7f28b84838b370611d75d3569/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4067236363

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/66cf80eb0cdc5217862aa9b9080e957b1073a1f9/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/17d1434fb82116f7f28b84838b370611d75d3569/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/66cf80eb0cdc5217862aa9b9080e957b1073a1f9}{stain/2022-fdo-paper@66cf80e}
+from \href{https://github.com/stain/2022-fdo-paper/tree/17d1434fb82116f7f28b84838b370611d75d3569}{stain/2022-fdo-paper@17d1434}
 on February 1, 2023.

### 2023-02-02 latex/main.tex
No ★ in lualatex/styx fonts 😞

@@
 We argue that the Web, with its now ubiquitous use of REST API \autocite{fieldingArchitecturalStylesDesign2000a}, can be compared as a similar global middleware.
-Note that while early moves for developing Semantic Web Services \autocite{fenselSemanticWebServices2011} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of \emph{★★★ Linked Data} \autocite{OpenData}.
+Note that while early moves for developing Semantic Web Services \autocite{fenselSemanticWebServices2011} attempted to merge the Web Service and RDF aspects, we are here considering mainly the current programmatic Web and its mostly light-weight use of 3 out of possible \emph{5 stars Linked Data} \autocite{OpenData}.
 

### 2023-02-02 manuscript.tex
No ★ in lualatex/styx fonts 😞 [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/8f0433b6c923287d3a404c9434df00f9e47b172a.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/8f0433b6c923287d3a404c9434df00f9e47b172a/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4072669009

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/17d1434fb82116f7f28b84838b370611d75d3569/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/8f0433b6c923287d3a404c9434df00f9e47b172a/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/17d1434fb82116f7f28b84838b370611d75d3569}{stain/2022-fdo-paper@17d1434}
-on February 1, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/8f0433b6c923287d3a404c9434df00f9e47b172a}{stain/2022-fdo-paper@8f0433b}
+on February 2, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 1 February 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 2 February 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 1 February 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 2 February 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 1 February 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 2 February 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 1 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 2 February 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 1 February 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 2 February 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 1 February 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 2 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 1 February 2023).
+DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 2 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 1 February 2023).
+DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 2 February 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 1 February 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 2 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 1 February 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 2 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 1 February 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 2 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 1 February 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 2 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 1 February 2023).
+Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 2 February 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 1 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 2 February 2023).
 
@@
 martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 1 February 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 2 February 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 1 February 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 2 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 1 February 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 2 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 1 February 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 2 February 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 1 February 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 2 February 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 1 February 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 2 February 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 1 February 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 2 February 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 1 February 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 2 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 1 February 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 2 February 2023).
 

### 2023-02-02 latex/main.tex
updated FDO specs now mostly published

@@
 
-The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups are preparing more detailed \href{https://docs.google.com/spreadsheets/d/1O1PTLmVWqjQgvJsIcPgaZJe9BntHZt8_QndEG0eRwUY/edit}{requirement documents} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-01-05, these documents are in draft stages in Google Docs for internal review and have not yet formally been made public.
-As these drafts clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
+The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
+As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages in Google Docs for internal review.
+As these documents clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
 

### 2023-02-02 manuscript.tex
citations [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/436f3638b53adb35135c7589db39c8bbbb3d3df9.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/436f3638b53adb35135c7589db39c8bbbb3d3df9/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4073763121

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/8f0433b6c923287d3a404c9434df00f9e47b172a/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/436f3638b53adb35135c7589db39c8bbbb3d3df9/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/8f0433b6c923287d3a404c9434df00f9e47b172a}{stain/2022-fdo-paper@8f0433b}
+from \href{https://github.com/stain/2022-fdo-paper/tree/436f3638b53adb35135c7589db39c8bbbb3d3df9}{stain/2022-fdo-paper@436f363}
 on February 2, 2023.

### 2023-02-02 manuscript.tex
reference roadmap/glossary [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/601c8115352e75e7e5f7ba261f24d27cadfa887a.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/601c8115352e75e7e5f7ba261f24d27cadfa887a/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4073799158

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/436f3638b53adb35135c7589db39c8bbbb3d3df9/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/601c8115352e75e7e5f7ba261f24d27cadfa887a/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/436f3638b53adb35135c7589db39c8bbbb3d3df9}{stain/2022-fdo-paper@436f363}
+from \href{https://github.com/stain/2022-fdo-paper/tree/601c8115352e75e7e5f7ba261f24d27cadfa887a}{stain/2022-fdo-paper@601c811}
 on February 2, 2023.

### 2023-02-02 latex/_site/main.tex

--- 2023-02-02 latex/main.tex
@@
 
-The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
+The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
 As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages in Google Docs for internal review.
-As these documents clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
+As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide their brief summaries below:
 
@@
 It serves as a primer that introduces FDO concepts and the remaining documents.
+It is accompanied by an FDO Glossary \autocite{fdo-Glossary}.
 

### 2023-02-02 latex/_site/table1.tex

--- 2023-01-30 latex/table1.tex
@@
   >{\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data \cite{KlCFFFyL}.
+\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data \cite{delgadoInteroperabilityFrameworkDistributed2016a}.
 \label{tbl:fdo-web-interoperability-framework}}\tabularnewline
@@
 & The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs \cite{174AwcFUL} end up being mandated centrally by dominant (often commercial) providers \cite{11VxBOeZy}, which clients are required to use as-is with special code per service.
+In practice, REST APIs \cite{pubs/dissertation} end up being mandated centrally by dominant (often commercial) providers \cite{fieldingReflectionsRESTArchitectural2017a}, which clients are required to use as-is with special code per service.
 Use of Linked Data enables common tooling and semantic mapping across differences.
@@
 & ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 \cite{k0AfCGzw}, schema.org Actions \cite{dKAekUjL}, WSDL/SOAP \cite{LmudecyN}, but frequently just as human-readable developer documentation/examples.
+Operational composition, typically within a single API provider, documented by OpenAPI 3 \cite{OpenAPISpecificationV3}, schema.org Actions \cite{SchemaOrgActions}, WSDL/SOAP \cite{w3-wsdl20-primer}, but frequently just as human-readable developer documentation/examples.
 \\
@@
 No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) \cite{jRWurmQc} & Lightweight HTTP semantics for authenticity/navigation.
+JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) \cite{wittenburgFAIRDigitalObject2022b} & Lightweight HTTP semantics for authenticity/navigation.
 Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.

### 2023-02-02 latex/_site/table2.tex

--- 2023-01-24 latex/table2.tex
@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
+\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data \cite{KlCFFFyL} to equivalent concepts for FDO and Web.
 \label{tbl:metamodel-concepts}}\tabularnewline

### 2023-02-02 manuscript.tex
biblatex tests [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/1bb192ee3877d8cf686952ca8a79a62e840c6247.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/1bb192ee3877d8cf686952ca8a79a62e840c6247/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4073836159

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/601c8115352e75e7e5f7ba261f24d27cadfa887a/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/1bb192ee3877d8cf686952ca8a79a62e840c6247/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/601c8115352e75e7e5f7ba261f24d27cadfa887a}{stain/2022-fdo-paper@601c811}
+from \href{https://github.com/stain/2022-fdo-paper/tree/1bb192ee3877d8cf686952ca8a79a62e840c6247}{stain/2022-fdo-paper@1bb192e}
 on February 2, 2023.

### 2023-02-02 manuscript.tex
Merge remote-tracking branch 'origin/overleaf-2023-02-02-1106' into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/f187ed4e20c6670aebdc79236fe4e40e7391e19d.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/f187ed4e20c6670aebdc79236fe4e40e7391e19d/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4073850598

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/1bb192ee3877d8cf686952ca8a79a62e840c6247/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/f187ed4e20c6670aebdc79236fe4e40e7391e19d/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/1bb192ee3877d8cf686952ca8a79a62e840c6247}{stain/2022-fdo-paper@1bb192e}
+from \href{https://github.com/stain/2022-fdo-paper/tree/f187ed4e20c6670aebdc79236fe4e40e7391e19d}{stain/2022-fdo-paper@f187ed4}
 on February 2, 2023.

### 2023-02-02 latex/table1.tex
more citations

@@
   >{\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data \cite{KlCFFFyL}.
+\caption{Considering FDO and Web according to the quality levels of the Interoperability Framework for Fast Data \autocite{delgadoInteroperabilityFrameworkDistributed2016a}.
 \label{tbl:fdo-web-interoperability-framework}}\tabularnewline
@@
 & The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs \cite{174AwcFUL} end up being mandated centrally by dominant (often commercial) providers \cite{11VxBOeZy}, which clients are required to use as-is with special code per service.
+In practice, REST APIs \autocite{pubs/dissertation} end up being mandated centrally by dominant (often commercial) providers \autocite{fieldingReflectionsRESTArchitectural2017a}, which clients are required to use as-is with special code per service.
 Use of Linked Data enables common tooling and semantic mapping across differences.
@@
 & ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 \cite{k0AfCGzw}, schema.org Actions \cite{dKAekUjL}, WSDL/SOAP \cite{LmudecyN}, but frequently just as human-readable developer documentation/examples.
+Operational composition, typically within a single API provider, documented by OpenAPI 3 \autocite{OpenAPISpecificationV3}, schema.org Actions \autocite{SchemaOrgActions}, WSDL/SOAP \autocite{w3-wsdl20-primer}, but frequently just as human-readable developer documentation/examples.
 \\
@@
 No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) \cite{jRWurmQc} & Lightweight HTTP semantics for authenticity/navigation.
+JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) \autocite{wittenburgFAIRDigitalObject2022b} & Lightweight HTTP semantics for authenticity/navigation.
 Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
@@
 \\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP \cite{13TcbsZF6} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
+\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP \autocite{foundationDigitalObjectInterface} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
 \textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.

### 2023-02-02 latex/table2.tex
more citations

@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
+\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data \cite{delgadoInteroperabilityFrameworkDistributed2016a} to equivalent concepts for FDO and Web.
 \label{tbl:metamodel-concepts}}\tabularnewline

### 2023-02-02 latex/table3.tex
more citations

@@
+\begin{landscape}
 \begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3676}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1544}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}
-  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1618}}@{}}
-\caption{Checking FDO guidelines \cite{RwvirqWg, yygVPoL0} against its current implementations as DOIP \cite{13TcbsZF6} and Linked Data Platform (LDP) \cite{7szz7dwO}, with suggestions for required additions.
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.25}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.25}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2}}@{}}
+\caption{Checking FDO guidelines \autocite{boninoFAIRDigitalObject,fdo-RequirementSpec} against its current implementations as DOIP \autocite{foundationDigitalObjectInterface} and Linked Data Platform (LDP) \autocite{FDOFramework}, with suggestions for required additions.
 \label{tbl:fdo-checks}}\tabularnewline
@@
 Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites \cite{j3SqQLBR}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
+& JSON-LD used by half of all websites \autocite{UsageStatisticsJSONLD}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
 Example-driven documentation.
@@
 Unsupported methods reported using LDP constraints to human-readable text.
-& Declare supported methods in advance, e.g.~OpenAPI \cite{k0AfCGzw} \\
+& Declare supported methods in advance, e.g.~OpenAPI \autocite{OpenAPISpecificationV3} \\
 G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
@@
 \\
-G9: \emph{standard compliance} & Handle \cite{2gXY2Nwh}, DOIP \cite{13TcbsZF6}.
+G9: \emph{standard compliance} & Handle \autocite{rfc3650}, DOIP \autocite{foundationDigitalObjectInterface}.
 FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements \cite{TQdku4YF} & HTTP, LDP.
+& Formalise standard process of FDO requirements \autocite{fdo-DocProcessStd} & HTTP, LDP.
 FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
 FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve \cite{ljVViWCl}.
+& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve \autocite{fdo-PIDProfileAttributes}.
 Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
@@
 FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns \cite{i9Ang0rM} are used instead.
+& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns \autocite{martinekuanWebAPIDesign} are used instead.
 & Document how operation resources can be discovered from an LPD container.
@@
 FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
-Metadata would need to be bundled within custom FDO types/attributes.
+Metadata would need to be bundled within custom FDO types or attributes.
 Unclear how levels are separated within single FDO representation (need FDOF8?).
@@
 & Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} \cite{CNP39tah} & Rather document better how individual ad-hoc schemas can be started for prototypes.
+& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} \autocite{smithOBOFoundryCoordinated2007a} & Rather document better how individual ad-hoc schemas can be started for prototypes.
 \\
@@
 Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-\cite{6LvydTsC}).
+\autocite{DataInformationView}).
 & & LDP Containers required by specification, also user-created (eg.
 \texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 \cite{1EW4hLN8U}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE \cite{gtGuSevK} \\
+& Clarify relation to other collections like DCAT 3 \autocite{w3-vocab-dcat-3}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE \autocite{gtGuSevK} \\
 FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
@@
 \end{longtable}
+\end{landscape}

### 2023-02-02 latex/main.tex
Updates from Overleaf

@@
 \subsubsection*{Next steps for FDO}\label{sec:next-step-fdo}}
+\begin{refsection}
 
-The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
+The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
 As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages in Google Docs for internal review.
-As these documents clarify the future aims and focus of FAIR Digital Objects, we provide their brief summaries below:
+As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide their brief summaries below:
 
@@
 It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
+
+\defbibheading{fdobibliography}{\subsubsection*{FDO Specifications}}
+\printshorthands[heading=fdobibliography]
+\end{refsection}
 
@@
 \def\UrlFont{\small}
+
 \printbibliography

### 2023-02-02 manuscript.tex
Merge overleaf-2023-02-02-1340 into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/a122b2bad38efe2c132d867ed52b0f97f1ddabca.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/a122b2bad38efe2c132d867ed52b0f97f1ddabca/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4075061948

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/f187ed4e20c6670aebdc79236fe4e40e7391e19d/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/a122b2bad38efe2c132d867ed52b0f97f1ddabca/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/f187ed4e20c6670aebdc79236fe4e40e7391e19d}{stain/2022-fdo-paper@f187ed4}
+from \href{https://github.com/stain/2022-fdo-paper/tree/a122b2bad38efe2c132d867ed52b0f97f1ddabca}{stain/2022-fdo-paper@a122b2b}
 on February 2, 2023.

### 2023-02-02 manuscript.tex
Move FDO spec list to end [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/b789df55763555b00f2320b2c7f1ecd7412e3bc8.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/b789df55763555b00f2320b2c7f1ecd7412e3bc8/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4075295111

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/a122b2bad38efe2c132d867ed52b0f97f1ddabca/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/b789df55763555b00f2320b2c7f1ecd7412e3bc8/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/a122b2bad38efe2c132d867ed52b0f97f1ddabca}{stain/2022-fdo-paper@a122b2b}
+from \href{https://github.com/stain/2022-fdo-paper/tree/b789df55763555b00f2320b2c7f1ecd7412e3bc8}{stain/2022-fdo-paper@b789df5}
 on February 2, 2023.

### 2023-02-02 latex/table2.tex
autocite

--- 2023-01-24 latex/table2.tex
@@
   >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
-\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data (\protect\hyperlink{ref-KlCFFFyL}{Delgado, 2016}) to equivalent concepts for FDO and Web.
+\caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data \autocite{delgadoInteroperabilityFrameworkDistributed2016a} to equivalent concepts for FDO and Web.
 \label{tbl:metamodel-concepts}}\tabularnewline

### 2023-02-02 latex/table5.tex
Cool URIs

@@
+\begin{landscape}
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0625}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0833}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}
-  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2292}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2}}
+  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1}}
+  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2}}
+  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2}}
+  >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2}}@{}}
 \caption{Assessing RDA's FAIR Data Maturity Model \cite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \cite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \cite{13TcbsZF6}, Linked Data Platform (LDP) \cite{7szz7dwO} and examples from Linked Data practices in general.
@@
 & FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
+PID can be any URI & ``Cool'' URIs \autocite{berners-lee-cool-uris}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
 RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.

### 2023-02-02 latex/table3.tex

@@
 \texttt{BasicContainer}).
-& Clarify relation to other collections like DCAT 3 \autocite{w3-vocab-dcat-3}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE \autocite{gtGuSevK} \\
+& Clarify relation to other collections like DCAT 3 \autocite{w3-vocab-dcat-3}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE \autocite{ORESpecificationAbstract} \\
 FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.

### 2023-02-02 latex/table3.tex
Updates from Overleaf

@@
 Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites \autocite{UsageStatisticsJSONLD}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
+& JSON-LD used by half of all websites \autocite{UsageStatisticsJSONLD}, however previous bad experiences with Semantic Web may deter adopters & Ensure simplicity for end developers, rather than semantic overengineering.
 Example-driven documentation.
 \\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
+G3: \emph{follows FAIR principles} & See Table \vref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
+& Touched briefly, see Table \vref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
 \\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
+G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \vref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
 Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
+& HTTP CRUD operations, Open API (see Table \vref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
 \\

### 2023-02-02 latex/table5.tex
Updates from Overleaf

@@
 & FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs \autocite{berners-lee-cool-uris}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
+PID can be any URI & ``Cool'' URIs {[}{`Hypertext Style: Cool URIs don't change.'};\{https://www.w3.org/TR/cooluris/\}{]}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
 RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
@@
 RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
-RDA-A2-01M & Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
+RDA-A2-01M\label{RDA-A2-01M}
+& Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
 RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\

### 2023-02-02 latex/main.tex
Move FDO spec list to end

@@
 }
+\usepackage{varioref}
 
@@
 \usepackage[autostyle]{csquotes}
-\usepackage[backend=biber,style=authoryear]{biblatex}
+\usepackage[backend=biber,alldates=iso,labeldate=year,style=authoryear]{biblatex}
 \addbibresource{bibliography.bib}
@@
 Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}.
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \ref{sec:next-step-fdo}).
+The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \cref{sec:next-step-fdo}).
 
@@
 
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover in the section \ref{sec:ld}.
+From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover \vpageref{sec:ld}.
 
@@
 \subsubsection*{Next steps for FDO}\label{sec:next-step-fdo}}
-\begin{refsection}
 
@@
 This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
-The requirements shown in Table \ref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
+The requirements shown in Table \vref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
 
@@
 It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-
-\defbibheading{fdobibliography}{\subsubsection*{FDO Specifications}}
-\printshorthands[heading=fdobibliography]
-\end{refsection}
 
@@
 The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (see section \ref{sec:next-step-fdo}).
+We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (\vpageref*{sec:next-step-fdo}).
 
@@
 We have chosen to investigate using this framework as it covers the higher levels of the OSI Model \autocite{stallingsHandbookComputercommunicationsStandards1990} better with regards to automated machine-to-machine interaction (and thus interoperability), which is a crucial aspect of the FAIR principles.
-In Table \ref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
+In Table \vref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
 
@@
 
-Based on the analysis shown in table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
+Based on the analysis shown in Table \vref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
 
@@
 
-The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \ref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \autocite{rfc9110}.
+The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \vref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \autocite{rfc9110}.
 
@@
 From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
-Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \ref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
+Notable are that neither DOIP or HTTP have strong support for transactions (explored further \vpageref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
 
@@
 The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject} sets out recommendations for FDO implementations.
-In Table \ref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \autocite{FDOFramework}.
+In Table \vref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \autocite{FDOFramework}.
 
@@
 
-Note that the draft update to FDO specification \autocite{fdo-RequirementSpec} (see section \ref{sec:next-step-fdo}) clarifies these definitions with equivalent identifiers\footnote{Newer \autocite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
+Note that the draft update to FDO specification \autocite{fdo-RequirementSpec} clarifies these definitions with equivalent identifiers\footnote{Newer \autocite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
 
@@
 Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement \autocite{fdo-DOIPEndorsement} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (see section \ref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
+The DOIP Endorsement \autocite{fdo-DOIPEndorsement} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (details \vpageref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
 Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
@@
 \item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in table \ref{sec:fair-compare}).
+  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in Table \vref{RDA-A2-01M}).
 \item
@@
 
-For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inhexieritance.
+For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
 
@@
 
-Based on the analysis in Table \ref{tbl:fdo-web-middleware}, we make the following observations:
+Based on the analysis in Table \vref{tbl:fdo-web-middleware}, we make the following observations:
 
@@
 Here we've used the RDA's FAIR Data Maturity Model \autocite{groupFAIRDataMaturity2020} as it has decomposed the FAIR principles to a structured list of FAIR indicators \autocite{bahimFAIRDataMaturity2020a}, importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \ref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remainining JSON/RDF structures always considered as metadata.
+In our interpretation for Table \vref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remaining JSON/RDF structures always considered as metadata.
 
@@
 
+\defbibheading{fdobibliography}{\subsection*{FDO Specifications}}
+\printshorthands[heading=fdobibliography]
+
 \end{document}

### 2023-02-02 manuscript.tex
Updates from Overleaf [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/7dd90cbf606ec18cbbd096a79baebdf28c4abed6.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/7dd90cbf606ec18cbbd096a79baebdf28c4abed6/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4075917909

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/b789df55763555b00f2320b2c7f1ecd7412e3bc8/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/7dd90cbf606ec18cbbd096a79baebdf28c4abed6/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/b789df55763555b00f2320b2c7f1ecd7412e3bc8}{stain/2022-fdo-paper@b789df5}
+from \href{https://github.com/stain/2022-fdo-paper/tree/7dd90cbf606ec18cbbd096a79baebdf28c4abed6}{stain/2022-fdo-paper@7dd90cb}
 on February 2, 2023.

### 2023-02-02 latex/table4.tex
more citations

@@
+\begin{landscape}
 \begin{longtable}[]{@{}
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1642}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \cite{13TcbsZF6}) and Web technologies (using Linked Data) as middleware infrastructures \cite{hRzcHhPD}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
+\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \autocite{13TcbsZF6}) and Web technologies (using Linked Data) as middleware infrastructures \autocite{hRzcHhPD}
 \label{tbl:fdo-web-middleware}}\tabularnewline
@@
 Setup is typically through TCP and TLS which adds latency.
-& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks \cite{F8VZ86hu}.
+& HTTP traffic is about 10\% of global Internet traffic, excluding video and social networks \autocite{sandvineGlobalInternetPhenomena}.
 HTTP 1 connections are serial and reusable, and concurrent connections is common.
-HTTP/2 adds asynchronous responses and multiplexed streams \cite{KbbW0kGT} but still has TCP+TLS startup costs.
-For reduced latency, HTTP/3 \cite{Amit86Tp} use QUIC \cite{1GWvyEmUf} rather than TCP, already adapted heavily (30\% of EMEA traffic) of which \href{https://engineering.fb.com/2020/10/21/networking-traffic/how-facebook-is-bringing-quic-to-billions/}{Instagram \& Facebook video} is the majority of traffic.
+HTTP/2 adds asynchronous responses and multiplexed streams \autocite{rfc7540} but still has TCP+TLS startup costs.
+For reduced latency, HTTP/3 \autocite{rfc9114} use QUIC \autocite{rfc9000} rather than TCP, already adapted heavily (30\% of EMEA traffic) of which Instagram \& Facebook video is the majority of traffic \autocite{joras2020}.
 \\
@@
 A multitude of schemas and serializations for metadata exists, conventions might be implied by a declared profile or certain media types.
-Metadata is not always machine findable, may need pre-agreed API URI Templates \cite{11hxwwuRt}, content-negotiation \cite{n2f3TfoI} or FAIR Signposting \cite{snykkm7R}.
+Metadata is not always machine findable, may need pre-agreed API URI Templates \autocite{rfc6570}, content-negotiation \autocite{ContentNegotiationHTTP} or FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022}.
 \\
@@
 There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
-& URLs are not required to persist, although encouraged \cite{rbG9uRKw}.
+& URLs are not required to persist, although encouraged \autocite{berners-lee-cool-uris}.
 Persistence requires convention to use PIDs/PURLs and HTTP \texttt{410\ Gone}.
 An URL may change its content, change in type may sometimes force new URLs if exposing extensions like \texttt{.json}.
-Memento \cite{DNjcMi4a} expose versioned snapshots.
-WebDAV \texttt{VERSION-CONTROL} method \cite{hRUsaTiz} (used by SVN).
+Memento \autocite{rfc7089} expose versioned snapshots.
+WebDAV \texttt{VERSION-CONTROL} method \autocite{rfc3253} (used by SVN).
 \\
@@
 & Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
-WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} \cite{xvMAyRAc} with \texttt{LOCK} and \texttt{UNLOCK} methods.
+WebDAV \href{https://datatracker.ietf.org/doc/html/rfc4918\#section-6}{locking mechanisms} \autocite{rfc4918} with \texttt{LOCK} and \texttt{UNLOCK} methods.
 \\
@@
 Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
-\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade \cite{1B16WfR7q} to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
-Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns \cite{iME5l1kk}) and PIDs via Persistent URLs (PURL).
+\texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade \autocite{rfc2817} to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
+Implementation details may leak into URIs (e.g.~\texttt{search.aspx}), countered by deliberate design of URI patterns \autocite{berners-lee-cool-uris}) and PIDs via Persistent URLs (PURL).
 \\
@@
 FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} \cite{15EZ2D0Rm}) & Syntactically Media Type with multiple suffixes \cite{nlH2rm1s} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) \cite{nCzRO0pK}.
-In metadata, semantic type systems (RDFS \cite{ZwAcGQKY}), OWL2 \cite{1p4IWJpI}, SKOS \cite{15gQDya5B}).
-OpenAPI 3 \cite{k0AfCGzw} \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} \cite{11oVTMsuZ}, JSON \texttt{\$schema} \cite{15EZ2D0Rm}), JSON-LD \texttt{@context} \cite{by6CfOmv}.
+Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} \autocite{Draftbhuttonjsonschema00}) & Syntactically Media Type with multiple suffixes \autocite{Draftietfmediamansuffixes00MediaTypes} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) \autocite{rfc6906}.
+In metadata, semantic type systems (RDFS \autocite{w3-rdf-schema}), OWL2 \autocite{w3-owl2-overview}, SKOS \autocite{w3-skos-primer}).
+OpenAPI 3 \autocite{OpenAPISpecificationV3} \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
+XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} \autocite{w3-xmlschema11}, JSON \texttt{\$schema} \autocite{Draftbhuttonjsonschema00}), JSON-LD \texttt{@context} \autocite{w3-json-ld}.
 Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
@@
 \textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} \cite{KbbW0kGT}, Web Sockets \cite{qgqADk45}, Linked Data Notifications \cite{zlGYiuWC}, AtomPub \cite{IbaebabD}, SWORD \cite{7DGllKzG}, Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \cite{HE7Ikwwl}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \cite{11hxwwuRt}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \cite{k0AfCGzw}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \cite{174AwcFUL}, an important element of the REST architectural style.} incl.~Hydra \cite{13PaKBrZe}, schema.org Actions \cite{dKAekUjL}, JSON HAL \cite{nyrPYHoP}) \& Link headers (RFC8288) \cite{ozdWB3O7} \\
+Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} \autocite{rfc7540}, Web Sockets \autocite{WebSocketsStandard}, Linked Data Notifications \autocite{w3-ldn}, AtomPub \autocite{rfc5023}, SWORD \autocite{SWORDSpecification}, Micropub \autocite{w3-micropub}, more typically ad-hoc jobs/futures REST resources \\
+\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \autocite{HE7Ikwwl}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \autocite{rfc6570}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \autocite{OpenAPISpecificationV3}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \autocite{fieldingArchitecturalStylesDesign2000a}, an important element of the REST architectural style.} incl.~Hydra \autocite{HydraW3CCommunity}, schema.org Actions \autocite{SchemaOrgActions}, JSON HAL \autocite{Draftkellyjsonhal08}) \& Link headers (RFC8288) \autocite{rfc8288} \\
 \textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 \cite{tIkSXrb5} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) \cite{PL512B0X}, MPEG-DASH (ISO/IEC 23009-1:2019) \cite{ikO7e0mh} \\
+DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 \autocite{rfc7230} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) \autocite{rfc8216}, MPEG-DASH (ISO/IEC 23009-1:2019) \autocite{iso23009} \\
 \bottomrule
 \end{longtable}
+\end{landscape}

### 2023-02-02 latex/table5.tex
more citations

@@
 & Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-\cite{1Gsq23e51}).
+\cite{NCBOBioPortal}).
 Indexing by search engines if exposing HTML w/schema.org.
@@
 Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP \cite{8HJqcF1Q} & HTTP/HTTPS, FTP (now less common), GridFTP \cite{18LsXvKGJ} (for large data), ARK \cite{voDQ6diM} \\
+RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP \cite{rfc9110} & HTTP/HTTPS, FTP (now less common), GridFTP \cite{allcockGlobusStripedGridFTP} (for large data), ARK \cite{ARKIdentifierScheme} \\
 RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol \cite{I9Kx0Hjy}.
-One server implementation \cite{1CvYRVhy2}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved \cite{xVJMmGVr}  as an open source license -- it includes usage restrictions and requires Service Agreements.
+RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol \cite{rfc3652}.
+One server implementation \cite{HandleNetRegistry}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license} is not OSI-approved \cite{LicensesStandardsOpen}  as an open source license -- it includes usage restrictions and requires Service Agreements.
 It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
 One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
+& \LDP is open W3C recommendation \autocite{w3-ldp}.
 \href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
@@
 & FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \cite{1BlxdxOwa} \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs \cite{o4h92LpS}), \texttt{rdfs:seeAlso}, FAIR signposting \cite{snykkm7R} \texttt{describedby} \\
+RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \cite{iso16684} \\
+RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs \cite{DataW3C}), \texttt{rdfs:seeAlso}, FAIR signposting \cite{vandesompelFAIRSignpostingProfile2022} \texttt{describedby} \\
 RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
@@
 RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles \cite{QpDmajc4}, \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role \cite{10HA0mMBx}, \texttt{prov:specializationOf} \cite{ZUgwvAHE}, OAI-ORE Proxy \cite{ID1pTGry}) \\
+RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles \cite{w3-prov-links}, \href{https://schema.org/Role}{schema.org/Role} \\
+RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role \cite{hollandIntroducingRole2014}, \texttt{prov:specializationOf} \cite{w3-prov-o}, OAI-ORE Proxy \cite{ORESpecificationAbstract}) \\
 RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired \cite{Fi62cJAo} but not assigned PIDs yet.
+Kernel metadata attributes desired \cite{fdo-KernelAttributes} but not assigned PIDs yet.
 & Unspecified.
@@
 & Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org \cite{10pzKCEUE}, Dublin Core Terms \cite{wAbZpOsr}, DCAT \cite{lnH0hEuh}, FOAF \cite{XmUY0cZB} \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata \cite{Fi62cJAo} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \cite{lnH0hEuh} \\
+Rough consensus on kernel metadata: schema.org \cite{10pzKCEUE}, Dublin Core Terms \cite{DCMIMetadataTerms}, DCAT \cite{w3-vocab-dcat-2}, FOAF \cite{FOAFVocabularySpecification} \\
+RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata \cite{fdo-KernelAttributes} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \cite{w3-vocab-dcat-2} \\
 RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\

### 2023-02-02 latex/table5.tex
all the citations. Bioschemas by article.

--- 2023-02-02 latex/table5.tex
@@
   >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model \cite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \cite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \cite{13TcbsZF6}, Linked Data Platform (LDP) \cite{7szz7dwO} and examples from Linked Data practices in general.
+\caption{Assessing RDA's FAIR Data Maturity Model \autocite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \autocite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \autocite{13TcbsZF6}, Linked Data Platform (LDP) \autocite{7szz7dwO} and examples from Linked Data practices in general.
 (--- indicates \emph{Unspecified}, may be possible with additional conventions)
@@
 & Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-\cite{1Gsq23e51}).
+\autocite{NCBOBioPortal}).
 Indexing by search engines if exposing HTML w/schema.org.
@@
 Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP \cite{8HJqcF1Q} & HTTP/HTTPS, FTP (now less common), GridFTP \cite{18LsXvKGJ} (for large data), ARK \cite{voDQ6diM} \\
+RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP \autocite{rfc9110} & HTTP/HTTPS, FTP (now less common), GridFTP \autocite{allcockGlobusStripedGridFTP} (for large data), ARK \autocite{ARKIdentifierScheme} \\
 RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol \cite{I9Kx0Hjy}.
-One server implementation \cite{1CvYRVhy2}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved \cite{xVJMmGVr}  as an open source license -- it includes usage restrictions and requires Service Agreements.
+RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol \autocite{rfc3652}.
+One server implementation \autocite{HandleNetRegistry}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license} is not OSI-approved \autocite{LicensesStandardsOpen}  as an open source license -- it includes usage restrictions and requires Service Agreements.
 It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
 One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).
-& \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation.
+& \LDP is open W3C recommendation \autocite{w3-ldp}.
 \href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
@@
 & FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \cite{1BlxdxOwa} \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs \cite{o4h92LpS}), \texttt{rdfs:seeAlso}, FAIR signposting \cite{snykkm7R} \texttt{describedby} \\
+RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \autocite{iso16684} \\
+RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs \autocite{DataW3C}), \texttt{rdfs:seeAlso}, FAIR signposting \autocite{vandesompelFAIRSignpostingProfile2022} \texttt{describedby} \\
 RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
@@
 RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles \cite{QpDmajc4}, \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role \cite{10HA0mMBx}, \texttt{prov:specializationOf} \cite{ZUgwvAHE}, OAI-ORE Proxy \cite{ID1pTGry}) \\
+RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles \autocite{w3-prov-links}, \href{https://schema.org/Role}{schema.org/Role} \\
+RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role \autocite{hollandIntroducingRole2014}, \texttt{prov:specializationOf} \autocite{w3-prov-o}, OAI-ORE Proxy \autocite{ORESpecificationAbstract}) \\
 RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired \cite{Fi62cJAo} but not assigned PIDs yet.
+Kernel metadata attributes desired \autocite{fdo-KernelAttributes} but not assigned PIDs yet.
 & Unspecified.
@@
 & Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org \cite{10pzKCEUE}, Dublin Core Terms \cite{wAbZpOsr}, DCAT \cite{lnH0hEuh}, FOAF \cite{XmUY0cZB} \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata \cite{Fi62cJAo} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \cite{lnH0hEuh} \\
+Rough consensus on kernel metadata: schema.org \autocite{10pzKCEUE}, Dublin Core Terms \autocite{DCMIMetadataTerms}, DCAT \autocite{w3-vocab-dcat-2}, FOAF \autocite{FOAFVocabularySpecification} \\
+RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata \autocite{fdo-KernelAttributes} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \autocite{w3-vocab-dcat-2} \\
 RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
@@
 Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O \cite{9T4j3N4e}, PAV \cite{1E4OfHyhI}, Dublin Core Terms \cite{hKblceo6} \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen \cite{zZTtJruI}) & --- & Common, e.g.~DCAT 2 \cite{zt8AAjcZ}, BioSchemas \cite{1F8S9FeHf} \\
+RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O \autocite{w3-prov-o}, PAV \autocite{ciccaresePAVOntologyProvenance2013e}, Dublin Core Terms \autocite{DCMIMetadataTerms} \\
+RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen \autocite{hardistySpecimenDataRefinery2022a}) & --- & Common, e.g.~DCAT 2 \autocite{w3-vocab-dcat-2}, BioSchemas \autocite{BioschemasBioschemas} \\
 RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal \cite{1Gsq23e51}, Darwin Core \cite{9jaVzCbL} \\
+RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal \autocite{NCBOBioPortal}, Darwin Core \autocite{wieczorekDarwinCoreEvolving2012} \\
 RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\

### 2023-02-02 manuscript.tex
all the citations. Bioschemas by article. [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/1cbe9b42e4c7f0fecfd786fcff5fbd152fe3abdb.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/1cbe9b42e4c7f0fecfd786fcff5fbd152fe3abdb/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4076291242

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/7dd90cbf606ec18cbbd096a79baebdf28c4abed6/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/1cbe9b42e4c7f0fecfd786fcff5fbd152fe3abdb/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/7dd90cbf606ec18cbbd096a79baebdf28c4abed6}{stain/2022-fdo-paper@7dd90cb}
+from \href{https://github.com/stain/2022-fdo-paper/tree/1cbe9b42e4c7f0fecfd786fcff5fbd152fe3abdb}{stain/2022-fdo-paper@1cbe9b4}
 on February 2, 2023.

### 2023-02-02 manuscript.tex
bioschemas in ceur [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/752f19bbb05f04e1d3a3cb2f18122c2e8ba953c7.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/752f19bbb05f04e1d3a3cb2f18122c2e8ba953c7/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4076315629

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/1cbe9b42e4c7f0fecfd786fcff5fbd152fe3abdb/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/752f19bbb05f04e1d3a3cb2f18122c2e8ba953c7/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/1cbe9b42e4c7f0fecfd786fcff5fbd152fe3abdb}{stain/2022-fdo-paper@1cbe9b4}
+from \href{https://github.com/stain/2022-fdo-paper/tree/752f19bbb05f04e1d3a3cb2f18122c2e8ba953c7}{stain/2022-fdo-paper@752f19b}
 on February 2, 2023.

### 2023-02-02 manuscript.tex
more FDO framework [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/0bcef85d5aab4a6e854dce169024b4d2b2f36598.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/0bcef85d5aab4a6e854dce169024b4d2b2f36598/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4076361303

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/752f19bbb05f04e1d3a3cb2f18122c2e8ba953c7/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/0bcef85d5aab4a6e854dce169024b4d2b2f36598/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/752f19bbb05f04e1d3a3cb2f18122c2e8ba953c7}{stain/2022-fdo-paper@752f19b}
+from \href{https://github.com/stain/2022-fdo-paper/tree/0bcef85d5aab4a6e854dce169024b4d2b2f36598}{stain/2022-fdo-paper@0bcef85}
 on February 2, 2023.

### 2023-02-02 manuscript.tex
more authors, not just editors [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/a11b88f74960da7ec707d4300ddf7812a555c940.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/a11b88f74960da7ec707d4300ddf7812a555c940/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4076408274

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/0bcef85d5aab4a6e854dce169024b4d2b2f36598/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/a11b88f74960da7ec707d4300ddf7812a555c940/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/0bcef85d5aab4a6e854dce169024b4d2b2f36598}{stain/2022-fdo-paper@0bcef85}
+from \href{https://github.com/stain/2022-fdo-paper/tree/a11b88f74960da7ec707d4300ddf7812a555c940}{stain/2022-fdo-paper@a11b88f}
 on February 2, 2023.

### 2023-02-02 latex/main.tex

@@
 Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}.
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see section \cref{sec:next-step-fdo}).
+The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (\vref*{sec:next-step-fdo}).
 
@@
 
+\section*{Method}
+
 \hypertarget{sec:comparing}{%
-\section*{Comparing FDO and existing approaches}\label{sec:comparing}}
+\subsection*{Comparing FDO and existing approaches}\label{sec:comparing}}
 
@@
 For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data in general.
+
+\section*{Methods}
 
@@
 
-%\include{bibliography}
-
-\section*{Methods}
-
-\section*{Results}
-
-\section*{Discussion}
-
-\section*{Conclusions}
-
 \section*{Acknowledgments}
 
-So long and thanks for all the fish.
+This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreement numbers H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreement numbers HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s  \emph{Horizon Europe funding guarantee} grant numbers 10038963 (EuroScienceGateway), 10038930 (BioDT).
 

### 2023-02-02 manuscript.tex
Merge overleaf-2023-02-02-1646 into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/53c6e80932ef4d1f170229e8bff81ba829d2be4d.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/53c6e80932ef4d1f170229e8bff81ba829d2be4d/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4076682981

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/a11b88f74960da7ec707d4300ddf7812a555c940/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/53c6e80932ef4d1f170229e8bff81ba829d2be4d/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/a11b88f74960da7ec707d4300ddf7812a555c940}{stain/2022-fdo-paper@a11b88f}
+from \href{https://github.com/stain/2022-fdo-paper/tree/53c6e80932ef4d1f170229e8bff81ba829d2be4d}{stain/2022-fdo-paper@53c6e80}
 on February 2, 2023.

### 2023-02-03 manuscript.tex
UoM thesis template from https://github.com/Non-Invasive-Bioelectronics-Lab/uom_thesis_template [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/d60e53cb0e45701fd619b51499c45a63f889736e.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/d60e53cb0e45701fd619b51499c45a63f889736e/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4082416357

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/53c6e80932ef4d1f170229e8bff81ba829d2be4d/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/d60e53cb0e45701fd619b51499c45a63f889736e/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/53c6e80932ef4d1f170229e8bff81ba829d2be4d}{stain/2022-fdo-paper@53c6e80}
-on February 2, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/d60e53cb0e45701fd619b51499c45a63f889736e}{stain/2022-fdo-paper@d60e53c}
+on February 3, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 2 February 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 3 February 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 2 February 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 3 February 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 2 February 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 3 February 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 2 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 3 February 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 2 February 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 3 February 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 2 February 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 3 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 2 February 2023).
+DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 3 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 2 February 2023).
+DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 3 February 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 2 February 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 3 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 2 February 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 3 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 2 February 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 3 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 2 February 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 3 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 2 February 2023).
+Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 3 February 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 2 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 3 February 2023).
 
@@
 martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 2 February 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 3 February 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 2 February 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 3 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 2 February 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 3 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 2 February 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 3 February 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 2 February 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 3 February 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 2 February 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 3 February 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 2 February 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 3 February 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 2 February 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 3 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 2 February 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 3 February 2023).
 

### 2023-02-03 manuscript.tex
Revert "UoM thesis template from https://github.com/Non-Invasive-Bioelectronics-Lab/uom_thesis_template" [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/add3947b9250886170d178ca451b7469ea5fd0fe.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/add3947b9250886170d178ca451b7469ea5fd0fe/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4082422816

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/d60e53cb0e45701fd619b51499c45a63f889736e/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/add3947b9250886170d178ca451b7469ea5fd0fe/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/d60e53cb0e45701fd619b51499c45a63f889736e}{stain/2022-fdo-paper@d60e53c}
+from \href{https://github.com/stain/2022-fdo-paper/tree/add3947b9250886170d178ca451b7469ea5fd0fe}{stain/2022-fdo-paper@add3947}
 on February 3, 2023.

### 2023-02-16 latex/main.tex
Updates from Overleaf

@@
 
-\documentclass[fleqn,10pt,lineno]{wlpeerjlua}
+\documentclass[fleqn,10pt,NOlineno]{wlpeerjlua}
 
@@
   \url{https://doi.org/#1}}
+\usepackage{biblatex2bibitem}  
 %\urlstyle{tt}
@@
 
-\hypertarget{sec:introduction}{%
-\section*{Introduction}\label{sec:introduction}}
+\section*{Introduction}\label{sec:introduction}
 
@@
 
-\hypertarget{sec:background}{%
-\section*{Background and related work}\label{sec:background}}
+\section*{Background and related work}\label{sec:background}
 
@@
 
-\hypertarget{sec:fdo}{%
-\subsection*{FAIR Digital Object}\label{sec:fdo}}
+\subsection*{FAIR Digital Object}\label{sec:fdo}
 
@@
 
-\hypertarget{fdo-approaches}{%
-\subsubsection*{FDO approaches}\label{fdo-approaches}}
+\subsubsection*{FDO approaches}\label{sec:fdo-approaches}
 
@@
 
-\hypertarget{sec:next-step-fdo}{%
-\subsubsection*{Next steps for FDO}\label{sec:next-step-fdo}}
+\subsubsection*{Next steps for FDO}\label{sec:next-step-fdo}
 
@@
 
-\hypertarget{sec:ld}{%
-\subsection*{From the Semantic Web to Linked Data}\label{sec:ld}}
+\subsection*{From the Semantic Web to Linked Data}\label{sec:ld}
 
@@
 
-\hypertarget{sec:semweb}{%
-\subsubsection*{A brief history of the Semantic Web}\label{sec:semweb}}
+\subsubsection*{A brief history of the Semantic Web}\label{sec:semweb}
 
@@
 
-\hypertarget{sec:ld-web}{%
-\subsubsection*{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}}
+\subsubsection*{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}
 
@@
 
-\hypertarget{sec:comparing}{%
-\subsection*{Comparing FDO and existing approaches}\label{sec:comparing}}
+\subsection*{Comparing FDO and existing approaches}\label{sec:comparing}
 
@@
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
 \tightlist
@@
 \section*{Methods}
-
-\hypertarget{sec:interoperability-compare}{%
-\subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}}
+\subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}
 
@@
 In Table \vref{tbl:fdo-web-interoperability-framework} we use the interoperability framework to compare the current FAIR Digital Object approach against the Web and its Linked Data practices.
-
-\include{table1}
 
@@
 This is clearly not the case on the Web.
-
 In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed.
@@
 
-\hypertarget{mapping-of-metamodel-concepts}{%
-\subsubsection*{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}}
+\include{table1}
+
+\subsubsection*{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}
 
 The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \vref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \autocite{rfc9110}.
-
-\include{table2}
 
@@
 
-\hypertarget{sec:doip-fdo-compare}{%
-\subsection*{Assessing FDO implementations}\label{sec:doip-fdo-compare}}
+\include{table2}
+
+\subsection*{Assessing FDO implementations}\label{sec:doip-fdo-compare}
 
@@
 In Table \vref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \autocite{FDOFramework}.
-
-\include{table3}
 
@@
 
-\hypertarget{sec:middleware}{%
-\subsection*{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}}
+\include{table3}
+
+\subsection*{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}
 
@@
 For this purpose, we here utillise the Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a} that formalise multiple dimensions of openness, scalability, transparency, as well as characteristics known from Object-oriented programming such as modularity, encapsulation and inheritance.
-
-\include{table4}
 
@@
 
-\hypertarget{sec:fair-compare}{%
-\subsection*{Assessing FDO against FAIR}\label{sec:fair-compare}}
+\include{table4}
 
-In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \autocite{fdo-RequirementSpec} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness.} .
+\subsection*{Assessing FDO against FAIR}\label{sec:fair-compare}
+
+In addition to having ``FAIR'' in its name, the FAIR Digital Object guidelines \autocite{fdo-RequirementSpec} also include \emph{G3: FDOs must offer compliance with the FAIR principles through measurable indicators of FAIRness}.
 
@@
 In our interpretation for Table \vref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remaining JSON/RDF structures always considered as metadata.
-
-\include{table5}
 
@@
 
-\hypertarget{eosc-interoperability-framework}{%
-\section*{EOSC Interoperability Framework}\label{eosc-interoperability-framework}}
+\include{table5}
+
+\section*{EOSC Interoperability Framework}\label{eosc-interoperability-framework}
 
@@
 The EOSC Interoperability Framework \autocite[][section 3.6]{corchoEOSCInteroperabilityFramework2021b} recommends:
-
-\include{table6}
 
@@
 
-\hypertarget{sec:discussion}{%
-\section*{Discussion}\label{sec:discussion}}
+\include{table6}
+
+\section*{Discussion}\label{sec:discussion}
 
@@
 \item
-  Finish tables
-\item
   Bullet points per table
-\item
-  Send in Google Docs
 \end{itemize}
 
-\hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection*{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
+\subsection*{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}
 
@@
 
+% To generate bibitems for manual latex editing, enable
+%\printbibitembibliography
+
 \end{document}

### 2023-02-16 manuscript.tex
Merge overleaf-2023-02-16-1646 into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/ce601f32ec6369a4f226edbda85eddab7b543321.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/ce601f32ec6369a4f226edbda85eddab7b543321/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4196206564

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/add3947b9250886170d178ca451b7469ea5fd0fe/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/ce601f32ec6369a4f226edbda85eddab7b543321/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/add3947b9250886170d178ca451b7469ea5fd0fe}{stain/2022-fdo-paper@add3947}
-on February 3, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/ce601f32ec6369a4f226edbda85eddab7b543321}{stain/2022-fdo-paper@ce601f3}
+on February 16, 2023.
 
@@
 It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\protect\hyperlink{ref-68fzID2k}{{`DOI Handbook - Resolution'}}).
+For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI® Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail .
 
@@
 FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan}) are used instead.
+& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan, 2022}) are used instead.
 & Document how operation resources can be discovered from an LPD container.
@@
 RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol (\protect\hyperlink{ref-I9Kx0Hjy}{Sun et al., 2003}).
-One server implementation (\protect\hyperlink{ref-1CvYRVhy2}{{`Handle.Net Registry'}}), free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved (\protect\hyperlink{ref-xVJMmGVr}{{`Licenses \& Standards \textbar{} Open Source Initiative'}}) as an open source license -- it includes usage restrictions and requires Service Agreements.
+One server implementation (\protect\hyperlink{ref-1CvYRVhy2}{{`Handle.Net Registry'}}), free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved (\protect\hyperlink{ref-xVJMmGVr}{{`Licenses'}, 2022}) as an open source license -- it includes usage restrictions and requires Service Agreements.
 It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
+. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 3 February 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 16 February 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 3 February 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 16 February 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 3 February 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 16 February 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 3 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 16 February 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 3 February 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 16 February 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 3 February 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 16 February 2023).
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI Handbook - Resolution. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}} (accessed 3 February 2023).
-
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
-DOI Resolution Documentation. \emph{Available at} \href{https://www.doi.org/factsheets/DOIProxy.html}{\emph{https://www.doi.org/factsheets/DOIProxy.html}} (accessed 3 February 2023).
+DOI® Resolution Documentation.
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 16 February 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 3 February 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 16 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 3 February 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 16 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-zZTtJruI}{}}%
-Hardisty A, Brack P, Goble C, Livermore L, Scott B, Groom Q, Owen S, Soiland-Reyes S. 2022. The Specimen Data Refinery: A Canonical Workflow Framework and FAIR
-Digital Object Approach to Speeding up Digital Mobilisation of Natural History
-Collections. \emph{Data Intelligence} 4:320--341. DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}.
+Hardisty A, Brack P, Goble C, Livermore L, Scott B, Groom Q, Owen S, Soiland-Reyes S.
+2022.
+The Specimen Data Refinery: A Canonical Workflow Framework and FAIR Digital Object Approach to Speeding up Digital Mobilisation of Natural History Collections.
+\emph{Data Intelligence} 4:320--341.
+DOI: \href{https://doi.org/10.1162/dint_a_00134}{10.1162/dint\_a\_00134}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 3 February 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 16 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 3 February 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 16 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses \& Standards \textbar{} Open Source Initiative. \emph{Available at} \href{https://opensource.org/licenses}{\emph{https://opensource.org/licenses}} (accessed 3 February 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 16 February 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 3 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 16 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-i9Ang0rM}{}}%
-martinekuan.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 3 February 2023).
+martinekuan.
+2022.Web API design best practices - Azure Architecture Center.
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 16 February 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 3 February 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 16 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 3 February 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 16 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 3 February 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 16 February 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 3 February 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 16 February 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 3 February 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 16 February 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 3 February 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 16 February 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 3 February 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 16 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 3 February 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 16 February 2023).
 

### 2023-02-24 manuscript.tex
no build [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/c994f327c59fdf703497a5be0faab917538dd310.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/c994f327c59fdf703497a5be0faab917538dd310/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4264548342

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/ce601f32ec6369a4f226edbda85eddab7b543321/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/c994f327c59fdf703497a5be0faab917538dd310/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/ce601f32ec6369a4f226edbda85eddab7b543321}{stain/2022-fdo-paper@ce601f3}
-on February 16, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/c994f327c59fdf703497a5be0faab917538dd310}{stain/2022-fdo-paper@c994f32}
+on February 24, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 16 February 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 February 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 16 February 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 February 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 16 February 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 February 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 16 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 February 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 16 February 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 February 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 16 February 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 February 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 16 February 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 24 February 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 16 February 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 16 February 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 16 February 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 16 February 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 16 February 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 24 February 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 16 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 February 2023).
 
@@
 2022.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 16 February 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 February 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 16 February 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 February 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 16 February 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 16 February 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 February 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 16 February 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 February 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 16 February 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 February 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 16 February 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 February 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 16 February 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 February 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 16 February 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 February 2023).
 

### 2023-03-21 manuscript.tex
Merge overleaf-2023-03-21-1120 into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/130d4210c6612bf03e9bd5333bdd51e40cf4a852.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/130d4210c6612bf03e9bd5333bdd51e40cf4a852/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4478433440

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/c994f327c59fdf703497a5be0faab917538dd310/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/130d4210c6612bf03e9bd5333bdd51e40cf4a852/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/c994f327c59fdf703497a5be0faab917538dd310}{stain/2022-fdo-paper@c994f32}
-on February 24, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/130d4210c6612bf03e9bd5333bdd51e40cf4a852}{stain/2022-fdo-paper@130d421}
+on March 21, 2023.
 
@@
 It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI® Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail .
+For example, DOIs are mostly resolved on the web (\protect\hyperlink{ref-1H9iQhQYq}{{`DOI® Resolution Documentation'}}) using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail (\protect\hyperlink{ref-68fzID2k}{{`DOI® Handbook'}}).
 
@@
 Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
+\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra , schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
 \textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-. \emph{Available at} \href{https://www.doi.org/doi_handbook/3_Resolution.html}{\emph{https://www.doi.org/doi\_handbook/3\_Resolution.html}}
+\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
+. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}}
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 February 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 21 March 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 February 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 21 March 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 February 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 21 March 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 21 March 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 February 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 21 March 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 February 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 21 March 2023).
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 21 March 2023).
+
 \leavevmode\vadjust pre{\hypertarget{ref-1H9iQhQYq}{}}%
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 24 February 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 21 March 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 February 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 21 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 February 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 21 March 2023).
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 February 2023).
-
 \leavevmode\vadjust pre{\hypertarget{ref-rbG9uRKw}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 February 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 21 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 24 February 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 21 March 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 February 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 21 March 2023).
 
@@
 2022.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 February 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 21 March 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 February 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 21 March 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 February 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 21 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 February 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 21 March 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 February 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 21 March 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 24 February 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 21 March 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 February 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 21 March 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 February 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 21 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 February 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 21 March 2023).
 

### 2023-03-21 latex/main.tex
Updates from Overleaf
Citation fixing in progress

@@
 \usepackage[autostyle]{csquotes}
-\usepackage[backend=biber,alldates=iso,labeldate=year,style=authoryear]{biblatex}
+\usepackage[backend=biber,alldates=iso,labeldate=year,style=authoryear,maxnames=2,maxbibnames=7,maxcitenames=2]{biblatex}
 \addbibresource{bibliography.bib}
@@
 
-Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object \autocite{schultesFAIRPrinciplesDigital2019a} (FDO) concept as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
+Specifically, the EOSC Interoperability framework highlights the emerging FAIR Digital Object (FDO) concept \autocite{schultesFAIRPrinciplesDigital2019a} as a possible foundation for building a semantically interoperable ecosystem to fully realise the FAIR principles beyond individual repositories and infrastructures.
 The FDO approach has great potential, as it proposes strong requirements for identifiers, types, access and formalises interactive operations on objects.
@@
 In other discourse, Linked Data \autocite{bizerLinkedDataStory2009a} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \autocite{boninodasilvasantosFAIRDataPoints2016a,hasnainAssessingFAIRData2018a}.
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \autocite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \autocite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}).
+Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \autocite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \autocite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}.
 While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \autocite{turcoaneLinkedDataJSONLD2014a,wilkinsonWorkflowsWhenParts2022b}.
@@
 This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced in 1995 \autocite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework \autocite{x1255FrameworkDiscovery} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) v1 \autocite{DigitalObjectInterface}.
+DO was the inspiration for the ITU X.1255 framework \autocite{x1255FrameworkDiscovery} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) \autocite{DigitalObjectInterface}.
 
@@
 Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}.
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (\vref*{sec:next-step-fdo}).
+The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see \emph{Next steps for FDO} \vpageref{sec:next-step-fdo}).
 
@@
 These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
-FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \autocite{Draftbhuttonjsonschema00} and PIDs are assigned using the Handle system.
+FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \autocite{Draftbhuttonjsonschema} and PIDs are assigned using the Handle system.
 Several type registries have been established.

### 2023-03-21 latex/table5.tex
Citation fixing in progress

@@
 RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol \autocite{rfc3652}.
-One server implementation \autocite{HandleNetRegistry}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license} is not OSI-approved \autocite{LicensesStandardsOpen}  as an open source license -- it includes usage restrictions and requires Service Agreements.
+RDA-A1.1-01M & Metadata is accessible through a free access protocol 
+    & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} 
+        system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.}
+    protocol \autocite{rfc3652}. 
+    One server implementation \autocite{HandleNetRegistry}, free\footnote{The 
+        \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license} is not OSI-approved \autocite{LicensesStandardsOpen}  as an open source license -- it includes usage restrictions and requires Service Agreements.
 It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
     One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source). 
-& \LDP is open W3C recommendation \autocite{w3-ldp}.
-\href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.
-& DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}.
-\\
+    & LDP is open W3C recommendation \autocite{w3-ldp}. \href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}. 
+    & DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}. \\
 RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\

### 2023-03-21 manuscript.tex
Citation fixing in progress [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/9f1d51557301a61b06b5b4dd934b2b478512567f.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/9f1d51557301a61b06b5b4dd934b2b478512567f/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4479098000

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/130d4210c6612bf03e9bd5333bdd51e40cf4a852/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/9f1d51557301a61b06b5b4dd934b2b478512567f/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/130d4210c6612bf03e9bd5333bdd51e40cf4a852}{stain/2022-fdo-paper@130d421}
+from \href{https://github.com/stain/2022-fdo-paper/tree/9f1d51557301a61b06b5b4dd934b2b478512567f}{stain/2022-fdo-paper@9f1d515}
 on March 21, 2023.

### 2023-03-21 latex/main.tex
more citations

@@
 The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite{SemanticWebXML2000}.
+At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite[][slide 10]{SemanticWebXML2000}.
 However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.

### 2023-03-21 latex/table5.tex
more citations

@@
 RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O \autocite{w3-prov-o}, PAV \autocite{ciccaresePAVOntologyProvenance2013e}, Dublin Core Terms \autocite{DCMIMetadataTerms} \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen \autocite{hardistySpecimenDataRefinery2022a}) & --- & Common, e.g.~DCAT 2 \autocite{w3-vocab-dcat-2}, BioSchemas \autocite{BioschemasBioschemas} \\
+RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen \autocite{hardistySpecimenDataRefinery2022a}) & --- & Common, e.g.~DCAT 2 \autocite{w3-vocab-dcat-2}, BioSchemas \autocite{Bioschemas} \\
 RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\

### 2023-03-21 manuscript.tex
more citations [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/c84452f9fd52ac995ffa76f1c51091daa85f9396.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/c84452f9fd52ac995ffa76f1c51091daa85f9396/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4479367514

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/9f1d51557301a61b06b5b4dd934b2b478512567f/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/c84452f9fd52ac995ffa76f1c51091daa85f9396/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/9f1d51557301a61b06b5b4dd934b2b478512567f}{stain/2022-fdo-paper@9f1d515}
+from \href{https://github.com/stain/2022-fdo-paper/tree/c84452f9fd52ac995ffa76f1c51091daa85f9396}{stain/2022-fdo-paper@c84452f}
 on March 21, 2023.
@@
 Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} (\protect\hyperlink{ref-KbbW0kGT}{Belshe \& Peon, 2015}), Web Sockets (\protect\hyperlink{ref-qgqADk45}{{`WebSockets Standard'}}), Linked Data Notifications (\protect\hyperlink{ref-zlGYiuWC}{{`Linked Data Notifications'}}), AtomPub (\protect\hyperlink{ref-IbaebabD}{Gregorio \& de hOra, 2007}), SWORD (\protect\hyperlink{ref-7DGllKzG}{{`SWORD 3.0 Specification'}}), Micropub, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra , schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
+\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} (\protect\hyperlink{ref-HE7Ikwwl}{Fielding \& Reschke, 2014a}), (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates (\protect\hyperlink{ref-11hxwwuRt}{Gregorio et al., 2012}), \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} (\protect\hyperlink{ref-k0AfCGzw}{{`OpenAPI Specification v3.1.0 \textbar{} Introduction, Definitions, \& More'}}), HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State (\protect\hyperlink{ref-174AwcFUL}{Fielding, 2000}), an important element of the REST architectural style.} incl.~Hydra (\protect\hyperlink{ref-13PaKBrZe}{{`Hydra W3C Community Group'}}), schema.org Actions {[}{`Schema.org Actions - schema.org'}), JSON HAL (\protect\hyperlink{ref-nyrPYHoP}{{`draft-kelly-json-hal-08'}}) \& Link headers (RFC8288) (\protect\hyperlink{ref-ozdWB3O7}{Nottingham, 2017}) \\
 \textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}}
-
 \leavevmode\vadjust pre{\hypertarget{ref-18P9txgeB}{}}%
@@
 DOI: \href{https://doi.org/10.1007/978-3-642-21034-1_20}{10.1007/978-3-642-21034-1\_20}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 21 March 2023).
 

### 2023-03-21 latex/table1.tex
table magic

@@
 \emph{Quality}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 FDO w/ DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 Web w/ Linked Data
@@
 \endhead
-\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact, align, collaborate or cooperate} & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
+\textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact, align, collaborate or cooperate}
+  & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
 FDO encourages research communities to develop common type registries to be shared across instances.
@@
   & The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs \autocite{pubs/dissertation} end up being mandated centrally by dominant (often commercial) providers \autocite{fieldingReflectionsRESTArchitectural2017a}, which clients are required to use as-is with special code per service.
+In practice, REST APIs \autocite{fieldingArchitecturalStylesDesign2000a} end up being mandated centrally by dominant (often commercial) providers \autocite{fieldingReflectionsRESTArchitectural2017a}, which clients are required to use as-is with special code per service.
 Use of Linked Data enables common tooling and semantic mapping across differences.
 \\
-\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows} & FDO types and operations enable detailed choreography (see CWFP).
+\textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows}
+  & FDO types and operations enable detailed choreography (see CWFP).
 \texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).
@@
 \\
-\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies} & FDO semantic enable navigation and typing.
+\textbf{Semantic}: \emph{ensuring consistent understanding of messages, interoperability of rules, knowledge and ontologies}
+  & FDO semantic enable navigation and typing.
 Every FDO has a type.
@@
 No single type semantic, Type FDOs can link to existing vocabularies \& ontologies.
-JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) \autocite{wittenburgFAIRDigitalObject2022b} & Lightweight HTTP semantics for authenticity/navigation.
+JSON-LD used within some FDO objects (e.g.~DISSCO Digital Specimen, NIST Material Science schema) \autocite{wittenburgFAIRDigitalObject2022b}
+  & Lightweight HTTP semantics for authenticity/navigation.
 Semantic Type not commonly expressed on PID/header level, may be declared within Linked Data metadata.
@@
 \\
-\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation} & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema.
-Multiple byte stream attachments of any media type.
+\textbf{Syntactic}: \emph{serialising messages for digital exchange, structure representation}
+  & DOIP serialise FDOs as JSON, metadata commonly use JSON, typed with JSON Schema. Multiple byte stream attachments of any media type.
   & Textual HTTP headers (including any signposting), single byte stream of any media type, e.g.~Linked Data formats (JSON-LD, Turtle, RDF/XML) or embedded in document (HTML with RDFa, JSON-LD or Microdata).
@@
 \\
-\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols} & DOIP \autocite{foundationDigitalObjectInterface} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
-\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability} & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
+\textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols}
+  & DOIP \autocite{foundationDigitalObjectInterface} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP}
+  & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
+\textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability}
+  & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
 Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable).

### 2023-03-21 latex/table3.tex
table magic

--- 2023-02-02 latex/table3.tex
@@
 \textbf{FDO Guideline}
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 DOIP 2.0
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 FDO suggestions
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 Linked Data Platform
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 LDP suggestion
@@
 \endhead
-G1: \emph{invest for many decades} & Handle system stable for 20 years, DOIP 2.0 since 2017.
-& Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced & HTTP stable for 30 years, Semantic Web for 20 years.
-\texttt{http://} URIs replaced by \texttt{https://}.
+G1: \emph{invest for many decades}
+  & Handle system stable for 20 years, DOIP 2.0 since 2017.
+  & Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced
+  & HTTP stable for 30 years, Semantic Web for 20 years. \texttt{http://} URIs replaced by \texttt{https://}.
   & Keep flexibility of RDF serialisation formats which may change more frequently \\
-G2: \emph{trustworthiness} & DOI/Handle trusted by all major academic publishers and data repositories.
-DOIP relatively unknown, in effect only one implementation.
-& Further promote DOIP and justify its benefits.
-Build tutorials and OSI open source implementations.
-Standardise DOIP-over-HTTP alternative.
-& JSON-LD used by half of all websites \autocite{UsageStatisticsJSONLD}, however previous bad experiences with Semantic Web may deter adapters & Ensure simplicity for end developers, rather than semantic overengineering.
-Example-driven documentation.
-\\
-G3: \emph{follows FAIR principles} & See Table \ref{tbl:fair-data-maturity-model} & Ensure all FAIR principles are covered, build complete examples.
-& Touched briefly, see Table \ref{tbl:fair-data-maturity-model} & Add explicit expression to show each FAIR pcinciple fulfilled.
-\\
-G4: \emph{machine actionability} & CRUD and extension operations dynamically listed (see Table \ref{tbl:fdo-web-middleware}) & Specify which operations should work for a given type, to reduce need for dynamic lookup.
-Specify input/output expectations formally (e.g.~JSON Schema).
-& HTTP CRUD operations, Open API (see Table \ref{tbl:fdo-web-middleware}) & Document operations so client can make subsequent HTTP calls.
-\\
-G5: \emph{abstraction principle} & Handle PIDs as abstraction base.
-DOIP operations can use any transport protocol.
+G2: \emph{trustworthiness}
+  & DOI/Handle trusted by all major academic publishers and data repositories. DOIP relatively unknown, in effect only one implementation.
+  & Further promote DOIP and justify its benefits. Build tutorials and OSI open source implementations. Standardise DOIP-over-HTTP alternative.
+  & JSON-LD used by half of all websites \autocite{UsageStatisticsJSONLD}, however previous bad experiences with Semantic Web may deter adopters
+  & Ensure simplicity for end developers, rather than semantic overengineering. Example-driven documentation. \\
+G3: \emph{follows FAIR principles}
+  & See Table \vref{tbl:fair-data-maturity-model}
+  & Ensure all FAIR principles are covered, build complete examples.
+  & Touched briefly, see Table \vref{tbl:fair-data-maturity-model}
+  & Add explicit expression to show each FAIR pcinciple fulfilled. \\
+G4: \emph{machine actionability}
+  & CRUD and extension operations dynamically listed (see Table \vref{tbl:fdo-web-middleware})
+  & Specify which operations should work for a given type, to reduce need for dynamic lookup. Specify input/output expectations formally (e.g.~JSON Schema).
+  & HTTP CRUD operations, Open API (see Table \vref{tbl:fdo-web-middleware})
+  & Document operations so client can make subsequent HTTP calls. \\
+G5: \emph{abstraction principle}
+  & Handle PIDs as abstraction base. DOIP operations can use any transport protocol.
   & Document transport protocols as FDOs, recommend which transport to use.
-& URI as abstraction base.
-Does not specify PID requirements.
-& Give stronger deployment recommendations.
-\\
-G6: \emph{stable binding between entities} & Machine-navigation through PIDs and operations specified per type.
-Unclear when metadata field is a PID or plain text.
+  & URI as abstraction base. Does not specify PID requirements.
+  & Give stronger deployment recommendations. \\
+G6: \emph{stable binding between entities}
+  & Machine-navigation through PIDs and operations specified per type. Unclear when metadata field is a PID or plain text.
   & Make datatype of fields explicit to support navigation.
-& Machine-navigation through URIs via properties and types.
-Unclear when URI should be followed or is just identifier, but always distinct from plain text.
+  & Machine-navigation through URIs via properties and types. Unclear when URI should be followed or is just identifier, but always distinct from plain text.
   & \\
-G7: \emph{encapsulation} & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
+G7: \emph{encapsulation}
+  & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
   & Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
-& HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly.
-Unsupported methods reported using LDP constraints to human-readable text.
+  & HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly. Unsupported methods reported using LDP constraints to human-readable text.
   & Declare supported methods in advance, e.g.~OpenAPI \autocite{OpenAPISpecificationV3} \\
-G8: \emph{technology independence} & In theory independent, in reality depends on single implementations of Handle system and DOIP & Encourage open source DOIP testbeds and lighter reference implementations & Multiple HTTP implementations, multiple LDP implementations.
-No FDOF implementations.
-& Develop demonstrator of FDOF usage based on existing LDP server.
-\\
-G9: \emph{standard compliance} & Handle \autocite{rfc3650}, DOIP \autocite{foundationDigitalObjectInterface}.
-FDO requirements not standardised yet.
-& Formalise standard process of FDO requirements \autocite{fdo-DocProcessStd} & HTTP, LDP.
-FDOF not yet standardised & Formalise FDOF from FDOF-SEM working group \\
-FDOF1: \emph{PID as basis} & Extensive use of Handle system.
-& Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve \autocite{fdo-PIDProfileAttributes}.
-Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
+G8: \emph{technology independence}
+  & In theory independent, in reality depends on single implementations of Handle system and DOIP
+  & Encourage open source DOIP testbeds and lighter reference implementations
+  & Multiple HTTP implementations, multiple LDP implementations. No FDOF implementations.
+  & Develop demonstrator of FDOF usage based on existing LDP server. \\
+G9: \emph{standard compliance}
+  & Handle \autocite{rfc3650}, DOIP \autocite{foundationDigitalObjectInterface}. FDO requirements not standardised yet.
+  & Formalise standard process of FDO requirements \autocite{fdo-DocProcessStd}
+  & HTTP, LDP. FDOF not yet standardised
+  & Formalise FDOF from FDOF-SEM working group \\
+FDOF1: \emph{PID as basis}
+  & Extensive use of Handle system.
+  & Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve \autocite{fdo-PIDProfileAttributes}. Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
   & HTTP URLs as basis for identifiers, but they are frequently not persistent.
-& Add strong guidance for PID services like w3id and persistence policies.
-\\
-FDOF2: \emph{PID record w/ type} & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA} & Document requirements for PID Record & w3id/purl PIDs redirect without giving any metadata.
-Datacite DOIs content-negotiate to give registered metadata.
+  & Add strong guidance for PID services like w3id and persistence policies. \\
+FDOF2: \emph{PID record w/ type}
+  & Unspecified how to resolve from Handle to DOIP Service (!), in practice \texttt{10320/loc}, \texttt{0.TYPE/DOIPService}, \texttt{URL}, \texttt{URL\_REPLICA}
+  & Document requirements for PID Record
+  & w3id/purl PIDs redirect without giving any metadata. Datacite DOIs content-negotiate to give registered metadata.
   & Add FAIR Signposting at PID provider for minimal PID record \\
-FDOF3: \emph{PID resolvable to bytestream \& metadata} & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
+FDOF3: \emph{PID resolvable to bytestream \& metadata}
+  & Byte stream resolvable (\texttt{0.DOIP/Retrieve}), \texttt{includeElementData} option can retrieve bytestream or full object structure.
 No method/attribute defined for separate metadata, only directly in PID Record.
 Unclear meaning of multiple items and bytestream chunks.
-& Clarify expectations for multiple items.
-Recommend chunks to not be used.
-& URIs resolvable by default.
-Multiple ways to resolve metadata, unclear preference.
-& Add FAIR Signposting and preference order.
-\\
-FDOF4: \emph{Additional attributes} & Freetext attribute keys.
-Attributes should be defined for FDO type (?).
-& Require that attribute keys should be PIDs (or have predefined mapping to PIDs).
-Explicitly allow attributes not already defined in type.
-& All attributes individually identified.
-Any Linked Data attributes can be used by URI or with mapped prefix.
-& Clarify type expectations of required/recommended/optional attributes.
-\\
-FDOF5: \emph{Interface: operation by PID} & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
-& Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
-Clarify that operations can be mapped to protocol directly.
-& CRUD operations used directly in HTTP (e.g.~\texttt{PUT}).
-Unclear how to provide PID for additional operations.
-& Specify how additional operations should be called over HTTP.
-\\
-FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns \autocite{martinekuanWebAPIDesign} are used instead.
-& Document how operation resources can be discovered from an LPD container.
-Document search API.
-\\
-FDOF7: \emph{FDOF Types related to operations} & Not yet formalised, by DOIP discoverable on a given FDO rather than type.
-PR-TypingFDOs leaves this open.
-& Add explicit relation between type and operations & \texttt{OPTIONS} per LDP Resource, but not by type.
-Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
-& Always make LDP types explicit in FDO profile.
-\\
-FDOF8: \emph{Metadata as FDO, semantic assertions} & DOIP includes all metadata in PID Record.
-Separate Metadata FDO need custom property.
+  & Clarify expectations for multiple items. Recommend chunks to not be used.
+  & URIs resolvable by default. Multiple ways to resolve metadata, unclear preference.
+  & Add FAIR Signposting and preference order. \\
+FDOF4: \emph{Additional attributes}
+  & Freetext attribute keys. Attributes should be defined for FDO type (?).
+  & Require that attribute keys should be PIDs (or have predefined mapping to PIDs). Explicitly allow attributes not already defined in type.
+  & All attributes individually identified. Any Linked Data attributes can be used by URI or with mapped prefix.
+  & Clarify type expectations of required/recommended/optional attributes. \\
+FDOF5: \emph{Interface: operation by PID}
+  & Extended operations use PID, but ``pid-like'' DOIP operations/types are not registered as handles.
+  & Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs. Clarify that operations can be mapped to protocol directly.
+  & CRUD operations used directly in HTTP (e.g.~\texttt{PUT}). Unclear how to provide PID for additional operations.
+  & Specify how additional operations should be called over HTTP. \\
+FDOF6: \emph{CRUD operations + extensions}
+  & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
+  & Document
+  & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns \autocite{martinekuanWebAPIDesign} are used instead.
+  & Document how operation resources can be discovered from an LPD container. Document search API. \\
+FDOF7: \emph{FDOF Types related to operations}
+  & Not yet formalised, by DOIP discoverable on a given FDO rather than type. PR-TypingFDOs leaves this open.
+  & Add explicit relation between type and operations
+  & \texttt{OPTIONS} per LDP Resource, but not by type. Common types (\texttt{ldp:Resource}, \texttt{ldp:Container}) indicate LDP support, but are not required.
+  & Always make LDP types explicit in FDO profile. \\
+FDOF8: \emph{Metadata as FDO, semantic assertions}
+  & DOIP includes all metadata in PID Record. Separate Metadata FDO need custom property.
   & Specify a \texttt{0.FDO/metadata} or similar to point to Metadata FDOs.
-& Assertions are always with semantics, using RDF vocabularies.
-Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
+  & Assertions are always with semantics, using RDF vocabularies. Unspecified how to find additional metadata resources, \texttt{rdfs:seeAlso} is common.
   & Use FAIR Signposting \texttt{describedby} link relation to additional metadata PIDs \\
-FDOF9: \emph{Different metadata levels} & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
+FDOF9: \emph{Different metadata levels}
+  & Defines open-ended ``Response Attributes'' without namespaces, but mandated as ``None'' for all CRUD operations.
 Metadata would need to be bundled within custom FDO types or attributes.
 Unclear how levels are separated within single FDO representation (need FDOF8?).
-& Declare which metadata are expected within response attribute or within FDO object.
-Require PIDs for custom attributes.
-Define how alternate metadata levels can be represented separately.
+  & Declare which metadata are expected within response attribute or within FDO object. Require PIDs for custom attributes. Define how alternate metadata levels can be represented separately.
   & Undefined how to handle multiple metadata granularities or domains, alternative LDP containers can present different views on same stored objects.
   & Define how to navigate to alternate views and their semantic implications, e.g.~\texttt{owl:sameAs} \\
-FDOF10: \emph{Metadata schemas by community} & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
+FDOF10: \emph{Metadata schemas by community}
+  & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
   & Require types to be FDOs with registered PIDs, implement shared types.
-& Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} \autocite{smithOBOFoundryCoordinated2007a} & Rather document better how individual ad-hoc schemas can be started for prototypes.
-\\
-FDOF11: \emph{FDO collections w/ semantic relations} & Collection type undefined by DOIP.
-Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g.
-\autocite{DataInformationView}).
-& & LDP Containers required by specification, also user-created (eg.
-\texttt{BasicContainer}).
+  & Plethora of existing RDF vocabularies/ontologies managed by larger communities, e.g.~\href{https://obofoundry.org/}{OBO Foundry} \autocite{smithOBOFoundryCoordinated2007a}
+  & Rather document better how individual ad-hoc schemas can be started for prototypes. \\
+FDOF11: \emph{FDO collections w/ semantic relations}
+  & Collection type undefined by DOIP. Informal use of \texttt{HAS\_PARTS} Handle attribute (e.g. \autocite{DataInformationView}).
+  &
+  & LDP Containers required by specification, also user-created (eg. \texttt{BasicContainer}).
   & Clarify relation to other collections like DCAT 3 \autocite{w3-vocab-dcat-3}, \href{https://schema.org/Dataset}{Schema.org Dataset}, OAI-ORE \autocite{ORESpecificationAbstract} \\
-FDOF12: \emph{Deleted FDO preserve PID w/ tombstone} & Tombstone for deleted resource undefined by DOIP.
-\texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone'' & Formalise tombstone requirements with new FDO type & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common.
-No requirement for tombstone serialisation & Formalise tombstone requirements and serialisation \\
+FDOF12: \emph{Deleted FDO preserve PID w/ tombstone}
+  & Tombstone for deleted resource undefined by DOIP. \texttt{0.DOIP/Status.104} status code does not distinguish ``Not Found'' or ``Gone''
+  & Formalise tombstone requirements with new FDO type
+  & \texttt{410\ Gone} recommended, but \texttt{404\ Not\ Found} common. No requirement for tombstone serialisation
+  & Formalise tombstone requirements and serialisation \\
 \bottomrule

### 2023-03-21 latex/table4.tex
table magic

@@
 \endhead
-\textbf{Openness}: \emph{framework enable extension of applications} & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
+\textbf{Openness}: \emph{framework enable extension of applications}
+  & FDOs can be cross-linked using PIDs, pointing to multiple FDO endpoints.
 Custom DOIP operations can be exposed, although it is unclear if these can be outside the FDO server.
@@
 \\
-\textbf{Scalability}: \emph{application should be effective at many different scales} & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes) & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
+\textbf{Scalability}: \emph{application should be effective at many different scales}
+  & No defined methods for caching or mirroring, although this could be handled by backend, depending on exposed FDO operations (e.g.~Cordra can scale to multiple backend nodes)
+  & Cache control headers reduce repeated transfer and assist explicit and transparent proxies for speed-up.
 HTTP \texttt{GET} can be scaled to world-population-wide with Content-Delivery Networks (CDNs), while write-access scalability is typically manage by backend.
 \\
-\textbf{Performance}: \emph{efficient and predictable execution} & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
+\textbf{Performance}: \emph{efficient and predictable execution}
+  & DOIP has been shown moderately scalable to 100 millions of objects, create operation at 900 requests/second .
 DOIP protocol is reusable for many operations, multiple requests may be answered out of order (by \texttt{requestId}).
@@
 \\
-\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.} & Each FDO is accessed separately along with its components (typically from the same endpoint).
+\textbf{Distribution transparency}: \emph{application perceived as a consistent whole rather than independent elements.}
+  & Each FDO is accessed separately along with its components (typically from the same endpoint).
 FDOs should provide the mandatory kernel metadata fields.
@@
 \\
-\textbf{Access transparency}: \emph{local/remote elements accessed similarly} & FDOs should be accessed through PID indirection, this means difficult to make private test setup.
-Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
+\textbf{Access transparency}: \emph{local/remote elements accessed similarly}
+  & FDOs should be accessed through PID indirection, this means difficult to make private test setup. Commonly a fixed DOIP server is used directly, which permits local non-PID identifiers.
   & Global HTTP protocol frequently used locally and behind firewalls, but at risk of non-global URIs (e.g.~\texttt{http://localhost/object/1}) and SSL issues (e.g.~self-signed certificates, local CAs) \\
-\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location} & FDOs always accessed through PIDs.
-Multiple locations possible in Handle system, can expose geo-info.
-& PIDs and URL redirects.
-DNS aliases and IP routing can hide location.
-Geo-localised servers common for large cloud deployments.
-\\
-\textbf{Concurrency transparency}: \emph{concurrent processing without interference} & No explicit concurrency measures.
-FDO kernel metadata can include checksum and date.
+\textbf{Location transparency}: \emph{elements accessed without knowledge of physical location}
+  & FDOs always accessed through PIDs. Multiple locations possible in Handle system, can expose geo-info.
+  & PIDs and URL redirects. DNS aliases and IP routing can hide location. Geo-localised servers common for large cloud deployments. \\
+\textbf{Concurrency transparency}: \emph{concurrent processing without interference}
+  & No explicit concurrency measures. FDO kernel metadata can include checksum and date.
   & HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
@@
 \\
-\textbf{Failure transparency}: \emph{service provisioning resilient to failures} & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
+\textbf{Failure transparency}: \emph{service provisioning resilient to failures}
+  & DOIP status codes, e.g.~\texttt{0.DOIP/Status.104}, additional codes can be added as custom attributes
+  & HTTP \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-6.5}{status codes} e.g.~\texttt{404\ Not\ Found}, specific meaning of standard codes can be \href{https://swagger.io/docs/specification/describing-responses/}{documented in Open API}.
 Custom codes uncommon.
 \\
-\textbf{Migration transparency}: \emph{allow relocating elements without interfering application} & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently).
-No redirection from DOIP service.
-& HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints.
-\\
-\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users} & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
+\textbf{Migration transparency}: \emph{allow relocating elements without interfering application}
+  & Update of PID record URLs, indirection through \texttt{0.TYPE/DOIPServiceInfo} (not always used consistently). No redirection from DOIP service.
+  & HTTP \texttt{30x} status codes provide temporary or permanent redirections, commonly used for PURLs but also by endpoints. \\
+\textbf{Persistence transparency}: \emph{conceal deactivation/reactivation of elements from their users}
+  & FDO requires use of PIDs for object persistence, including a tombstone response for deleted objects.
 There is no guarantee that an FDO is immutable or will even stay the same type (note: CORDRA extends DOIP with \href{https://www.cordra.org/documentation/design/object-versioning.html}{version tracking}).
@@
 \\
-\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions} & No transaction capabilities declared by FDO or DOIP.
-Internal synchronisation possible in backend for Extended operations.
+\textbf{Transaction transparency}: \emph{coordinate execution of atomic/isolated transactions}
+  & No transaction capabilities declared by FDO or DOIP. Internal synchronisation possible in backend for Extended operations.
   & Limited transaction capabilities (e.g.~\texttt{If-Unmodified-Since}) on same resource.
@@
 \\
-\textbf{Modularity}: \emph{application as collection of connected/distributed elements} & FDOs are inheritedly modular using global PID spaces and their cross-references.
+\textbf{Modularity}: \emph{application as collection of connected/distributed elements}
+  & FDOs are inheritedly modular using global PID spaces and their cross-references.
 In practice, FDOs of a given type are exposed through a single server shared within a particular community/institution.
@@
 \\
-\textbf{Encapsulation}: \emph{separate interface from implementation.
-Specify interface as contract, multiple implementations possible} & Indirection by PID gives separation.
+\textbf{Encapsulation}: \emph{separate interface from implementation. Specify interface as contract, multiple implementations possible}
+  & Indirection by PID gives separation.
 FDO principles are protocol independent, although it may be unclear which protocol to use for which FDO (although \texttt{0.DOIP/Transport} can be specified after already contacting DOIP).
-Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API}) & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
+Cordra supports \href{https://www.cordra.org/documentation/api/doip.html}{native DOIP}, \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP} and \href{https://www.cordra.org/documentation/api/rest-api.html}{Cordra REST API})
+  & HTTP/1.1 semantics can seemlessly upgrade to HTTP/2 and HTTP/3.
 \texttt{http} vs \texttt{https} URIs exposes encryption detail\footnote{The \texttt{http} protocol (port 80) can in theory also upgrade \autocite{rfc2817} to TLS encryption, as commonly used by \href{https://www.rfc-editor.org/rfc/rfc8010.html\#section-8.2}{Internet Printing Protocol} for \texttt{ipp} URIs, but on the Web, best practice is explicit \texttt{https} (port 443) URLs to ensure following links stay secure.}.
@@
 \\
-\textbf{Inheritance}: \emph{Deriving specialised interface from another type} & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
+\textbf{Inheritance}: \emph{Deriving specialised interface from another type}
+  & DOIP types nested with parents, implying shared FDO structures (unclear if operations are inherited).
 FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} \autocite{Draftbhuttonjsonschema00}) & Syntactically Media Type with multiple suffixes \autocite{Draftietfmediamansuffixes00MediaTypes} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) \autocite{rfc6906}.
+Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} \autocite{Draftbhuttonjsonschema00})
+  & Syntactically Media Type with multiple suffixes \autocite{Draftietfmediamansuffixes00MediaTypes} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) \autocite{rfc6906}.
 In metadata, semantic type systems (RDFS \autocite{w3-rdf-schema}), OWL2 \autocite{w3-owl2-overview}, SKOS \autocite{w3-skos-primer}).
@@
 \\
-\textbf{Signal interfaces}: \emph{asynchronous handling of messages} & DOIP 2.0 is synchronous, in FDO async operations undefined.
-Could be handled as custom jobs/futures FDOs & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} \autocite{rfc7540}, Web Sockets \autocite{WebSocketsStandard}, Linked Data Notifications \autocite{w3-ldn}, AtomPub \autocite{rfc5023}, SWORD \autocite{SWORDSpecification}, Micropub \autocite{w3-micropub}, more typically ad-hoc jobs/futures REST resources \\
-\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages} & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"}) & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \autocite{HE7Ikwwl}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \autocite{rfc6570}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \autocite{OpenAPISpecificationV3}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \autocite{fieldingArchitecturalStylesDesign2000a}, an important element of the REST architectural style.} incl.~Hydra \autocite{HydraW3CCommunity}, schema.org Actions \autocite{SchemaOrgActions}, JSON HAL \autocite{Draftkellyjsonhal08}) \& Link headers (RFC8288) \autocite{rfc8288} \\
-\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams} & Undefined in FDO.
-DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics) & HTTP 1.1 \autocite{rfc7230} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) \autocite{rfc8216}, MPEG-DASH (ISO/IEC 23009-1:2019) \autocite{iso23009} \\
+\textbf{Signal interfaces}: \emph{asynchronous handling of messages}
+  & DOIP 2.0 is synchronous, in FDO async operations undefined. Could be handled as custom jobs/futures FDOs
+  & HTTP/2 \href{https://datatracker.ietf.org/doc/html/rfc7540\#section-5}{multiplexed streams} \autocite{rfc7540}, Web Sockets \autocite{WebSocketsStandard}, Linked Data Notifications \autocite{w3-ldn}, AtomPub \autocite{rfc5023}, SWORD \autocite{SWORDSpecification}, Micropub \autocite{w3-micropub}, more typically ad-hoc jobs/futures REST resources \\
+\textbf{Operation interfaces}: \emph{defining operations possible on an instance, interface of request/response messages}
+  & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"})
+  & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \autocite{HE7Ikwwl}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \autocite{rfc6570}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \autocite{OpenAPISpecificationV3}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \autocite{fieldingArchitecturalStylesDesign2000a}, an important element of the REST architectural style.} incl.~Hydra \autocite{HydraW3CCommunity}, schema.org Actions \autocite{SchemaOrgActions}, JSON HAL \autocite{Draftkellyjsonhal08}) \& Link headers (RFC8288) \autocite{rfc8288} \\
+\textbf{Stream interfaces}: \emph{operations that can handle continuous information streams}
+  & Undefined in FDO. DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics)
+  & HTTP 1.1 \autocite{rfc7230} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) \autocite{rfc8216}, MPEG-DASH (ISO/IEC 23009-1:2019) \autocite{iso23009} \\
 \bottomrule

### 2023-03-21 latex/table5.tex
table magic

@@
 FAIR ID
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\raggedright
 Indicator
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 FDO guidelines
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 FDO/DOIP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 FDO/LDP
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
+\end{minipage}
+  & \begin{minipage}[b]{\linewidth}\centering
 Linked Data examples
@@
 \endhead
-RDA-F1-01M & Metadata is identified by a persistent identifier & FDOF4 & Optional \emph{Metadata FDO} w/separate PID & Content-negotiation to URL, not required to be PID & Metadata typically don't have own PID \\
-RDA-F1-01D & Data is identified by a persistent identifier & FDOF1 & PIDs required (FDOF1).
-Handle, DOI.
-& FDOF-IR (Identifier Record).
-PID can be any URI & ``Cool'' URIs \autocite{berners-lee-cool-uris}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
-RDA-F1-02M & Metadata is identified by a globally unique identifier & FDOR4 FDOF8 & Optional \emph{Metadata FDO}, unspecified how to indicate & Content-negotiation to URL & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}.
-FAIR Signposting.
-\\
-RDA-F1-02D & Data is identified by a globally unique identifier & FDOF1 & All FDOs have PIDs (FDOR1), DOIP uses Handle system & FDOF-IR (Identifier Record) & Always accessed by URL \\
-RDA-F2-01M & Rich metadata is provided to allow discovery & FDOF2 FDOF4 FDOF8 FDOF9 & FDO has key-value metadata.
-Unclear how to link to additional metadata.
-& FDOF-IR links to multiple metadata records & RDF-based metadata by content negotiation or FAIR Signposting.
-Embedded in landing page (RDFa).
-\\
-RDA-F3-01M & Metadata includes the identifier for the data & --- & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID & PID only required in FDOF-IR record.
+RDA-F1-01M
+  & Metadata is identified by a persistent identifier
+  & FDOF4
+  & Optional \emph{Metadata FDO} w/separate PID
+  & Content-negotiation to URL, not required to be PID
+  & Metadata typically don't have own PID \\
+RDA-F1-01D
+  & Data is identified by a persistent identifier
+  & FDOF1
+  & PIDs required (FDOF1). Handle, DOI.
+  & FDOF-IR (Identifier Record). PID can be any URI
+  & ``Cool'' URIs \autocite{berners-lee-cool-uris}, PURL services incl.~\texttt{purl.org}, \texttt{w3id.org} \\
+RDA-F1-02M
+  & Metadata is identified by a globally unique identifier
+  & FDOR4 FDOF8
+  & Optional \emph{Metadata FDO}, unspecified how to indicate
+  & Content-negotiation to URL
+  & Not required, content-negotiation can redirect to URL or \texttt{Content-Location}. FAIR Signposting. \\
+RDA-F1-02D
+  & Data is identified by a globally unique identifier
+  & FDOF1
+  & All FDOs have PIDs (FDOR1), DOIP uses Handle system
+  & FDOF-IR (Identifier Record)
+  & Always accessed by URL \\
+RDA-F2-01M
+  & Rich metadata is provided to allow discovery
+  & FDOF2 FDOF4 FDOF8 FDOF9
+  & FDO has key-value metadata. Unclear how to link to additional metadata.
+  & FDOF-IR links to multiple metadata records
+  & RDF-based metadata by content negotiation or FAIR Signposting. Embedded in landing page (RDFa). \\
+RDA-F3-01M
+  & Metadata includes the identifier for the data
+  & ---
+  & \texttt{id} and \texttt{type} are required metadata elements PIDs, also implicit as requests must use PID
+  & PID only required in FDOF-IR record.
   & PID inclusion typical, but often inconsistent (e.g.~\texttt{www.example.com} vs \texttt{example.com}) or missing (use of \texttt{\textless{}\textgreater{}} as \emph{this} subject) \\
-RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries).
-Handle registry only searchable by PID.
-& Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g.
-\autocite{NCBOBioPortal}).
-Indexing by search engines if exposing HTML w/schema.org.
-\\
-RDA-A1-01M & Metadata contains information to enable the user to get access to the data & FDOF3 FDOF6 & Directly by DOIP, but not included in FDO metadata.
-\texttt{handle.net} HTTP resolution may redirect to landing page & Any property can point to URIs, but unclear if it is data & Common with clickable ``follow your nose'' URLs \\
-RDA-A1-02M & Metadata can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso & HTTP content-negotiation to HTML is common \\
-RDA-A1-02D & Data can be accessed manually (i.e.~with human intervention) & --- & (Cordra HTML landing page from \texttt{handle.net} URIs) & Optional content-negotiation & Direct download, HTML landing pages common for DOIs \\
-RDA-A1-03M & Metadata identifier resolves to a metadata record & FDOF8+FDOF2 & --- & --- & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
-RDA-A1-03D & Data identifier resolves to a digital object & FDOF2 & Required, but frequently not directly resolvable & Recommended, but any URI acceptable & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
-RDA-A1-04M & Metadata is accessed through standardised protocol & G9 FDOF3 & Retrievable from PID (FDOF3).
-Informal DOIP standard maintained by DONA Foundation & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation) & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
-RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP \autocite{rfc9110} & HTTP/HTTPS, FTP (now less common), GridFTP \autocite{allcockGlobusStripedGridFTP} (for large data), ARK \autocite{ARKIdentifierScheme} \\
-RDA-A1-05D & Data can be accessed automatically (i.e.~by a computer program) & G4 FDOF3 FDOF6 & Required, but few client libraries & & Ubiquitous, hundreds of HTTP libraries \\
-RDA-A1.1-01M & Metadata is accessible through a free access protocol 
-    & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} 
-        system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.}
-    protocol \autocite{rfc3652}. 
-    One server implementation \autocite{HandleNetRegistry}, free\footnote{The 
-        \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license} is not OSI-approved \autocite{LicensesStandardsOpen}  as an open source license -- it includes usage restrictions and requires Service Agreements.
+RDA-F4-01M
+  & Metadata is offered in such a way that it can be harvested and indexed
+  & FDOF10
+  & No, registries not required (except Data Type Registries). Handle registry only searchable by PID.
+  & Not specified
+  & Not specified, several registries/catalogues for vocabularies/types (e.g. \autocite{NCBOBioPortal}). Indexing by search engines if exposing HTML w/schema.org. \\
+RDA-A1-01M
+  & Metadata contains information to enable the user to get access to the data
+  & FDOF3 FDOF6
+  & Directly by DOIP, but not included in FDO metadata. \texttt{handle.net} HTTP resolution may redirect to landing page
+  & Any property can point to URIs, but unclear if it is data
+  & Common with clickable ``follow your nose'' URLs \\
+RDA-A1-02M
+  & Metadata can be accessed manually (i.e.~with human intervention)
+  & ---
+  & (Cordra HTML landing page from \texttt{handle.net} URIs)
+  & Optional content-negotiation, e.g.~by Apache Marmotta, OpenLink Virtuoso
+  & HTTP content-negotiation to HTML is common \\
+RDA-A1-02D
+  & Data can be accessed manually (i.e.~with human intervention)
+  & ---
+  & (Cordra HTML landing page from \texttt{handle.net} URIs)
+  & Optional content-negotiation
+  & Direct download, HTML landing pages common for DOIs \\
+RDA-A1-03M
+  & Metadata identifier resolves to a metadata record
+  & FDOF8+FDOF2
+  & ---
+  & ---
+  & \texttt{Content-Location} or HTTP redirection may indicate metadata URI \\
+RDA-A1-03D
+  & Data identifier resolves to a digital object
+  & FDOF2
+  & Required, but frequently not directly resolvable
+  & Recommended, but any URI acceptable
+  & Resolvable HTTP/HTTPS URIs are most common, now infrequent URNs are not directly resolvable \\
+RDA-A1-04M
+  & Metadata is accessed through standardised protocol
+  & G9 FDOF3
+  & Retrievable from PID (FDOF3). Informal DOIP standard maintained by DONA Foundation
+  & LDP standard maintained by W3C, HTTP standards maintained by IETF, FDO components resolved by informal proposals (custom vocabulary, extra HTTP methods) or HTTP content negotiation)
+  & Formal HTTP standards maintained by IETF, HTTP content negotiation, informal FAIR Signposting \\
+RDA-A1-04D
+  & Data is accessible through standardised protocol
+  & G9
+  & (see above)
+  & HTTP \autocite{rfc9110}
+  & HTTP/HTTPS, FTP (now less common), GridFTP \autocite{allcockGlobusStripedGridFTP} (for large data), ARK \autocite{ARKIdentifierScheme} \\
+RDA-A1-05D
+  & Data can be accessed automatically (i.e.~by a computer program)
+  & G4 FDOF3 FDOF6
+  & Required, but few client libraries
+  & & Ubiquitous, hundreds of HTTP libraries \\
+RDA-A1.1-01M
+  & Metadata is accessible through a free access protocol    
+  & G1 G8 G9
+  & Partially realised: Handle system is open\footnote{
+        The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.}
+    protocol \autocite{rfc3652}. One server implementation \autocite{HandleNetRegistry}, free\footnote{
+        The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license} is not OSI-approved \autocite{LicensesStandardsOpen}  as an open source license -- it includes usage restrictions and requires Service Agreements.
 It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
@@
   & DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}. \\
-RDA-A1.1-01D & Data is accessible through a free access protocol & G9 & (see above) & URI, DNS, HTTP, TLS & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
-RDA-A1.2-01D & Data is accessible through an access protocol that supports authentication and authorisation & (FDOR9) & TLS certificates, \texttt{authentication} field (details unspecified) & Implied & HTTP authentication, TLS certificates \\
+RDA-A1.1-01D
+  & Data is accessible through a free access protocol
+  & G9
+  & (see above)
+  & URI, DNS, HTTP, TLS
+  & URI, DNS, HTTP, TLS. Non-free DRM may be used (e.g.~subscription video streaming) \\
+RDA-A1.2-01D
+  & Data is accessible through an access protocol that supports authentication and authorisation
+  & (FDOR9)
+  & TLS certificates, \texttt{authentication} field (details unspecified)
+  & Implied
+  & HTTP authentication, TLS certificates \\
 RDA-A2-01M\label{RDA-A2-01M}
-& Metadata is guaranteed to remain available after data is no longer available & FDOF12 & --- & Unspecified, however FDOF-IR links to separate metadata records & --- \\
-RDA-I1-01M & Metadata uses knowledge representation expressed in standardised format & FDOF8 & Required, but not currently defined & --- & Always implied by use of RDF syntaxes. \\
-RDA-I1-01D & Data uses knowledge representation expressed in standardised format & --- & --- & --- & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
-RDA-I1-02M & Metadata uses machine-understandable knowledge representation & FDOF8 & Required & Optional RDF metadata with any vocabulary & Always implied by use of RDF syntaxes. \\
-RDA-I1-02D & Data uses machine-understandable knowledge representation & G4 G7 FDOR2 & No requirements on binary data formats & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources & Common, specially for scientific data formats \\
-RDA-I2-01M & Metadata uses FAIR-compliant vocabularies & G3 FDOF10 & Informally required & Unspecified, implied by use of RDF?
+  & Metadata is guaranteed to remain available after data is no longer available
+  & FDOF12
+  & ---
+  & Unspecified, however FDOF-IR links to separate metadata records
+  & --- \\
+RDA-I1-01M
+  & Metadata uses knowledge representation expressed in standardised format
+  & FDOF8
+  & Required, but not currently defined
+  & ---
+  & Always implied by use of RDF syntaxes. \\
+RDA-I1-01D
+  & Data uses knowledge representation expressed in standardised format
+  & ---
+  & ---
+  & ---
+  & Common (e.g.~HDF5, JSON, XML), yet common scientific data formats frequently not standardised \\
+RDA-I1-02M
+  & Metadata uses machine-understandable knowledge representation
+  & FDOF8
+  & Required
+  & Optional RDF metadata with any vocabulary
+  & Always implied by use of RDF syntaxes. \\
+RDA-I1-02D
+  & Data uses machine-understandable knowledge representation
+  & G4 G7 FDOR2
+  & No requirements on binary data formats
+  & Only indirectly, \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-basic-container}{LDP Basic Container} reference only information resources
+  & Common, specially for scientific data formats \\
+RDA-I2-01M
+  & Metadata uses FAIR-compliant vocabularies
+  & G3 FDOF10
+  & Informally required
+  & Unspecified, implied by use of RDF?
   & FAIR practices for LD vocabularies increasingly common, sometimes inconsistent (e.g.~PURLs that don't resolve) or incomplete (e.g.~unknown license) \\
-RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \autocite{iso16684} \\
-RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs \autocite{DataW3C}), \texttt{rdfs:seeAlso}, FAIR signposting \autocite{vandesompelFAIRSignpostingProfile2022} \texttt{describedby} \\
-RDA-I3-01D & Data includes references to other data & G6 FDOR3 FDOR11 & --- & --- & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
-RDA-I3-02M & Metadata includes references to other data & G6 FDOR3 FDOR8 & Implied from custom FDO type's attribute & LDP Direct Container members can be any resources & URI objects are frequently data references, may be indirect via PID \\
-RDA-I3-02D & Data includes qualified references to other data & FDOR3 FDOR11 & Only indirectly through FDO metadata & Indirectly through LDP membership & Uncommon: Link relations, FAIR Signposting \\
-RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles \autocite{w3-prov-links}, \href{https://schema.org/Role}{schema.org/Role} \\
-RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role \autocite{hollandIntroducingRole2014}, \texttt{prov:specializationOf} \autocite{w3-prov-o}, OAI-ORE Proxy \autocite{ORESpecificationAbstract}) \\
-RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required.
-Kernel metadata attributes desired \autocite{fdo-KernelAttributes} but not assigned PIDs yet.
-& Unspecified.
-Multiple metadata records can allow multiple semantic profiles.
+RDA-I2-01D
+  & Data uses FAIR-compliant vocabularies
+  & ---
+  & ---
+  & ---
+  & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \autocite{iso16684} \\
+RDA-I3-01M
+  & Metadata includes references to other metadata
+  & FDOR8
+  & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference
+  & ---
+  & By definition (Linked Data reference existing URIs \autocite{DataW3C}), \texttt{rdfs:seeAlso}, FAIR signposting \autocite{vandesompelFAIRSignpostingProfile2022} \texttt{describedby} \\
+RDA-I3-01D
+  & Data includes references to other data
+  & G6 FDOR3 FDOR11
+  & ---
+  & ---
+  & URL hyperlinks common in several formats (HTML, PDF, JSON, XML). \\
+RDA-I3-02M
+  & Metadata includes references to other data
+  & G6 FDOR3 FDOR8
+  & Implied from custom FDO type's attribute
+  & LDP Direct Container members can be any resources
+  & URI objects are frequently data references, may be indirect via PID \\
+RDA-I3-02D
+  & Data includes qualified references to other data
+  & FDOR3 FDOR11
+  & Only indirectly through FDO metadata
+  & Indirectly through LDP membership
+  & Uncommon: Link relations, FAIR Signposting \\
+RDA-I3-03M
+  & Metadata includes qualified references to other metadata
+  & (FDOR3)
+  & Qualification by attribute keys defined per FDO Type
+  & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container}
+  & Qualifications by property, PROV bundles \autocite{w3-prov-links}, \href{https://schema.org/Role}{schema.org/Role} \\
+RDA-I3-04M
+  & Metadata include qualified references to other data
+  & (FDOR3)
+  & Qualification by attribute keys defined per FDO type
+  & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container}
+  & Qualifications by property, n-ary indirection (schema.org Role \autocite{hollandIntroducingRole2014}, \texttt{prov:specializationOf} \autocite{w3-prov-o}, OAI-ORE Proxy \autocite{ORESpecificationAbstract}) \\
+RDA-R1-01M
+  & Plurality of accurate and relevant attributes are provided to allow reuse
+  & FDOF4
+  & Required. Kernel metadata attributes desired \autocite{fdo-KernelAttributes} but not assigned PIDs yet.
+  & Unspecified. Multiple metadata records can allow multiple semantic profiles.
   & Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
 Rough consensus on kernel metadata: schema.org \autocite{10pzKCEUE}, Dublin Core Terms \autocite{DCMIMetadataTerms}, DCAT \autocite{w3-vocab-dcat-2}, FOAF \autocite{FOAFVocabularySpecification} \\
-RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata \autocite{fdo-KernelAttributes} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \autocite{w3-vocab-dcat-2} \\
-RDA-R1.1-02M & Metadata refers to a standard reuse licence & --- & --- & --- & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
-RDA-R1.1-03M & Metadata refers to a machine-understandable reuse licence & --- & --- & --- & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
-RDA-R1.2-01M & Metadata includes provenance information according to community-specific standards & FDOR9 FDOR10 & Unspecified (some CORDRA types add getProvenance methods).
-PID Kernel attributes?
-Unspecified W3C PROV-O, PAV & & \\
-RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O \autocite{w3-prov-o}, PAV \autocite{ciccaresePAVOntologyProvenance2013e}, Dublin Core Terms \autocite{DCMIMetadataTerms} \\
-RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen \autocite{hardistySpecimenDataRefinery2022a}) & --- & Common, e.g.~DCAT 2 \autocite{w3-vocab-dcat-2}, BioSchemas \autocite{Bioschemas} \\
-RDA-R1.3-01D & Data complies with a community standard & (FDOR3) & --- & --- & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
-RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal \autocite{NCBOBioPortal}, Darwin Core \autocite{wieczorekDarwinCoreEvolving2012} \\
-RDA-R1.3-02D & Data is expressed in compliance with a machine-understandable community standard & (FDOR2) & No, FDO is typed but data can be any bytestream & --- & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
+RDA-R1.1-01M
+  & Metadata includes information about the licence under which the data can be reused
+  & ---
+  & \texttt{licenseConditions} URL/PID in kernel metadata \autocite{fdo-KernelAttributes}
+  & ---
+  & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \autocite{w3-vocab-dcat-2} \\
+RDA-R1.1-02M
+  & Metadata refers to a standard reuse licence
+  & ---
+  & ---
+  & ---
+  & \href{https://spdx.org/licenses/}{SPDX} and \href{https://creativecommons.org/}{Creative Commons} URIs common, identifiers often inconsistent \\
+RDA-R1.1-03M
+  & Metadata refers to a machine-understandable reuse licence
+  & ---
+  & ---
+  & ---
+  & \href{https://spdx.dev/resources/use/\#documents}{SPDX documents} uncommon \\
+RDA-R1.2-01M
+  & Metadata includes provenance information according to community-specific standards
+  & FDOR9 FDOR10
+  & Unspecified (some CORDRA types add getProvenance methods). PID Kernel attributes? Unspecified W3C PROV-O, PAV
+  & & \\
+RDA-R1.2-02M
+  & Metadata includes provenance information according to a cross-community language
+  & FDOR9 FDOR8
+  & ---
+  & ---
+  & W3C PROV-O \autocite{w3-prov-o}, PAV \autocite{ciccaresePAVOntologyProvenance2013e}, Dublin Core Terms \autocite{DCMIMetadataTerms} \\
+RDA-R1.3-01M
+  & Metadata complies with a community standard
+  & FDOR10 FROR8
+  & (Emerging, e.g.~DiSSCo Digital Specimen \autocite{hardistySpecimenDataRefinery2022a})
+  & ---
+  & Common, e.g.~DCAT 2 \autocite{w3-vocab-dcat-2}, BioSchemas \autocite{Bioschemas} \\
+RDA-R1.3-01D
+  & Data complies with a community standard
+  & (FDOR3)
+  & ---
+  & ---
+  & Common, HTTP use registered IANA \href{https://www.iana.org/assignments/media-types/media-types.xhtml}{media types}, additional scientific file formats frequently not standardised or identified \\
+RDA-R1.3-02M
+  & Metadata is expressed in compliance with a machine-understandable community standard
+  & FDOF4 FDOF10
+  & Recommended
+  & ---
+  & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal \autocite{NCBOBioPortal}, Darwin Core \autocite{wieczorekDarwinCoreEvolving2012} \\
+RDA-R1.3-02D
+  & Data is expressed in compliance with a machine-understandable community standard
+  & (FDOR2)
+  & No, FDO is typed but data can be any bytestream
+  & ---
+  & Occassionally, (e.g.~\href{https://github.com/The-Sequence-Ontology/Specifications/blob/master/gff3.md}{GFF3}, \href{https://fits.gsfc.nasa.gov/fits_standard.html}{FITS}, \href{https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml}{ESRI}) \\
 \bottomrule

### 2023-03-21 latex/table4.tex
updated bibliography entries

@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \autocite{13TcbsZF6}) and Web technologies (using Linked Data) as middleware infrastructures \autocite{hRzcHhPD}
+\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \autocite{foundationDigitalObjectInterface}) and Web technologies (using Linked Data) as middleware infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}
 \label{tbl:fdo-web-middleware}}\tabularnewline
@@
 FDO establishes need for multiple Data Type Registries (e.g.~managed by a community for a particular domain).
-Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} \autocite{Draftbhuttonjsonschema00})
+Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} \autocite{Draftbhuttonjsonschema})
   & Syntactically Media Type with multiple suffixes \autocite{Draftietfmediamansuffixes00MediaTypes} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) \autocite{rfc6906}.
@@
 OpenAPI 3 \autocite{OpenAPISpecificationV3} \href{https://spec.openapis.org/oas/v3.1.0\#composition-and-inheritance-polymorphism}{inheritance and Polymorphism}.
-XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} \autocite{w3-xmlschema11}, JSON \texttt{\$schema} \autocite{Draftbhuttonjsonschema00}), JSON-LD \texttt{@context} \autocite{w3-json-ld}.
+XML \texttt{xsd:schemaLocation} or \texttt{xsd:type} \autocite{w3-xmlschema11}, JSON \texttt{\$schema} \autocite{Draftbhuttonjsonschema}), JSON-LD \texttt{@context} \autocite{w3-json-ld}.
 Large number of domain-specific and general ontologies define semantic types, but finding and selecting remains a challenge.
@@
   & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"})
-  & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \autocite{HE7Ikwwl}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \autocite{rfc6570}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \autocite{OpenAPISpecificationV3}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \autocite{fieldingArchitecturalStylesDesign2000a}, an important element of the REST architectural style.} incl.~Hydra \autocite{HydraW3CCommunity}, schema.org Actions \autocite{SchemaOrgActions}, JSON HAL \autocite{Draftkellyjsonhal08}) \& Link headers (RFC8288) \autocite{rfc8288} \\
+  & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \autocite{rfc7231}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \autocite{rfc6570}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \autocite{OpenAPISpecificationV3}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \autocite{fieldingArchitecturalStylesDesign2000a}, an important element of the REST architectural style.} incl.~Hydra \autocite{HydraW3CCommunity}, schema.org Actions \autocite{SchemaOrgActions}, JSON HAL \autocite{Draftkellyjsonhal08}) \& Link headers (RFC8288) \autocite{rfc8288} \\
 \textbf{Stream interfaces}: \emph{operations that can handle continuous information streams}

### 2023-03-21 latex/table5.tex
updated bibliography entries

@@
   >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model \autocite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \autocite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \autocite{13TcbsZF6}, Linked Data Platform (LDP) \autocite{7szz7dwO} and examples from Linked Data practices in general.
+\caption{Assessing RDA's FAIR Data Maturity Model \autocite{groupFAIRDataMaturity2020,bahimFAIRDataMaturity2020a} (first 2 columns) against the FDO guidelines \autocite{boninoFAIRDigitalObject}, FDO implemented with the protocol DOIPv2 \autocite{foundationDigitalObjectInterface}, Linked Data Platform (LDP) \autocite{FDOFramework} and examples from Linked Data practices in general.
 (--- indicates \emph{Unspecified}, may be possible with additional conventions)
@@
   & Large number of general and domain-specific vocabularies can make it hard to find relevant attributes.
-Rough consensus on kernel metadata: schema.org \autocite{10pzKCEUE}, Dublin Core Terms \autocite{DCMIMetadataTerms}, DCAT \autocite{w3-vocab-dcat-2}, FOAF \autocite{FOAFVocabularySpecification} \\
+Rough consensus on kernel metadata: schema.org \autocite{SchemaOrg}, Dublin Core Terms \autocite{DCMIMetadataTerms}, DCAT \autocite{w3-vocab-dcat-2}, FOAF \autocite{FOAFVocabularySpecification} \\
 RDA-R1.1-01M

### 2023-03-21 manuscript.tex
more bib [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/f1ceda5536de40bba679ff7801af69b4405a2cc5.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/f1ceda5536de40bba679ff7801af69b4405a2cc5/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4481402299

--- 2023-03-21 manuscript.tex
@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/9f1d51557301a61b06b5b4dd934b2b478512567f/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/f1ceda5536de40bba679ff7801af69b4405a2cc5/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/9f1d51557301a61b06b5b4dd934b2b478512567f}{stain/2022-fdo-paper@9f1d515}
+from \href{https://github.com/stain/2022-fdo-paper/tree/f1ceda5536de40bba679ff7801af69b4405a2cc5}{stain/2022-fdo-paper@f1ceda5}
 on March 21, 2023.

### 2023-03-22 latex/main.tex
table magic
cite rdf

@@
   \url{https://doi.org/#1}}
-\usepackage{biblatex2bibitem}  
+%\usepackage{biblatex2bibitem}  
 %\urlstyle{tt}
@@
 
-Established Semantic Web services and standards include SPARQL \autocite{w3-sparql11-overview} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
+Established Semantic Web services and standards include SPARQL \autocite{w3-sparql11-overview} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} \autocite{w3-rdf11-concepts} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
 

### 2023-03-22 manuscript.tex
cite rdf [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/17c99087d5562cf949ae5952e872d5d04271aae4.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/17c99087d5562cf949ae5952e872d5d04271aae4/checks
https://github.com/stain/2022-fdo-paper/actions/runs/4493389704

--- 2023-03-21 manuscript.tex
@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/c84452f9fd52ac995ffa76f1c51091daa85f9396/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/17c99087d5562cf949ae5952e872d5d04271aae4/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/c84452f9fd52ac995ffa76f1c51091daa85f9396}{stain/2022-fdo-paper@c84452f}
-on March 21, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/17c99087d5562cf949ae5952e872d5d04271aae4}{stain/2022-fdo-paper@17c9908}
+on March 22, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 21 March 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 22 March 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 21 March 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 22 March 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 21 March 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 22 March 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 21 March 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 22 March 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 21 March 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 22 March 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 21 March 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 22 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 21 March 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 22 March 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 21 March 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 22 March 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 21 March 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 22 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 21 March 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 22 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 21 March 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 22 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 21 March 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 22 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 21 March 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 22 March 2023).
 
@@
 Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 21 March 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 22 March 2023).
 
@@
 2022.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 21 March 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 22 March 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 21 March 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 22 March 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 21 March 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 22 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 21 March 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 22 March 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 21 March 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 22 March 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 21 March 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 22 March 2023).
 
@@
 The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 21 March 2023).
+\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 22 March 2023).
 
@@
 The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 21 March 2023).
+\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 22 March 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 21 March 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 22 March 2023).
 

### 2023-05-18 latex/main.tex
comments

@@
 This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \autocite{weigelRDARecommendationPID2018}.
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
+This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}\autocite{islam_2023}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
 
@@
 
+% Signposting / RO-Crate
+% ProtocolBuffer
+% https://www.cordra.org/documentation/api/doip-api-for-http-clients.html
+% https://pages.educs-hosting.net/konsortswd/konsortswd-api/
+
 \begin{itemize}

### 2023-05-18 latex/table4.tex
comments

@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-\caption{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \autocite{foundationDigitalObjectInterface}) and Web technologies (using Linked Data) as middleware infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}
+
+\caption[Comparing FAIR Digital Object and Web technologies as middleware infrastructures]{Comparing FAIR Digital Object with the DOIP 2.0 protocol \autocite{foundationDigitalObjectInterface} against Web technologies using Linked Data, considered as middleware infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}
 \label{tbl:fdo-web-middleware}}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
+Quality  & FDO with DOIP  & Web with Linked Data \\
 \midrule
@@
 \toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
+Quality  & FDO with DOIP  & Web with Linked Data \\
 \midrule
@@
   & HTTP operations are classified as being stateless/idempotent or not (e.g.~\texttt{PUT} changes state, but can be repeated on failure), although these constraints are occassionally violated by Web applications.
-Cache control, \texttt{ETag} (\textasciitilde{} checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
+Cache control, \texttt{ETag} (e.g.
+checksum) and modification date in HTTP headers allows detection of concurrent changes on a single resource.
 \\
@@
 Semantics of type system currently undefined for FDO and DOIP, syntactic types can also piggyback of FDO type's schema (e.g.~\href{(https://www.cordra.org/documentation/design/schemas.html\#schema-references)}{CORDRA \texttt{\$ref}} use of \href{https://json-schema.org/draft/2020-12/json-schema-core.html\#references}{JSON Schema references} \autocite{Draftbhuttonjsonschema})
-  & Syntactically Media Type with multiple suffixes \autocite{Draftietfmediamansuffixes00MediaTypes} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) \autocite{rfc6906}.
+  & Syntactically Media Type with multiple suffixes \autocite{Draftietfmediamansuffixes00MediaTypes} (mainly used with \texttt{+json}), declaration of subtypes as profiles (RFC6906) \citetitle{rfc6906}.
 In metadata, semantic type systems (RDFS \autocite{w3-rdf-schema}), OWL2 \autocite{w3-owl2-overview}, SKOS \autocite{w3-skos-primer}).
@@
   & CRUD predefined in DOIP, custom operations through \texttt{0.DOIP/Op.ListOperations} (can be FDOs of type \texttt{0.TYPE/DOIPOperation}, more typically local identifiers like \texttt{"getProvenance"})
-  & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \autocite{rfc7231}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \autocite{rfc6570}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \autocite{OpenAPISpecificationV3}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \autocite{fieldingArchitecturalStylesDesign2000a}, an important element of the REST architectural style.} incl.~Hydra \autocite{HydraW3CCommunity}, schema.org Actions \autocite{SchemaOrgActions}, JSON HAL \autocite{Draftkellyjsonhal08}) \& Link headers (RFC8288) \autocite{rfc8288} \\
+  & CRUD predefined in \href{https://datatracker.ietf.org/doc/html/rfc7231\#section-4.3}{HTTP methods} \autocite{rfc7231}, (\href{https://www.iana.org/assignments/http-methods/http-methods.xhtml}{extended by registration}), URI Templates \autocite{rfc6570}, \href{https://spec.openapis.org/oas/v3.1.0.html\#operation-object}{OpenAPI operations} \autocite{OpenAPISpecificationV3}, HATEOAS\footnote{HATEOAS: Hypermedia as the Engine of Application State \autocite{fieldingArchitecturalStylesDesign2000a}, an important element of the REST architectural style.} incl.~Hydra \autocite{HydraW3CCommunity}, schema.org Actions \autocite{SchemaOrgActions}, JSON HAL \autocite{Draftkellyjsonhal08} \& Link headers (RFC8288) \autocite{rfc8288} \\
 \textbf{Stream interfaces}: \emph{operations that can handle continuous information streams}
   & Undefined in FDO. DOIP can support multiple byte stream elements (need custom FDO type to determine stream semantics)
-  & HTTP 1.1 \autocite{rfc7230} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) \autocite{rfc8216}, MPEG-DASH (ISO/IEC 23009-1:2019) \autocite{iso23009} \\
+  & HTTP 1.1 \autocite{rfc7230} \href{https://datatracker.ietf.org/doc/html/rfc7230\#section-4.1}{chunked transfer}, HLS (RFC8216) \autocite{rfc8216}, MPEG-DASH \autocite{iso23009} \\
 \bottomrule

### 2023-05-18 manuscript.tex
comments [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/0e5545c8430c3f287c3b3a8c7618cf039d63b9d9.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/0e5545c8430c3f287c3b3a8c7618cf039d63b9d9/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5014395287

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/17c99087d5562cf949ae5952e872d5d04271aae4/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/0e5545c8430c3f287c3b3a8c7618cf039d63b9d9/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/17c99087d5562cf949ae5952e872d5d04271aae4}{stain/2022-fdo-paper@17c9908}
-on March 22, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/0e5545c8430c3f287c3b3a8c7618cf039d63b9d9}{stain/2022-fdo-paper@0e5545c}
+on May 18, 2023.
 
@@
 FDOF6: \emph{CRUD operations + extensions} & \texttt{0.DOIP/Op.Create}, \texttt{Op.Retrieve}, \texttt{Op.Update}, \texttt{Op.Delete} but also \texttt{0.DOIP/Op.Search}.
-& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan, 2022}) are used instead.
+& Document & \texttt{PUT}, \texttt{GET}, \texttt{POST}, \texttt{DELETE}, \texttt{PATCH}, \texttt{HEAD} -- extension operations (e.g.~WebDAV \texttt{COPY}) not used, resource patterns (\protect\hyperlink{ref-i9Ang0rM}{martinekuan, 2023}) are used instead.
 & Document how operation resources can be discovered from an LPD container.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 22 March 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 18 May 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 22 March 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 18 May 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 22 March 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 18 May 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 22 March 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 18 May 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 22 March 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 18 May 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 22 March 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 22 March 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 18 May 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 22 March 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 18 May 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 22 March 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 22 March 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 22 March 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 22 March 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 22 March 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles.
-\emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 22 March 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 18 May 2023).
 
@@
 martinekuan.
-2022.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 22 March 2023).
+2023.Web API design best practices - Azure Architecture Center.
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 18 May 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 22 March 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 18 May 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 22 March 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 22 March 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 18 May 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 22 March 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 18 May 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 22 March 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme.
-\emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 22 March 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 18 May 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles.
-\emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 22 March 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 18 May 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 22 March 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 18 May 2023).
 

### 2023-05-18 manuscript.tex
Merge remote-tracking branch 'origin/main' [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/55b211690e436edabc36739a1bca1546f2b6d680.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/55b211690e436edabc36739a1bca1546f2b6d680/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5014914749

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/0e5545c8430c3f287c3b3a8c7618cf039d63b9d9/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/55b211690e436edabc36739a1bca1546f2b6d680/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/0e5545c8430c3f287c3b3a8c7618cf039d63b9d9}{stain/2022-fdo-paper@0e5545c}
+from \href{https://github.com/stain/2022-fdo-paper/tree/55b211690e436edabc36739a1bca1546f2b6d680}{stain/2022-fdo-paper@55b2116}
 on May 18, 2023.

### 2023-05-18 latex/table1.tex
simplified table latex

@@
 \textbf{Symbiotic}: \emph{to what extent multiple applications can agree to interact, align, collaborate or cooperate}
-  & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of or compatibility between FDO types.
+  & The purpose of FDO is to enable federated machine actionable digital objects for scholarly purposes, in practice this also requires agreement of compatibility between FDO types.
 FDO encourages research communities to develop common type registries to be shared across instances.
@@
   & The Web is loosely coupled and encourages collaboration and linking by URL.
-In practice, REST APIs \autocite{fieldingArchitecturalStylesDesign2000a} end up being mandated centrally by dominant (often commercial) providers \autocite{fieldingReflectionsRESTArchitectural2017a}, which clients are required to use as-is with special code per service.
+In practice, REST APIs \autocite{fieldingArchitecturalStylesDesign2000a} end up being mandated centrally by dominant (often commercial) providers \autocite{fieldingReflectionsRESTArchitectural2017a}, and the clients are required to use each API as-is with special code per service.
 Use of Linked Data enables common tooling and semantic mapping across differences.
@@
 \textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows}
-  & FDO types and operations enable detailed choreography (see CWFP).
+  & FDO types and operations enable detailed choreography (Canonical Workflow Frameworks for Research \autocite{cwfr}).
 \texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).

### 2023-05-18 latex/table3.tex
simplified table latex

@@
 \begin{landscape}
+  \begin{small}
   \begin{longtable}[]{@{}
@@
     >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2}}@{}}
-\caption{Checking FDO guidelines \autocite{boninoFAIRDigitalObject,fdo-RequirementSpec} against its current implementations as DOIP \autocite{foundationDigitalObjectInterface} and Linked Data Platform (LDP) \autocite{FDOFramework}, with suggestions for required additions.
+    \caption[Checking FDO guidelines against its implementations]{Checking FDO guidelines \cite{boninoFAIRDigitalObject,fdo-RequirementSpec} against its current implementations as DOIP \cite{foundationDigitalObjectInterface} and Linked Data Platform (LDP) \cite{FDOFramework}, with suggestions for required additions.
   \label{tbl:fdo-checks}}\tabularnewline
   \toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
+  \textbf{FDO Guideline} & 
+  DOIP 2.0 & 
+  FDO suggestions & 
+  Linked Data Platform & 
+  LDP suggestion \\
   \midrule
@@
   \toprule
-\begin{minipage}[b]{\linewidth}\centering
-\textbf{FDO Guideline}
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-DOIP 2.0
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-FDO suggestions
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-Linked Data Platform
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-LDP suggestion
-\end{minipage} \\
+  \textbf{FDO Guideline} & 
+  DOIP 2.0 & 
+  FDO suggestions & 
+  Linked Data Platform & 
+  LDP suggestion \\
   \midrule
@@
 \end{longtable}
+\end{small}
 \end{landscape}

### 2023-05-18 latex/table4.tex
simplified table latex

@@
 \begin{landscape}
+\begin{small}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-
-\caption[Comparing FAIR Digital Object and Web technologies as middleware infrastructures]{Comparing FAIR Digital Object with the DOIP 2.0 protocol \autocite{foundationDigitalObjectInterface} against Web technologies using Linked Data, considered as middleware infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}
+	\caption[Comparing FAIR Digital Object and Web technologies as middleware infrastructures]{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \cite{13TcbsZF6}) and Web technologies (using Linked Data) as middleware infrastructures \cite{hRzcHhPD}
 \label{tbl:fdo-web-middleware}}\tabularnewline
 \toprule
-Quality  & FDO with DOIP  & Web with Linked Data \\
+\emph{Quality} & 
+FDO w/ DOIP & 
+Web w/ Linked Data \\
 \midrule
@@
 \toprule
-Quality  & FDO with DOIP  & Web with Linked Data \\
+\emph{Quality} & 
+FDO w/ DOIP & 
+Web w/ Linked Data \\
 \midrule

### 2023-05-18 latex/table5.tex
simplified table latex

@@
     >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2}}@{}}
-\caption{Assessing RDA's FAIR Data Maturity Model \autocite{groupFAIRDataMaturity2020,bahimFAIRDataMaturity2020a} (first 2 columns) against the FDO guidelines \autocite{boninoFAIRDigitalObject}, FDO implemented with the protocol DOIPv2 \autocite{foundationDigitalObjectInterface}, Linked Data Platform (LDP) \autocite{FDOFramework} and examples from Linked Data practices in general.
+    \caption[Assessing RDA's FAIR Data Maturity Model against the FDO guidelines]{Assessing RDA's FAIR Data Maturity Model \cite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \cite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \cite{13TcbsZF6}, Linked Data Platform (LDP) \cite{7szz7dwO} and examples from Linked Data practices in general.
 (--- indicates \emph{Unspecified}, may be possible with additional conventions)
@@
   \toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-FAIR ID
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\raggedright
-Indicator
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-FDO guidelines
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-FDO/LDP
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-Linked Data examples
-\end{minipage} \\
+  FAIR ID &
+  Indicator &
+  FDO guidelines &
+  FDO/DOIP &
+  FDO/LDP &
+  Linked Data examples \\
   \midrule

### 2023-05-18 latex/table6.tex
converted to peerj overleaf template
Fixing references to use bibliography
more regex!
all the citations. Bioschemas by article.
simplified table latex

--- 2023-01-24 latex/table5.tex
@@
+\begin{landscape}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
+\caption[Assessing EOSC Interoperability Framework, against FDO \& Linked Data]{Assessing EOSC Interoperability Framework \autocite[section 3.6]{corchoEOSCInteroperabilityFramework2021b} against the FDO guidelines \autocite{RwvirqWg} and Linked Data practices.}
+\label{tbl:eosc}\tabularnewline
 \toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-Layer
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Recommendation
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-FDO
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Linked Data
-\end{minipage} \\
+Layer &
+Recommendation &
+FDO &
+Linked Data \\
 \midrule
 \endhead
-Technical & Open Specification & FDO specifications are semi-open, process gradually more transparent & Open and transparent standard processes through W3C \& IETF \\
-Technical & Common security \& privacy framework & Unspecified & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
-Technical & Easy SLAs for service providers & Unspecified & None \\
-Technical & Access data in different formats & None formalised, custom operations or relations & Content-negotiation, \texttt{rel=alternate} relations \\
-Technical & Coarse-grained/fine-grained search tools & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
-Technical & Clear PID policy & Strong FDO requirements, tends towards Handle system. & Not required, different communities set policies \\
-Semantic & Clear definitions for concepts/metadata/schemas & Required by FDO requirements, but not yet formalised & Ontologies, SKOS, OWL \\
-Semantic & Semantic artefacts w/ open licenses & All artefacts are PIDs w/ license required by kernel metadata? & Open License is best practice for ontology publishing \\
-Semantic & Documentation for each semantic artefact & No direct rendering from FDO, no requirement for human-readable description & Ontology rendering, content-negotiation \\
-Semantic & Repositories of artefacts & Required, but not formalised & Bioontologies, etc \\
-Semantic & Repositories w/ clear governance & Recommended & Largely self-governed repositories, if well-established may have clear governance. \\
-Semantic & Minimal metadata model for federated discovery & Kernel metadata (\protect\hyperlink{ref-Fi62cJAo}{Working Group \& Working Group, 2022}) & DCAT, ++ \\
-Semantic & Crosswalks from minimal metadata model & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
-Semantic & Extensibility options for diciplinary metadata & Communities encouraged to establish own types & Extensible by design, domain-specific metadata may be at different granularity \\
-Semantic & Clear protocols/building blocks for federation/harvesting of artefact catalogues & Collection types not yet defined & SWORD, OAI-PMH \\
-Organisational & Interoperability-focused rules of participation recommendations & Recommended & Implied only by some communities, tendency to specialise \\
-Organisational & Usage recommendations of standardised data formats & None & None -- but common for metadata (e.g.~JSON-LD) \\
-Organisational & Usage recommendations of vocabularies & Recommended by community & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
-Organisational & Usage recommendations of metadata & Recommended by community & RO-Crate, Bioschemas \\
-Organisational & Management of permanent organization names/functions & Handle owner, but unclear contact. Contact info in DOIP service provider & ROR. DCAT contacts. \\
-Legal & Standardised human and machine-readable licenses & None & \href{https://spdx.org/licenses/}{SPDX} \\
-Legal & Permissive licenses for metadata (CC0, CC-BY-4.0) & Undefined & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
-Legal & Different licenses for different parts & Each part as separate FDO can have separate license & DCAT, RO-Crate, Named graphs for splitting metadata \\
-Legal & Mark expired/inexistent copyright & Undefined & Unclear, semantics assume copyright valid \\
-Legal & Mark orphaned data & Tombstone for deleted data, but no owner of DOIP server means FDO disappears & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
-Legal & List recommended licenses & Undefined & Best practice recommendations \\
-Legal & Track license evolution for dataset & Undefined & Versioning with PAV/PROV/DCAT \\
-Legal & Policy/guidance for patent/trade secrets violation & Undefined & Undefined, legal owner may be specified \\
-Legal & GDPR compliance for personal data & Undefined & Undefined \\
-Legal & Restrict access/use if legally required & By transport protocol (undefined by FDO/DOIP) & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
-Legal & Harmonised terms-of-use & Undefined & Undefined \\
-Legal & Alignment between EOSC and national legislation & Not applicable & Not applicable \\
+Technical      & Open Specification 
+  & FDO specifications are semi-open, process gradually more transparent 
+  & Open and transparent standard processes through W3C \& IETF \\
+Technical      & Common security \& privacy framework 
+  & Unspecified 
+  & TLS for encryption, multiple approaches for single-sign-on (e.g.~ORCID, Life Science Login). Privacy largely unspecified. \\
+Technical      & Easy SLAs for service providers 
+  & Unspecified 
+  & None \\
+Technical      & Access data in different formats 
+  & None formalised, custom operations or relations 
+  & Content-negotiation, \texttt{rel=alternate} relations \\
+Technical      & Coarse-grained/fine-grained search tools 
+  & Freetext \texttt{0.DOIP/Op.Search} on local DOIP, no federation 
+  & Coarse-grained e.g.~\href{https://datasetsearch.research.google.com/}{Google Dataset Search}, fine-grained (e.g.~federated SPARQL) require detailed vocabulary/metadata insight \\
+Technical      & Clear PID policy 
+  & Strong FDO requirements, tends towards Handle system. 
+  & Not required, different communities set policies \\
+Semantic       & Clear definitions for concepts/metadata/schemas 
+  & Required by FDO requirements, but not yet formalised 
+  & Ontologies, SKOS, OWL \\
+Semantic       & Semantic artefacts w/ open licenses 
+  & All artefacts are PIDs, license not yet required by kernel metadata
+  & Open License is best practice for ontology publishing \\
+Semantic       & Documentation for each semantic artefact 
+  & No direct rendering from FDO, no requirement for human-readable description 
+  & Ontology rendering, content-negotiation \\
+Semantic       & Repositories of artefacts 
+  & Required, but not formalised 
+  & Bioontologies, otherwise not usually federated \\
+Semantic       & Repositories w/ clear governance 
+  & Recommended 
+  & Largely self-governed repositories, if well-established may have clear governance. \\
+Semantic       & Minimal metadata model for federated discovery 
+  & Kernel metadata \autocite{fdo-KernelAttributes} based on RDA recommendations \autocite{weigelRDARecommendationPID2018}.
+  & DCAT, schema.org, Dublin Core \\
+Semantic       & Crosswalks from minimal metadata model 
+  & FDO Typing recommends referencing existing type definitions, but not as separate crosswalks 
+  & Multiple crosswalks for common metadata models, but frequently not in semantic format \\
+Semantic       & Extensibility options for diciplinary metadata 
+  & Communities encouraged to establish own types 
+  & Extensible by design, domain-specific metadata may be at different granularity \\
+Semantic       & Clear protocols/building blocks for federation/harvesting of artefact catalogues 
+  & Collection types not yet defined 
+  & SWORD, OAI-PMH \\
+Organisational  & Interoperability-focused rules of participation recommendations 
+  & Recommended 
+  & Implied only by some communities, tendency to specialise \\
+Organisational  & Usage recommendations of standardised data formats 
+  & None 
+  & None -- but common for metadata (e.g.~JSON-LD) \\
+Organisational  & Usage recommendations of vocabularies 
+  & Recommended by community 
+  & Common (see \href{https://rdmkit.elixir-europe.org/metadata_management}{RDMKit}) \\
+Organisational  & Usage recommendations of metadata 
+  & Recommended by community 
+  & RO-Crate, Bioschemas \\
+Organisational  & Management of permanent organization names/functions 
+  & Handle owner, but unclear contact. Contact info in DOIP service provider 
+  & ROR. DCAT contacts. \\
+Legal          & Standardised human and machine-readable licenses 
+  & None 
+  & \href{https://spdx.org/licenses/}{SPDX}, but not that frequently used \\
+Legal          & Permissive licenses for metadata (CC0, CC-BY-4.0) 
+  & Undefined 
+  & Both CC0, CC-BY-4.0 common, e.g.~in DCAT \\
+Legal          & Different licenses for different parts 
+  & Each part as separate FDO can have separate license 
+  & DCAT, RO-Crate, Named graphs for splitting metadata \\
+Legal          & Mark expired/inexistent copyright 
+  & Undefined 
+  & Unclear, semantics assume copyright valid \\
+Legal          & Mark orphaned data 
+  & Tombstone for deleted data, but no owner of DOIP server means FDO disappears 
+  & Frequently data and endpoint has no known maintainer, archiving in common repositories becoming common \\
+Legal          & List recommended licenses 
+  & Undefined 
+  & Best practice recommendations \\
+Legal          & Track license evolution for dataset 
+  & Undefined 
+  & Versioning with PAV/PROV/DCAT \\
+Legal          & Policy/guidance for patent/trade secrets violation 
+  & Undefined 
+  & Undefined, legal owner may be specified \\
+Legal          & GDPR compliance for personal data 
+  & Undefined 
+  & Undefined \\
+Legal          & Restrict access/use if legally required 
+  & By transport protocol (undefined by FDO/DOIP) 
+  & Diverging approaches, typically landing pages w/ auth\&auth or click-thru \\
+Legal          & Harmonised terms-of-use 
+  & Undefined 
+  & Undefined \\
+Legal          & Alignment between EOSC and national legislation 
+  & Not applicable 
+  & Not applicable \\
 \bottomrule
 \end{longtable}
+\end{landscape}

### 2023-05-18 latex/main.tex
more on EOSC IF

@@
 
-In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data. For the later, we do so by looking through the lens of its development overtime.
+In the following, we discuss the related work with respect to FAIR Digital Objects and Linked Data.
+We do so by looking through the lens of development of these technologies over time, including future directions.
 
@@
 This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \autocite{weigelRDARecommendationPID2018}.
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record}\autocite{islam_2023}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
+This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record} \autocite{islam_2023}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
 
@@
 ResourceSync \autocite{ResourceSyncFrameworkSpecification} is suggested as one type of service to list FDOs.
-This document highlights potential practices by repositories and their clients, but adds no particular requirements (e.g.~how should failed upload checks be reported?).
+This document highlights potential practices by repositories and their clients, without adding any particular requirements.
 
@@
 
-It is worth pointing out at that, except for the DOIP endorsement, all of these documents are abstract, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
+It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
 
@@
 
-The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \autocite{berners-leeWeavingWebOriginal1999}, at a time the Web had been widely established for information exchange, as a global set of hypermedia documents that eare cross-related using universal links in the form of URLs.
+The \textbf{Semantic Web} was developed as a vision by Tim Berners-Lee \autocite{berners-leeWeavingWebOriginal1999}, at a time that the Web had already become widely established for information exchange, being a global set of hypermedia documents which are cross-related using universal links in the form of URLs.
 The foundations of the Web (e.g.~URLs, HTTP, SSL/TLS, HTML, CSS, ECMAScript/JavaScript, media types) were standardised by \href{https://www.w3.org/standards/}{W3C}, \href{https://www.ecma-international.org/}{Ecma}, \href{https://www.ietf.org/standards/}{IETF} and later \href{https://whatwg.org/}{WHATWG}.
@@
 RDF 1.1 formats use Unicode to support IRIs \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
-By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \autocite{sauermannCoolURIsSemantic2011}.}, the Semantic Web builds a self-described system of types and properties, the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
+By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \autocite{sauermannCoolURIsSemantic2011}.}, the Semantic Web builds a self-described system of types and properties, where the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
+By applying these principles as well to any kind of resource that could be described at a URL, this then forms a global distributed Semantic Web.
 
@@
 Vocabularies themselves were formalised using RDFS or simply as human-readable HTML web pages defining each term.
-The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) had a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource.
-By using \emph{content negotiation} in HTTP, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs.
+The main approach of this \emph{Web of Data} was that a URI identified a \emph{resource} (e.g.~an author) with a HTML \emph{representation} for human readers, along with a RDF representation for machine-readable data of the same resource.
+By using \emph{content negotiation} in HTTP\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/HTTP/Content_negotiation}}, the same identifier could be used in both views, avoiding \texttt{index.html} vs \texttt{index.rdf} exposure in the URLs.
 The concept of \emph{namespaces} gave a way to give a group of RDF resources with the same URI base from a Semantic Web-aware service a common \emph{prefix}, avoiding repeated long URLs.
 
-The mid-2000s saw a large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
+The mid-2000s saw large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL \autocite{w3-owl2-overview}, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
 More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
@@
 
-The move towards \emph{open science} data sharing practices from the late 2000s encouraged knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
+The move towards \emph{Open Science} data sharing practices did from the late 2000s encourage knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
 This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \autocite{grothAPIcentricLinkedData2014b}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
@@
 This meant that links stayed mainly within a single information system, with little URI reuse even with large term overlaps \autocite{kamdarSystematicAnalysisTerm2017a}.
-Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \autocite{kleinScholarlyContextNot2014a}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately they do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \autocite{polleresMoreDecentralizedVision2020a}.
+Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \autocite{kleinScholarlyContextNot2014a}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \autocite{polleresMoreDecentralizedVision2020a}.
 Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), hash vs slash, HTTP status codes and PID redirection strategies \autocite{sauermannCoolURIsSemantic2011}.
@@
 Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} (Open Graph \autocite{OpenGraphProtocol}, RDFa \autocite{w3-rdfa-primer}, Microdata \autocite{HTMLStandard}) we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
-JSON-LD is as of 2022-05-13 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 44.4\% of the top 10 million websites \autocite{UsageStatisticsJSONLD}.
+In addition to several such embedded \emph{microformats} \autocite{OpenGraphProtocol,w3-rdfa-primer,HTMLStandard} or we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
+JSON-LD is as of 2023-05-18 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 45\% of the top 10 million websites \autocite{UsageStatisticsJSONLD}.
 
-Recently there has been a renewed emphasis to improve the \emph{Developer Experience} \autocite{DesigningLinkedData2018} for consumption of Linked Data, for instance RDF Shapes (expressed in SHACL \autocite{w3-shacl} or ShEx \autocite{ShapeExpressionsShEx}) can be used to validate RDF Data \autocite{gayoValidatingRDFData2017a,thorntonUsingShapeExpressions2019a} before consuming it programmatically, or reshaping data to fit other models.
-While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying technologies, which hampers adaption by non-LD experts \autocite{klimekSurveyToolsLinked2019a}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
+Recently there has been a renewed emphasis to improve the \emph{Developer Experience} \autocite{DesigningLinkedData2018} for consumption of Linked Data, for instance RDF Shapes -- expressed in SHACL \autocite{w3-shacl} or ShEx \autocite{ShapeExpressionsShEx} -- can be used to validate RDF Data \autocite{gayoValidatingRDFData2017a,thorntonUsingShapeExpressions2019a} before consuming it programmatically, or reshaping data to fit other models.
+While a varied set of tools for Linked Data consumptions have been identified, most of them still require developers to gain significant knowledge of the underlying Semantic Web technologies, which hampers adaption by non-LD experts \autocite{klimekSurveyToolsLinked2019a}, which then tend to prefer non-semantic two-dimensional formats such as CSV files.
 
@@
 
-\subsection*{Comparing FDO and existing approaches}\label{sec:comparing}
+Our main motivation for this article is to investigate how the promises of FAIR Digital Objects may differ from the learnt experiences of Linked Data and the Web.
+We also reflect back from FDO's motivation of machine-actionability to consider the Web as a distributed computational system.
 
@@
 \item
-  An Interoperability Framework and Distributed Platform for Fast Data Applications \autocite{delgadoInteroperabilityFrameworkDistributed2016a}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view
+  An Interoperability Framework and Distributed Platform for Fast Data Applications \autocite{delgadoInteroperabilityFrameworkDistributed2016a}, which proposes quality measurements for comparing how frameworks support interoperability, particularly from a service architectural view.
 \item
-  The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject}, validated against its implementations for completeness.
+  The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject}, validated against its current implementations for completeness.
 \item
-  A Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods
+  A Comparison Framework for Middleware Infrastructures \autocite{zarrasComparisonFrameworkMiddleware2004a}, which suggest dimensions like openness, performance and transparency, mainly focused on remote computational methods.
 \item
-  Cross-checks against RDA's FAIR Data Maturity Model \autocite{bahimFAIRDataMaturity2020a} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness
+  Cross-checks against RDA's FAIR Data Maturity Model \autocite{bahimFAIRDataMaturity2020a} to find how the FAIR principles are achieved in FDO, in particular considering access, sharing and openness.
 \item
-  EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective
+  EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} which gives recommendations for technical, semantic, organisational and legal interoperability, particularly from a metadata perspective.
 \end{enumerate}
@@
 The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (\vpageref*{sec:next-step-fdo}).
+We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref*{sec:next-step-fdo}).
 
 Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data in general.
+For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data in general practice.
 
-\section*{Methods}
 \subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}
@@
 
-Based on the analysis shown in Table \vref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
+Based on the analysis shown in Table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
 
@@
 Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the plethora of documentation and broad developer usage that is available for Web APIs can be developed for FDO.
+There is a question as to whether the extebsuve documentation and broad developer usage that is available for Web APIs could potentially be utillised for FDO.
 
@@
 Whereas the Web allows many different syntactic formats (e.g.~from HTML to XML, PDFs), FDO realised with DOIP requires JSON.
-On the semantic front, FDO requires that every object have a well-defined type and structured form.
+On the semantic front, FDO mandates that every object have a well-defined type and structured form.
 This is clearly not the case on the Web.
@@
 The Web standards bodies (e.g.~IETF and W3C) follow the OpenStand principles \autocite{ModernStandardsParadigm} to embrace openness, transparency, and broad consensus.
-In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first FDO conference \autocite{looFirstInternationalConference2022}.
+In contrast, FDO has a small number of implementations and corresponding protocols, although with a growing community, as evidenced at the first international FDO conference \autocite{looFirstInternationalConference2022}.
 This is not to say that it is not worth developing further Handle+DOIP implementations in the future, but we note that the current FDO functionality can easily be implemented using Web technologies, even as DOIP-over-HTTP \autocite{DOIPAPIHTTPa}.
@@
 
-\textbf{TODO}: Introduce EOSC IF
+The European Open Science Cloud (EOSC) is a large EU initiative to promote Open Science by implementing a joint research infrastructure by federating existing and new services and focusing on interoperability, accessability, best practices as well as technical infrastructure \autocite{10.2777/940154}.
+The EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} details the principles for creating a common way to achieve interoperability between all digital aspects of research activities in EOSC, including data, protocols and software.
+The recommendations are realized through 4 layers, Technical (e.g.
+protocols), Semantic (e.g.
+metadata models), Organisational (e.g.
+recommendations) and Legal (e.g.
+agreements), with a particular aim to address the FAIR interoperability principles and building on the concept of FAIR Digital Objects.
 
-The EOSC Interoperability Framework \autocite[][section 3.6]{corchoEOSCInteroperabilityFramework2021b} recommends:
+In Table \vref{tbl:eosc} we review the EOSC Interoperability Framework (EOSC IF) recommendations, and evaluate to what extent they are addressed by the principles of FDO and Linked Data or their common implementations.
 
-Observations:
+Firstly, we observe that the EOSC IF recommendations are at a high level, mainly affecting governance and practices by communities.
+This \emph{Organizational} level is also highlighted by the FDO recommendations, for instance the FDO Typing \autocite{fdo-TypingFDOs} propose a governance structure to recognize community-endorsed services.
+While these community aspects are not mandated by Linked Data practices, best practices have become established for aspects like ontology development \autocite{10.1186/s13326-021-00240-6}.
+EOSC IF's technical layer is likewise at a architecturally high level, such as service-level agreements, but also highlight PID policies which is strongly required by FDO, while Linked Data communities choose PID practices separately.
+The recommendations for the Semantic layer, is largely already implemented by Linked Data practices, yet for FDO mostly consist of encouragements.
+For instance \emph{clear definitions of semantic concepts} is required by FDO guidelines, but how to technically define them has not been formalised by FDO specifications.
 
-\begin{itemize}
-\tightlist
-\item
-  The recommendations from EOSC IF are at a higher level that mainly affect governance and practices by communities
-\item
-  Technical aspects highlighted by EOSC IF
-\item
-  Search/indexing is important FAIR aspect for Findability, but is poorly supported by current FDO and Linked Data.
-There is a strong role for organizations like EOSC to provide broader registries than more specialised metadata federations like OpenAIRE.
-\item
-  FDO principles have strong recommendations for community development of organisational aspects.
-\item
-  Both FDO and LD are weak on legal aspects like licensing, privacy and usage policies -- these are essential for cross-institutional and cross-repository access of FAIR objects
-\end{itemize}
+The Legal layer of interoperability is perhaps the one most emphasised by EOSC, by enabling collaboration across organizational barriers to joinly build a research infrastructure, but this is an area that both FDO and Linked Data are relatively weak in directly supporting.
+The EOSC IF recommendations in this layer are still largely related to governance practices and metadata, for instance licensing, privacy and usage policies; yet these are essential for cross-institutional and cross-repository access of FAIR objects.
+Likewise, search and indexing is important FAIR aspect for Findability, but is poorly supported globally by FDO and Linked Data.
+Efforts such as Open Research Knowledge Graph (ORKG) \autocite{10.1007/978-3-030-30760-8_31}, DataCite's PID Graph \autocite{10.5438/jwvf-8a66} and Google Knowledge Graph \autocite{singhal2012} have improved programmatic findability to some degree, however not significantly for domain-specific semantic artefacts, currently scattered across multiple semantic catalogues \autocite{10.48550/arXiv.2305.06746}.
+There is a strong role for organizations like EOSC to provide such broader registries, moving beyond scholarly output metadata federations.
+The EOSC Marketplace\footnote{\url{https://marketplace.eosc-portal.eu/}} has for instance recently been expanded to include training material, software and data sources.
 

### 2023-05-18 manuscript.tex
more on EOSC IF [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/d023a0b7859ad2f63bd2b2fe25943b6c268c9733.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/d023a0b7859ad2f63bd2b2fe25943b6c268c9733/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5016916629

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/55b211690e436edabc36739a1bca1546f2b6d680/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/d023a0b7859ad2f63bd2b2fe25943b6c268c9733/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/55b211690e436edabc36739a1bca1546f2b6d680}{stain/2022-fdo-paper@55b2116}
+from \href{https://github.com/stain/2022-fdo-paper/tree/d023a0b7859ad2f63bd2b2fe25943b6c268c9733}{stain/2022-fdo-paper@d023a0b}
 on May 18, 2023.

### 2023-06-12 latex/main.tex
more on conclusion

@@
 This is clearly not the case on the Web.
+
 In terms of connectivity and the deployment of applications, the Web has a plethora of software, services, and protocols that are widely deployed.
@@
 Here we've used the RDA's FAIR Data Maturity Model \autocite{groupFAIRDataMaturity2020} as it has decomposed the FAIR principles to a structured list of FAIR indicators \autocite{bahimFAIRDataMaturity2020a}, importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \vref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remaining JSON/RDF structures always considered as metadata.
+In our interpretation for Table \vref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remaining JSON or RDF structures always considered as metadata.
 
@@
 \item
-  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
+  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little concrete metadata guidance.
 \item
-  FDO/DOIP are stronger on identifier indicators
+  FDO/DOIP are stronger on identifier indicators, while Linked Data approach for identifiers relies on best practices. 
 \item
@@
 
-\section*{EOSC Interoperability Framework}\label{eosc-interoperability-framework}
+\section*{EOSC Interoperability Framework}\label{sec:eosc-interoperability-framework}
 
@@
 
-\textbf{TODO}
+We have evaluated the FAIR Digital Object concept using multiple frameworks, and contrasted FDO against existing experiences from Linked Data on the Web.
+In this section we discuss the implications of this evaluation, and propose how these two approaches can be better combined.
+
+\subsection*{Framework evaluation}
+
+Having considered FDO and the Web architecture as interoperability frameworks (\vref{sec:interoperability-compare}), we observe that neither are magic bullets, but each bring different aspects of interoperability.
+The Web comes with a large degree of flexibility and openness, however this means interoperability can suffer as services have different APIs and data models, although with common patterns.
+This is also true for Linked Data on the Web, with many overlapping ontologies and frequent inconsistencies in resolution mechanisms; although somewhat alleviated in recent years by schema.org becoming common metadata model for semantic markup inline in Web pages.
+The Web is based on a common HTTP protocol which has remained stable architecturally throughout its 32 years of largely backwards-compatible evolution.
+FDO on the other side sets down multiple rigid rules for identifiers, types, methods etc.
+that are advanterous for interoperability and predictability for FAIR consumption.
+Yet there is a large degree of freedom in how the FDO rules can be implemented by a given community, for instance there is no common metadata model or identifier resolution mechanism, and DOIP is just one possible transport method for FDOs, which itself does not enforce these rules.
+
+When evaluating FDO implementations against the FDO guidelines (\vref{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used.
+Achieving fully interoperable FAIR digital objects would require further convergence on implementation practices, and it is not given that his need to diverge from the established Web architecture.
+
+Considering this, by comparing FDO and Web as middleware (\vref{sec:middleware}) we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g.
+lack of concurrency transparancy.
+Recent updates to HTTP have added many features needed for large-scale usage such as video streaming services (e.g.
+caching, multiplexing, cloud deployments), and having the option to transparantly apply these also to FDOs seems like a strong incentive.
+Many programmatic features are however missing or needing custom extensions in both aspects, such as transactions, asynchronous operations and streaming.
+
+By assessing FDO against the FAIR principles (\vref{sec:fair-compare}) we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
+While there are implementations of each of these in general Linked Data examples, there is no single set of implementation guides that fully realizes the FAIR principles.
+\emph{FAIRification} efforts like the FAIR Cookbook \autocite{faircookbook} and FAIR Implementation Profiles \autocite{FIP} are bringing existing practices together, but there remains a potential role for FDO in giving a coherent set of implementation practices that can practically achieve FAIR.
+Significant effort, also within EOSC, is now moving towards FAIR Assessment \autocite{Devaraju_2021}, which in practice need to make additional assumptions on how FAIR principles are implemented, but these are not always formalized.
+Given that most of the existing FAIR guides and assessment tools are focused on Web and Linked Data, it would be reasonable for FDO to then provide a profile of such implementation choices that achieve best of both worlds.
+
+EOSC has been largely supportive of FDO, FAIR and related services.
+By contrasting the EOSC Interoperability Framework (\vref{sec:eosc-interoperability-framework}) with FDO, we found that there are important dimensions that are not solved at a technical level, but through organization collaboration, legal requirements and building community practices.
+FDO recommendations highlight community aspects, but at the same time the largest FAIR communities in many science domains are already producing and consuming Linked Data.
+Just as the Linked Data community has a challenge in convincing more research fields to use Semantic Web technologies, FDO currently need to build many new communities in areas that have shown interest in this approach (e.g.
+material science).
+It may be advantegous for both these effort to be aligned and jointly promoted under an EOSC umbrella.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Ramnifications of ide
-\item
-  Bullet points per table
-\end{itemize}
-
-\subsection*{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}
+\subsection*{(What does FDO mean for Linked Data?)}\label{sec:what-does-it-mean-for-linked-data}
 
@@
 While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
+Indeed, this may be one reason why the initial FDO effort steered away from Linked Data approaches, but now seems in a danger of opening the many same degrees of freedom within FDO.
 
-We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
+We therefore identify the need for a new explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
 Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
+It is beyond the scope of this work to detail such a profile, but its main priorities could be:
+
+\begin{itemize}
+  \tightlist
+  \item Use HTTP(S) as protocol
+  \item Use URIs as identifiers, with persistent identifier promises
+  \item Provide a consistent identifier resolution that does not require heuristics
+  \item Common core metadata model
+  \item References are always URIs, and should be to other FDOs
+  \item Types, attributes and actions are self-defined by their identifier
+\end{itemize}
 

### 2023-06-12 latex/table5.tex
more on conclusion

@@
     >{\centering\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2}}@{}}
-    \caption[Assessing RDA's FAIR Data Maturity Model against the FDO guidelines]{Assessing RDA's FAIR Data Maturity Model \cite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \cite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \cite{13TcbsZF6}, Linked Data Platform (LDP) \cite{7szz7dwO} and examples from Linked Data practices in general.
+    \caption[Assessing RDA's FAIR Data Maturity Model against the FDO guidelines]{Assessing RDA's FAIR Data Maturity Model \cite{groupFAIRDataMaturity2020,bahimFAIRDataMaturity2020a} (first 2 columns) against the FDO guidelines \cite{boninoFAIRDigitalObject}, FDO implemented with the protocol DOIPv2 \cite{foundationDigitalObjectInterface}, Linked Data Platform (LDP) \cite{FDOFramework} and examples from Linked Data practices in general.
 (--- indicates \emph{Unspecified}, may be possible with additional conventions)

### 2023-06-12 latex/table6.tex
more on conclusion

@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3056}}@{}}
-\caption[Assessing EOSC Interoperability Framework, against FDO \& Linked Data]{Assessing EOSC Interoperability Framework \autocite[section 3.6]{corchoEOSCInteroperabilityFramework2021b} against the FDO guidelines \autocite{RwvirqWg} and Linked Data practices.}
+\caption[Assessing EOSC Interoperability Framework, against FDO \& Linked Data]{Assessing EOSC Interoperability Framework \autocite[section 3.6]{corchoEOSCInteroperabilityFramework2021b} against the FDO guidelines \autocite{boninoFAIRDigitalObject} and Linked Data practices.}
 \label{tbl:eosc}\tabularnewline

### 2023-06-12 manuscript.tex
more on conclusion [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/fff5377281a23bdf35c2012ec242e6571e1a2692.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/fff5377281a23bdf35c2012ec242e6571e1a2692/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5242943619

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/d023a0b7859ad2f63bd2b2fe25943b6c268c9733/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/fff5377281a23bdf35c2012ec242e6571e1a2692/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/d023a0b7859ad2f63bd2b2fe25943b6c268c9733}{stain/2022-fdo-paper@d023a0b}
-on May 18, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/fff5377281a23bdf35c2012ec242e6571e1a2692}{stain/2022-fdo-paper@fff5377}
+on June 12, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 18 May 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 12 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 18 May 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 12 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 18 May 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 12 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 18 May 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 12 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 18 May 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 12 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 18 May 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 18 May 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 12 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 18 May 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 12 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 18 May 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 18 May 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 18 May 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 18 May 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 18 May 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 18 May 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 12 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 18 May 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 12 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 18 May 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 12 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 18 May 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 18 May 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 12 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 18 May 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 12 June 2023b).
 
@@
 Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 18 May 2023).
+\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 18 May 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 12 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 18 May 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 12 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 18 May 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 12 June 2023).
 

### 2023-06-12 latex/main.tex
More on conclusion

@@
 The EOSC IF recommendations in this layer are still largely related to governance practices and metadata, for instance licensing, privacy and usage policies; yet these are essential for cross-institutional and cross-repository access of FAIR objects.
+
 Likewise, search and indexing is important FAIR aspect for Findability, but is poorly supported globally by FDO and Linked Data.
@@
 
-Having considered FDO and the Web architecture as interoperability frameworks (\vref{sec:interoperability-compare}), we observe that neither are magic bullets, but each bring different aspects of interoperability.
+Having considered FDO and the Web architecture as interoperability frameworks (\vref*{sec:interoperability-compare}), we observe that neither are magic bullets, but each bring different aspects of interoperability.
 The Web comes with a large degree of flexibility and openness, however this means interoperability can suffer as services have different APIs and data models, although with common patterns.
@@
 
-When evaluating FDO implementations against the FDO guidelines (\vref{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used.
+When evaluating FDO implementations against the FDO guidelines (\vref*{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used.
 Achieving fully interoperable FAIR digital objects would require further convergence on implementation practices, and it is not given that his need to diverge from the established Web architecture.
 
-Considering this, by comparing FDO and Web as middleware (\vref{sec:middleware}) we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g.
+Considering this, by comparing FDO and Web as middleware (\vref*{sec:middleware}) we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g.
 lack of concurrency transparancy.
@@
 
-By assessing FDO against the FAIR principles (\vref{sec:fair-compare}) we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
+By assessing FDO against the FAIR principles (\vref*{sec:fair-compare}) we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
 While there are implementations of each of these in general Linked Data examples, there is no single set of implementation guides that fully realizes the FAIR principles.
 \emph{FAIRification} efforts like the FAIR Cookbook \autocite{faircookbook} and FAIR Implementation Profiles \autocite{FIP} are bringing existing practices together, but there remains a potential role for FDO in giving a coherent set of implementation practices that can practically achieve FAIR.
-Significant effort, also within EOSC, is now moving towards FAIR Assessment \autocite{Devaraju_2021}, which in practice need to make additional assumptions on how FAIR principles are implemented, but these are not always formalized.
-Given that most of the existing FAIR guides and assessment tools are focused on Web and Linked Data, it would be reasonable for FDO to then provide a profile of such implementation choices that achieve best of both worlds.
+Significant effort, also within EOSC, is now moving towards FAIR metrics \autocite{Devaraju_2021}, which in practice need to make additional assumptions on how FAIR principles are implemented, but these are not always formalized \autocite{10.5281/zenodo.7463421} nor can they be taken to be universally correct \autocite{10.5281/zenodo.7848102}.
+Given that most of the existing FAIR guides and assessment tools are focused on Web and Linked Data, it would be reasonable for FDO to then provide a profile of such implementation choices that can achieve best of both worlds.
 
 EOSC has been largely supportive of FDO, FAIR and related services.
-By contrasting the EOSC Interoperability Framework (\vref{sec:eosc-interoperability-framework}) with FDO, we found that there are important dimensions that are not solved at a technical level, but through organization collaboration, legal requirements and building community practices.
+By contrasting the EOSC Interoperability Framework (\vref*{sec:eosc-interoperability-framework}) with FDO, we found that there are important dimensions that are not solved at a technical level, but through organization collaboration, legal requirements and building community practices.
 FDO recommendations highlight community aspects, but at the same time the largest FAIR communities in many science domains are already producing and consuming Linked Data.
-Just as the Linked Data community has a challenge in convincing more research fields to use Semantic Web technologies, FDO currently need to build many new communities in areas that have shown interest in this approach (e.g.
+Just as the Linked Data community has a challenge in convincing more research fields to use Semantic Web technologies, FDO currently need to build many new communities in areas that have shown interest in that approach (e.g.
 material science).
-It may be advantegous for both these effort to be aligned and jointly promoted under an EOSC umbrella.
+It may be advantegous for both these effort to be aligned and jointly promoted under the EOSC umbrella.
 
@@
 
-\subsection*{(What does FDO mean for Linked Data?)}\label{sec:what-does-it-mean-for-linked-data}
+\subsection*{What does FDO mean for Linked Data?}\label{sec:what-does-it-mean-for-linked-data}
 
 The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite[][slide 10]{SemanticWebXML2000}.
-However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
+At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite[][slide 10]{SemanticWebXML2000}, as we have previously proposed \autocite{10.3897/rio.8.e94501}.
+However, this deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed can make it difficult to achieve the FDO goals without significant effort.
 
@@
   \item Use URIs as identifiers, with persistent identifier promises
-  \item Provide a consistent identifier resolution that does not require heuristics
+  \item Provide consistent identifier resolution that does not require heuristics
   \item Common core metadata model
-  \item References are always URIs, and should be to other FDOs
+  \item References are always URIs, and should be persistent identifiers
   \item Types, attributes and actions are self-defined by their identifier
 \end{itemize}
+
+The FAIR and Linked Data communities likewise need to recognize the need for simpler, more pragmatic approaches that make it easier for FAIR practitioners to adapt the technologies with "just enough" semantics.
+We have previously proposed the combination of RO-Crate \autocite{10.3233/ds-210053} and Signposting \autocite{vandesompelFAIRSignpostingProfile2022} as a mean to implement FDO \autocite{10.3897/rio.8.e93937} over HTTP using a common Linked Data metadata model.
+However it may be sufficient to use HTTP-based FAIR Signposting alone to achieve the above list, if one considers only a small metadata model, and rather reference from the signposting which metadata resources are additionally available.
+This will allow any Linked Data resource to gradually participate in the FDO ecosystem, with minimal effort and non-intrusive implementation changes.
+Likewise, DOIP implementations like Cordra typically already use HTTP APIs \autocite{DOIPAPIHTTPa}, which can be augmented with signposting headers without necessarily moving to a Linked Data metadata model.
 

### 2023-06-12 latex/main.tex
we have a conclusion!

@@
 It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
+See bibliography \vref{sec:fdo-bibliography} for the citation per document above.
 
@@
 Achieving fully interoperable FAIR digital objects would require further convergence on implementation practices, and it is not given that his need to diverge from the established Web architecture.
+It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional restrictions, equably implementable, such as using persistent identifiers or pre-defining an object's type.
 
@@
 This will allow any Linked Data resource to gradually participate in the FDO ecosystem, with minimal effort and non-intrusive implementation changes.
-Likewise, DOIP implementations like Cordra typically already use HTTP APIs \autocite{DOIPAPIHTTPa}, which can be augmented with signposting headers without necessarily moving to a Linked Data metadata model.
+FDO implementations like Cordra typically already use HTTP APIs that align with DOIP \autocite{DOIPAPIHTTPa}, these can be augmented with signposting headers without necessarily moving to a Linked Data metadata model.
 
@@
 
+In this work we have considered FAIR Digital Objects (FDO) as a potential distributed object system and compared FDO with established Web approaches like Linked Data.
+We have described the background of the Semantic Web and FAIR Digital Objects, and evaluated both using multiple conceptual frameworks.
+We find that both directions can significantly benefit from each-other and should be aligned further.
+Namely Linked Data proponents need to make their technologies more approachable, agreeing on predictable and consistent implementations of FAIR principles.
+The FDO recommendations show that FAIR thinking in this regard need to move beyond data publishing and into machine actionability across digital objects, and with broader community consensus.
+While flexibility is a necessary ingredient alongside rigidity, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as the early Semantic Web adopters.
+
 \section*{Acknowledgments}
 
-This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreement numbers H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreement numbers HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s  \emph{Horizon Europe funding guarantee} grant numbers 10038963 (EuroScienceGateway), 10038930 (BioDT).
+This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).
 
@@
 
-\defbibheading{fdobibliography}{\subsection*{FDO Specifications}}
+\defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}
 \printshorthands[heading=fdobibliography]

### 2023-06-12 manuscript.tex
we have a conclusion! [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/36ab554e83d48c8aa533b66d262946c219fd6997.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/36ab554e83d48c8aa533b66d262946c219fd6997/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5244663917

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/fff5377281a23bdf35c2012ec242e6571e1a2692/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/36ab554e83d48c8aa533b66d262946c219fd6997/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/fff5377281a23bdf35c2012ec242e6571e1a2692}{stain/2022-fdo-paper@fff5377}
+from \href{https://github.com/stain/2022-fdo-paper/tree/36ab554e83d48c8aa533b66d262946c219fd6997}{stain/2022-fdo-paper@36ab554}
 on June 12, 2023.

### 2023-06-12 latex/main.tex
title, abstract etc.

@@
 
-\title{Evaluating FAIR Digital Object as a distributed object system}
+\title{Evaluating FAIR Digital Object and Linked Data as distributed object systems}
 
@@
 
-% \keywords{Keyword1, Keyword2, Keyword3}
+%\keywords{FAIR, FDO, digital object, Linked Data, EOSC, distributed objects, frameworks, interoperability, middleware}
 
 \begin{abstract}
-FAIR Digital Object is an emerging concept from EOSC.
-This is important.
-Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
-Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
+FAIR Digital Object (FDO) is an emerging concept that is highlighted by European Open Science Cloud (EOSC) as a potential candidate for building a ecosystem of machine-actionable research outputs.
+In this work we systematically evaluate FDO and its implementations as a global distributed object system, by using five different conceptual frameworks that cover interoperability, middleware, FAIR principles, EOSC requirements and FDO guidelines themself.
+We compare the FDO approach with established Linked Data practices and the existing Web architecture, and provide a brief history of the Semantic Web while discussing why these technologies may have been difficult to adopt for FDO purposes.
+We conclude with recommendations for both communities to
 \end{abstract}
@@
 The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages in Google Docs for internal review.
+As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages for internal review.
 As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide their brief summaries below:
@@
 It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-See bibliography \vref{sec:fdo-bibliography} for the citation per document above.
+See bibliography \vref*{sec:fdo-bibliography} for the citation per document above.
 
@@
 In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
+One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
 
@@
 We have previously proposed the combination of RO-Crate \autocite{10.3233/ds-210053} and Signposting \autocite{vandesompelFAIRSignpostingProfile2022} as a mean to implement FDO \autocite{10.3897/rio.8.e93937} over HTTP using a common Linked Data metadata model.
+
 However it may be sufficient to use HTTP-based FAIR Signposting alone to achieve the above list, if one considers only a small metadata model, and rather reference from the signposting which metadata resources are additionally available.
 This will allow any Linked Data resource to gradually participate in the FDO ecosystem, with minimal effort and non-intrusive implementation changes.
-FDO implementations like Cordra typically already use HTTP APIs that align with DOIP \autocite{DOIPAPIHTTPa}, these can be augmented with signposting headers without necessarily moving to a Linked Data metadata model.
+FDO implementations like Cordra typically already use HTTP APIs that align with DOIP \autocite{DOIPAPIHTTPa}, these can be augmented with Signposting headers without necessarily moving to a Linked Data metadata model.
 
@@
 
-In this work we have considered FAIR Digital Objects (FDO) as a potential distributed object system and compared FDO with established Web approaches like Linked Data.
+In this work we have considered FAIR Digital Objects (FDO) as a potential distributed object system and compared FDO with established Web approaches focusing on Linked Data.
 We have described the background of the Semantic Web and FAIR Digital Objects, and evaluated both using multiple conceptual frameworks.
-We find that both directions can significantly benefit from each-other and should be aligned further.
+
+We find that both FDO and Linked Data approaches can significantly benefit from each-other and should be aligned further.
 Namely Linked Data proponents need to make their technologies more approachable, agreeing on predictable and consistent implementations of FAIR principles.
+
 The FDO recommendations show that FAIR thinking in this regard need to move beyond data publishing and into machine actionability across digital objects, and with broader community consensus. 
-While flexibility is a necessary ingredient alongside rigidity, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as the early Semantic Web adopters.
+As flexibility for extensions is a necessary ingredient alongside rigidity for core concepts, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as the early Semantic Web adopters.
+
+By implementing the goals of FAIR Digital Objects with the mature technology stack developed for Linked Data, EOSC research infrastructures and researchers in general can create and use FAIR machine-actionable research outputs for decades to come.
 
@@
 
-This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).
+This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020\-INFRAEOSC\-2018\-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON\-INFRA\-2021\-EMERGENCY\-01 101046203 (BY\-COVID), HORIZON\-INFRA\-2021\-EOSC\-01 101057388 (EuroScienceGateway), HORIZON\-INFRA\-2021\-EOSC\-01\-05 101057344 (FAIR\-IMPACT), HORIZON\-INFRA\-2021\-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR\-IMPACT), 10038930 (BioDT).
 

### 2023-06-12 manuscript.tex
title, abstract etc. [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/fe73d9fbfc49fdc4a9fc913c0ccd95c6b0ef7ab3.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/fe73d9fbfc49fdc4a9fc913c0ccd95c6b0ef7ab3/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5245186527

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/36ab554e83d48c8aa533b66d262946c219fd6997/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/fe73d9fbfc49fdc4a9fc913c0ccd95c6b0ef7ab3/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/36ab554e83d48c8aa533b66d262946c219fd6997}{stain/2022-fdo-paper@36ab554}
+from \href{https://github.com/stain/2022-fdo-paper/tree/fe73d9fbfc49fdc4a9fc913c0ccd95c6b0ef7ab3}{stain/2022-fdo-paper@fe73d9f}
 on June 12, 2023.

### 2023-06-12 latex/main.tex

@@
 We compare the FDO approach with established Linked Data practices and the existing Web architecture, and provide a brief history of the Semantic Web while discussing why these technologies may have been difficult to adopt for FDO purposes.
-We conclude with recommendations for both communities to
+We conclude with recommendations for both Linked Data and FDO communities to further their adaptation and alignment.
 \end{abstract}
@@
 The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
-This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced in 1995 \autocite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the ITU X.1255 framework \autocite{x1255FrameworkDiscovery} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) \autocite{DigitalObjectInterface}.
+This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced by \textcite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
+DO was the inspiration for the \textcite{x1255FrameworkDiscovery} framework which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) \autocite{DigitalObjectInterface}.
 
@@
 \item
-  Building on the Digital Object concept, using the simplified DOIP v2 specification \autocite{foundationDigitalObjectInterface}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0, see \autocite{DOIPExamplesCordraa}} (usually TCP/IP over TLS).
+  Building on the Digital Object concept, using the simplified \textcite{foundationDigitalObjectInterface} specification, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0, see \textcite{DOIPExamplesCordraa}} (usually TCP/IP over TLS).
 The main DOIP operations are retrieving, creating and updating digital objects.
@@
 In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
+One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~\emph{should a gene identifier signify just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?}) lead to large differences in representational granularity, and thus needed different identifiers.
 
@@
 Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \autocite{w3-skos-primer} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \autocite{huHowMatchableAre2011a} .
+Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \autocite{huHowMatchableAre2011a}.
 
@@
 
-\include{table1}
+\input{table1}
 
@@
 
-\include{table2}
+\input{table2}
 
@@
 
-\include{table6}
+\input{table6}
 
@@
 
-This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020\-INFRAEOSC\-2018\-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON\-INFRA\-2021\-EMERGENCY\-01 101046203 (BY\-COVID), HORIZON\-INFRA\-2021\-EOSC\-01 101057388 (EuroScienceGateway), HORIZON\-INFRA\-2021\-EOSC\-01\-05 101057344 (FAIR\-IMPACT), HORIZON\-INFRA\-2021\-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR\-IMPACT), 10038930 (BioDT).
+\begin{small}
+This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).
+\end{small}
+
+\section*{Author contributions}
+\begin{small}
+
+Contributions to this article according to the CASRAI CRediT taxonomy\footnote{\url{https://credit.niso.org/}}:
+
+\begin{description}
+\tightlist
+\item[Stian Soiland-Reyes]
+Conceptualization, Funding acquisition, Investigation, Methodology,
+Writing -- original draft, Writing -- review \& editing
+\item[Carole Goble]
+Funding acquisition, Supervision, Writing -- review \& editing
+\item[Paul Groth]
+Conceptualization, Methodology, Supervision, Writing -- review \& editing
+\end{description}
+\end{small}
 

### 2023-06-12 latex/table2.tex
table magic
40 pages!

@@
-\begin{longtable}[]{@{}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2923}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2769}}
-  >{\centering\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4308}}@{}}
+\begin{table}[h!]
+\centering
 \caption{Mapping the Metamodel concepts from the Interoperability Framework for Fast Data \autocite{delgadoInteroperabilityFrameworkDistributed2016a} to equivalent concepts for FDO and Web.
 \label{tbl:metamodel-concepts}}\tabularnewline
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endfirsthead
-\toprule
-\begin{minipage}[b]{\linewidth}\centering
-Metamodel concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-FDO/DOIP concept
-\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
-Web/HTTP concept
-\end{minipage} \\
-\midrule
-\endhead
+ \begin{tabular}{ m{5em}  m{15em} m{15em} } 
+ \hline
+Metamodel concept & 
+FDO/DOIP concept & 
+Web/HTTP concept \\ 
+ \hline
 Resource	  & FDO/DO	            & Resource \\
@@
 Response	  & DOIP Response	      & Response \\
-Message & Segment, \texttt{requestId} & Message, Representation \\
-Channel & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures. & TCP/IP, TLS, UDP \\
+Message	    & Segment, \texttt{requestId} 
+                                  & Message, Representation \\
+Channel	    & DOIP Transport protocol (e.g.~TCP/IP, TLS). JSWS signatures.
+                                  & TCP/IP, TLS, UDP \\
 Protocol  	& DOIP 2.0, ++	      & HTTP/1.1, HTTP/2, HTTP/3 \\
 Link	      & PID/Handle	        & URL \\ 
-\bottomrule
-\end{longtable}
+ \hline
+ \end{tabular}\end{table}

### 2023-06-12 manuscript.tex
40 pages! [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/76ebc8f6ebdf030e2b62b590e21331952649784f.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/76ebc8f6ebdf030e2b62b590e21331952649784f/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5248371671

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/fe73d9fbfc49fdc4a9fc913c0ccd95c6b0ef7ab3/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/76ebc8f6ebdf030e2b62b590e21331952649784f/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/fe73d9fbfc49fdc4a9fc913c0ccd95c6b0ef7ab3}{stain/2022-fdo-paper@fe73d9f}
+from \href{https://github.com/stain/2022-fdo-paper/tree/76ebc8f6ebdf030e2b62b590e21331952649784f}{stain/2022-fdo-paper@76ebc8f}
 on June 12, 2023.

### 2023-06-12 manuscript.tex
Merge branch 'main' of github.com:stain/2022-fdo-paper into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/d1c36597d366bf2fb734ce61283e9810ef5c22ee.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/d1c36597d366bf2fb734ce61283e9810ef5c22ee/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5248604853

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/76ebc8f6ebdf030e2b62b590e21331952649784f/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/d1c36597d366bf2fb734ce61283e9810ef5c22ee/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/76ebc8f6ebdf030e2b62b590e21331952649784f}{stain/2022-fdo-paper@76ebc8f}
+from \href{https://github.com/stain/2022-fdo-paper/tree/d1c36597d366bf2fb734ce61283e9810ef5c22ee}{stain/2022-fdo-paper@d1c3659}
 on June 12, 2023.
@@
 
-\textbf{TODO}: Rewrite from notes below to actual abstract.
-
-FAIR Digital Object is an emerging concept from EOSC.
-This is important.
-Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
-Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
-
-What is the message of this paper?
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  These are the overlaps
-\item
-  This is what FDO is requiring but not commonly deployed
-\item
-  DOI indirection is emphasized. It is used in SW, but the idea for stability. Embrace it!
-\item
-  PROV had the idea of using indirection to help us represent provenance of objects - digital twin
-\item
-  Lessons for Semantic Web community. What is missing in FDO and in SW.
-\item
-  Contribution is also about thinking about Semantic Web as different architectural levels. Interoperability level, Middleware level. Governance level. Data level.
-\item
-  Go back to the old SW layer cake. Explains why we picked these frameworks.
-\item
-  Lessons for SW: Parts of the stack that is less. FDO contributions.
-\end{enumerate}
-
-Semantic Web in a way already implements FDO, but other things that SW perhaps should drop in emphasis to better support FDO and FAIR vision. More about indirection, visibility.
-
-Have all the ingredients, but not cooking properly.
-
-Systematic through frameworks.
-
-Emerging stack - how does it compare to what we've already done? What are the implications for our design and research? What new technology is needed?
+FAIR Digital Object (FDO) is an emerging concept that is highlighted by European Open Science Cloud (EOSC) as a potential candidate for building a ecosystem of machine-actionable research outputs.
+In this work we systematically evaluate FDO and its implementations as a global distributed object system, by using five different conceptual frameworks that cover interoperability, middleware, FAIR principles, EOSC requirements and FDO guidelines themself.
+We compare the FDO approach with established Linked Data practices and the existing Web architecture, and provide a brief history of the Semantic Web while discussing why these technologies may have been difficult to adopt for FDO purposes.
+We conclude with recommendations for both communities to further their adaptation and alignment.
 
@@
 Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully (\protect\hyperlink{ref-jHwttlTs}{Verborgh \& Vander Sande, 2020}), opting instead for custom in-house models and JSON-derived formats from RESTful Web services (\protect\hyperlink{ref-99MS5xE0}{Meroño-Peñuela, Lisena \& Martínez-Ortiz, 2021a}; \protect\hyperlink{ref-J2H8yssV}{Neumann, Laranjeiro \& Bernardino, 2021}).
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022}).
+While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability (\protect\hyperlink{ref-DNTCUjpC}{Turcoane, 2014}; \protect\hyperlink{ref-Pa67pUtR}{Wilkinson et al., 2022a}).
 
@@
 
-\textbf{TODO}
+Discussion
+
+We have evaluated the FAIR Digital Object concept using multiple
+frameworks, and contrasted FDO against existing experiences from Linked
+Data on the Web. In this section we discuss the implications of this
+evaluation, and propose how these two approaches can be better combined.
+
+\hypertarget{framework-evaluation}{%
+\section*{Framework evaluation}\label{framework-evaluation}}
+\addcontentsline{toc}{section}{Framework evaluation}
+
+Having considered FDO and the Web architecture as interoperability
+frameworks
+(\protect\hyperlink{sec:interoperability-compare}{{[}sec:interoperability-compare{]}}), we observe that neither are
+magic bullets, but each bring different aspects of interoperability. The
+Web comes with a large degree of flexibility and openness, however this
+means interoperability can suffer as services have different APIs and
+data models, although with common patterns. This is also true for Linked
+Data on the Web, with many overlapping ontologies and frequent
+inconsistencies in resolution mechanisms; although somewhat alleviated
+in recent years by schema.org becoming common metadata model for
+semantic markup inline in Web pages. The Web is based on a common HTTP
+protocol which has remained stable architecturally throughout its 32
+years of largely backwards-compatible evolution. FDO on the other side
+sets down multiple rigid rules for identifiers, types, methods etc. that
+are advanterous for interoperability and predictability for FAIR
+consumption. Yet there is a large degree of freedom in how the FDO rules
+can be implemented by a given community, for instance there is no common
+metadata model or identifier resolution mechanism, and DOIP is just one
+possible transport method for FDOs, which itself does not enforce these
+rules.
+
+When evaluating FDO implementations against the FDO guidelines
+(\protect\hyperlink{sec:doip-fdo-compare}{{[}sec:doip-fdo-compare{]}}) we see that several technical pieces
+and community practices still need to be developed and further defined,
+for instance the FDO type system, how to declare FDO actions, how to
+resolve persistent identifiers, or how to know which pattern of FDO
+composition is used. Achieving fully interoperable FAIR digital objects
+would require further convergence on implementation practices, and it is
+not given that his need to diverge from the established Web
+architecture. It is not clear from FDO guidelines if moving from
+HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects
+will benefit FAIR practitioners, when both approaches require additional
+restrictions, equably implementable, such as using persistent
+identifiers or pre-defining an object's type.
+
+Considering this, by comparing FDO and Web as middleware
+(\protect\hyperlink{sec:middleware}{{[}sec:middleware{]}}) we saw that programmatic access to digital
+objects, a core promise of FDO, is not particularly improved by the use
+of the protocol DOIP as compared to HTTP, e.g.~lack of concurrency
+transparancy. Recent updates to HTTP have added many features needed for
+large-scale usage such as video streaming services (e.g.~caching,
+multiplexing, cloud deployments), and having the option to transparantly
+apply these also to FDOs seems like a strong incentive. Many
+programmatic features are however missing or needing custom extensions
+in both aspects, such as transactions, asynchronous operations and
+streaming.
+
+By assessing FDO against the FAIR principles
+(\protect\hyperlink{sec:fair-compare}{{[}sec:fair-compare{]}}) we found that both FDO implementations
+are underspecified in several aspects (licences, provenance, data
+references, data vocabularies, metadata persistence). While there are
+implementations of each of these in general Linked Data examples, there
+is no single set of implementation guides that fully realizes the FAIR
+principles. \emph{FAIRification} efforts like the FAIR Cookbook
+(\protect\hyperlink{ref-faircookbook}{\textbf{faircookbook?}}) and FAIR Implementation Profiles (\protect\hyperlink{ref-FIP}{\textbf{FIP?}}) are bringing
+existing practices together, but there remains a potential role for FDO
+in giving a coherent set of implementation practices that can
+practically achieve FAIR. Significant effort, also within EOSC, is now
+moving towards FAIR metrics (\protect\hyperlink{ref-Devaraju_2021}{\textbf{Devaraju\_2021?}}), which in practice need to
+make additional assumptions on how FAIR principles are implemented, but
+these are not always formalized (\protect\hyperlink{ref-HkdfgHs}{Wilkinson et al., 2022b}) nor can they
+be taken to be universally correct (\protect\hyperlink{ref-JazFAhMJ}{Verburg et al., 2023}). Given that
+most of the existing FAIR guides and assessment tools are focused on Web
+and Linked Data, it would be reasonable for FDO to then provide a
+profile of such implementation choices that can achieve best of both
+worlds.
+
+EOSC has been largely supportive of FDO, FAIR and related services. By
+contrasting the EOSC Interoperability Framework
+(\protect\hyperlink{sec:eosc-interoperability-framework}{{[}sec:eosc-interoperability-framework{]}}) with FDO, we found
+that there are important dimensions that are not solved at a technical
+level, but through organization collaboration, legal requirements and
+building community practices. FDO recommendations highlight community
+aspects, but at the same time the largest FAIR communities in many
+science domains are already producing and consuming Linked Data. Just as
+the Linked Data community has a challenge in convincing more research
+fields to use Semantic Web technologies, FDO currently need to build
+many new communities in areas that have shown interest in that approach
+(e.g.~material science). It may be advantegous for both these effort to
+be aligned and jointly promoted under the EOSC umbrella.
+
+\hypertarget{sec:what-does-it-mean-for-linked-data}{%
+\section*{What does FDO mean for Linked Data?}\label{sec:what-does-it-mean-for-linked-data}}
+\addcontentsline{toc}{section}{What does FDO mean for Linked Data?}
+
+The FAIR Digital Object approach raises many important points for Linked
+Data practictioners. At first glance, the explicit requirements of FDOs
+may seem to be easy to furfill by different parts of the Semantic Web
+Cake (\protect\hyperlink{ref-SemanticWebXML2000}{\textbf{SemanticWebXML2000?}} slide 10), as we have previously proposed
+(\protect\hyperlink{ref-eBBPb4Yf}{Soiland-Reyes et al., 2022a}). However, this deeper investigation, based on
+multiple frameworks, highlights that the openness and variability of how
+Linked Data is deployed can make it difficult to achieve the FDO goals
+without significant effort.
+
+While RDF and Linked Data have been suggested as prime candidates for
+making FAIR data, we argue that when different developers have too many
+degrees of freedom (such as serialization formats, vocabularies,
+identifiers, navigation), interoperability is hampered -- this makes it
+hard for machines to reliably consume multiple FAIR resources across
+repositories and data providers. Indeed, this may be one reason why the
+initial FDO effort steered away from Linked Data approaches, but now
+seems in a danger of opening the many same degrees of freedom within
+FDO.
+
+We therefore identify the need for a new explicit FDO profile of Linked
+Data that sets pragmatic constraints and stronger recommendations for
+consistent and developer-friendly deployment of digital objects. Such a
+combination of efforts will utillise both the benefits of mature
+Semantic Web technologies (e.g.~federated knowledge graph queries and
+rich validation) and data management practices that follow FDO guidance
+in order to grow a rigid (yet flexible) ecosystem of machine-actionable
+scholarly objects. It is beyond the scope of this work to detail such a
+profile, but its main priorities could be:
 
@@
 \item
-  Ramnifications of ide
-\item
-  Finish tables
-\item
-  Bullet points per table
-\item
-  Send in Google Docs
+  Use HTTP(S) as protocol
+\item
+  Use URIs as identifiers, with persistent identifier promises
+\item
+  Provide consistent identifier resolution that does not require
+  heuristics
+\item
+  Common core metadata model
+\item
+  References are always URIs, and should be persistent identifiers
+\item
+  Types, attributes and actions are self-defined by their identifier
 \end{itemize}
 
-\hypertarget{what-does-it-mean-for-linked-data}{%
-\subsection{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}}
-
-The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake (\protect\hyperlink{ref-kYjzRJ0H}{{`Semantic Web - XML2000 - slide "Architecture"'}}).
-However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
-
-While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
-
-We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
+The FAIR and Linked Data communities likewise need to recognize the need
+for simpler, more pragmatic approaches that make it easier for FAIR
+practitioners to adapt the technologies with "just enough" semantics.
+We have previously proposed the combination of RO-Crate
+(\protect\hyperlink{ref-oCW2bFIv}{Soiland-Reyes et al., 2022c}) and Signposting
+(\protect\hyperlink{ref-vandesompelFAIRSignpostingProfile2022}{\textbf{vandesompelFAIRSignpostingProfile2022?}}) as a mean to implement FDO
+(\protect\hyperlink{ref-1GD9fs68i}{Soiland-Reyes et al., 2022b}) over HTTP using a common Linked Data metadata
+model.
+
+However it may be sufficient to use HTTP-based FAIR Signposting alone to
+achieve the above list, if one considers only a small metadata model,
+and rather reference from the signposting which metadata resources are
+additionally available. This will allow any Linked Data resource to
+gradually participate in the FDO ecosystem, with minimal effort and
+non-intrusive implementation changes. FDO implementations like Cordra
+typically already use HTTP APIs that align with DOIP (\protect\hyperlink{ref-DOIPAPIHTTPa}{\textbf{DOIPAPIHTTPa?}}),
+these can be augmented with Signposting headers without necessarily
+moving to a Linked Data metadata model.
 
@@
 
-\ldots{}
+In this work we have considered FAIR Digital Objects (FDO) as a potential distributed object system and compared FDO with established Web approaches focusing on Linked Data.
+We have described the background of the Semantic Web and FAIR Digital Objects, and evaluated both using multiple conceptual frameworks.
+
+We find that both FDO and Linked Data approaches can significantly benefit from each-other and should be aligned further.
+Namely Linked Data proponents need to make their technologies more approachable, agreeing on predictable and consistent implementations of FAIR principles.
+
+The FDO recommendations show that FAIR thinking in this regard need to move beyond data publishing and into machine actionability across digital objects, and with broader community consensus.
+As flexibility for extensions is a necessary ingredient alongside rigidity for core concepts, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as the early Semantic Web adopters.
+
+By implementing the goals of FAIR Digital Objects with the mature technology stack developed for Linked Data, EOSC research infrastructures and researchers in general can create and use FAIR machine-actionable research outputs for decades to come.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kYjzRJ0H}{}}%
-Semantic Web - XML2000 - slide "Architecture".
-\emph{Available at} \href{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}{\emph{https://www.w3.org/2000/Talks/1206-xml2k-tbl/slide10-0.html}} (accessed 12 June 2023).
-
 \leavevmode\vadjust pre{\hypertarget{ref-YUzogaV}{}}%
@@
 DOI: \href{https://doi.org/10.1038/nbt1346}{10.1038/nbt1346}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-eBBPb4Yf}{}}%
+Soiland-Reyes S, Castro LJ, Garijo D, Portier M, Goble C, Groth P.
+2022a.
+Updating Linked Data practices for FAIR Digital Object principles.
+\emph{Research Ideas and Outcomes} 8.
+DOI: \href{https://doi.org/10.3897/rio.8.e94501}{10.3897/rio.8.e94501}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-1GD9fs68i}{}}%
+Soiland-Reyes S, Sefton P, Castro LJ, Coppens F, Garijo D, Leo S, Portier M, Groth P.
+2022b.
+Creating lightweight FAIR Digital Objects with RO-Crate.
+\emph{Research Ideas and Outcomes} 8.
+DOI: \href{https://doi.org/10.3897/rio.8.e93937}{10.3897/rio.8.e93937}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-oCW2bFIv}{}}%
+Soiland-Reyes S, Sefton P, Crosas M, Castro LJ, Coppens F, Fernández JM, Garijo D, Grüning B, La Rosa M, Leo S, Ó Carragáin E, Portier M, Trisovic A, RO-Crate Community, Groth P, Goble C.
+2022c.
+Packaging research artefacts with RO-Crate.
+\emph{Data Science} 5:97--138.
+DOI: \href{https://doi.org/10.3233/ds-210053}{10.3233/ds-210053}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-JazFAhMJ}{}}%
+Verburg M, Huber R, Jonquet C, Garijo D.
+2023.
+FAIR-IMPACT project response to "FAIR Assessment Tools: Towards an "Apples to Apples" Comparisons".
+DOI: \href{https://doi.org/10.5281/zenodo.7848102}{10.5281/zenodo.7848102}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-11oVTMsuZ}{}}%
@@
 Wilkinson SR, Eisenhauer G, Kapadia AJ, Knight K, Logan J, Widener P, Wolf M.
-2022.
+2022a.
 F*** workflows: when parts of FAIR are missing.
@@
 DOI: \href{https://doi.org/10.48550/arxiv.2209.09022}{10.48550/arxiv.2209.09022}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-HkdfgHs}{}}%
+Wilkinson MD, Sansone S-A, Marjan G, Nordling J, Dennis R, Hecker D.
+2022b.
+FAIR Assessment Tools: Towards an "Apples to Apples" Comparisons.
+DOI: \href{https://doi.org/10.5281/zenodo.7463421}{10.5281/zenodo.7463421}.
 

### 2023-06-13 latex/main.tex

@@
 
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \autocite{w3-rdf11-primer} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \autocite{w3-rdf-syntax} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
+Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \autocite{w3-rdf11-primer} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \autocite{w3-rdf-syntax99} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
 

### 2023-06-13 manuscript.tex
Merge overleaf-2023-06-13-1634 into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/12c36282be545d60358eeed915d0e52941a16774.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/12c36282be545d60358eeed915d0e52941a16774/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5258086952

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/d1c36597d366bf2fb734ce61283e9810ef5c22ee/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/12c36282be545d60358eeed915d0e52941a16774/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/d1c36597d366bf2fb734ce61283e9810ef5c22ee}{stain/2022-fdo-paper@d1c3659}
-on June 12, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/12c36282be545d60358eeed915d0e52941a16774}{stain/2022-fdo-paper@12c3628}
+on June 13, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 12 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 13 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 12 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 13 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 12 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 13 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 12 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 13 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 12 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 13 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 12 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 13 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 12 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 13 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 12 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 13 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 12 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 13 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 12 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 13 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 12 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 13 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 12 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 13 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 12 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 13 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 12 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 13 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 12 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 13 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 12 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 13 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 12 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 13 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 12 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 13 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 12 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 13 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 12 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 13 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 12 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 13 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 12 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 13 June 2023).
 

### 2023-06-13 manuscript.tex
bibliography fixes from thesis [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/d590b21d8a5e9afb5e0df414044044c16f5afcb9.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/d590b21d8a5e9afb5e0df414044044c16f5afcb9/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5261086824

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/12c36282be545d60358eeed915d0e52941a16774/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/d590b21d8a5e9afb5e0df414044044c16f5afcb9/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/12c36282be545d60358eeed915d0e52941a16774}{stain/2022-fdo-paper@12c3628}
+from \href{https://github.com/stain/2022-fdo-paper/tree/d590b21d8a5e9afb5e0df414044044c16f5afcb9}{stain/2022-fdo-paper@d590b21}
 on June 13, 2023.

### 2023-06-14 manuscript.tex
abstract [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/a3cbfec2e6874029aa0f01da4498910caeb6a6a8.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/a3cbfec2e6874029aa0f01da4498910caeb6a6a8/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5269378322

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/d590b21d8a5e9afb5e0df414044044c16f5afcb9/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/a3cbfec2e6874029aa0f01da4498910caeb6a6a8/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/d590b21d8a5e9afb5e0df414044044c16f5afcb9}{stain/2022-fdo-paper@d590b21}
-on June 13, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/a3cbfec2e6874029aa0f01da4498910caeb6a6a8}{stain/2022-fdo-paper@a3cbfec}
+on June 14, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 13 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 14 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 13 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 14 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 13 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 14 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 13 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 14 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 13 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 14 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 13 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 14 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 13 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 14 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 13 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 14 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 13 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 14 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 13 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 14 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 13 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 14 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 13 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 14 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 13 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 14 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 13 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 14 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 13 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 14 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 13 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 14 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 13 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 14 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 13 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 14 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 13 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 14 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 13 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 14 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 13 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 14 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 13 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 14 June 2023).
 

### 2023-06-15 manuscript.tex
author order on RFCs [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/c01187b0544337b5638fa36c73801550ce98a89a.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/c01187b0544337b5638fa36c73801550ce98a89a/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5276855021

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/a3cbfec2e6874029aa0f01da4498910caeb6a6a8/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/c01187b0544337b5638fa36c73801550ce98a89a/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/a3cbfec2e6874029aa0f01da4498910caeb6a6a8}{stain/2022-fdo-paper@a3cbfec}
-on June 14, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/c01187b0544337b5638fa36c73801550ce98a89a}{stain/2022-fdo-paper@c01187b}
+on June 15, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 14 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 15 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 14 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 15 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 14 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 15 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 14 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 15 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 14 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 15 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 14 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 15 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 14 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 15 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 14 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 15 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 14 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 15 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 14 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 15 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 14 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 15 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 14 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 15 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 14 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 15 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 14 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 15 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 14 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 15 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 14 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 15 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 14 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 15 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 14 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 15 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 14 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 15 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 14 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 15 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 14 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 15 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 14 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 15 June 2023).
 

### 2023-06-15 latex/main.tex
abstract
more acknowledgements
more on acknowledgements

@@
 In this work we systematically evaluate FDO and its implementations as a global distributed object system, by using five different conceptual frameworks that cover interoperability, middleware, FAIR principles, EOSC requirements and FDO guidelines themself.
+
 We compare the FDO approach with established Linked Data practices and the existing Web architecture, and provide a brief history of the Semantic Web while discussing why these technologies may have been difficult to adopt for FDO purposes.
@@
 
+We would like to acknowledge the FAIR Digital Object Forum \autocite{FAIRDigitalObjects} community and working groups, where SSR and CG are members.
+
 \section*{Author contributions}
@@
 \item[Stian Soiland-Reyes]
-Conceptualization, Funding acquisition, Investigation, Methodology,
+Conceptualization, Formal Analysis, Funding acquisition, Investigation, Methodology,
 Writing -- original draft, Writing -- review \& editing

### 2023-06-15 latex/main.tex
corporate authors

@@
 It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web \autocite{DOIResolutionDocumentation} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \autocite{DOIHandbookResolution}.
+For example, DOIs are mostly resolved on the web \autocite{DOIHandbookResolution} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \autocite{DOIHandbookResolution}.
 

### 2023-06-15 latex/table1.tex
more bib tweaks.

@@
 \toprule
-\begin{minipage}[b]{\linewidth}\raggedright
-\emph{Quality}
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-FDO w/ DOIP
-\end{minipage}
-  & \begin{minipage}[b]{\linewidth}\centering
-Web w/ Linked Data
-\end{minipage} \\
+\emph{Quality} & 
+FDO w/ DOIP & 
+Web w/ Linked Data \\
 \midrule
@@
 \textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols}
-  & DOIP \autocite{foundationDigitalObjectInterface} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP}
+  & \textcite{foundationDigitalObjectInterface} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP}
   & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\

### 2023-06-15 manuscript.tex
more bib tweaks. [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/f701f88bde71d6a44507f5904a8e4ac2ad6d69ce.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/f701f88bde71d6a44507f5904a8e4ac2ad6d69ce/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5278088301

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/c01187b0544337b5638fa36c73801550ce98a89a/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/f701f88bde71d6a44507f5904a8e4ac2ad6d69ce/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/c01187b0544337b5638fa36c73801550ce98a89a}{stain/2022-fdo-paper@c01187b}
+from \href{https://github.com/stain/2022-fdo-paper/tree/f701f88bde71d6a44507f5904a8e4ac2ad6d69ce}{stain/2022-fdo-paper@f701f88}
 on June 15, 2023.

### 2023-06-16 latex/main.tex
Updates from Overleaf

--- 2023-06-13 latex/main.tex
@@
 \begin{abstract}
-FAIR Digital Object (FDO) is an emerging concept that is highlighted by European Open Science Cloud (EOSC) as a potential candidate for building a ecosystem of machine-actionable research outputs.
+FAIR Digital Objects (FDO) is an emerging concept that is highlighted by European Open Science Cloud (EOSC) as a potential candidate for building a ecosystem of machine-actionable research outputs.
 In this work we systematically evaluate FDO and its implementations as a global distributed object system, by using five different conceptual frameworks that cover interoperability, middleware, FAIR principles, EOSC requirements and FDO guidelines themself.
@@
 The FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been widely recognised by the research community and policy makers as a goal to strive for.
+They have been recognised by the research community and policy makers as a goal to strive for.
 In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \autocite{monsCloudyIncreasinglyFAIR2017b}.
@@
 In other discourse, Linked Data \autocite{bizerLinkedDataStory2009a} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \autocite{boninodasilvasantosFAIRDataPoints2016a,hasnainAssessingFAIRData2018a}.
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \autocite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \autocite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}.
-While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \autocite{turcoaneLinkedDataJSONLD2014a,wilkinsonWorkflowsWhenParts2022b}.
+Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a ``FAIR Linked Data approach'' fully \autocite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \autocite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}.
+While such focus on simplicity allows for rapid development and highly specialised services, it raises wider concerns about interoperability \autocite{turcoaneLinkedDataJSONLD2014a,wilkinsonWorkflowsWhenParts2022b}.
 
@@
 
-These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.
+These approaches - FDO and Linked Data - thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
+Given their importance, in this article, we aim to compare FAIR Digital Objects, with Linked Data and the Web architecture in the context of the discourse around FAIR data.
 
-Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.
+Concretely, the contribution of this paper is {\bf a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks} that capture different perspectives on interoperability and readiness for implementation.
 
@@
 The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
+These are mostly realised using the reference implementation ( \href{https://cordra.org/}{Cordra}).
 FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \autocite{Draftbhuttonjsonschema} and PIDs are assigned using the Handle system.
@@
 \item
-  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
+  Following a Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
 \item
@@
 
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover \vpageref{sec:ld}.
+From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we will cover \vpageref{sec:ld}.
 
@@
 
-The mid-2000s saw large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL \autocite{w3-owl2-overview}, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
+The mid-2000s saw large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL \autocite{w3-owl2-overview}, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm.
+%(e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
+More human-readable syntaxes for RDF such as Turtle evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
 
-Established Semantic Web services and standards include SPARQL \autocite{w3-sparql11-overview} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} \autocite{w3-rdf11-concepts} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
+Established Semantic Web services and standards include: SPARQL \autocite{w3-sparql11-overview} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} \autocite{w3-rdf11-concepts} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and  graph visualisation.
 
@@
 
-The \textbf{Linked Data} concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect back into focus.
-Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
+The \textbf{Linked Data} concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect of the Semantic Web back into focus.
+Crucial toLinked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
 This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
@@
 Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} \autocite{OpenGraphProtocol,w3-rdfa-primer,HTMLStandard} or we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
+In addition to several such embedded \emph{microformats} \autocite{OpenGraphProtocol,w3-rdfa-primer,HTMLStandard}, we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
 JSON-LD is as of 2023-05-18 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 45\% of the top 10 million websites \autocite{UsageStatisticsJSONLD}.
@@
 
+Linked Data provides a technologies that have evolved over time for the purposes of data interoperability.
+The need to embrace the web and developer experience have been central lessons learned.
+In contrast, FDO is a new specification with many different potential paths forward.
+
 \section*{Method}
 
-Our main motivation for this article is to investigate how the promises of FAIR Digital Objects may differ from the learnt experiences of Linked Data and the Web.
-We also reflect back from FDO's motivation of machine-actionability to consider the Web as a distributed computational system.
+Our main motivation for this article is to investigate how FAIR Digital Objects may differ from the learnt experiences of Linked Data and the Web.
+We also aim to reflect back from FDO's motivation of machine-actionability to consider the Web as a distributed computational system.
 
-To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:
+To better understand the relationship between the FDO framework and other existing approaches, we use the following for analysis:
 
@@
 Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data in general practice.
+For conceptual comparisions we consider FAIR Digital Objects and the Web broadly.
+Ror implementations,we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice.
 
@@
 Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the extebsuve documentation and broad developer usage that is available for Web APIs could potentially be utillised for FDO.
+There is a question as to whether the extensive documentation and broad developer usage that is available for Web APIs could potentially be utilised for FDO.
 
@@
 
-It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
+It is also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
 For example, DOIs are mostly resolved on the web \autocite{DOIResolutionDocumentation} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \autocite{DOIHandbookResolution}.
@@
 
-In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
+In this section, we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
 As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.
@@
 The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite[][slide 10]{SemanticWebXML2000}, as we have previously proposed \autocite{10.3897/rio.8.e94501}.
+At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite[][slide 10]{SemanticWebXML2000}, as  has previously been proposed \autocite{10.3897/rio.8.e94501}.
 However, this deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed can make it difficult to achieve the FDO goals without significant effort.
@@
 We therefore identify the need for a new explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
-It is beyond the scope of this work to detail such a profile, but its main priorities could be:
+Such a combination of efforts could utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a ecosystem of machine-actionable objects.
+It is beyond the scope of this work to detail such a profile, but we suggest the following potential key aspects:
 
@@
 
-In this work we have considered FAIR Digital Objects (FDO) as a potential distributed object system and compared FDO with established Web approaches focusing on Linked Data.
+In this work, we have considered FAIR Digital Objects (FDO) as a potential distributed object system for FAIR data and compared it with established Web approaches focusing on Linked Data.
 We have described the background of the Semantic Web and FAIR Digital Objects, and evaluated both using multiple conceptual frameworks.
@@
 We find that both FDO and Linked Data approaches can significantly benefit from each-other and should be aligned further.
-Namely Linked Data proponents need to make their technologies more approachable, agreeing on predictable and consistent implementations of FAIR principles.
+Namely, Linked Data proponents need to make their technologies more approachable, agreeing on predictable and consistent implementations of FAIR principles.
 
@@
 \begin{small}
-This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).
+This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT);  HORIZON-CL4-2021-HUMAN-01-01 101070305 (ENEXA)  and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).
 \end{small}

### 2023-06-16 latex/main.tex

@@
 \begin{abstract}
-FAIR Digital Objects (FDO) is an emerging concept that is highlighted by European Open Science Cloud (EOSC) as a potential candidate for building a ecosystem of machine-actionable research outputs.
+FAIR Digital Object (FDO) is an emerging concept that is highlighted by European Open Science Cloud (EOSC) as a potential candidate for building an ecosystem of machine-actionable research outputs.
 In this work we systematically evaluate FDO and its implementations as a global distributed object system, by using five different conceptual frameworks that cover interoperability, middleware, FAIR principles, EOSC requirements and FDO guidelines themself.
+
 We compare the FDO approach with established Linked Data practices and the existing Web architecture, and provide a brief history of the Semantic Web while discussing why these technologies may have been difficult to adopt for FDO purposes.
@@
 It is also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web \autocite{DOIResolutionDocumentation} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \autocite{DOIHandbookResolution}.
+For example, DOIs are mostly resolved on the web using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \autocite{DOIHandbookResolution}.
 

### 2023-06-16 manuscript.tex
Merge remote-tracking branch 'origin/overleaf-2023-06-16-1051' into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/040dee6cbdcfc6eba52dae0267b6458d68162dba.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/040dee6cbdcfc6eba52dae0267b6458d68162dba/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5289153470

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/f701f88bde71d6a44507f5904a8e4ac2ad6d69ce/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/040dee6cbdcfc6eba52dae0267b6458d68162dba/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/f701f88bde71d6a44507f5904a8e4ac2ad6d69ce}{stain/2022-fdo-paper@f701f88}
-on June 15, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/040dee6cbdcfc6eba52dae0267b6458d68162dba}{stain/2022-fdo-paper@040dee6}
+on June 16, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 15 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 16 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 15 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 16 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 15 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 16 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 15 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 16 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 15 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 16 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 15 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 16 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 15 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 16 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 15 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 16 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 15 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 16 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 15 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 16 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 15 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 16 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 15 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 16 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 15 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 16 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 15 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 16 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 15 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 16 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 15 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 16 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 15 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 16 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 15 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 16 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 15 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 16 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 15 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 16 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 15 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 16 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 15 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 16 June 2023).
 

### 2023-06-16 latex/main.tex
Paul's excellent revisions

@@
 
-These approaches - FDO and Linked Data - thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
+These approaches -- FDO and Linked Data -- thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
 Given their importance, in this article, we aim to compare FAIR Digital Objects, with Linked Data and the Web architecture in the context of the discourse around FAIR data.
@@
 \item
-  Following a Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
+  Following a Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP in Materal Sciences archives \autocite{10.1002/jcc.26842}.
 \item
-  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
+  Approaching the FDO principles from existing Linked Data practices on the Web,e.g.~WorkflowHub use of RO-Crate and schema.org \autocite{10.3897/rio.8.e93937}.
 \end{itemize}
@@
 As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages for internal review.
-As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide their brief summaries below:
+As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide a brief summary of each:
 
@@
 The \textbf{Linked Data} concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect of the Semantic Web back into focus.
-Crucial toLinked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
+Crucial to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
 This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
@@
 The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed  \vpageref*{sec:next-step-fdo}).
+We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref{sec:next-step-fdo}).
 
 Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For conceptual comparisions we consider FAIR Digital Objects and the Web broadly.
-Ror implementations,we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice.
+For conceptual comparisons we consider FAIR Digital Objects and the Web broadly.
+For implementations,we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice.
 

### 2023-06-16 manuscript.tex
Paul's excellent revisions [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/192b16d94263109406c3457018cf55bec4a36281.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/192b16d94263109406c3457018cf55bec4a36281/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5289312684

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/040dee6cbdcfc6eba52dae0267b6458d68162dba/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/192b16d94263109406c3457018cf55bec4a36281/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/040dee6cbdcfc6eba52dae0267b6458d68162dba}{stain/2022-fdo-paper@040dee6}
+from \href{https://github.com/stain/2022-fdo-paper/tree/192b16d94263109406c3457018cf55bec4a36281}{stain/2022-fdo-paper@192b16d}
 on June 16, 2023.

### 2023-06-16 latex/main.tex
Responded to Paul's comments. Some new text and citations.

@@
 The FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been recognised by the research community and policy makers as a goal to strive for.
+They have been recognised by the research community and policy makers as a goal to strive for \autocite{h2020fair2016}.
 In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \autocite{monsCloudyIncreasinglyFAIR2017b}.
@@
 \textbf{Configuration Types} \autocite{fdo-ConfigurationTypes} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022}
+Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022}.
 
@@
 
+See bibliography \vref*{sec:fdo-bibliography} for the citation per document above.
 It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-See bibliography \vref*{sec:fdo-bibliography} for the citation per document above.
+Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDO is not yet a single ecosystem.
 
@@
 The distinction between locator-full and locator-less identifiers have weakened in recent years \autocite{InfoURIRegistry}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \autocite{jutyIdentifiersOrgMIRIAM2011}.
-RDF 1.1 formats use Unicode to support IRIs \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
+RDF 1.1 formats use Unicode to support \emph{IRI}s \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
 By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \autocite{sauermannCoolURIsSemantic2011}.}, the Semantic Web builds a self-described system of types and properties, where the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
@@
 
+RDF is one way to implement \emph{knowledge graphs}, a system of named edges and nodes\footnote{In RDF, each triple represent an edge that is named using its property URI, and the nodes are subject/object as URIs, blank nodes or (for objects) typed literal values \autocite{w3-rdf11-primer}.} \autocite{nurdiati2008}, which when used to represent a sufficiently detailed model of the world, can then be queried and processed to answer detailed research questions.
 The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins \autocite{gobleStateNationData2008c,williamsOpenPHACTSSemantic2012c}.
@@
 Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \autocite{w3-skos-primer} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \autocite{huHowMatchableAre2011a}.
+Exemplifying the need for such cross-references, automated ontology mappings have identified large potential overlaps like 372 definitions of \texttt{Person} \autocite{huHowMatchableAre2011a}.
 
-The move towards \emph{Open Science} data sharing practices did from the late 2000s encourage knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
+The move towards \emph{Open Science} data sharing practices did from the late 2000s encourage knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{\emph{RDF Datasets}} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
 This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \autocite{grothAPIcentricLinkedData2014b}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
@@
 Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \autocite{kleinScholarlyContextNot2014a}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \autocite{polleresMoreDecentralizedVision2020a}.
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), hash vs slash, HTTP status codes and PID redirection strategies \autocite{sauermannCoolURIsSemantic2011}.
+Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), and deployment challenges \autocite{sauermannCoolURIsSemantic2011} (e.g.
+hash vs slash in namespaces, HTTP status codes and PID redirection strategies).
 
@@
 
-Linked Data provides a technologies that have evolved over time for the purposes of data interoperability.
-The need to embrace the web and developer experience have been central lessons learned.
-In contrast, FDO is a new specification with many different potential paths forward.
+Linked Data provides a technologies that have evolved over time to satisfy its primary purpose of data interoperability.
+The need to embrace the Web and developer experience have been central lessons learned.
+In contrast, FDO is a new approach with many different potential paths forward, and having a partial overlap with the aims of Linked Data.
 
@@
 
+EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
+Therefore we here find it important to analyse how  FDO and Linked Data can help satisfy the EOSC requirements.
+
 In Table \vref{tbl:eosc} we review the EOSC Interoperability Framework (EOSC IF) recommendations, and evaluate to what extent they are addressed by the principles of FDO and Linked Data or their common implementations.
@@
 The FAIR and Linked Data communities likewise need to recognize the need for simpler, more pragmatic approaches that make it easier for FAIR practitioners to adapt the technologies with "just enough" semantics.
-We have previously proposed the combination of RO-Crate \autocite{10.3233/ds-210053} and Signposting \autocite{vandesompelFAIRSignpostingProfile2022} as a mean to implement FDO \autocite{10.3897/rio.8.e93937} over HTTP using a common Linked Data metadata model.
 
-However it may be sufficient to use HTTP-based FAIR Signposting alone to achieve the above list, if one considers only a small metadata model, and rather reference from the signposting which metadata resources are additionally available.
-This will allow any Linked Data resource to gradually participate in the FDO ecosystem, with minimal effort and non-intrusive implementation changes.
+%We have previously proposed the combination of RO-Crate \autocite{10.3233/ds-210053} and Signposting \autocite{vandesompelFAIRSignpostingProfile2022} as a mean to implement FDO \autocite{10.3897/rio.8.e93937} over HTTP using a common Linked Data metadata model.
+
+%However it may be sufficient to use HTTP-based FAIR Signposting alone to achieve the above list, if one considers only a small metadata model, and rather reference from the signposting which metadata resources are additionally available.
+%This will allow any Linked Data resource to gradually participate in the FDO ecosystem, with minimal effort and non-intrusive implementation changes.
 FDO implementations like Cordra typically already use HTTP APIs that align with DOIP \autocite{DOIPAPIHTTPa}, these can be augmented with Signposting headers without necessarily moving to a Linked Data metadata model.

### 2023-06-16 manuscript.tex
Responded to Paul's comments. Some new text and citations. [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/39b67330882bd6a00851c8595baf225d32892b11.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/39b67330882bd6a00851c8595baf225d32892b11/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5291552036

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/192b16d94263109406c3457018cf55bec4a36281/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/39b67330882bd6a00851c8595baf225d32892b11/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/192b16d94263109406c3457018cf55bec4a36281}{stain/2022-fdo-paper@192b16d}
+from \href{https://github.com/stain/2022-fdo-paper/tree/39b67330882bd6a00851c8595baf225d32892b11}{stain/2022-fdo-paper@39b6733}
 on June 16, 2023.

### 2023-06-16 latex/main.tex
language tweak. linenos for review

@@
 
-\documentclass[fleqn,10pt,NOlineno]{wlpeerjlua}
+\documentclass[fleqn,10pt,lineno]{wlpeerjlua}
 
@@
 
-EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
-Therefore we here find it important to analyse how  FDO and Linked Data can help satisfy the EOSC requirements.
+As covered in our introduction, EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
+Therefore we here find it important to summarize how FDO and Linked Data can help satisfy the EOSC requirements.
 

### 2023-06-17 latex/table1.tex
reference http rfcs

@@
 \textbf{Pragmatic}: \emph{using interaction contracts so processes can be choreographed in workflows}
-  & FDO types and operations enable detailed choreography (Canonical Workflow Frameworks for Research \autocite{cwfr}).
+  & FDO types and operations enable detailed choreography (Canonical Workflows; \textcite{cwfr}).
 \texttt{0.TYPE/DOIPOperation} has lightweight definition of operation, \texttt{0.DOIP/Request} or \texttt{0.DOIP/Response} may give FDO Type or any other kind of ``specifics'' (incl.~human readable docs).
@@
   & ``Follow your nose'' crawler navigation, which may lead to frequent dead ends.
-Operational composition, typically within a single API provider, documented by OpenAPI 3 \autocite{OpenAPISpecificationV3}, schema.org Actions \autocite{SchemaOrgActions}, WSDL/SOAP \autocite{w3-wsdl20-primer}, but frequently just as human-readable developer documentation/examples.
+Operational composition, typically within a single API provider, documented by OpenAPI 3 \autocite{OpenAPISpecificationV3}, schema.org Actions \autocite{SchemaOrgActions}, WSDL/SOAP \autocite{w3-wsdl20-primer}, but frequently just as human-readable developer documentation with examples.
 \\
@@
 \textbf{Connective}: \emph{transferring messages to another application, e.g.~wrapping in other protocols}
-  & \textcite{foundationDigitalObjectInterface} is transport-independent, commonly TLS TCP/IP port 9000), \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{DOIP over HTTP}
-  & HTTP/1.1 (TCP/IP port 80), HTTP/1.1+TLS (TCP/IP 443), HTTP/2 (as HTTP/1* but binary), HTTP/3 (like HTTP/2+TLS but UDP) \\
+  & \textcite{foundationDigitalObjectInterface} is transport-independent, commonly TLS TCP/IP port 9000, DOIP over HTTP \autocite{DOIPAPIHTTPa}
+  & HTTP/1.1 TCP/IP port 80 \autocite{rfc2616}, HTTP/1.1+TLS TCP/IP 443 \autocite{rfc2818}, HTTP/2 (as HTTP/1* but binary) \autocite{rfc7540}, HTTP/3 (like HTTP/2+TLS but UDP) \autocite{rfc9114} \\
 \textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability}

### 2023-06-17 latex/main.tex
minor typos

@@
 These approaches -- FDO and Linked Data -- thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to compare FAIR Digital Objects, with Linked Data and the Web architecture in the context of the discourse around FAIR data.
+Given their importance, in this article, we aim to compare FAIR Digital Objects with Linked Data and the Web architecture in the context of the discourse around FAIR data.
 
@@
 The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation ( \href{https://cordra.org/}{Cordra}).
+These are mostly realised using the reference implementation Cordra \autocite{tupelo-schneckrobertBriefIntroductionCordra2022}.
 FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \autocite{Draftbhuttonjsonschema} and PIDs are assigned using the Handle system.
@@
 \item
-  Approaching the FDO principles from existing Linked Data practices on the Web,e.g.~WorkflowHub use of RO-Crate and schema.org \autocite{10.3897/rio.8.e93937}.
+  Approaching the FDO principles from existing Linked Data practices on the Web, e.g.~WorkflowHub use of RO-Crate and schema.org \autocite{10.3897/rio.8.e93937}.
 \end{itemize}
@@
 The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages for internal review.
+As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review.
 As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide a brief summary of each:
@@
 See bibliography \vref*{sec:fdo-bibliography} for the citation per document above.
-It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDO is not yet a single ecosystem.
+It is worth pointing out that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
+Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are thus not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDOs are not yet in a single ecosystem.
 
@@
 
-The \textbf{Linked Data} concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect of the Semantic Web back into focus.
+The \textbf{Linked Data} (LD) concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect of the Semantic Web back into focus.
 Crucial to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
@@
 A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \autocite{verborghSemanticWebIdentity2020a}.
-One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \autocite{wolstencroftRightFieldEmbeddingOntology2011b}, the use of Linked Data as part of REST Web APIs \autocite{pageRESTLinkedData2011}, and as shown by the big uptake by publishers to annotate the Web using schema.org \autocite{bernsteinNewLookSemantic2016a}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
+One bridging gap here by the Linked Data movement has been ``Linked Data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \autocite{wolstencroftRightFieldEmbeddingOntology2011b}, the use of Linked Data as part of REST Web APIs \autocite{pageRESTLinkedData2011}, and as shown by the big uptake by publishers to annotate the Web using schema.org \autocite{bernsteinNewLookSemantic2016a}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
 
-Linked Data provides a technologies that have evolved over time to satisfy its primary purpose of data interoperability.
-The need to embrace the Web and developer experience have been central lessons learned.
+Linked Data provides technologies that have evolved over time to satisfy its primary purpose of data interoperability.
+The needs to embrace the Web and developer experience have been central lessons learned.
 In contrast, FDO is a new approach with many different potential paths forward, and having a partial overlap with the aims of Linked Data.
@@
 The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref{sec:next-step-fdo}).
+We have left out further considerations on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref{sec:next-step-fdo}).
 
@@
 For conceptual comparisons we consider FAIR Digital Objects and the Web broadly.
-For implementations,we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice.
+For implementations, we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice\footnote{For further background on FDO implemented with Linked Data see \autocite{FDOFramework,10.3897/rio.8.e94501}}.
 

### 2023-06-17 latex/main.tex
final tweaks, ready to go!

@@
 \usepackage{pdflscape} 
+%\usepackage[pass]{geometry}
 
@@
 The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject} sets out recommendations for FDO implementations.
-In Table \vref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \autocite{FDOFramework}.
+Note that the proposed update to FDO specification \autocite{fdo-RequirementSpec} clarifies these definitions with equivalent identifiers\footnote{Newer \autocite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
 
-Note that the draft update to FDO specification \autocite{fdo-RequirementSpec} clarifies these definitions with equivalent identifiers\footnote{Newer \autocite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
+In Table \vref{tbl:fdo-checks} we evaluate completeness of the guidelines in two current FDO realizations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}.
 
@@
 Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement \autocite{fdo-DOIPEndorsement} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (details \vpageref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
+The DOIP Endorsement \autocite{fdo-DOIPEndorsement} set out that to comply, DOIP must be used according to the set of FDO requirement documents (details \vpageref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
+Likewise, a Linked Data approach will need to follow the same requirements to actually comply as an FDO implementation.
 
@@
   While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} \autocite{w3-prov-o}
+Linked Data frequently have multiple representations, but often not sufficiently linked (link relation \texttt{alternate} \autocite{rfc8288}) or related (\texttt{prov:specializationOf} from \textcite{w3-prov-o}).
 \item
-  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
+  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives. LDP has specific native support for containers.
 \item
@@
 
-As covered in our introduction, EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
+As covered in our introduction\vref{sec:introduction}, EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
 Therefore we here find it important to summarize how FDO and Linked Data can help satisfy the EOSC requirements.
@@
 While these community aspects are not mandated by Linked Data practices, best practices have become established for aspects like ontology development \autocite{10.1186/s13326-021-00240-6}.
-EOSC IF's technical layer is likewise at a architecturally high level, such as service-level agreements, but also highlight PID policies which is strongly required by FDO, while Linked Data communities choose PID practices separately.
-The recommendations for the Semantic layer, is largely already implemented by Linked Data practices, yet for FDO mostly consist of encouragements.
+EOSC IF's \emph{Technical} layer is likewise at a architecturally high level, such as service-level agreements, but also highlight PID policies which is strongly required by FDO, while Linked Data communities choose PID practices separately.
+The recommendations for the \emph{Semantic} layer is largely already implemented by Linked Data practices, yet for FDO mostly consist of encouragements.
 For instance \emph{clear definitions of semantic concepts} is required by FDO guidelines, but how to technically define them has not been formalised by FDO specifications.
 
-The Legal layer of interoperability is perhaps the one most emphasised by EOSC, by enabling collaboration across organizational barriers to joinly build a research infrastructure, but this is an area that both FDO and Linked Data are relatively weak in directly supporting.
-The EOSC IF recommendations in this layer are still largely related to governance practices and metadata, for instance licensing, privacy and usage policies; yet these are essential for cross-institutional and cross-repository access of FAIR objects.
+The \emph{Legal} layer of interoperability is perhaps the one most emphasised by EOSC, by enabling collaboration across organizational barriers to joinly build a research infrastructure, but this is an area that both FDO and Linked Data are relatively weak in directly supporting.
+The EOSC IF recommendations in this layer are largely related to governance practices and metadata, for instance licensing, privacy and usage policies; these are also essential for cross-institutional and cross-repository access of FAIR objects.
 
@@
 When evaluating FDO implementations against the FDO guidelines (\vref*{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used.
-Achieving fully interoperable FAIR digital objects would require further convergence on implementation practices, and it is not given that his need to diverge from the established Web architecture.
-It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional restrictions, equably implementable, such as using persistent identifiers or pre-defining an object's type.
+Achieving fully interoperable FAIR Digital Objects would require further convergence on implementation practices, and it is not given that this needs to diverge from the established Web architecture.
+It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional equably implementable restrictions and conventions, such as using persistent identifiers or pre-defining an object's type.
 
 Considering this, by comparing FDO and Web as middleware (\vref*{sec:middleware}) we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g.
-lack of concurrency transparancy.
+lack of concurrency and transparancy.
 Recent updates to HTTP have added many features needed for large-scale usage such as video streaming services (e.g.
-caching, multiplexing, cloud deployments), and having the option to transparantly apply these also to FDOs seems like a strong incentive.
-Many programmatic features are however missing or needing custom extensions in both aspects, such as transactions, asynchronous operations and streaming.
+caching, multiplexing, cloud deployments), and having the option to transparently apply these also to FDOs seems like a strong incentive.
+Many programmatic features for distributed objects are however missing or needing custom extensions in both aspects, such as transactions, asynchronous operations and streaming.
 
-By assessing FDO against the FAIR principles (\vref*{sec:fair-compare}) we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
+By assessing FDO against the FAIR principles (\vref*{sec:fair-compare}) we found that both FDO implementations are underspecified ian several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
 While there are implementations of each of these in general Linked Data examples, there is no single set of implementation guides that fully realizes the FAIR principles.
@@
 material science).
-It may be advantegous for both these effort to be aligned and jointly promoted under the EOSC umbrella.
+It may be advantageous for both these effort to be aligned and jointly promoted under the EOSC umbrella.
 
@@
 We therefore identify the need for a new explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts could utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a ecosystem of machine-actionable objects.
+Such a combination of efforts could utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow an ecosystem of machine-actionable objects.
 It is beyond the scope of this work to detail such a profile, but we suggest the following potential key aspects:
@@
   \item Types, attributes and actions are self-defined by their identifier
+  \item Use Web approaches directly where possible, rather than wrap in a new model
 \end{itemize}
@@
 The FDO recommendations show that FAIR thinking in this regard need to move beyond data publishing and into machine actionability across digital objects, and with broader community consensus. 
-As flexibility for extensions is a necessary ingredient alongside rigidity for core concepts, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as the early Semantic Web adopters.
+As flexibility for extensions is a necessary ingredient alongside rigidity for core concepts, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as learnt by early Semantic Web adopters.
 
@@
 
+\clearpage
 \defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}

### 2023-06-17 manuscript.tex
final tweaks, ready to go! [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/86cfd89b9215552dadcb75065d595d3e7c1d6c42.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/86cfd89b9215552dadcb75065d595d3e7c1d6c42/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5298397464

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/39b67330882bd6a00851c8595baf225d32892b11/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/86cfd89b9215552dadcb75065d595d3e7c1d6c42/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/39b67330882bd6a00851c8595baf225d32892b11}{stain/2022-fdo-paper@39b6733}
-on June 16, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/86cfd89b9215552dadcb75065d595d3e7c1d6c42}{stain/2022-fdo-paper@86cfd89}
+on June 17, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 16 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 17 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 16 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 17 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 16 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 17 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 16 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 17 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 16 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 17 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 16 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 17 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 16 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 17 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 16 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 17 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 16 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 17 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 16 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 17 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 16 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 17 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 16 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 17 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 16 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 17 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 16 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 17 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 16 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 17 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 16 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 17 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 16 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 17 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 16 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 17 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 16 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 17 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 16 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 17 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 16 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 17 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 16 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 17 June 2023).
 

### 2023-06-17 manuscript.tex
No need for _site clone [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/f615459f766ae8625991bc1f1e3f980a17b606ba.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/f615459f766ae8625991bc1f1e3f980a17b606ba/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5298666326

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/86cfd89b9215552dadcb75065d595d3e7c1d6c42/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/f615459f766ae8625991bc1f1e3f980a17b606ba/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/86cfd89b9215552dadcb75065d595d3e7c1d6c42}{stain/2022-fdo-paper@86cfd89}
+from \href{https://github.com/stain/2022-fdo-paper/tree/f615459f766ae8625991bc1f1e3f980a17b606ba}{stain/2022-fdo-paper@f615459}
 on June 17, 2023.

### 2023-06-18 latex/main.tex
fix typo

@@
 
-By assessing FDO against the FAIR principles (\vref*{sec:fair-compare}) we found that both FDO implementations are underspecified ian several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
+By assessing FDO against the FAIR principles (\vref*{sec:fair-compare}) we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
 While there are implementations of each of these in general Linked Data examples, there is no single set of implementation guides that fully realizes the FAIR principles.

### 2023-06-18 manuscript.tex
fix typo [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/4210fb857855271dcb3adb645fca755e429315e6.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/4210fb857855271dcb3adb645fca755e429315e6/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5303217057

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/f615459f766ae8625991bc1f1e3f980a17b606ba/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/4210fb857855271dcb3adb645fca755e429315e6/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/f615459f766ae8625991bc1f1e3f980a17b606ba}{stain/2022-fdo-paper@f615459}
-on June 17, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/4210fb857855271dcb3adb645fca755e429315e6}{stain/2022-fdo-paper@4210fb8}
+on June 18, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 17 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 18 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 17 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 18 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 17 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 18 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 17 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 18 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 17 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 18 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 17 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 18 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 17 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 18 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 17 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 18 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 17 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 18 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 17 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 18 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 17 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 18 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 17 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 18 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 17 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 18 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 17 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 18 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 17 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 18 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 17 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 18 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 17 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 18 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 17 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 18 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 17 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 18 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 17 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 18 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 17 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 18 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 17 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 18 June 2023).
 

### 2023-06-19 latex/main.tex
Results header missing
EOSC header wrong level

--- 2023-06-17 latex/main.tex
@@
 
-\title{Evaluating FAIR Digital Object and Linked Data as distributed object systems}
+\title{Evaluating FAIR Digital Object as a distributed object system}
 
@@
 
-%\keywords{FAIR, FDO, digital object, Linked Data, EOSC, distributed objects, frameworks, interoperability, middleware}
+% \keywords{Keyword1, Keyword2, Keyword3}
 
 \begin{abstract}
-FAIR Digital Object (FDO) is an emerging concept that is highlighted by European Open Science Cloud (EOSC) as a potential candidate for building an ecosystem of machine-actionable research outputs.
-In this work we systematically evaluate FDO and its implementations as a global distributed object system, by using five different conceptual frameworks that cover interoperability, middleware, FAIR principles, EOSC requirements and FDO guidelines themself.
-
-We compare the FDO approach with established Linked Data practices and the existing Web architecture, and provide a brief history of the Semantic Web while discussing why these technologies may have been difficult to adopt for FDO purposes.
-We conclude with recommendations for both Linked Data and FDO communities to further their adaptation and alignment.
+FAIR Digital Object is an emerging concept from EOSC.
+This is important.
+Worthwile to understand how semantic technologies and semantic web vision relate to this emerging landscape.
+Here we do this systematically by comparing the technologies introduced under the banner of FAIR digital Object and Semantic Web.
 \end{abstract}
@@
 The FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e} encourage sharing of scientific data with machine-readable metadata and the use of interoperable formats, and are being adapted by a wide range of research infrastructures.
-They have been recognised by the research community and policy makers as a goal to strive for \autocite{h2020fair2016}.
+They have been widely recognised by the research community and policy makers as a goal to strive for.
 In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \autocite{monsCloudyIncreasinglyFAIR2017b}.
@@
 In other discourse, Linked Data \autocite{bizerLinkedDataStory2009a} has been seen as an established set of principles based on Semantic Web technologies that can achieve the vision of the FAIR principles \autocite{boninodasilvasantosFAIRDataPoints2016a,hasnainAssessingFAIRData2018a}.
-Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a ``FAIR Linked Data approach'' fully \autocite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \autocite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}.
-While such focus on simplicity allows for rapid development and highly specialised services, it raises wider concerns about interoperability \autocite{turcoaneLinkedDataJSONLD2014a,wilkinsonWorkflowsWhenParts2022b}.
+Yet regular researchers and developers of emerging platforms for computation and data management are reluctant to adapt such a FAIR Linked Data approach fully \autocite{verborghSemanticWebIdentity2020a}, opting instead for custom in-house models and JSON-derived formats from RESTful Web services \autocite{merono-penuelaConclusionFutureChallenges2021a,neumannAnalysisPublicREST2021a}.
+While such focus on simplicity gives rapid development and highly specialised services, it raises wider concerns on interoperability \autocite{turcoaneLinkedDataJSONLD2014a,wilkinsonWorkflowsWhenParts2022b}.
 
@@
 
-These approaches -- FDO and Linked Data -- thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to compare FAIR Digital Objects with Linked Data and the Web architecture in the context of the discourse around FAIR data.
+These approaches, thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
+Given their importance, in this article, we aim to examine the relationships between FAIR and FAIR Digital Objects, contrasted with Linked Data and the Web in general.
 
-Concretely, the contribution of this paper is {\bf a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks} that capture different perspectives on interoperability and readiness for implementation.
+Concretely, the contribution of this paper is a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks that capture different perspectives on interoperability and readiness for implementation.
 
@@
 The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
-This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced by \textcite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
-DO was the inspiration for the \textcite{x1255FrameworkDiscovery} framework which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) \autocite{DigitalObjectInterface}.
+This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced in 1995 \autocite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
+DO was the inspiration for the ITU X.1255 framework \autocite{x1255FrameworkDiscovery} which introduced an abstract \emph{Digital Entity Interface Protocol} for managing such objects programmatically, first realised by the Digital Object Interface Protocol (DOIP) \autocite{DigitalObjectInterface}.
 
@@
 \item
-  Building on the Digital Object concept, using the simplified \textcite{foundationDigitalObjectInterface} specification, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0, see \textcite{DOIPExamplesCordraa}} (usually TCP/IP over TLS).
+  Building on the Digital Object concept, using the simplified DOIP v2 specification \autocite{foundationDigitalObjectInterface}, which detail how to exchange JSON objects through a text-based protocol\footnote{For a brief introduction to DOIP 2.0, see \autocite{DOIPExamplesCordraa}} (usually TCP/IP over TLS).
 The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation Cordra \autocite{tupelo-schneckrobertBriefIntroductionCordra2022}.
+These are mostly realised using the reference implementation \href{https://cordra.org/}{Cordra}.
 FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \autocite{Draftbhuttonjsonschema} and PIDs are assigned using the Handle system.
@@
 \item
-  Following a Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP in Materal Sciences archives \autocite{10.1002/jcc.26842}.
+  Following the traditional Linked Data approach, but using the DOIP protocol, e.g.~using JSON-LD and schema.org within DOIP (NIST for material science).
 \item
-  Approaching the FDO principles from existing Linked Data practices on the Web, e.g.~WorkflowHub use of RO-Crate and schema.org \autocite{10.3897/rio.8.e93937}.
+  Approaching the FDO principles from existing Linked Data practices on the Web (e.g.~WorkflowHub use of RO-Crate and schema.org).
 \end{itemize}
 
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we will cover \vpageref{sec:ld}.
+From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we'll cover \vpageref{sec:ld}.
 
@@
 The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review.
-As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide a brief summary of each:
+As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages in Google Docs for internal review.
+As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide their brief summaries below:
 
@@
 \textbf{Configuration Types} \autocite{fdo-ConfigurationTypes} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022}.
+Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022}
 
@@
 
-See bibliography \vref*{sec:fdo-bibliography} for the citation per document above.
-It is worth pointing out that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are thus not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDOs are not yet in a single ecosystem.
+It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
 
@@
 
-Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \autocite{w3-rdf11-primer} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \autocite{w3-rdf-syntax99} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
+Through W3C, the Semantic Web was realised with the Resource Description Framework (RDF) \autocite{w3-rdf11-primer} that used \emph{triples} of subject-predicate-object statements, with its initial serialisation format \autocite{w3-rdf-syntax} being RDF/XML (XML was at the time seen as a natural data-focused evolution from the document-centric SGML and HTML).
 
@@
 The distinction between locator-full and locator-less identifiers have weakened in recent years \autocite{InfoURIRegistry}, for instance DOI identifiers now are commonly expressed with the prefix \texttt{https://doi.org/} rather than as URNs with \texttt{info:doi:} given that the URL/URN gap has been bridged by HTTP resolvers and the use of Persistent Identifiers (PIDs) \autocite{jutyIdentifiersOrgMIRIAM2011}.
-RDF 1.1 formats use Unicode to support \emph{IRI}s \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
+RDF 1.1 formats use Unicode to support IRIs \autocite{rfc3987}, which extends URIs to include international characters and domain names.} as the primary identifier of the \emph{subject} (what the statement is about), \emph{predicate} (relation/attribute of the subject) and \emph{object} (what is pointed to).
 By using URIs not just for documents\footnote{URIs can also identify \emph{non-information resources} for any kind of physical object (e.g.~people), such identifiers can resolve with \texttt{303\ See\ Other} redirections to a corresponding \emph{information resources} \autocite{sauermannCoolURIsSemantic2011}.}, the Semantic Web builds a self-described system of types and properties, where the meaning of a relation can be resolved by following its hyperlink to the definition within a \emph{vocabulary}.
@@
 
-The mid-2000s saw large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL \autocite{w3-owl2-overview}, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm.
-%(e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
-More human-readable syntaxes for RDF such as Turtle evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
+The mid-2000s saw large academic interest and growth of the Semantic Web, with the development of more formal representation system for ontologies, such as OWL \autocite{w3-owl2-overview}, allowing complex class hierarchies and logic inference rules following \emph{open world} paradigm (e.g.~a \emph{ex:Parent} is equivalent to a subclass of \emph{foaf:Person} which must \emph{ex:hasChild} at least one \emph{foaf:Person}, then if we know \emph{:Alice a ex:Parent} we can infer \emph{:Alice ex:hasChild {[}a foaf:Person{]}} even if we don't know who that child is).
+More human-readable syntaxes of RDF such as Turtle (shown in this paragraph) evolved at this time, and conferences such as \href{https://iswc2022.semanticweb.org/}{ISWC} \autocite{horrocksSemanticWebISWC2002} gained traction, with a large interest in knowledge representation and logic systems based on Semantic Web technologies evolving at the same time.
 
-Established Semantic Web services and standards include: SPARQL \autocite{w3-sparql11-overview} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} \autocite{w3-rdf11-concepts} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and  graph visualisation.
+Established Semantic Web services and standards include SPARQL \autocite{w3-sparql11-overview} (pattern-based triple queries), \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{named graphs} \autocite{w3-rdf11-concepts} (triples expanded to \emph{quads} to indicate statement source or represent conflicting views), triple/quad stores (graph databases such as OpenLink Virtuoso, GraphDB, 4Store), mature RDF libraries (including Redland RDF, Apache Jena, Eclipse RDF4J, RDFLib, RDF.rb, rdflib.js), and numerous graph visualisation (many of which struggle with usability for more than 20 nodes).
 
-RDF is one way to implement \emph{knowledge graphs}, a system of named edges and nodes\footnote{In RDF, each triple represent an edge that is named using its property URI, and the nodes are subject/object as URIs, blank nodes or (for objects) typed literal values \autocite{w3-rdf11-primer}.} \autocite{nurdiati2008}, which when used to represent a sufficiently detailed model of the world, can then be queried and processed to answer detailed research questions.
 The creation of RDF-based knowledge graphs grew particularly in fields like bioinformatics, e.g.~for describing genomes and proteins \autocite{gobleStateNationData2008c,williamsOpenPHACTSSemantic2012c}.
 In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~\emph{should a gene identifier signify just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?}) lead to large differences in representational granularity, and thus needed different identifiers.
+One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~should a \emph{gene identifier} signify which protein a DNA sequence would make, or just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?) lead to large differences in representational granularity, and thus needed different identifiers.
 
@@
 Such mappings can be expressed themselves using lightweight and flexible RDF vocabularies such as SKOS \autocite{w3-skos-primer} (e.g.~\texttt{dct:title\ skos:closeMatch\ schema:name} to indicate near equivalence of two properties).
-Exemplifying the need for such cross-references, automated ontology mappings have identified large potential overlaps like 372 definitions of \texttt{Person} \autocite{huHowMatchableAre2011a}.
+Automated ontology mappings have identified large potential overlaps (e.g.~372 definitions of \texttt{Person}) \autocite{huHowMatchableAre2011a} .
 
-The move towards \emph{Open Science} data sharing practices did from the late 2000s encourage knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{\emph{RDF Datasets}} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
+The move towards \emph{Open Science} data sharing practices did from the late 2000s encourage knowledge providers to distribute collections of RDF descriptions as downloadable \emph{datasets} \footnote{\emph{Datasets} that distribute RDF graphs should not be confused with \href{https://www.w3.org/TR/rdf11-concepts/\#section-dataset}{RDF Datasets} used for partitioning \emph{named graphs}.}, so that their clients can avoid thousands of HTTP requests for individual resources.
 This enabled local processing, mapping and data integration across datasets (e.g.~Open PHACTS \autocite{grothAPIcentricLinkedData2014b}), rather than relying on the providers' RDF and SPARQL endpoints (which could become overloaded when handling many concurrent, complex queries).
@@
 Just like \emph{link rot} affect regular Web pages and their citations from scholarly communication \autocite{kleinScholarlyContextNot2014a}, for a majority of described RDF resources in the \href{https://lod-cloud.net/}{Linked Open Data} (LOD) Cloud's gathering of more than thousand datasets, unfortunately do not actually link to (still) downloadable (\emph{dereferenceable}) Linked Data \autocite{polleresMoreDecentralizedVision2020a}.
-Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), and deployment challenges \autocite{sauermannCoolURIsSemantic2011} (e.g.
-hash vs slash in namespaces, HTTP status codes and PID redirection strategies).
+Another challenge facing potential adopters is the plethora of choices, not just to navigate, understand and select to reuse the many possible vocabularies and ontologies \autocite{carrieroLandscapeOntologyReuse2020a}, but also technological choices on RDF serialisation (at least \href{https://www.w3.org/TR/rdf11-primer/\#section-graph-syntax}{7 formats}), type system (RDFS \autocite{w3-rdf-schema}, OWL \autocite{w3-owl2-overview}, OBO \autocite{tirmiziMappingOBOOWL2011a}, SKOS \autocite{w3-skos-primer}), hash vs slash, HTTP status codes and PID redirection strategies \autocite{sauermannCoolURIsSemantic2011}.
 
@@
 
-The \textbf{Linked Data} (LD) concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect of the Semantic Web back into focus.
-Crucial to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
+The \textbf{Linked Data} concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect back into focus.
+Crucially to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
 This means a loosening of the semantic restrictions previously applied, and an emphasis on building navigable data resources, rather than elaborate graph representations.
@@
 Vocabularies like \href{https://schema.org/}{schema.org} evolved not long after, intended for lightweight semantic markup of existing Web pages, primarily to improve search engines' understanding of types and embedded data.
-In addition to several such embedded \emph{microformats} \autocite{OpenGraphProtocol,w3-rdfa-primer,HTMLStandard}, we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
+In addition to several such embedded \emph{microformats} \autocite{OpenGraphProtocol,w3-rdfa-primer,HTMLStandard} or we find JSON-LD \autocite{w3-json-ld} as a Web-focused RDF serialisation that aims for improved programmatic generation and consumption, including from Web applications.
 JSON-LD is as of 2023-05-18 used\footnote{Presumably this large uptake of JSON-LD is mainly for the purpose of Search Engine Optimisation (SEO), with typically small amounts of metadata which may not constitute Linked Data as introduced above, however this deployment nevertheless constitute machine-actionable structured data.} by 45\% of the top 10 million websites \autocite{UsageStatisticsJSONLD}.
@@
 A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \autocite{verborghSemanticWebIdentity2020a}.
-One bridging gap here by the Linked Data movement has been ``Linked Data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \autocite{wolstencroftRightFieldEmbeddingOntology2011b}, the use of Linked Data as part of REST Web APIs \autocite{pageRESTLinkedData2011}, and as shown by the big uptake by publishers to annotate the Web using schema.org \autocite{bernsteinNewLookSemantic2016a}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
-
-Linked Data provides technologies that have evolved over time to satisfy its primary purpose of data interoperability.
-The needs to embrace the Web and developer experience have been central lessons learned.
-In contrast, FDO is a new approach with many different potential paths forward, and having a partial overlap with the aims of Linked Data.
+One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \autocite{wolstencroftRightFieldEmbeddingOntology2011b}, the use of Linked Data as part of REST Web APIs \autocite{pageRESTLinkedData2011}, and as shown by the big uptake by publishers to annotate the Web using schema.org \autocite{bernsteinNewLookSemantic2016a}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
 
@@
 
-Our main motivation for this article is to investigate how FAIR Digital Objects may differ from the learnt experiences of Linked Data and the Web.
-We also aim to reflect back from FDO's motivation of machine-actionability to consider the Web as a distributed computational system.
+Our main motivation for this article is to investigate how the promises of FAIR Digital Objects may differ from the learnt experiences of Linked Data and the Web.
+We also reflect back from FDO's motivation of machine-actionability to consider the Web as a distributed computational system.
 
-To better understand the relationship between the FDO framework and other existing approaches, we use the following for analysis:
+To better understand the relationship between the FDO framework and other exisiting approaches, we use the following for analysis:
 
@@
 The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further considerations on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref{sec:next-step-fdo}).
+We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref*{sec:next-step-fdo}).
 
 Some of these frameworks invite a comparison on a conceptual level, while others relate better to implementations and current practices.
-For conceptual comparisons we consider FAIR Digital Objects and the Web broadly.
-For implementations, we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice\footnote{For further background on FDO implemented with Linked Data see \autocite{FDOFramework,10.3897/rio.8.e94501}}.
+For these we consider FAIR Digital Objects and the Web conceptually, and for implementations we contrast between the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data in general practice.
+
+\section*{Results}
 
@@
 Similiarly, FDO does not enable automatic composition because operation semantics are not well defined.
-There is a question as to whether the extensive documentation and broad developer usage that is available for Web APIs could potentially be utilised for FDO.
+There is a question as to whether the extebsuve documentation and broad developer usage that is available for Web APIs could potentially be utillised for FDO.
 
@@
 
-It is also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
-For example, DOIs are mostly resolved on the web using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \autocite{DOIHandbookResolution}.
+It's also a question as to whether a highly constrained protocol revolving around persistent identifiers is in fact necessary.
+For example, DOIs are mostly resolved on the web \autocite{DOIResolutionDocumentation} using HTTP redirects with the common \texttt{https://doi.org/} prefix, hiding their Handle nature as an implementation detail \autocite{DOIHandbookResolution}.
 
-\input{table1}
+\include{table1}
 
@@
 
-\input{table2}
+\include{table2}
 
@@
 
-In this section, we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
+In this section we take the perspective that FDO principles are in effect proposing a global infrastructure of machine-actionable digital objects.
 As such we can consider implementations of FDO as \textbf{middleware infrastructures} for programmatic usage, and can evaluate them based on expectations for client and server developers.
@@
 Here we've used the RDA's FAIR Data Maturity Model \autocite{groupFAIRDataMaturity2020} as it has decomposed the FAIR principles to a structured list of FAIR indicators \autocite{bahimFAIRDataMaturity2020a}, importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \vref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remaining JSON or RDF structures always considered as metadata.
+In our interpretation for Table \vref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remaining JSON/RDF structures always considered as metadata.
 
@@
 \item
-  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little concrete metadata guidance.
+  Linked Data in general is strong on metadata indicators, but LDP approach is weak as it has little metadata guidance
 \item
-  FDO/DOIP are stronger on identifier indicators, while Linked Data approach for identifiers relies on best practices. 
+  FDO/DOIP are stronger on identifier indicators
 \item
@@
 
-\section*{EOSC Interoperability Framework}\label{sec:eosc-interoperability-framework}
+\subsection*{EOSC Interoperability Framework}\label{eosc-interoperability-framework}
 
@@
 agreements), with a particular aim to address the FAIR interoperability principles and building on the concept of FAIR Digital Objects.
-
-As covered in our introduction, EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
-Therefore we here find it important to summarize how FDO and Linked Data can help satisfy the EOSC requirements.
 
@@
 The EOSC IF recommendations in this layer are still largely related to governance practices and metadata, for instance licensing, privacy and usage policies; yet these are essential for cross-institutional and cross-repository access of FAIR objects.
-
 Likewise, search and indexing is important FAIR aspect for Findability, but is poorly supported globally by FDO and Linked Data.
@@
 
-\input{table6}
+\include{table6}
 
@@
 
-We have evaluated the FAIR Digital Object concept using multiple frameworks, and contrasted FDO against existing experiences from Linked Data on the Web.
-In this section we discuss the implications of this evaluation, and propose how these two approaches can be better combined.
-
-\subsection*{Framework evaluation}
-
-Having considered FDO and the Web architecture as interoperability frameworks (\vref*{sec:interoperability-compare}), we observe that neither are magic bullets, but each bring different aspects of interoperability.
-The Web comes with a large degree of flexibility and openness, however this means interoperability can suffer as services have different APIs and data models, although with common patterns.
-This is also true for Linked Data on the Web, with many overlapping ontologies and frequent inconsistencies in resolution mechanisms; although somewhat alleviated in recent years by schema.org becoming common metadata model for semantic markup inline in Web pages.
-The Web is based on a common HTTP protocol which has remained stable architecturally throughout its 32 years of largely backwards-compatible evolution.
-FDO on the other side sets down multiple rigid rules for identifiers, types, methods etc.
-that are advanterous for interoperability and predictability for FAIR consumption.
-Yet there is a large degree of freedom in how the FDO rules can be implemented by a given community, for instance there is no common metadata model or identifier resolution mechanism, and DOIP is just one possible transport method for FDOs, which itself does not enforce these rules.
-
-When evaluating FDO implementations against the FDO guidelines (\vref*{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used.
-Achieving fully interoperable FAIR digital objects would require further convergence on implementation practices, and it is not given that his need to diverge from the established Web architecture.
-It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional restrictions, equably implementable, such as using persistent identifiers or pre-defining an object's type.
-
-Considering this, by comparing FDO and Web as middleware (\vref*{sec:middleware}) we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g.
-lack of concurrency transparancy.
-Recent updates to HTTP have added many features needed for large-scale usage such as video streaming services (e.g.
-caching, multiplexing, cloud deployments), and having the option to transparantly apply these also to FDOs seems like a strong incentive.
-Many programmatic features are however missing or needing custom extensions in both aspects, such as transactions, asynchronous operations and streaming.
-
-By assessing FDO against the FAIR principles (\vref*{sec:fair-compare}) we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
-While there are implementations of each of these in general Linked Data examples, there is no single set of implementation guides that fully realizes the FAIR principles.
-\emph{FAIRification} efforts like the FAIR Cookbook \autocite{faircookbook} and FAIR Implementation Profiles \autocite{FIP} are bringing existing practices together, but there remains a potential role for FDO in giving a coherent set of implementation practices that can practically achieve FAIR.
-Significant effort, also within EOSC, is now moving towards FAIR metrics \autocite{Devaraju_2021}, which in practice need to make additional assumptions on how FAIR principles are implemented, but these are not always formalized \autocite{10.5281/zenodo.7463421} nor can they be taken to be universally correct \autocite{10.5281/zenodo.7848102}.
-Given that most of the existing FAIR guides and assessment tools are focused on Web and Linked Data, it would be reasonable for FDO to then provide a profile of such implementation choices that can achieve best of both worlds.
-
-EOSC has been largely supportive of FDO, FAIR and related services.
-By contrasting the EOSC Interoperability Framework (\vref*{sec:eosc-interoperability-framework}) with FDO, we found that there are important dimensions that are not solved at a technical level, but through organization collaboration, legal requirements and building community practices.
-FDO recommendations highlight community aspects, but at the same time the largest FAIR communities in many science domains are already producing and consuming Linked Data.
-Just as the Linked Data community has a challenge in convincing more research fields to use Semantic Web technologies, FDO currently need to build many new communities in areas that have shown interest in that approach (e.g.
-material science).
-It may be advantegous for both these effort to be aligned and jointly promoted under the EOSC umbrella.
+\textbf{TODO}
 
@@
 
-\subsection*{What does FDO mean for Linked Data?}\label{sec:what-does-it-mean-for-linked-data}
+\begin{itemize}
+\tightlist
+\item
+  Ramnifications of ide
+\item
+  Bullet points per table
+\end{itemize}
+
+\subsection*{(What does it mean for Linked Data?)}\label{what-does-it-mean-for-linked-data}
 
 The FAIR Digital Object approach raises many important points for Linked Data practictioners.
-At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite[][slide 10]{SemanticWebXML2000}, as  has previously been proposed \autocite{10.3897/rio.8.e94501}.
-However, this deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed can make it difficult to achieve the FDO goals without significant effort.
+At first glance, the explicit requirements of FDOs may seem to be easy to furfill by different parts of the Semantic Web Cake \autocite[][slide 10]{SemanticWebXML2000}.
+However, our deeper investigation, based on multiple frameworks, highlights that the openness and variability of how Linked Data is deployed makes it difficult to achieve the FDO goals without significant effort.
 
 While RDF and Linked Data have been suggested as prime candidates for making FAIR data, we argue that when different developers have too many degrees of freedom (such as serialization formats, vocabularies, identifiers, navigation), interoperability is hampered -- this makes it hard for machines to reliably consume multiple FAIR resources across repositories and data providers.
-Indeed, this may be one reason why the initial FDO effort steered away from Linked Data approaches, but now seems in a danger of opening the many same degrees of freedom within FDO.
 
-We therefore identify the need for a new explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts could utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a ecosystem of machine-actionable objects.
-It is beyond the scope of this work to detail such a profile, but we suggest the following potential key aspects:
-
-\begin{itemize}
-  \tightlist
-  \item Use HTTP(S) as protocol
-  \item Use URIs as identifiers, with persistent identifier promises
-  \item Provide consistent identifier resolution that does not require heuristics
-  \item Common core metadata model
-  \item References are always URIs, and should be persistent identifiers
-  \item Types, attributes and actions are self-defined by their identifier
-\end{itemize}
-
-The FAIR and Linked Data communities likewise need to recognize the need for simpler, more pragmatic approaches that make it easier for FAIR practitioners to adapt the technologies with "just enough" semantics.
-
-%We have previously proposed the combination of RO-Crate \autocite{10.3233/ds-210053} and Signposting \autocite{vandesompelFAIRSignpostingProfile2022} as a mean to implement FDO \autocite{10.3897/rio.8.e93937} over HTTP using a common Linked Data metadata model.
-
-%However it may be sufficient to use HTTP-based FAIR Signposting alone to achieve the above list, if one considers only a small metadata model, and rather reference from the signposting which metadata resources are additionally available.
-%This will allow any Linked Data resource to gradually participate in the FDO ecosystem, with minimal effort and non-intrusive implementation changes.
-FDO implementations like Cordra typically already use HTTP APIs that align with DOIP \autocite{DOIPAPIHTTPa}, these can be augmented with Signposting headers without necessarily moving to a Linked Data metadata model.
+We therefore identify the need for an explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
+Such a combination of efforts will utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a rigid (yet flexible) ecosystem of machine-actionable scholarly objects.
 
@@
 
-In this work, we have considered FAIR Digital Objects (FDO) as a potential distributed object system for FAIR data and compared it with established Web approaches focusing on Linked Data.
-We have described the background of the Semantic Web and FAIR Digital Objects, and evaluated both using multiple conceptual frameworks.
-
-We find that both FDO and Linked Data approaches can significantly benefit from each-other and should be aligned further.
-Namely, Linked Data proponents need to make their technologies more approachable, agreeing on predictable and consistent implementations of FAIR principles.
-
-The FDO recommendations show that FAIR thinking in this regard need to move beyond data publishing and into machine actionability across digital objects, and with broader community consensus. 
-As flexibility for extensions is a necessary ingredient alongside rigidity for core concepts, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as the early Semantic Web adopters.
-
-By implementing the goals of FAIR Digital Objects with the mature technology stack developed for Linked Data, EOSC research infrastructures and researchers in general can create and use FAIR machine-actionable research outputs for decades to come.
-
 \section*{Acknowledgments}
 
-\begin{small}
-This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT);  HORIZON-CL4-2021-HUMAN-01-01 101070305 (ENEXA)  and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).
-\end{small}
-
-We would like to acknowledge the FAIR Digital Object Forum \autocite{FAIRDigitalObjects} community and working groups, where SSR and CG are members.
-
-\section*{Author contributions}
-\begin{small}
-
-Contributions to this article according to the CASRAI CRediT taxonomy\footnote{\url{https://credit.niso.org/}}:
-
-\begin{description}
-\tightlist
-\item[Stian Soiland-Reyes]
-Conceptualization, Formal Analysis, Funding acquisition, Investigation, Methodology,
-Writing -- original draft, Writing -- review \& editing
-\item[Carole Goble]
-Funding acquisition, Supervision, Writing -- review \& editing
-\item[Paul Groth]
-Conceptualization, Methodology, Supervision, Writing -- review \& editing
-\end{description}
-\end{small}
+This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreement numbers H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreement numbers HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s  \emph{Horizon Europe funding guarantee} grant numbers 10038963 (EuroScienceGateway), 10038930 (BioDT).
 
@@
 
-\defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}
+\defbibheading{fdobibliography}{\subsection*{FDO Specifications}}
 \printshorthands[heading=fdobibliography]

### 2023-06-19 latex/table1.tex

@@
   & \textcite{foundationDigitalObjectInterface} is transport-independent, commonly TLS TCP/IP port 9000, DOIP over HTTP \autocite{DOIPAPIHTTPa}
-  & HTTP/1.1 TCP/IP port 80 \autocite{rfc2616}, HTTP/1.1+TLS TCP/IP 443 \autocite{rfc2818}, HTTP/2 (as HTTP/1* but binary) \autocite{rfc7540}, HTTP/3 (like HTTP/2+TLS but UDP) \autocite{rfc9114} \\
+  & HTTP/1.1, TCP/IP port 80 \autocite{rfc2616}; HTTP/1.1+TLS, TCP/IP 443 \autocite{rfc2818}; HTTP/2, as HTTP/1* but binary \autocite{rfc7540}; HTTP/3, like HTTP/2+TLS but UDP \autocite{rfc9114} \\
 \textbf{Environmental}: \emph{how applications are deployed and affected by its environment, portability}
-  & Main DOIP implementation is \href{https://www.cordra.org/}{Cordra}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
+  & Main DOIP implementation is \href{https://www.cordra.org/}{\emph{Cordra}}, which can be single-instance or \href{https://www.cordra.org/documentation/configuration/distributed-deployment.html}{distributed}.
 Cordra \href{https://www.cordra.org/documentation/configuration/storage-backends.html}{storage backends} include file system, S3, MongoDB (itself scalable).
 Unique DOIP protocol can be hard to add to existing Web application frameworks, although proxy services have been developed (e.g.~B2SHARE adapter).
-  & HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to (for static content) multi-cloud Content-Delivery Networks (CDN).
+  & HTTP services widely deployed in a myriad of ways, ranging from single instance servers, horizontally \& vertically scaled application servers, to multi-cloud Content-Delivery Networks (CDN).
 Current scalable cloud technologies for Web hosting may not support HTTP features previously seen as important for Semantic Web, e.g.~content negotiation and semantic HTTP status codes.

### 2023-06-19 latex/table3.tex
more bib tweaks.

@@
+\newgeometry{left=4cm,%
+             right=2cm,%
+             top=2.25cm,%
+             bottom=2.25cm,%
+             headheight=12pt}
+
 \begin{landscape}
@@
   & Ensure FDO types will not be protocol-bound as DOIP might be updated/replaced
-  & HTTP stable for 30 years, Semantic Web for 20 years. \texttt{http://} URIs replaced by \texttt{https://}.
+  & HTTP stable for 30 years, Semantic Web for 20 years. \texttt{http://} URIs mostly replaced by \texttt{https://}.
   & Keep flexibility of RDF serialisation formats which may change more frequently \\
@@
   & Touched briefly, see Table \vref{tbl:fair-data-maturity-model}
-  & Add explicit expression to show each FAIR pcinciple fulfilled. \\
+  & Add explicit expression to show each FAIR principle fulfilled. \\
 G4: \emph{machine actionability}
@@
   & Operations discovered at runtime (\texttt{0.DOIP/Op.ListOperations}).
-  & Allow method discovery by type FDOs in advance (see PR-TypingFDOs-2.0-20220608).
+  & Allow method discovery by type FDOs in advance \autocite{fdo-TypingFDOs}.
   & HTTP methods discovered at runtime (\texttt{OPTIONS}), indempotent methods attempted directly. Unsupported methods reported using LDP constraints to human-readable text.
@@
   & Formalise standard process of FDO requirements \autocite{fdo-DocProcessStd}
-  & HTTP, LDP. FDOF not yet standardised
-  & Formalise FDOF from FDOF-SEM working group \\
+  & HTTP, LDP. However FDOF is not yet standardised.
+  & Formalise FDOF from FDOF-SEM working group. \\
 FDOF1: \emph{PID as basis}
   & Extensive use of Handle system.
-  & Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve \autocite{fdo-PIDProfileAttributes}. Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as PIDs.
+  & Clarify how local testing handles can be used during development, how ``temporary'' FDOs should evolve \autocite{fdo-PIDProfileAttributes}.
+Register \texttt{0.DOIP/*} and \texttt{0.FDO/*} as actual PIDs.
   & HTTP URLs as basis for identifiers, but they are frequently not persistent.
-  & Add strong guidance for PID services like w3id and persistence policies. \\
+  & Add strong guidance for PID services like w3id and persistence policies \autocite{McMurry_2017}. \\
 FDOF2: \emph{PID record w/ type}
@@
   & w3id/purl PIDs redirect without giving any metadata. Datacite DOIs content-negotiate to give registered metadata.
-  & Add FAIR Signposting at PID provider for minimal PID record \\
+  & Add FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022} at PID provider for minimal PID record \\
 FDOF3: \emph{PID resolvable to bytestream \& metadata}
@@
 FDOF4: \emph{Additional attributes}
-  & Freetext attribute keys. Attributes should be defined for FDO type (?).
+  & Freetext attribute keys. Attributes should be defined for FDO type.
   & Require that attribute keys should be PIDs (or have predefined mapping to PIDs). Explicitly allow attributes not already defined in type.
@@
 Metadata would need to be bundled within custom FDO types or attributes.
-Unclear how levels are separated within single FDO representation (need FDOF8?).
+Unclear how levels are separated within single FDO representation (may need FDOF8).
   & Declare which metadata are expected within response attribute or within FDO object. Require PIDs for custom attributes. Define how alternate metadata levels can be represented separately.
@@
 FDOF10: \emph{Metadata schemas by community}
-  & Metadata schemas are in practice managed on single CORDA server as local types, using JSON Schema.
+  & Metadata schemas are in practice managed on single Cordra server as local types, using JSON Schema.
   & Require types to be FDOs with registered PIDs, implement shared types.
@@
 \end{landscape}
+\restoregeometry

### 2023-06-19 latex/table4.tex

@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4179}}@{}}
-	\caption[Comparing FAIR Digital Object and Web technologies as middleware infrastructures]{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \cite{13TcbsZF6}) and Web technologies (using Linked Data) as middleware infrastructures \cite{hRzcHhPD}
+	\caption[Comparing FAIR Digital Object and Web technologies as middleware infrastructures]{Comparing FAIR Digital Object (with the DOIP 2.0 protocol \cite{foundationDigitalObjectInterface}) and Web technologies (using Linked Data) as middleware infrastructures \cite{zarrasComparisonFrameworkMiddleware2004a}
 \label{tbl:fdo-web-middleware}}\tabularnewline

### 2023-06-19 latex/table5.tex

@@
   & No, registries not required (except Data Type Registries). Handle registry only searchable by PID.
-  & Not specified
+  & ---
   & Not specified, several registries/catalogues for vocabularies/types (e.g. \autocite{NCBOBioPortal}). Indexing by search engines if exposing HTML w/schema.org. \\
@@
   & Required, but few client libraries
-  & & Ubiquitous, hundreds of HTTP libraries \\
+  & HTTP \texttt{GET}, content-negotiation for \texttt{fdof/object}
+  & Ubiquitous, hundreds of HTTP libraries \\
 RDA-A1.1-01M
@@
 It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}.
-    One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source).    
+    One DOIPv2 implementation (\href{https://www.cordra.org/}{Cordra}): free under BSD-like license (not recognised as Open Source).    
   & LDP is open W3C recommendation \autocite{w3-ldp}. \href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}.    
@@
   & FDOR9 FDOR10
-  & Unspecified (some CORDRA types add getProvenance methods). PID Kernel attributes? Unspecified W3C PROV-O, PAV
-  & & \\
+  & Unspecified (some Cordra types add getProvenance methods). PID Kernel attributes? 
+  & --- & W3C PROV-O, PAV \\
 RDA-R1.2-02M

### 2023-06-19 manuscript.tex
Merge branch 'main' of github.com:stain/2022-fdo-paper [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/6f7e00b05e91c1799cc3ea6c819f517ab11786ee.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/6f7e00b05e91c1799cc3ea6c819f517ab11786ee/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5314612773

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/4210fb857855271dcb3adb645fca755e429315e6/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/6f7e00b05e91c1799cc3ea6c819f517ab11786ee/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/4210fb857855271dcb3adb645fca755e429315e6}{stain/2022-fdo-paper@4210fb8}
-on June 18, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/6f7e00b05e91c1799cc3ea6c819f517ab11786ee}{stain/2022-fdo-paper@6f7e00b}
+on June 19, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 18 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 19 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 18 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 19 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 18 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 19 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 18 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 19 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 18 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 19 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 18 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 19 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 18 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 19 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 18 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 19 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 18 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 19 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 18 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 19 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 18 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 19 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 18 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 19 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 18 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 19 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 18 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 19 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 18 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 19 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 18 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 19 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 18 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 19 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 18 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 19 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 18 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 19 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 18 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 19 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 18 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 19 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 18 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 19 June 2023).
 

### 2023-06-22 html/extract_tables.sh
HTML version of tables

@@ -1,0 +1,18 @@
+#!/bin/sh
+set -e
+rm -f *tmp.html
+pandoc --version > /dev/null || ( >&2 echo "Try installing pandoc 2.19.2 with citeproc" ; exit 1 )
+
+for file in ../latex/table*tex ; do 
+	table=`basename $file | sed s/.tex//`
+	html=$table.tmp.html
+	pandoc --columns=500 -i $file --bibliography ../latex/bibliography.bib --citeproc -o $html
+	caption="$table: `grep '<caption>' table*.tmp.* | sed -E 's,</?[^>]*>,,g'`"
+	if grep -q endfirsthead $file; then 
+		# Remove double caption
+		sed -i '/tbody/,/tr>/d' $html
+	fi
+	cat prefix $html suffix | sed "s/{title}/$caption/" > $table.html
+	rm $html
+	echo Extracted $table.html from $file
+done

### 2023-06-22 html/extract_tables.sh
tables extracted from Latex (!)

@@ -6,12 +6,23 @@ pandoc --version > /dev/null || ( >&2 echo "Try installing pandoc 2.19.2 with ci
 for file in ../latex/table*tex ; do 
 	table=`basename $file | sed s/.tex//`
 	html=$table.tmp.html
-	pandoc --columns=500 -i $file --bibliography ../latex/bibliography.bib --citeproc -o $html
+	pandoc -M link-citations=true -M csl=https://s11.no/2021/house-rules/s11-citation-style.csl --columns=500 -i $file --bibliography ../latex/bibliography.bib --citeproc -o $html
 	caption="$table: `grep '<caption>' table*.tmp.* | sed -E 's,</?[^>]*>,,g'`"
 	if grep -q endfirsthead $file; then 
 		# Remove double caption
-		sed -i '/tbody/,/tr>/d' $html
+		sed -i '/<tbody/,/tr>/d' $html
 	fi
+	cat prefix $html suffix |\
+		# Add Cross-table references
+		sed 's/#tbl:fair-data-maturity-model/table5.html/' |\
+		sed 's/\[tbl:fair-data-maturity-model\]/Table 5/' |\
+		sed 's/#tbl:fdo-web-middleware/table4.html/' |\
+		sed 's/\[tbl:fdo-web-middleware\]/Table 4/' |\
+		# Insert linebreak in reference
+		sed 's,&lt;https://doi.org,<br />,' |\
+		# Insert title
+		sed "s/{title}/$caption/"   > $table.html
+	# Add CSS and charset
 	cat prefix $html suffix | sed "s/{title}/$caption/" > $table.html
 	rm $html
 	echo Extracted $table.html from $file

### 2023-06-22 html/extract_tables.sh
HTML tables

@@ -6,21 +6,17 @@ pandoc --version > /dev/null || ( >&2 echo "Try installing pandoc 2.19.2 with ci
 for file in ../latex/table*tex ; do 
 	table=`basename $file | sed s/.tex//`
 	html=$table.tmp.html
-	pandoc -M link-citations=true -M csl=https://s11.no/2021/house-rules/s11-citation-style.csl --columns=500 -i $file --bibliography ../latex/bibliography.bib --citeproc -o $html
+	pandoc -M link-citations=true --columns=600 -i $file --bibliography ../latex/bibliography.bib --citeproc -o $html
 	caption="$table: `grep '<caption>' table*.tmp.* | sed -E 's,</?[^>]*>,,g'`"
 	if grep -q endfirsthead $file; then 
 		# Remove double caption
 		sed -i '/<tbody/,/tr>/d' $html
 	fi
 	cat prefix $html suffix |\
-		# Add Cross-table references
-		sed 's/#tbl:fair-data-maturity-model/table5.html/' |\
-		sed 's/\[tbl:fair-data-maturity-model\]/Table 5/' |\
-		sed 's/#tbl:fdo-web-middleware/table4.html/' |\
-		sed 's/\[tbl:fdo-web-middleware\]/Table 4/' |\
-		# Insert linebreak in reference
-		sed 's,&lt;https://doi.org,<br />,' |\
-		# Insert title
+		sed 's/#tbl:fair-data-maturity-model/table5.html/g' |\
+		sed 's/\[tbl:fair-data-maturity-model\]/Table 5/g' |\
+		sed 's/#tbl:fdo-web-middleware/table4.html/g' |\
+		sed 's/\[tbl:fdo-web-middleware\]/Table 4/g' |\
 		sed "s/{title}/$caption/"   > $table.html
 	# Add CSS and charset
 	cat prefix $html suffix | sed "s/{title}/$caption/" > $table.html

### 2023-06-22 manuscript.tex
Merge branch 'main' of github.com:stain/2022-fdo-paper into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/6d9233550171f4d9f517b3a485faa60d68c64d34.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/6d9233550171f4d9f517b3a485faa60d68c64d34/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5340366129

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/6f7e00b05e91c1799cc3ea6c819f517ab11786ee/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/6d9233550171f4d9f517b3a485faa60d68c64d34/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/6f7e00b05e91c1799cc3ea6c819f517ab11786ee}{stain/2022-fdo-paper@6f7e00b}
-on June 19, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/6d9233550171f4d9f517b3a485faa60d68c64d34}{stain/2022-fdo-paper@6d92335}
+on June 22, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 19 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 21 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 19 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 21 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 19 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 21 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 19 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 21 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 19 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 21 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 19 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 21 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 19 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 21 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 19 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 21 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 19 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 21 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 19 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 21 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 19 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 21 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 19 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 21 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 19 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 21 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 19 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 21 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 19 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 21 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 19 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 21 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 19 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 21 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 19 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 21 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 19 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 21 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 19 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 21 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 19 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 21 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 19 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 21 June 2023).
 

### 2023-06-22 manuscript.tex
don't bundle s11's CSS as it refers to fonts [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/e60f158440d0393345a67a362a123fa00df7f8e1.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/e60f158440d0393345a67a362a123fa00df7f8e1/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5340376993

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/6d9233550171f4d9f517b3a485faa60d68c64d34/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/e60f158440d0393345a67a362a123fa00df7f8e1/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/6d9233550171f4d9f517b3a485faa60d68c64d34}{stain/2022-fdo-paper@6d92335}
+from \href{https://github.com/stain/2022-fdo-paper/tree/e60f158440d0393345a67a362a123fa00df7f8e1}{stain/2022-fdo-paper@e60f158}
 on June 22, 2023.

### 2023-06-22 manuscript.tex
local styles [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/b1393612506612279d1adce2c9bc387fde40c48f.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/b1393612506612279d1adce2c9bc387fde40c48f/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5340383730

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/e60f158440d0393345a67a362a123fa00df7f8e1/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/b1393612506612279d1adce2c9bc387fde40c48f/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/e60f158440d0393345a67a362a123fa00df7f8e1}{stain/2022-fdo-paper@e60f158}
+from \href{https://github.com/stain/2022-fdo-paper/tree/b1393612506612279d1adce2c9bc387fde40c48f}{stain/2022-fdo-paper@b139361}
 on June 22, 2023.

### 2023-06-22 manuscript.tex
arxiv [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/251c1d7a7c7a10ef9206e0a11ed9bb3d24dcf8b1.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/251c1d7a7c7a10ef9206e0a11ed9bb3d24dcf8b1/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5340408964

@@
+\emph{An updated preprint of this manuscript is available at \url{https://doi.org/10.48550/arXiv.2306.07436}
+-- these pages may not reflect the latest changes.}.
+
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/b1393612506612279d1adce2c9bc387fde40c48f/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/251c1d7a7c7a10ef9206e0a11ed9bb3d24dcf8b1/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/b1393612506612279d1adce2c9bc387fde40c48f}{stain/2022-fdo-paper@b139361}
+from \href{https://github.com/stain/2022-fdo-paper/tree/251c1d7a7c7a10ef9206e0a11ed9bb3d24dcf8b1}{stain/2022-fdo-paper@251c1d7}
 on June 22, 2023.

### 2023-06-22 manuscript.tex
title [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/5d37849fa90947eb93bbfe067f9489cc79aa07dd.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/5d37849fa90947eb93bbfe067f9489cc79aa07dd/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5340432870

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/251c1d7a7c7a10ef9206e0a11ed9bb3d24dcf8b1/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/5d37849fa90947eb93bbfe067f9489cc79aa07dd/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/251c1d7a7c7a10ef9206e0a11ed9bb3d24dcf8b1}{stain/2022-fdo-paper@251c1d7}
+from \href{https://github.com/stain/2022-fdo-paper/tree/5d37849fa90947eb93bbfe067f9489cc79aa07dd}{stain/2022-fdo-paper@5d37849}
 on June 22, 2023.

### 2023-06-22 manuscript.tex
build docx/latex again even if we dont' want it [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/5a0aa3288f4449ccea093fdf768529d4c690d942.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/5a0aa3288f4449ccea093fdf768529d4c690d942/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5343021478

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/5d37849fa90947eb93bbfe067f9489cc79aa07dd/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/5a0aa3288f4449ccea093fdf768529d4c690d942/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/5d37849fa90947eb93bbfe067f9489cc79aa07dd}{stain/2022-fdo-paper@5d37849}
+from \href{https://github.com/stain/2022-fdo-paper/tree/5a0aa3288f4449ccea093fdf768529d4c690d942}{stain/2022-fdo-paper@5a0aa32}
 on June 22, 2023.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 21 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 22 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 21 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 22 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 21 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 22 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 21 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 22 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 21 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 22 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 21 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 22 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 21 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 22 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 21 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 22 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 21 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 22 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 21 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 22 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 21 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 22 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 21 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 22 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 21 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 22 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 21 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 22 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 21 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 22 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 21 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 22 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 21 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 22 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 21 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 22 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 21 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 22 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 21 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 22 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 21 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 22 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 21 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 22 June 2023).
 

### 2023-06-23 latex/main.tex
Updates from Overleaf

--- 2023-06-18 latex/main.tex
@@
 \usepackage{pdflscape}
-%\usepackage[pass]{geometry}
 
@@
 These approaches -- FDO and Linked Data -- thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to compare FAIR Digital Objects with Linked Data and the Web architecture in the context of the discourse around FAIR data.
+Given their importance, in this article, we aim to compare FAIR Digital Objects, with Linked Data and the Web architecture in the context of the discourse around FAIR data.
 
@@
 The main DOIP operations are retrieving, creating and updating digital objects.
-These are mostly realised using the reference implementation Cordra \autocite{tupelo-schneckrobertBriefIntroductionCordra2022}.
+These are mostly realised using the reference implementation ( \href{https://cordra.org/}{Cordra}).
 FDO types are registered in the local Cordra instance, where they are specified using JSON Schema \autocite{Draftbhuttonjsonschema} and PIDs are assigned using the Handle system.
@@
 \item
-  Approaching the FDO principles from existing Linked Data practices on the Web, e.g.~WorkflowHub use of RO-Crate and schema.org \autocite{10.3897/rio.8.e93937}.
+  Approaching the FDO principles from existing Linked Data practices on the Web,e.g.~WorkflowHub use of RO-Crate and schema.org \autocite{10.3897/rio.8.e93937}.
 \end{itemize}
@@
 The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
-As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review.
+As of 2023-02-02, most of these documents are open for public review, while some are still in draft stages for internal review.
 As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide a brief summary of each:
@@
 See bibliography \vref*{sec:fdo-bibliography} for the citation per document above.
-It is worth pointing out that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
-Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are thus not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDOs are not yet in a single ecosystem.
+It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
+Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDO is not yet a single ecosystem.
 
@@
 
-The \textbf{Linked Data} (LD) concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect of the Semantic Web back into focus.
+The \textbf{Linked Data} concept \autocite{bizerLinkedDataStory2009a} was kickstarted as a set of best practices \autocite{LinkedDataDesign} to bring the Web aspect of the Semantic Web back into focus.
 Crucial to Linked Data is the \emph{reuse of existing URIs}, rather than making new identifiers.
@@
 A valid concern is that the Semantic Web research community has still not fully embraced the Web, and that the ``final 20\%'' engineering effort is frequently overlooked in favour of chasing new trends such as Big Data and AI, rather than making powerful Linked Data technologies available to the wider groups of Web developers \autocite{verborghSemanticWebIdentity2020a}.
-One bridging gap here by the Linked Data movement has been ``Linked Data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \autocite{wolstencroftRightFieldEmbeddingOntology2011b}, the use of Linked Data as part of REST Web APIs \autocite{pageRESTLinkedData2011}, and as shown by the big uptake by publishers to annotate the Web using schema.org \autocite{bernsteinNewLookSemantic2016a}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
+One bridging gap here by the Linked Data movement has been ``linked data by stealth'' approaches such as structured data entry spreadsheets powered by ontologies \autocite{wolstencroftRightFieldEmbeddingOntology2011b}, the use of Linked Data as part of REST Web APIs \autocite{pageRESTLinkedData2011}, and as shown by the big uptake by publishers to annotate the Web using schema.org \autocite{bernsteinNewLookSemantic2016a}, with vocabulary use patterns documented by copy-pastable JSON-LD examples, rather than by formalised ontologies or developer requirements to understand the full Semantic Web stack.
 
-Linked Data provides technologies that have evolved over time to satisfy its primary purpose of data interoperability.
-The needs to embrace the Web and developer experience have been central lessons learned.
+Linked Data provides a technologies that have evolved over time to satisfy its primary purpose of data interoperability.
+The need to embrace the Web and developer experience have been central lessons learned.
 In contrast, FDO is a new approach with many different potential paths forward, and having a partial overlap with the aims of Linked Data.
@@
 The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further considerations on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref{sec:next-step-fdo}).
+We have left out further comparisons on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref{sec:next-step-fdo}).
 
@@
 For conceptual comparisons we consider FAIR Digital Objects and the Web broadly.
-For implementations, we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice\footnote{For further background on FDO implemented with Linked Data see \autocite{FDOFramework,10.3897/rio.8.e94501}}.
+For implementations,we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice.
+
+\section*{Results}
 
@@
 The FAIR Digital Object guidelines \autocite{boninoFAIRDigitalObject} sets out recommendations for FDO implementations.
-Note that the proposed update to FDO specification \autocite{fdo-RequirementSpec} clarifies these definitions with equivalent identifiers\footnote{Newer \autocite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
+In Table \vref{tbl:fdo-checks} we evaluate the two current implementations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \autocite{FDOFramework}.
 
-In Table \vref{tbl:fdo-checks} we evaluate completeness of the guidelines in two current FDO realizations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}.
+Note that the draft update to FDO specification \autocite{fdo-RequirementSpec} clarifies these definitions with equivalent identifiers\footnote{Newer \autocite{fdo-RequirementSpec} renames \texttt{FDOF*} to \texttt{FDOR*} but follows same ordering.} and relates them to further FDO requirements such as FDO Data Type Registries.
 
@@
 Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement \autocite{fdo-DOIPEndorsement} set out that to comply, DOIP must be used according to the set of FDO requirement documents (details \vpageref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to actually comply as an FDO implementation.
+The DOIP Endorsement \autocite{fdo-DOIPEndorsement} sets out that to comply, DOIP must be used according to the set of FDO requirement documents (details \vpageref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
+Likewise, a Linked Data approach will need to follow the same requirements to comply as an FDO implementation.
 
@@
   While FDO recommends nested metadata FDOs (FDOF8, FDOF9), in practice this is not found (or linked with custom keys), particularly due to lack of namespaces and the favouring of local types rather than type/property re-use.
-Linked Data frequently have multiple representations, but often not sufficiently linked (link relation \texttt{alternate} \autocite{rfc8288}) or related (\texttt{prov:specializationOf} from \textcite{w3-prov-o}).
+Linked Data frequently have multiple representations, but often not sufficiently linked, perhaps \texttt{prov:specializationOf} \autocite{w3-prov-o}
 \item
-  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives. LDP has specific native support for containers.
+  FDO collections are not yet defined for DOIP, while Linked Data seemingly have too many alternatives, LDP has specific native support for containers.
 \item
@@
 
-\section*{EOSC Interoperability Framework}\label{sec:eosc-interoperability-framework}
+\subsection*{EOSC Interoperability Framework}\label{sec:eosc-interoperability-framework}
 
@@
 
-As covered in our introduction\vref{sec:introduction}, EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
-Therefore we here find it important to summarize how FDO and Linked Data can help satisfy the EOSC requirements.
+EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
+Therefore we here find it important to analyse how  FDO and Linked Data can help satisfy the EOSC requirements.
 
@@
 While these community aspects are not mandated by Linked Data practices, best practices have become established for aspects like ontology development \autocite{10.1186/s13326-021-00240-6}.
-EOSC IF's \emph{Technical} layer is likewise at a architecturally high level, such as service-level agreements, but also highlight PID policies which is strongly required by FDO, while Linked Data communities choose PID practices separately.
-The recommendations for the \emph{Semantic} layer is largely already implemented by Linked Data practices, yet for FDO mostly consist of encouragements.
+EOSC IF's technical layer is likewise at a architecturally high level, such as service-level agreements, but also highlight PID policies which is strongly required by FDO, while Linked Data communities choose PID practices separately.
+The recommendations for the Semantic layer, is largely already implemented by Linked Data practices, yet for FDO mostly consist of encouragements.
 For instance \emph{clear definitions of semantic concepts} is required by FDO guidelines, but how to technically define them has not been formalised by FDO specifications.
 
-The \emph{Legal} layer of interoperability is perhaps the one most emphasised by EOSC, by enabling collaboration across organizational barriers to joinly build a research infrastructure, but this is an area that both FDO and Linked Data are relatively weak in directly supporting.
-The EOSC IF recommendations in this layer are largely related to governance practices and metadata, for instance licensing, privacy and usage policies; these are also essential for cross-institutional and cross-repository access of FAIR objects.
+The Legal layer of interoperability is perhaps the one most emphasised by EOSC, by enabling collaboration across organizational barriers to joinly build a research infrastructure, but this is an area that both FDO and Linked Data are relatively weak in directly supporting.
+The EOSC IF recommendations in this layer are still largely related to governance practices and metadata, for instance licensing, privacy and usage policies; yet these are essential for cross-institutional and cross-repository access of FAIR objects.
 
@@
 When evaluating FDO implementations against the FDO guidelines (\vref*{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used.
-Achieving fully interoperable FAIR Digital Objects would require further convergence on implementation practices, and it is not given that this needs to diverge from the established Web architecture.
-It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional equably implementable restrictions and conventions, such as using persistent identifiers or pre-defining an object's type.
+Achieving fully interoperable FAIR digital objects would require further convergence on implementation practices, and it is not given that his need to diverge from the established Web architecture.
+It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional restrictions, equably implementable, such as using persistent identifiers or pre-defining an object's type.
 
 Considering this, by comparing FDO and Web as middleware (\vref*{sec:middleware}) we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g.
-lack of concurrency and transparancy.
+lack of concurrency transparancy.
 Recent updates to HTTP have added many features needed for large-scale usage such as video streaming services (e.g.
-caching, multiplexing, cloud deployments), and having the option to transparently apply these also to FDOs seems like a strong incentive.
-Many programmatic features for distributed objects are however missing or needing custom extensions in both aspects, such as transactions, asynchronous operations and streaming.
+caching, multiplexing, cloud deployments), and having the option to transparantly apply these also to FDOs seems like a strong incentive.
+Many programmatic features are however missing or needing custom extensions in both aspects, such as transactions, asynchronous operations and streaming.
 
@@
 material science).
-It may be advantageous for both these effort to be aligned and jointly promoted under the EOSC umbrella.
+It may be advantegous for both these effort to be aligned and jointly promoted under the EOSC umbrella.
 
@@
 We therefore identify the need for a new explicit FDO profile of Linked Data that sets pragmatic constraints and stronger recommendations for consistent and developer-friendly deployment of digital objects.
-Such a combination of efforts could utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow an ecosystem of machine-actionable objects.
+Such a combination of efforts could utillise both the benefits of mature Semantic Web technologies (e.g.~federated knowledge graph queries and rich validation) and data management practices that follow FDO guidance in order to grow a ecosystem of machine-actionable objects.
 It is beyond the scope of this work to detail such a profile, but we suggest the following potential key aspects:
@@
   \item Types, attributes and actions are self-defined by their identifier
-  \item Use Web approaches directly where possible, rather than wrap in a new model
 \end{itemize}
@@
 
-\hypertarget{conclusion}{%
-\section*{Conclusion}\label{conclusion}}
+\section*{Conclusion}\label{conclusion}
 
@@
 The FDO recommendations show that FAIR thinking in this regard need to move beyond data publishing and into machine actionability across digital objects, and with broader community consensus. 
-As flexibility for extensions is a necessary ingredient alongside rigidity for core concepts, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as learnt by early Semantic Web adopters.
+As flexibility for extensions is a necessary ingredient alongside rigidity for core concepts, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as the early Semantic Web adopters.
 
@@
 
-\clearpage
 \defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}

### 2023-06-23 manuscript.tex
Merge overleaf-2023-06-23-1355 into main [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/18c218220087b13fe0db935b3c1a320647449330.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/18c218220087b13fe0db935b3c1a320647449330/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5357134142

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/5a0aa3288f4449ccea093fdf768529d4c690d942/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/18c218220087b13fe0db935b3c1a320647449330/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/5a0aa3288f4449ccea093fdf768529d4c690d942}{stain/2022-fdo-paper@5a0aa32}
-on June 22, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/18c218220087b13fe0db935b3c1a320647449330}{stain/2022-fdo-paper@18c2182}
+on June 23, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 22 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 23 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 22 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 23 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 22 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 23 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 22 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 23 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 22 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 23 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 22 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 23 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 22 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 23 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 22 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 23 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 22 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 23 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 22 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 23 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 22 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 23 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 22 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 23 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 22 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 23 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 22 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 23 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 22 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 23 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 22 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 23 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 22 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 23 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 22 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 23 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 22 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 23 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 22 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 23 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 22 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 23 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 22 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 23 June 2023).
 

### 2023-06-23 manuscript.tex
cite Van de Sompel 2015 for signposting [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/196b2c0a09b766bbcc026f7504f34a805280a9fc.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/196b2c0a09b766bbcc026f7504f34a805280a9fc/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5357287070

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/18c218220087b13fe0db935b3c1a320647449330/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/196b2c0a09b766bbcc026f7504f34a805280a9fc/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/18c218220087b13fe0db935b3c1a320647449330}{stain/2022-fdo-paper@18c2182}
+from \href{https://github.com/stain/2022-fdo-paper/tree/196b2c0a09b766bbcc026f7504f34a805280a9fc}{stain/2022-fdo-paper@196b2c0}
 on June 23, 2023.

### 2023-06-23 manuscript.tex
now better description (if a bit long!) [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/98a96b025d3d28782653c0302c3b390cab049c0c.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/98a96b025d3d28782653c0302c3b390cab049c0c/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5357449612

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/196b2c0a09b766bbcc026f7504f34a805280a9fc/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/98a96b025d3d28782653c0302c3b390cab049c0c/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/196b2c0a09b766bbcc026f7504f34a805280a9fc}{stain/2022-fdo-paper@196b2c0}
+from \href{https://github.com/stain/2022-fdo-paper/tree/98a96b025d3d28782653c0302c3b390cab049c0c}{stain/2022-fdo-paper@98a96b0}
 on June 23, 2023.

### 2023-06-24 latex/main.tex

--- 2023-06-18 latex/main.tex
@@
 \textbf{Configuration Types} \autocite{fdo-ConfigurationTypes} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022}.
+Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompel2015,vandesompelFAIRSignpostingProfile2022}.
 
@@
 For implementations, we contrast the main FDO realisation using the DOIPv2 protocol \autocite{foundationDigitalObjectInterface} against Linked Data as implemented in general practice\footnote{For further background on FDO implemented with Linked Data see \autocite{FDOFramework,10.3897/rio.8.e94501}}.
+
+\section*{Results}
 
@@
 For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
+Signposting \autocite{vandesompel2015} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
 In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.

### 2023-06-24 manuscript.tex
citations etc [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/9677b4fb32c5e4e6351e27304de42c115a379172.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/9677b4fb32c5e4e6351e27304de42c115a379172/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5366463172

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/98a96b025d3d28782653c0302c3b390cab049c0c/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/9677b4fb32c5e4e6351e27304de42c115a379172/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/98a96b025d3d28782653c0302c3b390cab049c0c}{stain/2022-fdo-paper@98a96b0}
-on June 23, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/9677b4fb32c5e4e6351e27304de42c115a379172}{stain/2022-fdo-paper@9677b4f}
+on June 24, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 23 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 23 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 23 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 23 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 23 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 23 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 23 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 24 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 23 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 24 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 23 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 23 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 23 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 23 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 23 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 24 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 23 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 23 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 23 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 23 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 23 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 23 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 23 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 23 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 23 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 June 2023).
 

### 2023-06-24 manuscript.tex
html rendering [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/f697e5dfb9f4502a9486a03fbf56a5d69e0a6b56.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/f697e5dfb9f4502a9486a03fbf56a5d69e0a6b56/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5366467035

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/9677b4fb32c5e4e6351e27304de42c115a379172/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/f697e5dfb9f4502a9486a03fbf56a5d69e0a6b56/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/9677b4fb32c5e4e6351e27304de42c115a379172}{stain/2022-fdo-paper@9677b4f}
+from \href{https://github.com/stain/2022-fdo-paper/tree/f697e5dfb9f4502a9486a03fbf56a5d69e0a6b56}{stain/2022-fdo-paper@f697e5d}
 on June 24, 2023.

### 2023-06-25 manuscript.tex
finished SKOS but not mapping [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/9ff6f0da1c2aa50a163efffac0fd16d3f3c6b144.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/9ff6f0da1c2aa50a163efffac0fd16d3f3c6b144/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5367047177

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/f697e5dfb9f4502a9486a03fbf56a5d69e0a6b56/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/9ff6f0da1c2aa50a163efffac0fd16d3f3c6b144/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/f697e5dfb9f4502a9486a03fbf56a5d69e0a6b56}{stain/2022-fdo-paper@f697e5d}
-on June 24, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/9ff6f0da1c2aa50a163efffac0fd16d3f3c6b144}{stain/2022-fdo-paper@9ff6f0d}
+on June 25, 2023.
 

### 2023-06-26 latex/main.tex
Updates from Overleaf

@@
 \textbf{Configuration Types} \autocite{fdo-ConfigurationTypes} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompel2015,vandesompelFAIRSignpostingProfile2022}.
+Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022}.
 
@@
 The Interoperability Framework for Fast Data also provides a brief \emph{metamodel} which we use in Table \vref{tbl:metamodel-concepts} to map and examplify corresponding concepts in FDO's DOIP realization and the Web using HTTP semantics \autocite{rfc9110}.
+A SKOS mapping \autocite{w3-skos-primer} corresponding to this table is available as an RO-Crate \autocite{soilandreyes-fdo-comparison-2023}.
 
@@
 For instance in Linked Data, given a URL of a CSV file, what is its persistant identifier or license information?
-Signposting \autocite{vandesompel2015} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
+FAIR Signposting \autocite{vandesompelFAIRSignpostingProfile2022} can improve findability of metadata using HTTP Link relations, which enable an FDO-like overlay for any HTTP resource.
 In DOIP, responses for bytestreams can include the data identifier: if that is a PID (not enforced by DOIP), its metadata is accessible.

### 2023-06-26 manuscript.tex
RO-Crate as published on https://doi.org/10.5281/zenodo.8075229 [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/1fa38a4c54cf973f779ddf5f0320fe35cb36674f.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/1fa38a4c54cf973f779ddf5f0320fe35cb36674f/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5381412896

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/9ff6f0da1c2aa50a163efffac0fd16d3f3c6b144/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/1fa38a4c54cf973f779ddf5f0320fe35cb36674f/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/9ff6f0da1c2aa50a163efffac0fd16d3f3c6b144}{stain/2022-fdo-paper@9ff6f0d}
-on June 25, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/1fa38a4c54cf973f779ddf5f0320fe35cb36674f}{stain/2022-fdo-paper@1fa38a4}
+on June 26, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 24 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 26 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 24 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 26 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 24 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 26 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 24 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 26 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 24 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 26 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 24 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 26 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 24 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 26 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 24 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 26 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 24 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 26 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 24 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 26 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 24 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 26 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 24 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 26 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 24 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 26 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 24 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 26 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 24 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 26 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 24 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 26 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 24 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 26 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 24 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 26 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 24 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 26 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 24 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 26 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 24 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 26 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 24 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 26 June 2023).
 

### 2023-06-26 manuscript.tex
more pandoc [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/4b65e98117b1d8009b403ace5f9174d4e25dec32.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/4b65e98117b1d8009b403ace5f9174d4e25dec32/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5381409955

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/1fa38a4c54cf973f779ddf5f0320fe35cb36674f/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/4b65e98117b1d8009b403ace5f9174d4e25dec32/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/1fa38a4c54cf973f779ddf5f0320fe35cb36674f}{stain/2022-fdo-paper@1fa38a4}
+from \href{https://github.com/stain/2022-fdo-paper/tree/4b65e98117b1d8009b403ace5f9174d4e25dec32}{stain/2022-fdo-paper@4b65e98}
 on June 26, 2023.

### 2023-06-26 manuscript.tex
Updates from Overleaf [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/cbf2f8a9960d492b42521c3808995ff47b2311f3.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/cbf2f8a9960d492b42521c3808995ff47b2311f3/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5381473174

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/4b65e98117b1d8009b403ace5f9174d4e25dec32/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/cbf2f8a9960d492b42521c3808995ff47b2311f3/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/4b65e98117b1d8009b403ace5f9174d4e25dec32}{stain/2022-fdo-paper@4b65e98}
+from \href{https://github.com/stain/2022-fdo-paper/tree/cbf2f8a9960d492b42521c3808995ff47b2311f3}{stain/2022-fdo-paper@cbf2f8a}
 on June 26, 2023.

### 2023-06-26 latex/main.tex
footnotes

--- 2023-06-24 latex/main.tex
@@
 
-From this mapping we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
+From this mapping\footnote{An equivalent SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{stian_soiland_reyes_2023_8075230}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
 Notable are that neither DOIP or HTTP have strong support for transactions (explored further \vpageref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.

### 2023-06-26 manuscript.tex
footnotes [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/51ed4a0f879b511074d60b75dcd5ef10837b4eba.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/51ed4a0f879b511074d60b75dcd5ef10837b4eba/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5381976871

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/cbf2f8a9960d492b42521c3808995ff47b2311f3/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/51ed4a0f879b511074d60b75dcd5ef10837b4eba/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/cbf2f8a9960d492b42521c3808995ff47b2311f3}{stain/2022-fdo-paper@cbf2f8a}
+from \href{https://github.com/stain/2022-fdo-paper/tree/51ed4a0f879b511074d60b75dcd5ef10837b4eba}{stain/2022-fdo-paper@51ed4a0}
 on June 26, 2023.

### 2023-06-26 latex/main.tex
Data Curation as well

@@
 \item[Stian Soiland-Reyes]
-Conceptualization, Formal Analysis, Funding acquisition, Investigation, Methodology,
+Conceptualization, Data curation, Formal Analysis, Funding acquisition, Investigation, Methodology,
 Writing -- original draft, Writing -- review \& editing

### 2023-06-26 manuscript.tex
Data Curation as well [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/a4a64794fe7e965163e5f2ac58d0c097945ec124.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/a4a64794fe7e965163e5f2ac58d0c097945ec124/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5382530328

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/51ed4a0f879b511074d60b75dcd5ef10837b4eba/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/a4a64794fe7e965163e5f2ac58d0c097945ec124/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/51ed4a0f879b511074d60b75dcd5ef10837b4eba}{stain/2022-fdo-paper@51ed4a0}
+from \href{https://github.com/stain/2022-fdo-paper/tree/a4a64794fe7e965163e5f2ac58d0c097945ec124}{stain/2022-fdo-paper@a4a6479}
 on June 26, 2023.

### 2023-06-26 latex/main.tex
avoid _ in bib

@@
 
-From this mapping\footnote{An equivalent SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{stian_soiland_reyes_2023_8075230}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
+From this mapping\footnote{An equivalent SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{soilandreyes2023}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
 Notable are that neither DOIP or HTTP have strong support for transactions (explored further \vpageref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.

### 2023-06-26 manuscript.tex
avoid _ in bib [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/d3100d949030a0fa261397bd204b045bdfde8861.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/d3100d949030a0fa261397bd204b045bdfde8861/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5382597999

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/a4a64794fe7e965163e5f2ac58d0c097945ec124/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/d3100d949030a0fa261397bd204b045bdfde8861/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/a4a64794fe7e965163e5f2ac58d0c097945ec124}{stain/2022-fdo-paper@a4a6479}
+from \href{https://github.com/stain/2022-fdo-paper/tree/d3100d949030a0fa261397bd204b045bdfde8861}{stain/2022-fdo-paper@d3100d9}
 on June 26, 2023.

### 2023-06-26 manuscript.tex
no md here [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/2bddf6d7be9adedcbbfc3d6028a6d845d6e72b0f.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/2bddf6d7be9adedcbbfc3d6028a6d845d6e72b0f/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5382646945

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/d3100d949030a0fa261397bd204b045bdfde8861/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/2bddf6d7be9adedcbbfc3d6028a6d845d6e72b0f/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/d3100d949030a0fa261397bd204b045bdfde8861}{stain/2022-fdo-paper@d3100d9}
+from \href{https://github.com/stain/2022-fdo-paper/tree/2bddf6d7be9adedcbbfc3d6028a6d845d6e72b0f}{stain/2022-fdo-paper@2bddf6d}
 on June 26, 2023.

### 2023-06-26 manuscript.tex
typo [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/be8675b6907eabfc339d356014312f2b2aa8292e.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/be8675b6907eabfc339d356014312f2b2aa8292e/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5383793295

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/2bddf6d7be9adedcbbfc3d6028a6d845d6e72b0f/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/be8675b6907eabfc339d356014312f2b2aa8292e/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/2bddf6d7be9adedcbbfc3d6028a6d845d6e72b0f}{stain/2022-fdo-paper@2bddf6d}
+from \href{https://github.com/stain/2022-fdo-paper/tree/be8675b6907eabfc339d356014312f2b2aa8292e}{stain/2022-fdo-paper@be8675b}
 on June 26, 2023.

### 2023-06-27 manuscript.tex
ro-crate of this article [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/7252a9e9e0e603e16c53d583f5a2b6ad2a2a60ac.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/7252a9e9e0e603e16c53d583f5a2b6ad2a2a60ac/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5384402598

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/be8675b6907eabfc339d356014312f2b2aa8292e/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/7252a9e9e0e603e16c53d583f5a2b6ad2a2a60ac/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/be8675b6907eabfc339d356014312f2b2aa8292e}{stain/2022-fdo-paper@be8675b}
-on June 26, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/7252a9e9e0e603e16c53d583f5a2b6ad2a2a60ac}{stain/2022-fdo-paper@7252a9e}
+on June 27, 2023.
 

### 2023-06-28 latex/main.tex
ro-crate of this article
typos

@@
 We conclude with recommendations for both Linked Data and FDO communities to further their adaptation and alignment.
+
+An \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} is archived at \url{https://doi.org/10.5281/zenodo.8075229}
 \end{abstract}
@@
 
-FDO is an evolving concept. A set of FDO Demonstrators \autocite{wittenburgFAIRDigitalObject2022b} highlight how current adapters are approaching implementations of FDO from different angles:
+FDO is an evolving concept. A set of FDO Demonstrators \autocite{wittenburgFAIRDigitalObject2022b} highlights how current adapters are approaching implementations of FDO from different angles:
 
@@
 In theory, the use of RDF by the life sciences would enable interoperability between the many data repositories and support combined views of the many aspects of bio-entities -- however in practice most institutions ended up making their own ontologies and identifiers, for what to the untrained eye would mean roughly the same.
-One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~\emph{should a gene identifier signify just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?}) lead to large differences in representational granularity, and thus needed different identifiers.
+One can argue that the toll of adding the semantic logic system of rich ontologies meant that small, but fundamental, differences in opinion (e.g.~\emph{should a gene identifier signify just the particular DNA sequence letters, or those letters as they appear in a particular position on a human chromosome?}) led to large differences in representational granularity, and thus needed different identifiers.
 

### 2023-06-28 manuscript.tex
typos [ci skip]
This build is based on
https://github.com/stain/2022-fdo-paper/commit/bacb841e135816c2ab2983795ce61290c98c0fb1.

This commit was created by the following CI build and job:
https://github.com/stain/2022-fdo-paper/commit/bacb841e135816c2ab2983795ce61290c98c0fb1/checks
https://github.com/stain/2022-fdo-paper/actions/runs/5406436632

@@
 This manuscript
-(\href{https://stain.github.io/2022-fdo-paper/v/7252a9e9e0e603e16c53d583f5a2b6ad2a2a60ac/}{permalink})
+(\href{https://stain.github.io/2022-fdo-paper/v/bacb841e135816c2ab2983795ce61290c98c0fb1/}{permalink})
 was automatically generated
-from \href{https://github.com/stain/2022-fdo-paper/tree/7252a9e9e0e603e16c53d583f5a2b6ad2a2a60ac}{stain/2022-fdo-paper@7252a9e}
-on June 27, 2023.
+from \href{https://github.com/stain/2022-fdo-paper/tree/bacb841e135816c2ab2983795ce61290c98c0fb1}{stain/2022-fdo-paper@bacb841}
+on June 28, 2023.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1BlxdxOwa}{}}%
-14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 26 June 2023b).
+14:00-17:00.ISO 16684-1:2019. \emph{Available at} \href{https://www.iso.org/standard/75163.html}{\emph{https://www.iso.org/standard/75163.html}} (accessed 28 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-11dInoyVF}{}}%
-5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 26 June 2023).
+5-star Open Data. \emph{Available at} \href{http://5stardata.info/en/}{\emph{http://5stardata.info/en/}} (accessed 28 June 2023).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 2.
-\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 26 June 2023b).
+\emph{Available at} \href{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}{\emph{https://www.w3.org/TR/2020/REC-vocab-dcat-2-20200204/}} (accessed 28 June 2023b).
 
@@
 Data Catalog Vocabulary (DCAT) - Version 3.
-\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 26 June 2023).
+\emph{Available at} \href{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}{\emph{https://www.w3.org/TR/2022/WD-vocab-dcat-3-20220510/}} (accessed 28 June 2023).
 
@@
 Data Information View.
-\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 26 June 2023).
+\emph{Available at} \href{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}{\emph{https://handle-esgf.dkrz.de/lp/21.14100/2fcf49d3-0608-3373-a47f-0e721b7eaa87}} (accessed 28 June 2023).
 
@@
 2020.
-\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 26 June 2023).
+\emph{Available at} \href{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}{\emph{https://www.dublincore.org/specifications/dublin-core/dcmi-terms/2020-01-20/}} (accessed 28 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-68fzID2k}{}}%
-DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 26 June 2023).
+DOI® Handbook. \emph{Available at} \href{https://www.doi.org/the-identifier/resources/handbook/}{\emph{https://www.doi.org/the-identifier/resources/handbook/}} (accessed 28 June 2023).
 
@@
 DOI® Resolution Documentation.
-\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 26 June 2023).
+\emph{Available at} \href{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}{\emph{https://www.doi.org/the-identifier/resources/factsheets/doi-resolution-documentation}} (accessed 28 June 2023).
 
@@
 DOIP API for HTTP Clients --- Cordra documentation.
-\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 26 June 2023).
+\emph{Available at} \href{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}{\emph{https://www.cordra.org/documentation/api/doip-api-for-http-clients.html}} (accessed 28 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-1CvYRVhy2}{}}%
-Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 26 June 2023).
+Handle.Net Registry. \emph{Available at} \href{https://www.handle.net/download_hnr.html}{\emph{https://www.handle.net/download\_hnr.html}} (accessed 28 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-13PaKBrZe}{}}%
-Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 26 June 2023).
+Hydra W3C Community Group. \emph{Available at} \href{https://www.hydra-cg.com/}{\emph{https://www.hydra-cg.com/}} (accessed 28 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nW4HY8Nq}{}}%
-"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 26 June 2023).
+"info" URI Registry (Frozen). \emph{Available at} \href{https://oclc-research.github.io/infoURI-Frozen/}{\emph{https://oclc-research.github.io/infoURI-Frozen/}} (accessed 28 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xVJMmGVr}{}}%
-Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 26 June 2023).
+Licenses. 2022. \emph{Available at} \href{https://opensource.org/licenses/}{\emph{https://opensource.org/licenses/}} (accessed 28 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-QpDmajc4}{}}%
-Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 26 June 2023).
+Linking Across Provenance Bundles. \emph{Available at} \href{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}{\emph{https://www.w3.org/TR/2013/NOTE-prov-links-20130430/}} (accessed 28 June 2023).
 
@@
 2023.Web API design best practices - Azure Architecture Center.
-\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 26 June 2023).
+\emph{Available at} \href{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}{\emph{https://learn.microsoft.com/en-us/azure/architecture/best-practices/api-design}} (accessed 28 June 2023).
 
@@
 ORE Specification - Abstract Data Model.
-\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 26 June 2023).
+\emph{Available at} \href{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}{\emph{http://www.openarchives.org/ore/1.0/datamodel\#Proxies}} (accessed 28 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-gtGuSevK}{}}%
-ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 26 June 2023).
+ORE User Guide - Primer. \emph{Available at} \href{http://www.openarchives.org/ore/1.0/primer}{\emph{http://www.openarchives.org/ore/1.0/primer}} (accessed 28 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-9T4j3N4e}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 26 June 2023a).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/2013/REC-prov-o-20130430/}{\emph{https://www.w3.org/TR/2013/REC-prov-o-20130430/}} (accessed 28 June 2023a).
 
 \leavevmode\vadjust pre{\hypertarget{ref-ZUgwvAHE}{}}%
-PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 26 June 2023b).
+PROV-O: The PROV Ontology. \emph{Available at} \href{https://www.w3.org/TR/prov-o/\#specializationOf}{\emph{https://www.w3.org/TR/prov-o/\#specializationOf}} (accessed 28 June 2023b).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-voDQ6diM}{}}%
-The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 26 June 2023).
+The ARK Identifier Scheme. \emph{Available at} \href{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}{\emph{https://datatracker.ietf.org/doc/id/draft-kunze-ark.html}} (accessed 28 June 2023).
 
 \leavevmode\vadjust pre{\hypertarget{ref-hxtgCGjY}{}}%
-The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 26 June 2023).
+The Modern Standards Paradigm - Five Key Principles. \emph{Available at} \href{https://open-stand.org/about-us/principles/}{\emph{https://open-stand.org/about-us/principles/}} (accessed 28 June 2023).
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-10HA0mMBx}{}}%
-Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 26 June 2023).
+Unknown.Introducing 'Role'. \emph{Available at} \href{http://blog.schema.org/2014/06/introducing-role.html}{\emph{http://blog.schema.org/2014/06/introducing-role.html}} (accessed 28 June 2023).
 

### 2023-11-15 latex/main.tex
Typo/grammar fixes

@@
 In particular, the European Open Science Cloud (\href{https://www.eosc.eu/}{EOSC}) has promoted adaptation of FAIR data sharing of data resources across electronic research infrastructures \autocite{monsCloudyIncreasinglyFAIR2017b}.
-The EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of rich digital objects.
+The EOSC Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b} puts particular emphasis on how interoperability can be achieved technically, semantically, organisationally, and legally -- laying out a vision of how data, publication, software and services can work together to form an ecosystem of digital objects that are extensively described.
+Such descriptions for interoperability connect a range of information -- from protocols and presentations, to hardware designs and scientific workflows, including extensive metadata of the information itself.
 
@@
 These approaches -- FDO and Linked Data -- thus, form two of the major avenues for allowing developers and the wider research community to achieve the goal of FAIR data.
-Given their importance, in this article, we aim to compare FAIR Digital Objects with Linked Data and the Web architecture in the context of the discourse around FAIR data.
+Given their importance, in this article, we aim to compare FDO with Linked Data and the Web architecture in the context of the discourse around FAIR data.
 
 Concretely, the contribution of this paper is {\bf a systematic comparison between FDO and Linked Data using 5 different conceptual frameworks} that capture different perspectives on interoperability and readiness for implementation.
+
+The rest of this paper is organised as follows: We begin with a background primer on FDO and Linked Data to provide a foundation for the work.
+In the Methods section, we introduce the conceptual frameworks we use for comparison.
+Subsequently, in the Results section, we systematically step through the outcomes of applying these frameworks to both FDO and Linked Data.
+For each framework, we derive key observations.
+We end with a discussion of these results and their implications for both approaches and conclude.
 
@@
 
-The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
+The concept of \textbf{FAIR Digital Objects} \autocite{schultesFAIRPrinciplesDigital2019a} has been introduced as a way to expose research data as active objects that conform to the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
 This builds on the \emph{Digital Object} (DO) concept \autocite{kahnFrameworkDistributedDigital2006b}, first introduced by \textcite{kahnFrameworkDistributedDigital1995a} as a system of \emph{repositories} containing \emph{digital objects} identified by \emph{handles} and described by \emph{metadata} which may have references to other handles.
@@
 The premise of systematically building an ecosystem of such digital objects is to give researchers a way to organise complex digital entities, associated with identifiers, metadata, and supporting automated processing \autocite{wittenburgDigitalObjectsDrivers2019a}.
+As mentioned previously, this ecoystem is envisioned to consist of a wide variety of digital entities and contextual information ranging from software to articles to even descriptions of experimental infrastructures \autocite{Azeroual2022PuttingFP}.
+Recently, it has been noted that the practical use of FDOs to achieve interoperability requires governance in particular with respect to assessing such interoperability \autocite{Wilkinson2023}.
 
-Recently, FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
+FDOs have been recognised by the European Open Science Cloud (\href{https://eosc.eu/}{EOSC}) as a suggested part of its Interoperability Framework \autocite{corchoEOSCInteroperabilityFramework2021b}, in particular for deploying active and interoperable FAIR resources that are \emph{machine actionable}.
 Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}.
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents (see \emph{Next steps for FDO} \vpageref{sec:next-step-fdo}).
+The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents.
 
@@
 As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review.
-As these documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}, we provide a brief summary of each:
+We provide an overview of these documents in Appendix \ref{sec:appendixfdo}.
+These documents clarify the future aims and focus of FAIR Digital Objects \autocite{fdo-Roadmap}.
+Except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
+Going forward a key strategy of the Forum is the use of profiles to help define specific attributes in metadata that are necessary for domains or application contexts.
+However, these are not yet fully implemented in the implementations considered here.
 
-\textbf{FAIR Digital Object Overview and Specifications} \autocite{fdo-Overview} is a comprehensive overview of FAIR Digital Object specifications listed below.
-It serves as a primer that introduces FDO concepts and the remaining documents.
-It is accompanied by an FDO Glossary \autocite{fdo-Glossary}.
-
-The \textbf{FDO Forum Document Standards} \autocite{fdo-DocProcessStd} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
-In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
-
-The \textbf{FDO Requirement Specifications} \autocite{fdo-RequirementSpec} is an update of \autocite{boninoFAIRDigitalObject} as the foundational definition of FDO.
-This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
-The requirements shown in Table \vref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
-
-The \textbf{Machine actionability} \autocite{fdo-MachineActionDef} sets out to define what is meant by \emph{machine actionability} for FDOs.
-\emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
-The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.
-
-\textbf{Configuration Types} \autocite{fdo-ConfigurationTypes} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
-Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompel2015,vandesompelFAIRSignpostingProfile2022}.
-
-\textbf{PID Profiles \& Attributes} \autocite{fdo-PIDProfileAttributes} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
-This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \autocite{weigelRDARecommendationPID2018}.
-This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record} \autocite{islam_2023}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
-
-\textbf{Kernel Attributes \& Metadata} \autocite{fdo-KernelAttributes} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
-This document expands on RDA Recommendation on PID Kernel Information \autocite{weigelRDARecommendationPID2018}.
-It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
-
-\textbf{Granularity, Versioning, Mutability} \autocite{fdo-Granularity} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
-The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
-
-\textbf{DOIP Endorsement Request} \autocite{fdo-DOIPEndorsement} is an endorsement of the DOIP v2.0 \autocite{foundationDigitalObjectInterface} specification as a potential FDO implementation, as it has been applied by several institutions \autocite{wittenburgFAIRDigitalObject2022b}.
-The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
-
-\textbf{Upload of FDO} \autocite{fdo-FDO-Upload} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
-ResourceSync \autocite{ResourceSyncFrameworkSpecification} is suggested as one type of service to list FDOs.
-This document highlights potential practices by repositories and their clients, without adding any particular requirements.
-
-\textbf{Typing FAIR Digital Objects} \autocite{fdo-TypingFDOs} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
-This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
-Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work.
-How to define such FDO operations is not specified.
-
-\textbf{Implementation of Attributes, Types, Profiles and Registries} \autocite{fdo-ImplAttributesTypesProfiles} details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
-This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
-Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} \autocite{schwardmannTwoExamplesHow2022}.
-
-See bibliography \vref*{sec:fdo-bibliography} for the citation per document above.
-It is worth pointing out that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.
 Existing FDO implementations \autocite{wittenburgFAIRDigitalObject2022b} are thus not fully consolidated in choices such as protocols, type systems and serialisations -- this divergence and corresponding additional technical requirements mean that FDOs are not yet in a single ecosystem.
@@
 
-\section*{Method}
+\section*{Methods}\label{sec:methods}
 
@@
 
+Conceptual framework 1, 3, 5 consider more general views of interoperability between systems, whereas frameworks 2 and 4 are developed specifically for addressing FAIR principles.
+
 The reason for this wide-ranged comparison is to exercise the different dimensions that together form FAIR Digital Objects: Data, Metadata, Service, Access, Operations, Computation.
-We have left out further considerations on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community (as detailed \vpageref{sec:next-step-fdo}).
+We have left out further considerations on type systems, persistent identifiers and social aspects as principles and practices within these dimensions are still taking form within the FDO community.
 
@@
 
-\section*{Results}
+For all our comparisons, our process was to perform a mapping between the relevant specifications and/or implementation and the given conceptual model through detailed reading of the defining documents.
+We aim in all cases for traceability between the given specification and our mapping such that readers can validate our analysis.
+
+\section*{Results}\label{sec:results}
 
@@
 
+\subsubsection*{Observations}
 Based on the analysis shown in Table \ref{tbl:fdo-web-interoperability-framework}, we draw the following conclusions:
@@
 
-From this mapping\footnote{An equivalent SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{soilandreyes2023}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
+From this mapping\footnote{A SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{soilandreyes2023}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
 Notable are that neither DOIP or HTTP have strong support for transactions (explored further \vpageref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
@@
 
-In Table \vref{tbl:fdo-checks} we evaluate completeness of the guidelines in two current FDO realizations, using DOIPv2 \autocite{foundationDigitalObjectInterface} and using Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}.
+In Table \vref{tbl:fdo-checks}, we evaluate completeness of the guidelines in two current FDO realisations: 1) DOIPv2 \autocite{foundationDigitalObjectInterface} and 2) Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}.
+We provide our analysis of each realization with respect to the FDO Guidline and also provide suggestions for that realization to meet the given guideline.
 
-A key observation from this is that simply using DOIP does not achieve many of the FDO guidelines.
-Rather the guidelines set out how a protocol like DOIPs should be used to achieve FAIR Digital Object goals.
-The DOIP Endorsement \autocite{fdo-DOIPEndorsement} set out that to comply, DOIP must be used according to the set of FDO requirement documents (details \vpageref{sec:next-step-fdo}), and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
-Likewise, a Linked Data approach will need to follow the same requirements to actually comply as an FDO implementation.
+A key insight from this is that simply using DOIP does not achieve many of the FDO guidelines.
+Rather the guidelines set out how a protocol like DOIP should be used to achieve FAIR Digital Object goals.
+The DOIP Endorsement \autocite{fdo-DOIPEndorsement} set out that to comply, DOIP must be used according to the set of FDO requirement documents and notes \emph{Achieving FDO compliance requires more than DOIP and full compliance is thus left to system designers}.
+Likewise, a Linked Data approach will need to follow the same FDO requirements to actually comply as an FDO implementation.
 
-From our evaluation, we can observe:
+\subsubsection*{Observations}
 
@@
 
+\subsubsection*{Observations}
+
 Based on the analysis in Table \vref{tbl:fdo-web-middleware}, we make the following observations:
@@
 
-From this evaluation we observe:
+\subsubsection*{Observations}
 
@@
 In Table \vref{tbl:eosc} we review the EOSC Interoperability Framework (EOSC IF) recommendations, and evaluate to what extent they are addressed by the principles of FDO and Linked Data or their common implementations.
+
+\subsubsection*{Observations}
 
@@
 
-\section*{Conclusion}\label{conclusion}
+\section*{Conclusion}\label{sec:conclusion}
 
@@
 We would like to acknowledge the FAIR Digital Object Forum \autocite{FAIRDigitalObjects} community and working groups, where SSR and CG are members.
+
+Views and opinions expressed in this work are those of the authors only and do not necessarily reflect those of the funded projects, FAIR Digital Object Forum, European Union nor the European Commission.
 
@@
 \printbibliography
+\clearpage
 
-\clearpage
+\appendix
+\section{An overview of upcoming FDO specifications} \label{sec:appendixfdo}
+
+\textbf{FAIR Digital Object Overview and Specifications} \autocite{fdo-Overview} is a comprehensive overview of FAIR Digital Object specifications listed below.
+It serves as a primer that introduces FDO concepts and the remaining documents.
+It is accompanied by an FDO Glossary \autocite{fdo-Glossary}.
+
+The \textbf{FDO Forum Document Standards} \autocite{fdo-DocProcessStd} documents the recommendation process within the forum, starting at \emph{Working Draft} (WD) status within the closed working group and later within the open forum, then \emph{Proposed Recommendation} (PR) published for public review, finalised as \emph{FDO Forum Recommendation} (REC) following any revisions.
+In addition, the forum may choose to \emph{endorse} existing third-party notes and specifications.
+
+The \textbf{FDO Requirement Specifications} \autocite{fdo-RequirementSpec} is an update of \autocite{boninoFAIRDigitalObject} as the foundational definition of FDO.
+This sets the criteria for classifying an digital entity as a FAIR Digital Object, allowing for multiple implementations.
+The requirements shown in Table \vref{tbl:fdo-checks} are largely equivalent, but in this specification clarified with references to other FDO documents.
+
+The \textbf{Machine actionability} \autocite{fdo-MachineActionDef} sets out to define what is meant by \emph{machine actionability} for FDOs.
+\emph{Machine readable} is defined as elements of bit-sequences defined by structural specification, \emph{machine interpretable} elements that can be identified and related with semantic artefacts, while \emph{machine actionable} are elements with a type with operations in a symbolic grammar.
+The document largely describes requirements for resolving an FDO to metadata, and how types should be related to possible operations.
+
+\textbf{Configuration Types} \autocite{fdo-ConfigurationTypes} classifies different granularities for organising FDOs in terms of PIDs, PID Records, Metadata and bit sequences, e.g.~as a single FDO or several daisy-chained FDOs.
+Different patterns used by current DOIP deployments are considered, as well as FAIR Signposting \autocite{vandesompel2015,vandesompelFAIRSignpostingProfile2022}.
+
+\textbf{PID Profiles \& Attributes} \autocite{fdo-PIDProfileAttributes} specifies that PIDs must be formally associated with a \emph{PID Profile}, a separate FDO that defines attributes required and recommended by FDOs following said profile.
+This forms the \emph{kernel attributes}, building on recommendations from RDA's \emph{PID Information Types} working group \autocite{weigelRDARecommendationPID2018}.
+This document makes a clear distinction between a minimal set of attributes needed for PID resolution and FDO navigation, which needs to be part of the \emph{PID Record} \autocite{islam_2023}, compared with a richer set of more specific attributes as part of the \emph{metadata} for an FDO, possibly represented as a separate FDO.
+
+\textbf{Kernel Attributes \& Metadata} \autocite{fdo-KernelAttributes} elaborates on categories of FDO Mandatory, FDO Optional and Community Attributes, recommending kernel attributes like \texttt{dateCreated}, \texttt{ScientificDomain}, \texttt{PersistencePolicy}, \texttt{digitalObjectMutability}, etc.
+This document expands on RDA Recommendation on PID Kernel Information \autocite{weigelRDARecommendationPID2018}.
+It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
+
+\textbf{Granularity, Versioning, Mutability} \autocite{fdo-Granularity} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
+The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
+
+\textbf{DOIP Endorsement Request} \autocite{fdo-DOIPEndorsement} is an endorsement of the DOIP v2.0 \autocite{foundationDigitalObjectInterface} specification as a potential FDO implementation, as it has been applied by several institutions \autocite{wittenburgFAIRDigitalObject2022b}.
+The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).
+
+\textbf{Upload of FDO} \autocite{fdo-FDO-Upload} illustrates the operations for uploading an FDO to a repository, what checks it should do (for instance conformance with the PID Profile, if PIDs resolve).
+ResourceSync \autocite{ResourceSyncFrameworkSpecification} is suggested as one type of service to list FDOs.
+This document highlights potential practices by repositories and their clients, without adding any particular requirements.
+
+\textbf{Typing FAIR Digital Objects} \autocite{fdo-TypingFDOs} defines what \emph{type} means for FDOs, primarily to enable machine actionability and to define an FDO's purpose.
+This document lays out requirements for how \emph{FDO Types} should themselves be specified as FDOs, and how an \emph{FDO Type Framework} allows organising and locating types.
+Operations applicable to an FDO is not predefined for a type, however operations naturally will require certain FDO types to work.
+How to define such FDO operations is not specified.
+
+\textbf{Implementation of Attributes, Types, Profiles and Registries} \autocite{fdo-ImplAttributesTypesProfiles} details how to establish FDO registries for types and FDO profiles, with their association with PID systems.
+This document suggest policies and governance structures, together with guidelines for implementations, but without mandating any explicit technology choices.
+Differences in use of attributes are examplified using FDO PIDs for scientific instruments, and the proto-FDO approach of \href{https://de.dariah.eu/}{DARIAH-DE} \autocite{schwardmannTwoExamplesHow2022}.
+
+See bibliography below for the citation per document above.
+
+%\clearpage
 \defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}

### 2023-11-16 latex/main.tex
Typos, remove section crossreferences

@@
 %% Definitions needed for pandoc's silly output
-\usepackage{hyperref}
-\hypersetup{
-  pdflang={en-GB},
-  hidelinks
-}
-\usepackage{varioref}
 
@@
 
+%\usepackage[usetoc]{titleref}
+\usepackage{hyperref}
+\hypersetup{
+  pdflang={en-GB},
+  hidelinks
+  }
+
+%\usepackage{varioref}
+\newcommand{\vref}{\ref}
+\newcommand{\vpageref}{\ref}
+
 %%% END own imports
@@
 
-\section*{Introduction}\label{sec:introduction}
+\section*{Introduction}
+\label{sec:introduction}
 
@@
 The rest of this paper is organised as follows: We begin with a background primer on FDO and Linked Data to provide a foundation for the work.
-In the Methods section, we introduce the conceptual frameworks we use for comparison.
+In the Method section, we introduce the conceptual frameworks we use for comparison.
 Subsequently, in the Results section, we systematically step through the outcomes of applying these frameworks to both FDO and Linked Data.
@@
 
-\section*{Background and related work}\label{sec:background}
+\section*{Background and related work}
+\label{sec:background}
 
@@
 
-\subsection*{FAIR Digital Object}\label{sec:fdo}
+\subsection*{FAIR Digital Object}
+\label{sec:fdo}
 
@@
 Development of the FDO concept continued within Research Data Alliance (\href{https://www.rd-alliance.org/}{RDA}) groups and EOSC projects like \href{https://www.go-fair.org/}{GO-FAIR}, concluding with a set of guidelines for implementing FDO \autocite{boninoFAIRDigitalObject}.
-The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents.
+The \href{https://fairdo.org/}{FAIR Digital Objects Forum} has since taken over the maturing of FDO through focused working groups which have currently drafted several more detailed specification documents \autocite{fdo-Specs}.
 
-\subsubsection*{FDO approaches}\label{sec:fdo-approaches}
+\subsubsection*{FDO approaches}
+\label{sec:fdo-approaches}
 
@@
 
-From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we will cover \vpageref{sec:ld}.
+From this it becomes apparent that there is a potentially large overlap between the goals and approaches of FAIR Digital Objects and Linked Data, which we will cover in section \emph{From the Semantic Web to Linked Data}.
+%\vref{sec:ld}
 
-\subsubsection*{Next steps for FDO}\label{sec:next-step-fdo}
+\subsubsection*{Next steps for FDO}
+\label{sec:next-step-fdo}
 
-The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed requirement documents \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
+The FAIR Digital Object Forum \autocite{FAIRDigitalObjects} working groups have prepared detailed \href{https://fairdo.org/specifications/}{requirement documents} \autocite{fdo-Specs} setting out the path for realising FDOs, named \emph{FDO Recommendations}.
 As of 2023-06-17, most of these documents are open for public review, while some are still in draft stages for internal review.
@@
 
-\subsection*{From the Semantic Web to Linked Data}\label{sec:ld}
+\subsection*{From the Semantic Web to Linked Data}
+\addcontentsline{toc}{subsection}{From the Semantic Web to Linked Data}
+\label{sec:ld}
 
@@
 
-\subsubsection*{A brief history of the Semantic Web}\label{sec:semweb}
+\subsubsection*{A brief history of the Semantic Web}
+\label{sec:semweb}
 
@@
 
-\subsubsection*{Linked Data: Rebuilding the Web of Data}\label{sec:ld-web}
+\subsubsection*{Linked Data: Rebuilding the Web of Data}
+\label{sec:ld-web}
 
@@
 
-\section*{Methods}\label{sec:methods}
+\section*{Method}
+\label{sec:methods}
 
@@
 
-\section*{Results}\label{sec:results}
+\section*{Results}
+\label{sec:results}
 
-\subsection*{Considering FDO/Web as interoperability framework for Fast Data}\label{sec:interoperability-compare}
+\subsection*{Considering FDO/Web as interoperability framework for Fast Data}
+\label{sec:interoperability-compare}
 
@@
 
-\subsubsection*{Mapping of Metamodel concepts}\label{mapping-of-metamodel-concepts}
+\subsubsection*{Mapping of Metamodel concepts}
+\label{mapping-of-metamodel-concepts}
 
@@
 From this mapping\footnote{A SKOS mapping \autocite{w3-skos-primer} is provided as part of the \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} \autocite{soilandreyes2023}.} we can identify the conceptual similarities between DOIP and HTTP, often with common terminology.
-Notable are that neither DOIP or HTTP have strong support for transactions (explored further \vpageref{sec:middleware}), as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
+Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \emph{Comparing FDO and Web as middleware infrastructures}),
+%\vpageref{sec:middleware}
+as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
 
@@
 
-\subsection*{Assessing FDO implementations}\label{sec:doip-fdo-compare}
+\subsection*{Assessing FDO implementations}
+\label{sec:doip-fdo-compare}
 
@@
 In Table \vref{tbl:fdo-checks}, we evaluate completeness of the guidelines in two current FDO realisations: 1) DOIPv2 \autocite{foundationDigitalObjectInterface} and 2) Linked Data Platform \autocite{w3-ldp}, as proposed by \textcite{FDOFramework}.
-We provide our analysis of each realization with respect to the FDO Guidline and also provide suggestions for that realization to meet the given guideline.
+We provide our analysis of each realisation with respect to the FDO Guideline and also provide suggestions for that realisation to meet the given guideline.
 
@@
 
-\subsection*{Comparing FDO and Web as middleware infrastructures}\label{sec:middleware}
+\subsection*{Comparing FDO and Web as middleware infrastructures}
+\label{sec:middleware}
 
@@
 
-\subsection*{Assessing FDO against FAIR}\label{sec:fair-compare}
+\subsection*{Assessing FDO against FAIR}
+\label{sec:fair-compare}
 
@@
 
-\subsection*{EOSC Interoperability Framework}\label{sec:eosc-interoperability-framework}
+\subsection*{EOSC Interoperability Framework}
+\label{sec:eosc-interoperability-framework}
 
@@
 
-As covered in our introduction\vref{sec:introduction}, EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
+As covered in our introduction,
+%\vref{sec:introduction}, 
+EOSC proposes FAIR Digital Objects as a way to improve interoperability, for instance invoked by scientific workflows, carried by metadata frameworks and semantic artefacts.
 Therefore we here find it important to summarize how FDO and Linked Data can help satisfy the EOSC requirements.
@@
 
-\section*{Discussion}\label{sec:discussion}
+\section*{Discussion}
+\label{sec:discussion}
 
@@
 
-Having considered FDO and the Web architecture as interoperability frameworks (\vref*{sec:interoperability-compare}), we observe that neither are magic bullets, but each bring different aspects of interoperability.
+Having considered FDO and the Web architecture as interoperability frameworks 
+%(\vref*{sec:interoperability-compare})
+we observe that neither are magic bullets, but each bring different aspects of interoperability.
 The Web comes with a large degree of flexibility and openness, however this means interoperability can suffer as services have different APIs and data models, although with common patterns.
@@
 
-When evaluating FDO implementations against the FDO guidelines (\vref*{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used.
+When evaluating FDO implementations against the FDO guidelines 
+%(\vref*{sec:doip-fdo-compare}) 
+we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used.
 Achieving fully interoperable FAIR Digital Objects would require further convergence on implementation practices, and it is not given that this needs to diverge from the established Web architecture.
@@
 
-Considering this, by comparing FDO and Web as middleware (\vref*{sec:middleware}) we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g.
+Considering this, by comparing FDO and Web as middleware 
+%(\vref*{sec:middleware}) 
+we saw that programmatic access to digital objects, a core promise of FDO, is not particularly improved by the use of the protocol DOIP as compared to HTTP, e.g.
 lack of concurrency and transparancy.
@@
 
-By assessing FDO against the FAIR principles (\vref*{sec:fair-compare}) we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
+By assessing FDO against the FAIR principles 
+%(\vref*{sec:fair-compare}) 
+we found that both FDO implementations are underspecified in several aspects (licences, provenance, data references, data vocabularies, metadata persistence).
 While there are implementations of each of these in general Linked Data examples, there is no single set of implementation guides that fully realizes the FAIR principles.
@@
 
-EOSC has been largely supportive of FDO, FAIR and related services.
-By contrasting the EOSC Interoperability Framework (\vref*{sec:eosc-interoperability-framework}) with FDO, we found that there are important dimensions that are not solved at a technical level, but through organization collaboration, legal requirements and building community practices.
+EOSC has been largely supportive of FDO, FAIR and related services. By contrasting the EOSC Interoperability Framework 
+%(\vref*{sec:eosc-interoperability-framework})
+with FDO, we found that there are important dimensions that are not solved at a technical level, but through organization collaboration, legal requirements and building community practices.
 FDO recommendations highlight community aspects, but at the same time the largest FAIR communities in many science domains are already producing and consuming Linked Data.
@@
 
-\subsection*{What does FDO mean for Linked Data?}\label{sec:what-does-it-mean-for-linked-data}
+\subsection*{What does FDO mean for Linked Data?}
+\label{sec:what-does-it-mean-for-linked-data}
 
@@
 
-\section*{Conclusion}\label{sec:conclusion}
+\section*{Conclusion}
+\label{sec:conclusion}
 
@@
 \appendix
-\section{An overview of upcoming FDO specifications} \label{sec:appendixfdo}
+\section{An overview of upcoming FDO specifications}
+\label{sec:appendixfdo}
 
@@
 %\clearpage
-\defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}
+\defbibheading{fdobibliography}{\subsection*{FDO Specifications}
+\label{sec:fdo-bibliography}}
 \printshorthands[heading=fdobibliography]

### 2023-11-22 latex/main.tex
cite preprint, thesis and RO-Crate

@@
 
-\begin{small}
-This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT);  HORIZON-CL4-2021-HUMAN-01-01 101070305 (ENEXA)  and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).
-\end{small}
+%\begin{small}
+%This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT);  HORIZON-CL4-2021-HUMAN-01-01 101070305 (ENEXA)  and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).
+%\end{small}
 
@@
 Views and opinions expressed in this work are those of the authors only and do not necessarily reflect those of the funded projects, FAIR Digital Object Forum, European Union nor the European Commission.
+
+An earlier version of this article was published as a preprint \autocite{soilandreyes2023evaluating}, which is embedded in the corresponding RO-Crate for this article \autocite{soilandreyes2023} and the PhD thesis \autocite{soilandreyesthesis}.
 
@@
 This document expands on RDA Recommendation on PID Kernel Information \autocite{weigelRDARecommendationPID2018}.
-It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.
-
-\textbf{Granularity, Versioning, Mutability} \autocite{fdo-Granularity} considers how granularity decisions for forming FDOs must be agreed by different communities depending on their pragmatic usage requirements.
-The affect on versioning, mutability and changes to PIDs are considered, based on use cases and existing PID practices.
-
-\textbf{DOIP Endorsement Request} \autocite{fdo-DOIPEndorsement} is an endorsement of the DOIP v2.0 \autocite{foundationDigitalObjectInterface} specification as a potential FDO implementation, as it has been applied by several institutions \autocite{wittenburgFAIRDigitalObject2022b}.
+It is worth noting that both documents are relatively abstract and do not establish PIDs or namespaces for the kernel attributes.zenodo.8075229has been applied by several institutions \autocite{wittenburgFAIRDigitalObject2022b}.
 The document proposes that DOIP shall be assessed for completeness against FDO -- in this initial draft this is justified as \emph{``we can state that DOIP is compliant with the FDO specification documents in process''} (the documents listed above).

### 2023-11-22 latex/main.tex
bibliography fixes as requested by peerj
changed phd title again :)

@@
 \usepackage[autostyle]{csquotes}
-\usepackage[backend=biber,alldates=iso,labeldate=year,style=authoryear,maxnames=2,maxbibnames=7,maxcitenames=2]{biblatex}
+\usepackage[backend=biber,alldates=iso,labeldate=year,style=authoryear,maxnames=2,maxbibnames=999,maxcitenames=2]{biblatex}
 \addbibresource{bibliography.bib}
@@
 Notable are that neither DOIP or HTTP have strong support for transactions (explored further in section \emph{Comparing FDO and Web as middleware infrastructures}),
-%\vpageref{sec:middleware}
+%\vref{sec:middleware}
 as well that HTTP has poor direct support for processes, as the Web is primarily stateless by design.
@@
 \item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in Table \vref{RDA-A2-01M}).
+  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in Table \vref{RDA-A2-01M} introduced by section \emph{Assessing FDO against FAIR}).
+%\vref{sec:fair-compare}
 \item
@@
 
-Here we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform \autocite{FDOFramework} comply with the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
+In Table \vref{tbl:fair-data-maturity-model} we evaluate to what extent the FDO guidelines and its implementation with DOIP and Linked Data Platform \autocite{FDOFramework} comply with the FAIR principles \autocite{wilkinsonFAIRGuidingPrinciples2016e}.
 Here we've used the RDA's FAIR Data Maturity Model \autocite{groupFAIRDataMaturity2020} as it has decomposed the FAIR principles to a structured list of FAIR indicators \autocite{bahimFAIRDataMaturity2020a}, importantly considering \emph{Data} and \emph{Metadata} separately.
-In our interpretation for Table \vref{tbl:fair-data-maturity-model} we have for simplicity chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remaining JSON or RDF structures always considered as metadata.
+In our interpretations we have, for simplicity, chosen to interpret ``data'' in FDOs as the associated bytestream of arbitrary formats, with remaining JSON or RDF structures always considered as metadata.
 

### 2023-11-30 latex/main.tex
PeerJ Editorials

@@
 
-An \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} is archived at \url{https://doi.org/10.5281/zenodo.8075229}
+An \href{https://w3id.org/ro/doi/10.5281/zenodo.8075229}{RO-Crate for this article} is archived at \url{https://doi.org/10.5281/zenodo.8075229}. 
 \end{abstract}
@@
 \thispagestyle{empty}
+
+\emph{An earlier version of this article was published as a preprint \autocite{soilandreyes2023evaluating}, which is embedded in the corresponding RO-Crate for this article \autocite{soilandreyes2023} and the PhD thesis \autocite{soilandreyesthesis}}.
 
@@
 \item
-  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (see RDA-A2-01M in Table \vref{RDA-A2-01M} introduced by section \emph{Assessing FDO against FAIR}).
+  Tombstones for deleted resources are not well supported, nor specified, for either approach, although the continued availability of metadata when data is removed is a requirement for FAIR principles (RDA-A2-01M).
 %\vref{sec:fair-compare}
@@
 
-An earlier version of this article was published as a preprint \autocite{soilandreyes2023evaluating}, which is embedded in the corresponding RO-Crate for this article \autocite{soilandreyes2023} and the PhD thesis \autocite{soilandreyesthesis}.
-
 \section*{Author contributions}
